### 2023-08-20 Paper_code.Rmd
Added all code for the paper and data related to it

---
title: "Code used to generate paper results"
author: "Julie Devis"
date: "2023-08-28"
output: 
  html_document:
    theme: cosmo
    df_print : paged
    highlight : default
    toc: yes
    number_sections : TRUE
    toc_depth : 4
---

# Loading data and packages

```{r library, message = FALSE}
library(readxl)
library(readr)
library(CTexploreR)
library(Vennerable)
library(biomaRt)
library(tidyverse)
library(SummarizedExperiment)
library(factoextra)
library(UpSetR)
library(ComplexHeatmap)
library(circlize)
library(SingleCellExperiment)
library(org.Hs.eg.db)
library(clusterProfiler)
library(DOSE)
library(tibble)
library(BiocParallel)
library(patchwork)
library(Biostrings)
```

Gene names/synonyms required for databases cleaning

```{r ensembl_synonym, warning=FALSE, message=FALSE}
ensembl <- biomaRt::useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
attributes_vector <- c("ensembl_gene_id", "external_gene_name",
                       "external_synonym", "gene_biotype",
                       "chromosome_name", "band", "start_position", "end_position",
                       "strand")
ensembl_gene_synonym <- as_tibble(getBM(attributes = attributes_vector, mart = ensembl))

ensembl_gene_synonym <- ensembl_gene_synonym %>%
  mutate(external_synonym = sub(pattern = "ORF", external_synonym, 
                                replacement = "orf"))

attributes_vector <- c("ensembl_gene_id", "external_gene_name")
ensembl_gene_names <- as_tibble(getBM(attributes = attributes_vector, mart = ensembl))
```

```{r load_data, warning=FALSE, message=FALSE}
GTEX_data <- CTdata::GTEX_data()
normal_tissues_multimapping_data <- CTdata::normal_tissues_multimapping_data()
CCLE_data <- CTdata::CCLE_data()
TCGA_TPM <- CTdata::TCGA_TPM()
DAC_treated_cells_multimapping <- CTdata::DAC_treated_cells_multimapping()
load("../CTdata/eh_data/testis_sce.rda") 
load("../CTdata/eh_data/scRNAseq_HPA.rda")
load("../CTdata/eh_data/CT_mean_methylation_in_tissues.rda")
```

Common figures parameters

```{r fig_param}
legends_param <- list(
  labels_gp = gpar(col = "black", fontsize = 6),
  title_gp = gpar(col = "black", fontsize = 6),
  row_names_gp = gpar(fontsize = 4),
  annotation_name_side = "left")

legend_colors <- c("#5E4FA2", "#3288BD", "#66C2A5", "#ABDDA4", "#E6F598",
                   "#FFFFBF", "#FEE08B", "#FDAE61", "#F46D43", "#D53E4F",
                   "#9E0142")
chr_colors <- c("X-linked" = "deeppink", "Not X" = "royalblue1")

meth_colors <- c("Methylation" = "lightgreen", "Not methylation" = "gray")
```

# Database comparison

## Lists cleaning

CT lists from other databases have been checked (using GTEx and our 
`GTEx_expression()` funtion and GeneCards) in order to remove duplicated
gene names or deprecated ones and allow comparison between databases.

### CTdatabase

Online list copied in a csv file, several lists exist so we combined them.

We checked gene names that were a concetanation of two genes (choice using
biomaRt synonyms to get the official one), checked which ones had the right names,
removed duplicated genes, verified lost genes and added back those that should 
be there.

```{r CTdatabase_cleaning, message = FALSE}
CTdatabase <- read_delim("data/CTdatabase1.csv", delim = ";", 
                         escape_double = FALSE, trim_ws = TRUE)
colnames(CTdatabase) <- c("Family", "Gene_Name", "Chromosomal_localization",
                          "CT_identifier")
CTdatabase_bis <- read_csv2("data/CTdatabase2.csv")
CTdatabase <- left_join(CTdatabase, CTdatabase_bis, 
                        by = c("Gene_Name" = "Gene_Symbol"))

CTdatabase_single <- CTdatabase %>%
  mutate(Gene_Name = sub(pattern = "/.*$", Gene_Name, replacement = ""))
CTdatabase_single <- CTdatabase_single %>%
  mutate(Gene_Name = sub(pattern = ",.*$", Gene_Name, replacement = ""))

CTdatabase_official_names <- 
  unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, 
                       external_gene_name)) %>%
  filter(external_gene_name %in% CTdatabase_single$Gene_Name) %>%
  mutate(Gene_Name = external_gene_name) %>%
  mutate(external_synonym = NA)
CTdatabase_synonym <- 
  ensembl_gene_synonym %>%
  filter(external_synonym %in% CTdatabase_single$Gene_Name) %>%
  mutate(Gene_Name = external_synonym) %>%
  dplyr::select(ensembl_gene_id, external_gene_name, Gene_Name, external_synonym)
CTdatabase_cleaned <- 
  rbind(CTdatabase_official_names, CTdatabase_synonym) %>% 
  left_join(CTdatabase_single)

duplicated_genes <- CTdatabase_cleaned$Gene_Name[duplicated(CTdatabase_cleaned$Gene_Name)]
bad_ids <- ensembl_gene_synonym %>%
  filter(external_gene_name %in% duplicated_genes | external_synonym %in% duplicated_genes) %>%
  filter(chromosome_name %in% grep(pattern = "H", x = chromosome_name, value = TRUE)) %>%
  pull(ensembl_gene_id)
CTdatabase_cleaned <- CTdatabase_cleaned %>%
  dplyr::filter(!ensembl_gene_id %in% bad_ids)
CTdatabase_cleaned <- CTdatabase_cleaned %>%
  filter(!ensembl_gene_id == "ENSG00000052126")
CTdatabase_cleaned <- CTdatabase_cleaned %>% 
  filter(!(ensembl_gene_id == "ENSG00000183305" & Gene_Name == "MAGEA2"))
CTdatabase_cleaned <- CTdatabase_cleaned %>% 
  filter(!ensembl_gene_id == "ENSG00000204648")
CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "CSAG3B")
CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "CSAG2", "external_synonym"] <- "CSAG3B"
CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "CT45A4")
CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "CT45A3", "external_synonym"] <- "CT45A4"
CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "LAGE-1b")
CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "CTAG2", "external_synonym"] <- "LAGE-1b"
CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "CT16.2")
CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "PAGE5", "external_synonym"] <- "CT16.2"
CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "SPANXB2")
CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "SPANXB1", "external_synonym"] <- "SPANXB2"
CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "SPANXE")
CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "SPANXD", "external_synonym"] <- "SPANXE"
CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "XAGE1C")
CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "XAGE1D")
CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "XAGE1E")
CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "XAGE2B")
CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "XAGE2", "external_synonym"] <- "XAGE2B"
CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "CTAGE-2")
CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "CTAGE1", "external_synonym"] <- "CTAGE-2"

CTdatabase_cleaned <- ensembl_gene_synonym %>%
  mutate(Gene_Name = external_synonym) %>%
  filter(external_synonym == "CXorf61") %>%
  dplyr::select(ensembl_gene_id, external_gene_name, Gene_Name, external_synonym) %>%
  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "Cxorf61", 
                    c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
  rbind(CTdatabase_cleaned)
CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name)) %>%
  filter(external_gene_name == "CCNA1") %>%
  mutate(Gene_Name = external_gene_name) %>%
  mutate(external_synonym = NA) %>% 
  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "cyclin A1", 
                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
  rbind(CTdatabase_cleaned)
CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
  filter(external_gene_name == "GOLGA6L2") %>%
  filter(ensembl_gene_id == "ENSG00000174450") %>%
  mutate(Gene_Name = external_gene_name) %>%
  mutate(external_synonym = NA) %>% 
  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "GOLGAGL2 FA", 
                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
  rbind(CTdatabase_cleaned)
CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
  filter(external_gene_name == "LYPD6B") %>%
  mutate(Gene_Name = external_gene_name) %>%
  mutate(external_synonym = NA) %>% 
  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "LOC130576", 
                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
  rbind(CTdatabase_cleaned)
CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
  filter(external_gene_name == "CT62") %>%
  mutate(Gene_Name = external_gene_name) %>%
  mutate(external_synonym = NA) %>% 
  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "LOC196993", 
                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
  rbind(CTdatabase_cleaned)
CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
  filter(external_gene_name == "CT75") %>%
  filter(ensembl_gene_id == "ENSG00000291155") %>%
  mutate(Gene_Name = external_gene_name) %>%
  mutate(external_synonym = NA) %>% 
  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "LOC440934", 
                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
  rbind(CTdatabase_cleaned)
CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
  filter(external_gene_name == "LINC01192") %>%
  mutate(Gene_Name = external_gene_name) %>%
  mutate(external_synonym = NA) %>% 
  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "LOC647107", 
                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
  rbind(CTdatabase_cleaned)
CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
  filter(external_gene_name == "TSPY1") %>%
  mutate(Gene_Name = external_gene_name) %>%
  mutate(external_synonym = NA) %>% 
  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "LOC728137", 
                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
  rbind(CTdatabase_cleaned)
CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
  filter(external_gene_name == "SSX2B") %>%
  mutate(Gene_Name = external_gene_name) %>%
  mutate(external_synonym = NA) %>% 
  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "SSX2b", 
                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
  rbind(CTdatabase_cleaned)
```

### Jamin's list

Excel file coming from supplemental data.

```{r Jamin_clean, message = FALSE}
Jamin_core_CT <- read_excel("data/Jamin_core_CT.xlsx")
Jamin_core_CT[Jamin_core_CT$Gene == "KIAA1211", "Gene"] <- "CRACD"
Jamin_core_CT[Jamin_core_CT$Gene == "CXorf67", "Gene"] <- "EZHIP"
```

### Wang's CTatlas

Excel file coming from supplemental data.

```{r Wang_cleaning, message = FALSE}
Wang_CT <- read_excel("data/Wang_Suppl_Data_3.xlsx", 
    sheet = "Supplementary Data 3B", skip = 1)
colnames(Wang_CT)[1] <- "ensembl_gene_id"

Wang_CT <- ensembl_gene_names %>% 
  filter(ensembl_gene_id %in% Wang_CT$ensembl_gene_id) %>%
  right_join(Wang_CT)

Wang_CT[Wang_CT$ensembl_gene_id == "ENSG00000181013", "external_gene_name"] <- "C17orf47"
Wang_CT[Wang_CT$ensembl_gene_id == "ENSG00000204293", "external_gene_name"] <- "OR8B2"
Wang_CT[Wang_CT$external_gene_name == "", "external_gene_name"] <- "RNASE11"
Wang_CT[Wang_CT$external_gene_name == "CHCT1", "external_gene_name"] <- "C17orf64"
Wang_CT[Wang_CT$external_gene_name == "PRSS40A", "external_gene_name"] <- "TISP43"
Wang_CT[Wang_CT$external_gene_name == "TEX56P", "external_gene_name"] <- "C6orf201"
Wang_CT[Wang_CT$external_gene_name == "SLC25A51P4", "external_gene_name"] <- "RP11-113D6.10"
Wang_CT[Wang_CT$external_gene_name == "TCP10L3", "external_gene_name"] <- "TCP10"
Wang_CT[Wang_CT$external_gene_name == "SCAND3", "external_gene_name"] <- "ZBED9"
```

### Carter's list

```{r Carter_clean, message = FALSE}
Carter_CT_list <- read_table("data/Carter_CT_list.txt", skip = 1)
Carter_CT <- Carter_CT_list[Carter_CT_list$CT_Expression,]

Carter_CT[Carter_CT$Gene == "ENSG00000261649", "Gene_Name"] <- "GOLGA6L7"
Carter_CT[Carter_CT$Gene == "ENSG00000239620", "Gene_Name"] <- "PRR20G"
Carter_CT[Carter_CT$Gene == "ENSG00000168148", "Gene_Name"] <- "H3-4"
Carter_CT[Carter_CT$Gene == "ENSG00000204296", "Gene_Name"] <- "TSBP1"
Carter_CT[Carter_CT$Gene == "ENSG00000180219", "Gene_Name"] <- "GARIN6"
Carter_CT[Carter_CT$Gene == "ENSG00000172717", "Gene_Name"] <- "GARIN2"
Carter_CT[Carter_CT$Gene == "ENSG00000174015", "Gene_Name"] <- "CBY2"
Carter_CT[Carter_CT$Gene == "ENSG00000224960", "Gene_Name"] <- "PPP4R3C"
```

## CTexploreR data for selection pipeline

To characterise the differences between our database and other, we need the 
category we created in CTexploreR.

Hereunder is what we used for our selection pipeline (coming from 
`make_CT_list.R` in `CTdata`), mainly how we created the data.

```{r make_CT, message = FALSE, warning = FALSE}
# GTEX data with the tissue specificity category determined
all_genes <- as_tibble(rowData(GTEX_data), rownames = "ensembl_gene_id")
all_genes$TPM_testis <- assay(GTEX_data)[, "Testis"]

# Add multimapping_analysis assessing testis-specificity of genes "lowly_expressed"

all_genes <- all_genes %>%
  left_join(as_tibble(rowData(normal_tissues_multimapping_data), 
                      rownames = "ensembl_gene_id"))

# Summarise both specificity

all_genes <- all_genes %>%
  mutate(testis_specificity = case_when(
    GTEX_category == "testis_specific" |
      multimapping_analysis == "testis_specific" ~ "testis_specific",
    GTEX_category == "testis_preferential" ~ "testis_preferential"))

# Add testis scRNA seq highest expressed cell type

all_genes <- all_genes %>%
  left_join(as_tibble(rowData(testis_sce)) %>%
              dplyr::select(external_gene_name, testis_cell_type))

# Add higher in somatic scRNAseq data of normal tissues from the Human Protein Atlas

all_genes <- all_genes %>%
  left_join(as_tibble(rowData(scRNAseq_HPA), rownames = "ensembl_gene_id") %>%
              dplyr::select(ensembl_gene_id, external_gene_name,
                            Higher_in_somatic_cell_type))

# CCLE database analysis added

all_genes <- all_genes %>%
  left_join(as_tibble(rowData(CCLE_data), rownames = "ensembl_gene_id"))
all_genes[is.na(all_genes$CCLE_category), "CCLE_category"] <- "not_in_CCLE"

# TCGA database analysis added

all_genes <- all_genes %>%
  left_join(as_tibble(rowData(TCGA_TPM), rownames = "ensembl_gene_id") %>%
              dplyr::select(ensembl_gene_id, external_gene_name, percent_pos_tum,
                            percent_neg_tum, max_TPM_in_TCGA, TCGA_category))

all_genes[all_genes$lowly_expressed_in_GTEX == TRUE &
            all_genes$multimapping_analysis == "testis_specific",
          "TCGA_category"] <- "multimapping_issue"

# Add DAC analysis
induced <- as_tibble(rowData(DAC_treated_cells_multimapping), 
                     rownames = "ensembl_gene_id") %>%
  filter(induced == TRUE) %>%
  pull(external_gene_name)

all_genes <- all_genes %>%
  mutate(DAC_induced = case_when(external_gene_name %in% induced ~ TRUE,
                                 !external_gene_name %in% induced ~ FALSE))

# Add the most biologically relevant transcript (canonical and coordinates)

ensembl <- biomaRt::useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
attributes_vector <- c("ensembl_gene_id",
                       "external_gene_name",
                       "ensembl_transcript_id",
                       "external_transcript_name",
                       "chromosome_name",
                       "strand",
                       "transcript_start",
                       "transcript_end",
                       "transcription_start_site",
                       "transcript_length",
                       "transcript_biotype",
                       "transcript_is_canonical")
transcripts_infos <- as_tibble(biomaRt::getBM(attributes = attributes_vector,
                                              mart = ensembl))
canonical_transcripts <- transcripts_infos %>%
  filter(transcript_is_canonical == 1) %>%
  filter(chromosome_name %in% c(1:22, "X", "Y", "MT")) %>%
  filter(transcript_biotype == "protein_coding" | transcript_biotype == "lncRNA")
all_genes <- all_genes %>%
  left_join(canonical_transcripts %>%
              dplyr::select(ensembl_gene_id,
                            external_transcript_name, ensembl_transcript_id,
                            chromosome_name, strand, transcript_start,
                            transcript_end, transcription_start_site,
                            transcript_length, transcript_biotype))

```

From there, we filtered based on the testis_specificity ("testis_specific" or 
"testis_preferential"), testis_cell_type (not "Macrophage", "Endothelial",
"Myoid", "Sertoli", "Leydig"), Higher_in_somatic_cell_type (TRUE), 
CCLE_category ("activated") and TCGA_category ("activated" or 
"multimapping_issue") to have our CT genes. Then the most relevant 
transcript was validated manually (and check that there is a transcript 
activated in testis and the same is activated in tumor).

## CTexploreR VS CTdatabase

```{r CTdatabase_ours}
CTdatabase_ours <- Venn(list(CTdatabase = CTdatabase_cleaned$external_gene_name,
                             CTexploreR = CT_genes$external_gene_name))
plot(CTdatabase_ours)
```

## CTexploreR VS omics databases

```{r upset_plot}
core_ours <- Venn(list(Jamin = Jamin_core_CT$Gene, 
                       CTexploreR = CT_genes$external_gene_name))

Wang_ours <- Venn(list(Wang = Wang_CT$external_gene_name, 
                       CTexploreR = CT_genes$external_gene_name))

Carter_ours <- Venn(list(Carter_CT = Carter_CT$Gene_Name, 
                         CTexploreR = CT_genes$external_gene_name))

gene_list <- list(Jamin = Jamin_core_CT$Gene, 
                  CTdatabase = CTdatabase_cleaned$external_gene_name, 
                  CTatlas = Wang_CT$external_gene_name, 
                  CTexploreR = CT_genes$external_gene_name, 
                  Carter = Carter_CT$Gene_Name)

upset_omics <- fromList(gene_list[-2])
upset(upset_omics)
```

## Characterisation of differences with all databases

```{r all_databases}
common <- unique(c(core_ours@IntersectionSets[["11"]], 
                   CTdatabase_ours@IntersectionSets[["11"]], 
                   Wang_ours@IntersectionSets[["11"]], 
                   Carter_ours@IntersectionSets[["11"]]))

length(common)

length(common)/dim(CT_genes)[1] * 100

lost_list <- unique(c(core_ours@IntersectionSets[["10"]], 
                 CTdatabase_ours@IntersectionSets[["10"]],
                 Wang_ours@IntersectionSets[["10"]],
                 Carter_ours@IntersectionSets[["10"]]))

lost <- all_genes %>%
  filter(external_gene_name %in% lost_list)

not_specific <- filter(lost, is.na(testis_specificity))

somatic_testis <- filter(lost, testis_cell_type %in% c( "Macrophage", 
                                                        "Endothelial", "Myoid",
                                                        "Sertoli", "Leydig"))

somatic_tissue <- filter(lost, Higher_in_somatic_cell_type == TRUE)

not_TCGA_activated <- filter(lost, TCGA_category != "activated" & 
                               TCGA_category != "multimapping_issue")

not_CCLE_activated <- filter(lost, CCLE_category  != "activated")

transcript_prob <- lost %>% 
  filter(testis_specificity == "testis_specific" |
           testis_specificity == "testis_preferential") %>%
  filter(TCGA_category == "activated" | TCGA_category == "multimapping_issue") %>% 
  filter(CCLE_category == "activated") %>% 
  dim()
```

We have lost `r length(lost_list)` genes in total. Among them, 
`r dim(not_specific)[1]/length(lost_list)*100`% are not considered testis 
specific, `r dim(somatic_testis)[1]/length(lost_list)*100`% are expressed in 
testis somatic cells, `r dim(somatic_tissue)[1]/length(lost_list)*100`% are 
expressed in somatic cells, `r dim(not_TCGA_activated)[1]/length(lost_list)*100`% 
are not activated in TCGA samples, 
`r dim(not_CCLE_activated)[1]/length(lost_list)*100`% are not activated in CCLE 
cell lines and `r transcript_prob[1]/length(lost_list)*100`% is lost due to 
transcripts problems.

# CT genes characteristics

```{r basic}
table(CT_genes$testis_specificity)
table(CT_genes$transcript_biotype)
```

Most genes are testis specific 
(`r table(CT_genes$testis_specificity)[[2]]/dim(CT_genes)[1]*100`%).
Most genes are mainly protein coding genes 
(`r table(CT_genes$transcript_biotype)[[3]]/dim(CT_genes)[1]*100`%).

## X chromosome and regulated by methylation

In CTexploreR, genes have been characterised as regulated by methylation or not.

```{r X_meth}
table(CT_genes$X_linked)
table(CT_genes$regulated_by_methylation)
table(CT_genes$X_linked, CT_genes$regulated_by_methylation)
```

Genes are enriched on the X chromosome 
(`r table(CT_genes$X_linked)[[2]]/dim(CT_genes)[1]*100`%). Also, 
`r table(CT_genes$regulated_by_methylation)[[2]]` genes have been flagged as 
regulated by methylation 
(`r table(CT_genes$regulated_by_methylation)[[2]]/dim(CT_genes)[1]*100`%). 
It is interesting to study the link between these two characteristics

On the chromosome X, there is a clear enrichment of CT genes regulated by 
methylation (83/93 chrX genes or 83/160 genes regulated by methylation).

Let's check that with a statistical test, I here want to see if, on there is an 
enrichment of genes regulated by methylation on the X chromosome. I need to do a
Pearson Chi square test (to know if observed proportion differ from expected 
proportion). It is a statistical method to determine if two categorical variables
have a significant correlation between them. I can directly put a matrix 
(my table) in the function

```{r chisq}
chisq.test(table(CT_genes$X_linked, CT_genes$regulated_by_methylation))
```

```{r X_meth_fig}
CT_genes$chr_factor <- factor(CT_genes$chr,
                       levels = c("1", "2", "3", "4", "5", "6", "7", "8", "9",
                                  "10", "11", "12", "13", "14", "15", "16", "17", 
                                  "18", "19", "20", "21", "22", "X", "Y"))

CT_genes %>% 
  mutate(regulated_by_methylation = ifelse(regulated_by_methylation,
                                           "Regulated by methylation",
                                           "Not regulated by methylation")) %>% 
  ggplot(aes(x = chr_factor, fill = testis_specificity, color = X_linked)) +
  geom_bar(stat = 'count') +
  scale_fill_manual(values = c("goldenrod1", "mediumpurple1")) +
  scale_color_manual(values = c("royalblue1", "deeppink"), guide = "none") +
  facet_wrap(~ regulated_by_methylation, ncol = 2) +
  theme_bw() +
  xlab("Chromosome") +
  ylab("Number of genes") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "top",
        axis.title = element_text(size = 10, color = "gray10"))
```

There is indeed a significative link between regulation by methylation and being
on the X chromosome. There is thus an enrichment of CT genes regulated by 
methylation on the X chromosome (and inversely).

## Tumour and methylation

Using CTexploreR functions, we can explore all CT genes or focus on some 
potential targets.

### All CT

For these heatmaps, the code comes from the function but has been copies to add 
some annotations.

```{r common_annot}
chr_mat <- as.matrix(CT_genes$X_linked)
chr_mat <- ifelse(chr_mat == TRUE, "X-linked", "Not X")
rownames(chr_mat) <- CT_genes$external_gene_name
row_ha_chr <- rowAnnotation(chr_factor = chr_mat,
                            annotation_legend_param = legends_param,
                            simple_anno_size = unit(0.5, "cm"),
                            col = list(chr_factor = chr_colors),
                            annotation_name_gp = gpar(fontsize = 8),
                            annotation_name_side = "top")

regulation_mat <- as.matrix(CT_genes$regulated_by_methylation)
regulation_mat <- ifelse(regulation_mat == TRUE, "Methylation",
                         "Not methylation")
rownames(regulation_mat) <- CT_genes$external_gene_name

row_ha_reg <- rowAnnotation(regulation = regulation_mat,
                            annotation_legend_param = legends_param,
                            simple_anno_size = unit(0.5, "cm"),
                            col = list(regulation = meth_colors),
                            annotation_name_gp = gpar(fontsize = 8),
                            annotation_name_side = "top")

left_annot <- c(row_ha_chr, row_ha_reg, gap = unit(1, "mm"))

split <- data.frame(CT_genes$regulated_by_methylation, CT_genes$X_linked)
```

```{r TCGA_heatmap, message = FALSE}
database <- TCGA_TPM
database$tumor <- sub(pattern = 'TCGA-', x = database$project_id, '')
database$type <- "Tumor"
database$type[database$shortLetterCode == "NT"] <- "Peritumoral"
database <- database[, order(database$tumor, database$type)]

genes <- CT_genes$external_gene_name
database <- database[rowData(database)$external_gene_name %in% genes, ]
database <- database[match(genes, rowData(database)$external_gene_name), ]

column_ha_type <- HeatmapAnnotation(
  Type = database$type,
  border = TRUE,
  col = list(Type = c("Tumor" = "salmon",
                      "Peritumoral" = "mediumseagreen")),
  annotation_name_gp = gpar(fontsize = 8),
  annotation_legend_param = legends_param)

column_ha_tumor <- HeatmapAnnotation(
  Tumor = database$tumor,
  border = TRUE,
  col = list(Tumor = c("BRCA" = "midnightblue", "COAD" = "darkorchid2",
                 "ESCA" = "gold", "HNSC" = "deeppink2",
                 "LUAD" = "seagreen", "LUSC" = "seagreen3",
                 "SKCM" = "red3")),
  annotation_name_gp = gpar(fontsize = 8),
  annotation_legend_param = legends_param)

split_by <- factor(database$tumor)
col_annot <- column_ha_tumor

mat <- log1p(assay(database))
rownames(mat) <- rowData(database)$external_gene_name

Heatmap(mat[, , drop = FALSE],
        name = "logTPM",
        column_title = paste0("Expression in TCGA samples (all)"),
        column_split = split_by,
        row_split = split,
        row_title_gp = gpar(fontsize = 0),
        col = colorRamp2(seq(0, max(mat), length = 11),
                         legend_colors),
        clustering_method_rows = "ward.D",
        clustering_method_columns = "ward.D",
        cluster_rows = TRUE,
        show_column_names = FALSE,
        cluster_columns = TRUE,
        show_column_dend = FALSE,
        show_row_dend = FALSE,
        row_names_gp = gpar(fontsize = 4),
        heatmap_legend_param = legends_param,
        top_annotation = col_annot,
        left_annotation = left_annot)

```

```{r CCLE_heatmap}
database <- CCLE_data
database$type <- tolower(database$type)
genes <- CT_genes$external_gene_name
database <- database[rowData(database)$external_gene_name %in% genes, ]
database <- database[match(genes, rowData(database)$external_gene_name), ]

mat <- log1p(assay(database))
rownames(mat) <- rowData(database)$external_gene_name

df_col <- data.frame("cell_line" = colData(database)$cell_line_name,
                     "type" = as.factor(colData(database)$type))
rownames(df_col) <- rownames(colData(database))
df_col <- df_col[order(df_col$type), ]

column_ha_type <- HeatmapAnnotation(
  type = df_col$type,
  border = TRUE,
  annotation_name_gp = gpar(fontsize = 8),
  annotation_legend_param = legends_param,
  col = list(type = c("lung" = "seagreen3", "skin" = "red3",
                 "bile_duct" = "mediumpurple1", "bladder" = "mistyrose2",
                 "colorectal" = "plum", "lymphoma" = "steelblue1",
                 "uterine" = "darkorange4", "myeloma" = "turquoise3", 
                 "kidney" = "thistle4",
                 "pancreatic" = "darkmagenta", "brain" = "palegreen2",
                 "gastric" = "wheat3", "breast" = "midnightblue",
                 "bone" = "sienna1", "head_and_neck" = "deeppink2",
                 "ovarian" = "tan3", "sarcoma" = "lightcoral",
                 "leukemia" = "steelblue4", "esophageal"= "khaki",
                 "neuroblastoma" = "olivedrab1")))

Heatmap(mat[, rownames(df_col), drop = FALSE],
        name = "logTPM",
        column_title = "Gene Expression in tumor cell lines (CCLE)",
        column_split = factor(df_col$type),
        row_split = split,
        row_title_gp = gpar(fontsize = 0),
        col = colorRamp2(seq(0, max(mat), length = 11),
                          legend_colors),
        clustering_method_rows = "ward.D",
        clustering_method_columns = "ward.D",
        cluster_rows = TRUE,
        show_row_dend = FALSE,
        show_column_names = FALSE,
        cluster_columns = TRUE,
        show_column_dend = FALSE,
        row_names_gp = gpar(fontsize = 4),
        heatmap_legend_param = legends_param,
        top_annotation = c(column_ha_type),
        left_annotation = left_annot)
```

```{r all_meth}
genes <- CT_genes$external_gene_name
database <- CT_mean_methylation_in_tissues[rownames(CT_mean_methylation_in_tissues) %in% genes]

mat <- na.omit(assay(database))
clustering_option <- TRUE

row_ha_chr_meth <- rowAnnotation(chr = chr_mat[rownames(mat),],
                            annotation_legend_param = legends_param,
                            simple_anno_size = unit(0.5, "cm"),
                            col = list(chr = chr_colors),
                            annotation_name_gp = gpar(fontsize = 8),
                            annotation_name_side = "top")

row_ha_reg_meth <- rowAnnotation(regulation = regulation_mat[rownames(mat),],
                            annotation_legend_param = legends_param,
                            simple_anno_size = unit(0.5, "cm"),
                            col = list(regulation = meth_colors),
                            annotation_name_gp = gpar(fontsize = 8),
                            annotation_name_side = "top")

left_annot_meth <- c(row_ha_chr_meth, row_ha_reg_meth, gap = unit(1, "mm"))

split_meth <- data.frame(filter(CT_genes, external_gene_name %in% rownames(mat))$regulated_by_methylation, 
                    filter(CT_genes, external_gene_name %in% rownames(mat))$X_linked)

Heatmap(mat,
        column_title = 'Promoter mean methylation level by tissue',
        name = 'Meth',
        col = colorRamp2(c(1:100),
                         colorRampPalette(c("moccasin","dodgerblue4"))(100)),
        na_col = "gray80",
        cluster_rows = clustering_option,
        cluster_columns = FALSE,
        row_split = split_meth,
        row_title_gp = gpar(fontsize = 0),
        show_row_names = TRUE,
        show_heatmap_legend = TRUE,
        show_row_dend = FALSE,
        row_names_gp = gpar(fontsize = 3),
        column_names_gp = gpar(fontsize = 8),
        column_names_side = "bottom",
        row_names_side = "right",
        left_annotation = left_annot_meth)

```

### MAGE genes

```{r MAGE_TCGA}
MAGE_genes <- filter(CT_genes, family == "MAGE")$external_gene_name

TCGA_expression(tumor = "all", 
                genes = MAGE_genes,
                units = "log_TPM")

CCLE_expression(genes = MAGE_genes,
                 type = c("lung", "skin", "bile_duct", "bladder", 
                          "colorectal", "lymphoma", "uterine",
                          "myeloma", "kidney", "pancreatic", "brain", 
                          "gastric", "breast", "bone", "head_and_neck", 
                          "ovarian", "sarcoma", "leukemia", "esophageal",
                          "neuroblastoma"), units = "log_TPM")

normal_tissues_mean_methylation(MAGE_genes, na.omit = TRUE)

```

## Single Cell analysis : timing of expression

In CTexploreR, we there is single cell data from the testis, we thus can 
analyse CT genes expression during spermatogenesis.

There is only data for 251 of our 298 CT genes.

NB : SSC, Spermatogonia and early spermatocytes are *premeiotic cells*. Late 
spermatocytes (between both meiosis), round spermatid, elongated spermatid and 
sperm are *postmeiotic cells*.

```{r repartition_sc}
genes_avail <- 
  CT_genes$external_gene_name[CT_genes$external_gene_name %in% unique(rownames(testis_sce))]
table(CT_genes$testis_cell_type)
table(CT_genes$testis_cell_type)/length(genes_avail)*100
```

41% of genes are mainly expressed pre-meioticly.

```{r sc_heatmap, message = FALSE}
germ_cells <- c("SSC", "Spermatogonia", "Early_spermatocyte",
                "Late_spermatocyte","Round_spermatid", "Elongated_spermatid",
                "Sperm1", "Sperm2")
somatic_cells <- c("Macrophage", "Endothelial", "Myoid", "Sertoli", "Leydig")

testis_sce_CT <- testis_sce[genes_avail, ]

mat <- SingleCellExperiment::logcounts(testis_sce_CT)

df_col <- data.frame(clusters = colData(testis_sce_CT)$clusters,
                     type = colData(testis_sce_CT)$type,
                     Donor = colData(testis_sce_CT)$Donor)
rownames(df_col) <- colnames(testis_sce_CT)

df_col <- df_col[order(df_col$type),]
df_col$lineage <- "Germ cells"
df_col$lineage[df_col$type %in% somatic_cells] <- "Somatic cells"

column_ha_type = HeatmapAnnotation(
  type = df_col$type,
  border = TRUE,
  col = list(type = c("SSC" = "floralwhite", "Spermatogonia" = "moccasin",
                   "Early_spermatocyte" = "gold", 
                   "Late_spermatocyte" = "orange",
                   "Round_spermatid" = "red2", 
                   "Elongated_spermatid" = "darkred",
                   "Sperm1" = "violet", "Sperm2" = "purple", 
                   "Sertoli" = "gray", 
                   "Leydig" = "cadetblue2", "Myoid" = "springgreen3", 
                   "Macrophage" = "gray10",
                   "Endothelial" = "steelblue")),

  annotation_name_gp = gpar(fontsize = 8),
  annotation_legend_param = legends_param)

column_ha_lineage = HeatmapAnnotation(
  lineage = df_col$lineage,
  border = TRUE,
  col = list(lineage = c("Germ cells" = "salmon", "Somatic cells" = "cyan4")),
  annotation_name_gp = gpar(fontsize = 8),
  annotation_legend_param = legends_param)

scale_lims <- c(0, quantile(rowMax(mat), 0.75))
top_annot <- c(column_ha_lineage, column_ha_type)

# Until here is what's in the function, hereunder is my addition/change in Heatmap()

CT_genes_avail <- filter(CT_genes, external_gene_name %in% genes_avail)

chr_mat <- as.matrix(CT_genes_avail$X_linked)
chr_mat <- ifelse(chr_mat == TRUE, "X-linked", "Not X")
rownames(chr_mat) <- CT_genes_avail$external_gene_name
row_ha_chr <- rowAnnotation(chr = chr_mat,
                            annotation_legend_param = legends_param,
                            simple_anno_size = unit(0.5, "cm"),
                            col = list(chr = chr_colors),
                            annotation_name_gp = gpar(fontsize = 8),
                            annotation_name_side = "top")

regulation_mat <- as.matrix(CT_genes_avail$regulated_by_methylation)
regulation_mat <- ifelse(regulation_mat == TRUE, "Methylation",
                         "Not methylation")
rownames(regulation_mat) <- CT_genes_avail$external_gene_name

row_ha_reg <- rowAnnotation(regulation = regulation_mat,
                            annotation_legend_param = legends_param,
                            simple_anno_size = unit(0.5, "cm"),
                            col = list(regulation = meth_colors),
                            annotation_name_gp = gpar(fontsize = 8),
                            annotation_name_side = "top")

left_annot <- c(row_ha_chr, row_ha_reg, gap = unit(1, "mm"))

split <- data.frame(CT_genes_avail$regulated_by_methylation, CT_genes_avail$X_linked)

Heatmap(mat[genes_avail, rownames(df_col), drop = FALSE],
        name = "logCounts",
        column_title = "Expression in testis cells (scRNAseq)",
        column_split = df_col$type,
        row_split = split,
        row_title_gp = gpar(fontsize = 0),
        show_column_names = FALSE,
        show_column_dend = FALSE,
        clustering_method_rows = "ward.D",
        clustering_method_columns = "ward.D",
        cluster_rows = TRUE,
        cluster_columns = FALSE,
        show_row_dend = FALSE,
        row_names_gp = gpar(fontsize = 4),
        col = colorRamp2(seq(scale_lims[1], scale_lims[2], length = 11),                 
                         legend_colors),
        top_annotation = top_annot,
        left_annotation = left_annot,
        heatmap_legend_param = legends_param)

```

We used these data to determine in which testis cell type
each gene is mostly expressed.

```{r main_cell_type_expression_barplot}
CT_genes$main_cell_type_expression <- factor(CT_genes$testis_cell_type,
                                                levels = c("SSC", 
                                               "Spermatogonia", 
                                               "Early_spermatocyte",
                                               "Late_spermatocyte",
                                               "Round_spermatid",
                                               "Elongated_spermatid",
                                               "Sperm1",
                                               "Sperm2",
                                               "Sertoli"))

CT_genes %>% 
  mutate(regulated_by_methylation = ifelse(regulated_by_methylation,
                                           "Regulated by methylation",
                                           "Not regulated by methylation")) %>%
  filter(!is.na(testis_cell_type)) %>%
  filter(main_cell_type_expression != "Sertoli") %>% 
  ggplot(aes(x = chr, fill = main_cell_type_expression,
             color = main_cell_type_expression)) +
  scale_fill_manual(values = c("lightyellow2", "moccasin", "gold", "orange","red2",
                               "darkred", "violet", "purple")) +
  scale_color_manual(values = c("lightyellow3", "navajowhite2", "gold", "orange","red2",
                               "darkred", "violet", "purple")) +
  geom_bar(stat = 'count') +
  facet_grid(main_cell_type_expression ~ regulated_by_methylation) +
  xlab("Chromosome") +
  ylab("Number of genes") +
  theme_bw() +
  theme(
    axis.text.x = element_text(color = "black", angle = 90, vjust = 0.5, size = 10),
    axis.title = element_text(size = 10, color = "gray10"),
    strip.text.y = element_blank())

```

14 genes are on the X chromosome but escape X chromosome inactivation and are
activates post-meioticly.

```{r escapees}
CT_genes %>% 
  filter(X_linked) %>% 
  filter(testis_cell_type %in% c("Late_spermatocyte","Round_spermatid", 
                                 "Elongated_spermatid", "Sperm1", "Sperm2"))

```

## Gene function

```{r GO_ORA}
go_ora <- enrichGO(gene = CT_genes$ensembl_gene_id,
                   keyType = "ENSEMBL",
                   OrgDb = org.Hs.eg.db,
                   ont = "all",
                   readable = TRUE)
as_tibble(go_ora)

as_tibble(go_ora) %>% 
  arrange(desc(Count)) %>% 
  head(12) %>% 
  mutate(Ratio = Count/183) %>% 
  ggplot(aes(x = Ratio, y = Description, fill = Description)) +
  geom_col() +
  theme_bw() +
  ylab("GO term") +
  xlab("Gene Ratio") +
  theme(axis.text.y = element_blank(),
        legend.position = "none",
        axis.ticks.y = element_blank(),
        axis.title = element_text(size = 10, color = "gray10"))+ 
  geom_text(aes(0, y = Description, label = Description),
            hjust = 0,
            nudge_x = 0.005,
            colour = "floralwhite",
            size = 4)

```

As we can see here, most of genes are indeed linked to functions from reproduction.
I represented here the 12 categories with the most genes, all from biological 
processes. However, they are enriched in 89 different GO terms.

We've also added a column tumor suppressor and oncogene in CTexploreR. 
These information come from Cancermine.

```{r C_TS}
table(CT_genes$oncogene) 
table(CT_genes$tumor_suppressor)

table(CT_genes$oncogene, CT_genes$tumor_suppressor) 
```

# Package application : What are the mechanisms involved in CT demethylation in cancer

## Determination of PMDs in male fibroblasts celles

WGBS data from GM23248 cells (male so X chromosome is reliable) downloaded from 
ENCODE.
Data was processed using a WGBS pipeline [MethPipe]
(http://smithlabresearch.org/software/methpipe/) including [TrimGalore](https://www.bioinformatics.babraham.ac.uk/projects/trim_galore/) 
to remove low quality reads and trim the adapter from the sequences.

The pipeline has been verified by comparing PMDs already defined by Lister et 
al. on IMR90 and PMDs obtained with MethPipe.

```{bash download_GM23248, eval = FALSE}
wget https://www.encodeproject.org/files/ENCFF142GZA/@@download/ENCFF142GZA.fastq.gz
wget https://www.encodeproject.org/files/ENCFF845RUF/@@download/ENCFF845RUF.fastq.gz
wget https://www.encodeproject.org/files/ENCFF968GRR/@@download/ENCFF968GRR.fastq.gz
wget https://www.encodeproject.org/files/ENCFF975YNS/@@download/ENCFF975YNS.fastq.gz
```

```{bash trim/methpipe_GM23248, eval = FALSE}
#!/bin/bash
mkdir /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/rep1_ENCLB777FMV;
cd /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/rep1_ENCLB777FMV
/opt/TrimGalore/trim_galore --fastqc --paired /data/cbio/2021-mth-julie-devis/WGBS-processing/GM23248_data/ENCFF142GZA.fastq /data/cbio/2021-mth-julie-devis/WGBS-processing/GM23248_data/ENCFF845RUF.fastq

mkdir /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/rep2_ENCLB436YZI;
cd /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/rep2_ENCLB436YZI
/opt/TrimGalore/trim_galore --fastqc --paired /data/cbio/2021-mth-julie-devis/WGBS-processing/GM23248_data/ENCFF975YNS.fastq /data/cbio/2021-mth-julie-devis/WGBS-processing/GM23248_data/ENCFF968GRR.fastq

#!/bin/bash

cd /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/ENCODE_GM23248/rep1_ENCLB777FMV;

#Alignement
/home/users/jdevis/walt/bin/walt \
-i /data/cbio/2021-mth-julie-devis/WGBS-processing/genomes/hg38_by_Chr.dbindex \
-1 ENCFF142GZA_val_1.fq \
-2 ENCFF845RUF_val_2.fq \
-o rep1_ENCLB777FMV_mapping.mr;

#Duplicate-remover
LC_ALL=C sort -k 1,1 -k 2,2n -k 3,3n -k 6,6 -o rep1_ENCLB777FMV.mr.sorted_start rep1_ENCLB777FMV_mapping.mr;
/opt/methpipe/build/duplicate-remover -S rep1_ENCLB777FMV_dremove_stat.txt -o rep1_ENCLB777FMV.mr.dremove rep1_ENCLB777FMV.mr.sorted_start;

#bsrate
/opt/methpipe/build/bsrate -c /data/cbio/GENOMES/GRCh38_by_Chr/hgdownload.cse.ucsc.edu/chromosomes \
-o rep1_ENCLB777FMV.bsrate rep1_ENCLB777FMV.mr.dremove;

#Methylation levels
/opt/methpipe/build/methcounts -n \
-c /data/cbio/GENOMES/GRCh38_by_Chr/hgdownload.cse.ucsc.edu/chromosomes \
-o rep1_ENCLB777FMV.meth rep1_ENCLB777FMV.mr.dremove;

#Merge CpG
/opt/methpipe/build/symmetric-cpgs -o rep1_ENCLB777FMV_CpG_merge.meth rep1_ENCLB777FMV.meth;

#Stats
/opt/methpipe/build/levels -o rep1_ENCLB777FMV.levels rep1_ENCLB777FMV.meth;

cd /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/ENCODE_GM23248/rep2_ENCLB436YZI;

#Alignement
/home/users/jdevis/walt/bin/walt \
-i /data/cbio/2021-mth-julie-devis/WGBS-processing/genomes/hg38_by_Chr.dbindex \
-1 ENCFF975YNS_val_1.fq \
-2 ENCFF968GRR_val_2.fq \
-o rep2_ENCLB436YZI_mapping.mr;

#Duplicate-remover
LC_ALL=C sort -k 1,1 -k 2,2n -k 3,3n -k 6,6 -o rep2_ENCLB436YZI.mr.sorted_start rep2_ENCLB436YZI_mapping.mr;
/opt/methpipe/build/duplicate-remover -S rep2_ENCLB436YZI_dremove_stat.txt -o rep2_ENCLB436YZI.mr.dremove rep2_ENCLB436YZI.mr.sorted_start;

#bsrate
/opt/methpipe/build/bsrate -c /data/cbio/GENOMES/GRCh38_by_Chr/hgdownload.cse.ucsc.edu/chromosomes \
-o rep2_ENCLB436YZI.bsrate rep2_ENCLB436YZI.mr.dremove;

#Methylation levels
/opt/methpipe/build/methcounts -n \
-c /data/cbio/GENOMES/GRCh38_by_Chr/hgdownload.cse.ucsc.edu/chromosomes \
-o rep2_ENCLB436YZI.meth rep2_ENCLB436YZI.mr.dremove;

#Merge CpG
/opt/methpipe/build/symmetric-cpgs -o rep2_ENCLB436YZI_CpG_merge.meth rep2_ENCLB436YZI.meth;

#Stats
/opt/methpipe/build/levels -o rep2_ENCLB436YZI.levels rep2_ENCLB436YZI.meth;

#Merge both replicats
cd /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/ENCODE_GM23248;

/opt/methpipe/build/merge-methcounts rep1_ENCLB777FMV/rep1_ENCLB777FMV_CpG_merge.meth \
rep2_ENCLB436YZI/rep2_ENCLB436YZI_CpG_merge.meth \
-o GM23248_merged.meth;

#pmd
/opt/methpipe/build/pmd -v -i 1000 -o GM23248.pmd GM23248_merged.meth
```

```{r import_meth_GM23248, eval = FALSE}
GM23248_meth <- read_delim("GM23248_merged.meth", "\t", escape_double = FALSE, col_names = FALSE, trim_ws = TRUE)
names(GM23248_meth) <- c("chromosome", "position", "strand", "sequence_context","methylation_level","number_of_reads")
```

We can use the PMD file coming from MethPipe and change it into a GRanges object
so that we can easily do overlapping analysis

```{r load_PMD, message = FALSE}
my_PMD_GM23248 <- read_delim("data/GM23248.pmd", "\t", 
                             escape_double = FALSE, col_names = FALSE, 
                             trim_ws = TRUE)
names(my_PMD_GM23248) <- c("chr", "start","end","name_quality","bins","strand")
my_PMD_GM23248 <- separate(my_PMD_GM23248, "name_quality", 
                           into = c("name", "likelihood_start", "certainty_start",
                                    "likelihood_end", "certainty_end"), 
                           sep = ":")
my_PMD_GM23248_GR <- makeGRangesFromDataFrame(my_PMD_GM23248 [, -10], keep.extra.columns = TRUE)
```

Not PMD and NA regions aren’t dissociated. Coverage of windows of 10kb were 
calculated along the genome. A threshold of 1000 reads per 10kb is necessary to 
not be considered NA.

```{r all_chr_10kb, eval = FALSE}
for (chromosome_name in paste0("chr",c(1:22,"X","Y"))){

  roi <- GM23248_meth %>% 
    filter(chromosome == chromosome_name)

  res <- data.frame(region_i = floor(min(roi$position)/10000):floor(max(roi$position)/10000),
                    reads_tot = rep(NA,length(floor(min(roi$position)/10000):floor(max(roi$position)/10000))),
                    chromosome = rep(unique(roi$chromosome),length(floor(min(roi$position)/10000):floor(max(roi$position)/10000))))

  res <- res %>% 
    mutate(start_region = region_i*10000) %>% 
    mutate(end_region = region_i*10000+10000) %>% 
    column_to_rownames("region_i")

  for (i in floor(min(roi$position)/10000):floor(max(roi$position)/10000)){
    value <-  i*10000
    tmp <- roi %>% 
      filter(position >= value & position < (value + 10000))
    res[as.character(i) , "reads_tot"] <- sum(tmp$number_of_reads)
    }

  res_GR <- makeGRangesFromDataFrame(res, start.field = "start_region", 
                                     end.field = "end_region", keep.extra.columns = TRUE)

  overlap_boul <- overlapsAny(res_GR, my_PMD_GM23248_GR, type = "within")

  res$in_PMD <- overlap_boul

  saveRDS(res, file = paste0("data/10kb_reads_PMD_", chromosome_name, ".rds"))

}

windows_reads_PMD_chr1 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr1.rds")
windows_reads_PMD_chr2 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr2.rds")
windows_reads_PMD_chr3 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr3.rds")
windows_reads_PMD_chr4 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr4.rds")
windows_reads_PMD_chr5 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr5.rds")
windows_reads_PMD_chr6 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr6.rds")
windows_reads_PMD_chr7 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr7.rds")
windows_reads_PMD_chr8 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr8.rds")
windows_reads_PMD_chr9 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr9.rds")
windows_reads_PMD_chr10 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr10.rds")
windows_reads_PMD_chr11 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr11.rds")
windows_reads_PMD_chr12 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr12.rds")
windows_reads_PMD_chr13 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr13.rds")
windows_reads_PMD_chr14 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr14.rds")
windows_reads_PMD_chr15 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr15.rds")
windows_reads_PMD_chr16 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr16.rds")
windows_reads_PMD_chr17 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr17.rds")
windows_reads_PMD_chr18 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr18.rds")
windows_reads_PMD_chr19 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr19.rds")
windows_reads_PMD_chr20 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr20.rds")
windows_reads_PMD_chr21 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr21.rds")
windows_reads_PMD_chr22 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr22.rds")
windows_reads_PMD_chrX <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chrX.rds")
windows_reads_PMD_chrY <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chrY.rds")

windows_reads_PMD <- rbind(windows_reads_PMD_chr1, windows_reads_PMD_chr2, windows_reads_PMD_chr3, windows_reads_PMD_chr4, windows_reads_PMD_chr5, windows_reads_PMD_chr6, windows_reads_PMD_chr7, windows_reads_PMD_chr8, windows_reads_PMD_chr9, windows_reads_PMD_chr10, windows_reads_PMD_chr11, windows_reads_PMD_chr12, windows_reads_PMD_chr13, windows_reads_PMD_chr14, windows_reads_PMD_chr15, windows_reads_PMD_chr16, windows_reads_PMD_chr17, windows_reads_PMD_chr18, windows_reads_PMD_chr19, windows_reads_PMD_chr20, windows_reads_PMD_chr21, windows_reads_PMD_chr22, windows_reads_PMD_chrX, windows_reads_PMD_chrY)

rownames(windows_reads_PMD) <- NULL

saveRDS(windows_reads_PMD, file = "data/10kb_reads_PMD_all.rds")
```

Here we load the coverage for all the 10k windows and determine which one are 
non-PMD due to coverage issues. These region are to be removed from the analysis.

```{r NA_windows}
windows_reads_PMD <- readRDS("data/10kb_reads_PMD_all.rds")

NA_nonPMD_GR <- windows_reads_PMD %>% 
  filter(!in_PMD) %>% 
  filter(reads_tot <= 1000) %>% 
  makeGRangesFromDataFrame(start.field = "start_region", end.field = "end_region",
                           keep.extra.columns = TRUE) %>% 
  GenomicRanges::reduce()
```

## Are CT genes associated with PMDs ?

### CT regulated by methylation and PMD overlap

```{r CT_genes_prep}
CT_genes <- CT_genes %>% 
  mutate(start_prom = case_when(strand == 1 ~ transcription_start_site - 1000,
                                strand == -1 ~ transcription_start_site - 500)) %>% 
  mutate(end_prom = case_when(strand == 1 ~ transcription_start_site + 500,
                              strand == -1 ~ transcription_start_site + 1000))

CT_genes$strand <- ifelse(CT_genes$strand == "1", "+", "-")
seqlevelsStyle(CT_genes$chr) <- "UCSC"

CT_meth_regulated <- CT_genes %>% 
  filter(regulated_by_methylation)
```

```{r CT_prom}
meth_reg_prom_GR <- makeGRangesFromDataFrame(CT_meth_regulated, start.field = "start_prom",
                                    end.field = "end_prom",
                                    keep.extra.columns = TRUE)
```

```{r overlap_PMD_CT}
CTmeth_PMD_GR <- subsetByOverlaps(meth_reg_prom_GR, my_PMD_GM23248_GR)
CTmeth_not_PMD_GR <- subsetByOverlaps(subsetByOverlaps(meth_reg_prom_GR, 
                                                       my_PMD_GM23248_GR, 
                                                   invert = TRUE),
                                      NA_nonPMD_GR, invert = TRUE)
CTmeth_NA_PMD_GR <-subsetByOverlaps(subsetByOverlaps(meth_reg_prom_GR, 
                                                     my_PMD_GM23248_GR, 
                                                 invert = TRUE), 
                                    NA_nonPMD_GR)
```

Now that these overlapping have been done, I can report them into my data as a boulean column to use analyse it.

```{r PMD_in_CT}
CT_meth_regulated <- CT_meth_regulated %>% 
  mutate(PMD = case_when(CT_meth_regulated$external_gene_name %in% CTmeth_PMD_GR$external_gene_name ~ TRUE,
                         CT_meth_regulated$external_gene_name %in% CTmeth_not_PMD_GR$external_gene_name ~ FALSE,
                         CT_meth_regulated$external_gene_name %in% CTmeth_NA_PMD_GR$external_gene_name ~ NA))
table(is.na(CT_meth_regulated$PMD))
table(CT_meth_regulated$PMD)

PMD_perc <- table(CT_meth_regulated$PMD)[[2]] / 
  sum(table(CT_meth_regulated$PMD))*100
PMD_perc

X_in_PMD_table <- CT_meth_regulated %>% 
  filter(X_linked) %>% 
  dplyr::select(PMD) %>% 
  table()
X_in_PMD_table

PMD_chr_repartition <- CT_meth_regulated %>% 
  filter(!is.na(PMD)) %>% 
  dplyr::select(X_linked) %>% 
  table()
PMD_chr_repartition
```

### CT not regulated by methylation and PMD overlap

```{r CT_not_meth_PMD}
CT_not_meth_regulated <- CT_genes %>% 
  filter(!regulated_by_methylation)

prom_GR_not_reg <- makeGRangesFromDataFrame(CT_not_meth_regulated, 
                                            start.field = "start_prom",
                                    end.field = "end_prom",
                                    keep.extra.columns = TRUE)
not_reg_PMD_GR <- subsetByOverlaps(prom_GR_not_reg, my_PMD_GM23248_GR)
not_reg_not_PMD_GR <- subsetByOverlaps(subsetByOverlaps(prom_GR_not_reg,
                                                        my_PMD_GM23248_GR, 
                                                   invert = TRUE), 
                                  NA_nonPMD_GR, invert = TRUE)
not_reg_NA_PMD_GR <-subsetByOverlaps(subsetByOverlaps(prom_GR_not_reg,
                                                      my_PMD_GM23248_GR, 
                                                 invert = TRUE),
                                NA_nonPMD_GR)

CT_not_meth_regulated <- CT_not_meth_regulated %>% 
  mutate(PMD = case_when(CT_not_meth_regulated$external_gene_name %in% 
                           not_reg_PMD_GR$external_gene_name ~ TRUE,
                         CT_not_meth_regulated$external_gene_name %in% 
                           not_reg_not_PMD_GR$external_gene_name ~ FALSE,
                         CT_not_meth_regulated$external_gene_name %in% 
                           not_reg_NA_PMD_GR$external_gene_name ~ NA))

table(is.na(CT_not_meth_regulated$PMD))
table(CT_not_meth_regulated$PMD)

CT_not_meth_reg_PMD_perc <- table(CT_not_meth_regulated$PMD)[[2]] / 
  sum(table(CT_not_meth_regulated$PMD))*100
CT_not_meth_reg_PMD_perc
```

### Enrichment analysis

The numbers are useful but we don't know if it is significant or not. 
In order to do so, we're going to compare this percentage with an empiric 
distribution. For that we need groups of genes comparable.

This distribution is going to be realised by 1000 times computing the % of genes
in PMD in `r PMD_chr_repartition[[1]]` tissue-specific genes on autosomes and 
`r PMD_chr_repartition[[2]]` tissue-specific genes on the chromosome X. We use 
tissue-specific genes because this characteristic can make them more prone to be
in a repressive region like PMDs.

We then have to create a pool of non-CT tissue-specific genes with their 
localisation, for that we use the list of tissue specific genes that can be 
found on the Human Protein Atlas (3106 tissue enriched genes: At least four-fold
higher mRNA level in a particular tissue compared to any other tissue).

It is necessary to use biomaRt because we need transcription start sites to be 
able to realise the same analysis on enriched genes than on CT genes. Also, we 
need to remove CT genes and genes that are in the non-covered regions.

```{r enriched_genes}
enriched_genes <- read.delim("data/HPA_tissue_enriched_genes.tsv")

attributes_vector <- c("ensembl_gene_id","external_gene_name", "chromosome_name",
                       "strand", "transcription_start_site", 
                       "external_transcript_name", "transcript_is_canonical")
biomaRt_query <- getBM(attributes = attributes_vector, mart = ensembl)
enriched_genes_biomart <- biomaRt_query %>% 
  filter(ensembl_gene_id %in% enriched_genes$Ensembl)
enriched_genes_biomart <- enriched_genes_biomart %>% 
  mutate(start_prom = case_when(strand == 1 ~ transcription_start_site - 1000,
                                strand == -1 ~ transcription_start_site - 500)) %>% 
  mutate(end_prom = case_when(strand == 1 ~ transcription_start_site + 500,
                              strand == -1 ~ transcription_start_site + 1000)) %>% 
  filter(!external_gene_name %in% CT_genes$external_gene_name) %>% 
  filter(transcript_is_canonical == 1) %>%
  filter(chromosome_name%in%c(1:22, "X", "Y"))

enriched_genes_biomart$strand <- ifelse(enriched_genes_biomart$strand == "1", "+", "-")
seqlevelsStyle(enriched_genes_biomart$chromosome_name) <- "UCSC"

enriched_genes_GR <- makeGRangesFromDataFrame(enriched_genes_biomart, 
                                              start.field = "start_prom",
                                              end.field = "end_prom",
                                              keep.extra.columns = TRUE)
enriched_NA_GR <-subsetByOverlaps(subsetByOverlaps(enriched_genes_GR, 
                                                   my_PMD_GM23248_GR, 
                                                   invert = TRUE),
                                  NA_nonPMD_GR)

enriched_genes_analysable <- as.data.frame(
  enriched_genes_GR[!enriched_genes_GR$ensembl_gene_id %in% enriched_NA_GR$ensembl_gene_id, ])
```

```{r enrichment_analysis}
random_percentage_chrprop <- function(i, data, nchrX, nautosomes){
  random_sample1 <- sample_n(filter(data, seqnames == "chrX"), nchrX)
  random_sample2 <- sample_n(filter(data, seqnames != "chrX"), nautosomes)
  random_sample <- rbind(random_sample1,random_sample2)
  random_sample_GR <- makeGRangesFromDataFrame(random_sample, 
                                               keep.extra.columns = TRUE)
  length(subsetByOverlaps(random_sample_GR, my_PMD_GM23248_GR)) / 
    length(random_sample_GR)*100
}

distrib_enriched <- unlist(bplapply(1:1000, random_percentage_chrprop, 
                                    data = enriched_genes_analysable, 
                                    nchrX = PMD_chr_repartition[[2]], 
                                    nautosomes = PMD_chr_repartition[[1]]))
hist(distrib_enriched,
     col="violetred1",
     border = "violetred3",
     breaks = 20,
     main = paste0("Percentage of promoters in PMDs when randomly taking ",
                   sum(table(CT_meth_regulated$PMD)), 
                   " tissue-specific 
genes 1000 times (taking chromosome proportion into account)"),
     cex.main = 0.75,
     xlab = "Percentages",
     ylab = "Frequency",
     xlim = c(20,60))
abline(v = PMD_perc, col = "slateblue1", lwd = 3)
text(PMD_perc, 120, "CT genes regulated 
by methylation", col = "slateblue4", cex = 0.8)
abline(v = CT_not_meth_reg_PMD_perc, col = "darkseagreen1", lwd = 3)
text(CT_not_meth_reg_PMD_perc, 120, "CT genes not regulated 
by methylation", col = "darkseagreen", cex = 0.8)

```

```{r add_PMD}
PMD_CT_results <- rbind(dplyr::select(CT_not_meth_regulated, 
                                      c(PMD, external_gene_name)),
                    dplyr::select(CT_meth_regulated, 
                                  c(PMD, external_gene_name)))
CT_genes <- left_join(CT_genes, PMD_CT_results)
```

## Methylation in an aging model : CD4+ cells

Now that we know that CT genes are indeed enriched in PMDs, we have to go further. Indeed, we know that proliferation has an impact on PMD and their methylation : there are losses of methylation with time. As CT genes are located in PMDs, we can hypothesise that their methylation might be impacted by the losses happening during proliferation. We thus want to focus on the impact on CT genes of the methylation changes happening on PMDss in proliferative or aging cells.

### Data processing

Data coming from Heyn 2012 paper.

```{bash CD4, eval = FALSE}
#!/bin/bash

#Newborn
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR330578
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR330579
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394135
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394136
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394137
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394138
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394139
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394140
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394141
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394142

#Centenarian
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR330574
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR330576
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR330577
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR330575

#Middle-age
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR389249
/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR389248

for f in `cat CD4_SRA.txt`;

do mkdir /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/$f; 
cd /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/$f;

#Trimming
/opt/TrimGalore/trim_galore --fastqc --path_to_cutadapt /home/users/jdevis/cutadapt --paired /data/cbio/2021-mth-julie-devis/WGBS-processing/rawdata/$f\_1.fastq /data/cbio/2021-mth-julie-devis/WGBS-processing/rawdata/$f\_2.fastq;

#Alignement
/home/users/jdevis/walt/bin/walt \
-i /data/cbio/2021-mth-julie-devis/WGBS-processing/genomes/hg38_by_Chr.dbindex \
-1 $f\_1_val_1.fq \
-2 $f\_2_val_2.fq \
-o $f\_mapping.mr;

#Duplicate-remover
LC_ALL=C sort -k 1,1 -k 2,2n -k 3,3n -k 6,6 -o $f.mr.sorted_start $f\_mapping.mr;
/opt/methpipe/build/duplicate-remover -S $f\_dremove_stat.txt -o $f.mr.dremove $f.mr.sorted_start;

#bsrate
/opt/methpipe/build/bsrate -c /data/cbio/GENOMES/GRCh38_by_Chr/hgdownload.cse.ucsc.edu/chromosomes \
-o $f.bsrate $f.mr.dremove;

#Methylation levels
/opt/methpipe/build/methcounts -n \
-c /data/cbio/GENOMES/GRCh38_by_Chr/hgdownload.cse.ucsc.edu/chromosomes \
-o $f.meth $f.mr.dremove;

#Merge CpG
/opt/methpipe/build/symmetric-cpgs -o $f\_CpG_merge.meth $f.meth;

#Stats
/opt/methpipe/build/levels -o $f.levels $f.meth

done

```

Bisulfite-Seq data coming from a male newborn, a middle age donor and a male centenarian.

Quality of the data is good, with a mean of 13 reads per cytosine. However, middle age data has no info about the sex of the donor. By comparing the number of reads by chromosome in this sample, a female one and the male ones, we saw that the proportion of reads on chromosomes X and Y showed that the middle age data come from a woman. We then have to be very careful when analysing the X chromosome with the middle age data. To facilitate the analysis and compare all samples, I’ll remove the adult data on chrX.

```{r load_CD4, message = FALSE, warning = FALSE}
columns <- c("chromosome", "position", "strand", "sequence context","methylation_level","number_of_reads")
CD4_newborn <- read_delim("data/CD4_newborn.meth", 
    delim = "\t", escape_double = FALSE, col_names = FALSE, trim_ws = TRUE)
names(CD4_newborn) <- columns
CD4_middle <- read_delim("data/CD4_middle.meth", 
    delim = "\t", escape_double = FALSE, col_names = FALSE, trim_ws = TRUE)
names(CD4_middle) <- columns
CD4_middle <- filter(CD4_middle, chromosome != "chrX")
CD4_cent <- read_delim("data/CD4_cent.meth", 
    delim = "\t", escape_double = FALSE, col_names = FALSE, trim_ws = TRUE)
names(CD4_cent) <- columns
```
```{r change_CD4, message = FALSE, warning = FALSE}
CD4_newborn$age <- "newborn"
CD4_middle$age <- "middle_age"
CD4_cent$age <- "centenarian"
```
```{r CD4_GR, message = FALSE, warning = FALSE}
CD4_newborn_GR <- makeGRangesFromDataFrame(CD4_newborn, start.field = "position", 
                                           end.field = "position", keep.extra.columns = TRUE)
CD4_middle_GR <- makeGRangesFromDataFrame(CD4_middle, start.field = "position", 
                                          end.field = "position", keep.extra.columns = TRUE)
CD4_cent_GR <- makeGRangesFromDataFrame(CD4_cent, start.field = "position", 
                                        end.field = "position", keep.extra.columns = TRUE)
```
```{r rm, echo = FALSE}
rm(CD4_newborn)
rm(CD4_middle)
rm(CD4_cent)
```

### CT genes methylation level

Here I extract methylation for all CpGs located in CT genes promoter 
(regulated by methylation because it's these that we can find in PMDs).

```{r CT_CD4, warning = FALSE}
CT_prom_GR <- makeGRangesFromDataFrame(CT_genes, start.field = "start_prom",
                                    end.field = "end_prom",
                                    keep.extra.columns = TRUE)

newborn <- mergeByOverlaps(CD4_newborn_GR, CT_prom_GR, ignore.strand = TRUE)[,1]
newborn$external_gene_name <- mergeByOverlaps(CD4_newborn_GR, CT_prom_GR, 
                                              ignore.strand = TRUE)$external_gene_name
newborn <- as.data.frame(newborn)
middle <- mergeByOverlaps(CD4_middle_GR, CT_prom_GR, ignore.strand = TRUE)[,1]
middle$external_gene_name <- mergeByOverlaps(CD4_middle_GR, CT_prom_GR, 
                                             ignore.strand = TRUE)$external_gene_name
middle <- as.data.frame(middle)
cent <- mergeByOverlaps(CD4_cent_GR, CT_prom_GR, ignore.strand = TRUE)[,1]
cent$external_gene_name <- mergeByOverlaps(CD4_cent_GR, CT_prom_GR, 
                                           ignore.strand = TRUE)$external_gene_name
cent <- as.data.frame(cent)

CTmeth_inCD4 <- rbind(newborn, middle, cent)
rm(newborn, middle, cent)
CTmeth_inCD4$age <- factor(CTmeth_inCD4$age, 
                           levels = c("newborn","middle_age","centenarian"))
CTmeth_inCD4[CTmeth_inCD4$number_of_reads == 0,]$methylation_level <- NA

length(unique(CTmeth_inCD4$external_gene_name))

```

We can see here that we have information for all 
`r length(unique(CTmeth_inCD4$external_gene_name))` CT genes. We now have to 
calculate the mean methylation level of each promoter to visualise it and 
compare to PMDs.

Some df that we obtain are empty, their CpG are not covered by the CD4 data, 
and so need to be removed. I also want to remove those who have data in only 
one age category, as we want to compare different age, they are of no need. 
When one of the age is missing, I'm adding it as NA to facilitate the rest 
(mostly gene on chrX and we removed data from adult)

```{r CT_level}
reorganize_data_genes <- function(gene_name, data){
  reorganized_data <- data %>%
    filter(external_gene_name == gene_name) %>% 
    dplyr::select("start","methylation_level","age") %>% 
    na.omit() %>%  # to remove CpG rows with NA meth_level
    pivot_wider(names_from = start, values_from = methylation_level) %>% 
    column_to_rownames("age")
}

CT_per_gene_CD4_datasets <- sapply(unique(CTmeth_inCD4$external_gene_name), 
                                   reorganize_data_genes, data = CTmeth_inCD4)

p <- rep(NA, length(CT_per_gene_CD4_datasets))
for(i in 1:length(CT_per_gene_CD4_datasets)){
 p[i] <- nrow(CT_per_gene_CD4_datasets[[i]]) > 1 & 
   is.data.frame(CT_per_gene_CD4_datasets[[i]])
}
CT_per_gene_CD4_datasets <- CT_per_gene_CD4_datasets[p]

k <- rep(NA, length(CT_per_gene_CD4_datasets))
for(i in 1:length(CT_per_gene_CD4_datasets)){
  k[i] <- nrow(CT_per_gene_CD4_datasets[[i]]) == 2
}
for (gene in names(CT_per_gene_CD4_datasets[k])){
  tmp <- rep(NA, ncol(CT_per_gene_CD4_datasets[[gene]]))
  CT_per_gene_CD4_datasets[[gene]] <- rbind(CT_per_gene_CD4_datasets[[gene]], tmp)
  rownames(CT_per_gene_CD4_datasets[[gene]])[3] <- 
    c("newborn","middle_age", "centenarian")[(!c("newborn","middle_age","centenarian") %in% 
                                                rownames(CT_per_gene_CD4_datasets[[gene]]))]
  # naming the new row
}

for (gene in names(CT_per_gene_CD4_datasets)){
  CT_per_gene_CD4_datasets[[gene]] <- arrange(CT_per_gene_CD4_datasets[[gene]], 
                                              rownames(CT_per_gene_CD4_datasets[[gene]]))
  # Put rows in the right order
}

rm(p, i, k, tmp)

CT_meth_means_pre <- sapply(CT_per_gene_CD4_datasets, rowMeans, na.rm = TRUE)

CT_meth_means <- CT_meth_means_pre %>%
  t() %>% 
  as.data.frame() %>%
  rownames_to_column("external_gene_name")

CT_meth_means$middle_age[is.na(CT_meth_means$middle_age)] <- NA

n_means_meth <- CT_meth_means %>% 
  filter(external_gene_name %in% CT_meth_regulated$external_gene_name) %>% 
  nrow()

CT_genes <- left_join(CT_genes, CT_meth_means)
```

On the `r dim(CT_genes)[1]` CT genes, I have enough data to analyse 
`r ncol(CT_meth_means_pre)` of them (`r n_means_meth`/`r dim(CT_meth_regulated)[1]` 
meth regulated).

```{r all_CT_plot}
CT_genes_to_plot <- CT_genes %>% 
  dplyr::select(external_gene_name, chr, X_linked, regulated_by_methylation, 
                PMD, newborn, middle_age, centenarian) %>%
  filter(!is.na(PMD)) %>% 
  pivot_longer(names_to = "age", values_to = "methylation_level", 
               - c("external_gene_name","chr","X_linked", 
                   "regulated_by_methylation", "PMD"))
CT_genes_to_plot$age <- factor(CT_genes_to_plot$age, 
                               levels = c("newborn","middle_age","centenarian"))
CT_genes_to_plot %>%
  mutate(PMD = case_when(PMD == TRUE ~ "PMD",
                         PMD == FALSE ~ "not_PMD")) %>%
  mutate(regulated_by_methylation = case_when(regulated_by_methylation == TRUE ~ "meth_reg",
                                              regulated_by_methylation == FALSE ~ "not_meth_reg")) %>% 
  ggplot(aes(x = age, y = methylation_level, fill = age)) +
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(aes(color = age), alpha = 0.3) +
  facet_grid(regulated_by_methylation ~ PMD) + 
  scale_fill_manual(values = c("rosybrown1", "palevioletred2", "violetred3")) +
  scale_color_manual(values = c("rosybrown3","palevioletred3","violetred4")) +
  theme_bw()

```

### Comparison CT and PMDs

```{r PMD_CD4, warning = FALSE}
PMD_newborn <- mergeByOverlaps(CD4_newborn_GR, my_PMD_GM23248_GR, ignore.strand = TRUE)[,1]
PMD_newborn$name <- mergeByOverlaps(CD4_newborn_GR, my_PMD_GM23248_GR, ignore.strand = TRUE)$name
PMD_newborn <- as.data.frame(PMD_newborn)
PMD_middle <- mergeByOverlaps(CD4_middle_GR, my_PMD_GM23248_GR, ignore.strand = TRUE)[,1]
PMD_middle$name <- mergeByOverlaps(CD4_middle_GR, my_PMD_GM23248_GR, ignore.strand = TRUE)$name
PMD_middle <- as.data.frame(PMD_middle)
PMD_cent<- mergeByOverlaps(CD4_cent_GR, my_PMD_GM23248_GR, ignore.strand = TRUE)[,1]
PMD_cent$name <- mergeByOverlaps(CD4_cent_GR, my_PMD_GM23248_GR, ignore.strand = TRUE)$name
PMD_cent <- as.data.frame(PMD_cent)

PMD_CD4 <- rbind(PMD_newborn, PMD_middle, PMD_cent)
rm(PMD_newborn, PMD_middle, PMD_cent)
PMD_CD4$age <- factor(PMD_CD4$age, levels = c("newborn","middle_age","centenarian"))
PMD_CD4[PMD_CD4$number_of_reads == 0,]$methylation_level <- NA
```

Let's compare the evolution of CT methylation with that of the PMDs they're in.

```{r CT_PMD_plot}
CT_reg_PMD_plot <- CT_genes_to_plot %>% 
  filter(regulated_by_methylation & PMD) %>% 
  ggplot(aes(x = methylation_level, color = age)) +
  geom_density() + 
  scale_color_manual(values = c("rosybrown1", "palevioletred2", "violetred3"))+
  xlab ("Promoter mean methylation level (CT meth reg and in PMDs)") +
  theme_minimal()

CT_reg_PMD_GR <- CT_prom_GR[!is.na(CT_prom_GR@elementMetadata$PMD) &
                              CT_prom_GR@elementMetadata$PMD & 
                              CT_prom_GR@elementMetadata$regulated_by_methylation, ]

CT_PMD_meth_plot <- PMD_CD4 %>% 
  filter(name %in% subsetByOverlaps(my_PMD_GM23248_GR,CT_reg_PMD_GR)$name) %>% 
  group_by(name, age) %>% 
  summarise("mean_methylation" = mean(methylation_level, na.rm = TRUE)) %>% 
  ggplot(aes(x = mean_methylation, color = age)) + 
  geom_density() +
  scale_color_manual(values = c("rosybrown1", "palevioletred2", "violetred3")) +
  theme_minimal() +
  xlab("CT's PMDs mean methylation level")

(CT_reg_PMD_plot + scale_x_continuous(limits = c(0,1))) / (CT_PMD_meth_plot + scale_x_continuous(limits = c(0,1)))

```

### Comparison CT and non CT in PMDs

We don't know if what we see happening in CT genes is common with other genes 
located in PMDs. If we can also see a demethylation of these genes promoters 
and that it is bigger than CT demethylation, it means that there would be a 
protection of CT genes against demethylation. We thus want to compare CT 
methylation to the one of other genes in PMDs.

#### Non-CT methylation
Get methylation data for all other genes than CT in PMDs.

```{r nonCT_PMD, warning = FALSE}
non_CT_genes <- biomaRt_query %>% 
  mutate(start_prom = case_when(strand == 1 ~ transcription_start_site - 1000,
                                strand == -1 ~ transcription_start_site - 500)) %>% 
  mutate(end_prom = case_when(strand == 1 ~ transcription_start_site + 500,
                              strand == -1 ~ transcription_start_site + 1000)) %>% 
  filter(!external_gene_name %in% CT_genes$external_gene_name) %>% 
  filter(transcript_is_canonical == 1) %>%
  filter(chromosome_name%in%c(1:22, "X", "Y"))

non_CT_genes$strand <- ifelse(non_CT_genes$strand == "1", "+", "-")
seqlevelsStyle(non_CT_genes$chromosome_name) <- "UCSC"
non_CT_GR <- makeGRangesFromDataFrame(non_CT_genes,
                                      start.field = "start_prom",
                                      end.field = "end_prom",
                                      keep.extra.columns = TRUE)
non_CT_in_PMD <- subsetByOverlaps(non_CT_GR, my_PMD_GM23248_GR)

newborn <- mergeByOverlaps(CD4_newborn_GR, non_CT_in_PMD, ignore.strand = TRUE)[,1]
newborn$ensembl_gene_id <- mergeByOverlaps(CD4_newborn_GR, non_CT_in_PMD, 
                                           ignore.strand = TRUE)$ensembl_gene_id
newborn <- as.data.frame(newborn)
middle <- mergeByOverlaps(CD4_middle_GR, non_CT_in_PMD, ignore.strand = TRUE)[,1]
middle$ensembl_gene_id <- mergeByOverlaps(CD4_middle_GR, non_CT_in_PMD, 
                                          ignore.strand = TRUE)$ensembl_gene_id
middle <- as.data.frame(middle)
cent <- mergeByOverlaps(CD4_cent_GR, non_CT_in_PMD, ignore.strand = TRUE)[,1]
cent$ensembl_gene_id <- mergeByOverlaps(CD4_cent_GR, non_CT_in_PMD, 
                                        ignore.strand = TRUE)$ensembl_gene_id
cent <- as.data.frame(cent)

nonCTmeth_inCD4 <- rbind(newborn, middle, cent)
rm(newborn, middle, cent)
nonCTmeth_inCD4$age <- factor(nonCTmeth_inCD4$age, 
                           levels = c("newborn","middle_age","centenarian"))
nonCTmeth_inCD4[nonCTmeth_inCD4$number_of_reads == 0,]$methylation_level <- NA

reorganize_data_ensembl <- function(gene_name, data){
  reorganized_data <- data %>%
    filter(ensembl_gene_id == gene_name) %>% 
    dplyr::select("start","methylation_level","age") %>% 
    na.omit() %>%  # to remove CpG rows with NA meth_level
    pivot_wider(names_from = start, values_from = methylation_level) %>% 
    column_to_rownames("age")
}

nonCT_per_gene_CD4_datasets <- bplapply(unique(nonCTmeth_inCD4$ensembl_gene_id),
                                      reorganize_data_ensembl, data = nonCTmeth_inCD4)
names(nonCT_per_gene_CD4_datasets) <- unique(nonCTmeth_inCD4$ensembl_gene_id)

p <- rep(NA, length(nonCT_per_gene_CD4_datasets))
for(i in 1:length(nonCT_per_gene_CD4_datasets)) {
 p[i] <- nrow(nonCT_per_gene_CD4_datasets[[i]]) > 1 &
   is.data.frame(nonCT_per_gene_CD4_datasets[[i]])
}
nonCT_per_gene_CD4_datasets <- nonCT_per_gene_CD4_datasets[p]

k <- rep(NA, length(nonCT_per_gene_CD4_datasets))
for(i in 1:length(nonCT_per_gene_CD4_datasets)) {
  k[i] <- nrow(nonCT_per_gene_CD4_datasets[[i]]) == 2
}

for (gene in names(nonCT_per_gene_CD4_datasets[k])) {
  tmp <- rep(NA, ncol(nonCT_per_gene_CD4_datasets[[gene]]))
  nonCT_per_gene_CD4_datasets[[gene]] <- rbind(nonCT_per_gene_CD4_datasets[[gene]], tmp)
  rownames(nonCT_per_gene_CD4_datasets[[gene]])[3] <-
    c("newborn","middle_age", "centenarian")[(!c("newborn","middle_age","centenarian") %in%
                                                rownames(nonCT_per_gene_CD4_datasets[[gene]]))]
  # naming the new row
}

for (gene in names(nonCT_per_gene_CD4_datasets)) {
  nonCT_per_gene_CD4_datasets[[gene]] <- arrange(nonCT_per_gene_CD4_datasets[[gene]],
                                              rownames(nonCT_per_gene_CD4_datasets[[gene]]))
  # Put rows in the right order
}

rm(p, i, k, tmp)

#load("data_rmd/nonCT_per_gene_CD4_datasets.rda")

nonCT_meth_means <- sapply(nonCT_per_gene_CD4_datasets, rowMeans, na.rm = TRUE)
nonCT_meth_means <- nonCT_meth_means %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("ensembl_gene_id")

nonCT_meth_means$middle_age[is.na(nonCT_meth_means$middle_age)] <- NA
```

#### CT and non CT comparison

```{r non-CT_plot}
threshold <- CT_genes %>% 
  filter(regulated_by_methylation & PMD) %>% 
  dplyr::select(newborn) %>% 
  min(na.rm = TRUE)

nonCT_meth_means_thr <- nonCT_meth_means %>% 
  filter(newborn >= threshold)

nonCT_meth_means_to_plot <-  nonCT_meth_means_thr %>% 
  pivot_longer(names_to = "age", values_to = "methylation_level", 
               - "ensembl_gene_id")
nonCT_meth_means_to_plot$age <- factor(nonCT_meth_means_to_plot$age, 
                               levels = c("newborn","middle_age","centenarian"))

med_meth_nb_CT <- CT_genes_to_plot %>% 
  filter(regulated_by_methylation & PMD & age == "newborn") %>% 
  pull(methylation_level) %>% 
  median(na.rm = TRUE)
med_meth_nb_CT

med_meth_nb_nonCT <-nonCT_meth_means_to_plot %>% 
  filter(age == "newborn") %>% 
  pull(methylation_level) %>% 
  median(na.rm = TRUE)
med_meth_nb_nonCT

med_meth_cent_CT <- CT_genes_to_plot %>% 
  filter(regulated_by_methylation & PMD & age == "centenarian") %>% 
  pull(methylation_level) %>% 
  median(na.rm = TRUE)
med_meth_cent_CT

med_meth_cent_nonCT <-nonCT_meth_means_to_plot %>% 
  filter(age == "centenarian") %>% 
  pull(methylation_level) %>% 
  median(na.rm = TRUE)
med_meth_cent_nonCT

CT_reg_PMD_boxplot <- CT_genes_to_plot %>%
  filter(regulated_by_methylation & PMD) %>% 
  ggplot(aes(x = age, y = methylation_level, fill = age)) +
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(aes(color = age), alpha = 0.3) +
  scale_fill_manual(values = c("rosybrown1", "palevioletred2", "violetred3")) +
  scale_color_manual(values = c("rosybrown3","palevioletred3","violetred4")) +
  theme_bw()

nonCT_boxplot <- nonCT_meth_means_to_plot %>%
  ggplot(aes(x = age, y = methylation_level, fill = age)) +
  geom_jitter(aes(color = age), alpha = 0.3) +
  geom_boxplot(outlier.shape = NA) +
  scale_fill_manual(values = c("rosybrown1", "palevioletred2", "violetred3")) +
  scale_color_manual(values = c("rosybrown3","palevioletred3","violetred4")) +
  theme_bw()

(CT_reg_PMD_boxplot + ggtitle("CT genes in PMDs") + 
    scale_y_continuous(limits = c(0,1)) + theme(legend.position = "none")) + 
  (nonCT_boxplot + ggtitle("non CT genes in PMDs") + 
     scale_y_continuous(limits = c(0,1)))
```

#### Differential analysis

However, this is only an observation based on the data, we need to do a 
quantification of this decrease and see its signification. For this, we have to 
do a differential methylation analysis.

First step is to check that there is indeed a loss between newborn and 
centenarian for each gene. To do so, we use a unidirectional (newborn > 
centenarian) paired statistical test on the methylation level of cytosines in 
both ages. Data can't really be normalise so we use a Wilcoxon test 
(non-parametric t test). Our significativity threshold is 5%, which means that 
a test is significative if the adjusted (B-H method) pvalue is inferior to 0.05.

Second step is to quantify the loss of methylation between newborn and 
centenarian. To do so, we calculate a delta_meth for each gene, which is 
simply the difference between newborn and cent level.

```{r diff_analysis}
CT_to_pval <- CTmeth_inCD4 %>%
  filter(external_gene_name %in% filter(CT_genes, 
                                        PMD & regulated_by_methylation)$external_gene_name) %>%
  dplyr::select(external_gene_name, seqnames, start, age, methylation_level) %>% 
  pivot_wider(names_from = "age", values_from = "methylation_level", values_fill = NA) %>% 
  filter(!is.na(newborn) & !is.na(centenarian)) %>% 
  left_join(dplyr::select(CT_genes, external_gene_name, ensembl_gene_id))

non_CT_to_pval <- nonCTmeth_inCD4 %>% 
  filter(ensembl_gene_id %in% nonCT_meth_means_thr$ensembl_gene_id) %>%
  dplyr::select(ensembl_gene_id, seqnames, start, age, methylation_level) %>%
  pivot_wider(names_from = "age", values_from = "methylation_level", values_fill = NA) %>% 
  filter(!is.na(newborn) & !is.na(centenarian))

test_nb_cent <- function(gene_i, data){
    wilcox.test(
      as.matrix(filter(data, ensembl_gene_id == gene_i)$newborn), 
      as.matrix(filter(data, ensembl_gene_id == gene_i)$centenarian), 
      paired = TRUE, 
      alternative = "greater", 
      exact = FALSE)$p.value
}

pval_CT <- sapply(unique(CT_to_pval$ensembl_gene_id), test_nb_cent, CT_to_pval)
pval_nonCT <- sapply(unique(non_CT_to_pval$ensembl_gene_id), test_nb_cent, non_CT_to_pval)

CT_stat_analysis <- CT_genes %>% 
  filter(ensembl_gene_id %in% CT_to_pval$ensembl_gene_id) %>% 
  dplyr::select(ensembl_gene_id, external_gene_name, chr, X_linked, 
                 newborn, middle_age, centenarian) %>% 
  filter(!is.na(newborn) & !is.na(centenarian)) %>% 
  mutate("padj" = p.adjust(pval_CT, method = "BH"))
sign_table_CT <- table(CT_stat_analysis$padj < 0.05)

nonCT_stat_analysis <- nonCT_meth_means_thr %>% 
  filter(ensembl_gene_id %in% non_CT_to_pval$ensembl_gene_id) %>% 
  mutate("padj" = p.adjust(pval_nonCT, method = "BH"))
sign_table_nonCT <- table(nonCT_stat_analysis$padj < 0.05)

table(CT_stat_analysis$padj <= 0.05)[[2]]/length(CT_stat_analysis$padj)
table(nonCT_stat_analysis$padj <= 0.05)[[2]]/length(nonCT_stat_analysis$padj)

CT_stat_analysis <- CT_stat_analysis %>% 
  mutate("delta_meth" = centenarian - newborn)

nonCT_stat_analysis <- nonCT_stat_analysis %>% 
  mutate("delta_meth" = centenarian - newborn)

med_delta_CT <- CT_stat_analysis %>% 
  filter(padj <= 0.05) %>% 
  pull(delta_meth) %>% 
  median()
med_delta_CT

med_delta_nonCT <- nonCT_stat_analysis %>% 
  filter(padj <= 0.05) %>% 
  pull(delta_meth) %>% 
  median()
med_delta_nonCT

CT_stat_analysis %>% 
  filter(padj <= 0.05) %>%
  mutate(type = "CT_genes") %>%
  dplyr::select(ensembl_gene_id, centenarian, middle_age, newborn, padj, delta_meth, type) %>% 
  rbind(nonCT_stat_analysis %>%
          filter(padj <= 0.05) %>%
          mutate(type = "nonCT_genes")) %>%

  ggplot(aes(y = delta_meth, x = type, fill = type)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(color = type), alpha = 0.2, size = 0.5) +
  scale_fill_manual(values = c("plum","lightgoldenrod1")) +
  scale_color_manual(values = c("darkorchid3","gold1")) +
  ggtitle("Delta_meth of PMD genes with significant demethylation") +
  xlab("") + 
  ylab("Delta_meth centenarian - newborn") +
  theme_bw() +
  theme(legend.position = "none")

```

There is indeed a difference in the amplitude of delta_meth between CT 
and nonCT genes in PMD.

We then want to test the difference between delta_meth for CT and nonCT genes
in PMD, when pval is significative. For that we use a unidirectionnal (CT > 
non-CT) Mann-Whitney test, non-parametric (because no normality in delta_meth 
non-CT) version of student t test.

```{r delta_test}
wilcox.test(filter(CT_stat_analysis, padj <= 0.05)$delta_meth,
            filter(nonCT_stat_analysis, padj <= 0.05)$delta_meth,
            alternative = "greater")
```

pvalue here is significative. This shows then that delta_meth are smaller in nonCT genes and demethylation in nonCT genes is then higher than in CT genes.

However, the difference between delta_meth distributions could be the result of 
the biggest number of genes in nonCT (`r sign_table_nonCT[[1]]` VS 
`r sign_table_CT[[1]]` not sign, or `r sign_table_nonCT[[2]]` VS 
`r sign_table_CT[[2]]` padj sign). In order to see that, let's do a down 
sampling : taking `r sign_table_CT[[2]]` nonCT genes with a pval <= 0.05 and 
`r sign_table_CT[[1]]` with a pval > 0.05 a thousand times and calculate the 
mean of each distributions. I can visualise that after and see where CT genes 
goes compare to that.

```{r down_sampling}
random_delta_distrib_means <- function(i, data, n_sign, n_notsign){
  random_sample1 <- sample_n(filter(data, padj <= 0.05), n_sign)
  random_sample2 <- sample_n(filter(data, padj > 0.05), n_notsign)
  random_sample <- rbind(random_sample1, random_sample2)

  mean(random_sample$delta_meth)
}

distrib_means <- unlist(bplapply(1:1000, random_delta_distrib_means,
                                data = nonCT_stat_analysis, 
                                n_sign = sign_table_CT[[2]], 
                                n_notsign = sign_table_CT[[1]]))

mean(CT_stat_analysis$delta_meth)

table(distrib_means >= mean(CT_stat_analysis$delta_meth))

hist(distrib_means,
     col="dodgerblue1",
     border = "dodgerblue3",
     breaks = 20,
     main = paste0("Means of 1000 delta_meth distributions of nonCT genes when randomly 
     taking ", sign_table_CT[[2]], "significative and ", sign_table_CT[[1]], " not significative genes"),
     cex.main = 0.75,
     xlab = "Means",
     ylab = "Frequency",
     xlim = c(-0.2, 0))
abline(v = mean(CT_stat_analysis$delta_meth), col = "violetred1", lwd = 3)
text(mean(CT_stat_analysis$delta_meth), 125, "Mean of CT genes 
delta_meth distribution", col = "violetred3")

```

This shows that even with the same amount of genes, delta_meth are 
significantly smaller for CT genes than for nonCT genes, the effect is then not
due to the amount of genes.

This analysis shows that CT genes located in PMDs undergo a demethylation 
related to aging that is smaller than the other genes in PMDs. Thus, age and 
demethylation that is associated to it couldn't explain demethylation and 
activation of CT genes. There is then probably something that protects CT genes 
from PMD demethylation.

## Why is there a protection on CT genes : CpG density hypothesis

CT genes are thus protected against demethylation in PMDs. Next question is to 
determine what confers this protection to CT genes. what One of our hypothesis 
for that protection is that CT genes are protected thanks to their density in CpG.
Indeed, CT genes promoters are rich in CpGs while PMDs are poor. 
It is possible that this density protects CT genes from demethylation.

Promoter's density in CpG is already computed in CTexploreR but we have to do 
it also for non-CT genes.

```{r compute_cpg}
CT_density_analysis <- CT_genes %>% 
  filter(ensembl_gene_id %in% CT_stat_analysis$ensembl_gene_id) %>%
  mutate("padj" = CT_stat_analysis$padj) %>% 
  mutate("delta_meth" = CT_stat_analysis$delta_meth)

nonCT_density_analysis <- as.data.frame(non_CT_in_PMD) %>% 
  filter(ensembl_gene_id %in% nonCT_stat_analysis$ensembl_gene_id)%>%
  left_join(nonCT_stat_analysis)
colnames(nonCT_density_analysis)[1:3] <- c("chr", "start_prom", "end_prom")
nonCT_density_analysis$chr <- as.character(nonCT_density_analysis$chr)

cpg_density <- function(gene, data){
  temp <- data %>% 
    filter(ensembl_gene_id == gene) %>% 
    dplyr::select(ensembl_gene_id, start_prom, end_prom, chr)

  seqlevelsStyle(temp$chr) <- "NCBI"

  chrom <- genome[startsWith(names(genome), paste0(temp$chr, " "))][[1]]

  interest <- subseq(chrom, temp$start_prom, temp$end_prom)

  p <- matchPattern("CG", interest)
  length(p) / length(interest)
}

genome <- readDNAStringSet("../../cluster/GENOMES/Homo_sapiens.GRCh38.dna.primary_assembly.fa")

CT_cpg_density <- sapply(CT_density_analysis$ensembl_gene_id, cpg_density, 
                         CT_density_analysis)

nonCT_cpg_density <- unlist(bplapply(nonCT_density_analysis$ensembl_gene_id, cpg_density,
                            nonCT_density_analysis))
```

There is already the CpG_density column from CT_genes so careful, what I've done
here is similar to what was already in it, it confirms.

### Comparison CT and non CT with similar densities

```{r cpg_plot}
CT_density_analysis <- CT_density_analysis %>% 
  mutate("CpG_density_analysis" = CT_cpg_density*100)

nonCT_density_analysis <- nonCT_density_analysis %>% 
  mutate("CpG_density_analysis" = nonCT_cpg_density*100)

CT_density_analysis %>% 
  filter(padj <= 0.05) %>%
  mutate(type = "CT_genes") %>%
  dplyr::select(ensembl_gene_id, centenarian, middle_age, newborn, padj, 
                delta_meth, CpG_density_analysis, type) %>% 
  rbind(nonCT_density_analysis %>%
          filter(padj <= 0.05) %>%
          mutate(type = "nonCT_genes") %>% 
          dplyr::select(ensembl_gene_id, centenarian, middle_age, newborn, padj, 
                delta_meth, CpG_density_analysis, type)) %>%

  ggplot(aes(x = CpG_density_analysis, color = type)) +
  geom_density() + 
  scale_color_manual(values = c("darkorchid3","gold1")) +
  xlab("CpG density in promoters") +
  ylab("Quantity (density plot)") +
  ggtitle("Distribution of CpG density on CT and non CT promoters
significantly demethylated in PMDs")+
  theme_minimal()

```

This shows/confirms that nonCT promoters in PMDs are poorer in CpG.

```{r density_deltameth}
CT_density_analysis %>% 
  filter(padj <= 0.05) %>%
  mutate(type = "CT_genes") %>%

  ggplot(aes(x = delta_meth, y = CpG_density_analysis, color = type)) +
  geom_point() +
  geom_smooth(method = "loess", span = 2.5, color = "darkorchid4") +
  scale_color_manual(values = c("darkorchid3")) +
  xlab("Delta_meth of significatively 
demethylated genes") +
  ylab("CpG density of promoter") +
  facet_grid(~ type) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(0, 8)) +
  scale_x_continuous(limits = c(-0.63, 0.05)) +

  nonCT_density_analysis %>%
          filter(padj <= 0.05) %>%
          mutate(type = "nonCT_genes") %>%

  ggplot(aes(x = delta_meth, y = CpG_density_analysis, color = type)) +
  geom_point() +
  geom_smooth(method = "loess", span = 0.75, color = "goldenrod2") +
  scale_color_manual(values = c("gold1")) +
  facet_grid(~ type) +
  theme_bw() + 
  ylab("") +
  xlab("Delta_meth of significatively 
demethylated genes") +
  scale_y_continuous(limits = c(0, 8)) +
  scale_x_continuous(limits = c(-0.63, 0.05)) +  
  theme(legend.position = "none")
```

We see here a diagonal tendancy with a positive correlation, it shows that there
is a link between CpG density and delta_meth. This tendancy is visible in CT and
nonCT : the more CpG there are in promoters, less demethylation we observe. 
Isolated CpGs are then probably those that are easier demethylated in PMDs. We 
thus propose that the strong density in CpG that characterize CT promoters is the 
origin of their resistance to demethylation

Let's compare only similar CpG densities, and do the same categories as in 
CTexploreR.

```{r cpg_category}
CT_density_analysis <- CT_density_analysis %>% 
  mutate(density_state = 
             case_when (CT_density_analysis$CpG_density_analysis >= 4 ~ "rich",
                        CT_density_analysis$CpG_density_analysis <= 2 ~ "poor",
                        CT_density_analysis$CpG_density_analysis > 2 & 
                          CT_density_analysis$CpG_density_analysis < 4 ~ "normal"))

nonCT_density_analysis <- nonCT_density_analysis %>% 
  mutate(density_state = 
             case_when (nonCT_density_analysis$CpG_density_analysis >= 4 ~ "rich",
                        nonCT_density_analysis$CpG_density_analysis <= 2 ~ "poor",
                        nonCT_density_analysis$CpG_density_analysis > 2 &
                          nonCT_density_analysis$CpG_density_analysis < 4 ~ "normal"))
```

By comparing only delta_meth from genes normal/rich in CpG, we can erase the 
effect of CpG density from the demethylation analysis.

```{r delta_meth_rich_plot}
CT_density_analysis %>% 
  filter(padj <= 0.05) %>%
  filter(density_state != "poor") %>% 
  mutate(type = "CT_genes") %>%
  dplyr::select(ensembl_gene_id, delta_meth, type) %>% 
  rbind(nonCT_density_analysis %>%
          filter(padj <= 0.05) %>%
          filter(density_state != "poor") %>%
          mutate(type = "nonCT_genes")%>%
          dplyr::select(ensembl_gene_id, delta_meth, type))%>%

  ggplot(aes(y = delta_meth, x = type, fill = type)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(color = type), alpha = 0.2, size = 0.5) +
  scale_fill_manual(values = c("plum","lightgoldenrod1")) +
  scale_color_manual(values = c("darkorchid3","gold1")) +
  ggtitle("Delta_meth of PMD genes with significant demethylation
and rich/normal in CpG") +
  xlab("") + 
  ylab("Delta_meth centenarian - newborn") +
  theme_bw() +
  theme(legend.position = "none")

rich_med_delta_CT <- CT_density_analysis %>% 
  filter(padj <= 0.05 & density_state != "poor") %>% 
  pull(delta_meth) %>% 
  median()
rich_med_delta_CT

rich_med_delta_nonCT <-nonCT_density_analysis %>% 
  filter(padj <= 0.05 & density_state != "poor") %>% 
  pull(delta_meth) %>% 
  median()
rich_med_delta_nonCT
```

Medians are now way closer to each other and we can see that distributions are 
closer to each other. For CT genes, the median is still a loss of 
`r abs(rich_med_delta_CT)*100`%, however for nonCT the median loss has reduce 
to `r abs(rich_med_delta_nonCT)*100`% ! The two distributions are now clearly 
comparable.
It looks like the demethylation difference that we observed earlier between CT 
and nonCT genes in PMD during aging is now gone when we only analyse gene whose 
promoters are not poor in CpG.

Let's do the same test as previously to see if there is a difference between 
both distributions (Mann-Withney) and if it isn't due to the size of the sample.

```{r delta_meth_rich_stat}
wilcox.test(filter(CT_density_analysis, padj <= 0.05 & density_state != "poor")$delta_meth,
            filter(nonCT_density_analysis, padj <= 0.05 & density_state != "poor")$delta_meth,
            alternative = "greater")

rich_CT_sign <- table(filter(CT_density_analysis, density_state != "poor")$padj <= 0.05)

distrib_means_rich <- unlist(bplapply(1:1000, random_delta_distrib_means,
                                data = filter(nonCT_density_analysis, density_state != "poor"), 
                                n_sign = rich_CT_sign[[2]], n_notsign = rich_CT_sign[[1]]))

mean_delta_rich <- mean(filter(CT_density_analysis, density_state != "poor")$delta_meth)
mean_delta_rich

table(distrib_means_rich >= mean(filter(CT_density_analysis, 
                                        density_state != "poor")$delta_meth))

hist(distrib_means_rich,
     col="dodgerblue1",
     border = "dodgerblue3",
     breaks = 20,
     main = paste0("Means of 1000 delta_meth distributions of nonCT genes when randomly 
     taking ", rich_CT_sign[[2]], " significative and ", rich_CT_sign[[1]], " not significative genes"),
     cex.main = 0.75,
     xlab = "Means",
     ylab = "Frequency",
     xlim = c(-0.15, 0))
abline(v = mean_delta_rich, col = "violetred1", lwd = 3)
text(mean_delta_rich, 125, "Mean of rich CT genes 
delta_meth distribution", col = "violetred3")
```

pvalue is not significative anymore which means that there is ano  difference 
between both distribution.

Also, we can see when doing 1000 mean of delta_meth with the same amount of 
genes that the mean for CT genes is in the middle of the distribution, which
means that there is no real difference between CT and nonCT demethylation when 
taking the CpG density into account

It thus looks like CpG density plays a big role in the protection as when we're 
taking that into account in our analysis, the difference between CT and nonCT 
disappears.

When analysing genes with comlparable CpG density, there is no difference in the 
demethylation of CT and nonCT genes in PMD. This shows that the richness in CpG 
of the promoter gives a certain protection against PMD demethylation during aging.

### solo-CpG methylation

In summary, although CT genes are found in PMDs, they lose little methylation 
and are protected from continued demethylation during aging. This protection 
seems to be explained by the CpG density of the promoters. The demethylation of 
CT genes visible in tumours is therefore not associated with cell proliferation.
However, the question remains: why does a high CpG density protect against
continued demethylation? Our hypothesis is that there may be a self-maintenance 
system, i.e. when a CpG loses its methylation in a CpG-rich region, the other 
still methylated CpG methylated CpGs attract the methylation maintenance 
machinery, thereby promoting the remethylation of the demethylated site. It 
should be noted that our observations are consistent with some analyses performed
by Zhou et al. (2018) which showed that that the hypomethylation of PMDs was 
better observed by focusing only on the methylation of certain CpGs, called 
"solo-CpGs", corresponding to isolated CpGs at least 35 bp away from a 
neighbouring CpG and flanked by a A or a T (WCGW).

We're going to use Zhou's coordinate of solo-WCGW to look a bit more into that.

```{r wcgw_bed, message = FALSE}
solo_WCGW <- read_delim("data/zhou_solo_WCGW_hg38.bed", 
    delim = "\t", escape_double = FALSE, 
    col_names = FALSE, trim_ws = TRUE)
solo_WCGW <- solo_WCGW[, -5]
colnames(solo_WCGW) <- c("seqnames", "start", "end", "context")
```

```{r solo}
CT_with_solo <- CTmeth_inCD4 %>%
  filter(external_gene_name %in% filter(CT_genes, 
                                        PMD & regulated_by_methylation)$external_gene_name) %>%
  dplyr::select(external_gene_name, seqnames, start, age, methylation_level) %>% 
  left_join(dplyr::select(CT_genes, external_gene_name, ensembl_gene_id)) %>% 
  left_join(solo_WCGW)

CT_with_solo <- CT_with_solo %>% 
  mutate(solo_state = case_when(is.na(CT_with_solo$context) ~ "not-solo",
                                !is.na(CT_with_solo$context) ~ "solo-CpG")) %>% 
  mutate(type = "CT_genes")

nonCT_with_solo <- nonCTmeth_inCD4 %>% 
  filter(ensembl_gene_id %in% nonCT_meth_means_thr$ensembl_gene_id) %>%
  dplyr::select(ensembl_gene_id, seqnames, start, age, methylation_level) %>%
  left_join(solo_WCGW)

nonCT_with_solo <- nonCT_with_solo %>% 
  mutate(solo_state = case_when(is.na(nonCT_with_solo$context) ~ "not-solo",
                                !is.na(nonCT_with_solo$context) ~ "solo-CpG"))%>% 
  mutate(type = "nonCT_genes")

CT_with_solo %>% 
  dplyr::select(colnames(nonCT_with_solo)) %>% 
  rbind(nonCT_with_solo) %>%

  ggplot(aes(y = methylation_level, x = age, 
             fill = solo_state)) +
  geom_jitter(aes(color = age), alpha = 0.2, size = 0.5) +
  geom_boxplot(outlier.shape = NA) +
  scale_fill_manual(values = c("lightskyblue1", "darkseagreen2")) +
  scale_color_manual(values = c("rosybrown3","palevioletred3","violetred4")) +
  ggtitle("CpG methylation in all ages, depending
on their solo state, in CT and nonCT genes") +
  facet_grid(type ~ solo_state) +
  theme_bw() 
```

What we see here confirms what zhou said and what we hypothesize, we can see 
that solo-cpg are losing more methylation following aging.

# SessionInfo

```{r}
sessionInfo()
```

### 2023-08-21 Paper_code.Rmd
Added numbers to gene function

@@
 library(clusterProfiler)
+library(msigdbr)
 library(DOSE)
@@
 ```{r GO_ORA}
+
+msigdbr(species = "Homo sapiens" , category = "C5") %>% 
+  filter(gene_symbol %in% CT_genes$external_gene_name) %>% 
+  pull(gene_symbol) %>% 
+  unique() %>% 
+  length()
+
+msigdbr(species = "Homo sapiens" , category = "H") %>% 
+  filter(gene_symbol %in% CT_genes$external_gene_name) %>% 
+  pull(gene_symbol) %>% 
+  unique() %>% 
+  length()
+
 go_ora <- enrichGO(gene = CT_genes$ensembl_gene_id,
@@
 
-Now that these overlapping have been done, I can report them into my data as a boulean column to use analyse it.
+Now that these overlapping have been done, I can report them into my data as a 
+boulean column to use analyse it.
 
@@
 
-Now that we know that CT genes are indeed enriched in PMDs, we have to go further.
-Indeed, we know that proliferation has an impact on PMD and their methylation : there are losses of methylation with time.
-As CT genes are located in PMDs, we can hypothesise that their methylation might be impacted by the losses happening during proliferation.
-We thus want to focus on the impact on CT genes of the methylation changes happening on PMDss in proliferative or aging cells.
+Now that we know that CT genes are indeed enriched in PMDs, we have to go 
+further. Indeed, we know that proliferation has an impact on PMD and their 
+methylation : there are losses of methylation with time. As CT genes are located
+in PMDs, we can hypothesise that their methylation might be impacted by the 
+losses happening during proliferation. We thus want to focus on the impact on CT
+genes of the methylation changes happening on PMDss in proliferative or aging 
+cells.
 
@@
 
-Bisulfite-Seq data coming from a male newborn, a middle age donor and a male centenarian.
-
-Quality of the data is good, with a mean of 13 reads per cytosine.
-However, middle age data has no info about the sex of the donor.
-By comparing the number of reads by chromosome in this sample, a female one and the male ones, we saw that the proportion of reads on chromosomes X and Y showed that the middle age data come from a woman.
-We then have to be very careful when analysing the X chromosome with the middle age data.
-To facilitate the analysis and compare all samples, I’ll remove the adult data on chrX.
+Bisulfite-Seq data coming from a male newborn, a middle age donor and a male 
+centenarian.
+
+Quality of the data is good, with a mean of 13 reads per cytosine. However, 
+middle age data has no info about the sex of the donor. By comparing the number 
+of reads by chromosome in this sample, a female one and the male ones, we saw 
+that the proportion of reads on chromosomes X and Y showed that the middle age 
+data come from a woman. We then have to be very careful when analysing the X 
+chromosome with the middle age data. To facilitate the analysis and compare all
+samples, I’ll remove the adult data on chrX.
 
@@
 
-#load("data_rmd/nonCT_per_gene_CD4_datasets.rda")
+# load("data_rmd/nonCT_per_gene_CD4_datasets.rda")
 
@@
 
-pvalue here is significative. This shows then that delta_meth are smaller in nonCT genes and demethylation in nonCT genes is then higher than in CT genes.
+pvalue here is significative. This shows then that delta_meth are smaller in 
+nonCT genes and demethylation in nonCT genes is then higher than in CT genes.
 

### 2023-08-23 Paper_code-Database_CT.Rmd
Separated full code in two files

--- 2023-08-21 Paper_code.Rmd
@@
 ---
-title: "Code used to generate paper results"
+title: "Code used to generate paper results : Database comparison and CT characteristics"
 author: "Julie Devis"
@@
 library(DOSE)
-library(tibble)
 library(BiocParallel)
@@
 
-# Package application : What are the mechanisms involved in CT demethylation in cancer
-
-## Determination of PMDs in male fibroblasts celles
-
-WGBS data from GM23248 cells (male so X chromosome is reliable) downloaded from 
-ENCODE.
-Data was processed using a WGBS pipeline [MethPipe]
-(http://smithlabresearch.org/software/methpipe/) including [TrimGalore](https://www.bioinformatics.babraham.ac.uk/projects/trim_galore/) 
-to remove low quality reads and trim the adapter from the sequences.
-
-The pipeline has been verified by comparing PMDs already defined by Lister et 
-al. on IMR90 and PMDs obtained with MethPipe.
-
-```{bash download_GM23248, eval = FALSE}
-wget https://www.encodeproject.org/files/ENCFF142GZA/@@download/ENCFF142GZA.fastq.gz
-wget https://www.encodeproject.org/files/ENCFF845RUF/@@download/ENCFF845RUF.fastq.gz
-wget https://www.encodeproject.org/files/ENCFF968GRR/@@download/ENCFF968GRR.fastq.gz
-wget https://www.encodeproject.org/files/ENCFF975YNS/@@download/ENCFF975YNS.fastq.gz
-```
-
-```{bash trim/methpipe_GM23248, eval = FALSE}
-#!/bin/bash
-mkdir /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/rep1_ENCLB777FMV;
-cd /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/rep1_ENCLB777FMV
-/opt/TrimGalore/trim_galore --fastqc --paired /data/cbio/2021-mth-julie-devis/WGBS-processing/GM23248_data/ENCFF142GZA.fastq /data/cbio/2021-mth-julie-devis/WGBS-processing/GM23248_data/ENCFF845RUF.fastq
-
-mkdir /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/rep2_ENCLB436YZI;
-cd /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/rep2_ENCLB436YZI
-/opt/TrimGalore/trim_galore --fastqc --paired /data/cbio/2021-mth-julie-devis/WGBS-processing/GM23248_data/ENCFF975YNS.fastq /data/cbio/2021-mth-julie-devis/WGBS-processing/GM23248_data/ENCFF968GRR.fastq
-
-#!/bin/bash
-
-cd /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/ENCODE_GM23248/rep1_ENCLB777FMV;
-
-#Alignement
-/home/users/jdevis/walt/bin/walt \
--i /data/cbio/2021-mth-julie-devis/WGBS-processing/genomes/hg38_by_Chr.dbindex \
--1 ENCFF142GZA_val_1.fq \
--2 ENCFF845RUF_val_2.fq \
--o rep1_ENCLB777FMV_mapping.mr;
-
-#Duplicate-remover
-LC_ALL=C sort -k 1,1 -k 2,2n -k 3,3n -k 6,6 -o rep1_ENCLB777FMV.mr.sorted_start rep1_ENCLB777FMV_mapping.mr;
-/opt/methpipe/build/duplicate-remover -S rep1_ENCLB777FMV_dremove_stat.txt -o rep1_ENCLB777FMV.mr.dremove rep1_ENCLB777FMV.mr.sorted_start;
-
-#bsrate
-/opt/methpipe/build/bsrate -c /data/cbio/GENOMES/GRCh38_by_Chr/hgdownload.cse.ucsc.edu/chromosomes \
--o rep1_ENCLB777FMV.bsrate rep1_ENCLB777FMV.mr.dremove;
-
-#Methylation levels
-/opt/methpipe/build/methcounts -n \
--c /data/cbio/GENOMES/GRCh38_by_Chr/hgdownload.cse.ucsc.edu/chromosomes \
--o rep1_ENCLB777FMV.meth rep1_ENCLB777FMV.mr.dremove;
-
-#Merge CpG
-/opt/methpipe/build/symmetric-cpgs -o rep1_ENCLB777FMV_CpG_merge.meth rep1_ENCLB777FMV.meth;
-
-#Stats
-/opt/methpipe/build/levels -o rep1_ENCLB777FMV.levels rep1_ENCLB777FMV.meth;
-
-cd /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/ENCODE_GM23248/rep2_ENCLB436YZI;
-
-#Alignement
-/home/users/jdevis/walt/bin/walt \
--i /data/cbio/2021-mth-julie-devis/WGBS-processing/genomes/hg38_by_Chr.dbindex \
--1 ENCFF975YNS_val_1.fq \
--2 ENCFF968GRR_val_2.fq \
--o rep2_ENCLB436YZI_mapping.mr;
-
-#Duplicate-remover
-LC_ALL=C sort -k 1,1 -k 2,2n -k 3,3n -k 6,6 -o rep2_ENCLB436YZI.mr.sorted_start rep2_ENCLB436YZI_mapping.mr;
-/opt/methpipe/build/duplicate-remover -S rep2_ENCLB436YZI_dremove_stat.txt -o rep2_ENCLB436YZI.mr.dremove rep2_ENCLB436YZI.mr.sorted_start;
-
-#bsrate
-/opt/methpipe/build/bsrate -c /data/cbio/GENOMES/GRCh38_by_Chr/hgdownload.cse.ucsc.edu/chromosomes \
--o rep2_ENCLB436YZI.bsrate rep2_ENCLB436YZI.mr.dremove;
-
-#Methylation levels
-/opt/methpipe/build/methcounts -n \
--c /data/cbio/GENOMES/GRCh38_by_Chr/hgdownload.cse.ucsc.edu/chromosomes \
--o rep2_ENCLB436YZI.meth rep2_ENCLB436YZI.mr.dremove;
-
-#Merge CpG
-/opt/methpipe/build/symmetric-cpgs -o rep2_ENCLB436YZI_CpG_merge.meth rep2_ENCLB436YZI.meth;
-
-#Stats
-/opt/methpipe/build/levels -o rep2_ENCLB436YZI.levels rep2_ENCLB436YZI.meth;
-
-#Merge both replicats
-cd /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/ENCODE_GM23248;
-
-/opt/methpipe/build/merge-methcounts rep1_ENCLB777FMV/rep1_ENCLB777FMV_CpG_merge.meth \
-rep2_ENCLB436YZI/rep2_ENCLB436YZI_CpG_merge.meth \
--o GM23248_merged.meth;
-
-#pmd
-/opt/methpipe/build/pmd -v -i 1000 -o GM23248.pmd GM23248_merged.meth
-```
-
-```{r import_meth_GM23248, eval = FALSE}
-GM23248_meth <- read_delim("GM23248_merged.meth", "\t", escape_double = FALSE, col_names = FALSE, trim_ws = TRUE)
-names(GM23248_meth) <- c("chromosome", "position", "strand", "sequence_context","methylation_level","number_of_reads")
-```
-
-We can use the PMD file coming from MethPipe and change it into a GRanges object
-so that we can easily do overlapping analysis
-
-```{r load_PMD, message = FALSE}
-my_PMD_GM23248 <- read_delim("data/GM23248.pmd", "\t", 
-                             escape_double = FALSE, col_names = FALSE, 
-                             trim_ws = TRUE)
-names(my_PMD_GM23248) <- c("chr", "start","end","name_quality","bins","strand")
-my_PMD_GM23248 <- separate(my_PMD_GM23248, "name_quality", 
-                           into = c("name", "likelihood_start", "certainty_start",
-                                    "likelihood_end", "certainty_end"), 
-                           sep = ":")
-my_PMD_GM23248_GR <- makeGRangesFromDataFrame(my_PMD_GM23248 [, -10], keep.extra.columns = TRUE)
-```
-
-Not PMD and NA regions aren’t dissociated. Coverage of windows of 10kb were 
-calculated along the genome. A threshold of 1000 reads per 10kb is necessary to 
-not be considered NA.
-
-```{r all_chr_10kb, eval = FALSE}
-for (chromosome_name in paste0("chr",c(1:22,"X","Y"))){
-
-  roi <- GM23248_meth %>% 
-    filter(chromosome == chromosome_name)
-
-  res <- data.frame(region_i = floor(min(roi$position)/10000):floor(max(roi$position)/10000),
-                    reads_tot = rep(NA,length(floor(min(roi$position)/10000):floor(max(roi$position)/10000))),
-                    chromosome = rep(unique(roi$chromosome),length(floor(min(roi$position)/10000):floor(max(roi$position)/10000))))
-
-  res <- res %>% 
-    mutate(start_region = region_i*10000) %>% 
-    mutate(end_region = region_i*10000+10000) %>% 
-    column_to_rownames("region_i")
-
-  for (i in floor(min(roi$position)/10000):floor(max(roi$position)/10000)){
-    value <-  i*10000
-    tmp <- roi %>% 
-      filter(position >= value & position < (value + 10000))
-    res[as.character(i) , "reads_tot"] <- sum(tmp$number_of_reads)
-    }
-
-  res_GR <- makeGRangesFromDataFrame(res, start.field = "start_region", 
-                                     end.field = "end_region", keep.extra.columns = TRUE)
-
-  overlap_boul <- overlapsAny(res_GR, my_PMD_GM23248_GR, type = "within")
-
-  res$in_PMD <- overlap_boul
-
-  saveRDS(res, file = paste0("data/10kb_reads_PMD_", chromosome_name, ".rds"))
-
-}
-
-windows_reads_PMD_chr1 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr1.rds")
-windows_reads_PMD_chr2 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr2.rds")
-windows_reads_PMD_chr3 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr3.rds")
-windows_reads_PMD_chr4 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr4.rds")
-windows_reads_PMD_chr5 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr5.rds")
-windows_reads_PMD_chr6 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr6.rds")
-windows_reads_PMD_chr7 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr7.rds")
-windows_reads_PMD_chr8 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr8.rds")
-windows_reads_PMD_chr9 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr9.rds")
-windows_reads_PMD_chr10 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr10.rds")
-windows_reads_PMD_chr11 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr11.rds")
-windows_reads_PMD_chr12 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr12.rds")
-windows_reads_PMD_chr13 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr13.rds")
-windows_reads_PMD_chr14 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr14.rds")
-windows_reads_PMD_chr15 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr15.rds")
-windows_reads_PMD_chr16 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr16.rds")
-windows_reads_PMD_chr17 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr17.rds")
-windows_reads_PMD_chr18 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr18.rds")
-windows_reads_PMD_chr19 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr19.rds")
-windows_reads_PMD_chr20 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr20.rds")
-windows_reads_PMD_chr21 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr21.rds")
-windows_reads_PMD_chr22 <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chr22.rds")
-windows_reads_PMD_chrX <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chrX.rds")
-windows_reads_PMD_chrY <- readRDS("/data/cbio/2021-mth-julie-devis/analyses-rmd/data/10kb_reads_PMD_chrY.rds")
-
-windows_reads_PMD <- rbind(windows_reads_PMD_chr1, windows_reads_PMD_chr2, windows_reads_PMD_chr3, windows_reads_PMD_chr4, windows_reads_PMD_chr5, windows_reads_PMD_chr6, windows_reads_PMD_chr7, windows_reads_PMD_chr8, windows_reads_PMD_chr9, windows_reads_PMD_chr10, windows_reads_PMD_chr11, windows_reads_PMD_chr12, windows_reads_PMD_chr13, windows_reads_PMD_chr14, windows_reads_PMD_chr15, windows_reads_PMD_chr16, windows_reads_PMD_chr17, windows_reads_PMD_chr18, windows_reads_PMD_chr19, windows_reads_PMD_chr20, windows_reads_PMD_chr21, windows_reads_PMD_chr22, windows_reads_PMD_chrX, windows_reads_PMD_chrY)
-
-rownames(windows_reads_PMD) <- NULL
-
-saveRDS(windows_reads_PMD, file = "data/10kb_reads_PMD_all.rds")
-```
-
-Here we load the coverage for all the 10k windows and determine which one are 
-non-PMD due to coverage issues. These region are to be removed from the analysis.
-
-```{r NA_windows}
-windows_reads_PMD <- readRDS("data/10kb_reads_PMD_all.rds")
-
-NA_nonPMD_GR <- windows_reads_PMD %>% 
-  filter(!in_PMD) %>% 
-  filter(reads_tot <= 1000) %>% 
-  makeGRangesFromDataFrame(start.field = "start_region", end.field = "end_region",
-                           keep.extra.columns = TRUE) %>% 
-  GenomicRanges::reduce()
-```
-
-## Are CT genes associated with PMDs ?
-
-### CT regulated by methylation and PMD overlap
-
-```{r CT_genes_prep}
-CT_genes <- CT_genes %>% 
-  mutate(start_prom = case_when(strand == 1 ~ transcription_start_site - 1000,
-                                strand == -1 ~ transcription_start_site - 500)) %>% 
-  mutate(end_prom = case_when(strand == 1 ~ transcription_start_site + 500,
-                              strand == -1 ~ transcription_start_site + 1000))
-
-CT_genes$strand <- ifelse(CT_genes$strand == "1", "+", "-")
-seqlevelsStyle(CT_genes$chr) <- "UCSC"
-
-CT_meth_regulated <- CT_genes %>% 
-  filter(regulated_by_methylation)
-```
-
-```{r CT_prom}
-meth_reg_prom_GR <- makeGRangesFromDataFrame(CT_meth_regulated, start.field = "start_prom",
-                                    end.field = "end_prom",
-                                    keep.extra.columns = TRUE)
-```
-
-```{r overlap_PMD_CT}
-CTmeth_PMD_GR <- subsetByOverlaps(meth_reg_prom_GR, my_PMD_GM23248_GR)
-CTmeth_not_PMD_GR <- subsetByOverlaps(subsetByOverlaps(meth_reg_prom_GR, 
-                                                       my_PMD_GM23248_GR, 
-                                                   invert = TRUE),
-                                      NA_nonPMD_GR, invert = TRUE)
-CTmeth_NA_PMD_GR <-subsetByOverlaps(subsetByOverlaps(meth_reg_prom_GR, 
-                                                     my_PMD_GM23248_GR, 
-                                                 invert = TRUE), 
-                                    NA_nonPMD_GR)
-```
-
-Now that these overlapping have been done, I can report them into my data as a 
-boulean column to use analyse it.
-
-```{r PMD_in_CT}
-CT_meth_regulated <- CT_meth_regulated %>% 
-  mutate(PMD = case_when(CT_meth_regulated$external_gene_name %in% CTmeth_PMD_GR$external_gene_name ~ TRUE,
-                         CT_meth_regulated$external_gene_name %in% CTmeth_not_PMD_GR$external_gene_name ~ FALSE,
-                         CT_meth_regulated$external_gene_name %in% CTmeth_NA_PMD_GR$external_gene_name ~ NA))
-table(is.na(CT_meth_regulated$PMD))
-table(CT_meth_regulated$PMD)
-
-PMD_perc <- table(CT_meth_regulated$PMD)[[2]] / 
-  sum(table(CT_meth_regulated$PMD))*100
-PMD_perc
-
-X_in_PMD_table <- CT_meth_regulated %>% 
-  filter(X_linked) %>% 
-  dplyr::select(PMD) %>% 
-  table()
-X_in_PMD_table
-
-PMD_chr_repartition <- CT_meth_regulated %>% 
-  filter(!is.na(PMD)) %>% 
-  dplyr::select(X_linked) %>% 
-  table()
-PMD_chr_repartition
-```
-
-### CT not regulated by methylation and PMD overlap
-
-```{r CT_not_meth_PMD}
-CT_not_meth_regulated <- CT_genes %>% 
-  filter(!regulated_by_methylation)
-
-prom_GR_not_reg <- makeGRangesFromDataFrame(CT_not_meth_regulated, 
-                                            start.field = "start_prom",
-                                    end.field = "end_prom",
-                                    keep.extra.columns = TRUE)
-not_reg_PMD_GR <- subsetByOverlaps(prom_GR_not_reg, my_PMD_GM23248_GR)
-not_reg_not_PMD_GR <- subsetByOverlaps(subsetByOverlaps(prom_GR_not_reg,
-                                                        my_PMD_GM23248_GR, 
-                                                   invert = TRUE), 
-                                  NA_nonPMD_GR, invert = TRUE)
-not_reg_NA_PMD_GR <-subsetByOverlaps(subsetByOverlaps(prom_GR_not_reg,
-                                                      my_PMD_GM23248_GR, 
-                                                 invert = TRUE),
-                                NA_nonPMD_GR)
-
-CT_not_meth_regulated <- CT_not_meth_regulated %>% 
-  mutate(PMD = case_when(CT_not_meth_regulated$external_gene_name %in% 
-                           not_reg_PMD_GR$external_gene_name ~ TRUE,
-                         CT_not_meth_regulated$external_gene_name %in% 
-                           not_reg_not_PMD_GR$external_gene_name ~ FALSE,
-                         CT_not_meth_regulated$external_gene_name %in% 
-                           not_reg_NA_PMD_GR$external_gene_name ~ NA))
-
-table(is.na(CT_not_meth_regulated$PMD))
-table(CT_not_meth_regulated$PMD)
-
-CT_not_meth_reg_PMD_perc <- table(CT_not_meth_regulated$PMD)[[2]] / 
-  sum(table(CT_not_meth_regulated$PMD))*100
-CT_not_meth_reg_PMD_perc
-```
-
-### Enrichment analysis
-
-The numbers are useful but we don't know if it is significant or not. 
-In order to do so, we're going to compare this percentage with an empiric 
-distribution. For that we need groups of genes comparable.
-
-This distribution is going to be realised by 1000 times computing the % of genes
-in PMD in `r PMD_chr_repartition[[1]]` tissue-specific genes on autosomes and 
-`r PMD_chr_repartition[[2]]` tissue-specific genes on the chromosome X. We use 
-tissue-specific genes because this characteristic can make them more prone to be
-in a repressive region like PMDs.
-
-We then have to create a pool of non-CT tissue-specific genes with their 
-localisation, for that we use the list of tissue specific genes that can be 
-found on the Human Protein Atlas (3106 tissue enriched genes: At least four-fold
-higher mRNA level in a particular tissue compared to any other tissue).
-
-It is necessary to use biomaRt because we need transcription start sites to be 
-able to realise the same analysis on enriched genes than on CT genes. Also, we 
-need to remove CT genes and genes that are in the non-covered regions.
-
-```{r enriched_genes}
-enriched_genes <- read.delim("data/HPA_tissue_enriched_genes.tsv")
-
-attributes_vector <- c("ensembl_gene_id","external_gene_name", "chromosome_name",
-                       "strand", "transcription_start_site", 
-                       "external_transcript_name", "transcript_is_canonical")
-biomaRt_query <- getBM(attributes = attributes_vector, mart = ensembl)
-enriched_genes_biomart <- biomaRt_query %>% 
-  filter(ensembl_gene_id %in% enriched_genes$Ensembl)
-enriched_genes_biomart <- enriched_genes_biomart %>% 
-  mutate(start_prom = case_when(strand == 1 ~ transcription_start_site - 1000,
-                                strand == -1 ~ transcription_start_site - 500)) %>% 
-  mutate(end_prom = case_when(strand == 1 ~ transcription_start_site + 500,
-                              strand == -1 ~ transcription_start_site + 1000)) %>% 
-  filter(!external_gene_name %in% CT_genes$external_gene_name) %>% 
-  filter(transcript_is_canonical == 1) %>%
-  filter(chromosome_name%in%c(1:22, "X", "Y"))
-
-enriched_genes_biomart$strand <- ifelse(enriched_genes_biomart$strand == "1", "+", "-")
-seqlevelsStyle(enriched_genes_biomart$chromosome_name) <- "UCSC"
-
-enriched_genes_GR <- makeGRangesFromDataFrame(enriched_genes_biomart, 
-                                              start.field = "start_prom",
-                                              end.field = "end_prom",
-                                              keep.extra.columns = TRUE)
-enriched_NA_GR <-subsetByOverlaps(subsetByOverlaps(enriched_genes_GR, 
-                                                   my_PMD_GM23248_GR, 
-                                                   invert = TRUE),
-                                  NA_nonPMD_GR)
-
-enriched_genes_analysable <- as.data.frame(
-  enriched_genes_GR[!enriched_genes_GR$ensembl_gene_id %in% enriched_NA_GR$ensembl_gene_id, ])
-```
-
-```{r enrichment_analysis}
-random_percentage_chrprop <- function(i, data, nchrX, nautosomes){
-  random_sample1 <- sample_n(filter(data, seqnames == "chrX"), nchrX)
-  random_sample2 <- sample_n(filter(data, seqnames != "chrX"), nautosomes)
-  random_sample <- rbind(random_sample1,random_sample2)
-  random_sample_GR <- makeGRangesFromDataFrame(random_sample, 
-                                               keep.extra.columns = TRUE)
-  length(subsetByOverlaps(random_sample_GR, my_PMD_GM23248_GR)) / 
-    length(random_sample_GR)*100
-}
-
-distrib_enriched <- unlist(bplapply(1:1000, random_percentage_chrprop, 
-                                    data = enriched_genes_analysable, 
-                                    nchrX = PMD_chr_repartition[[2]], 
-                                    nautosomes = PMD_chr_repartition[[1]]))
-hist(distrib_enriched,
-     col="violetred1",
-     border = "violetred3",
-     breaks = 20,
-     main = paste0("Percentage of promoters in PMDs when randomly taking ",
-                   sum(table(CT_meth_regulated$PMD)), 
-                   " tissue-specific 
-genes 1000 times (taking chromosome proportion into account)"),
-     cex.main = 0.75,
-     xlab = "Percentages",
-     ylab = "Frequency",
-     xlim = c(20,60))
-abline(v = PMD_perc, col = "slateblue1", lwd = 3)
-text(PMD_perc, 120, "CT genes regulated 
-by methylation", col = "slateblue4", cex = 0.8)
-abline(v = CT_not_meth_reg_PMD_perc, col = "darkseagreen1", lwd = 3)
-text(CT_not_meth_reg_PMD_perc, 120, "CT genes not regulated 
-by methylation", col = "darkseagreen", cex = 0.8)
-
-```
-
-```{r add_PMD}
-PMD_CT_results <- rbind(dplyr::select(CT_not_meth_regulated, 
-                                      c(PMD, external_gene_name)),
-                    dplyr::select(CT_meth_regulated, 
-                                  c(PMD, external_gene_name)))
-CT_genes <- left_join(CT_genes, PMD_CT_results)
-```
-
-## Methylation in an aging model : CD4+ cells
-
-Now that we know that CT genes are indeed enriched in PMDs, we have to go 
-further. Indeed, we know that proliferation has an impact on PMD and their 
-methylation : there are losses of methylation with time. As CT genes are located
-in PMDs, we can hypothesise that their methylation might be impacted by the 
-losses happening during proliferation. We thus want to focus on the impact on CT
-genes of the methylation changes happening on PMDss in proliferative or aging 
-cells.
-
-### Data processing
-
-Data coming from Heyn 2012 paper.
-
-```{bash CD4, eval = FALSE}
-#!/bin/bash
-
-#Newborn
-/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR330578
-/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR330579
-/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394135
-/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394136
-/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394137
-/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394138
-/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394139
-/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394140
-/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394141
-/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR394142
-
-#Centenarian
-/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR330574
-/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR330576
-/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR330577
-/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR330575
-
-#Middle-age
-/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR389249
-/opt/sratoolkit/bin/fastq-dump --split-files --outdir ../rawdata --gzip SRR389248
-
-for f in `cat CD4_SRA.txt`;
-
-do mkdir /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/$f; 
-cd /data/cbio/2021-mth-julie-devis/WGBS-processing/processed_data/$f;
-
-#Trimming
-/opt/TrimGalore/trim_galore --fastqc --path_to_cutadapt /home/users/jdevis/cutadapt --paired /data/cbio/2021-mth-julie-devis/WGBS-processing/rawdata/$f\_1.fastq /data/cbio/2021-mth-julie-devis/WGBS-processing/rawdata/$f\_2.fastq;
-
-#Alignement
-/home/users/jdevis/walt/bin/walt \
--i /data/cbio/2021-mth-julie-devis/WGBS-processing/genomes/hg38_by_Chr.dbindex \
--1 $f\_1_val_1.fq \
--2 $f\_2_val_2.fq \
--o $f\_mapping.mr;
-
-#Duplicate-remover
-LC_ALL=C sort -k 1,1 -k 2,2n -k 3,3n -k 6,6 -o $f.mr.sorted_start $f\_mapping.mr;
-/opt/methpipe/build/duplicate-remover -S $f\_dremove_stat.txt -o $f.mr.dremove $f.mr.sorted_start;
-
-#bsrate
-/opt/methpipe/build/bsrate -c /data/cbio/GENOMES/GRCh38_by_Chr/hgdownload.cse.ucsc.edu/chromosomes \
--o $f.bsrate $f.mr.dremove;
-
-#Methylation levels
-/opt/methpipe/build/methcounts -n \
--c /data/cbio/GENOMES/GRCh38_by_Chr/hgdownload.cse.ucsc.edu/chromosomes \
--o $f.meth $f.mr.dremove;
-
-#Merge CpG
-/opt/methpipe/build/symmetric-cpgs -o $f\_CpG_merge.meth $f.meth;
-
-#Stats
-/opt/methpipe/build/levels -o $f.levels $f.meth
-
-done
-
-```
-
-Bisulfite-Seq data coming from a male newborn, a middle age donor and a male 
-centenarian.
-
-Quality of the data is good, with a mean of 13 reads per cytosine. However, 
-middle age data has no info about the sex of the donor. By comparing the number 
-of reads by chromosome in this sample, a female one and the male ones, we saw 
-that the proportion of reads on chromosomes X and Y showed that the middle age 
-data come from a woman. We then have to be very careful when analysing the X 
-chromosome with the middle age data. To facilitate the analysis and compare all
-samples, I’ll remove the adult data on chrX.
-
-```{r load_CD4, message = FALSE, warning = FALSE}
-columns <- c("chromosome", "position", "strand", "sequence context","methylation_level","number_of_reads")
-CD4_newborn <- read_delim("data/CD4_newborn.meth", 
-    delim = "\t", escape_double = FALSE, col_names = FALSE, trim_ws = TRUE)
-names(CD4_newborn) <- columns
-CD4_middle <- read_delim("data/CD4_middle.meth", 
-    delim = "\t", escape_double = FALSE, col_names = FALSE, trim_ws = TRUE)
-names(CD4_middle) <- columns
-CD4_middle <- filter(CD4_middle, chromosome != "chrX")
-CD4_cent <- read_delim("data/CD4_cent.meth", 
-    delim = "\t", escape_double = FALSE, col_names = FALSE, trim_ws = TRUE)
-names(CD4_cent) <- columns
-```
-```{r change_CD4, message = FALSE, warning = FALSE}
-CD4_newborn$age <- "newborn"
-CD4_middle$age <- "middle_age"
-CD4_cent$age <- "centenarian"
-```
-```{r CD4_GR, message = FALSE, warning = FALSE}
-CD4_newborn_GR <- makeGRangesFromDataFrame(CD4_newborn, start.field = "position", 
-                                           end.field = "position", keep.extra.columns = TRUE)
-CD4_middle_GR <- makeGRangesFromDataFrame(CD4_middle, start.field = "position", 
-                                          end.field = "position", keep.extra.columns = TRUE)
-CD4_cent_GR <- makeGRangesFromDataFrame(CD4_cent, start.field = "position", 
-                                        end.field = "position", keep.extra.columns = TRUE)
-```
-```{r rm, echo = FALSE}
-rm(CD4_newborn)
-rm(CD4_middle)
-rm(CD4_cent)
-```
-
-### CT genes methylation level
-
-Here I extract methylation for all CpGs located in CT genes promoter 
-(regulated by methylation because it's these that we can find in PMDs).
-
-```{r CT_CD4, warning = FALSE}
-CT_prom_GR <- makeGRangesFromDataFrame(CT_genes, start.field = "start_prom",
-                                    end.field = "end_prom",
-                                    keep.extra.columns = TRUE)
-
-newborn <- mergeByOverlaps(CD4_newborn_GR, CT_prom_GR, ignore.strand = TRUE)[,1]
-newborn$external_gene_name <- mergeByOverlaps(CD4_newborn_GR, CT_prom_GR, 
-                                              ignore.strand = TRUE)$external_gene_name
-newborn <- as.data.frame(newborn)
-middle <- mergeByOverlaps(CD4_middle_GR, CT_prom_GR, ignore.strand = TRUE)[,1]
-middle$external_gene_name <- mergeByOverlaps(CD4_middle_GR, CT_prom_GR, 
-                                             ignore.strand = TRUE)$external_gene_name
-middle <- as.data.frame(middle)
-cent <- mergeByOverlaps(CD4_cent_GR, CT_prom_GR, ignore.strand = TRUE)[,1]
-cent$external_gene_name <- mergeByOverlaps(CD4_cent_GR, CT_prom_GR, 
-                                           ignore.strand = TRUE)$external_gene_name
-cent <- as.data.frame(cent)
-
-CTmeth_inCD4 <- rbind(newborn, middle, cent)
-rm(newborn, middle, cent)
-CTmeth_inCD4$age <- factor(CTmeth_inCD4$age, 
-                           levels = c("newborn","middle_age","centenarian"))
-CTmeth_inCD4[CTmeth_inCD4$number_of_reads == 0,]$methylation_level <- NA
-
-length(unique(CTmeth_inCD4$external_gene_name))
-
-```
-
-We can see here that we have information for all 
-`r length(unique(CTmeth_inCD4$external_gene_name))` CT genes. We now have to 
-calculate the mean methylation level of each promoter to visualise it and 
-compare to PMDs.
-
-Some df that we obtain are empty, their CpG are not covered by the CD4 data, 
-and so need to be removed. I also want to remove those who have data in only 
-one age category, as we want to compare different age, they are of no need. 
-When one of the age is missing, I'm adding it as NA to facilitate the rest 
-(mostly gene on chrX and we removed data from adult)
-
-```{r CT_level}
-reorganize_data_genes <- function(gene_name, data){
-  reorganized_data <- data %>%
-    filter(external_gene_name == gene_name) %>% 
-    dplyr::select("start","methylation_level","age") %>% 
-    na.omit() %>%  # to remove CpG rows with NA meth_level
-    pivot_wider(names_from = start, values_from = methylation_level) %>% 
-    column_to_rownames("age")
-}
-
-CT_per_gene_CD4_datasets <- sapply(unique(CTmeth_inCD4$external_gene_name), 
-                                   reorganize_data_genes, data = CTmeth_inCD4)
-
-p <- rep(NA, length(CT_per_gene_CD4_datasets))
-for(i in 1:length(CT_per_gene_CD4_datasets)){
- p[i] <- nrow(CT_per_gene_CD4_datasets[[i]]) > 1 & 
-   is.data.frame(CT_per_gene_CD4_datasets[[i]])
-}
-CT_per_gene_CD4_datasets <- CT_per_gene_CD4_datasets[p]
-
-k <- rep(NA, length(CT_per_gene_CD4_datasets))
-for(i in 1:length(CT_per_gene_CD4_datasets)){
-  k[i] <- nrow(CT_per_gene_CD4_datasets[[i]]) == 2
-}
-for (gene in names(CT_per_gene_CD4_datasets[k])){
-  tmp <- rep(NA, ncol(CT_per_gene_CD4_datasets[[gene]]))
-  CT_per_gene_CD4_datasets[[gene]] <- rbind(CT_per_gene_CD4_datasets[[gene]], tmp)
-  rownames(CT_per_gene_CD4_datasets[[gene]])[3] <- 
-    c("newborn","middle_age", "centenarian")[(!c("newborn","middle_age","centenarian") %in% 
-                                                rownames(CT_per_gene_CD4_datasets[[gene]]))]
-  # naming the new row
-}
-
-for (gene in names(CT_per_gene_CD4_datasets)){
-  CT_per_gene_CD4_datasets[[gene]] <- arrange(CT_per_gene_CD4_datasets[[gene]], 
-                                              rownames(CT_per_gene_CD4_datasets[[gene]]))
-  # Put rows in the right order
-}
-
-rm(p, i, k, tmp)
-
-CT_meth_means_pre <- sapply(CT_per_gene_CD4_datasets, rowMeans, na.rm = TRUE)
-
-CT_meth_means <- CT_meth_means_pre %>%
-  t() %>% 
-  as.data.frame() %>%
-  rownames_to_column("external_gene_name")
-
-CT_meth_means$middle_age[is.na(CT_meth_means$middle_age)] <- NA
-
-n_means_meth <- CT_meth_means %>% 
-  filter(external_gene_name %in% CT_meth_regulated$external_gene_name) %>% 
-  nrow()
-
-CT_genes <- left_join(CT_genes, CT_meth_means)
-```
-
-On the `r dim(CT_genes)[1]` CT genes, I have enough data to analyse 
-`r ncol(CT_meth_means_pre)` of them (`r n_means_meth`/`r dim(CT_meth_regulated)[1]` 
-meth regulated).
-
-```{r all_CT_plot}
-CT_genes_to_plot <- CT_genes %>% 
-  dplyr::select(external_gene_name, chr, X_linked, regulated_by_methylation, 
-                PMD, newborn, middle_age, centenarian) %>%
-  filter(!is.na(PMD)) %>% 
-  pivot_longer(names_to = "age", values_to = "methylation_level", 
-               - c("external_gene_name","chr","X_linked", 
-                   "regulated_by_methylation", "PMD"))
-CT_genes_to_plot$age <- factor(CT_genes_to_plot$age, 
-                               levels = c("newborn","middle_age","centenarian"))
-CT_genes_to_plot %>%
-  mutate(PMD = case_when(PMD == TRUE ~ "PMD",
-                         PMD == FALSE ~ "not_PMD")) %>%
-  mutate(regulated_by_methylation = case_when(regulated_by_methylation == TRUE ~ "meth_reg",
-                                              regulated_by_methylation == FALSE ~ "not_meth_reg")) %>% 
-  ggplot(aes(x = age, y = methylation_level, fill = age)) +
-  geom_boxplot(outlier.shape = NA)+
-  geom_jitter(aes(color = age), alpha = 0.3) +
-  facet_grid(regulated_by_methylation ~ PMD) + 
-  scale_fill_manual(values = c("rosybrown1", "palevioletred2", "violetred3")) +
-  scale_color_manual(values = c("rosybrown3","palevioletred3","violetred4")) +
-  theme_bw()
-
-```
-
-### Comparison CT and PMDs
-
-```{r PMD_CD4, warning = FALSE}
-PMD_newborn <- mergeByOverlaps(CD4_newborn_GR, my_PMD_GM23248_GR, ignore.strand = TRUE)[,1]
-PMD_newborn$name <- mergeByOverlaps(CD4_newborn_GR, my_PMD_GM23248_GR, ignore.strand = TRUE)$name
-PMD_newborn <- as.data.frame(PMD_newborn)
-PMD_middle <- mergeByOverlaps(CD4_middle_GR, my_PMD_GM23248_GR, ignore.strand = TRUE)[,1]
-PMD_middle$name <- mergeByOverlaps(CD4_middle_GR, my_PMD_GM23248_GR, ignore.strand = TRUE)$name
-PMD_middle <- as.data.frame(PMD_middle)
-PMD_cent<- mergeByOverlaps(CD4_cent_GR, my_PMD_GM23248_GR, ignore.strand = TRUE)[,1]
-PMD_cent$name <- mergeByOverlaps(CD4_cent_GR, my_PMD_GM23248_GR, ignore.strand = TRUE)$name
-PMD_cent <- as.data.frame(PMD_cent)
-
-PMD_CD4 <- rbind(PMD_newborn, PMD_middle, PMD_cent)
-rm(PMD_newborn, PMD_middle, PMD_cent)
-PMD_CD4$age <- factor(PMD_CD4$age, levels = c("newborn","middle_age","centenarian"))
-PMD_CD4[PMD_CD4$number_of_reads == 0,]$methylation_level <- NA
-```
-
-Let's compare the evolution of CT methylation with that of the PMDs they're in.
-
-```{r CT_PMD_plot}
-CT_reg_PMD_plot <- CT_genes_to_plot %>% 
-  filter(regulated_by_methylation & PMD) %>% 
-  ggplot(aes(x = methylation_level, color = age)) +
-  geom_density() + 
-  scale_color_manual(values = c("rosybrown1", "palevioletred2", "violetred3"))+
-  xlab ("Promoter mean methylation level (CT meth reg and in PMDs)") +
-  theme_minimal()
-
-CT_reg_PMD_GR <- CT_prom_GR[!is.na(CT_prom_GR@elementMetadata$PMD) &
-                              CT_prom_GR@elementMetadata$PMD & 
-                              CT_prom_GR@elementMetadata$regulated_by_methylation, ]
-
-CT_PMD_meth_plot <- PMD_CD4 %>% 
-  filter(name %in% subsetByOverlaps(my_PMD_GM23248_GR,CT_reg_PMD_GR)$name) %>% 
-  group_by(name, age) %>% 
-  summarise("mean_methylation" = mean(methylation_level, na.rm = TRUE)) %>% 
-  ggplot(aes(x = mean_methylation, color = age)) + 
-  geom_density() +
-  scale_color_manual(values = c("rosybrown1", "palevioletred2", "violetred3")) +
-  theme_minimal() +
-  xlab("CT's PMDs mean methylation level")
-
-(CT_reg_PMD_plot + scale_x_continuous(limits = c(0,1))) / (CT_PMD_meth_plot + scale_x_continuous(limits = c(0,1)))
-
-```
-
-### Comparison CT and non CT in PMDs
-
-We don't know if what we see happening in CT genes is common with other genes 
-located in PMDs. If we can also see a demethylation of these genes promoters 
-and that it is bigger than CT demethylation, it means that there would be a 
-protection of CT genes against demethylation. We thus want to compare CT 
-methylation to the one of other genes in PMDs.
-
-#### Non-CT methylation
-Get methylation data for all other genes than CT in PMDs.
-
-```{r nonCT_PMD, warning = FALSE}
-non_CT_genes <- biomaRt_query %>% 
-  mutate(start_prom = case_when(strand == 1 ~ transcription_start_site - 1000,
-                                strand == -1 ~ transcription_start_site - 500)) %>% 
-  mutate(end_prom = case_when(strand == 1 ~ transcription_start_site + 500,
-                              strand == -1 ~ transcription_start_site + 1000)) %>% 
-  filter(!external_gene_name %in% CT_genes$external_gene_name) %>% 
-  filter(transcript_is_canonical == 1) %>%
-  filter(chromosome_name%in%c(1:22, "X", "Y"))
-
-non_CT_genes$strand <- ifelse(non_CT_genes$strand == "1", "+", "-")
-seqlevelsStyle(non_CT_genes$chromosome_name) <- "UCSC"
-non_CT_GR <- makeGRangesFromDataFrame(non_CT_genes,
-                                      start.field = "start_prom",
-                                      end.field = "end_prom",
-                                      keep.extra.columns = TRUE)
-non_CT_in_PMD <- subsetByOverlaps(non_CT_GR, my_PMD_GM23248_GR)
-
-newborn <- mergeByOverlaps(CD4_newborn_GR, non_CT_in_PMD, ignore.strand = TRUE)[,1]
-newborn$ensembl_gene_id <- mergeByOverlaps(CD4_newborn_GR, non_CT_in_PMD, 
-                                           ignore.strand = TRUE)$ensembl_gene_id
-newborn <- as.data.frame(newborn)
-middle <- mergeByOverlaps(CD4_middle_GR, non_CT_in_PMD, ignore.strand = TRUE)[,1]
-middle$ensembl_gene_id <- mergeByOverlaps(CD4_middle_GR, non_CT_in_PMD, 
-                                          ignore.strand = TRUE)$ensembl_gene_id
-middle <- as.data.frame(middle)
-cent <- mergeByOverlaps(CD4_cent_GR, non_CT_in_PMD, ignore.strand = TRUE)[,1]
-cent$ensembl_gene_id <- mergeByOverlaps(CD4_cent_GR, non_CT_in_PMD, 
-                                        ignore.strand = TRUE)$ensembl_gene_id
-cent <- as.data.frame(cent)
-
-nonCTmeth_inCD4 <- rbind(newborn, middle, cent)
-rm(newborn, middle, cent)
-nonCTmeth_inCD4$age <- factor(nonCTmeth_inCD4$age, 
-                           levels = c("newborn","middle_age","centenarian"))
-nonCTmeth_inCD4[nonCTmeth_inCD4$number_of_reads == 0,]$methylation_level <- NA
-
-reorganize_data_ensembl <- function(gene_name, data){
-  reorganized_data <- data %>%
-    filter(ensembl_gene_id == gene_name) %>% 
-    dplyr::select("start","methylation_level","age") %>% 
-    na.omit() %>%  # to remove CpG rows with NA meth_level
-    pivot_wider(names_from = start, values_from = methylation_level) %>% 
-    column_to_rownames("age")
-}
-
-nonCT_per_gene_CD4_datasets <- bplapply(unique(nonCTmeth_inCD4$ensembl_gene_id),
-                                      reorganize_data_ensembl, data = nonCTmeth_inCD4)
-names(nonCT_per_gene_CD4_datasets) <- unique(nonCTmeth_inCD4$ensembl_gene_id)
-
-p <- rep(NA, length(nonCT_per_gene_CD4_datasets))
-for(i in 1:length(nonCT_per_gene_CD4_datasets)) {
- p[i] <- nrow(nonCT_per_gene_CD4_datasets[[i]]) > 1 &
-   is.data.frame(nonCT_per_gene_CD4_datasets[[i]])
-}
-nonCT_per_gene_CD4_datasets <- nonCT_per_gene_CD4_datasets[p]
-
-k <- rep(NA, length(nonCT_per_gene_CD4_datasets))
-for(i in 1:length(nonCT_per_gene_CD4_datasets)) {
-  k[i] <- nrow(nonCT_per_gene_CD4_datasets[[i]]) == 2
-}
-
-for (gene in names(nonCT_per_gene_CD4_datasets[k])) {
-  tmp <- rep(NA, ncol(nonCT_per_gene_CD4_datasets[[gene]]))
-  nonCT_per_gene_CD4_datasets[[gene]] <- rbind(nonCT_per_gene_CD4_datasets[[gene]], tmp)
-  rownames(nonCT_per_gene_CD4_datasets[[gene]])[3] <-
-    c("newborn","middle_age", "centenarian")[(!c("newborn","middle_age","centenarian") %in%
-                                                rownames(nonCT_per_gene_CD4_datasets[[gene]]))]
-  # naming the new row
-}
-
-for (gene in names(nonCT_per_gene_CD4_datasets)) {
-  nonCT_per_gene_CD4_datasets[[gene]] <- arrange(nonCT_per_gene_CD4_datasets[[gene]],
-                                              rownames(nonCT_per_gene_CD4_datasets[[gene]]))
-  # Put rows in the right order
-}
-
-rm(p, i, k, tmp)
-
-# load("data_rmd/nonCT_per_gene_CD4_datasets.rda")
-
-nonCT_meth_means <- sapply(nonCT_per_gene_CD4_datasets, rowMeans, na.rm = TRUE)
-nonCT_meth_means <- nonCT_meth_means %>%
-  t() %>%
-  as.data.frame() %>%
-  rownames_to_column("ensembl_gene_id")
-
-nonCT_meth_means$middle_age[is.na(nonCT_meth_means$middle_age)] <- NA
-```
-
-#### CT and non CT comparison
-
-```{r non-CT_plot}
-threshold <- CT_genes %>% 
-  filter(regulated_by_methylation & PMD) %>% 
-  dplyr::select(newborn) %>% 
-  min(na.rm = TRUE)
-
-nonCT_meth_means_thr <- nonCT_meth_means %>% 
-  filter(newborn >= threshold)
-
-nonCT_meth_means_to_plot <-  nonCT_meth_means_thr %>% 
-  pivot_longer(names_to = "age", values_to = "methylation_level", 
-               - "ensembl_gene_id")
-nonCT_meth_means_to_plot$age <- factor(nonCT_meth_means_to_plot$age, 
-                               levels = c("newborn","middle_age","centenarian"))
-
-med_meth_nb_CT <- CT_genes_to_plot %>% 
-  filter(regulated_by_methylation & PMD & age == "newborn") %>% 
-  pull(methylation_level) %>% 
-  median(na.rm = TRUE)
-med_meth_nb_CT
-
-med_meth_nb_nonCT <-nonCT_meth_means_to_plot %>% 
-  filter(age == "newborn") %>% 
-  pull(methylation_level) %>% 
-  median(na.rm = TRUE)
-med_meth_nb_nonCT
-
-med_meth_cent_CT <- CT_genes_to_plot %>% 
-  filter(regulated_by_methylation & PMD & age == "centenarian") %>% 
-  pull(methylation_level) %>% 
-  median(na.rm = TRUE)
-med_meth_cent_CT
-
-med_meth_cent_nonCT <-nonCT_meth_means_to_plot %>% 
-  filter(age == "centenarian") %>% 
-  pull(methylation_level) %>% 
-  median(na.rm = TRUE)
-med_meth_cent_nonCT
-
-CT_reg_PMD_boxplot <- CT_genes_to_plot %>%
-  filter(regulated_by_methylation & PMD) %>% 
-  ggplot(aes(x = age, y = methylation_level, fill = age)) +
-  geom_boxplot(outlier.shape = NA)+
-  geom_jitter(aes(color = age), alpha = 0.3) +
-  scale_fill_manual(values = c("rosybrown1", "palevioletred2", "violetred3")) +
-  scale_color_manual(values = c("rosybrown3","palevioletred3","violetred4")) +
-  theme_bw()
-
-nonCT_boxplot <- nonCT_meth_means_to_plot %>%
-  ggplot(aes(x = age, y = methylation_level, fill = age)) +
-  geom_jitter(aes(color = age), alpha = 0.3) +
-  geom_boxplot(outlier.shape = NA) +
-  scale_fill_manual(values = c("rosybrown1", "palevioletred2", "violetred3")) +
-  scale_color_manual(values = c("rosybrown3","palevioletred3","violetred4")) +
-  theme_bw()
-
-(CT_reg_PMD_boxplot + ggtitle("CT genes in PMDs") + 
-    scale_y_continuous(limits = c(0,1)) + theme(legend.position = "none")) + 
-  (nonCT_boxplot + ggtitle("non CT genes in PMDs") + 
-     scale_y_continuous(limits = c(0,1)))
-```
-
-#### Differential analysis
-
-However, this is only an observation based on the data, we need to do a 
-quantification of this decrease and see its signification. For this, we have to 
-do a differential methylation analysis.
-
-First step is to check that there is indeed a loss between newborn and 
-centenarian for each gene. To do so, we use a unidirectional (newborn > 
-centenarian) paired statistical test on the methylation level of cytosines in 
-both ages. Data can't really be normalise so we use a Wilcoxon test 
-(non-parametric t test). Our significativity threshold is 5%, which means that 
-a test is significative if the adjusted (B-H method) pvalue is inferior to 0.05.
-
-Second step is to quantify the loss of methylation between newborn and 
-centenarian. To do so, we calculate a delta_meth for each gene, which is 
-simply the difference between newborn and cent level.
-
-```{r diff_analysis}
-CT_to_pval <- CTmeth_inCD4 %>%
-  filter(external_gene_name %in% filter(CT_genes, 
-                                        PMD & regulated_by_methylation)$external_gene_name) %>%
-  dplyr::select(external_gene_name, seqnames, start, age, methylation_level) %>% 
-  pivot_wider(names_from = "age", values_from = "methylation_level", values_fill = NA) %>% 
-  filter(!is.na(newborn) & !is.na(centenarian)) %>% 
-  left_join(dplyr::select(CT_genes, external_gene_name, ensembl_gene_id))
-
-non_CT_to_pval <- nonCTmeth_inCD4 %>% 
-  filter(ensembl_gene_id %in% nonCT_meth_means_thr$ensembl_gene_id) %>%
-  dplyr::select(ensembl_gene_id, seqnames, start, age, methylation_level) %>%
-  pivot_wider(names_from = "age", values_from = "methylation_level", values_fill = NA) %>% 
-  filter(!is.na(newborn) & !is.na(centenarian))
-
-test_nb_cent <- function(gene_i, data){
-    wilcox.test(
-      as.matrix(filter(data, ensembl_gene_id == gene_i)$newborn), 
-      as.matrix(filter(data, ensembl_gene_id == gene_i)$centenarian), 
-      paired = TRUE, 
-      alternative = "greater", 
-      exact = FALSE)$p.value
-}
-
-pval_CT <- sapply(unique(CT_to_pval$ensembl_gene_id), test_nb_cent, CT_to_pval)
-pval_nonCT <- sapply(unique(non_CT_to_pval$ensembl_gene_id), test_nb_cent, non_CT_to_pval)
-
-CT_stat_analysis <- CT_genes %>% 
-  filter(ensembl_gene_id %in% CT_to_pval$ensembl_gene_id) %>% 
-  dplyr::select(ensembl_gene_id, external_gene_name, chr, X_linked, 
-                 newborn, middle_age, centenarian) %>% 
-  filter(!is.na(newborn) & !is.na(centenarian)) %>% 
-  mutate("padj" = p.adjust(pval_CT, method = "BH"))
-sign_table_CT <- table(CT_stat_analysis$padj < 0.05)
-
-nonCT_stat_analysis <- nonCT_meth_means_thr %>% 
-  filter(ensembl_gene_id %in% non_CT_to_pval$ensembl_gene_id) %>% 
-  mutate("padj" = p.adjust(pval_nonCT, method = "BH"))
-sign_table_nonCT <- table(nonCT_stat_analysis$padj < 0.05)
-
-table(CT_stat_analysis$padj <= 0.05)[[2]]/length(CT_stat_analysis$padj)
-table(nonCT_stat_analysis$padj <= 0.05)[[2]]/length(nonCT_stat_analysis$padj)
-
-CT_stat_analysis <- CT_stat_analysis %>% 
-  mutate("delta_meth" = centenarian - newborn)
-
-nonCT_stat_analysis <- nonCT_stat_analysis %>% 
-  mutate("delta_meth" = centenarian - newborn)
-
-med_delta_CT <- CT_stat_analysis %>% 
-  filter(padj <= 0.05) %>% 
-  pull(delta_meth) %>% 
-  median()
-med_delta_CT
-
-med_delta_nonCT <- nonCT_stat_analysis %>% 
-  filter(padj <= 0.05) %>% 
-  pull(delta_meth) %>% 
-  median()
-med_delta_nonCT
-
-CT_stat_analysis %>% 
-  filter(padj <= 0.05) %>%
-  mutate(type = "CT_genes") %>%
-  dplyr::select(ensembl_gene_id, centenarian, middle_age, newborn, padj, delta_meth, type) %>% 
-  rbind(nonCT_stat_analysis %>%
-          filter(padj <= 0.05) %>%
-          mutate(type = "nonCT_genes")) %>%
-
-  ggplot(aes(y = delta_meth, x = type, fill = type)) +
-  geom_boxplot(outlier.shape = NA) +
-  geom_jitter(aes(color = type), alpha = 0.2, size = 0.5) +
-  scale_fill_manual(values = c("plum","lightgoldenrod1")) +
-  scale_color_manual(values = c("darkorchid3","gold1")) +
-  ggtitle("Delta_meth of PMD genes with significant demethylation") +
-  xlab("") + 
-  ylab("Delta_meth centenarian - newborn") +
-  theme_bw() +
-  theme(legend.position = "none")
-
-```
-
-There is indeed a difference in the amplitude of delta_meth between CT 
-and nonCT genes in PMD.
-
-We then want to test the difference between delta_meth for CT and nonCT genes
-in PMD, when pval is significative. For that we use a unidirectionnal (CT > 
-non-CT) Mann-Whitney test, non-parametric (because no normality in delta_meth 
-non-CT) version of student t test.
-
-```{r delta_test}
-wilcox.test(filter(CT_stat_analysis, padj <= 0.05)$delta_meth,
-            filter(nonCT_stat_analysis, padj <= 0.05)$delta_meth,
-            alternative = "greater")
-```
-
-pvalue here is significative. This shows then that delta_meth are smaller in 
-nonCT genes and demethylation in nonCT genes is then higher than in CT genes.
-
-However, the difference between delta_meth distributions could be the result of 
-the biggest number of genes in nonCT (`r sign_table_nonCT[[1]]` VS 
-`r sign_table_CT[[1]]` not sign, or `r sign_table_nonCT[[2]]` VS 
-`r sign_table_CT[[2]]` padj sign). In order to see that, let's do a down 
-sampling : taking `r sign_table_CT[[2]]` nonCT genes with a pval <= 0.05 and 
-`r sign_table_CT[[1]]` with a pval > 0.05 a thousand times and calculate the 
-mean of each distributions. I can visualise that after and see where CT genes 
-goes compare to that.
-
-```{r down_sampling}
-random_delta_distrib_means <- function(i, data, n_sign, n_notsign){
-  random_sample1 <- sample_n(filter(data, padj <= 0.05), n_sign)
-  random_sample2 <- sample_n(filter(data, padj > 0.05), n_notsign)
-  random_sample <- rbind(random_sample1, random_sample2)
-
-  mean(random_sample$delta_meth)
-}
-
-distrib_means <- unlist(bplapply(1:1000, random_delta_distrib_means,
-                                data = nonCT_stat_analysis, 
-                                n_sign = sign_table_CT[[2]], 
-                                n_notsign = sign_table_CT[[1]]))
-
-mean(CT_stat_analysis$delta_meth)
-
-table(distrib_means >= mean(CT_stat_analysis$delta_meth))
-
-hist(distrib_means,
-     col="dodgerblue1",
-     border = "dodgerblue3",
-     breaks = 20,
-     main = paste0("Means of 1000 delta_meth distributions of nonCT genes when randomly 
-     taking ", sign_table_CT[[2]], "significative and ", sign_table_CT[[1]], " not significative genes"),
-     cex.main = 0.75,
-     xlab = "Means",
-     ylab = "Frequency",
-     xlim = c(-0.2, 0))
-abline(v = mean(CT_stat_analysis$delta_meth), col = "violetred1", lwd = 3)
-text(mean(CT_stat_analysis$delta_meth), 125, "Mean of CT genes 
-delta_meth distribution", col = "violetred3")
-
-```
-
-This shows that even with the same amount of genes, delta_meth are 
-significantly smaller for CT genes than for nonCT genes, the effect is then not
-due to the amount of genes.
-
-This analysis shows that CT genes located in PMDs undergo a demethylation 
-related to aging that is smaller than the other genes in PMDs. Thus, age and 
-demethylation that is associated to it couldn't explain demethylation and 
-activation of CT genes. There is then probably something that protects CT genes 
-from PMD demethylation.
-
-## Why is there a protection on CT genes : CpG density hypothesis
-
-CT genes are thus protected against demethylation in PMDs. Next question is to 
-determine what confers this protection to CT genes. what One of our hypothesis 
-for that protection is that CT genes are protected thanks to their density in CpG.
-Indeed, CT genes promoters are rich in CpGs while PMDs are poor. 
-It is possible that this density protects CT genes from demethylation.
-
-Promoter's density in CpG is already computed in CTexploreR but we have to do 
-it also for non-CT genes.
-
-```{r compute_cpg}
-CT_density_analysis <- CT_genes %>% 
-  filter(ensembl_gene_id %in% CT_stat_analysis$ensembl_gene_id) %>%
-  mutate("padj" = CT_stat_analysis$padj) %>% 
-  mutate("delta_meth" = CT_stat_analysis$delta_meth)
-
-nonCT_density_analysis <- as.data.frame(non_CT_in_PMD) %>% 
-  filter(ensembl_gene_id %in% nonCT_stat_analysis$ensembl_gene_id)%>%
-  left_join(nonCT_stat_analysis)
-colnames(nonCT_density_analysis)[1:3] <- c("chr", "start_prom", "end_prom")
-nonCT_density_analysis$chr <- as.character(nonCT_density_analysis$chr)
-
-cpg_density <- function(gene, data){
-  temp <- data %>% 
-    filter(ensembl_gene_id == gene) %>% 
-    dplyr::select(ensembl_gene_id, start_prom, end_prom, chr)
-
-  seqlevelsStyle(temp$chr) <- "NCBI"
-
-  chrom <- genome[startsWith(names(genome), paste0(temp$chr, " "))][[1]]
-
-  interest <- subseq(chrom, temp$start_prom, temp$end_prom)
-
-  p <- matchPattern("CG", interest)
-  length(p) / length(interest)
-}
-
-genome <- readDNAStringSet("../../cluster/GENOMES/Homo_sapiens.GRCh38.dna.primary_assembly.fa")
-
-CT_cpg_density <- sapply(CT_density_analysis$ensembl_gene_id, cpg_density, 
-                         CT_density_analysis)
-
-nonCT_cpg_density <- unlist(bplapply(nonCT_density_analysis$ensembl_gene_id, cpg_density,
-                            nonCT_density_analysis))
-```
-
-There is already the CpG_density column from CT_genes so careful, what I've done
-here is similar to what was already in it, it confirms.
-
-### Comparison CT and non CT with similar densities
-
-```{r cpg_plot}
-CT_density_analysis <- CT_density_analysis %>% 
-  mutate("CpG_density_analysis" = CT_cpg_density*100)
-
-nonCT_density_analysis <- nonCT_density_analysis %>% 
-  mutate("CpG_density_analysis" = nonCT_cpg_density*100)
-
-CT_density_analysis %>% 
-  filter(padj <= 0.05) %>%
-  mutate(type = "CT_genes") %>%
-  dplyr::select(ensembl_gene_id, centenarian, middle_age, newborn, padj, 
-                delta_meth, CpG_density_analysis, type) %>% 
-  rbind(nonCT_density_analysis %>%
-          filter(padj <= 0.05) %>%
-          mutate(type = "nonCT_genes") %>% 
-          dplyr::select(ensembl_gene_id, centenarian, middle_age, newborn, padj, 
-                delta_meth, CpG_density_analysis, type)) %>%
-
-  ggplot(aes(x = CpG_density_analysis, color = type)) +
-  geom_density() + 
-  scale_color_manual(values = c("darkorchid3","gold1")) +
-  xlab("CpG density in promoters") +
-  ylab("Quantity (density plot)") +
-  ggtitle("Distribution of CpG density on CT and non CT promoters
-significantly demethylated in PMDs")+
-  theme_minimal()
-
-```
-
-This shows/confirms that nonCT promoters in PMDs are poorer in CpG.
-
-```{r density_deltameth}
-CT_density_analysis %>% 
-  filter(padj <= 0.05) %>%
-  mutate(type = "CT_genes") %>%
-
-  ggplot(aes(x = delta_meth, y = CpG_density_analysis, color = type)) +
-  geom_point() +
-  geom_smooth(method = "loess", span = 2.5, color = "darkorchid4") +
-  scale_color_manual(values = c("darkorchid3")) +
-  xlab("Delta_meth of significatively 
-demethylated genes") +
-  ylab("CpG density of promoter") +
-  facet_grid(~ type) +
-  theme_bw() +
-  theme(legend.position = "none") +
-  scale_y_continuous(limits = c(0, 8)) +
-  scale_x_continuous(limits = c(-0.63, 0.05)) +
-
-  nonCT_density_analysis %>%
-          filter(padj <= 0.05) %>%
-          mutate(type = "nonCT_genes") %>%
-
-  ggplot(aes(x = delta_meth, y = CpG_density_analysis, color = type)) +
-  geom_point() +
-  geom_smooth(method = "loess", span = 0.75, color = "goldenrod2") +
-  scale_color_manual(values = c("gold1")) +
-  facet_grid(~ type) +
-  theme_bw() + 
-  ylab("") +
-  xlab("Delta_meth of significatively 
-demethylated genes") +
-  scale_y_continuous(limits = c(0, 8)) +
-  scale_x_continuous(limits = c(-0.63, 0.05)) +  
-  theme(legend.position = "none")
-```
-
-We see here a diagonal tendancy with a positive correlation, it shows that there
-is a link between CpG density and delta_meth. This tendancy is visible in CT and
-nonCT : the more CpG there are in promoters, less demethylation we observe. 
-Isolated CpGs are then probably those that are easier demethylated in PMDs. We 
-thus propose that the strong density in CpG that characterize CT promoters is the 
-origin of their resistance to demethylation
-
-Let's compare only similar CpG densities, and do the same categories as in 
-CTexploreR.
-
-```{r cpg_category}
-CT_density_analysis <- CT_density_analysis %>% 
-  mutate(density_state = 
-             case_when (CT_density_analysis$CpG_density_analysis >= 4 ~ "rich",
-                        CT_density_analysis$CpG_density_analysis <= 2 ~ "poor",
-                        CT_density_analysis$CpG_density_analysis > 2 & 
-                          CT_density_analysis$CpG_density_analysis < 4 ~ "normal"))
-
-nonCT_density_analysis <- nonCT_density_analysis %>% 
-  mutate(density_state = 
-             case_when (nonCT_density_analysis$CpG_density_analysis >= 4 ~ "rich",
-                        nonCT_density_analysis$CpG_density_analysis <= 2 ~ "poor",
-                        nonCT_density_analysis$CpG_density_analysis > 2 &
-                          nonCT_density_analysis$CpG_density_analysis < 4 ~ "normal"))
-```
-
-By comparing only delta_meth from genes normal/rich in CpG, we can erase the 
-effect of CpG density from the demethylation analysis.
-
-```{r delta_meth_rich_plot}
-CT_density_analysis %>% 
-  filter(padj <= 0.05) %>%
-  filter(density_state != "poor") %>% 
-  mutate(type = "CT_genes") %>%
-  dplyr::select(ensembl_gene_id, delta_meth, type) %>% 
-  rbind(nonCT_density_analysis %>%
-          filter(padj <= 0.05) %>%
-          filter(density_state != "poor") %>%
-          mutate(type = "nonCT_genes")%>%
-          dplyr::select(ensembl_gene_id, delta_meth, type))%>%
-
-  ggplot(aes(y = delta_meth, x = type, fill = type)) +
-  geom_boxplot(outlier.shape = NA) +
-  geom_jitter(aes(color = type), alpha = 0.2, size = 0.5) +
-  scale_fill_manual(values = c("plum","lightgoldenrod1")) +
-  scale_color_manual(values = c("darkorchid3","gold1")) +
-  ggtitle("Delta_meth of PMD genes with significant demethylation
-and rich/normal in CpG") +
-  xlab("") + 
-  ylab("Delta_meth centenarian - newborn") +
-  theme_bw() +
-  theme(legend.position = "none")
-
-rich_med_delta_CT <- CT_density_analysis %>% 
-  filter(padj <= 0.05 & density_state != "poor") %>% 
-  pull(delta_meth) %>% 
-  median()
-rich_med_delta_CT
-
-rich_med_delta_nonCT <-nonCT_density_analysis %>% 
-  filter(padj <= 0.05 & density_state != "poor") %>% 
-  pull(delta_meth) %>% 
-  median()
-rich_med_delta_nonCT
-```
-
-Medians are now way closer to each other and we can see that distributions are 
-closer to each other. For CT genes, the median is still a loss of 
-`r abs(rich_med_delta_CT)*100`%, however for nonCT the median loss has reduce 
-to `r abs(rich_med_delta_nonCT)*100`% ! The two distributions are now clearly 
-comparable.
-It looks like the demethylation difference that we observed earlier between CT 
-and nonCT genes in PMD during aging is now gone when we only analyse gene whose 
-promoters are not poor in CpG.
-
-Let's do the same test as previously to see if there is a difference between 
-both distributions (Mann-Withney) and if it isn't due to the size of the sample.
-
-```{r delta_meth_rich_stat}
-wilcox.test(filter(CT_density_analysis, padj <= 0.05 & density_state != "poor")$delta_meth,
-            filter(nonCT_density_analysis, padj <= 0.05 & density_state != "poor")$delta_meth,
-            alternative = "greater")
-
-rich_CT_sign <- table(filter(CT_density_analysis, density_state != "poor")$padj <= 0.05)
-
-distrib_means_rich <- unlist(bplapply(1:1000, random_delta_distrib_means,
-                                data = filter(nonCT_density_analysis, density_state != "poor"), 
-                                n_sign = rich_CT_sign[[2]], n_notsign = rich_CT_sign[[1]]))
-
-mean_delta_rich <- mean(filter(CT_density_analysis, density_state != "poor")$delta_meth)
-mean_delta_rich
-
-table(distrib_means_rich >= mean(filter(CT_density_analysis, 
-                                        density_state != "poor")$delta_meth))
-
-hist(distrib_means_rich,
-     col="dodgerblue1",
-     border = "dodgerblue3",
-     breaks = 20,
-     main = paste0("Means of 1000 delta_meth distributions of nonCT genes when randomly 
-     taking ", rich_CT_sign[[2]], " significative and ", rich_CT_sign[[1]], " not significative genes"),
-     cex.main = 0.75,
-     xlab = "Means",
-     ylab = "Frequency",
-     xlim = c(-0.15, 0))
-abline(v = mean_delta_rich, col = "violetred1", lwd = 3)
-text(mean_delta_rich, 125, "Mean of rich CT genes 
-delta_meth distribution", col = "violetred3")
-```
-
-pvalue is not significative anymore which means that there is ano  difference 
-between both distribution.
-
-Also, we can see when doing 1000 mean of delta_meth with the same amount of 
-genes that the mean for CT genes is in the middle of the distribution, which
-means that there is no real difference between CT and nonCT demethylation when 
-taking the CpG density into account
-
-It thus looks like CpG density plays a big role in the protection as when we're 
-taking that into account in our analysis, the difference between CT and nonCT 
-disappears.
-
-When analysing genes with comlparable CpG density, there is no difference in the 
-demethylation of CT and nonCT genes in PMD. This shows that the richness in CpG 
-of the promoter gives a certain protection against PMD demethylation during aging.
-
-### solo-CpG methylation
-
-In summary, although CT genes are found in PMDs, they lose little methylation 
-and are protected from continued demethylation during aging. This protection 
-seems to be explained by the CpG density of the promoters. The demethylation of 
-CT genes visible in tumours is therefore not associated with cell proliferation.
-However, the question remains: why does a high CpG density protect against
-continued demethylation? Our hypothesis is that there may be a self-maintenance 
-system, i.e. when a CpG loses its methylation in a CpG-rich region, the other 
-still methylated CpG methylated CpGs attract the methylation maintenance 
-machinery, thereby promoting the remethylation of the demethylated site. It 
-should be noted that our observations are consistent with some analyses performed
-by Zhou et al. (2018) which showed that that the hypomethylation of PMDs was 
-better observed by focusing only on the methylation of certain CpGs, called 
-"solo-CpGs", corresponding to isolated CpGs at least 35 bp away from a 
-neighbouring CpG and flanked by a A or a T (WCGW).
-
-We're going to use Zhou's coordinate of solo-WCGW to look a bit more into that.
-
-```{r wcgw_bed, message = FALSE}
-solo_WCGW <- read_delim("data/zhou_solo_WCGW_hg38.bed", 
-    delim = "\t", escape_double = FALSE, 
-    col_names = FALSE, trim_ws = TRUE)
-solo_WCGW <- solo_WCGW[, -5]
-colnames(solo_WCGW) <- c("seqnames", "start", "end", "context")
-```
-
-```{r solo}
-CT_with_solo <- CTmeth_inCD4 %>%
-  filter(external_gene_name %in% filter(CT_genes, 
-                                        PMD & regulated_by_methylation)$external_gene_name) %>%
-  dplyr::select(external_gene_name, seqnames, start, age, methylation_level) %>% 
-  left_join(dplyr::select(CT_genes, external_gene_name, ensembl_gene_id)) %>% 
-  left_join(solo_WCGW)
-
-CT_with_solo <- CT_with_solo %>% 
-  mutate(solo_state = case_when(is.na(CT_with_solo$context) ~ "not-solo",
-                                !is.na(CT_with_solo$context) ~ "solo-CpG")) %>% 
-  mutate(type = "CT_genes")
-
-nonCT_with_solo <- nonCTmeth_inCD4 %>% 
-  filter(ensembl_gene_id %in% nonCT_meth_means_thr$ensembl_gene_id) %>%
-  dplyr::select(ensembl_gene_id, seqnames, start, age, methylation_level) %>%
-  left_join(solo_WCGW)
-
-nonCT_with_solo <- nonCT_with_solo %>% 
-  mutate(solo_state = case_when(is.na(nonCT_with_solo$context) ~ "not-solo",
-                                !is.na(nonCT_with_solo$context) ~ "solo-CpG"))%>% 
-  mutate(type = "nonCT_genes")
-
-CT_with_solo %>% 
-  dplyr::select(colnames(nonCT_with_solo)) %>% 
-  rbind(nonCT_with_solo) %>%
-
-  ggplot(aes(y = methylation_level, x = age, 
-             fill = solo_state)) +
-  geom_jitter(aes(color = age), alpha = 0.2, size = 0.5) +
-  geom_boxplot(outlier.shape = NA) +
-  scale_fill_manual(values = c("lightskyblue1", "darkseagreen2")) +
-  scale_color_manual(values = c("rosybrown3","palevioletred3","violetred4")) +
-  ggtitle("CpG methylation in all ages, depending
-on their solo state, in CT and nonCT genes") +
-  facet_grid(type ~ solo_state) +
-  theme_bw() 
-```
-
-What we see here confirms what zhou said and what we hypothesize, we can see 
-that solo-cpg are losing more methylation following aging.
-
 # SessionInfo

### 2023-08-23 Paper_code-PMD_analysis.Rmd
Separated full code in two files

--- 2023-08-20 Paper_code.Rmd
@@
 ---
-title: "Code used to generate paper results"
+title: "Code used to generate paper results : PMD analysis"
 author: "Julie Devis"
@@
 
-# Loading data and packages
-
 ```{r library, message = FALSE}
-library(readxl)
 library(readr)
 library(CTexploreR)
-library(Vennerable)
 library(biomaRt)
@@
 library(SummarizedExperiment)
-library(factoextra)
-library(UpSetR)
-library(ComplexHeatmap)
-library(circlize)
-library(SingleCellExperiment)
-library(org.Hs.eg.db)
-library(clusterProfiler)
-library(DOSE)
-library(tibble)
 library(BiocParallel)
@@
 library(Biostrings)
-```
-
-Gene names/synonyms required for databases cleaning
-
-```{r ensembl_synonym, warning=FALSE, message=FALSE}
-ensembl <- biomaRt::useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
-attributes_vector <- c("ensembl_gene_id", "external_gene_name",
-                       "external_synonym", "gene_biotype",
-                       "chromosome_name", "band", "start_position", "end_position",
-                       "strand")
-ensembl_gene_synonym <- as_tibble(getBM(attributes = attributes_vector, mart = ensembl))
-
-ensembl_gene_synonym <- ensembl_gene_synonym %>%
-  mutate(external_synonym = sub(pattern = "ORF", external_synonym, 
-                                replacement = "orf"))
-
-attributes_vector <- c("ensembl_gene_id", "external_gene_name")
-ensembl_gene_names <- as_tibble(getBM(attributes = attributes_vector, mart = ensembl))
-```
-
-```{r load_data, warning=FALSE, message=FALSE}
-GTEX_data <- CTdata::GTEX_data()
-normal_tissues_multimapping_data <- CTdata::normal_tissues_multimapping_data()
-CCLE_data <- CTdata::CCLE_data()
-TCGA_TPM <- CTdata::TCGA_TPM()
-DAC_treated_cells_multimapping <- CTdata::DAC_treated_cells_multimapping()
-load("../CTdata/eh_data/testis_sce.rda") 
-load("../CTdata/eh_data/scRNAseq_HPA.rda")
-load("../CTdata/eh_data/CT_mean_methylation_in_tissues.rda")
-```
-
-Common figures parameters
-
-```{r fig_param}
-legends_param <- list(
-  labels_gp = gpar(col = "black", fontsize = 6),
-  title_gp = gpar(col = "black", fontsize = 6),
-  row_names_gp = gpar(fontsize = 4),
-  annotation_name_side = "left")
-
-legend_colors <- c("#5E4FA2", "#3288BD", "#66C2A5", "#ABDDA4", "#E6F598",
-                   "#FFFFBF", "#FEE08B", "#FDAE61", "#F46D43", "#D53E4F",
-                   "#9E0142")
-chr_colors <- c("X-linked" = "deeppink", "Not X" = "royalblue1")
-
-meth_colors <- c("Methylation" = "lightgreen", "Not methylation" = "gray")
-```
-
-# Database comparison
-
-## Lists cleaning
-
-CT lists from other databases have been checked (using GTEx and our 
-`GTEx_expression()` funtion and GeneCards) in order to remove duplicated
-gene names or deprecated ones and allow comparison between databases.
-
-### CTdatabase
-
-Online list copied in a csv file, several lists exist so we combined them.
-
-We checked gene names that were a concetanation of two genes (choice using
-biomaRt synonyms to get the official one), checked which ones had the right names,
-removed duplicated genes, verified lost genes and added back those that should 
-be there.
-
-```{r CTdatabase_cleaning, message = FALSE}
-CTdatabase <- read_delim("data/CTdatabase1.csv", delim = ";", 
-                         escape_double = FALSE, trim_ws = TRUE)
-colnames(CTdatabase) <- c("Family", "Gene_Name", "Chromosomal_localization",
-                          "CT_identifier")
-CTdatabase_bis <- read_csv2("data/CTdatabase2.csv")
-CTdatabase <- left_join(CTdatabase, CTdatabase_bis, 
-                        by = c("Gene_Name" = "Gene_Symbol"))
-
-CTdatabase_single <- CTdatabase %>%
-  mutate(Gene_Name = sub(pattern = "/.*$", Gene_Name, replacement = ""))
-CTdatabase_single <- CTdatabase_single %>%
-  mutate(Gene_Name = sub(pattern = ",.*$", Gene_Name, replacement = ""))
-
-CTdatabase_official_names <- 
-  unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, 
-                       external_gene_name)) %>%
-  filter(external_gene_name %in% CTdatabase_single$Gene_Name) %>%
-  mutate(Gene_Name = external_gene_name) %>%
-  mutate(external_synonym = NA)
-CTdatabase_synonym <- 
-  ensembl_gene_synonym %>%
-  filter(external_synonym %in% CTdatabase_single$Gene_Name) %>%
-  mutate(Gene_Name = external_synonym) %>%
-  dplyr::select(ensembl_gene_id, external_gene_name, Gene_Name, external_synonym)
-CTdatabase_cleaned <- 
-  rbind(CTdatabase_official_names, CTdatabase_synonym) %>% 
-  left_join(CTdatabase_single)
-
-duplicated_genes <- CTdatabase_cleaned$Gene_Name[duplicated(CTdatabase_cleaned$Gene_Name)]
-bad_ids <- ensembl_gene_synonym %>%
-  filter(external_gene_name %in% duplicated_genes | external_synonym %in% duplicated_genes) %>%
-  filter(chromosome_name %in% grep(pattern = "H", x = chromosome_name, value = TRUE)) %>%
-  pull(ensembl_gene_id)
-CTdatabase_cleaned <- CTdatabase_cleaned %>%
-  dplyr::filter(!ensembl_gene_id %in% bad_ids)
-CTdatabase_cleaned <- CTdatabase_cleaned %>%
-  filter(!ensembl_gene_id == "ENSG00000052126")
-CTdatabase_cleaned <- CTdatabase_cleaned %>% 
-  filter(!(ensembl_gene_id == "ENSG00000183305" & Gene_Name == "MAGEA2"))
-CTdatabase_cleaned <- CTdatabase_cleaned %>% 
-  filter(!ensembl_gene_id == "ENSG00000204648")
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "CSAG3B")
-CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "CSAG2", "external_synonym"] <- "CSAG3B"
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "CT45A4")
-CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "CT45A3", "external_synonym"] <- "CT45A4"
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "LAGE-1b")
-CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "CTAG2", "external_synonym"] <- "LAGE-1b"
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "CT16.2")
-CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "PAGE5", "external_synonym"] <- "CT16.2"
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "SPANXB2")
-CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "SPANXB1", "external_synonym"] <- "SPANXB2"
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "SPANXE")
-CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "SPANXD", "external_synonym"] <- "SPANXE"
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "XAGE1C")
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "XAGE1D")
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "XAGE1E")
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "XAGE2B")
-CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "XAGE2", "external_synonym"] <- "XAGE2B"
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "CTAGE-2")
-CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "CTAGE1", "external_synonym"] <- "CTAGE-2"
-
-CTdatabase_cleaned <- ensembl_gene_synonym %>%
-  mutate(Gene_Name = external_synonym) %>%
-  filter(external_synonym == "CXorf61") %>%
-  dplyr::select(ensembl_gene_id, external_gene_name, Gene_Name, external_synonym) %>%
-  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "Cxorf61", 
-                    c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
-  rbind(CTdatabase_cleaned)
-CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name)) %>%
-  filter(external_gene_name == "CCNA1") %>%
-  mutate(Gene_Name = external_gene_name) %>%
-  mutate(external_synonym = NA) %>% 
-  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "cyclin A1", 
-                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
-  rbind(CTdatabase_cleaned)
-CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
-  filter(external_gene_name == "GOLGA6L2") %>%
-  filter(ensembl_gene_id == "ENSG00000174450") %>%
-  mutate(Gene_Name = external_gene_name) %>%
-  mutate(external_synonym = NA) %>% 
-  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "GOLGAGL2 FA", 
-                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
-  rbind(CTdatabase_cleaned)
-CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
-  filter(external_gene_name == "LYPD6B") %>%
-  mutate(Gene_Name = external_gene_name) %>%
-  mutate(external_synonym = NA) %>% 
-  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "LOC130576", 
-                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
-  rbind(CTdatabase_cleaned)
-CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
-  filter(external_gene_name == "CT62") %>%
-  mutate(Gene_Name = external_gene_name) %>%
-  mutate(external_synonym = NA) %>% 
-  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "LOC196993", 
-                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
-  rbind(CTdatabase_cleaned)
-CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
-  filter(external_gene_name == "CT75") %>%
-  filter(ensembl_gene_id == "ENSG00000291155") %>%
-  mutate(Gene_Name = external_gene_name) %>%
-  mutate(external_synonym = NA) %>% 
-  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "LOC440934", 
-                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
-  rbind(CTdatabase_cleaned)
-CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
-  filter(external_gene_name == "LINC01192") %>%
-  mutate(Gene_Name = external_gene_name) %>%
-  mutate(external_synonym = NA) %>% 
-  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "LOC647107", 
-                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
-  rbind(CTdatabase_cleaned)
-CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
-  filter(external_gene_name == "TSPY1") %>%
-  mutate(Gene_Name = external_gene_name) %>%
-  mutate(external_synonym = NA) %>% 
-  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "LOC728137", 
-                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
-  rbind(CTdatabase_cleaned)
-CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
-  filter(external_gene_name == "SSX2B") %>%
-  mutate(Gene_Name = external_gene_name) %>%
-  mutate(external_synonym = NA) %>% 
-  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "SSX2b", 
-                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
-  rbind(CTdatabase_cleaned)
-```
-
-### Jamin's list
-
-Excel file coming from supplemental data.
-
-```{r Jamin_clean, message = FALSE}
-Jamin_core_CT <- read_excel("data/Jamin_core_CT.xlsx")
-Jamin_core_CT[Jamin_core_CT$Gene == "KIAA1211", "Gene"] <- "CRACD"
-Jamin_core_CT[Jamin_core_CT$Gene == "CXorf67", "Gene"] <- "EZHIP"
-```
-
-### Wang's CTatlas
-
-Excel file coming from supplemental data.
-
-```{r Wang_cleaning, message = FALSE}
-Wang_CT <- read_excel("data/Wang_Suppl_Data_3.xlsx", 
-    sheet = "Supplementary Data 3B", skip = 1)
-colnames(Wang_CT)[1] <- "ensembl_gene_id"
-
-Wang_CT <- ensembl_gene_names %>% 
-  filter(ensembl_gene_id %in% Wang_CT$ensembl_gene_id) %>%
-  right_join(Wang_CT)
-
-Wang_CT[Wang_CT$ensembl_gene_id == "ENSG00000181013", "external_gene_name"] <- "C17orf47"
-Wang_CT[Wang_CT$ensembl_gene_id == "ENSG00000204293", "external_gene_name"] <- "OR8B2"
-Wang_CT[Wang_CT$external_gene_name == "", "external_gene_name"] <- "RNASE11"
-Wang_CT[Wang_CT$external_gene_name == "CHCT1", "external_gene_name"] <- "C17orf64"
-Wang_CT[Wang_CT$external_gene_name == "PRSS40A", "external_gene_name"] <- "TISP43"
-Wang_CT[Wang_CT$external_gene_name == "TEX56P", "external_gene_name"] <- "C6orf201"
-Wang_CT[Wang_CT$external_gene_name == "SLC25A51P4", "external_gene_name"] <- "RP11-113D6.10"
-Wang_CT[Wang_CT$external_gene_name == "TCP10L3", "external_gene_name"] <- "TCP10"
-Wang_CT[Wang_CT$external_gene_name == "SCAND3", "external_gene_name"] <- "ZBED9"
-```
-
-### Carter's list
-
-```{r Carter_clean, message = FALSE}
-Carter_CT_list <- read_table("data/Carter_CT_list.txt", skip = 1)
-Carter_CT <- Carter_CT_list[Carter_CT_list$CT_Expression,]
-
-Carter_CT[Carter_CT$Gene == "ENSG00000261649", "Gene_Name"] <- "GOLGA6L7"
-Carter_CT[Carter_CT$Gene == "ENSG00000239620", "Gene_Name"] <- "PRR20G"
-Carter_CT[Carter_CT$Gene == "ENSG00000168148", "Gene_Name"] <- "H3-4"
-Carter_CT[Carter_CT$Gene == "ENSG00000204296", "Gene_Name"] <- "TSBP1"
-Carter_CT[Carter_CT$Gene == "ENSG00000180219", "Gene_Name"] <- "GARIN6"
-Carter_CT[Carter_CT$Gene == "ENSG00000172717", "Gene_Name"] <- "GARIN2"
-Carter_CT[Carter_CT$Gene == "ENSG00000174015", "Gene_Name"] <- "CBY2"
-Carter_CT[Carter_CT$Gene == "ENSG00000224960", "Gene_Name"] <- "PPP4R3C"
-```
-
-## CTexploreR data for selection pipeline
-
-To characterise the differences between our database and other, we need the 
-category we created in CTexploreR.
-
-Hereunder is what we used for our selection pipeline (coming from 
-`make_CT_list.R` in `CTdata`), mainly how we created the data.
-
-```{r make_CT, message = FALSE, warning = FALSE}
-# GTEX data with the tissue specificity category determined
-all_genes <- as_tibble(rowData(GTEX_data), rownames = "ensembl_gene_id")
-all_genes$TPM_testis <- assay(GTEX_data)[, "Testis"]
-
-# Add multimapping_analysis assessing testis-specificity of genes "lowly_expressed"
-
-all_genes <- all_genes %>%
-  left_join(as_tibble(rowData(normal_tissues_multimapping_data), 
-                      rownames = "ensembl_gene_id"))
-
-# Summarise both specificity
-
-all_genes <- all_genes %>%
-  mutate(testis_specificity = case_when(
-    GTEX_category == "testis_specific" |
-      multimapping_analysis == "testis_specific" ~ "testis_specific",
-    GTEX_category == "testis_preferential" ~ "testis_preferential"))
-
-# Add testis scRNA seq highest expressed cell type
-
-all_genes <- all_genes %>%
-  left_join(as_tibble(rowData(testis_sce)) %>%
-              dplyr::select(external_gene_name, testis_cell_type))
-
-# Add higher in somatic scRNAseq data of normal tissues from the Human Protein Atlas
-
-all_genes <- all_genes %>%
-  left_join(as_tibble(rowData(scRNAseq_HPA), rownames = "ensembl_gene_id") %>%
-              dplyr::select(ensembl_gene_id, external_gene_name,
-                            Higher_in_somatic_cell_type))
-
-# CCLE database analysis added
-
-all_genes <- all_genes %>%
-  left_join(as_tibble(rowData(CCLE_data), rownames = "ensembl_gene_id"))
-all_genes[is.na(all_genes$CCLE_category), "CCLE_category"] <- "not_in_CCLE"
-
-# TCGA database analysis added
-
-all_genes <- all_genes %>%
-  left_join(as_tibble(rowData(TCGA_TPM), rownames = "ensembl_gene_id") %>%
-              dplyr::select(ensembl_gene_id, external_gene_name, percent_pos_tum,
-                            percent_neg_tum, max_TPM_in_TCGA, TCGA_category))
-
-all_genes[all_genes$lowly_expressed_in_GTEX == TRUE &
-            all_genes$multimapping_analysis == "testis_specific",
-          "TCGA_category"] <- "multimapping_issue"
-
-# Add DAC analysis
-induced <- as_tibble(rowData(DAC_treated_cells_multimapping), 
-                     rownames = "ensembl_gene_id") %>%
-  filter(induced == TRUE) %>%
-  pull(external_gene_name)
-
-all_genes <- all_genes %>%
-  mutate(DAC_induced = case_when(external_gene_name %in% induced ~ TRUE,
-                                 !external_gene_name %in% induced ~ FALSE))
-
-# Add the most biologically relevant transcript (canonical and coordinates)
-
-ensembl <- biomaRt::useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
-attributes_vector <- c("ensembl_gene_id",
-                       "external_gene_name",
-                       "ensembl_transcript_id",
-                       "external_transcript_name",
-                       "chromosome_name",
-                       "strand",
-                       "transcript_start",
-                       "transcript_end",
-                       "transcription_start_site",
-                       "transcript_length",
-                       "transcript_biotype",
-                       "transcript_is_canonical")
-transcripts_infos <- as_tibble(biomaRt::getBM(attributes = attributes_vector,
-                                              mart = ensembl))
-canonical_transcripts <- transcripts_infos %>%
-  filter(transcript_is_canonical == 1) %>%
-  filter(chromosome_name %in% c(1:22, "X", "Y", "MT")) %>%
-  filter(transcript_biotype == "protein_coding" | transcript_biotype == "lncRNA")
-all_genes <- all_genes %>%
-  left_join(canonical_transcripts %>%
-              dplyr::select(ensembl_gene_id,
-                            external_transcript_name, ensembl_transcript_id,
-                            chromosome_name, strand, transcript_start,
-                            transcript_end, transcription_start_site,
-                            transcript_length, transcript_biotype))
-
-```
-
-From there, we filtered based on the testis_specificity ("testis_specific" or 
-"testis_preferential"), testis_cell_type (not "Macrophage", "Endothelial",
-"Myoid", "Sertoli", "Leydig"), Higher_in_somatic_cell_type (TRUE), 
-CCLE_category ("activated") and TCGA_category ("activated" or 
-"multimapping_issue") to have our CT genes. Then the most relevant 
-transcript was validated manually (and check that there is a transcript 
-activated in testis and the same is activated in tumor).
-
-## CTexploreR VS CTdatabase
-
-```{r CTdatabase_ours}
-CTdatabase_ours <- Venn(list(CTdatabase = CTdatabase_cleaned$external_gene_name,
-                             CTexploreR = CT_genes$external_gene_name))
-plot(CTdatabase_ours)
-```
-
-## CTexploreR VS omics databases
-
-```{r upset_plot}
-core_ours <- Venn(list(Jamin = Jamin_core_CT$Gene, 
-                       CTexploreR = CT_genes$external_gene_name))
-
-Wang_ours <- Venn(list(Wang = Wang_CT$external_gene_name, 
-                       CTexploreR = CT_genes$external_gene_name))
-
-Carter_ours <- Venn(list(Carter_CT = Carter_CT$Gene_Name, 
-                         CTexploreR = CT_genes$external_gene_name))
-
-gene_list <- list(Jamin = Jamin_core_CT$Gene, 
-                  CTdatabase = CTdatabase_cleaned$external_gene_name, 
-                  CTatlas = Wang_CT$external_gene_name, 
-                  CTexploreR = CT_genes$external_gene_name, 
-                  Carter = Carter_CT$Gene_Name)
-
-upset_omics <- fromList(gene_list[-2])
-upset(upset_omics)
-```
-
-## Characterisation of differences with all databases
-
-```{r all_databases}
-common <- unique(c(core_ours@IntersectionSets[["11"]], 
-                   CTdatabase_ours@IntersectionSets[["11"]], 
-                   Wang_ours@IntersectionSets[["11"]], 
-                   Carter_ours@IntersectionSets[["11"]]))
-
-length(common)
-
-length(common)/dim(CT_genes)[1] * 100
-
-lost_list <- unique(c(core_ours@IntersectionSets[["10"]], 
-                 CTdatabase_ours@IntersectionSets[["10"]],
-                 Wang_ours@IntersectionSets[["10"]],
-                 Carter_ours@IntersectionSets[["10"]]))
-
-lost <- all_genes %>%
-  filter(external_gene_name %in% lost_list)
-
-not_specific <- filter(lost, is.na(testis_specificity))
-
-somatic_testis <- filter(lost, testis_cell_type %in% c( "Macrophage", 
-                                                        "Endothelial", "Myoid",
-                                                        "Sertoli", "Leydig"))
-
-somatic_tissue <- filter(lost, Higher_in_somatic_cell_type == TRUE)
-
-not_TCGA_activated <- filter(lost, TCGA_category != "activated" & 
-                               TCGA_category != "multimapping_issue")
-
-not_CCLE_activated <- filter(lost, CCLE_category  != "activated")
-
-transcript_prob <- lost %>% 
-  filter(testis_specificity == "testis_specific" |
-           testis_specificity == "testis_preferential") %>%
-  filter(TCGA_category == "activated" | TCGA_category == "multimapping_issue") %>% 
-  filter(CCLE_category == "activated") %>% 
-  dim()
-```
-
-We have lost `r length(lost_list)` genes in total. Among them, 
-`r dim(not_specific)[1]/length(lost_list)*100`% are not considered testis 
-specific, `r dim(somatic_testis)[1]/length(lost_list)*100`% are expressed in 
-testis somatic cells, `r dim(somatic_tissue)[1]/length(lost_list)*100`% are 
-expressed in somatic cells, `r dim(not_TCGA_activated)[1]/length(lost_list)*100`% 
-are not activated in TCGA samples, 
-`r dim(not_CCLE_activated)[1]/length(lost_list)*100`% are not activated in CCLE 
-cell lines and `r transcript_prob[1]/length(lost_list)*100`% is lost due to 
-transcripts problems.
-
-# CT genes characteristics
-
-```{r basic}
-table(CT_genes$testis_specificity)
-table(CT_genes$transcript_biotype)
-```
-
-Most genes are testis specific 
-(`r table(CT_genes$testis_specificity)[[2]]/dim(CT_genes)[1]*100`%).
-Most genes are mainly protein coding genes 
-(`r table(CT_genes$transcript_biotype)[[3]]/dim(CT_genes)[1]*100`%).
-
-## X chromosome and regulated by methylation
-
-In CTexploreR, genes have been characterised as regulated by methylation or not.
-
-```{r X_meth}
-table(CT_genes$X_linked)
-table(CT_genes$regulated_by_methylation)
-table(CT_genes$X_linked, CT_genes$regulated_by_methylation)
-```
-
-Genes are enriched on the X chromosome 
-(`r table(CT_genes$X_linked)[[2]]/dim(CT_genes)[1]*100`%). Also, 
-`r table(CT_genes$regulated_by_methylation)[[2]]` genes have been flagged as 
-regulated by methylation 
-(`r table(CT_genes$regulated_by_methylation)[[2]]/dim(CT_genes)[1]*100`%). 
-It is interesting to study the link between these two characteristics
-
-On the chromosome X, there is a clear enrichment of CT genes regulated by 
-methylation (83/93 chrX genes or 83/160 genes regulated by methylation).
-
-Let's check that with a statistical test, I here want to see if, on there is an 
-enrichment of genes regulated by methylation on the X chromosome. I need to do a
-Pearson Chi square test (to know if observed proportion differ from expected 
-proportion). It is a statistical method to determine if two categorical variables
-have a significant correlation between them. I can directly put a matrix 
-(my table) in the function
-
-```{r chisq}
-chisq.test(table(CT_genes$X_linked, CT_genes$regulated_by_methylation))
-```
-
-```{r X_meth_fig}
-CT_genes$chr_factor <- factor(CT_genes$chr,
-                       levels = c("1", "2", "3", "4", "5", "6", "7", "8", "9",
-                                  "10", "11", "12", "13", "14", "15", "16", "17", 
-                                  "18", "19", "20", "21", "22", "X", "Y"))
-
-CT_genes %>% 
-  mutate(regulated_by_methylation = ifelse(regulated_by_methylation,
-                                           "Regulated by methylation",
-                                           "Not regulated by methylation")) %>% 
-  ggplot(aes(x = chr_factor, fill = testis_specificity, color = X_linked)) +
-  geom_bar(stat = 'count') +
-  scale_fill_manual(values = c("goldenrod1", "mediumpurple1")) +
-  scale_color_manual(values = c("royalblue1", "deeppink"), guide = "none") +
-  facet_wrap(~ regulated_by_methylation, ncol = 2) +
-  theme_bw() +
-  xlab("Chromosome") +
-  ylab("Number of genes") +
-  theme(axis.text.x = element_text(angle = 90, hjust = 1),
-        legend.position = "top",
-        axis.title = element_text(size = 10, color = "gray10"))
-```
-
-There is indeed a significative link between regulation by methylation and being
-on the X chromosome. There is thus an enrichment of CT genes regulated by 
-methylation on the X chromosome (and inversely).
-
-## Tumour and methylation
-
-Using CTexploreR functions, we can explore all CT genes or focus on some 
-potential targets.
-
-### All CT
-
-For these heatmaps, the code comes from the function but has been copies to add 
-some annotations.
-
-```{r common_annot}
-chr_mat <- as.matrix(CT_genes$X_linked)
-chr_mat <- ifelse(chr_mat == TRUE, "X-linked", "Not X")
-rownames(chr_mat) <- CT_genes$external_gene_name
-row_ha_chr <- rowAnnotation(chr_factor = chr_mat,
-                            annotation_legend_param = legends_param,
-                            simple_anno_size = unit(0.5, "cm"),
-                            col = list(chr_factor = chr_colors),
-                            annotation_name_gp = gpar(fontsize = 8),
-                            annotation_name_side = "top")
-
-regulation_mat <- as.matrix(CT_genes$regulated_by_methylation)
-regulation_mat <- ifelse(regulation_mat == TRUE, "Methylation",
-                         "Not methylation")
-rownames(regulation_mat) <- CT_genes$external_gene_name
-
-row_ha_reg <- rowAnnotation(regulation = regulation_mat,
-                            annotation_legend_param = legends_param,
-                            simple_anno_size = unit(0.5, "cm"),
-                            col = list(regulation = meth_colors),
-                            annotation_name_gp = gpar(fontsize = 8),
-                            annotation_name_side = "top")
-
-left_annot <- c(row_ha_chr, row_ha_reg, gap = unit(1, "mm"))
-
-split <- data.frame(CT_genes$regulated_by_methylation, CT_genes$X_linked)
-```
-
-```{r TCGA_heatmap, message = FALSE}
-database <- TCGA_TPM
-database$tumor <- sub(pattern = 'TCGA-', x = database$project_id, '')
-database$type <- "Tumor"
-database$type[database$shortLetterCode == "NT"] <- "Peritumoral"
-database <- database[, order(database$tumor, database$type)]
-
-genes <- CT_genes$external_gene_name
-database <- database[rowData(database)$external_gene_name %in% genes, ]
-database <- database[match(genes, rowData(database)$external_gene_name), ]
-
-column_ha_type <- HeatmapAnnotation(
-  Type = database$type,
-  border = TRUE,
-  col = list(Type = c("Tumor" = "salmon",
-                      "Peritumoral" = "mediumseagreen")),
-  annotation_name_gp = gpar(fontsize = 8),
-  annotation_legend_param = legends_param)
-
-column_ha_tumor <- HeatmapAnnotation(
-  Tumor = database$tumor,
-  border = TRUE,
-  col = list(Tumor = c("BRCA" = "midnightblue", "COAD" = "darkorchid2",
-                 "ESCA" = "gold", "HNSC" = "deeppink2",
-                 "LUAD" = "seagreen", "LUSC" = "seagreen3",
-                 "SKCM" = "red3")),
-  annotation_name_gp = gpar(fontsize = 8),
-  annotation_legend_param = legends_param)
-
-split_by <- factor(database$tumor)
-col_annot <- column_ha_tumor
-
-mat <- log1p(assay(database))
-rownames(mat) <- rowData(database)$external_gene_name
-
-Heatmap(mat[, , drop = FALSE],
-        name = "logTPM",
-        column_title = paste0("Expression in TCGA samples (all)"),
-        column_split = split_by,
-        row_split = split,
-        row_title_gp = gpar(fontsize = 0),
-        col = colorRamp2(seq(0, max(mat), length = 11),
-                         legend_colors),
-        clustering_method_rows = "ward.D",
-        clustering_method_columns = "ward.D",
-        cluster_rows = TRUE,
-        show_column_names = FALSE,
-        cluster_columns = TRUE,
-        show_column_dend = FALSE,
-        show_row_dend = FALSE,
-        row_names_gp = gpar(fontsize = 4),
-        heatmap_legend_param = legends_param,
-        top_annotation = col_annot,
-        left_annotation = left_annot)
-
-```
-
-```{r CCLE_heatmap}
-database <- CCLE_data
-database$type <- tolower(database$type)
-genes <- CT_genes$external_gene_name
-database <- database[rowData(database)$external_gene_name %in% genes, ]
-database <- database[match(genes, rowData(database)$external_gene_name), ]
-
-mat <- log1p(assay(database))
-rownames(mat) <- rowData(database)$external_gene_name
-
-df_col <- data.frame("cell_line" = colData(database)$cell_line_name,
-                     "type" = as.factor(colData(database)$type))
-rownames(df_col) <- rownames(colData(database))
-df_col <- df_col[order(df_col$type), ]
-
-column_ha_type <- HeatmapAnnotation(
-  type = df_col$type,
-  border = TRUE,
-  annotation_name_gp = gpar(fontsize = 8),
-  annotation_legend_param = legends_param,
-  col = list(type = c("lung" = "seagreen3", "skin" = "red3",
-                 "bile_duct" = "mediumpurple1", "bladder" = "mistyrose2",
-                 "colorectal" = "plum", "lymphoma" = "steelblue1",
-                 "uterine" = "darkorange4", "myeloma" = "turquoise3", 
-                 "kidney" = "thistle4",
-                 "pancreatic" = "darkmagenta", "brain" = "palegreen2",
-                 "gastric" = "wheat3", "breast" = "midnightblue",
-                 "bone" = "sienna1", "head_and_neck" = "deeppink2",
-                 "ovarian" = "tan3", "sarcoma" = "lightcoral",
-                 "leukemia" = "steelblue4", "esophageal"= "khaki",
-                 "neuroblastoma" = "olivedrab1")))
-
-Heatmap(mat[, rownames(df_col), drop = FALSE],
-        name = "logTPM",
-        column_title = "Gene Expression in tumor cell lines (CCLE)",
-        column_split = factor(df_col$type),
-        row_split = split,
-        row_title_gp = gpar(fontsize = 0),
-        col = colorRamp2(seq(0, max(mat), length = 11),
-                          legend_colors),
-        clustering_method_rows = "ward.D",
-        clustering_method_columns = "ward.D",
-        cluster_rows = TRUE,
-        show_row_dend = FALSE,
-        show_column_names = FALSE,
-        cluster_columns = TRUE,
-        show_column_dend = FALSE,
-        row_names_gp = gpar(fontsize = 4),
-        heatmap_legend_param = legends_param,
-        top_annotation = c(column_ha_type),
-        left_annotation = left_annot)
-```
-
-```{r all_meth}
-genes <- CT_genes$external_gene_name
-database <- CT_mean_methylation_in_tissues[rownames(CT_mean_methylation_in_tissues) %in% genes]
-
-mat <- na.omit(assay(database))
-clustering_option <- TRUE
-
-row_ha_chr_meth <- rowAnnotation(chr = chr_mat[rownames(mat),],
-                            annotation_legend_param = legends_param,
-                            simple_anno_size = unit(0.5, "cm"),
-                            col = list(chr = chr_colors),
-                            annotation_name_gp = gpar(fontsize = 8),
-                            annotation_name_side = "top")
-
-row_ha_reg_meth <- rowAnnotation(regulation = regulation_mat[rownames(mat),],
-                            annotation_legend_param = legends_param,
-                            simple_anno_size = unit(0.5, "cm"),
-                            col = list(regulation = meth_colors),
-                            annotation_name_gp = gpar(fontsize = 8),
-                            annotation_name_side = "top")
-
-left_annot_meth <- c(row_ha_chr_meth, row_ha_reg_meth, gap = unit(1, "mm"))
-
-split_meth <- data.frame(filter(CT_genes, external_gene_name %in% rownames(mat))$regulated_by_methylation, 
-                    filter(CT_genes, external_gene_name %in% rownames(mat))$X_linked)
-
-Heatmap(mat,
-        column_title = 'Promoter mean methylation level by tissue',
-        name = 'Meth',
-        col = colorRamp2(c(1:100),
-                         colorRampPalette(c("moccasin","dodgerblue4"))(100)),
-        na_col = "gray80",
-        cluster_rows = clustering_option,
-        cluster_columns = FALSE,
-        row_split = split_meth,
-        row_title_gp = gpar(fontsize = 0),
-        show_row_names = TRUE,
-        show_heatmap_legend = TRUE,
-        show_row_dend = FALSE,
-        row_names_gp = gpar(fontsize = 3),
-        column_names_gp = gpar(fontsize = 8),
-        column_names_side = "bottom",
-        row_names_side = "right",
-        left_annotation = left_annot_meth)
-
-```
-
-### MAGE genes
-
-```{r MAGE_TCGA}
-MAGE_genes <- filter(CT_genes, family == "MAGE")$external_gene_name
-
-TCGA_expression(tumor = "all", 
-                genes = MAGE_genes,
-                units = "log_TPM")
-
-CCLE_expression(genes = MAGE_genes,
-                 type = c("lung", "skin", "bile_duct", "bladder", 
-                          "colorectal", "lymphoma", "uterine",
-                          "myeloma", "kidney", "pancreatic", "brain", 
-                          "gastric", "breast", "bone", "head_and_neck", 
-                          "ovarian", "sarcoma", "leukemia", "esophageal",
-                          "neuroblastoma"), units = "log_TPM")
-
-normal_tissues_mean_methylation(MAGE_genes, na.omit = TRUE)
-
-```
-
-## Single Cell analysis : timing of expression
-
-In CTexploreR, we there is single cell data from the testis, we thus can 
-analyse CT genes expression during spermatogenesis.
-
-There is only data for 251 of our 298 CT genes.
-
-NB : SSC, Spermatogonia and early spermatocytes are *premeiotic cells*. Late 
-spermatocytes (between both meiosis), round spermatid, elongated spermatid and 
-sperm are *postmeiotic cells*.
-
-```{r repartition_sc}
-genes_avail <- 
-  CT_genes$external_gene_name[CT_genes$external_gene_name %in% unique(rownames(testis_sce))]
-table(CT_genes$testis_cell_type)
-table(CT_genes$testis_cell_type)/length(genes_avail)*100
-```
-
-41% of genes are mainly expressed pre-meioticly.
-
-```{r sc_heatmap, message = FALSE}
-germ_cells <- c("SSC", "Spermatogonia", "Early_spermatocyte",
-                "Late_spermatocyte","Round_spermatid", "Elongated_spermatid",
-                "Sperm1", "Sperm2")
-somatic_cells <- c("Macrophage", "Endothelial", "Myoid", "Sertoli", "Leydig")
-
-testis_sce_CT <- testis_sce[genes_avail, ]
-
-mat <- SingleCellExperiment::logcounts(testis_sce_CT)
-
-df_col <- data.frame(clusters = colData(testis_sce_CT)$clusters,
-                     type = colData(testis_sce_CT)$type,
-                     Donor = colData(testis_sce_CT)$Donor)
-rownames(df_col) <- colnames(testis_sce_CT)
-
-df_col <- df_col[order(df_col$type),]
-df_col$lineage <- "Germ cells"
-df_col$lineage[df_col$type %in% somatic_cells] <- "Somatic cells"
-
-column_ha_type = HeatmapAnnotation(
-  type = df_col$type,
-  border = TRUE,
-  col = list(type = c("SSC" = "floralwhite", "Spermatogonia" = "moccasin",
-                   "Early_spermatocyte" = "gold", 
-                   "Late_spermatocyte" = "orange",
-                   "Round_spermatid" = "red2", 
-                   "Elongated_spermatid" = "darkred",
-                   "Sperm1" = "violet", "Sperm2" = "purple", 
-                   "Sertoli" = "gray", 
-                   "Leydig" = "cadetblue2", "Myoid" = "springgreen3", 
-                   "Macrophage" = "gray10",
-                   "Endothelial" = "steelblue")),
-
-  annotation_name_gp = gpar(fontsize = 8),
-  annotation_legend_param = legends_param)
-
-column_ha_lineage = HeatmapAnnotation(
-  lineage = df_col$lineage,
-  border = TRUE,
-  col = list(lineage = c("Germ cells" = "salmon", "Somatic cells" = "cyan4")),
-  annotation_name_gp = gpar(fontsize = 8),
-  annotation_legend_param = legends_param)
-
-scale_lims <- c(0, quantile(rowMax(mat), 0.75))
-top_annot <- c(column_ha_lineage, column_ha_type)
-
-# Until here is what's in the function, hereunder is my addition/change in Heatmap()
-
-CT_genes_avail <- filter(CT_genes, external_gene_name %in% genes_avail)
-
-chr_mat <- as.matrix(CT_genes_avail$X_linked)
-chr_mat <- ifelse(chr_mat == TRUE, "X-linked", "Not X")
-rownames(chr_mat) <- CT_genes_avail$external_gene_name
-row_ha_chr <- rowAnnotation(chr = chr_mat,
-                            annotation_legend_param = legends_param,
-                            simple_anno_size = unit(0.5, "cm"),
-                            col = list(chr = chr_colors),
-                            annotation_name_gp = gpar(fontsize = 8),
-                            annotation_name_side = "top")
-
-regulation_mat <- as.matrix(CT_genes_avail$regulated_by_methylation)
-regulation_mat <- ifelse(regulation_mat == TRUE, "Methylation",
-                         "Not methylation")
-rownames(regulation_mat) <- CT_genes_avail$external_gene_name
-
-row_ha_reg <- rowAnnotation(regulation = regulation_mat,
-                            annotation_legend_param = legends_param,
-                            simple_anno_size = unit(0.5, "cm"),
-                            col = list(regulation = meth_colors),
-                            annotation_name_gp = gpar(fontsize = 8),
-                            annotation_name_side = "top")
-
-left_annot <- c(row_ha_chr, row_ha_reg, gap = unit(1, "mm"))
-
-split <- data.frame(CT_genes_avail$regulated_by_methylation, CT_genes_avail$X_linked)
-
-Heatmap(mat[genes_avail, rownames(df_col), drop = FALSE],
-        name = "logCounts",
-        column_title = "Expression in testis cells (scRNAseq)",
-        column_split = df_col$type,
-        row_split = split,
-        row_title_gp = gpar(fontsize = 0),
-        show_column_names = FALSE,
-        show_column_dend = FALSE,
-        clustering_method_rows = "ward.D",
-        clustering_method_columns = "ward.D",
-        cluster_rows = TRUE,
-        cluster_columns = FALSE,
-        show_row_dend = FALSE,
-        row_names_gp = gpar(fontsize = 4),
-        col = colorRamp2(seq(scale_lims[1], scale_lims[2], length = 11),                 
-                         legend_colors),
-        top_annotation = top_annot,
-        left_annotation = left_annot,
-        heatmap_legend_param = legends_param)
-
-```
-
-We used these data to determine in which testis cell type
-each gene is mostly expressed.
-
-```{r main_cell_type_expression_barplot}
-CT_genes$main_cell_type_expression <- factor(CT_genes$testis_cell_type,
-                                                levels = c("SSC", 
-                                               "Spermatogonia", 
-                                               "Early_spermatocyte",
-                                               "Late_spermatocyte",
-                                               "Round_spermatid",
-                                               "Elongated_spermatid",
-                                               "Sperm1",
-                                               "Sperm2",
-                                               "Sertoli"))
-
-CT_genes %>% 
-  mutate(regulated_by_methylation = ifelse(regulated_by_methylation,
-                                           "Regulated by methylation",
-                                           "Not regulated by methylation")) %>%
-  filter(!is.na(testis_cell_type)) %>%
-  filter(main_cell_type_expression != "Sertoli") %>% 
-  ggplot(aes(x = chr, fill = main_cell_type_expression,
-             color = main_cell_type_expression)) +
-  scale_fill_manual(values = c("lightyellow2", "moccasin", "gold", "orange","red2",
-                               "darkred", "violet", "purple")) +
-  scale_color_manual(values = c("lightyellow3", "navajowhite2", "gold", "orange","red2",
-                               "darkred", "violet", "purple")) +
-  geom_bar(stat = 'count') +
-  facet_grid(main_cell_type_expression ~ regulated_by_methylation) +
-  xlab("Chromosome") +
-  ylab("Number of genes") +
-  theme_bw() +
-  theme(
-    axis.text.x = element_text(color = "black", angle = 90, vjust = 0.5, size = 10),
-    axis.title = element_text(size = 10, color = "gray10"),
-    strip.text.y = element_blank())
-
-```
-
-14 genes are on the X chromosome but escape X chromosome inactivation and are
-activates post-meioticly.
-
-```{r escapees}
-CT_genes %>% 
-  filter(X_linked) %>% 
-  filter(testis_cell_type %in% c("Late_spermatocyte","Round_spermatid", 
-                                 "Elongated_spermatid", "Sperm1", "Sperm2"))
-
-```
-
-## Gene function
-
-```{r GO_ORA}
-go_ora <- enrichGO(gene = CT_genes$ensembl_gene_id,
-                   keyType = "ENSEMBL",
-                   OrgDb = org.Hs.eg.db,
-                   ont = "all",
-                   readable = TRUE)
-as_tibble(go_ora)
-
-as_tibble(go_ora) %>% 
-  arrange(desc(Count)) %>% 
-  head(12) %>% 
-  mutate(Ratio = Count/183) %>% 
-  ggplot(aes(x = Ratio, y = Description, fill = Description)) +
-  geom_col() +
-  theme_bw() +
-  ylab("GO term") +
-  xlab("Gene Ratio") +
-  theme(axis.text.y = element_blank(),
-        legend.position = "none",
-        axis.ticks.y = element_blank(),
-        axis.title = element_text(size = 10, color = "gray10"))+ 
-  geom_text(aes(0, y = Description, label = Description),
-            hjust = 0,
-            nudge_x = 0.005,
-            colour = "floralwhite",
-            size = 4)
-
-```
-
-As we can see here, most of genes are indeed linked to functions from reproduction.
-I represented here the 12 categories with the most genes, all from biological 
-processes. However, they are enriched in 89 different GO terms.
-
-We've also added a column tumor suppressor and oncogene in CTexploreR. 
-These information come from Cancermine.
-
-```{r C_TS}
-table(CT_genes$oncogene) 
-table(CT_genes$tumor_suppressor)
-
-table(CT_genes$oncogene, CT_genes$tumor_suppressor) 
 ```
@@
 
+ensembl <- biomaRt::useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
 attributes_vector <- c("ensembl_gene_id","external_gene_name", "chromosome_name",

### 2023-08-28 Paper_code-Database_CT.Rmd
Minor change

@@
 library(SummarizedExperiment)
-library(factoextra)
 library(UpSetR)

### 2023-08-28 Paper_code-Database_CT.Rmd
Update with CTexploreR last version

@@
 ```{r library, message = FALSE}
+.libPaths("/storage/research/dduv/cbio-lg/rlib/Bioc318/", include.site = FALSE)
 library(readxl)
@@
 DAC_treated_cells_multimapping <- CTdata::DAC_treated_cells_multimapping()
-load("../CTdata/eh_data/testis_sce.rda") 
-load("../CTdata/eh_data/scRNAseq_HPA.rda")
-load("../CTdata/eh_data/CT_mean_methylation_in_tissues.rda")
+testis_sce <- CTdata::testis_sce()
+scRNAseq_HPA <- CTdata::scRNAseq_HPA()
+CT_mean_methylation_in_tissues <- CTdata::CT_mean_methylation_in_tissues()
 ```

### 2023-08-28 Paper_code-PMD_analysis.Rmd
Update with CTexploreR last version

@@
 ```{r library, message = FALSE}
+.libPaths("/storage/research/dduv/cbio-lg/rlib/Bioc318/", include.site = FALSE)
 library(readr)
@@
 
-nonCT_per_gene_CD4_datasets <- bplapply(unique(nonCTmeth_inCD4$ensembl_gene_id),
-                                      reorganize_data_ensembl, data = nonCTmeth_inCD4)
-names(nonCT_per_gene_CD4_datasets) <- unique(nonCTmeth_inCD4$ensembl_gene_id)
-
-p <- rep(NA, length(nonCT_per_gene_CD4_datasets))
-for(i in 1:length(nonCT_per_gene_CD4_datasets)) {
- p[i] <- nrow(nonCT_per_gene_CD4_datasets[[i]]) > 1 &
-   is.data.frame(nonCT_per_gene_CD4_datasets[[i]])
-}
-nonCT_per_gene_CD4_datasets <- nonCT_per_gene_CD4_datasets[p]
-
-k <- rep(NA, length(nonCT_per_gene_CD4_datasets))
-for(i in 1:length(nonCT_per_gene_CD4_datasets)) {
-  k[i] <- nrow(nonCT_per_gene_CD4_datasets[[i]]) == 2
-}
-
-for (gene in names(nonCT_per_gene_CD4_datasets[k])) {
-  tmp <- rep(NA, ncol(nonCT_per_gene_CD4_datasets[[gene]]))
-  nonCT_per_gene_CD4_datasets[[gene]] <- rbind(nonCT_per_gene_CD4_datasets[[gene]], tmp)
-  rownames(nonCT_per_gene_CD4_datasets[[gene]])[3] <-
-    c("newborn","middle_age", "centenarian")[(!c("newborn","middle_age","centenarian") %in%
-                                                rownames(nonCT_per_gene_CD4_datasets[[gene]]))]
-  # naming the new row
-}
-
-for (gene in names(nonCT_per_gene_CD4_datasets)) {
-  nonCT_per_gene_CD4_datasets[[gene]] <- arrange(nonCT_per_gene_CD4_datasets[[gene]],
-                                              rownames(nonCT_per_gene_CD4_datasets[[gene]]))
-  # Put rows in the right order
-}
-
-rm(p, i, k, tmp)
-
-# load("data_rmd/nonCT_per_gene_CD4_datasets.rda")
+# nonCT_per_gene_CD4_datasets <- bplapply(unique(nonCTmeth_inCD4$ensembl_gene_id),
+#                                       reorganize_data_ensembl, data = nonCTmeth_inCD4)
+# names(nonCT_per_gene_CD4_datasets) <- unique(nonCTmeth_inCD4$ensembl_gene_id)
+# 
+# p <- rep(NA, length(nonCT_per_gene_CD4_datasets))
+# for(i in 1:length(nonCT_per_gene_CD4_datasets)) {
+#  p[i] <- nrow(nonCT_per_gene_CD4_datasets[[i]]) > 1 &
+#    is.data.frame(nonCT_per_gene_CD4_datasets[[i]])
+# }
+# nonCT_per_gene_CD4_datasets <- nonCT_per_gene_CD4_datasets[p]
+# 
+# k <- rep(NA, length(nonCT_per_gene_CD4_datasets))
+# for(i in 1:length(nonCT_per_gene_CD4_datasets)) {
+#   k[i] <- nrow(nonCT_per_gene_CD4_datasets[[i]]) == 2
+# }
+# 
+# for (gene in names(nonCT_per_gene_CD4_datasets[k])) {
+#   tmp <- rep(NA, ncol(nonCT_per_gene_CD4_datasets[[gene]]))
+#   nonCT_per_gene_CD4_datasets[[gene]] <- rbind(nonCT_per_gene_CD4_datasets[[gene]], tmp)
+#   rownames(nonCT_per_gene_CD4_datasets[[gene]])[3] <-
+#     c("newborn","middle_age", "centenarian")[(!c("newborn","middle_age","centenarian") %in%
+#                                                 rownames(nonCT_per_gene_CD4_datasets[[gene]]))]
+#   # naming the new row
+# }
+# 
+# for (gene in names(nonCT_per_gene_CD4_datasets)) {
+#   nonCT_per_gene_CD4_datasets[[gene]] <- arrange(nonCT_per_gene_CD4_datasets[[gene]],
+#                                               rownames(nonCT_per_gene_CD4_datasets[[gene]]))
+#   # Put rows in the right order
+# }
+# 
+# rm(p, i, k, tmp)
+
+load("data_rmd/nonCT_per_gene_CD4_datasets.rda")
 

### 2023-09-18 Paper_code-Database_CT.Rmd
small change in TCGA_expression

@@
 database <- database[match(genes, rowData(database)$external_gene_name), ]
-
-column_ha_type <- HeatmapAnnotation(
-  Type = database$type,
-  border = TRUE,
-  col = list(Type = c("Tumor" = "salmon",
-                      "Peritumoral" = "mediumseagreen")),
-  annotation_name_gp = gpar(fontsize = 8),
-  annotation_legend_param = legends_param)
+database <- database[, database$type == "Tumor"]
 

### 2023-10-19 Paper_code-Database_CT.Rmd
Added description of lost genes

@@
                              CTexploreR = CT_genes$external_gene_name))
-plot(CTdatabase_ours)
+gp <- VennThemes(compute.Venn(CTdatabase_ours))
+gp[["Face"]][["11"]]$fill <-  "mistyrose"
+gp[["Face"]][["01"]]$fill <-  "darkseagreen1"
+gp[["Face"]][["10"]]$fill <-  "lightsteelblue1"
+gp[["Set"]][["Set1"]]$col <-  "paleturquoise4"
+gp[["Set"]][["Set2"]]$col <-  "darkseagreen4"
+gp[["SetText"]][["Set1"]]$col <-  "paleturquoise4"
+gp[["SetText"]][["Set2"]]$col <-  "darkseagreen4"
+plot(CTdatabase_ours, gp = gp)
+
+```
+
+Lost genes analysis
+
+```{r CTdatabase_lost}
+CTdatabase_lost <- all_genes %>%
+  filter(external_gene_name %in% CTdatabase_ours@IntersectionSets[["10"]])
+
+# 10 Genes are lost because not in any  database
+
+CTdatabase_lost[is.na(CTdatabase_lost$testis_specificity), ]$testis_specificity <- "not_specific"
+table(CTdatabase_lost$testis_specificity)
+
+table(CTdatabase_lost$testis_cell_type)
+
+table(CTdatabase_lost$Higher_in_somatic_cell_type)
+
+table(CTdatabase_lost$TCGA_category)
+
+table(CTdatabase_lost$CCLE_category)
 ```
@@
 ## CTexploreR VS omics databases
-
 ```{r upset_plot}
@@
                          CTexploreR = CT_genes$external_gene_name))
-
 gene_list <- list(Jamin = Jamin_core_CT$Gene, 
@@
 
+Lost genes analysis
+
+```{r Jamin_lost}
+Jamin_lost <- all_genes %>%
+  filter(external_gene_name %in% core_ours@IntersectionSets[["10"]])
+
+Jamin_lost[is.na(Jamin_lost$testis_specificity), ]$testis_specificity <- "not_specific"
+table(Jamin_lost$testis_specificity)
+
+table(Jamin_lost$testis_cell_type)
+
+table(Jamin_lost$Higher_in_somatic_cell_type)
+
+table(Jamin_lost$TCGA_category)
+
+table(Jamin_lost$CCLE_category)
+```
+
+```{r Wang_lost}
+Wang_lost <- all_genes %>%
+  filter(external_gene_name %in% Wang_ours@IntersectionSets[["10"]])
+
+Wang_lost[is.na(Wang_lost$testis_specificity), ]$testis_specificity <- "not_specific"
+table(Wang_lost$testis_specificity)
+
+table(Wang_lost$testis_cell_type)
+
+table(Wang_lost$Higher_in_somatic_cell_type)
+
+table(Wang_lost$TCGA_category)
+
+table(Wang_lost$CCLE_category)
+
+```
+
+```{r Carter_lost}
+Carter_lost <- all_genes %>%
+  filter(external_gene_name %in% Carter_ours@IntersectionSets[["10"]])
+
+Carter_lost[is.na(Carter_lost$testis_specificity), ]$testis_specificity <- "not_specific"
+table(Carter_lost$testis_specificity)
+
+table(Carter_lost$testis_cell_type)
+
+table(Carter_lost$Higher_in_somatic_cell_type)
+
+table(Carter_lost$TCGA_category)
+
+table(Carter_lost$CCLE_category)
+```
+
 ## Characterisation of differences with all databases
@@
 
+GTEX_expression(not_specific$external_gene_name, units = "log_TPM")
+
 somatic_testis <- filter(lost, testis_cell_type %in% c( "Macrophage", 
@@
 
+testis_expression(somatic_testis$external_gene_name)
+
 somatic_tissue <- filter(lost, Higher_in_somatic_cell_type == TRUE)
+
+HPA_cell_type_expression(somatic_tissue$external_gene_name)
 
@@
 
+TCGA_expression(not_TCGA_activated$external_gene_name,
+                tumor = "all",
+                units = "log_TPM")
+
 not_CCLE_activated <- filter(lost, CCLE_category  != "activated")
+
+CCLE_expression(not_CCLE_activated$external_gene_name,
+                  type = c("lung", "skin", "colorectal",
+                           "gastric", "breast", "head_and_neck"),
+                units = "log_TPM")
 

### 2023-10-20 Paper_code-Database_CT.Rmd
worked on paper

@@
 
-testis_expression(somatic_testis$external_gene_name)
+testis_expression(somatic_testis$external_gene_name, cells = "all")
 
@@
 table(CT_genes$regulated_by_methylation)
+table(CT_genes$X_linked, CT_genes$testis_specificity)
 table(CT_genes$X_linked, CT_genes$regulated_by_methylation)
@@
 (`r table(CT_genes$regulated_by_methylation)[[2]]/dim(CT_genes)[1]*100`%). 
-It is interesting to study the link between these two characteristics
+It is interesting to study the link between these two characteristics.
 

### 2023-10-24 Paper_code-Database_CT.Rmd
Change in function figures

@@
 
+All CT genes function
+
 ```{r GO_ORA}
-
 msigdbr(species = "Homo sapiens" , category = "C5") %>% 
@@
   head(12) %>% 
-  mutate(Ratio = Count/183) %>% 
+  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/177,
+                           ONTOLOGY == "CC"~ Count/197,
+                           ONTOLOGY == "MF"~ Count/186)) %>% 
   ggplot(aes(x = Ratio, y = Description, fill = Description)) +
@@
 
+ora_to_plot <- as_tibble(simplify(go_ora))
+
+# Combine the two cilium movement from BP
+ora_to_plot[ora_to_plot$ID == "GO:0003341", "Count"] <- 
+  ora_to_plot[ora_to_plot$ID == "GO:0003341", "Count"] +
+  ora_to_plot[ora_to_plot$ID == "GO:0060294", "Count"]
+
+ora_to_plot <- ora_to_plot %>% 
+  arrange(desc(Count)) %>% 
+  head(11) %>% 
+  filter(ID != "GO:0060294") %>% 
+  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/177,
+                           ONTOLOGY == "CC"~ Count/197,
+                           ONTOLOGY == "MF"~ Count/186))
+
+ora_to_plot %>% 
+  ggplot(aes(x = Ratio, y = Description, fill = Description)) +
+  geom_col() +
+  theme_bw() +
+  ylab("GO term") +
+  xlab("Gene Ratio") +
+  theme(axis.text.y = element_blank(),
+        legend.position = "none",
+        axis.ticks.y = element_blank(),
+        axis.title = element_text(size = 10, color = "gray10"))+ 
+  geom_text(aes(0, y = Description, label = Description),
+            hjust = 0,
+            nudge_x = 0.005,
+            colour = "floralwhite",
+            size = 3.7)
 ```
@@
 
+Is there a difference between meth reg or not ?
+
+```{r GO_ORA_meth}
+
+go_ora_meth <- enrichGO(gene = 
+                          filter(CT_genes, regulated_by_methylation)$ensembl_gene_id,
+                        keyType = "ENSEMBL",
+                        OrgDb = org.Hs.eg.db,
+                        ont = "all",
+                        readable = TRUE)
+go_ora_meth <- simplify(go_ora_meth)
+
+go_ora_not_meth <- enrichGO(gene = 
+                          filter(CT_genes, !regulated_by_methylation)$ensembl_gene_id,
+                        keyType = "ENSEMBL",
+                        OrgDb = org.Hs.eg.db,
+                        ont = "all",
+                        readable = TRUE)
+
+go_ora_not_meth <- simplify(go_ora_not_meth)
+
+go_ora_meth_plot <- as_tibble(go_ora_meth) %>% 
+  arrange(desc(Count)) %>% 
+  head(11) %>% 
+  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/101,
+                           ONTOLOGY == "MF"~ Count/107)) %>%  
+  mutate(regulation = "Methylation")
+
+go_ora_not_meth_plot <- as_tibble(go_ora_not_meth) %>% 
+  arrange(desc(Count)) %>% 
+  head(11) %>% 
+  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/76,
+                           ONTOLOGY == "CC"~ Count/93))%>%  
+  mutate(regulation = "Not methylation")
+
+rbind(go_ora_meth_plot, go_ora_not_meth_plot) %>% 
+  ggplot(aes(x = Ratio, y = Description, fill = Description)) +
+  geom_col() +
+  theme_bw() +
+  ylab("GO term") +
+  xlab("Gene Ratio") +
+  facet_wrap(~ regulation) + 
+  theme(legend.position = "none",
+        axis.ticks.y = element_blank(),
+        axis.text.y = element_text(size = 6, color = "gray10"),
+        axis.title.x = element_text(size = 10, color = "gray10"),
+        axis.title.y = element_blank())+ 
+  geom_text(aes(0, y = Description, label = ID),
+            hjust = 0,
+            nudge_x = 0.005,
+            colour = "floralwhite",
+            size = 2)
+
+```
+
 We've also added a column tumor suppressor and oncogene in CTexploreR. 

### 2023-11-02 Paper_code-Database_CT.Rmd
Small changes in analysis

@@
 ```{r library, message = FALSE}
-.libPaths("/storage/research/dduv/cbio-lg/rlib/Bioc318/", include.site = FALSE)
 library(readxl)
@@
 
+What about new genes in CTexploreR
+
+```{r}
+new <- CT_genes %>% 
+  filter(!external_gene_name%in%common) %>% 
+  pull(external_gene_name)
+
+length(new)
+
+# Charles' idea
+# TCGA_expression(tumor = "all", genes = new, units = "log_TPM")
+```
+
 # CT genes characteristics
@@
 ```{r GO_ORA_meth}
-
 go_ora_meth <- enrichGO(gene = 
@@
 go_ora_not_meth <- simplify(go_ora_not_meth)
+as_tibble(go_ora_not_meth)
 

### 2023-11-02 Paper_code-PMD_analysis.Rmd
Small changes in analysis

@@
 ```{r library, message = FALSE}
-.libPaths("/storage/research/dduv/cbio-lg/rlib/Bioc318/", include.site = FALSE)
 library(readr)

### 2023-11-16 Paper_code-Database_CT.Rmd
worked on coactivation in tumors

@@
 
-```{r}
+```{r, warning = FALSE}
 new <- CT_genes %>% 
-  filter(!external_gene_name%in%common) %>% 
-  pull(external_gene_name)
-
-length(new)
-
-# Charles' idea
-# TCGA_expression(tumor = "all", genes = new, units = "log_TPM")
-```
+  filter(!external_gene_name%in%common)
+
+new
+
+table(new$testis_specificity)
+table(new$X_linked, new$regulated_by_methylation)
+TCGA_expression(tumor = "all", genes = new$external_gene_name, 
+                units = "log_TPM")
+TCGA_expression(tumor = "all", 
+                genes = filter(new, X_linked & regulated_by_methylation)$external_gene_name, 
+                units = "log_TPM")
+```
+There are `r dim(new)[1]` new CT genes in CTexploreR. These are mainly testis
+specific and on autosomes. Regulation by methylation is not the majority of 
+them. This makes sens as other databases have been focused on regulation by 
+methylation, so as here we don't segregate with that we can find these. There 
+is only 11 new "major" CT that are on the X chromosome and regulated by 
+methylation. CT45 are not that new.
+
+Expression in tumours doesn't strike that much.
 
@@
 
+Next graph was removed from paper
+
 ```{r all_meth}
@@
 ```
+
+### Coactivation in tumours
+
+We know that CT genes that are regulated by methylation are activated in 
+tumours due to methylation. However, those that are not it is more blurry. It 
+is interesting to look at their expression to see if there are groups of genes 
+that are activated together in tumours which would suggest another activation :
+a transcription factor ? a gene expresion program from meiosis ? timing of 
+expression ? function ?
+
+```{r not_reg}
+not_regulated <- filter(CT_genes, !regulated_by_methylation)$external_gene_name
+```
+
+As they can blur the clustering, I sometimes remove genes that are in the 
+background and  those that are expressed a lot.
+
+- BRCA :
+
+```{r BRCA}
+TCGA_expression(filter(CT_genes, !regulated_by_methylation)$external_gene_name,
+                tumor = "BRCA", units = "log_TPM")
+# Removing genes to see clearer
+
+BRCA_genes <- rowData(TCGA_TPM) %>% 
+  as_tibble() %>% 
+  filter(external_gene_name %in% not_regulated) %>% 
+  filter(percent_pos_BRCA > 0.01 & percent_pos_BRCA < 30) %>% 
+  pull(external_gene_name)
+
+BRCA_genes_ordered <- 
+  BRCA_genes[row_order(draw(TCGA_expression(BRCA_genes, tumor = "BRCA", 
+                                            units = "log_TPM")))]
+
+BRCA_cluster1 <- BRCA_genes_ordered[c(which(BRCA_genes_ordered == "AKAP14"):
+                       which(BRCA_genes_ordered == "C11orf97"))]
+BRCA_cluster2 <- BRCA_genes_ordered[c(which(BRCA_genes_ordered == "GTF2I-AS1"):
+                       which(BRCA_genes_ordered == "SLIT3-AS1"))]
+
+```
+
+We can see two small clusters.
+
+- LUAD :
+
+```{r LUAD}
+TCGA_expression(filter(CT_genes, !regulated_by_methylation)$external_gene_name,
+                tumor = "LUAD", units = "log_TPM")
+# Removing genes to see clearer
+LUAD_genes <- rowData(TCGA_TPM) %>% 
+  as_tibble() %>% 
+  filter(external_gene_name %in% not_regulated) %>% 
+  filter(percent_pos_LUAD > 0.01 & percent_pos_LUAD < 30) %>% 
+  pull(external_gene_name)
+
+LUAD_genes_ordered <- 
+  LUAD_genes[row_order(draw(TCGA_expression(LUAD_genes, tumor = "LUAD", 
+                                            units = "log_TPM")))]
+
+LUAD_cluster1 <- c("TSPY1", "TSPY2")
+LUAD_cluster2 <- c("SPANXC", "SPANXB1",  "SPANXD")
+LUAD_cluster3 <- LUAD_genes_ordered[c(1:22)]
+LUAD_cluster4 <- LUAD_genes_ordered[c(which(LUAD_genes_ordered == "GTF2I-AS1"):
+                       which(LUAD_genes_ordered == "SLIT3-AS1"))]
+```
+
+22 first genes are clearly quite clustering, there is also a small SPANX cluster,
+a small TSPY cluster ans a bigger one around the middle of the graph.
+
+- LUSC :
+
+```{r LUSC}
+TCGA_expression(filter(CT_genes, !regulated_by_methylation)$external_gene_name,
+                tumor = "LUSC", units = "log_TPM")
+# Removing genes to see clearer
+
+LUSC_genes <- rowData(TCGA_TPM) %>% 
+  as_tibble() %>% 
+  filter(external_gene_name %in% not_regulated) %>% 
+  filter(percent_pos_LUSC > 0.01 & percent_pos_LUSC < 30) %>% 
+  pull(external_gene_name)
+
+LUSC_genes_ordered <- 
+  LUSC_genes[row_order(draw(TCGA_expression(LUSC_genes, tumor = "LUSC", 
+                                            units = "log_TPM")))]
+
+LUSC_cluster1 <- c("TSPY1", "TSPY2",  "TSPY9", "TSPY10")
+LUSC_cluster2 <- c("SPANXC", "SPANXB1",  "SPANXD")
+LUSC_cluster3 <- LUSC_genes_ordered[c(1:15)]
+
+```
+
+We can see a cluster at the top and two smaller ones corresponding to SPANX and
+TSPY genes. As well as quite a big cluster with the 15 first lines
+
+- SKCM :
+
+```{r SKCM}
+SKCM_genes <- rowData(TCGA_TPM) %>% 
+  as_tibble() %>% 
+  filter(external_gene_name %in% not_regulated) %>% 
+  filter(percent_pos_SKCM > 0.01 & percent_pos_SKCM < 30) %>% 
+  pull(external_gene_name)
+
+TCGA_expression(filter(CT_genes, !regulated_by_methylation)$external_gene_name,
+                tumor = "SKCM", units = "log_TPM")
+# Lots of genes activated almost everywhere or nowhere that can mask clusters
+TCGA_expression(SKCM_genes,
+                tumor = "SKCM", units = "log_TPM")
+
+SKCM_cluster <- c("TSPY1", "TSPY2", "TSPY3", "TSPY9", "TSPY10")
+```
+
+It is clear that there is a TSPY cluster again.
+
+- Comparison
+
+TSPY and SPANX genes are found reactivated/coactivated in different types of 
+tumours. Let's compare the bigger groups found in LUAD, LUSC and BRCA
+
+```{r upste_clusters}
+upset(fromList(list(LUSC = LUSC_cluster3,
+               LUAD1 = LUAD_cluster3,
+               LUAD2 = LUAD_cluster4,
+               BRCA1 = BRCA_cluster1,
+               BRCA2 = BRCA_cluster2)))
+```
+
+There is quite a lot of similarities here between the clusters.
+
+Let's have a look at the timings of expression.
+
+```{r timing_cluster, message = FALSE}
+testis_expression(c("TSPY1", "TSPY2", "TSPY3", "TSPY9", "TSPY10"), 
+                  cells = "all")
+testis_expression(c("SPANXC", "SPANXB1",  "SPANXD"), 
+                  cells = "all")
+testis_expression(LUSC_cluster3, cells = "all")
+testis_expression(LUAD_cluster3, cells = "all")
+testis_expression(LUAD_cluster4, cells = "all")
+testis_expression(BRCA_cluster1, cells = "all")
+testis_expression(BRCA_cluster2, cells = "all")
+```
+
+They mainly seem to, in a cluster, be expressed at the same time.
 

### 2023-11-17 Paper_code-Database_CT.Rmd

@@
 
-```{r upste_clusters}
+```{r upset_clusters}
 upset(fromList(list(LUSC = LUSC_cluster3,
@@
 
-There is quite a lot of similarities here between the clusters.
+There is quite a lot of similarities here between the clusters. Let's look at
+the expression in tumours.
+
+```{r TCGA_clusters}
+TCGA_expression(c(LUSC_cluster3, LUAD_cluster3, LUAD_cluster4, BRCA_cluster1,
+                  BRCA_cluster2),
+                tumor = "all", units = "log_TPM")
+```
 

### 2023-11-17 Paper_code-Database_CT.Rmd
Added Bruggeman analysis

@@
 
-```{r, warning = FALSE}
+```{r new, warning = FALSE}
 new <- CT_genes %>% 
@@
 Expression in tumours doesn't strike that much.
+
+## Other databases : minor analysis
+
+There are ocassionnally other CT genes list that exist in litterature but there
+is not always details on the pipeline, the way it is maintained or handled. 
+There is usually a lack of structure.
+
+### Bruggeman 2018
+
+They started with the fact that most CT genes have been identified using whole 
+testis, this thus don’t make them germline specific and diminishes their 
+interest as targets for cancer immunotherapy. Their aim was thus to identify 
+true germ-cell specific cancer genes. They thus used transcriptomes of specific 
+germ cell-types present in healthy human spermatogenesis and compared it to data
+from GTEx and the TCGA. This allowed them to identify 756 germ cell specific 
+genes that are widely expressed in cancer. They state that "they’re certain that
+their genes are true GC-genes". However, they don't validate the known 
+enrichment in CT genes as their genes are distributed evenly across all 
+chromosomes
+
+The human male germ cells transcriptome they use comes from tissue samples from 
+6 males, cell subtypes were isolated using a laser capture microdissection 
+before RNA sequencing.
+
+Data was available through a web application that gives now an error, and is 
+also in an excel supplemental file.
+
+```{r Bruggeman}
+Bruggeman_CT <- read_excel("data/Bruggeman_suppl_data.xlsx", skip = 1,
+                           sheet = "1D")
+
+GTEX_expression(Bruggeman_CT$Gene, units = "log_TPM")
+TCGA_expression(Bruggeman_CT$Gene, units = "log_TPM", tumor = "all")
+testis_expression(Bruggeman_CT$Gene, cells = "all")
+```
+
+76 out of the 756 gene names aren't found in GTEx, and our data.
+
+GTEX expression : There is clearly a lack of testis specificty. Expression in 
+lymphocytes and fibroblasts is not surprising as they excluded it from their 
+analysis.
+
+TCGA expression : Most genes seems to be expressed in all tumours, which is the 
+reflection of an overall expression and not an activation.
+
+Testis cells expression : expression is indeed mainly in germ cells, even if 
+some genes are expressed in somatic cells. However, lots of genes are just not
+expressed in any of these cell types.
+
+```{r Bruggeman_ours}
+Bruggeman_ours <- Venn(list(Bruggeman = Bruggeman_CT$Gene,
+                            CTexploreR = CT_genes$external_gene_name))
+gp <- VennThemes(compute.Venn(Bruggeman_ours))
+gp[["Face"]][["11"]]$fill <-  "mistyrose"
+gp[["Face"]][["01"]]$fill <-  "darkseagreen1"
+gp[["Face"]][["10"]]$fill <-  "lightsteelblue1"
+gp[["Set"]][["Set1"]]$col <-  "paleturquoise4"
+gp[["Set"]][["Set2"]]$col <-  "darkseagreen4"
+gp[["SetText"]][["Set1"]]$col <-  "paleturquoise4"
+gp[["SetText"]][["Set2"]]$col <-  "darkseagreen4"
+plot(Bruggeman_ours, gp = gp)
+
+Bruggeman_others <- Venn(list(Bruggeman = Bruggeman_CT$Gene,
+                            Wang = Wang_CT$external_gene_name,
+                            CTdatabase = CTdatabase_cleaned$external_gene_name))
+plot(Bruggeman_others)
+```
+
+Despite the size of Bruggeman's list, there is a very small amount of genes
+in common with CTexploreR. This is not surprising seeing the patterns of 
+expression in normal and tumoral tissues above. I'll clarify the difference
+below.
+
+I however manage to recreate the Venn diagram of their fig 2, with slight 
+differences probably due to the cleaning of gene names I have done in Wang and 
+CTdatabase but not in Bruggeman.
+
+```{r Bruggeman_lost}
+Bruggeman_lost <- all_genes %>%
+  filter(external_gene_name %in% Bruggeman_ours@IntersectionSets[["10"]])
+# Missing the 76 genes
+
+Bruggeman_lost[is.na(Bruggeman_lost$testis_specificity), ]$testis_specificity <- "not_specific"
+table(Bruggeman_lost$testis_specificity)
+
+table(Bruggeman_lost$testis_cell_type)
+
+table(Bruggeman_lost$Higher_in_somatic_cell_type)
+
+table(Bruggeman_lost$TCGA_category)
+
+table(Bruggeman_lost$CCLE_category)
+```
+
+It is clear here that most genes are not testis specific. Even if they are 
+indeed mostly expressed in germ cells, they are not specific to it as they are 
+not testis specific. We can also see the small amount of genes activated in 
+TCGA samples and CCLE cell lines.
 

### 2023-11-29 Paper_code-Database_CT.Rmd
Added Bruggeman to the omics analysis

--- 2023-11-17 Paper_code-Database_CT.Rmd
@@
 ensembl_gene_names <- as_tibble(getBM(attributes = attributes_vector, mart = ensembl))
+
+attributes_vector <- c("external_gene_name",
+                       "external_synonym")
+gene_synonym <- as_tibble(getBM(attributes = attributes_vector, mart = ensembl))
 ```
@@
 
-We checked gene names that were a concetanation of two genes (choice using
+We checked gene names that were a concatenation of two genes (choice using
 biomaRt synonyms to get the official one), checked which ones had the right names,
@@
 
+### Burggeman's list
+
+Excel file from supplemental data.
+
+```{r Bruggeman_clean, message=FALSE}
+Bruggeman_data <- read_excel("data/Bruggeman_suppl_data.xlsx", skip = 1,
+                           sheet = "1D")
+
+Bruggeman_official_names <- gene_synonym %>% 
+  dplyr::select(external_gene_name) %>% 
+  unique() %>% 
+  filter(external_gene_name %in% Bruggeman_data$Gene) %>%
+  mutate(Gene_Name = external_gene_name) %>%
+  mutate(external_synonym = NA)
+
+Bruggeman_synonym <- gene_synonym %>%
+  filter(external_synonym %in% Bruggeman_data$Gene) %>%
+  mutate(Gene_Name = external_synonym) %>%
+  dplyr::select(external_gene_name, Gene_Name, external_synonym)
+
+Bruggeman_synonym <- Bruggeman_synonym[-which(Bruggeman_synonym$Gene_Name %in% 
+                          Bruggeman_official_names$Gene_Name),]
+
+Bruggeman_CT <- rbind(Bruggeman_official_names, Bruggeman_synonym)
+
+lost <- Bruggeman_data[which(!Bruggeman_data$Gene %in% c(Bruggeman_CT$Gene_Name)), "Gene"]
+colnames(lost) <- "external_gene_name"
+lost$Gene_Name <- rep(NA, 15)
+lost$external_synonym <- rep(NA, 15)
+
+lost[lost$external_gene_name == "C21orf59", "Gene_Name"] <- "CFAP298"
+lost[lost$external_gene_name == "C11orf57", "Gene_Name"] <- "NKAPD1"
+lost[lost$external_gene_name == "C7orf55", "Gene_Name"] <- "FMC1"
+lost[lost$external_gene_name == "C10orf12", "Gene_Name"] <- "LCOR"
+lost[lost$external_gene_name == "RPL19P12", "Gene_Name"] <- "RPL19P12"
+lost[lost$external_gene_name == "C16orf59", "Gene_Name"] <- "TEDC2"
+lost[lost$external_gene_name == "TTTY15", "Gene_Name"] <- "USP9Y"
+lost[lost$external_gene_name == "C17orf53", "Gene_Name"] <- "HROB"
+lost[lost$external_gene_name == "C12orf66", "Gene_Name"] <- "KICS2"
+lost[lost$external_gene_name == "C9orf84", "Gene_Name"] <- "SHOC1"
+lost[lost$external_gene_name == "C10orf25", "Gene_Name"] <- "ZNF22-AS1"
+lost[lost$external_gene_name == "C20orf197", "Gene_Name"] <- "LINC02910"
+lost[lost$external_gene_name == "C3orf67", "Gene_Name"] <- "CFAP20DC"
+lost[lost$external_gene_name == "C8orf37", "Gene_Name"] <- "CFAP418"
+lost[lost$external_gene_name == "C22orf34", "Gene_Name"] <- "MIR3667HG"
+
+Bruggeman_CT <- rbind(Bruggeman_CT, lost)
+
+missing_Bruggeman <- c("BMS1P4", "ADAM6", "ANXA2P3", "ARHGAP11B", "DPY19L2P2", 
+                       "HLA-L", "PA2G4P4", "PIPSL", "PRKY", "YBX3P1", 
+                       "RPL23AP53", "UQCRBP1", "RPL23P8", "MRS2P2", "PIN4P1", 
+                       "SLC6A10P", "GUSBP2", "PPIEL", "LRRC37BP1", "MSL3P1", 
+                       "PLEKHA8P1", "STAG3L1", "TCAM1P", "ZNF702P", "ZNF815P", 
+                       "ATP6AP1L", "RPL21P44", "SEC14L1P1", "ZNF876P", 
+                       "RPLP0P2", "FAM86JP", "FAM175A", "LACE1", "ATP5EP2", 
+                       "WDR92", "TCTE3", "METTL20", "KIAA2022", "ZNRD1-AS1", 
+                       "SGOL1", "FAM35DP", "MTL5", "TMEM14E", "MLLT4-AS1", 
+                       "CCDC173", "KIAA1524", "WDR78", "LINC00476", "LYRM5", 
+                       "HILS1", "CASC5", "KIAA1919", "CTAGE5", "FAM188B", 
+                       "TMEM194B", "FAM122C", "PPP1R2P3", "KIAA0391", "SGOL2", 
+                       "FAM19A3", "ZNF788", "RPL19P12")
+
+external_names_to_keep <- gene_synonym %>%
+   filter(external_synonym %in% missing_Bruggeman) %>%
+   filter(!external_gene_name %in% c("ATP5F1EP2", "POLR1HASP", "SHLD2P3", 
+                                   "TMEM14EP", "H1-9P", "ZNF788P")) %>% 
+   mutate(Gene_Name = external_gene_name)
+
+Bruggeman_CT[Bruggeman_CT$external_synonym %in% 
+                    external_names_to_keep$external_synonym, 
+                  "Gene_Name"] <- external_names_to_keep$Gene_Name
+
+Bruggeman_CT <- Bruggeman_CT %>% 
+  dplyr::select(Gene_Name)
+```
+
 ## CTexploreR data for selection pipeline
@@
                          CTexploreR = CT_genes$external_gene_name))
-gene_list <- list(Jamin = Jamin_core_CT$Gene, 
-                  CTdatabase = CTdatabase_cleaned$external_gene_name, 
+
+Bruggeman_ours <- Venn(list(Bruggeman = Bruggeman_CT$Gene_Name,
+                            CTexploreR = CT_genes$external_gene_name))
+
+gene_list <- list(CTexploreR = CT_genes$external_gene_name,
+                  Carter = Carter_CT$Gene_Name,
+                  Jamin = Jamin_core_CT$Gene, 
                   CTatlas = Wang_CT$external_gene_name,
-                  CTexploreR = CT_genes$external_gene_name, 
-                  Carter = Carter_CT$Gene_Name)
-
-upset_omics <- fromList(gene_list[-2])
+                  Bruggeman = Bruggeman_CT$Gene_Name)
+
+upset_omics <- fromList(gene_list)
 upset(upset_omics)
@@
 ```{r Jamin_lost}
+plot(core_ours, gp = gp)
+
 Jamin_lost <- all_genes %>%
@@
 ```{r Wang_lost}
+plot(Wang_ours, gp = gp)
+
 Wang_lost <- all_genes %>%
@@
 ```{r Carter_lost}
+plot(Carter_ours, gp = gp)
+
 Carter_lost <- all_genes %>%
@@
 table(Carter_lost$CCLE_category)
+```
+
+```{r Bruggeman_lost}
+plot(Bruggeman_ours, gp = gp)
+
+Bruggeman_lost <- all_genes %>%
+  filter(external_gene_name %in% Bruggeman_ours@IntersectionSets[["10"]])
+
+Bruggeman_lost[is.na(Bruggeman_lost$testis_specificity), ]$testis_specificity <- "not_specific"
+table(Bruggeman_lost$testis_specificity)
+
+table(Bruggeman_lost$testis_cell_type)
+
+table(Bruggeman_lost$Higher_in_somatic_cell_type)
+
+table(Bruggeman_lost$TCGA_category)
+
+table(Bruggeman_lost$CCLE_category)
 ```
@@
                    Wang_ours@IntersectionSets[["11"]], 
-                   Carter_ours@IntersectionSets[["11"]]))
+                   Carter_ours@IntersectionSets[["11"]],
+                   Bruggeman_ours@IntersectionSets[["11"]]))
 
@@
                       Wang_ours@IntersectionSets[["10"]],
-                 Carter_ours@IntersectionSets[["10"]]))
+                      Carter_ours@IntersectionSets[["10"]],
+                      Bruggeman_ours@IntersectionSets[["10"]]))
 

### 2024-04-25 Entire_PMD_analysis.Rmd
Updated PMD analysis as it will no longer be in paper
Changed everything to adapt to new gene list

--- 2023-11-02 Paper_code-PMD_analysis.Rmd
@@
 ```
+
+This should be completed using another model of senescene/immortalization by 
+confirming the difference between CTs and PMDs, and CTs and nonCT in PMDs.
 
@@
 ```{r CT_genes_prep}
+load("../CT_genes_latest_version.rda")
 CT_genes <- CT_genes %>% 
@@
      breaks = 20,
-     main = paste0("Percentage of promoters in PMDs when randomly taking ",
+     main = paste0("Percentage of promoters in PMDs when randomly taking",
                    sum(table(CT_meth_regulated$PMD)), 
@@
      ylab = "Frequency",
-     xlim = c(20,60))
+     xlim = c(20,70))
 abline(v = PMD_perc, col = "slateblue1", lwd = 3)
-text(PMD_perc, 120, "CT genes regulated 
+text(PMD_perc, 75, "CT genes regulated 
 by methylation", col = "slateblue4", cex = 0.8)
 abline(v = CT_not_meth_reg_PMD_perc, col = "darkseagreen1", lwd = 3)
-text(CT_not_meth_reg_PMD_perc, 120, "CT genes not regulated 
+text(CT_not_meth_reg_PMD_perc, 75, "CT genes not regulated 
 by methylation", col = "darkseagreen", cex = 0.8)
@@
 # nonCT_per_gene_CD4_datasets <- bplapply(unique(nonCTmeth_inCD4$ensembl_gene_id),
-#                                       reorganize_data_ensembl, data = nonCTmeth_inCD4)
+#                                       reorganize_data_ensembl, data = nonCTmeth_inCD4,
+#                                       BPPARAM = MulticoreParam(8L))
 # names(nonCT_per_gene_CD4_datasets) <- unique(nonCTmeth_inCD4$ensembl_gene_id)
@@
 # for (gene in names(nonCT_per_gene_CD4_datasets[k])) {
+#   print(gene)
 #   tmp <- rep(NA, ncol(nonCT_per_gene_CD4_datasets[[gene]]))
@@
 # rm(p, i, k, tmp)
+# save(nonCT_per_gene_CD4_datasets, 
+#      file = "data_rmd/nonCT_per_gene_CD4_datasets.rda")
 
@@
 
-nonCT_meth_means <- sapply(nonCT_per_gene_CD4_datasets, rowMeans, na.rm = TRUE)
-nonCT_meth_means <- nonCT_meth_means %>%
-  t() %>%
-  as.data.frame() %>%
-  rownames_to_column("ensembl_gene_id")
+nonCT_meth_means_tmp <- sapply(nonCT_per_gene_CD4_datasets, rowMeans, na.rm = TRUE)
+nonCT_meth_means <- Reduce(rbind, nonCT_meth_means_tmp) %>% 
+  data.frame()
+nonCT_meth_means$ensembl_gene_id <- names(nonCT_meth_means_tmp)
 

### 2024-04-25 Paper_code-Database_CT.Rmd
Changed everything to adapt to new gene list

@@
 CT_mean_methylation_in_tissues <- CTdata::CT_mean_methylation_in_tissues()
+####
+load("../CT_genes_latest_version.rda")
+CT_genes$testis_specificity <- "testis_specific"
 ```
@@
 colnames(lost) <- "external_gene_name"
-lost$Gene_Name <- rep(NA, 15)
-lost$external_synonym <- rep(NA, 15)
+lost$Gene_Name <- rep(NA, nrow(lost))
+lost$external_synonym <- rep(NA, nrow(lost))
 
@@
 lost[lost$external_gene_name == "C17orf53", "Gene_Name"] <- "HROB"
+lost[lost$external_gene_name == "C1orf112", "Gene_Name"] <- "FIRRM"
 lost[lost$external_gene_name == "C12orf66", "Gene_Name"] <- "KICS2"
@@
                        "TMEM194B", "FAM122C", "PPP1R2P3", "KIAA0391", "SGOL2", 
-                       "FAM19A3", "ZNF788", "RPL19P12")
+                       "FAM19A3", "ZNF788", "RPL19P12", "FIRRM")
 
@@
 
+We find `r length(CTdatabase_ours@IntersectionSets[["11"]])/nrow(CTdatabase_cleaned)*100`
+% of CTdatabase in CTexploreR, which is `r length(CTdatabase_ours@IntersectionSets[["11"]])/nrow(CT_genes)*100`
+% of our database.
+
 Lost genes analysis
@@
 
-# 10 Genes are lost because not in any database
+# 12 Genes are lost because not in any database
 
@@
 table(CTdatabase_lost$CCLE_category)
-```
+
+table(CTdatabase_lost$TCGA_category, CTdatabase_lost$CCLE_category)
+
+```
+
+`r (nrow(CTdatabase_lost) - table(CTdatabase_lost$testis_specificity)[["testis_specific"]])/nrow(CTdatabase_lost)*100`% of 
+these genes are not testis specific.
+
+`r (nrow(CTdatabase_lost) - table(CTdatabase_lost$TCGA_category, CTdatabase_lost$CCLE_category)["activated", "activated"])/nrow(CTdatabase_lost)*100`
+% are not properly activated in tumors and/or cancer cell lines.
 
 ## CTexploreR VS omics databases
+
 ```{r upset_plot}
@@
 
+4 in all, 59 in at least 3 databases
+
 Lost genes analysis
@@
 
+We find `r length(core_ours@IntersectionSets[["11"]])/nrow(Jamin_core_CT)*100`
+% of CTdatabase in CTexploreR, which is `r length(core_ours@IntersectionSets[["11"]])/nrow(CT_genes)*100`
+% of our database.
+
+`r (nrow(Jamin_lost) - table(Jamin_lost$testis_specificity)[["testis_specific"]])/nrow(Jamin_lost)*100`% of 
+these genes are not testis specific.
+
+`r (nrow(Jamin_lost) - table(Jamin_lost$TCGA_category, Jamin_lost$CCLE_category)["activated", "activated"])/nrow(Jamin_lost)*100`
+% are not properly activated in tumors and/or cancer cell lines.
+
 ```{r Wang_lost}
@@
 
+We find `r length(Wang_ours@IntersectionSets[["11"]])/nrow(Wang_CT)*100`
+% of CTdatabase in CTexploreR, which is `r length(Wang_ours@IntersectionSets[["11"]])/nrow(CT_genes)*100`
+% of our database.
+
+`r (nrow(Wang_lost) - table(Wang_lost$testis_specificity)[["testis_specific"]])/nrow(Wang_lost)*100`% of 
+these genes are not testis specific.
+
+`r (nrow(Wang_lost) - table(Wang_lost$TCGA_category, Wang_lost$CCLE_category)["activated", "activated"])/nrow(Wang_lost)*100`
+% are not properly activated in tumors and/or cancer cell lines.
+
 ```{r Carter_lost}
@@
 
+We find `r length(Carter_ours@IntersectionSets[["11"]])/nrow(Carter_CT)*100`
+% of CTdatabase in CTexploreR, which is `r length(Carter_ours@IntersectionSets[["11"]])/nrow(CT_genes)*100`
+% of our database.
+
+`r (nrow(Carter_lost) - table(Carter_lost$testis_specificity)[["testis_specific"]])/nrow(Carter_lost)*100`% of 
+these genes are not testis specific.
+
+`r (nrow(Carter_lost) - table(Carter_lost$TCGA_category, Carter_lost$CCLE_category)["activated", "activated"])/nrow(Carter_lost)*100`
+% are not properly activated in tumors and/or cancer cell lines.
+
 ```{r Bruggeman_lost}
@@
 ```
+
+We find `r length(Bruggeman_ours@IntersectionSets[["11"]])/nrow(Bruggeman_CT)*100`
+% of CTdatabase in CTexploreR, which is `r length(Bruggeman_ours@IntersectionSets[["11"]])/nrow(CT_genes)*100`
+% of our database.
+
+`r (nrow(Bruggeman_lost) - table(Bruggeman_lost$testis_specificity)[["testis_specific"]])/nrow(Bruggeman_lost)*100`% of 
+these genes are not testis specific.
+
+`r (nrow(Bruggeman_lost) - table(Bruggeman_lost$TCGA_category, Bruggeman_lost$CCLE_category)["activated", "activated"])/nrow(Bruggeman_lost)*100`
+% are not properly activated in tumors and/or cancer cell lines.
 
@@
 Most genes are testis specific 
-(`r table(CT_genes$testis_specificity)[[2]]/dim(CT_genes)[1]*100`%).
+(`r table(CT_genes$testis_specificity)[["testis_specific"]]/dim(CT_genes)[1]*100`%).
 Most genes are mainly protein coding genes 
-(`r table(CT_genes$transcript_biotype)[[3]]/dim(CT_genes)[1]*100`%).
+(`r table(CT_genes$transcript_biotype)["protein_coding"]/dim(CT_genes)[1]*100`%).
 
@@
 On the chromosome X, there is a clear enrichment of CT genes regulated by 
-methylation (83/93 chrX genes or 83/160 genes regulated by methylation).
+methylation (`r table(CT_genes$X_linked, CT_genes$regulated_by_methylation)[2, 2]`/`r table(CT_genes$X_linked)[[2]]` 
+chrX genes or `r table(CT_genes$X_linked, CT_genes$regulated_by_methylation)[2, 2]`/`r table(CT_genes$regulated_by_methylation)[[2]]` 
+genes regulated by methylation).
 
@@
                                            "Not regulated by methylation")) %>% 
-  ggplot(aes(x = chr_factor, fill = testis_specificity, color = X_linked)) +
+  ggplot(aes(x = chr_factor, fill = X_linked)) +
   geom_bar(stat = 'count') +
-  scale_fill_manual(values = c("goldenrod1", "mediumpurple1")) +
-  scale_color_manual(values = c("royalblue1", "deeppink"), guide = "none") +
+  scale_fill_manual(values = c("royalblue1", "deeppink")) +
   facet_wrap(~ regulated_by_methylation, ncol = 2) +
@@
   theme(axis.text.x = element_text(angle = 90, hjust = 1),
-        legend.position = "top",
+        legend.position = "none",
         axis.title = element_text(size = 10, color = "gray10"))
@@
 
-### Coactivation in tumours
-
-We know that CT genes that are regulated by methylation are activated in 
-tumours due to methylation. However, those that are not it is more blurry. It 
-is interesting to look at their expression to see if there are groups of genes 
-that are activated together in tumours which would suggest another activation :
-a transcription factor ? a gene expresion program from meiosis ? timing of 
-expression ? function ?
-
-```{r not_reg}
-not_regulated <- filter(CT_genes, !regulated_by_methylation)$external_gene_name
-```
-
-As they can blur the clustering, I sometimes remove genes that are in the 
-background and  those that are expressed a lot.
-
-- BRCA :
-
-```{r BRCA}
-TCGA_expression(filter(CT_genes, !regulated_by_methylation)$external_gene_name,
-                tumor = "BRCA", units = "log_TPM")
-# Removing genes to see clearer
-
-BRCA_genes <- rowData(TCGA_TPM) %>% 
-  as_tibble() %>% 
-  filter(external_gene_name %in% not_regulated) %>% 
-  filter(percent_pos_BRCA > 0.01 & percent_pos_BRCA < 30) %>% 
-  pull(external_gene_name)
-
-BRCA_genes_ordered <- 
-  BRCA_genes[row_order(draw(TCGA_expression(BRCA_genes, tumor = "BRCA", 
-                                            units = "log_TPM")))]
-
-BRCA_cluster1 <- BRCA_genes_ordered[c(which(BRCA_genes_ordered == "AKAP14"):
-                       which(BRCA_genes_ordered == "C11orf97"))]
-BRCA_cluster2 <- BRCA_genes_ordered[c(which(BRCA_genes_ordered == "GTF2I-AS1"):
-                       which(BRCA_genes_ordered == "SLIT3-AS1"))]
-
-```
-
-We can see two small clusters.
-
-- LUAD :
-
-```{r LUAD}
-TCGA_expression(filter(CT_genes, !regulated_by_methylation)$external_gene_name,
-                tumor = "LUAD", units = "log_TPM")
-# Removing genes to see clearer
-LUAD_genes <- rowData(TCGA_TPM) %>% 
-  as_tibble() %>% 
-  filter(external_gene_name %in% not_regulated) %>% 
-  filter(percent_pos_LUAD > 0.01 & percent_pos_LUAD < 30) %>% 
-  pull(external_gene_name)
-
-LUAD_genes_ordered <- 
-  LUAD_genes[row_order(draw(TCGA_expression(LUAD_genes, tumor = "LUAD", 
-                                            units = "log_TPM")))]
-
-LUAD_cluster1 <- c("TSPY1", "TSPY2")
-LUAD_cluster2 <- c("SPANXC", "SPANXB1",  "SPANXD")
-LUAD_cluster3 <- LUAD_genes_ordered[c(1:22)]
-LUAD_cluster4 <- LUAD_genes_ordered[c(which(LUAD_genes_ordered == "GTF2I-AS1"):
-                       which(LUAD_genes_ordered == "SLIT3-AS1"))]
-```
-
-22 first genes are clearly quite clustering, there is also a small SPANX cluster,
-a small TSPY cluster ans a bigger one around the middle of the graph.
-
-- LUSC :
-
-```{r LUSC}
-TCGA_expression(filter(CT_genes, !regulated_by_methylation)$external_gene_name,
-                tumor = "LUSC", units = "log_TPM")
-# Removing genes to see clearer
-
-LUSC_genes <- rowData(TCGA_TPM) %>% 
-  as_tibble() %>% 
-  filter(external_gene_name %in% not_regulated) %>% 
-  filter(percent_pos_LUSC > 0.01 & percent_pos_LUSC < 30) %>% 
-  pull(external_gene_name)
-
-LUSC_genes_ordered <- 
-  LUSC_genes[row_order(draw(TCGA_expression(LUSC_genes, tumor = "LUSC", 
-                                            units = "log_TPM")))]
-
-LUSC_cluster1 <- c("TSPY1", "TSPY2",  "TSPY9", "TSPY10")
-LUSC_cluster2 <- c("SPANXC", "SPANXB1",  "SPANXD")
-LUSC_cluster3 <- LUSC_genes_ordered[c(1:15)]
-
-```
-
-We can see a cluster at the top and two smaller ones corresponding to SPANX and
-TSPY genes. As well as quite a big cluster with the 15 first lines
-
-- SKCM :
-
-```{r SKCM}
-SKCM_genes <- rowData(TCGA_TPM) %>% 
-  as_tibble() %>% 
-  filter(external_gene_name %in% not_regulated) %>% 
-  filter(percent_pos_SKCM > 0.01 & percent_pos_SKCM < 30) %>% 
-  pull(external_gene_name)
-
-TCGA_expression(filter(CT_genes, !regulated_by_methylation)$external_gene_name,
-                tumor = "SKCM", units = "log_TPM")
-# Lots of genes activated almost everywhere or nowhere that can mask clusters
-TCGA_expression(SKCM_genes,
-                tumor = "SKCM", units = "log_TPM")
-
-SKCM_cluster <- c("TSPY1", "TSPY2", "TSPY3", "TSPY9", "TSPY10")
-```
-
-It is clear that there is a TSPY cluster again.
-
-- Comparison
-
-TSPY and SPANX genes are found reactivated/coactivated in different types of 
-tumours. Let's compare the bigger groups found in LUAD, LUSC and BRCA
-
-```{r upset_clusters}
-upset(fromList(list(LUSC = LUSC_cluster3,
-               LUAD1 = LUAD_cluster3,
-               LUAD2 = LUAD_cluster4,
-               BRCA1 = BRCA_cluster1,
-               BRCA2 = BRCA_cluster2)))
-```
-
-There is quite a lot of similarities here between the clusters. Let's look at
-the expression in tumours.
-
-```{r TCGA_clusters}
-TCGA_expression(c(LUSC_cluster3, LUAD_cluster3, LUAD_cluster4, BRCA_cluster1,
-                  BRCA_cluster2),
-                tumor = "all", units = "log_TPM")
-```
-
-Let's have a look at the timings of expression.
-
-```{r timing_cluster, message = FALSE}
-testis_expression(c("TSPY1", "TSPY2", "TSPY3", "TSPY9", "TSPY10"), 
-                  cells = "all")
-testis_expression(c("SPANXC", "SPANXB1",  "SPANXD"), 
-                  cells = "all")
-testis_expression(LUSC_cluster3, cells = "all")
-testis_expression(LUAD_cluster3, cells = "all")
-testis_expression(LUAD_cluster4, cells = "all")
-testis_expression(BRCA_cluster1, cells = "all")
-testis_expression(BRCA_cluster2, cells = "all")
-```
-
-They mainly seem to, in a cluster, be expressed at the same time.
-
 ### MAGE genes
@@
 
-There is only data for 251 of our 298 CT genes.
+There is only data for `r sum(CT_genes$external_gene_name %in% unique(rownames(testis_sce)))` of our `r nrow(CT_genes)` CT genes.
 
@@
 
-41% of genes are mainly expressed pre-meioticly.
+`r sum(table(CT_genes$testis_cell_type)[c("Early_spermatocyte", "Spermatogonia", "SSC")])/length(genes_avail)*100`
+% of genes are mainly expressed pre-meioticly.
 
@@
 
-14 genes are on the X chromosome but escape X chromosome inactivation and are
-activates post-meioticly.
-
 ```{r escapees}
@@
 ```
+
+These genes are on the X chromosome but escape X chromosome inactivation and are
+activates post-meioticly.
 
@@
 
-# Combine the two cilium movement from BP
-ora_to_plot[ora_to_plot$ID == "GO:0003341", "Count"] <- 
-  ora_to_plot[ora_to_plot$ID == "GO:0003341", "Count"] +
-  ora_to_plot[ora_to_plot$ID == "GO:0060294", "Count"]
-
 ora_to_plot <- ora_to_plot %>% 
   arrange(desc(Count)) %>% 
-  head(11) %>% 
-  filter(ID != "GO:0060294") %>% 
-  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/177,
-                           ONTOLOGY == "CC"~ Count/197,
+  head(9) %>% 
+  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/102,
                            ONTOLOGY == "MF"~ Count/186))
@@
 I represented here the 12 categories with the most genes, all from biological 
-processes. However, they are enriched in 89 different GO terms.
+processes. However, they are enriched in `r nrow(as_tibble(go_ora))` different GO terms.
 
@@
 go_ora_meth <- simplify(go_ora_meth)
+as_tibble(go_ora_meth)
 
@@
                         readable = TRUE)
-
 go_ora_not_meth <- simplify(go_ora_not_meth)
@@
   arrange(desc(Count)) %>% 
-  head(11) %>% 
-  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/101,
-                           ONTOLOGY == "MF"~ Count/107)) %>%  
+  head(10) %>% 
+  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/73,
+                           ONTOLOGY == "MF"~ Count/80,
+                           ONTOLOGY == "CC"~ Count/78)) %>%  
   mutate(regulation = "Methylation")
@@
   arrange(desc(Count)) %>% 
-  head(11) %>% 
-  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/76,
-                           ONTOLOGY == "CC"~ Count/93))%>%  
+  head(8) %>% 
+  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/29,
+                           ONTOLOGY == "MF"~ Count/30))%>%  
   mutate(regulation = "Not methylation")

### 2024-05-10 Paper_code-Database_CT.Rmd
Update and enhancement on database comparison

@@
 ```{r load_data, warning=FALSE, message=FALSE}
-GTEX_data <- CTdata::GTEX_data()
-normal_tissues_multimapping_data <- CTdata::normal_tissues_multimapping_data()
-CCLE_data <- CTdata::CCLE_data()
-TCGA_TPM <- CTdata::TCGA_TPM()
-DAC_treated_cells_multimapping <- CTdata::DAC_treated_cells_multimapping()
-testis_sce <- CTdata::testis_sce()
-scRNAseq_HPA <- CTdata::scRNAseq_HPA()
-CT_mean_methylation_in_tissues <- CTdata::CT_mean_methylation_in_tissues()
+# CCLE_data <- CTdata::CCLE_data()
+# TCGA_TPM <- CTdata::TCGA_TPM()
+# testis_sce <- CTdata::testis_sce()
 ####
-load("../CT_genes_latest_version.rda")
-CT_genes$testis_specificity <- "testis_specific"
+
+load("../CTdata/eh_data/CCLE_data.rda")
+load("../CTdata/eh_data/TCGA_TPM.rda")
+load("../CTdata/eh_data/testis_sce.rda")
+load("../CTdata/eh_data/mean_methylation_in_tissues.rda")
+
+load("../CTdata/eh_data/all_genes.rda")
+load("../CTdata/eh_data/HPA_cell_type_specificities.rda")
+load("../CTdata/eh_data/CT_genes.rda")
+
+CT_and_CTP_genes <- CT_genes
+CT_genes <- (filter(CT_genes, CT_gene_type == "CT_gene"))
 ```
@@
 meth_colors <- c("Methylation" = "lightgreen", "Not methylation" = "gray")
+```
+
+# Pipeline figure values
+
+Here is the code used to have all the values in the pipeline figure
+
+```{r pipeline figures value}
+table(all_genes$GTEX_category)
+table(all_genes$multimapping_analysis)
+
+table(all_genes$testis_specificity)
+specific_to_pref <- all_genes %>%
+  filter(GTEX_category == "testis_specific" |
+           multimapping_analysis == "testis_specific") %>% 
+  filter(testis_specificity == "testis_preferential")
+table(specific_to_pref$not_detected_in_somatic_HPA)
+table(specific_to_pref$CCLE_category)
+table(specific_to_pref$TCGA_category)
+
+all_genes %>% 
+  filter(testis_specificity != "not_testis_specific") %>% 
+  filter(CCLE_category == "activated" & 
+           (TCGA_category == "multimapping_issue" | 
+              TCGA_category == "activated")) %>% 
+  pull(testis_specificity) %>% 
+  table()
+
+table(all_genes$CT_gene_type) #Before and after TSS verif
 ```
@@
 To characterise the differences between our database and other, we need the 
-category we created in CTexploreR.
+category we created in CTexploreR. For this, we have the object `all_genes` in 
+CTdata that contains the CT analysis for all genes. More info in
 
 Hereunder is what we used for our selection pipeline (coming from 
-`make_CT_list.R` in `CTdata`), mainly how we created the data.
-
-```{r make_CT, message = FALSE, warning = FALSE}
-# GTEX data with the tissue specificity category determined
-all_genes <- as_tibble(rowData(GTEX_data), rownames = "ensembl_gene_id")
-all_genes$TPM_testis <- assay(GTEX_data)[, "Testis"]
-
-# Add multimapping_analysis assessing testis-specificity of genes "lowly_expressed"
-
-all_genes <- all_genes %>%
-  left_join(as_tibble(rowData(normal_tissues_multimapping_data), 
-                      rownames = "ensembl_gene_id"))
-
-# Summarise both specificity
-
-all_genes <- all_genes %>%
-  mutate(testis_specificity = case_when(
-    GTEX_category == "testis_specific" |
-      multimapping_analysis == "testis_specific" ~ "testis_specific",
-    GTEX_category == "testis_preferential" ~ "testis_preferential"))
-
-# Add testis scRNA seq highest expressed cell type
-
-all_genes <- all_genes %>%
-  left_join(as_tibble(rowData(testis_sce)) %>%
-              dplyr::select(external_gene_name, testis_cell_type))
-
-# Add higher in somatic scRNAseq data of normal tissues from the Human Protein Atlas
-
-all_genes <- all_genes %>%
-  left_join(as_tibble(rowData(scRNAseq_HPA), rownames = "ensembl_gene_id") %>%
-              dplyr::select(ensembl_gene_id, external_gene_name,
-                            Higher_in_somatic_cell_type))
-
-# CCLE database analysis added
-
-all_genes <- all_genes %>%
-  left_join(as_tibble(rowData(CCLE_data), rownames = "ensembl_gene_id"))
-all_genes[is.na(all_genes$CCLE_category), "CCLE_category"] <- "not_in_CCLE"
-
-# TCGA database analysis added
-
-all_genes <- all_genes %>%
-  left_join(as_tibble(rowData(TCGA_TPM), rownames = "ensembl_gene_id") %>%
-              dplyr::select(ensembl_gene_id, external_gene_name, percent_pos_tum,
-                            percent_neg_tum, max_TPM_in_TCGA, TCGA_category))
-
-all_genes[all_genes$lowly_expressed_in_GTEX == TRUE &
-            all_genes$multimapping_analysis == "testis_specific",
-          "TCGA_category"] <- "multimapping_issue"
-
-# Add DAC analysis
-induced <- as_tibble(rowData(DAC_treated_cells_multimapping), 
-                     rownames = "ensembl_gene_id") %>%
-  filter(induced == TRUE) %>%
-  pull(external_gene_name)
-
-all_genes <- all_genes %>%
-  mutate(DAC_induced = case_when(external_gene_name %in% induced ~ TRUE,
-                                 !external_gene_name %in% induced ~ FALSE))
-
-# Add the most biologically relevant transcript (canonical and coordinates)
-
-ensembl <- biomaRt::useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
-attributes_vector <- c("ensembl_gene_id",
-                       "external_gene_name",
-                       "ensembl_transcript_id",
-                       "external_transcript_name",
-                       "chromosome_name",
-                       "strand",
-                       "transcript_start",
-                       "transcript_end",
-                       "transcription_start_site",
-                       "transcript_length",
-                       "transcript_biotype",
-                       "transcript_is_canonical")
-transcripts_infos <- as_tibble(biomaRt::getBM(attributes = attributes_vector,
-                                              mart = ensembl))
-canonical_transcripts <- transcripts_infos %>%
-  filter(transcript_is_canonical == 1) %>%
-  filter(chromosome_name %in% c(1:22, "X", "Y", "MT")) %>%
-  filter(transcript_biotype == "protein_coding" | transcript_biotype == "lncRNA")
-all_genes <- all_genes %>%
-  left_join(canonical_transcripts %>%
-              dplyr::select(ensembl_gene_id,
-                            external_transcript_name, ensembl_transcript_id,
-                            chromosome_name, strand, transcript_start,
-                            transcript_end, transcription_start_site,
-                            transcript_length, transcript_biotype))
-
-```
-
-From there, we filtered based on the testis_specificity ("testis_specific" or 
-"testis_preferential"), testis_cell_type (not "Macrophage", "Endothelial",
-"Myoid", "Sertoli", "Leydig"), Higher_in_somatic_cell_type (TRUE), 
+`make_all_genes_prelim.R` and `130_make_all_genes_and_CT_genes.R` in `CTdata`).
+
+```{r all_genes, message = FALSE, warning = FALSE}
+all_genes
+```
+
+From there, we filtered based on the testis_specificity ("testis_specific", which
+is based on expression in health tissue and scRNA seq info from HPA), 
 CCLE_category ("activated") and TCGA_category ("activated" or 
-"multimapping_issue") to have our CT genes. Then the most relevant 
-transcript was validated manually (and check that there is a transcript 
-activated in testis and the same is activated in tumor).
+"multimapping_issue") to have our CT genes. Then, when wanting to validate TSS 
+manually, we realised that for some genes, reads were not properly aligned to
+exons which might reflect a poorly defined transcription in these regions and 
+are hence likely unreliable.
+
+Some genes were also characterized as Cancer-Testis preferential genes when 
+testis specificity was less stringent
 
@@
 We find `r length(CTdatabase_ours@IntersectionSets[["11"]])/nrow(CTdatabase_cleaned)*100`
-% of CTdatabase in CTexploreR, which is `r length(CTdatabase_ours@IntersectionSets[["11"]])/nrow(CT_genes)*100`
-% of our database.
+% of CTdatabase in CTexploreR, which is 
+`r length(CTdatabase_ours@IntersectionSets[["11"]])/nrow(CT_genes)*100`% of our 
+database.
 
@@
 
-# 12 Genes are lost because not in any database
-
-CTdatabase_lost[is.na(CTdatabase_lost$testis_specificity), ]$testis_specificity <- "not_specific"
+# 9 Genes are lost because not in any database
+
 table(CTdatabase_lost$testis_specificity)
 
-table(CTdatabase_lost$testis_cell_type)
-
-table(CTdatabase_lost$Higher_in_somatic_cell_type)
+table(CTdatabase_lost$CT_gene_type)
+
+table(CTdatabase_lost$not_detected_in_somatic_HPA)
 
@@
 
-```
-
-`r (nrow(CTdatabase_lost) - table(CTdatabase_lost$testis_specificity)[["testis_specific"]])/nrow(CTdatabase_lost)*100`% of 
-these genes are not testis specific.
+CTdatabase_lost_upset <- 
+  list(`Not testis specific` = 
+         filter(CTdatabase_lost,
+                testis_specificity != "testis_specific")$external_gene_name,
+       `Not tumour activated` = 
+         filter(CTdatabase_lost,
+                (TCGA_category != "activated" &
+                  TCGA_category != "multimapping_issue")|
+                   CCLE_category != "activated")$external_gene_name,
+       `CT preferential` =
+         filter(CTdatabase_lost,
+                CT_gene_type == "CTP_gene")$external_gene_name)
+
+upset_CTdatabase <- fromList(CTdatabase_lost_upset)
+upset(upset_CTdatabase)
+
+ (nrow(CTdatabase_lost) - table(CTdatabase_lost$testis_specificity)[["testis_specific"]])/nrow(CTdatabase_lost)*100
+```
+
+`r (nrow(CTdatabase_lost) - table(CTdatabase_lost$testis_specificity)[["testis_specific"]])/nrow(CTdatabase_lost)*100`
+% of these genes are not testis specific.
+
+However `r table(CTdatabase_lost$CT_gene_type)[["CTP_gene"]]` of these lost 
+genes are flagged as Cancer-Testis preferential in our analysis.
 
@@
 % are not properly activated in tumors and/or cancer cell lines.
+
+In their analysis, they had characterised gene specificity, some being not 
+available, not found in testis, testis-restricted, testis-selective and 
+testis/brain-restricted. Let's see how the lost genes qualify as they didn't
+mention those were strictly testis specific.
+
+```{r CTdatabase_classification}
+CTdatabase_cleaned %>% 
+  filter(external_gene_name %in% CTdatabase_lost$external_gene_name) %>% 
+  pull(Classification) %>% 
+  table()
+```
+
+We can see that most of them had no info or were testis-selective (I couldn't
+find on website or paper how they selected categories).
 
@@
 
-4 in all, 59 in at least 3 databases
+4 in all, 60 in at least 3 databases
 
@@
 
-Jamin_lost[is.na(Jamin_lost$testis_specificity), ]$testis_specificity <- "not_specific"
 table(Jamin_lost$testis_specificity)
 
-table(Jamin_lost$testis_cell_type)
-
-table(Jamin_lost$Higher_in_somatic_cell_type)
+table(Jamin_lost$CT_gene_type)
+
+table(Jamin_lost$not_detected_in_somatic_HPA)
 
@@
 table(Jamin_lost$CCLE_category)
+
+table(Jamin_lost$TCGA_category, Jamin_lost$CCLE_category)
+
+Jamin_lost_upset <- 
+  list(`Not testis specific` = 
+         filter(Jamin_lost,
+                testis_specificity != "testis_specific")$external_gene_name,
+       `Not tumour activated` = 
+         filter(Jamin_lost,
+                (TCGA_category != "activated" &
+                  TCGA_category != "multimapping_issue")|
+                   CCLE_category != "activated")$external_gene_name,
+       `CT preferential` =
+         filter(Jamin_lost,
+                CT_gene_type == "CTP_gene")$external_gene_name)
+
+upset_Jamin <- fromList(Jamin_lost_upset)
+upset(upset_Jamin)
 ```
@@
 
+However `r table(Jamin_lost$CT_gene_type)[["CTP_gene"]]` of these lost 
+genes are flagged as Cancer-Testis preferential in our analysis.
+
 `r (nrow(Jamin_lost) - table(Jamin_lost$TCGA_category, Jamin_lost$CCLE_category)["activated", "activated"])/nrow(Jamin_lost)*100`
@@
 
-Wang_lost[is.na(Wang_lost$testis_specificity), ]$testis_specificity <- "not_specific"
 table(Wang_lost$testis_specificity)
 
-table(Wang_lost$testis_cell_type)
-
-table(Wang_lost$Higher_in_somatic_cell_type)
+table(Wang_lost$CT_gene_type)
+
+table(Wang_lost$not_detected_in_somatic_HPA)
 
@@
 
+table(Wang_lost$TCGA_category, Wang_lost$CCLE_category)
+
+Wang_lost_upset <- 
+  list(`Not testis specific` = 
+         filter(Wang_lost,
+                testis_specificity != "testis_specific")$external_gene_name,
+       `Not tumour activated` = 
+         filter(Wang_lost,
+                (TCGA_category != "activated" &
+                  TCGA_category != "multimapping_issue")|
+                   CCLE_category != "activated")$external_gene_name,
+       `CT preferential` =
+         filter(Wang_lost,
+                CT_gene_type == "CTP_gene")$external_gene_name)
+
+upset_Wang <- fromList(Wang_lost_upset)
+upset(upset_Wang)
 ```
@@
 
+However `r table(Wang_lost$CT_gene_type)[["CTP_gene"]]` of these lost 
+genes are flagged as Cancer-Testis preferential in our analysis.
+
 `r (nrow(Wang_lost) - table(Wang_lost$TCGA_category, Wang_lost$CCLE_category)["activated", "activated"])/nrow(Wang_lost)*100`
@@
 
-Carter_lost[is.na(Carter_lost$testis_specificity), ]$testis_specificity <- "not_specific"
 table(Carter_lost$testis_specificity)
 
-table(Carter_lost$testis_cell_type)
-
-table(Carter_lost$Higher_in_somatic_cell_type)
+table(Carter_lost$CT_gene_type)
+
+table(Carter_lost$not_detected_in_somatic_HPA)
 
@@
 table(Carter_lost$CCLE_category)
+
+table(Carter_lost$TCGA_category, Carter_lost$CCLE_category)
+
+Carter_lost_upset <- 
+  list(`Not testis specific` = 
+         filter(Carter_lost,
+                testis_specificity != "testis_specific")$external_gene_name,
+       `Not tumour activated` = 
+         filter(Carter_lost,
+                (TCGA_category != "activated" &
+                  TCGA_category != "multimapping_issue")|
+                   CCLE_category != "activated")$external_gene_name,
+       `CT preferential` =
+         filter(Carter_lost,
+                CT_gene_type == "CTP_gene")$external_gene_name)
+
+upset_Carter <- fromList(Carter_lost_upset)
+upset(upset_Carter)
 ```
@@
 
+However `r table(Carter_lost$CT_gene_type)[["CTP_gene"]]` of these lost 
+genes are flagged as Cancer-Testis preferential in our analysis.
+
 `r (nrow(Carter_lost) - table(Carter_lost$TCGA_category, Carter_lost$CCLE_category)["activated", "activated"])/nrow(Carter_lost)*100`
@@
 
-Bruggeman_lost[is.na(Bruggeman_lost$testis_specificity), ]$testis_specificity <- "not_specific"
 table(Bruggeman_lost$testis_specificity)
 
-table(Bruggeman_lost$testis_cell_type)
-
-table(Bruggeman_lost$Higher_in_somatic_cell_type)
+table(Bruggeman_lost$CT_gene_type)
+
+table(Bruggeman_lost$not_detected_in_somatic_HPA)
 
@@
 table(Bruggeman_lost$CCLE_category)
+
+table(Bruggeman_lost$TCGA_category, Bruggeman_lost$CCLE_category)
+
+Bruggeman_lost_upset <- 
+  list(`Not testis specific` = 
+         filter(Bruggeman_lost,
+                testis_specificity != "testis_specific")$external_gene_name,
+       `Not tumour activated` = 
+         filter(Bruggeman_lost,
+                (TCGA_category != "activated" &
+                  TCGA_category != "multimapping_issue")|
+                   CCLE_category != "activated")$external_gene_name,
+       `CT preferential` =
+         filter(Bruggeman_lost,
+                CT_gene_type == "CTP_gene")$external_gene_name)
+
+upset_Bruggeman <- fromList(Bruggeman_lost_upset)
+upset(upset_Bruggeman)
 ```
@@
 these genes are not testis specific.
+
+However `r table(Bruggeman_lost$CT_gene_type)[["CTP_gene"]]` of these lost 
+genes are flagged as Cancer-Testis preferential in our analysis.
 
@@
 
-not_specific <- filter(lost, is.na(testis_specificity))
+not_specific <- filter(lost, testis_specificity == "not_testis_specific")
 
@@
 
-somatic_testis <- filter(lost, testis_cell_type %in% c( "Macrophage", 
-                                                        "Endothelial", "Myoid",
-                                                        "Sertoli", "Leydig"))
+somatic_testis <- filter(lost, not_detected_in_somatic_HPA == FALSE)
 
 testis_expression(somatic_testis$external_gene_name, cells = "all")
-
-somatic_tissue <- filter(lost, Higher_in_somatic_cell_type == TRUE)
-
-HPA_cell_type_expression(somatic_tissue$external_gene_name)
+HPA_cell_type_expression(somatic_testis$external_gene_name)
 
@@
 
+`r length(common)` genes in our CTexploreR database are found in at least one of
+the other database, which represents `r length(common)/dim(CT_genes)[1] * 100`%.
+
 We have lost `r length(lost_list)` genes in total. Among them, 
@@
 specific, `r dim(somatic_testis)[1]/length(lost_list)*100`% are expressed in 
-testis somatic cells, `r dim(somatic_tissue)[1]/length(lost_list)*100`% are 
-expressed in somatic cells, `r dim(not_TCGA_activated)[1]/length(lost_list)*100`% 
+somatic cells, `r dim(not_TCGA_activated)[1]/length(lost_list)*100`% 
 are not activated in TCGA samples, 
@@
 table(new$testis_specificity)
+table(new$X_linked)
+table(new$regulated_by_methylation)
 table(new$X_linked, new$regulated_by_methylation)
@@
 ```
-There are `r dim(new)[1]` new CT genes in CTexploreR. These are mainly testis
-specific and on autosomes. Regulation by methylation is not the majority of 
-them. This makes sens as other databases have been focused on regulation by 
-methylation, so as here we don't segregate with that we can find these. There 
+
+There are `r dim(new)[1]` new CT genes in CTexploreR. These are all testis
+specific and mainly on autosomes. Regulation by methylation is the majority of 
+them.  There 
 is only 11 new "major" CT that are on the X chromosome and regulated by 
@@
 genes <- CT_genes$external_gene_name
-database <- CT_mean_methylation_in_tissues[rownames(CT_mean_methylation_in_tissues) %in% genes]
+database <- mean_methylation_in_tissues[rownames(mean_methylation_in_tissues) %in% genes]
 
@@
 ```
-
-We used these data to determine in which testis cell type
-each gene is mostly expressed.
-
-```{r main_cell_type_expression_barplot}
-CT_genes$main_cell_type_expression <- factor(CT_genes$testis_cell_type,
-                                                levels = c("SSC", 
-                                               "Spermatogonia", 
-                                               "Early_spermatocyte",
-                                               "Late_spermatocyte",
-                                               "Round_spermatid",
-                                               "Elongated_spermatid",
-                                               "Sperm1",
-                                               "Sperm2",
-                                               "Sertoli"))
-
-CT_genes %>% 
-  mutate(regulated_by_methylation = ifelse(regulated_by_methylation,
-                                           "Regulated by methylation",
-                                           "Not regulated by methylation")) %>%
-  filter(!is.na(testis_cell_type)) %>%
-  filter(main_cell_type_expression != "Sertoli") %>% 
-  ggplot(aes(x = chr, fill = main_cell_type_expression,
-             color = main_cell_type_expression)) +
-  scale_fill_manual(values = c("lightyellow2", "moccasin", "gold", "orange","red2",
-                               "darkred", "violet", "purple")) +
-  scale_color_manual(values = c("lightyellow3", "navajowhite2", "gold", "orange","red2",
-                               "darkred", "violet", "purple")) +
-  geom_bar(stat = 'count') +
-  facet_grid(main_cell_type_expression ~ regulated_by_methylation) +
-  xlab("Chromosome") +
-  ylab("Number of genes") +
-  theme_bw() +
-  theme(
-    axis.text.x = element_text(color = "black", angle = 90, vjust = 0.5, size = 10),
-    axis.title = element_text(size = 10, color = "gray10"),
-    strip.text.y = element_blank())
-
-```
-
-```{r escapees}
-CT_genes %>% 
-  filter(X_linked) %>% 
-  filter(testis_cell_type %in% c("Late_spermatocyte","Round_spermatid", 
-                                 "Elongated_spermatid", "Sperm1", "Sperm2"))
-
-```
-
-These genes are on the X chromosome but escape X chromosome inactivation and are
-activates post-meioticly.
 

### 2024-05-14 Paper_code-CT_analysis.Rmd
Reorganisation of paper code (in prep of adding Axelle code) and fig in pdf

--- 2024-05-10 Paper_code-Database_CT.Rmd
@@
 library(CTexploreR)
-library(Vennerable)
-library(biomaRt)
 library(tidyverse)
 library(SummarizedExperiment)
-library(UpSetR)
 library(ComplexHeatmap)
@@
 library(SingleCellExperiment)
+library(msigdbr)
+library(DOSE)
 library(org.Hs.eg.db)
 library(clusterProfiler)
-library(msigdbr)
-library(DOSE)
+library(patchwork)
 library(BiocParallel)
-library(patchwork)
 library(Biostrings)
-```
-
-Gene names/synonyms required for databases cleaning
-
-```{r ensembl_synonym, warning=FALSE, message=FALSE}
-ensembl <- biomaRt::useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
-attributes_vector <- c("ensembl_gene_id", "external_gene_name",
-                       "external_synonym", "gene_biotype",
-                       "chromosome_name", "band", "start_position", "end_position",
-                       "strand")
-ensembl_gene_synonym <- as_tibble(getBM(attributes = attributes_vector, mart = ensembl))
-
-ensembl_gene_synonym <- ensembl_gene_synonym %>%
-  mutate(external_synonym = sub(pattern = "ORF", external_synonym, 
-                                replacement = "orf"))
-
-attributes_vector <- c("ensembl_gene_id", "external_gene_name")
-ensembl_gene_names <- as_tibble(getBM(attributes = attributes_vector, mart = ensembl))
-
-attributes_vector <- c("external_gene_name",
-                       "external_synonym")
-gene_synonym <- as_tibble(getBM(attributes = attributes_vector, mart = ensembl))
 ```
@@
 
+load("../CTdata/eh_data/all_genes.rda")
 load("../CTdata/eh_data/CCLE_data.rda")
@@
 load("../CTdata/eh_data/mean_methylation_in_tissues.rda")
-
-load("../CTdata/eh_data/all_genes.rda")
-load("../CTdata/eh_data/HPA_cell_type_specificities.rda")
 load("../CTdata/eh_data/CT_genes.rda")
@@
 ```
-
-# Database comparison
-
-## Lists cleaning
-
-CT lists from other databases have been checked (using GTEx and our 
-`GTEx_expression()` funtion and GeneCards) in order to remove duplicated
-gene names or deprecated ones and allow comparison between databases.
-
-### CTdatabase
-
-Online list copied in a csv file, several lists exist so we combined them.
-
-We checked gene names that were a concatenation of two genes (choice using
-biomaRt synonyms to get the official one), checked which ones had the right names,
-removed duplicated genes, verified lost genes and added back those that should 
-be there.
-
-```{r CTdatabase_cleaning, message = FALSE}
-CTdatabase <- read_delim("data/CTdatabase1.csv", delim = ";", 
-                         escape_double = FALSE, trim_ws = TRUE)
-colnames(CTdatabase) <- c("Family", "Gene_Name", "Chromosomal_localization",
-                          "CT_identifier")
-CTdatabase_bis <- read_csv2("data/CTdatabase2.csv")
-CTdatabase <- left_join(CTdatabase, CTdatabase_bis, 
-                        by = c("Gene_Name" = "Gene_Symbol"))
-
-CTdatabase_single <- CTdatabase %>%
-  mutate(Gene_Name = sub(pattern = "/.*$", Gene_Name, replacement = ""))
-CTdatabase_single <- CTdatabase_single %>%
-  mutate(Gene_Name = sub(pattern = ",.*$", Gene_Name, replacement = ""))
-
-CTdatabase_official_names <- 
-  unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, 
-                       external_gene_name)) %>%
-  filter(external_gene_name %in% CTdatabase_single$Gene_Name) %>%
-  mutate(Gene_Name = external_gene_name) %>%
-  mutate(external_synonym = NA)
-CTdatabase_synonym <- 
-  ensembl_gene_synonym %>%
-  filter(external_synonym %in% CTdatabase_single$Gene_Name) %>%
-  mutate(Gene_Name = external_synonym) %>%
-  dplyr::select(ensembl_gene_id, external_gene_name, Gene_Name, external_synonym)
-CTdatabase_cleaned <- 
-  rbind(CTdatabase_official_names, CTdatabase_synonym) %>% 
-  left_join(CTdatabase_single)
-
-duplicated_genes <- CTdatabase_cleaned$Gene_Name[duplicated(CTdatabase_cleaned$Gene_Name)]
-bad_ids <- ensembl_gene_synonym %>%
-  filter(external_gene_name %in% duplicated_genes | external_synonym %in% duplicated_genes) %>%
-  filter(chromosome_name %in% grep(pattern = "H", x = chromosome_name, value = TRUE)) %>%
-  pull(ensembl_gene_id)
-CTdatabase_cleaned <- CTdatabase_cleaned %>%
-  dplyr::filter(!ensembl_gene_id %in% bad_ids)
-CTdatabase_cleaned <- CTdatabase_cleaned %>%
-  filter(!ensembl_gene_id == "ENSG00000052126")
-CTdatabase_cleaned <- CTdatabase_cleaned %>% 
-  filter(!(ensembl_gene_id == "ENSG00000183305" & Gene_Name == "MAGEA2"))
-CTdatabase_cleaned <- CTdatabase_cleaned %>% 
-  filter(!ensembl_gene_id == "ENSG00000204648")
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "CSAG3B")
-CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "CSAG2", "external_synonym"] <- "CSAG3B"
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "CT45A4")
-CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "CT45A3", "external_synonym"] <- "CT45A4"
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "LAGE-1b")
-CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "CTAG2", "external_synonym"] <- "LAGE-1b"
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "CT16.2")
-CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "PAGE5", "external_synonym"] <- "CT16.2"
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "SPANXB2")
-CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "SPANXB1", "external_synonym"] <- "SPANXB2"
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "SPANXE")
-CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "SPANXD", "external_synonym"] <- "SPANXE"
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "XAGE1C")
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "XAGE1D")
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "XAGE1E")
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "XAGE2B")
-CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "XAGE2", "external_synonym"] <- "XAGE2B"
-CTdatabase_cleaned <- filter(CTdatabase_cleaned, Gene_Name != "CTAGE-2")
-CTdatabase_cleaned[CTdatabase_cleaned$Gene_Name == "CTAGE1", "external_synonym"] <- "CTAGE-2"
-
-CTdatabase_cleaned <- ensembl_gene_synonym %>%
-  mutate(Gene_Name = external_synonym) %>%
-  filter(external_synonym == "CXorf61") %>%
-  dplyr::select(ensembl_gene_id, external_gene_name, Gene_Name, external_synonym) %>%
-  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "Cxorf61", 
-                    c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
-  rbind(CTdatabase_cleaned)
-CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name)) %>%
-  filter(external_gene_name == "CCNA1") %>%
-  mutate(Gene_Name = external_gene_name) %>%
-  mutate(external_synonym = NA) %>% 
-  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "cyclin A1", 
-                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
-  rbind(CTdatabase_cleaned)
-CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
-  filter(external_gene_name == "GOLGA6L2") %>%
-  filter(ensembl_gene_id == "ENSG00000174450") %>%
-  mutate(Gene_Name = external_gene_name) %>%
-  mutate(external_synonym = NA) %>% 
-  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "GOLGAGL2 FA", 
-                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
-  rbind(CTdatabase_cleaned)
-CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
-  filter(external_gene_name == "LYPD6B") %>%
-  mutate(Gene_Name = external_gene_name) %>%
-  mutate(external_synonym = NA) %>% 
-  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "LOC130576", 
-                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
-  rbind(CTdatabase_cleaned)
-CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
-  filter(external_gene_name == "CT62") %>%
-  mutate(Gene_Name = external_gene_name) %>%
-  mutate(external_synonym = NA) %>% 
-  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "LOC196993", 
-                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
-  rbind(CTdatabase_cleaned)
-CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
-  filter(external_gene_name == "CT75") %>%
-  filter(ensembl_gene_id == "ENSG00000291155") %>%
-  mutate(Gene_Name = external_gene_name) %>%
-  mutate(external_synonym = NA) %>% 
-  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "LOC440934", 
-                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
-  rbind(CTdatabase_cleaned)
-CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
-  filter(external_gene_name == "LINC01192") %>%
-  mutate(Gene_Name = external_gene_name) %>%
-  mutate(external_synonym = NA) %>% 
-  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "LOC647107", 
-                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
-  rbind(CTdatabase_cleaned)
-CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
-  filter(external_gene_name == "TSPY1") %>%
-  mutate(Gene_Name = external_gene_name) %>%
-  mutate(external_synonym = NA) %>% 
-  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "LOC728137", 
-                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
-  rbind(CTdatabase_cleaned)
-CTdatabase_cleaned <- unique(dplyr::select(ensembl_gene_synonym, ensembl_gene_id, external_gene_name))%>%
-  filter(external_gene_name == "SSX2B") %>%
-  mutate(Gene_Name = external_gene_name) %>%
-  mutate(external_synonym = NA) %>% 
-  cbind(CTdatabase_single[CTdatabase_single$Gene_Name == "SSX2b", 
-                          c("Family", "Chromosomal_localization", "CT_identifier", "Classification")]) %>% 
-  rbind(CTdatabase_cleaned)
-```
-
-### Jamin's list
-
-Excel file coming from supplemental data.
-
-```{r Jamin_clean, message = FALSE}
-Jamin_core_CT <- read_excel("data/Jamin_core_CT.xlsx")
-Jamin_core_CT[Jamin_core_CT$Gene == "KIAA1211", "Gene"] <- "CRACD"
-Jamin_core_CT[Jamin_core_CT$Gene == "CXorf67", "Gene"] <- "EZHIP"
-```
-
-### Wang's CTatlas
-
-Excel file coming from supplemental data.
-
-```{r Wang_cleaning, message = FALSE}
-Wang_CT <- read_excel("data/Wang_Suppl_Data_3.xlsx", 
-    sheet = "Supplementary Data 3B", skip = 1)
-colnames(Wang_CT)[1] <- "ensembl_gene_id"
-
-Wang_CT <- ensembl_gene_names %>% 
-  filter(ensembl_gene_id %in% Wang_CT$ensembl_gene_id) %>%
-  right_join(Wang_CT)
-
-Wang_CT[Wang_CT$ensembl_gene_id == "ENSG00000181013", "external_gene_name"] <- "C17orf47"
-Wang_CT[Wang_CT$ensembl_gene_id == "ENSG00000204293", "external_gene_name"] <- "OR8B2"
-Wang_CT[Wang_CT$external_gene_name == "", "external_gene_name"] <- "RNASE11"
-Wang_CT[Wang_CT$external_gene_name == "CHCT1", "external_gene_name"] <- "C17orf64"
-Wang_CT[Wang_CT$external_gene_name == "PRSS40A", "external_gene_name"] <- "TISP43"
-Wang_CT[Wang_CT$external_gene_name == "TEX56P", "external_gene_name"] <- "C6orf201"
-Wang_CT[Wang_CT$external_gene_name == "SLC25A51P4", "external_gene_name"] <- "RP11-113D6.10"
-Wang_CT[Wang_CT$external_gene_name == "TCP10L3", "external_gene_name"] <- "TCP10"
-Wang_CT[Wang_CT$external_gene_name == "SCAND3", "external_gene_name"] <- "ZBED9"
-```
-
-### Carter's list
-
-```{r Carter_clean, message = FALSE}
-Carter_CT_list <- read_table("data/Carter_CT_list.txt", skip = 1)
-Carter_CT <- Carter_CT_list[Carter_CT_list$CT_Expression,]
-
-Carter_CT[Carter_CT$Gene == "ENSG00000261649", "Gene_Name"] <- "GOLGA6L7"
-Carter_CT[Carter_CT$Gene == "ENSG00000239620", "Gene_Name"] <- "PRR20G"
-Carter_CT[Carter_CT$Gene == "ENSG00000168148", "Gene_Name"] <- "H3-4"
-Carter_CT[Carter_CT$Gene == "ENSG00000204296", "Gene_Name"] <- "TSBP1"
-Carter_CT[Carter_CT$Gene == "ENSG00000180219", "Gene_Name"] <- "GARIN6"
-Carter_CT[Carter_CT$Gene == "ENSG00000172717", "Gene_Name"] <- "GARIN2"
-Carter_CT[Carter_CT$Gene == "ENSG00000174015", "Gene_Name"] <- "CBY2"
-Carter_CT[Carter_CT$Gene == "ENSG00000224960", "Gene_Name"] <- "PPP4R3C"
-```
-
-### Burggeman's list
-
-Excel file from supplemental data.
-
-```{r Bruggeman_clean, message=FALSE}
-Bruggeman_data <- read_excel("data/Bruggeman_suppl_data.xlsx", skip = 1,
-                           sheet = "1D")
-
-Bruggeman_official_names <- gene_synonym %>% 
-  dplyr::select(external_gene_name) %>% 
-  unique() %>% 
-  filter(external_gene_name %in% Bruggeman_data$Gene) %>%
-  mutate(Gene_Name = external_gene_name) %>%
-  mutate(external_synonym = NA)
-
-Bruggeman_synonym <- gene_synonym %>%
-  filter(external_synonym %in% Bruggeman_data$Gene) %>%
-  mutate(Gene_Name = external_synonym) %>%
-  dplyr::select(external_gene_name, Gene_Name, external_synonym)
-
-Bruggeman_synonym <- Bruggeman_synonym[-which(Bruggeman_synonym$Gene_Name %in% 
-                          Bruggeman_official_names$Gene_Name),]
-
-Bruggeman_CT <- rbind(Bruggeman_official_names, Bruggeman_synonym)
-
-lost <- Bruggeman_data[which(!Bruggeman_data$Gene %in% c(Bruggeman_CT$Gene_Name)), "Gene"]
-colnames(lost) <- "external_gene_name"
-lost$Gene_Name <- rep(NA, nrow(lost))
-lost$external_synonym <- rep(NA, nrow(lost))
-
-lost[lost$external_gene_name == "C21orf59", "Gene_Name"] <- "CFAP298"
-lost[lost$external_gene_name == "C11orf57", "Gene_Name"] <- "NKAPD1"
-lost[lost$external_gene_name == "C7orf55", "Gene_Name"] <- "FMC1"
-lost[lost$external_gene_name == "C10orf12", "Gene_Name"] <- "LCOR"
-lost[lost$external_gene_name == "RPL19P12", "Gene_Name"] <- "RPL19P12"
-lost[lost$external_gene_name == "C16orf59", "Gene_Name"] <- "TEDC2"
-lost[lost$external_gene_name == "TTTY15", "Gene_Name"] <- "USP9Y"
-lost[lost$external_gene_name == "C17orf53", "Gene_Name"] <- "HROB"
-lost[lost$external_gene_name == "C1orf112", "Gene_Name"] <- "FIRRM"
-lost[lost$external_gene_name == "C12orf66", "Gene_Name"] <- "KICS2"
-lost[lost$external_gene_name == "C9orf84", "Gene_Name"] <- "SHOC1"
-lost[lost$external_gene_name == "C10orf25", "Gene_Name"] <- "ZNF22-AS1"
-lost[lost$external_gene_name == "C20orf197", "Gene_Name"] <- "LINC02910"
-lost[lost$external_gene_name == "C3orf67", "Gene_Name"] <- "CFAP20DC"
-lost[lost$external_gene_name == "C8orf37", "Gene_Name"] <- "CFAP418"
-lost[lost$external_gene_name == "C22orf34", "Gene_Name"] <- "MIR3667HG"
-
-Bruggeman_CT <- rbind(Bruggeman_CT, lost)
-
-missing_Bruggeman <- c("BMS1P4", "ADAM6", "ANXA2P3", "ARHGAP11B", "DPY19L2P2", 
-                       "HLA-L", "PA2G4P4", "PIPSL", "PRKY", "YBX3P1", 
-                       "RPL23AP53", "UQCRBP1", "RPL23P8", "MRS2P2", "PIN4P1", 
-                       "SLC6A10P", "GUSBP2", "PPIEL", "LRRC37BP1", "MSL3P1", 
-                       "PLEKHA8P1", "STAG3L1", "TCAM1P", "ZNF702P", "ZNF815P", 
-                       "ATP6AP1L", "RPL21P44", "SEC14L1P1", "ZNF876P", 
-                       "RPLP0P2", "FAM86JP", "FAM175A", "LACE1", "ATP5EP2", 
-                       "WDR92", "TCTE3", "METTL20", "KIAA2022", "ZNRD1-AS1", 
-                       "SGOL1", "FAM35DP", "MTL5", "TMEM14E", "MLLT4-AS1", 
-                       "CCDC173", "KIAA1524", "WDR78", "LINC00476", "LYRM5", 
-                       "HILS1", "CASC5", "KIAA1919", "CTAGE5", "FAM188B", 
-                       "TMEM194B", "FAM122C", "PPP1R2P3", "KIAA0391", "SGOL2", 
-                       "FAM19A3", "ZNF788", "RPL19P12", "FIRRM")
-
-external_names_to_keep <- gene_synonym %>%
-   filter(external_synonym %in% missing_Bruggeman) %>%
-   filter(!external_gene_name %in% c("ATP5F1EP2", "POLR1HASP", "SHLD2P3", 
-                                   "TMEM14EP", "H1-9P", "ZNF788P")) %>% 
-   mutate(Gene_Name = external_gene_name)
-
-Bruggeman_CT[Bruggeman_CT$external_synonym %in% 
-                    external_names_to_keep$external_synonym, 
-                  "Gene_Name"] <- external_names_to_keep$Gene_Name
-
-Bruggeman_CT <- Bruggeman_CT %>% 
-  dplyr::select(Gene_Name)
-```
-
-## CTexploreR data for selection pipeline
-
-To characterise the differences between our database and other, we need the 
-category we created in CTexploreR. For this, we have the object `all_genes` in 
-CTdata that contains the CT analysis for all genes. More info in
-
-Hereunder is what we used for our selection pipeline (coming from 
-`make_all_genes_prelim.R` and `130_make_all_genes_and_CT_genes.R` in `CTdata`).
-
-```{r all_genes, message = FALSE, warning = FALSE}
-all_genes
-```
-
-From there, we filtered based on the testis_specificity ("testis_specific", which
-is based on expression in health tissue and scRNA seq info from HPA), 
-CCLE_category ("activated") and TCGA_category ("activated" or 
-"multimapping_issue") to have our CT genes. Then, when wanting to validate TSS 
-manually, we realised that for some genes, reads were not properly aligned to
-exons which might reflect a poorly defined transcription in these regions and 
-are hence likely unreliable.
-
-Some genes were also characterized as Cancer-Testis preferential genes when 
-testis specificity was less stringent
-
-## CTexploreR VS CTdatabase
-
-```{r CTdatabase_ours}
-CTdatabase_ours <- Venn(list(CTdatabase = CTdatabase_cleaned$external_gene_name,
-                             CTexploreR = CT_genes$external_gene_name))
-gp <- VennThemes(compute.Venn(CTdatabase_ours))
-gp[["Face"]][["11"]]$fill <-  "mistyrose"
-gp[["Face"]][["01"]]$fill <-  "darkseagreen1"
-gp[["Face"]][["10"]]$fill <-  "lightsteelblue1"
-gp[["Set"]][["Set1"]]$col <-  "paleturquoise4"
-gp[["Set"]][["Set2"]]$col <-  "darkseagreen4"
-gp[["SetText"]][["Set1"]]$col <-  "paleturquoise4"
-gp[["SetText"]][["Set2"]]$col <-  "darkseagreen4"
-plot(CTdatabase_ours, gp = gp)
-
-```
-
-We find `r length(CTdatabase_ours@IntersectionSets[["11"]])/nrow(CTdatabase_cleaned)*100`
-% of CTdatabase in CTexploreR, which is 
-`r length(CTdatabase_ours@IntersectionSets[["11"]])/nrow(CT_genes)*100`% of our 
-database.
-
-Lost genes analysis
-
-```{r CTdatabase_lost}
-CTdatabase_lost <- all_genes %>%
-  filter(external_gene_name %in% CTdatabase_ours@IntersectionSets[["10"]])
-
-# 9 Genes are lost because not in any database
-
-table(CTdatabase_lost$testis_specificity)
-
-table(CTdatabase_lost$CT_gene_type)
-
-table(CTdatabase_lost$not_detected_in_somatic_HPA)
-
-table(CTdatabase_lost$TCGA_category)
-
-table(CTdatabase_lost$CCLE_category)
-
-table(CTdatabase_lost$TCGA_category, CTdatabase_lost$CCLE_category)
-
-CTdatabase_lost_upset <- 
-  list(`Not testis specific` = 
-         filter(CTdatabase_lost,
-                testis_specificity != "testis_specific")$external_gene_name,
-       `Not tumour activated` = 
-         filter(CTdatabase_lost,
-                (TCGA_category != "activated" &
-                  TCGA_category != "multimapping_issue")|
-                   CCLE_category != "activated")$external_gene_name,
-       `CT preferential` =
-         filter(CTdatabase_lost,
-                CT_gene_type == "CTP_gene")$external_gene_name)
-
-upset_CTdatabase <- fromList(CTdatabase_lost_upset)
-upset(upset_CTdatabase)
-
- (nrow(CTdatabase_lost) - table(CTdatabase_lost$testis_specificity)[["testis_specific"]])/nrow(CTdatabase_lost)*100
-```
-
-`r (nrow(CTdatabase_lost) - table(CTdatabase_lost$testis_specificity)[["testis_specific"]])/nrow(CTdatabase_lost)*100`
-% of these genes are not testis specific.
-
-However `r table(CTdatabase_lost$CT_gene_type)[["CTP_gene"]]` of these lost 
-genes are flagged as Cancer-Testis preferential in our analysis.
-
-`r (nrow(CTdatabase_lost) - table(CTdatabase_lost$TCGA_category, CTdatabase_lost$CCLE_category)["activated", "activated"])/nrow(CTdatabase_lost)*100`
-% are not properly activated in tumors and/or cancer cell lines.
-
-In their analysis, they had characterised gene specificity, some being not 
-available, not found in testis, testis-restricted, testis-selective and 
-testis/brain-restricted. Let's see how the lost genes qualify as they didn't
-mention those were strictly testis specific.
-
-```{r CTdatabase_classification}
-CTdatabase_cleaned %>% 
-  filter(external_gene_name %in% CTdatabase_lost$external_gene_name) %>% 
-  pull(Classification) %>% 
-  table()
-```
-
-We can see that most of them had no info or were testis-selective (I couldn't
-find on website or paper how they selected categories).
-
-## CTexploreR VS omics databases
-
-```{r upset_plot}
-core_ours <- Venn(list(Jamin = Jamin_core_CT$Gene, 
-                       CTexploreR = CT_genes$external_gene_name))
-
-Wang_ours <- Venn(list(Wang = Wang_CT$external_gene_name, 
-                       CTexploreR = CT_genes$external_gene_name))
-
-Carter_ours <- Venn(list(Carter_CT = Carter_CT$Gene_Name, 
-                         CTexploreR = CT_genes$external_gene_name))
-
-Bruggeman_ours <- Venn(list(Bruggeman = Bruggeman_CT$Gene_Name,
-                            CTexploreR = CT_genes$external_gene_name))
-
-gene_list <- list(CTexploreR = CT_genes$external_gene_name,
-                  Carter = Carter_CT$Gene_Name,
-                  Jamin = Jamin_core_CT$Gene, 
-                  CTatlas = Wang_CT$external_gene_name,
-                  Bruggeman = Bruggeman_CT$Gene_Name)
-
-upset_omics <- fromList(gene_list)
-upset(upset_omics)
-```
-
-4 in all, 60 in at least 3 databases
-
-Lost genes analysis
-
-```{r Jamin_lost}
-plot(core_ours, gp = gp)
-
-Jamin_lost <- all_genes %>%
-  filter(external_gene_name %in% core_ours@IntersectionSets[["10"]])
-
-table(Jamin_lost$testis_specificity)
-
-table(Jamin_lost$CT_gene_type)
-
-table(Jamin_lost$not_detected_in_somatic_HPA)
-
-table(Jamin_lost$TCGA_category)
-
-table(Jamin_lost$CCLE_category)
-
-table(Jamin_lost$TCGA_category, Jamin_lost$CCLE_category)
-
-Jamin_lost_upset <- 
-  list(`Not testis specific` = 
-         filter(Jamin_lost,
-                testis_specificity != "testis_specific")$external_gene_name,
-       `Not tumour activated` = 
-         filter(Jamin_lost,
-                (TCGA_category != "activated" &
-                  TCGA_category != "multimapping_issue")|
-                   CCLE_category != "activated")$external_gene_name,
-       `CT preferential` =
-         filter(Jamin_lost,
-                CT_gene_type == "CTP_gene")$external_gene_name)
-
-upset_Jamin <- fromList(Jamin_lost_upset)
-upset(upset_Jamin)
-```
-
-We find `r length(core_ours@IntersectionSets[["11"]])/nrow(Jamin_core_CT)*100`
-% of CTdatabase in CTexploreR, which is `r length(core_ours@IntersectionSets[["11"]])/nrow(CT_genes)*100`
-% of our database.
-
-`r (nrow(Jamin_lost) - table(Jamin_lost$testis_specificity)[["testis_specific"]])/nrow(Jamin_lost)*100`% of 
-these genes are not testis specific.
-
-However `r table(Jamin_lost$CT_gene_type)[["CTP_gene"]]` of these lost 
-genes are flagged as Cancer-Testis preferential in our analysis.
-
-`r (nrow(Jamin_lost) - table(Jamin_lost$TCGA_category, Jamin_lost$CCLE_category)["activated", "activated"])/nrow(Jamin_lost)*100`
-% are not properly activated in tumors and/or cancer cell lines.
-
-```{r Wang_lost}
-plot(Wang_ours, gp = gp)
-
-Wang_lost <- all_genes %>%
-  filter(external_gene_name %in% Wang_ours@IntersectionSets[["10"]])
-
-table(Wang_lost$testis_specificity)
-
-table(Wang_lost$CT_gene_type)
-
-table(Wang_lost$not_detected_in_somatic_HPA)
-
-table(Wang_lost$TCGA_category)
-
-table(Wang_lost$CCLE_category)
-
-table(Wang_lost$TCGA_category, Wang_lost$CCLE_category)
-
-Wang_lost_upset <- 
-  list(`Not testis specific` = 
-         filter(Wang_lost,
-                testis_specificity != "testis_specific")$external_gene_name,
-       `Not tumour activated` = 
-         filter(Wang_lost,
-                (TCGA_category != "activated" &
-                  TCGA_category != "multimapping_issue")|
-                   CCLE_category != "activated")$external_gene_name,
-       `CT preferential` =
-         filter(Wang_lost,
-                CT_gene_type == "CTP_gene")$external_gene_name)
-
-upset_Wang <- fromList(Wang_lost_upset)
-upset(upset_Wang)
-```
-
-We find `r length(Wang_ours@IntersectionSets[["11"]])/nrow(Wang_CT)*100`
-% of CTdatabase in CTexploreR, which is `r length(Wang_ours@IntersectionSets[["11"]])/nrow(CT_genes)*100`
-% of our database.
-
-`r (nrow(Wang_lost) - table(Wang_lost$testis_specificity)[["testis_specific"]])/nrow(Wang_lost)*100`% of 
-these genes are not testis specific.
-
-However `r table(Wang_lost$CT_gene_type)[["CTP_gene"]]` of these lost 
-genes are flagged as Cancer-Testis preferential in our analysis.
-
-`r (nrow(Wang_lost) - table(Wang_lost$TCGA_category, Wang_lost$CCLE_category)["activated", "activated"])/nrow(Wang_lost)*100`
-% are not properly activated in tumors and/or cancer cell lines.
-
-```{r Carter_lost}
-plot(Carter_ours, gp = gp)
-
-Carter_lost <- all_genes %>%
-  filter(external_gene_name %in% Carter_ours@IntersectionSets[["10"]])
-
-table(Carter_lost$testis_specificity)
-
-table(Carter_lost$CT_gene_type)
-
-table(Carter_lost$not_detected_in_somatic_HPA)
-
-table(Carter_lost$TCGA_category)
-
-table(Carter_lost$CCLE_category)
-
-table(Carter_lost$TCGA_category, Carter_lost$CCLE_category)
-
-Carter_lost_upset <- 
-  list(`Not testis specific` = 
-         filter(Carter_lost,
-                testis_specificity != "testis_specific")$external_gene_name,
-       `Not tumour activated` = 
-         filter(Carter_lost,
-                (TCGA_category != "activated" &
-                  TCGA_category != "multimapping_issue")|
-                   CCLE_category != "activated")$external_gene_name,
-       `CT preferential` =
-         filter(Carter_lost,
-                CT_gene_type == "CTP_gene")$external_gene_name)
-
-upset_Carter <- fromList(Carter_lost_upset)
-upset(upset_Carter)
-```
-
-We find `r length(Carter_ours@IntersectionSets[["11"]])/nrow(Carter_CT)*100`
-% of CTdatabase in CTexploreR, which is `r length(Carter_ours@IntersectionSets[["11"]])/nrow(CT_genes)*100`
-% of our database.
-
-`r (nrow(Carter_lost) - table(Carter_lost$testis_specificity)[["testis_specific"]])/nrow(Carter_lost)*100`% of 
-these genes are not testis specific.
-
-However `r table(Carter_lost$CT_gene_type)[["CTP_gene"]]` of these lost 
-genes are flagged as Cancer-Testis preferential in our analysis.
-
-`r (nrow(Carter_lost) - table(Carter_lost$TCGA_category, Carter_lost$CCLE_category)["activated", "activated"])/nrow(Carter_lost)*100`
-% are not properly activated in tumors and/or cancer cell lines.
-
-```{r Bruggeman_lost}
-plot(Bruggeman_ours, gp = gp)
-
-Bruggeman_lost <- all_genes %>%
-  filter(external_gene_name %in% Bruggeman_ours@IntersectionSets[["10"]])
-
-table(Bruggeman_lost$testis_specificity)
-
-table(Bruggeman_lost$CT_gene_type)
-
-table(Bruggeman_lost$not_detected_in_somatic_HPA)
-
-table(Bruggeman_lost$TCGA_category)
-
-table(Bruggeman_lost$CCLE_category)
-
-table(Bruggeman_lost$TCGA_category, Bruggeman_lost$CCLE_category)
-
-Bruggeman_lost_upset <- 
-  list(`Not testis specific` = 
-         filter(Bruggeman_lost,
-                testis_specificity != "testis_specific")$external_gene_name,
-       `Not tumour activated` = 
-         filter(Bruggeman_lost,
-                (TCGA_category != "activated" &
-                  TCGA_category != "multimapping_issue")|
-                   CCLE_category != "activated")$external_gene_name,
-       `CT preferential` =
-         filter(Bruggeman_lost,
-                CT_gene_type == "CTP_gene")$external_gene_name)
-
-upset_Bruggeman <- fromList(Bruggeman_lost_upset)
-upset(upset_Bruggeman)
-```
-
-We find `r length(Bruggeman_ours@IntersectionSets[["11"]])/nrow(Bruggeman_CT)*100`
-% of CTdatabase in CTexploreR, which is `r length(Bruggeman_ours@IntersectionSets[["11"]])/nrow(CT_genes)*100`
-% of our database.
-
-`r (nrow(Bruggeman_lost) - table(Bruggeman_lost$testis_specificity)[["testis_specific"]])/nrow(Bruggeman_lost)*100`% of 
-these genes are not testis specific.
-
-However `r table(Bruggeman_lost$CT_gene_type)[["CTP_gene"]]` of these lost 
-genes are flagged as Cancer-Testis preferential in our analysis.
-
-`r (nrow(Bruggeman_lost) - table(Bruggeman_lost$TCGA_category, Bruggeman_lost$CCLE_category)["activated", "activated"])/nrow(Bruggeman_lost)*100`
-% are not properly activated in tumors and/or cancer cell lines.
-
-## Characterisation of differences with all databases
-
-```{r all_databases}
-common <- unique(c(core_ours@IntersectionSets[["11"]], 
-                   CTdatabase_ours@IntersectionSets[["11"]], 
-                   Wang_ours@IntersectionSets[["11"]], 
-                   Carter_ours@IntersectionSets[["11"]],
-                   Bruggeman_ours@IntersectionSets[["11"]]))
-
-length(common)
-
-length(common)/dim(CT_genes)[1] * 100
-
-lost_list <- unique(c(core_ours@IntersectionSets[["10"]],
-                      CTdatabase_ours@IntersectionSets[["10"]],
-                      Wang_ours@IntersectionSets[["10"]],
-                      Carter_ours@IntersectionSets[["10"]],
-                      Bruggeman_ours@IntersectionSets[["10"]]))
-
-lost <- all_genes %>%
-  filter(external_gene_name %in% lost_list)
-
-not_specific <- filter(lost, testis_specificity == "not_testis_specific")
-
-GTEX_expression(not_specific$external_gene_name, units = "log_TPM")
-
-somatic_testis <- filter(lost, not_detected_in_somatic_HPA == FALSE)
-
-testis_expression(somatic_testis$external_gene_name, cells = "all")
-HPA_cell_type_expression(somatic_testis$external_gene_name)
-
-not_TCGA_activated <- filter(lost, TCGA_category != "activated" & 
-                               TCGA_category != "multimapping_issue")
-
-TCGA_expression(not_TCGA_activated$external_gene_name,
-                tumor = "all",
-                units = "log_TPM")
-
-not_CCLE_activated <- filter(lost, CCLE_category  != "activated")
-
-CCLE_expression(not_CCLE_activated$external_gene_name,
-                  type = c("lung", "skin", "colorectal",
-                           "gastric", "breast", "head_and_neck"),
-                units = "log_TPM")
-
-transcript_prob <- lost %>% 
-  filter(testis_specificity == "testis_specific" |
-           testis_specificity == "testis_preferential") %>%
-  filter(TCGA_category == "activated" | TCGA_category == "multimapping_issue") %>% 
-  filter(CCLE_category == "activated") %>% 
-  dim()
-```
-
-`r length(common)` genes in our CTexploreR database are found in at least one of
-the other database, which represents `r length(common)/dim(CT_genes)[1] * 100`%.
-
-We have lost `r length(lost_list)` genes in total. Among them, 
-`r dim(not_specific)[1]/length(lost_list)*100`% are not considered testis 
-specific, `r dim(somatic_testis)[1]/length(lost_list)*100`% are expressed in 
-somatic cells, `r dim(not_TCGA_activated)[1]/length(lost_list)*100`% 
-are not activated in TCGA samples, 
-`r dim(not_CCLE_activated)[1]/length(lost_list)*100`% are not activated in CCLE 
-cell lines and `r transcript_prob[1]/length(lost_list)*100`% is lost due to 
-transcripts problems.
-
-What about new genes in CTexploreR
-
-```{r new, warning = FALSE}
-new <- CT_genes %>% 
-  filter(!external_gene_name%in%common)
-
-new
-
-table(new$testis_specificity)
-table(new$X_linked)
-table(new$regulated_by_methylation)
-table(new$X_linked, new$regulated_by_methylation)
-TCGA_expression(tumor = "all", genes = new$external_gene_name, 
-                units = "log_TPM")
-TCGA_expression(tumor = "all", 
-                genes = filter(new, X_linked & regulated_by_methylation)$external_gene_name, 
-                units = "log_TPM")
-```
-
-There are `r dim(new)[1]` new CT genes in CTexploreR. These are all testis
-specific and mainly on autosomes. Regulation by methylation is the majority of 
-them.  There 
-is only 11 new "major" CT that are on the X chromosome and regulated by 
-methylation. CT45 are not that new.
-
-Expression in tumours doesn't strike that much.
 
@@
 genes_avail <- 
-  CT_genes$external_gene_name[CT_genes$external_gene_name %in% unique(rownames(testis_sce))]
-table(CT_genes$testis_cell_type)
-table(CT_genes$testis_cell_type)/length(genes_avail)*100
+  CT_genes[CT_genes$external_gene_name %in% unique(rownames(testis_sce)), ]
+CT_genes_avail <- left_join(genes_avail, data.frame(rowData(testis_sce)))
+table(CT_genes_avail$testis_cell_type)
+table(CT_genes_avail$testis_cell_type)/dim(CT_genes_avail)[1]*100
 ```
 
-`r sum(table(CT_genes$testis_cell_type)[c("Early_spermatocyte", "Spermatogonia", "SSC")])/length(genes_avail)*100`
+`r sum(table(CT_genes_avail$testis_cell_type)[c("Early_spermatocyte", "Spermatogonia", "SSC")])/dim(genes_avail)[1]*100`
 % of genes are mainly expressed pre-meioticly.
@@
 
-testis_sce_CT <- testis_sce[genes_avail, ]
+testis_sce_CT <- testis_sce[CT_genes_avail$external_gene_name, ]
 
@@
 
-CT_genes_avail <- filter(CT_genes, external_gene_name %in% genes_avail)
-
 chr_mat <- as.matrix(CT_genes_avail$X_linked)
@@
 
-Heatmap(mat[genes_avail, rownames(df_col), drop = FALSE],
+Heatmap(mat[CT_genes_avail$external_gene_name, rownames(df_col), drop = FALSE],
         name = "logCounts",

### 2024-05-14 Paper_code-Database_Comparison.Rmd
Reorganisation of paper code (in prep of adding Axelle code) and fig in pdf

--- 2024-05-10 Paper_code-Database_CT.Rmd
@@
 library(UpSetR)
-library(ComplexHeatmap)
-library(circlize)
-library(SingleCellExperiment)
-library(org.Hs.eg.db)
-library(clusterProfiler)
-library(msigdbr)
-library(DOSE)
-library(BiocParallel)
-library(patchwork)
-library(Biostrings)
 ```
@@
 ```{r load_data, warning=FALSE, message=FALSE}
-# CCLE_data <- CTdata::CCLE_data()
-# TCGA_TPM <- CTdata::TCGA_TPM()
-# testis_sce <- CTdata::testis_sce()
-####
-
-load("../CTdata/eh_data/CCLE_data.rda")
-load("../CTdata/eh_data/TCGA_TPM.rda")
-load("../CTdata/eh_data/testis_sce.rda")
-load("../CTdata/eh_data/mean_methylation_in_tissues.rda")
-
 load("../CTdata/eh_data/all_genes.rda")
-load("../CTdata/eh_data/HPA_cell_type_specificities.rda")
 load("../CTdata/eh_data/CT_genes.rda")
@@
 CT_genes <- (filter(CT_genes, CT_gene_type == "CT_gene"))
-```
-
-Common figures parameters
-
-```{r fig_param}
-legends_param <- list(
-  labels_gp = gpar(col = "black", fontsize = 6),
-  title_gp = gpar(col = "black", fontsize = 6),
-  row_names_gp = gpar(fontsize = 4),
-  annotation_name_side = "left")
-
-legend_colors <- c("#5E4FA2", "#3288BD", "#66C2A5", "#ABDDA4", "#E6F598",
-                   "#FFFFBF", "#FEE08B", "#FDAE61", "#F46D43", "#D53E4F",
-                   "#9E0142")
-chr_colors <- c("X-linked" = "deeppink", "Not X" = "royalblue1")
-
-meth_colors <- c("Methylation" = "lightgreen", "Not methylation" = "gray")
-```
-
-# Pipeline figure values
-
-Here is the code used to have all the values in the pipeline figure
-
-```{r pipeline figures value}
-table(all_genes$GTEX_category)
-table(all_genes$multimapping_analysis)
-
-table(all_genes$testis_specificity)
-specific_to_pref <- all_genes %>%
-  filter(GTEX_category == "testis_specific" |
-           multimapping_analysis == "testis_specific") %>% 
-  filter(testis_specificity == "testis_preferential")
-table(specific_to_pref$not_detected_in_somatic_HPA)
-table(specific_to_pref$CCLE_category)
-table(specific_to_pref$TCGA_category)
-
-all_genes %>% 
-  filter(testis_specificity != "not_testis_specific") %>% 
-  filter(CCLE_category == "activated" & 
-           (TCGA_category == "multimapping_issue" | 
-              TCGA_category == "activated")) %>% 
-  pull(testis_specificity) %>% 
-  table()
-
-table(all_genes$CT_gene_type) #Before and after TSS verif
+
+upset_text_size <- c(1.5, 1.5, 1.5, 1.5, 1.8, 2)
+ #c(intersection size title, intersection size tick labels, set size title, 
+ # set size tick labels, set names, numbers above bars)
 ```
@@
 
+```{r, eval = FALSE, echo = FALSE}
+pdf(file = paste0("figs/Venn_CTdatabase.pdf"))
+plot(CTdatabase_ours, gp = gp)
+dev.off()
+```
+
 We find `r length(CTdatabase_ours@IntersectionSets[["11"]])/nrow(CTdatabase_cleaned)*100`
@@
 upset_CTdatabase <- fromList(CTdatabase_lost_upset)
-upset(upset_CTdatabase)
-
- (nrow(CTdatabase_lost) - table(CTdatabase_lost$testis_specificity)[["testis_specific"]])/nrow(CTdatabase_lost)*100
+
+upset(upset_CTdatabase,
+      text.scale = upset_text_size)
+```
+
+```{r, eval = FALSE, echo = FALSE}
+pdf(file = paste0("figs/upset_CTdatabase_lost.pdf"))
+upset(upset_CTdatabase,
+      text.scale = upset_text_size)
+dev.off()
 ```
@@
 upset(upset_omics)
+```
+
+```{r, eval = FALSE, echo = FALSE}
+pdf(file = paste0("figs/upset_omics.pdf"))
+upset(upset_omics)
+dev.off()
 ```
@@
 upset_Jamin <- fromList(Jamin_lost_upset)
-upset(upset_Jamin)
+upset(upset_Jamin,
+      text.scale = upset_text_size)
+```
+
+```{r, eval = FALSE, echo = FALSE}
+pdf(file = paste0("figs/Venn_Jamin.pdf"))
+plot(core_ours, gp = gp)
+dev.off()
+
+pdf(file = paste0("figs/upset_Jamin_lost.pdf"))
+upset(upset_Jamin,
+      text.scale = upset_text_size)
+dev.off()
 ```
@@
 upset_Wang <- fromList(Wang_lost_upset)
-upset(upset_Wang)
+upset(upset_Wang,
+      text.scale = upset_text_size)
+```
+
+```{r, eval = FALSE, echo = FALSE}
+pdf(file = paste0("figs/Venn_Wang.pdf"))
+plot(Wang_ours, gp = gp)
+dev.off()
+
+pdf(file = paste0("figs/upset_Wang_lost.pdf"))
+upset(upset_Wang,
+      text.scale = upset_text_size)
+dev.off()
 ```
@@
 upset_Carter <- fromList(Carter_lost_upset)
-upset(upset_Carter)
+upset(upset_Carter,
+      text.scale = upset_text_size)
+```
+
+```{r, eval = FALSE, echo = FALSE}
+pdf(file = paste0("figs/Venn_Carter.pdf"))
+plot(Carter_ours, gp = gp)
+dev.off()
+
+pdf(file = paste0("figs/upset_Carter_lost.pdf"))
+upset(upset_Carter,
+      text.scale = upset_text_size)
+dev.off()
 ```
@@
 upset_Bruggeman <- fromList(Bruggeman_lost_upset)
-upset(upset_Bruggeman)
+upset(upset_Bruggeman,
+      text.scale = upset_text_size)
+```
+
+```{r, eval = FALSE, echo = FALSE}
+pdf(file = paste0("figs/Venn_Bruggeman.pdf"))
+plot(Bruggeman_ours, gp = gp)
+dev.off()
+
+pdf(file = paste0("figs/upset_Bruggeman_lost.pdf"))
+upset(upset_Bruggeman,
+      text.scale = upset_text_size)
+dev.off()
 ```
@@
 
+all_lost_upset <- 
+  list(`Not testis specific` = 
+         filter(lost,
+                testis_specificity != "testis_specific")$external_gene_name,
+       `Not tumour activated` = 
+         filter(lost,
+                (TCGA_category != "activated" &
+                  TCGA_category != "multimapping_issue")|
+                   CCLE_category != "activated")$external_gene_name,
+       `CT preferential` =
+         filter(lost,
+                CT_gene_type == "CTP_gene")$external_gene_name)
+
+upset_all <- fromList(all_lost_upset)
+upset(upset_all,
+      text.scale = upset_text_size)
+
 not_specific <- filter(lost, testis_specificity == "not_testis_specific")
@@
   dim()
+```
+
+```{r, eval = FALSE, echo = FALSE}
+pdf(file = paste0("figs/upset_all_lost.pdf"))
+upset(upset_all,
+      text.scale = upset_text_size)
+dev.off()
 ```
@@
 
-# CT genes characteristics
-
-```{r basic}
-table(CT_genes$testis_specificity)
-table(CT_genes$transcript_biotype)
-```
-
-Most genes are testis specific 
-(`r table(CT_genes$testis_specificity)[["testis_specific"]]/dim(CT_genes)[1]*100`%).
-Most genes are mainly protein coding genes 
-(`r table(CT_genes$transcript_biotype)["protein_coding"]/dim(CT_genes)[1]*100`%).
-
-## X chromosome and regulated by methylation
-
-In CTexploreR, genes have been characterised as regulated by methylation or not.
-
-```{r X_meth}
-table(CT_genes$X_linked)
-table(CT_genes$regulated_by_methylation)
-table(CT_genes$X_linked, CT_genes$testis_specificity)
-table(CT_genes$X_linked, CT_genes$regulated_by_methylation)
-```
-
-Genes are enriched on the X chromosome 
-(`r table(CT_genes$X_linked)[[2]]/dim(CT_genes)[1]*100`%). Also, 
-`r table(CT_genes$regulated_by_methylation)[[2]]` genes have been flagged as 
-regulated by methylation 
-(`r table(CT_genes$regulated_by_methylation)[[2]]/dim(CT_genes)[1]*100`%). 
-It is interesting to study the link between these two characteristics.
-
-On the chromosome X, there is a clear enrichment of CT genes regulated by 
-methylation (`r table(CT_genes$X_linked, CT_genes$regulated_by_methylation)[2, 2]`/`r table(CT_genes$X_linked)[[2]]` 
-chrX genes or `r table(CT_genes$X_linked, CT_genes$regulated_by_methylation)[2, 2]`/`r table(CT_genes$regulated_by_methylation)[[2]]` 
-genes regulated by methylation).
-
-Let's check that with a statistical test, I here want to see if, on there is an 
-enrichment of genes regulated by methylation on the X chromosome. I need to do a
-Pearson Chi square test (to know if observed proportion differ from expected 
-proportion). It is a statistical method to determine if two categorical variables
-have a significant correlation between them. I can directly put a matrix 
-(my table) in the function
-
-```{r chisq}
-chisq.test(table(CT_genes$X_linked, CT_genes$regulated_by_methylation))
-```
-
-```{r X_meth_fig}
-CT_genes$chr_factor <- factor(CT_genes$chr,
-                       levels = c("1", "2", "3", "4", "5", "6", "7", "8", "9",
-                                  "10", "11", "12", "13", "14", "15", "16", "17", 
-                                  "18", "19", "20", "21", "22", "X", "Y"))
-
-CT_genes %>% 
-  mutate(regulated_by_methylation = ifelse(regulated_by_methylation,
-                                           "Regulated by methylation",
-                                           "Not regulated by methylation")) %>% 
-  ggplot(aes(x = chr_factor, fill = X_linked)) +
-  geom_bar(stat = 'count') +
-  scale_fill_manual(values = c("royalblue1", "deeppink")) +
-  facet_wrap(~ regulated_by_methylation, ncol = 2) +
-  theme_bw() +
-  xlab("Chromosome") +
-  ylab("Number of genes") +
-  theme(axis.text.x = element_text(angle = 90, hjust = 1),
-        legend.position = "none",
-        axis.title = element_text(size = 10, color = "gray10"))
-```
-
-There is indeed a significative link between regulation by methylation and being
-on the X chromosome. There is thus an enrichment of CT genes regulated by 
-methylation on the X chromosome (and inversely).
-
-## Tumour and methylation
-
-Using CTexploreR functions, we can explore all CT genes or focus on some 
-potential targets.
-
-### All CT
-
-For these heatmaps, the code comes from the function but has been copies to add 
-some annotations.
-
-```{r common_annot}
-chr_mat <- as.matrix(CT_genes$X_linked)
-chr_mat <- ifelse(chr_mat == TRUE, "X-linked", "Not X")
-rownames(chr_mat) <- CT_genes$external_gene_name
-row_ha_chr <- rowAnnotation(chr_factor = chr_mat,
-                            annotation_legend_param = legends_param,
-                            simple_anno_size = unit(0.5, "cm"),
-                            col = list(chr_factor = chr_colors),
-                            annotation_name_gp = gpar(fontsize = 8),
-                            annotation_name_side = "top")
-
-regulation_mat <- as.matrix(CT_genes$regulated_by_methylation)
-regulation_mat <- ifelse(regulation_mat == TRUE, "Methylation",
-                         "Not methylation")
-rownames(regulation_mat) <- CT_genes$external_gene_name
-
-row_ha_reg <- rowAnnotation(regulation = regulation_mat,
-                            annotation_legend_param = legends_param,
-                            simple_anno_size = unit(0.5, "cm"),
-                            col = list(regulation = meth_colors),
-                            annotation_name_gp = gpar(fontsize = 8),
-                            annotation_name_side = "top")
-
-left_annot <- c(row_ha_chr, row_ha_reg, gap = unit(1, "mm"))
-
-split <- data.frame(CT_genes$regulated_by_methylation, CT_genes$X_linked)
-```
-
-```{r TCGA_heatmap, message = FALSE}
-database <- TCGA_TPM
-database$tumor <- sub(pattern = 'TCGA-', x = database$project_id, '')
-database$type <- "Tumor"
-database$type[database$shortLetterCode == "NT"] <- "Peritumoral"
-database <- database[, order(database$tumor, database$type)]
-
-genes <- CT_genes$external_gene_name
-database <- database[rowData(database)$external_gene_name %in% genes, ]
-database <- database[match(genes, rowData(database)$external_gene_name), ]
-database <- database[, database$type == "Tumor"]
-
-column_ha_tumor <- HeatmapAnnotation(
-  Tumor = database$tumor,
-  border = TRUE,
-  col = list(Tumor = c("BRCA" = "midnightblue", "COAD" = "darkorchid2",
-                 "ESCA" = "gold", "HNSC" = "deeppink2",
-                 "LUAD" = "seagreen", "LUSC" = "seagreen3",
-                 "SKCM" = "red3")),
-  annotation_name_gp = gpar(fontsize = 8),
-  annotation_legend_param = legends_param)
-
-split_by <- factor(database$tumor)
-col_annot <- column_ha_tumor
-
-mat <- log1p(assay(database))
-rownames(mat) <- rowData(database)$external_gene_name
-
-Heatmap(mat[, , drop = FALSE],
-        name = "logTPM",
-        column_title = paste0("Expression in TCGA samples (all)"),
-        column_split = split_by,
-        row_split = split,
-        row_title_gp = gpar(fontsize = 0),
-        col = colorRamp2(seq(0, max(mat), length = 11),
-                         legend_colors),
-        clustering_method_rows = "ward.D",
-        clustering_method_columns = "ward.D",
-        cluster_rows = TRUE,
-        show_column_names = FALSE,
-        cluster_columns = TRUE,
-        show_column_dend = FALSE,
-        show_row_dend = FALSE,
-        row_names_gp = gpar(fontsize = 4),
-        heatmap_legend_param = legends_param,
-        top_annotation = col_annot,
-        left_annotation = left_annot)
-
-```
-
-```{r CCLE_heatmap}
-database <- CCLE_data
-database$type <- tolower(database$type)
-genes <- CT_genes$external_gene_name
-database <- database[rowData(database)$external_gene_name %in% genes, ]
-database <- database[match(genes, rowData(database)$external_gene_name), ]
-
-mat <- log1p(assay(database))
-rownames(mat) <- rowData(database)$external_gene_name
-
-df_col <- data.frame("cell_line" = colData(database)$cell_line_name,
-                     "type" = as.factor(colData(database)$type))
-rownames(df_col) <- rownames(colData(database))
-df_col <- df_col[order(df_col$type), ]
-
-column_ha_type <- HeatmapAnnotation(
-  type = df_col$type,
-  border = TRUE,
-  annotation_name_gp = gpar(fontsize = 8),
-  annotation_legend_param = legends_param,
-  col = list(type = c("lung" = "seagreen3", "skin" = "red3",
-                 "bile_duct" = "mediumpurple1", "bladder" = "mistyrose2",
-                 "colorectal" = "plum", "lymphoma" = "steelblue1",
-                 "uterine" = "darkorange4", "myeloma" = "turquoise3", 
-                 "kidney" = "thistle4",
-                 "pancreatic" = "darkmagenta", "brain" = "palegreen2",
-                 "gastric" = "wheat3", "breast" = "midnightblue",
-                 "bone" = "sienna1", "head_and_neck" = "deeppink2",
-                 "ovarian" = "tan3", "sarcoma" = "lightcoral",
-                 "leukemia" = "steelblue4", "esophageal"= "khaki",
-                 "neuroblastoma" = "olivedrab1")))
-
-Heatmap(mat[, rownames(df_col), drop = FALSE],
-        name = "logTPM",
-        column_title = "Gene Expression in tumor cell lines (CCLE)",
-        column_split = factor(df_col$type),
-        row_split = split,
-        row_title_gp = gpar(fontsize = 0),
-        col = colorRamp2(seq(0, max(mat), length = 11),
-                          legend_colors),
-        clustering_method_rows = "ward.D",
-        clustering_method_columns = "ward.D",
-        cluster_rows = TRUE,
-        show_row_dend = FALSE,
-        show_column_names = FALSE,
-        cluster_columns = TRUE,
-        show_column_dend = FALSE,
-        row_names_gp = gpar(fontsize = 4),
-        heatmap_legend_param = legends_param,
-        top_annotation = c(column_ha_type),
-        left_annotation = left_annot)
-```
-
-Next graph was removed from paper
-
-```{r all_meth}
-genes <- CT_genes$external_gene_name
-database <- mean_methylation_in_tissues[rownames(mean_methylation_in_tissues) %in% genes]
-
-mat <- na.omit(assay(database))
-clustering_option <- TRUE
-
-row_ha_chr_meth <- rowAnnotation(chr = chr_mat[rownames(mat),],
-                            annotation_legend_param = legends_param,
-                            simple_anno_size = unit(0.5, "cm"),
-                            col = list(chr = chr_colors),
-                            annotation_name_gp = gpar(fontsize = 8),
-                            annotation_name_side = "top")
-
-row_ha_reg_meth <- rowAnnotation(regulation = regulation_mat[rownames(mat),],
-                            annotation_legend_param = legends_param,
-                            simple_anno_size = unit(0.5, "cm"),
-                            col = list(regulation = meth_colors),
-                            annotation_name_gp = gpar(fontsize = 8),
-                            annotation_name_side = "top")
-
-left_annot_meth <- c(row_ha_chr_meth, row_ha_reg_meth, gap = unit(1, "mm"))
-
-split_meth <- data.frame(filter(CT_genes, external_gene_name %in% rownames(mat))$regulated_by_methylation, 
-                    filter(CT_genes, external_gene_name %in% rownames(mat))$X_linked)
-
-Heatmap(mat,
-        column_title = 'Promoter mean methylation level by tissue',
-        name = 'Meth',
-        col = colorRamp2(c(1:100),
-                         colorRampPalette(c("moccasin","dodgerblue4"))(100)),
-        na_col = "gray80",
-        cluster_rows = clustering_option,
-        cluster_columns = FALSE,
-        row_split = split_meth,
-        row_title_gp = gpar(fontsize = 0),
-        show_row_names = TRUE,
-        show_heatmap_legend = TRUE,
-        show_row_dend = FALSE,
-        row_names_gp = gpar(fontsize = 3),
-        column_names_gp = gpar(fontsize = 8),
-        column_names_side = "bottom",
-        row_names_side = "right",
-        left_annotation = left_annot_meth)
-
-```
-
-### MAGE genes
-
-```{r MAGE_TCGA}
-MAGE_genes <- filter(CT_genes, family == "MAGE")$external_gene_name
-
-TCGA_expression(tumor = "all", 
-                genes = MAGE_genes,
-                units = "log_TPM")
-
-CCLE_expression(genes = MAGE_genes,
-                 type = c("lung", "skin", "bile_duct", "bladder", 
-                          "colorectal", "lymphoma", "uterine",
-                          "myeloma", "kidney", "pancreatic", "brain", 
-                          "gastric", "breast", "bone", "head_and_neck", 
-                          "ovarian", "sarcoma", "leukemia", "esophageal",
-                          "neuroblastoma"), units = "log_TPM")
-
-normal_tissues_mean_methylation(MAGE_genes, na.omit = TRUE)
-
-```
-
-## Single Cell analysis : timing of expression
-
-In CTexploreR, we there is single cell data from the testis, we thus can 
-analyse CT genes expression during spermatogenesis.
-
-There is only data for `r sum(CT_genes$external_gene_name %in% unique(rownames(testis_sce)))` of our `r nrow(CT_genes)` CT genes.
-
-NB : SSC, Spermatogonia and early spermatocytes are *premeiotic cells*. Late 
-spermatocytes (between both meiosis), round spermatid, elongated spermatid and 
-sperm are *postmeiotic cells*.
-
-```{r repartition_sc}
-genes_avail <- 
-  CT_genes$external_gene_name[CT_genes$external_gene_name %in% unique(rownames(testis_sce))]
-table(CT_genes$testis_cell_type)
-table(CT_genes$testis_cell_type)/length(genes_avail)*100
-```
-
-`r sum(table(CT_genes$testis_cell_type)[c("Early_spermatocyte", "Spermatogonia", "SSC")])/length(genes_avail)*100`
-% of genes are mainly expressed pre-meioticly.
-
-```{r sc_heatmap, message = FALSE}
-germ_cells <- c("SSC", "Spermatogonia", "Early_spermatocyte",
-                "Late_spermatocyte","Round_spermatid", "Elongated_spermatid",
-                "Sperm1", "Sperm2")
-somatic_cells <- c("Macrophage", "Endothelial", "Myoid", "Sertoli", "Leydig")
-
-testis_sce_CT <- testis_sce[genes_avail, ]
-
-mat <- SingleCellExperiment::logcounts(testis_sce_CT)
-
-df_col <- data.frame(clusters = colData(testis_sce_CT)$clusters,
-                     type = colData(testis_sce_CT)$type,
-                     Donor = colData(testis_sce_CT)$Donor)
-rownames(df_col) <- colnames(testis_sce_CT)
-
-df_col <- df_col[order(df_col$type),]
-df_col$lineage <- "Germ cells"
-df_col$lineage[df_col$type %in% somatic_cells] <- "Somatic cells"
-
-column_ha_type = HeatmapAnnotation(
-  type = df_col$type,
-  border = TRUE,
-  col = list(type = c("SSC" = "floralwhite", "Spermatogonia" = "moccasin",
-                   "Early_spermatocyte" = "gold", 
-                   "Late_spermatocyte" = "orange",
-                   "Round_spermatid" = "red2", 
-                   "Elongated_spermatid" = "darkred",
-                   "Sperm1" = "violet", "Sperm2" = "purple", 
-                   "Sertoli" = "gray", 
-                   "Leydig" = "cadetblue2", "Myoid" = "springgreen3", 
-                   "Macrophage" = "gray10",
-                   "Endothelial" = "steelblue")),
-
-  annotation_name_gp = gpar(fontsize = 8),
-  annotation_legend_param = legends_param)
-
-column_ha_lineage = HeatmapAnnotation(
-  lineage = df_col$lineage,
-  border = TRUE,
-  col = list(lineage = c("Germ cells" = "salmon", "Somatic cells" = "cyan4")),
-  annotation_name_gp = gpar(fontsize = 8),
-  annotation_legend_param = legends_param)
-
-scale_lims <- c(0, quantile(rowMax(mat), 0.75))
-top_annot <- c(column_ha_lineage, column_ha_type)
-
-# Until here is what's in the function, hereunder is my addition/change in Heatmap()
-
-CT_genes_avail <- filter(CT_genes, external_gene_name %in% genes_avail)
-
-chr_mat <- as.matrix(CT_genes_avail$X_linked)
-chr_mat <- ifelse(chr_mat == TRUE, "X-linked", "Not X")
-rownames(chr_mat) <- CT_genes_avail$external_gene_name
-row_ha_chr <- rowAnnotation(chr = chr_mat,
-                            annotation_legend_param = legends_param,
-                            simple_anno_size = unit(0.5, "cm"),
-                            col = list(chr = chr_colors),
-                            annotation_name_gp = gpar(fontsize = 8),
-                            annotation_name_side = "top")
-
-regulation_mat <- as.matrix(CT_genes_avail$regulated_by_methylation)
-regulation_mat <- ifelse(regulation_mat == TRUE, "Methylation",
-                         "Not methylation")
-rownames(regulation_mat) <- CT_genes_avail$external_gene_name
-
-row_ha_reg <- rowAnnotation(regulation = regulation_mat,
-                            annotation_legend_param = legends_param,
-                            simple_anno_size = unit(0.5, "cm"),
-                            col = list(regulation = meth_colors),
-                            annotation_name_gp = gpar(fontsize = 8),
-                            annotation_name_side = "top")
-
-left_annot <- c(row_ha_chr, row_ha_reg, gap = unit(1, "mm"))
-
-split <- data.frame(CT_genes_avail$regulated_by_methylation, CT_genes_avail$X_linked)
-
-Heatmap(mat[genes_avail, rownames(df_col), drop = FALSE],
-        name = "logCounts",
-        column_title = "Expression in testis cells (scRNAseq)",
-        column_split = df_col$type,
-        row_split = split,
-        row_title_gp = gpar(fontsize = 0),
-        show_column_names = FALSE,
-        show_column_dend = FALSE,
-        clustering_method_rows = "ward.D",
-        clustering_method_columns = "ward.D",
-        cluster_rows = TRUE,
-        cluster_columns = FALSE,
-        show_row_dend = FALSE,
-        row_names_gp = gpar(fontsize = 4),
-        col = colorRamp2(seq(scale_lims[1], scale_lims[2], length = 11),                 
-                         legend_colors),
-        top_annotation = top_annot,
-        left_annotation = left_annot,
-        heatmap_legend_param = legends_param)
-
-```
-
-## Gene function
-
-All CT genes function
-
-```{r GO_ORA}
-msigdbr(species = "Homo sapiens" , category = "C5") %>% 
-  filter(gene_symbol %in% CT_genes$external_gene_name) %>% 
-  pull(gene_symbol) %>% 
-  unique() %>% 
-  length()
-
-msigdbr(species = "Homo sapiens" , category = "H") %>% 
-  filter(gene_symbol %in% CT_genes$external_gene_name) %>% 
-  pull(gene_symbol) %>% 
-  unique() %>% 
-  length()
-
-go_ora <- enrichGO(gene = CT_genes$ensembl_gene_id,
-                   keyType = "ENSEMBL",
-                   OrgDb = org.Hs.eg.db,
-                   ont = "all",
-                   readable = TRUE)
-as_tibble(go_ora)
-
-as_tibble(go_ora) %>% 
-  arrange(desc(Count)) %>% 
-  head(12) %>% 
-  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/177,
-                           ONTOLOGY == "CC"~ Count/197,
-                           ONTOLOGY == "MF"~ Count/186)) %>% 
-  ggplot(aes(x = Ratio, y = Description, fill = Description)) +
-  geom_col() +
-  theme_bw() +
-  ylab("GO term") +
-  xlab("Gene Ratio") +
-  theme(axis.text.y = element_blank(),
-        legend.position = "none",
-        axis.ticks.y = element_blank(),
-        axis.title = element_text(size = 10, color = "gray10"))+ 
-  geom_text(aes(0, y = Description, label = Description),
-            hjust = 0,
-            nudge_x = 0.005,
-            colour = "floralwhite",
-            size = 4)
-
-ora_to_plot <- as_tibble(simplify(go_ora))
-
-ora_to_plot <- ora_to_plot %>% 
-  arrange(desc(Count)) %>% 
-  head(9) %>% 
-  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/102,
-                           ONTOLOGY == "MF"~ Count/186))
-
-ora_to_plot %>% 
-  ggplot(aes(x = Ratio, y = Description, fill = Description)) +
-  geom_col() +
-  theme_bw() +
-  ylab("GO term") +
-  xlab("Gene Ratio") +
-  theme(axis.text.y = element_blank(),
-        legend.position = "none",
-        axis.ticks.y = element_blank(),
-        axis.title = element_text(size = 10, color = "gray10"))+ 
-  geom_text(aes(0, y = Description, label = Description),
-            hjust = 0,
-            nudge_x = 0.005,
-            colour = "floralwhite",
-            size = 3.7)
-```
-
-As we can see here, most of genes are indeed linked to functions from reproduction.
-I represented here the 12 categories with the most genes, all from biological 
-processes. However, they are enriched in `r nrow(as_tibble(go_ora))` different GO terms.
-
-Is there a difference between meth reg or not ?
-
-```{r GO_ORA_meth}
-go_ora_meth <- enrichGO(gene = 
-                          filter(CT_genes, regulated_by_methylation)$ensembl_gene_id,
-                        keyType = "ENSEMBL",
-                        OrgDb = org.Hs.eg.db,
-                        ont = "all",
-                        readable = TRUE)
-go_ora_meth <- simplify(go_ora_meth)
-as_tibble(go_ora_meth)
-
-go_ora_not_meth <- enrichGO(gene = 
-                          filter(CT_genes, !regulated_by_methylation)$ensembl_gene_id,
-                        keyType = "ENSEMBL",
-                        OrgDb = org.Hs.eg.db,
-                        ont = "all",
-                        readable = TRUE)
-go_ora_not_meth <- simplify(go_ora_not_meth)
-as_tibble(go_ora_not_meth)
-
-go_ora_meth_plot <- as_tibble(go_ora_meth) %>% 
-  arrange(desc(Count)) %>% 
-  head(10) %>% 
-  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/73,
-                           ONTOLOGY == "MF"~ Count/80,
-                           ONTOLOGY == "CC"~ Count/78)) %>%  
-  mutate(regulation = "Methylation")
-
-go_ora_not_meth_plot <- as_tibble(go_ora_not_meth) %>% 
-  arrange(desc(Count)) %>% 
-  head(8) %>% 
-  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/29,
-                           ONTOLOGY == "MF"~ Count/30))%>%  
-  mutate(regulation = "Not methylation")
-
-rbind(go_ora_meth_plot, go_ora_not_meth_plot) %>% 
-  ggplot(aes(x = Ratio, y = Description, fill = Description)) +
-  geom_col() +
-  theme_bw() +
-  ylab("GO term") +
-  xlab("Gene Ratio") +
-  facet_wrap(~ regulation) + 
-  theme(legend.position = "none",
-        axis.ticks.y = element_blank(),
-        axis.text.y = element_text(size = 6, color = "gray10"),
-        axis.title.x = element_text(size = 10, color = "gray10"),
-        axis.title.y = element_blank())+ 
-  geom_text(aes(0, y = Description, label = ID),
-            hjust = 0,
-            nudge_x = 0.005,
-            colour = "floralwhite",
-            size = 2)
-
-```
-
-We've also added a column tumor suppressor and oncogene in CTexploreR. 
-These information come from Cancermine.
-
-```{r C_TS}
-table(CT_genes$oncogene) 
-table(CT_genes$tumor_suppressor)
-
-table(CT_genes$oncogene, CT_genes$tumor_suppressor) 
-```
-
 # SessionInfo

### 2024-05-16 Paper_code-CT_analysis.Rmd
change due to change in CT selection

@@
 
-Here is the code used to have all the values in the pipeline figure
+Here is the code used to have all the values in the pipeline figure.
+
+As a reminder testis_specificity has been done summarizing testis-specificity 
+analysis from GTEX, multimapping, HPA_cell_type_specificities, CCLE and TCGA 
+categories.
+
+Testis-specific genes must be: classified as "testis-specific" in GTEX or 
+multimappig analysis +  must not be detected in any somatic cell type (in HPA 
+scRNAseq analysis) + can not be classified as "leaky" in CLLE and TCGA 
+categories + can not be detected (TPM > 0.5) in more than 75% of normal 
+peritumoral tissues from a TCGA tumor type.
+
+Testis_preferential genes must be: classified as "testis_preferential" in GTEX 
+or multimappig analysis or initially classified as "testis-specific" in GTEX or 
+multimappig analysis but downgraded to "testis_preferential because they were 
+detected in a somatic cell type (in HPA scRNAseq analysis) + can be detected in 
+some somatic cell type but the level has to be at least 10 times lower than the
+level detected in a germ cell type (in HPA scRNAseq analysis) must not be 
+classified as "leaky" in CLLE and TCGA categories.
 
@@
 table(all_genes$testis_specificity)
+table(all_genes$GTEX_category, all_genes$testis_specificity)
+
 specific_to_pref <- all_genes %>%
@@
   filter(testis_specificity == "testis_preferential")
-table(specific_to_pref$not_detected_in_somatic_HPA)
-table(specific_to_pref$CCLE_category)
-table(specific_to_pref$TCGA_category)
+# No more specific to pref as their ratio are all < 10
+# all_genes %>%
+#   filter(GTEX_category == "testis_specific" |
+#            multimapping_analysis == "testis_specific") %>% 
+#   filter(!not_detected_in_somatic_HPA)
+
+specific_to_none <- all_genes %>%
+  filter(GTEX_category == "testis_specific" |
+           multimapping_analysis == "testis_specific") %>% 
+  filter(testis_specificity == "not_testis_specific")
+table(specific_to_none$CCLE_category)
+table(specific_to_none$TCGA_category)
+table(specific_to_none$max_q75_in_NT < 0.5)
+# 95 lost at this step
+table(specific_to_none$not_detected_in_somatic_HPA)
+table(specific_to_none$HPA_ratio_germ_som >= 10)
+
+preferential_to_none <- all_genes %>%
+  filter(GTEX_category == "testis_preferential" |
+           multimapping_analysis == "testis_preferential") %>% 
+  filter(testis_specificity == "not_testis_specific")
+table(preferential_to_none$CCLE_category)
+table(preferential_to_none$TCGA_category)
+table(preferential_to_none$not_detected_in_somatic_HPA)
+# Almost all detected in somatic
+table(preferential_to_none$HPA_ratio_germ_som >= 10)
 

### 2024-05-16 Paper_code-Database_Comparison.Rmd
Updated database comparison with new CT_genes, changed some database cleaning

@@
   rbind(CTdatabase_cleaned)
+
+CTdatabase_cleaned <- CTdatabase_cleaned[!duplicated(CTdatabase_cleaned$external_gene_name), ]
+
+length(CTdatabase_cleaned$ensembl_gene_id[!CTdatabase_cleaned$ensembl_gene_id%in%all_genes$ensembl_gene_id])
 ```
+
+9 genes are not found in all_genes even when using gene names and ensembl_gene_id
 
@@
 Jamin_core_CT[Jamin_core_CT$Gene == "CXorf67", "Gene"] <- "EZHIP"
+Jamin_core_CT[Jamin_core_CT$Gene == "KIAA1109", "Gene"] <- "BLTP1"
+Jamin_core_CT[Jamin_core_CT$Gene == "C10orf82", "Gene"] <- "SPMIP5"
+Jamin_core_CT[Jamin_core_CT$Gene == "TTC30B", "Gene"] <- "IFT70B"
+Jamin_core_CT[Jamin_core_CT$Gene == "TTC30A", "Gene"] <- "IFT70A"
 ```
+
+All gene names have been linked to the right name in all_genes
 
@@
 Wang_CT[Wang_CT$ensembl_gene_id == "ENSG00000204293", "external_gene_name"] <- "OR8B2"
-Wang_CT[Wang_CT$external_gene_name == "", "external_gene_name"] <- "RNASE11"
-Wang_CT[Wang_CT$external_gene_name == "CHCT1", "external_gene_name"] <- "C17orf64"
-Wang_CT[Wang_CT$external_gene_name == "PRSS40A", "external_gene_name"] <- "TISP43"
-Wang_CT[Wang_CT$external_gene_name == "TEX56P", "external_gene_name"] <- "C6orf201"
-Wang_CT[Wang_CT$external_gene_name == "SLC25A51P4", "external_gene_name"] <- "RP11-113D6.10"
-Wang_CT[Wang_CT$external_gene_name == "TCP10L3", "external_gene_name"] <- "TCP10"
-Wang_CT[Wang_CT$external_gene_name == "SCAND3", "external_gene_name"] <- "ZBED9"
+
+length(Wang_CT$external_gene_name[!Wang_CT$external_gene_name %in% all_genes$external_gene_name])
 ```
+
+12 genes are not found in all_genes, by gene names or ensemble gene id and no 
+other name.
 
@@
 Carter_CT[Carter_CT$Gene == "ENSG00000224960", "Gene_Name"] <- "PPP4R3C"
+Carter_CT[Carter_CT$Gene == "ENSG00000221843", "Gene_Name"] <- "SPATA31H1"
+Carter_CT[Carter_CT$Gene == "ENSG00000177947", "Gene_Name"] <- "CIMAP1A"
+Carter_CT[Carter_CT$Gene == "ENSG00000172073", "Gene_Name"] <- "SPMIP9"
+Carter_CT[Carter_CT$Gene == "ENSG00000229894", "Gene_Name"] <- "GK3"
+Carter_CT[Carter_CT$Gene == "ENSG00000249693", "Gene_Name"] <- "SPMAP2L"
+Carter_CT[Carter_CT$Gene == "ENSG00000173728", "Gene_Name"] <- "SPMIP3"
+
+length(Carter_CT$Gene[!Carter_CT$Gene %in% all_genes$ensembl_gene_id])
+
 ```
+
+1 gene cannot be found in all_genes even by ensemble or gene name
 
@@
 
-lost <- Bruggeman_data[which(!Bruggeman_data$Gene %in% c(Bruggeman_CT$Gene_Name)), "Gene"]
+lost <- Bruggeman_data[which(!Bruggeman_data$Gene %in% 
+                               c(Bruggeman_CT$Gene_Name)), "Gene"]
 colnames(lost) <- "external_gene_name"
@@
 lost[lost$external_gene_name == "C22orf34", "Gene_Name"] <- "MIR3667HG"
+lost[lost$external_gene_name == "C9orf131", "Gene_Name"] <- "SPATA31G1"
 
@@
 
-missing_Bruggeman <- c("BMS1P4", "ADAM6", "ANXA2P3", "ARHGAP11B", "DPY19L2P2", 
-                       "HLA-L", "PA2G4P4", "PIPSL", "PRKY", "YBX3P1", 
-                       "RPL23AP53", "UQCRBP1", "RPL23P8", "MRS2P2", "PIN4P1", 
-                       "SLC6A10P", "GUSBP2", "PPIEL", "LRRC37BP1", "MSL3P1", 
-                       "PLEKHA8P1", "STAG3L1", "TCAM1P", "ZNF702P", "ZNF815P", 
-                       "ATP6AP1L", "RPL21P44", "SEC14L1P1", "ZNF876P", 
-                       "RPLP0P2", "FAM86JP", "FAM175A", "LACE1", "ATP5EP2", 
-                       "WDR92", "TCTE3", "METTL20", "KIAA2022", "ZNRD1-AS1", 
-                       "SGOL1", "FAM35DP", "MTL5", "TMEM14E", "MLLT4-AS1", 
-                       "CCDC173", "KIAA1524", "WDR78", "LINC00476", "LYRM5", 
-                       "HILS1", "CASC5", "KIAA1919", "CTAGE5", "FAM188B", 
-                       "TMEM194B", "FAM122C", "PPP1R2P3", "KIAA0391", "SGOL2", 
-                       "FAM19A3", "ZNF788", "RPL19P12", "FIRRM")
+missing_Bruggeman <- Bruggeman_CT[!Bruggeman_CT$Gene_Name %in% 
+                                    all_genes$external_gene_name, ]$Gene_Name
 
@@
    filter(external_synonym %in% missing_Bruggeman) %>%
-   filter(!external_gene_name %in% c("ATP5F1EP2", "POLR1HASP", "SHLD2P3", 
-                                   "TMEM14EP", "H1-9P", "ZNF788P")) %>% 
+   filter(external_gene_name %in% all_genes$external_gene_name) %>% 
    mutate(Gene_Name = external_gene_name)
@@
   dplyr::select(Gene_Name)
+
+length(Bruggeman_CT$Gene_Name[!Bruggeman_CT$Gene_Name %in% 
+                                all_genes$external_gene_name])
 ```
+
+38 genes are not found with their names in all genes
 
@@
 ```{r CTdatabase_ours}
-CTdatabase_ours <- Venn(list(CTdatabase = CTdatabase_cleaned$external_gene_name,
-                             CTexploreR = CT_genes$external_gene_name))
+CTdatabase_ours <- Venn(list(CTexploreR = CT_genes$external_gene_name,
+                             CTdatabase = CTdatabase_cleaned$external_gene_name))
 gp <- VennThemes(compute.Venn(CTdatabase_ours))
-gp[["Face"]][["11"]]$fill <-  "mistyrose"
-gp[["Face"]][["01"]]$fill <-  "darkseagreen1"
-gp[["Face"]][["10"]]$fill <-  "lightsteelblue1"
-gp[["Set"]][["Set1"]]$col <-  "paleturquoise4"
-gp[["Set"]][["Set2"]]$col <-  "darkseagreen4"
-gp[["SetText"]][["Set1"]]$col <-  "paleturquoise4"
-gp[["SetText"]][["Set2"]]$col <-  "darkseagreen4"
+gp[["Face"]][["11"]]$fill <-  "mediumaquamarine"
+gp[["Face"]][["10"]]$fill <-  "darkseagreen1"
+gp[["Face"]][["01"]]$fill <-  "lightsteelblue1"
+gp[["Set"]][["Set2"]]$col <-  "paleturquoise4"
+gp[["Set"]][["Set1"]]$col <-  "darkseagreen4"
+gp[["SetText"]][["Set2"]]$col <-  "paleturquoise4"
+gp[["SetText"]][["Set1"]]$col <-  "darkseagreen4"
 plot(CTdatabase_ours, gp = gp)
 
-```
-
-```{r, eval = FALSE, echo = FALSE}
-pdf(file = paste0("figs/Venn_CTdatabase.pdf"))
-plot(CTdatabase_ours, gp = gp)
-dev.off()
 ```
@@
 CTdatabase_lost <- all_genes %>%
-  filter(external_gene_name %in% CTdatabase_ours@IntersectionSets[["10"]])
+  filter(external_gene_name %in% CTdatabase_ours@IntersectionSets[["01"]])
 
-# 9 Genes are lost because not in any database
+# 9 Genes are lost because not in any database like before
 
@@
          filter(CTdatabase_lost,
-                testis_specificity != "testis_specific")$external_gene_name,
+                testis_specificity != "testis_specific" & 
+                  CT_gene_type == "other")$external_gene_name,
        `Not tumour activated` = 
@@
 ```{r, eval = FALSE, echo = FALSE}
-pdf(file = paste0("figs/upset_CTdatabase_lost.pdf"))
+pdf(file = paste0("figs/CTdatabase_comparison.pdf"))
+plot(CTdatabase_ours, gp = gp)
 upset(upset_CTdatabase,
@@
 ```{r upset_plot}
-core_ours <- Venn(list(Jamin = Jamin_core_CT$Gene, 
-                       CTexploreR = CT_genes$external_gene_name))
+core_ours <- Venn(list(CTexploreR = CT_genes$external_gene_name,
+                       Jamin = Jamin_core_CT$Gene))
 
-Wang_ours <- Venn(list(Wang = Wang_CT$external_gene_name, 
-                       CTexploreR = CT_genes$external_gene_name))
+Wang_ours <- Venn(list(CTexploreR = CT_genes$external_gene_name,
+                       Wang = Wang_CT$external_gene_name))
 
-Carter_ours <- Venn(list(Carter_CT = Carter_CT$Gene_Name, 
-                         CTexploreR = CT_genes$external_gene_name))
+Carter_ours <- Venn(list(CTexploreR = CT_genes$external_gene_name,
+                         Carter_CT = Carter_CT$Gene_Name))
 
-Bruggeman_ours <- Venn(list(Bruggeman = Bruggeman_CT$Gene_Name,
-                            CTexploreR = CT_genes$external_gene_name))
+Bruggeman_ours <- Venn(list(CTexploreR = CT_genes$external_gene_name,
+                            Bruggeman = Bruggeman_CT$Gene_Name))
 
@@
 
-```{r, eval = FALSE, echo = FALSE}
-pdf(file = paste0("figs/upset_omics.pdf"))
-upset(upset_omics)
-dev.off()
-```
-
 4 in all, 60 in at least 3 databases
@@
 Jamin_lost <- all_genes %>%
-  filter(external_gene_name %in% core_ours@IntersectionSets[["10"]])
+  filter(external_gene_name %in% core_ours@IntersectionSets[["01"]])
 
@@
          filter(Jamin_lost,
-                testis_specificity != "testis_specific")$external_gene_name,
+                testis_specificity != "testis_specific" &
+                  CT_gene_type == "other")$external_gene_name,
        `Not tumour activated` = 
@@
 ```{r, eval = FALSE, echo = FALSE}
-pdf(file = paste0("figs/Venn_Jamin.pdf"))
+pdf(file = paste0("figs/Jamin_comparison.pdf"))
 plot(core_ours, gp = gp)
-dev.off()
-
-pdf(file = paste0("figs/upset_Jamin_lost.pdf"))
 upset(upset_Jamin,
@@
 Wang_lost <- all_genes %>%
-  filter(external_gene_name %in% Wang_ours@IntersectionSets[["10"]])
+  filter(external_gene_name %in% Wang_ours@IntersectionSets[["01"]])
+
+# 12 genes lost because no info like before
 
@@
          filter(Wang_lost,
-                testis_specificity != "testis_specific")$external_gene_name,
+                testis_specificity != "testis_specific" &
+                  CT_gene_type == "other")$external_gene_name,
        `Not tumour activated` = 
@@
 ```{r, eval = FALSE, echo = FALSE}
-pdf(file = paste0("figs/Venn_Wang.pdf"))
+pdf(file = paste0("figs/Wang_comparison.pdf"))
 plot(Wang_ours, gp = gp)
-dev.off()
-
-pdf(file = paste0("figs/upset_Wang_lost.pdf"))
 upset(upset_Wang,
@@
 Carter_lost <- all_genes %>%
-  filter(external_gene_name %in% Carter_ours@IntersectionSets[["10"]])
+  filter(external_gene_name %in% Carter_ours@IntersectionSets[["01"]])
+
+# 1 lost because no info, like before
 
@@
          filter(Carter_lost,
-                testis_specificity != "testis_specific")$external_gene_name,
+                testis_specificity != "testis_specific" &
+                  CT_gene_type == "other")$external_gene_name,
        `Not tumour activated` = 
@@
 ```{r, eval = FALSE, echo = FALSE}
-pdf(file = paste0("figs/Venn_Carter.pdf"))
+pdf(file = paste0("figs/Carter_comparison.pdf"))
 plot(Carter_ours, gp = gp)
-dev.off()
-
-pdf(file = paste0("figs/upset_Carter_lost.pdf"))
 upset(upset_Carter,
@@
 Bruggeman_lost <- all_genes %>%
-  filter(external_gene_name %in% Bruggeman_ours@IntersectionSets[["10"]])
+  filter(external_gene_name %in% Bruggeman_ours@IntersectionSets[["01"]])
+
+# 38 lost, like before
 
@@
          filter(Bruggeman_lost,
-                testis_specificity != "testis_specific")$external_gene_name,
+                testis_specificity != "testis_specific" &
+                  CT_gene_type == "other")$external_gene_name,
        `Not tumour activated` = 
@@
 ```{r, eval = FALSE, echo = FALSE}
-pdf(file = paste0("figs/Venn_Bruggeman.pdf"))
+pdf(file = paste0("figs/Bruggeman_comparison.pdf"))
 plot(Bruggeman_ours, gp = gp)
-dev.off()
-
-pdf(file = paste0("figs/upset_Bruggeman_lost.pdf"))
 upset(upset_Bruggeman,
@@
 ```{r all_databases}
+all_database_gene_list <- list(CTexploreR = CT_genes$external_gene_name,
+                               Carter = Carter_CT$Gene_Name,
+                               Jamin = Jamin_core_CT$Gene,
+                               CTatlas = Wang_CT$external_gene_name,
+                               Bruggeman = Bruggeman_CT$Gene_Name,
+                               CTdatabase = CTdatabase_cleaned$external_gene_name)
+
+upset_all_database <- fromList(all_database_gene_list)
+upset(upset_all_database,
+      nsets = 6,
+      nintersects = 50,
+      text.scale = upset_text_size)
+
 common <- unique(c(core_ours@IntersectionSets[["11"]], 
@@
 
-lost_list <- unique(c(core_ours@IntersectionSets[["10"]],
-                      CTdatabase_ours@IntersectionSets[["10"]],
-                      Wang_ours@IntersectionSets[["10"]],
-                      Carter_ours@IntersectionSets[["10"]],
-                      Bruggeman_ours@IntersectionSets[["10"]]))
+lost_list <- unique(c(core_ours@IntersectionSets[["01"]],
+                      CTdatabase_ours@IntersectionSets[["01"]],
+                      Wang_ours@IntersectionSets[["01"]],
+                      Carter_ours@IntersectionSets[["01"]],
+                      Bruggeman_ours@IntersectionSets[["01"]]))
 
@@
          filter(lost,
-                testis_specificity != "testis_specific")$external_gene_name,
+                testis_specificity != "testis_specific" &
+                  CT_gene_type == "other")$external_gene_name,
        `Not tumour activated` = 
@@
 dev.off()
+
+pdf(file = paste0("figs/upset_all_databases.pdf"))
+upset(upset_all_database,
+      nsets = 6,
+      nintersects = 50)
+dev.off()
 ```

### 2024-05-16 Paper_code-CT_analysis.Rmd
Added all of Axelle's figures

@@
 library(Biostrings)
+library(AUCell)
 ```
@@
 load("../CTdata/eh_data/all_genes.rda")
+load("../CTdata/eh_data/GTEX_data.rda")
 load("../CTdata/eh_data/CCLE_data.rda")
 load("../CTdata/eh_data/TCGA_TPM.rda")
+load("../CTdata/eh_data/normal_tissues_multimapping_data.rda")
+load("../CTdata/eh_data/mean_methylation_in_tissues.rda")
+load("../CTdata/eh_data/DAC_treated_cells_multimapping.rda")
+load("../CTdata/eh_data/DAC_treated_cells.rda")
 load("../CTdata/eh_data/testis_sce.rda")
-load("../CTdata/eh_data/mean_methylation_in_tissues.rda")
+load("../CTdata/eh_data/FGC_sce.rda")
+load("../CTdata/eh_data/oocytes_sce.rda")
+
 load("../CTdata/eh_data/CT_genes.rda")
 
-CT_and_CTP_genes <- CT_genes
-CT_genes <- (filter(CT_genes, CT_gene_type == "CT_gene"))
+CT_genes_types <- CT_genes %>% 
+  dplyr::rename(gene = external_gene_name) %>% 
+  mutate(Regulation = case_when(regulated_by_methylation ~ "Methylation",
+                                !regulated_by_methylation ~ "Not methylation")) %>%
+  mutate(Chr = case_when(X_linked ~ "X-linked",
+                         !X_linked ~ "not X")) %>% 
+  mutate(chr_met = case_when(regulated_by_methylation & X_linked ~ "X_met",
+                             !regulated_by_methylation & X_linked ~ "X_not_met",
+                             regulated_by_methylation & !X_linked ~ "not_X_met",
+                             !regulated_by_methylation & !X_linked ~ "not_X_not_met")) %>% 
+  mutate(specificity = case_when(GTEX_category == "testis_specific" ~ "Testis specific",
+                                        GTEX_category == "testis_preferential" ~ "Testis preferential",
+                                        GTEX_category == "lowly_expressed" ~ "Undetectable in GTEX")) %>% 
+  mutate(specificity = factor(specificity, 
+                                     levels = c("Testis specific",  "Undetectable in GTEX", "Testis preferential"))) %>% 
+  mutate(chr_met = factor(chr_met, 
+                          levels = c("X_met", "not_X_met", "X_not_met", "not_X_not_met")))
+
 ```
@@
 legends_param <- list(
-  labels_gp = gpar(col = "black", fontsize = 6),
-  title_gp = gpar(col = "black", fontsize = 6),
-  row_names_gp = gpar(fontsize = 4),
-  annotation_name_side = "left")
+  labels_gp = gpar(col = "black", fontsize = 4),
+  title_gp = gpar(col = "black", fontsize = 5, fontface = "bold"),
+  simple_anno_size = unit(0.2, "cm"),
+  row_names_gp = gpar(fontsize = 3),
+  annotation_name_side = "left",
+  border = FALSE,
+  border_gp = gpar(lwd = 0.2),
+  grid_width = unit(0.2, "cm"),
+  grid_height = unit(0.05, "cm"),
+  legend_height = unit(1, "cm"),
+  use_raster = TRUE,
+               raster_device = "CairoPNG",
+               raster_quality = 10)
 
@@
                    "#9E0142")
-chr_colors <- c("X-linked" = "deeppink", "Not X" = "royalblue1")
-
-meth_colors <- c("Methylation" = "lightgreen", "Not methylation" = "gray")
-```
-
-# Pipeline figure values
+
+TCGA_colors <- c(
+    "BRCA" = "midnightblue", "COAD" = "darkorchid2",
+    "ESCA" = "gold", "HNSC" = "deeppink2",
+    "LUAD" = "seagreen", "LUSC" = "seagreen3",
+    "SKCM" = "red3")
+
+CCLE_colors <- c(
+    "Lung" = "seagreen3", "Skin" = "red3",
+    "Bile_Duct" = "mediumpurple1", "Bladder" = "mistyrose2",
+    "Colorectal" = "plum", "Lymphoma" = "steelblue1",
+    "Uterine" = "darkorange4", "Myeloma" = "turquoise3",
+    "Kidney" = "thistle4",
+    "Pancreatic" = "darkmagenta", "Brain" = "palegreen2",
+    "Gastric" = "wheat3", "Breast" = "midnightblue",
+    "Bone" = "sienna1", "Head_and_Neck" = "deeppink2",
+    "Ovarian" = "tan3", "Sarcoma" = "lightcoral",
+    "Leukemia" = "steelblue4", "Esophageal" = "khaki",
+    "Neuroblastoma" = "olivedrab1")
+
+DAC_colors <- c(
+    "B2-1" = "olivedrab2", "HCT116" = "lightcoral",
+    "HEK293T" = "seagreen3", "HMLER" = "mediumpurple1",
+    "IMR5-75" = "deeppink2", "NCH1681" = "steelblue2",
+    "NCH612" = "red3", "TS603" = "darkmagenta")
+
+Regulation_colors <- c("Methylation" = "indianred2", 
+                       "Not methylation"= "cyan4")
+
+chr_met_colors = c("X_met" = "mediumorchid2",
+                   "X_not_met" = "mediumseagreen",
+                   "not_X_met"= "darkorange1",
+                   "not_X_not_met" = "dodgerblue1")
+
+Chr_colors = c("X-linked" = "deeppink", "not X" = "lightgreen")
+
+Testis_specificity_colors = c("Testis specific" = "mediumpurple2",
+                              "Testis preferential" = "goldenrod1",
+                              "Undetectable in GTEX" = "gray")
+
+testis_colors <- c(
+    "SSC" = "floralwhite", "Spermatogonia" = "moccasin",
+    "Early_spermatocyte" = "gold",
+    "Late_spermatocyte" = "orange",
+    "Round_spermatid" = "red2",
+    "Elongated_spermatid" = "darkred", "Sperm" = "purple",
+    "Sertoli" = "gray",
+    "Leydig" = "cadetblue2", "Myoid" = "springgreen3",
+    "Macrophage" = "gray10",
+    "Endothelial" = "steelblue")
+
+Fetal_cell_colors = c( "F_PGC" = "pink", 
+                       "F_GC" = "pink4", 
+                       "F_oogonia" = "palevioletred3",
+                       "F_oocyte" = "mediumorchid4", 
+                       "M_PGC" = "lightblue1", 
+                       "M_GC" = "steelblue3",
+                       "M_pre_spermatogonia" = "darkblue")
+
+oocytes_colors <- c("Growing oocytes" = "paleturquoise3", 
+                  # "Fully grown oocytes" ="lightgreen", 
+                  # "Metaphase I" = "mediumvioletred", 
+                  "Metaphase II" = "mediumvioletred")
+
+sex_colors <- c("F" = "palevioletred1", "M"= "skyblue2")
+
+stage_colors <- c( "pre-meiotic" = "moccasin", 
+                   "meiotic"= "lightsalmon",
+                   "post-meiotic" = "maroon4",
+                   "somatic" = "gray")
+
+```
+
+# CT selection
+
+## Pipeline figure values
 
@@
 
-# CT genes characteristics
-
-```{r basic}
-table(CT_genes$testis_specificity)
-table(CT_genes$transcript_biotype)
-```
-
-Most genes are testis specific 
-(`r table(CT_genes$testis_specificity)[["testis_specific"]]/dim(CT_genes)[1]*100`%).
-Most genes are mainly protein coding genes 
-(`r table(CT_genes$transcript_biotype)["protein_coding"]/dim(CT_genes)[1]*100`%).
-
-## X chromosome and regulated by methylation
-
-In CTexploreR, genes have been characterised as regulated by methylation or not.
-
-```{r X_meth}
-table(CT_genes$X_linked)
-table(CT_genes$regulated_by_methylation)
-table(CT_genes$X_linked, CT_genes$testis_specificity)
-table(CT_genes$X_linked, CT_genes$regulated_by_methylation)
+## GTEX-TCGA-CCLE
+
+Illustrate the selection using heatmaps with GTEx data, TCGA and CCLE
+
+Function to print GTEX expression with specific annotations.
+```{r GTEX_function}
+print_GTEX_heatmap <- function(genes = NULL, units = "log_TPM", 
+                               row_fontsize = 2,
+                               column_fontsize = 6, 
+                               h_width = 3,
+                               h_height = 10,
+                               split_rows_by = "regulation") {
+
+  suppressMessages({
+    database <- GTEX_data
+  })
+
+  database <- database[rowData(database)$external_gene_name %in% genes, ]
+
+  mat <- assay(database)
+  rownames(mat) <- rowData(database)$external_gene_name
+
+  if (units == "log_TPM") mat <- log1p(mat)
+
+  gene_type <- CT_genes_types %>% 
+    filter(gene %in% genes)
+
+   if (split_rows_by == "regulation") {
+    gene_type <- gene_type %>% arrange(Regulation)
+    rowSplit <- gene_type$Regulation
+  }
+  if (split_rows_by == "testis_specificity") {
+    gene_type <- gene_type %>% arrange(Testis_specificity)
+    rowSplit <- gene_type$Testis_specificity
+  }
+  if (split_rows_by == "chr_met") {
+    gene_type <- gene_type %>% arrange(chr_met)
+    rowSplit <- gene_type$chr_met
+  }
+  if (split_rows_by == "specificity") {
+    gene_type <- gene_type %>% arrange(specificity)
+    rowSplit <- gene_type$specificity
+  }
+
+  if (split_rows_by == "GTEX_category") {
+    GTEX_category <- gene_type %>% arrange(GTEX_category)
+    rowSplit <- gene_type$GTEX_category
+  }
+
+  Regulation <- as.matrix(gene_type$Regulation)
+  rownames(Regulation) <- gene_type$gene
+  row_ha_regulation <- rowAnnotation(Regulation = Regulation,
+                                     annotation_legend_param = legends_param,
+                                     simple_anno_size = unit(0.2, "cm"),
+                                     col = list(Regulation = Regulation_colors),
+                                     annotation_name_gp = gpar(fontsize = 0))
+
+  chr_met <- as.matrix(gene_type$chr_met)
+  rownames(chr_met) <- gene_type$gene
+  row_ha_chr_met <- rowAnnotation(chr_met = chr_met,
+                                  annotation_legend_param = legends_param,
+                                  simple_anno_size = unit(0.2, "cm"),
+                                  col = list(chr_met = chr_met_colors),
+                                  annotation_name_gp = gpar(fontsize = 0))
+
+  Chr <- as.matrix(gene_type$Chr)
+  rownames(Chr) <- gene_type$gene
+  row_ha_chr <- rowAnnotation(Chr = Chr,
+                              annotation_legend_param = legends_param,
+                              simple_anno_size = unit(0.2, "cm"),
+                              col = list(Chr = Chr_colors),
+                              annotation_name_gp = gpar(fontsize = 0))
+
+  specificity <- as.matrix(gene_type$specificity)
+  rownames(specificity) <- gene_type$gene
+  row_ha_ts <- rowAnnotation(specificity = specificity,
+                             annotation_legend_param = legends_param,
+                             simple_anno_size = unit(0.2, "cm"),
+                             col = list(specificity = 
+                                          Testis_specificity_colors),
+                             annotation_name_gp = gpar(fontsize = 0))
+
+  heatmap_colors <- colorRamp2(
+    seq(0, 5, length = 11),                 
+    legend_colors)
+
+  h <- Heatmap(mat[gene_type$gene, ],
+               name = "log(TPM +1)",
+               column_title = "Normal tissues (GTEx)",
+               column_title_gp = gpar(fontsize = 6, face = "bold"),
+               row_split = rowSplit,
+               cluster_row_slices = FALSE,
+               row_title_gp = gpar(fontsize = 4, face = "bold"),
+               border = TRUE,
+               border_gp = gpar(lwd = 0.5),
+               col = heatmap_colors,
+               cluster_rows = TRUE,
+               cluster_columns = TRUE,
+               show_row_dend = FALSE,
+               show_column_dend = FALSE,
+               row_names_gp = gpar(fontsize = row_fontsize),
+               column_names_gp = gpar(fontsize = column_fontsize),
+               column_names_side = "top",
+               # right_annotation = c(row_ha_regulation, 
+               #                        row_ha_chr, 
+               #                        row_ha_ts,
+               #                        gap = unit(1, "mm")),
+               row_names_side = "left",
+               width = unit(h_width, "cm"),
+               height = unit(h_height, "cm"),
+               clustering_method_rows = "ward.D",
+               use_raster = TRUE,
+               raster_device = "CairoPNG",
+               raster_quality = 10,
+               heatmap_legend_param = legends_param)
+
+  return(h)
+}
+```
+
+Function to print TCGA expression with specific annotations and to downsample 
+the number of tumor samples per tumor type.
+
+```{r TCGA_function}
+print_TCGA_heatmap <- function(genes = NULL,
+                               tum_type = c("TCGA-SKCM", "TCGA-LUAD", 
+                                            "TCGA-LUSC", "TCGA-COAD", 
+                                            "TCGA-BRCA", "TCGA-HNSC",
+                                            "TCGA-ESCA"),
+                               downsample_to_n = 450,
+                               fontsize = 4, 
+                               clust_method = "ward.D",
+                               split_rows_by = "regulation", 
+                               h_width = 5, 
+                               h_height = 10,
+                               clustRow = TRUE,
+                               clustCol = TRUE) {
+
+  tcga <- TCGA_TPM
+  tcga <- tcga[,tcga$shortLetterCode != "NT"]
+
+  ## downsample tumors to have the same number of tumors in each type
+  TCGA <- tcga[, tcga$project_id == tum_type[1]]
+  TCGA <- TCGA[, 1:min(ncol(TCGA), downsample_to_n)]
+
+  for (tum in tum_type[-1]){
+    tmp <- tcga[, tcga$project_id == tum]
+    tmp <- tmp[, 1:min(ncol(tmp), downsample_to_n)]
+    TCGA <- cbind(TCGA, tmp)
+  }
+
+  TCGA$project_id <- gsub(pattern = "TCGA-", x = TCGA$project_id, 
+                          replacement = '')
+
+  split_by <- factor(TCGA$project_id)
+
+  column_ha_tumor <- HeatmapAnnotation(
+    TCGA_tumor = TCGA$project_id,
+    border = TRUE,
+    col = list(TCGA_tumor = TCGA_colors),
+    annotation_label = gpar(fontsize = 4),
+    annotation_name_gp = gpar(fontsize = 0),
+    simple_anno_size = unit(0.3, "cm"),
+    annotation_legend_param = legends_param)
+
+  ## Use gene names instead of ENSEMBL IDs
+  mat <- SummarizedExperiment::assay(TCGA)
+  rownames(mat) <- rowData(TCGA)$external_gene_name
+  mat <- mat[CT_genes$external_gene_name,]
+  mat <- log1p(mat)
+
+  gene_type <- CT_genes_types %>% 
+    filter(gene %in% genes)
+
+  rowSplit <- NULL
+  if(!is.null(split_rows_by)){
+   if (split_rows_by == "regulation") {
+    gene_type <- gene_type %>% arrange(Regulation)
+    rowSplit <- gene_type$Regulation
+  }
+  if (split_rows_by == "testis_specificity") {
+    gene_type <- gene_type %>% arrange(Testis_specificity)
+    rowSplit <- gene_type$testis_specificity
+  }
+  if (split_rows_by == "specificity") {
+    gene_type <- gene_type %>% arrange(specificity)
+    rowSplit <- gene_type$specificity
+  }
+
+  if (split_rows_by == "chr_met") {
+    gene_type <- gene_type %>% arrange(chr_met)
+    rowSplit <- gene_type$chr_met
+  }
+
+  if (split_rows_by == "GTEX_category") {
+    GTEX_category <- gene_type %>% arrange(GTEX_category)
+    rowSplit <- gene_type$GTEX_category
+  }
+  }
+
+  Regulation <- as.matrix(gene_type$Regulation)
+  rownames(Regulation) <- gene_type$gene
+  row_ha_regulation <- rowAnnotation(Regulation = Regulation,
+                                     annotation_legend_param = legends_param,
+                                     simple_anno_size = unit(0.2, "cm"),
+                                     col = list(Regulation = Regulation_colors),
+                                     annotation_name_gp = gpar(fontsize = 0))
+
+  chr_met <- as.matrix(gene_type$chr_met)
+  rownames(chr_met) <- gene_type$gene
+  row_ha_chr_met <- rowAnnotation(chr_met = chr_met,
+                                  annotation_legend_param = legends_param,
+                                  simple_anno_size = unit(0.2, "cm"),
+                                  col = list(chr_met = chr_met_colors),
+                                  annotation_name_gp = gpar(fontsize = 0))
+
+  Chr <- as.matrix(gene_type$Chr)
+  rownames(Chr) <- gene_type$gene
+  row_ha_chr <- rowAnnotation(Chr = Chr,
+                              annotation_legend_param = legends_param,
+                              simple_anno_size = unit(0.2, "cm"),
+                              col = list(Chr = Chr_colors),
+                              annotation_name_gp = gpar(fontsize = 0))
+
+  specificity <- as.matrix(gene_type$specificity)
+  rownames(specificity) <- gene_type$gene
+  row_ha_ts <- rowAnnotation(specificity = specificity,
+                             annotation_legend_param = legends_param,
+                             simple_anno_size = unit(0.2, "cm"),
+                             col = list(specificity = 
+                                          Testis_specificity_colors),
+                             annotation_name_gp = gpar(fontsize = 0))
+
+  # same range as in GTEX
+  heatmap_colors <- colorRamp2(
+    seq(0, 5, length = 11),                 
+    legend_colors)
+
+  Heatmap(mat[gene_type$gene, ],
+          name = "TCGA (logTPM)",
+          column_title = "TCGA tumor samples",
+          column_title_gp = gpar(fontsize = 6, face = "bold"),
+          clustering_method_rows = clust_method,
+          clustering_method_columns = clust_method,
+          column_split = split_by,
+          row_gap = unit(c(1), "mm"),
+          col = heatmap_colors,
+          cluster_rows = clustRow,
+          cluster_columns = clustCol,
+          show_row_names = FALSE,
+          show_column_names = FALSE,
+          show_row_dend = FALSE,
+          show_column_dend = FALSE,
+          column_names_gp = gpar(fontsize = 6),
+          column_names_centered = TRUE,
+          row_names_gp = gpar(fontsize = 0),
+          row_names_side = "left",
+          border = TRUE,
+          border_gp = gpar(lwd = 0.5),
+          column_names_side = c("top"),
+          column_names_rot = 90,
+          row_title_gp = gpar(fontsize = 0),
+          top_annotation = column_ha_tumor,
+          right_annotation = c(row_ha_ts,
+                               #row_ha_regulation,
+                               #row_ha_chr,
+                               gap = unit(1, "mm")),
+          row_split = rowSplit,
+          cluster_row_slices = FALSE,
+          width = unit(h_width, "cm"),
+          height = unit(h_height, "cm"),
+          use_raster = TRUE,
+          raster_device = "CairoPNG",
+          raster_quality = 10,
+          heatmap_legend_param = legends_param,
+          show_heatmap_legend = FALSE)
+}
+
+```
+
+Function to print CCLE expression with specific annotations. Allows to 
+downsample the number of tumor samples pet tumor type to show
+
+```{r CCLE_function}
+print_CCLE_heatmap <- function(genes = NULL,
+                               tum_type = c("Skin", "Lung", 
+                                            "Colorectal", 
+                                            "Breast", 
+                                            "Head_and_Neck",
+                                            "Esophageal"),
+                               h_width = 5,
+                               h_height = 10,
+                               downsample_to_n = 450,
+                               fontsize = 4, 
+                               clust_method = "ward.D",
+                               split_rows_by = "regulation", 
+                               clustRow = TRUE,
+                               clustCol = TRUE){
+
+  ## downsample tumors to have the same number of tumors in each type
+  ccle <- CCLE_data
+  CCLE <- ccle[, ccle$type == tum_type[1]]
+  CCLE <- CCLE[, 1:min(ncol(CCLE), downsample_to_n)]
+
+  for (tum in tum_type[-1]){
+    tmp <- ccle[, ccle$type == tum]
+    tmp <- tmp[, 1:min(ncol(tmp), downsample_to_n)]
+    CCLE <- cbind(CCLE, tmp)
+  }
+
+  split_by <- factor(CCLE$type)
+
+  column_ha_tumor <- HeatmapAnnotation(
+    CCLE_cell_type = CCLE$type,
+    border = TRUE,
+    col = list(CCLE_cell_type = CCLE_colors),
+    annotation_label = gpar(fontsize = 4),
+    annotation_name_gp = gpar(fontsize = 0),
+    simple_anno_size = unit(0.3, "cm"),
+    annotation_legend_param = legends_param)
+
+  ## Use gene names instead of ENSEMBL IDs
+  mat <- SummarizedExperiment::assay(CCLE)
+  rownames(mat) <- rowData(CCLE)$external_gene_name
+  mat <- mat[CT_genes$external_gene_name,]
+  mat <- log1p(mat)
+
+  gene_type <- CT_genes_types %>% 
+    filter(gene %in% genes)
+
+   if (split_rows_by == "regulation") {
+    gene_type <- gene_type %>% arrange(Regulation)
+    rowSplit <- gene_type$Regulation
+  }
+  if (split_rows_by == "testis_specificity") {
+    gene_type <- gene_type %>% arrange(testis_specificity)
+    rowSplit <- gene_type$testis_specificity
+  }
+  if (split_rows_by == "specificity") {
+    gene_type <- gene_type %>% arrange(specificity)
+    rowSplit <- gene_type$specificity
+  }
+  if (split_rows_by == "chr_met") {
+    gene_type <- gene_type %>% arrange(chr_met)
+    rowSplit <- gene_type$chr_met
+  }
+
+   if (split_rows_by == "GTEX_category") {
+    GTEX_category <- gene_type %>% arrange(GTEX_category)
+    rowSplit <- gene_type$GTEX_category
+  }
+
+  Regulation <- as.matrix(gene_type$Regulation)
+  rownames(Regulation) <- gene_type$gene
+  row_ha_regulation <- rowAnnotation(Regulation = Regulation,
+                                     annotation_legend_param = legends_param,
+                                     simple_anno_size = unit(0.2, "cm"),
+                                     col = list(Regulation = Regulation_colors),
+                                     annotation_name_gp = gpar(fontsize = 0))
+
+  chr_met <- as.matrix(gene_type$chr_met)
+  rownames(chr_met) <- gene_type$gene
+  row_ha_chr_met <- rowAnnotation(chr_met = chr_met,
+                                  annotation_legend_param = legends_param,
+                                  simple_anno_size = unit(0.2, "cm"),
+                                  col = list(chr_met = chr_met_colors),
+                                  annotation_name_gp = gpar(fontsize = 0))
+
+  Chr <- as.matrix(gene_type$Chr)
+  rownames(Chr) <- gene_type$gene
+  row_ha_chr <- rowAnnotation(Chr = Chr,
+                              annotation_legend_param = legends_param,
+                              simple_anno_size = unit(0.2, "cm"),
+                              col = list(Chr = Chr_colors),
+                              annotation_name_gp = gpar(fontsize = 0))
+
+  specificity <- as.matrix(gene_type$specificity)
+  rownames(specificity) <- gene_type$gene
+  row_ha_ts <- rowAnnotation(specificity = specificity,
+                                           annotation_legend_param = legends_param,
+                                           simple_anno_size = unit(0.2, "cm"),
+                                           col = list(specificity = 
+                                                        Testis_specificity_colors),
+                                           annotation_name_gp = gpar(fontsize = 0))
+
+  # heatmap_colors <- colorRamp2(
+  #   seq(0, quantile(rowMax(mat), 0.8), length = 11),                 
+  #   legend_colors)
+
+  # Same range as GTEX and TCGA
+  heatmap_colors <- colorRamp2(
+    seq(0, 5, length = 11),                 
+    legend_colors)
+
+  Heatmap(mat[gene_type$gene, ],
+          name = "CCLE (logTPM)",
+          column_title = paste0("CCLE tumor cell lines"),
+          column_title_gp = gpar(fontsize = 6, face = "bold"),
+          clustering_method_rows = clust_method,
+          clustering_method_columns = clust_method,
+          column_split = split_by,
+          row_gap = unit(c(1), "mm"),
+          col = heatmap_colors,
+          cluster_rows = clustRow,
+          cluster_columns = clustCol,
+          show_row_names = TRUE,
+          show_column_names = FALSE,
+          show_row_dend = FALSE,
+          show_column_dend = FALSE,
+          column_names_gp = gpar(fontsize = 6),
+          column_names_centered = TRUE,
+          row_names_gp = gpar(fontsize = fontsize),
+          row_names_side = "left",
+          border = TRUE,
+          border_gp = gpar(lwd = 0.5),
+          column_names_side = c("top"),
+          column_names_rot = 90,
+          row_title_gp = gpar(fontsize = 0),
+          top_annotation = column_ha_tumor,
+          # right_annotation = c(row_ha_ts,
+          #                      row_ha_regulation,
+          #                      #row_ha_chr,
+          #                      gap = unit(1, "mm")),
+          row_split = rowSplit,
+          width = unit(h_width, "cm"),
+          height = unit(h_width, "cm"),
+          use_raster = TRUE,
+          raster_device = "CairoPNG",
+          raster_quality = 10,
+          heatmap_legend_param = legends_param,
+          show_heatmap_legend = FALSE)
+}
+```
+
+Show CT genes in GTEX, CCLE and TCGA (downsampling the number of samples by 
+tumor type).
+
+```{r selection_heatmap}
+my_genes <- CT_genes %>% 
+  pull(external_gene_name)
+
+ht_TCGA <- print_TCGA_heatmap(my_genes, 
+                              fontsize = 4, 
+                              downsample_to_n = 50, 
+                              h_width = 3,
+                              h_height = 8,
+                              tum_type = c("TCGA-SKCM", "TCGA-LUAD", 
+                                           #"TCGA-ESCA", "TCGA-LUSC", 
+                                           #"TCGA-COAD",
+                                           #"TCGA-BRCA",
+                                           "TCGA-HNSC"),
+                              clust_method = "ward.D",
+                              split_rows_by = "specificity")
+
+ht_CCLE <- print_CCLE_heatmap(my_genes, 
+                              downsample_to_n = 50, 
+                              tum_type = c("Skin", "Lung", 
+                                           #"colorectal", 
+                                           #"breast", 
+                                           #"esophageal",
+                                           "Head_and_Neck"),
+                              #"Leukemia", "Lymphoma"),
+                              h_width = 3,
+                              h_height = 8,
+                              split_rows_by = "specificity")
+
+ht_GTEX <- print_GTEX_heatmap(my_genes, 
+                              row_fontsize = 1,
+                              column_fontsize = 5,
+                              h_width = 5,
+                              h_height = 8,
+                             split_rows_by = "specificity")
+h_list <- ht_GTEX + ht_CCLE + ht_TCGA
+draw(h_list, main_heatmap = "TCGA (logTPM)", ht_gap = unit(c(0.3, 0.01), "cm"),
+     merge_legend = TRUE)
+```
+
+```{r, eval = FALSE, echo=FALSE}
+pdf(file = paste0("figs/GTEX_Tumors.pdf"))
+draw(h_list, main_heatmap = "TCGA (logTPM)", 
+     ht_gap = unit(c(0.3, 0.01), "cm"),
+     merge_legend = TRUE)
+dev.off()
+```
+
+## Multimapping Fig
+
+Figure showing how we kept the lowly_expressed in GTEx
+
+```{r multimapping}
+my_genes <- CT_genes %>% 
+  filter(GTEX_category == 'lowly_expressed') %>% 
+  filter(testis_specificity == "testis_specific")
+
+without_MP <- log1p(
+  assay(normal_tissues_multimapping_data[my_genes$ensembl_gene_id,], 
+    "TPM_no_multimapping"))
+with_MP <-  log1p(
+  assay(normal_tissues_multimapping_data[my_genes$ensembl_gene_id,], 
+    "TPM_with_multimapping"))
+rownames(without_MP) <- rownames(with_MP) <- 
+  rowData(normal_tissues_multimapping_data)[my_genes$ensembl_gene_id,]$external_gene_name
+
+# Fix same range of colors for both heatmaps
+full_mat <- cbind(without_MP, with_MP)
+heatmap_colors <- colorRamp2(
+  seq(0, quantile(rowMax(full_mat), 0.8), length = 11), legend_colors)
+
+# Or same range of colors than GTEX heatmap
+# heatmap_colors <- colorRamp2(
+#   seq(0, 6, length = 11), legend_colors)
+
+colnames(without_MP) <- gsub("_", x = str_to_sentence(colnames(without_MP)), ' ')
+colnames(with_MP) <- gsub("_", x = str_to_sentence(colnames(with_MP)), ' ')
+ordered_tissues <- 
+  c("Testis", colnames(without_MP)[- which(colnames(without_MP) == "Testis")])
+h_without_MP <- Heatmap(without_MP[my_genes$external_gene_name, ordered_tissues],
+                        name = "no MP",
+                        column_title = "Multimapped reads are excluded",
+                        column_title_gp = gpar(fontsize = 10, face = "bold"),
+                        border = TRUE,
+                        border_gp = gpar(lwd = 0.5),
+                        col = heatmap_colors,
+                        cluster_rows = TRUE,
+                        cluster_columns = FALSE,
+                        show_row_dend = FALSE,
+                        show_column_dend = FALSE,
+                        row_names_gp = gpar(fontsize = 4),
+                        column_names_gp = gpar(fontsize = 6),
+                        column_names_side = "top",
+                        row_names_side = "left",
+                        heatmap_width = unit(8, "cm"),
+                        heatmap_height = unit(12, "cm"),
+                        clustering_method_rows = "ward.D",
+                        use_raster = TRUE,
+                        raster_device = "CairoPNG",
+                        raster_quality = 10,
+                        show_heatmap_legend = FALSE,
+                        heatmap_legend_param = legends_param)
+
+h_with_MP <- Heatmap(with_MP[my_genes$external_gene_name, ordered_tissues],
+                     name = "log(TPM)",
+                     column_title = "Multimapped reads are counted",
+                     column_title_gp = gpar(fontsize = 10, face = "bold"),
+                     border = TRUE,
+                     border_gp = gpar(lwd = 0.5),
+                     col = heatmap_colors,
+                     cluster_rows = TRUE,
+                     cluster_columns = FALSE,
+                     show_row_dend = FALSE,
+                     show_column_dend = FALSE,
+                     row_names_gp = gpar(fontsize = 4),
+                     column_names_gp = gpar(fontsize = 6),
+                     column_names_side = "top",
+                     row_names_side = "left",
+                     heatmap_width = unit(8, "cm"),
+                     heatmap_height = unit(12, "cm"),
+                     clustering_method_rows = "ward.D",
+                     use_raster = TRUE,
+                     raster_device = "CairoPNG",
+                     raster_quality = 10,
+                     heatmap_legend_param = legends_param)
+
+h_list <- h_without_MP + h_with_MP
+draw(h_list, main_heatmap = "log(TPM)", ht_gap = unit(0.3, "cm"),
+                 height = unit(5, "cm"))
+```
+
+```{r, eval = FALSE, echo = FALSE}
+pdf(file = paste0("figs/Multimapping.pdf"))
+draw(h_list, main_heatmap = "log(TPM)", ht_gap = unit(0.3, "cm"),
+                 height = unit(5, "cm"))
+dev.off()
+```
+
+## IGV
+
+Shows why we removed illegitimate transcripts
+
+```{r, echo=FALSE, fig.align='left', out.width = '100%', purl = TRUE}
+knitr::include_graphics("figs/Fig_IGV_no_clear_transcript.png")
+```
+
+# CT_genes specific
+
+From now on, we'll focus on CT genes that are specific
+
+```{r}
+CT_genes <- CT_genes %>%
+  filter(CT_gene_type == "CT_gene") 
+CT_genes_types <- CT_genes_types %>% 
+  filter(CT_gene_type == "CT_gene")
+```
+
+# Methylation analysis
+
+## Regulation by methylation
+
+```{r meth_value_heatmap}
+met_genes <- CT_genes %>% 
+  filter(regulated_by_methylation) %>% 
+  pull(external_gene_name)
+met_value_number_per_gene <- tibble(
+  gene = CT_genes$external_gene_name,
+  regulated_by_methylation = CT_genes$regulated_by_methylation,
+  NA_met_values_number = rowSums(
+    is.na(assay(mean_methylation_in_tissues[CT_genes$external_gene_name, ]))))
+n_met_values <- ncol(mean_methylation_in_tissues)
+CT_met_with_WGBS <- met_value_number_per_gene %>%
+  filter(regulated_by_methylation) %>% 
+  filter(NA_met_values_number !=  n_met_values)
+
+CT_met_without_WGBS <- met_value_number_per_gene %>%
+  filter(regulated_by_methylation) %>% 
+  filter(NA_met_values_number == n_met_values )
+
+CT_not_met_with_WGBS <- met_value_number_per_gene %>%
+  filter(!regulated_by_methylation) %>% 
+  filter(NA_met_values_number !=  n_met_values)
+
+CT_not_met_without_WGBS <- met_value_number_per_gene %>%
+  filter(!regulated_by_methylation) %>% 
+  filter(NA_met_values_number == n_met_values )
+
+mean_methylation_fig <- function(genes = NULL, 
+                                 values_only = FALSE,
+                                 row_fontsize = 4,
+                                 column_fontsize = 8, 
+                                 na.omit = FALSE) {
+
+  database <- mean_methylation_in_tissues
+  rowData(database)$external_gene_name <- rownames(database)
+  database <- database[genes, ]
+
+  if (na.omit) {
+    mat <- na.omit(assay(database))
+    clustering_option <- TRUE
+  } else {
+    mat <- assay(database)
+    clustering_option <- FALSE
+  }
+
+  gene_type <- CT_genes_types %>% 
+    filter(gene %in% genes) %>% 
+    mutate(group = case_when(gene %in% CT_met_with_WGBS$gene ~ "group1",
+                             gene %in% CT_met_without_WGBS$gene ~ "group2",
+                             gene %in% CT_not_met_with_WGBS$gene ~ "group3",
+                             gene %in% CT_not_met_without_WGBS$gene ~ "group4")) %>% 
+    left_join(CT_genes %>% dplyr::rename(gene = external_gene_name) %>% 
+                dplyr::select(gene, somatic_met_level, sperm_met_level, 
+                              germline_methylation, somatic_methylation)) %>% 
+    arrange(group, sperm_met_level)
+
+  # sort genes regulated by met differently than not regulated by met
+  gene_type_met <- gene_type %>%
+    filter(regulated_by_methylation)
+
+  gene_type_not_met <- gene_type %>%
+    filter(!regulated_by_methylation) %>%
+    arrange(germline_methylation, somatic_met_level)
+
+  gene_type <- rbind(gene_type_met, gene_type_not_met)
+  colnames(mat) <- str_to_sentence(colnames(mat))
+
+  h <- Heatmap(mat[gene_type$gene, sort(colnames(mat))],
+               column_title = "Promoter methylation",
+               name = "Mean \nmethylation",
+               column_title_gp = gpar(fontsize = 6),
+               border = TRUE,
+               border_gp = gpar(lwd = 0.5),
+               row_title_gp = gpar(col = c("indianred2", "cyan4"), fontsize = 6),
+               row_split = factor(gene_type$Regulation,
+                                  levels = c("Methylation", "Not methylation")),
+               #row_split = factor(gene_type$group),
+               cluster_row_slices = FALSE,
+               col = colorRamp2(seq_len(100),
+                                colorRampPalette(c("moccasin", "dodgerblue4"))(100)),
+               na_col = "gray80",
+               cluster_rows = FALSE,
+               cluster_columns = FALSE,
+               show_row_names = TRUE,
+               show_heatmap_legend = TRUE,
+               show_row_dend = FALSE,
+               column_names_gp = gpar(fontsize = column_fontsize),
+               row_names_gp = gpar(fontsize = row_fontsize),
+               row_names_side = "left",
+               heatmap_width = unit(4, "cm"),
+               heatmap_height = unit(10, "cm"),
+               use_raster = TRUE,
+               raster_device = "CairoPNG",
+               raster_quality = 8,
+               column_names_side = "top",
+               heatmap_legend_param = legends_param)
+
+  return(h)
+}
+```
+
+```{r DAC_heatmap}
+DAC_fig <- function(genes = gene_type$gene, multimapping = TRUE, 
+                    row_fontsize = 2,
+                    units = "log1p",
+                    values_only = FALSE, scale_lims = NULL) {
+
+  if (multimapping) {
+    # database <- CTdata::DAC_treated_cells_multimapping()
+    database <- DAC_treated_cells_multimapping
+
+  } else {
+    #database <- CTdata::DAC_treated_cells()
+    database <- DAC_treated_cells
+  }
+
+  database$treatment[database$treatment == "DAC"] <- "5-aza"
+  database$treatment <- factor(database$treatment, levels = c("CTL", "5-aza"))
+   gene_type <- CT_genes_types %>% 
+    filter(gene %in% genes) %>% 
+    mutate(group = case_when(gene %in% CT_met_with_WGBS$gene ~ "group1",
+                             gene %in% CT_met_without_WGBS$gene ~ "group2",
+                             gene %in% CT_not_met_with_WGBS$gene ~ "group3",
+                             gene %in% CT_not_met_without_WGBS$gene ~ "group4")) %>% 
+    left_join(CT_genes %>% dplyr::rename(gene = external_gene_name) %>% 
+                dplyr::select(gene, somatic_met_level, sperm_met_level, 
+                              germline_methylation, somatic_methylation)) %>% 
+    arrange(group, sperm_met_level)
+
+  # sort genes regulated by met differently than not regulated by met
+  gene_type_met <- gene_type %>%
+    filter(regulated_by_methylation)
+
+  gene_type_not_met <- gene_type %>%
+    filter(!regulated_by_methylation) %>%
+    arrange(germline_methylation, somatic_met_level)
+
+  gene_type <- rbind(gene_type_met, gene_type_not_met)
+  rownames(database) <- rowData(database)$external_gene_name
+
+  mat <- assay(database, units)
+
+  df_col <- data.frame("cell" = colData(database)$cell,
+                       "treatment" = colData(database)$treatment)
+  rownames(df_col) <- rownames(colData(database))
+  df_col <- df_col[order(df_col$cell, df_col$treatment), ]
+
+  column_ha_cell <- HeatmapAnnotation(
+    cell = df_col$cell,
+    border = FALSE, 
+    #border_gp = gpar(lwd = 0.1),
+    simple_anno_size = unit(0.2, "cm"),
+    annotation_legend_param = legends_param,
+    col = list("cell" = DAC_colors),
+    annotation_name_gp = gpar(fontsize = 0))
+
+  column_ha_treatment <- HeatmapAnnotation(
+    treatment = df_col$treatment,
+    col = list(treatment = c("CTL" = "royalblue4", "5-aza" = "maroon3")),
+    annotation_legend_param = legends_param,
+    border = FALSE, 
+    simple_anno_size = unit(0.2, "cm"),
+    annotation_name_gp = gpar(fontsize = 0))
+
+  Regulation <- as.matrix(gene_type$Regulation)
+  rownames(Regulation) <- gene_type$gene
+  row_ha_regulation <- rowAnnotation(Regulation = Regulation,
+                                     annotation_legend_param = legends_param,
+                                     simple_anno_size = unit(0.2, "cm"),
+                                     col = list(Regulation = Regulation_colors),
+                                     annotation_name_gp = gpar(fontsize = 0))
+
+  chr_met <- as.matrix(gene_type$chr_met)
+  rownames(chr_met) <- gene_type$gene
+  row_ha_chr_met <- rowAnnotation(chr_met = chr_met,
+                                  annotation_legend_param = legends_param,
+                                  simple_anno_size = unit(0.2, "cm"),
+                                  col = list(chr_met = chr_met_colors),
+                                  annotation_name_gp = gpar(fontsize = 0))
+
+  Chr <- as.matrix(gene_type$Chr)
+  rownames(Chr) <- gene_type$gene
+  row_ha_chr <- rowAnnotation(Chr = Chr,
+                              annotation_legend_param = legends_param,
+                              simple_anno_size = unit(0.2, "cm"),
+                              col = list(Chr = Chr_colors),
+                              annotation_name_gp = gpar(fontsize = 0))
+
+  specificity <- as.matrix(gene_type$specificity)
+  rownames(specificity) <- gene_type$gene
+  row_ha_ts <- rowAnnotation(specificity = specificity,
+                             annotation_legend_param = legends_param,
+                             simple_anno_size = unit(0.2, "cm"),
+                             col = list(specificity = 
+                                          Testis_specificity_colors),
+                             annotation_name_gp = gpar(fontsize = 0))
+  if (is.null(scale_lims)) scale_lims <- c(0, quantile(rowMaxs(mat), 0.75))
+
+  h <- Heatmap(mat[gene_type$gene, rownames(df_col), drop = FALSE],
+               name = "logTPM",
+               row_split = factor(gene_type$Regulation, 
+                                  levels = c("Methylation", "Not methylation")),
+               row_title_gp = gpar(col = Regulation_colors, fontsize = 6),
+               border = TRUE,
+               border_gp = gpar(lwd = 0.5),
+               column_title = "Expression in cells treated with 5-Aza",
+               column_title_gp = gpar(fontsize = 6),
+               column_split = factor(df_col$cell),
+               col = colorRamp2(seq(scale_lims[1], scale_lims[2], length = 11),
+                                legend_colors),
+               cluster_rows = TRUE,
+               show_row_dend = FALSE,
+               show_row_names = FALSE,
+               clustering_method_rows = "ward.D",
+               show_column_names = FALSE,
+               cluster_columns = FALSE,
+               row_names_gp = gpar(fontsize = row_fontsize),
+               top_annotation = c(column_ha_cell, column_ha_treatment),
+               # right_annotation = c(row_ha_ts,
+               #                      #row_ha_regulation,
+               #                      row_ha_chr,
+               #                      gap = unit(1, "mm")),
+               heatmap_width = unit(6, "cm"),
+               heatmap_height = unit(10, "cm"),
+               use_raster = TRUE,
+               raster_device = "CairoPNG",
+               raster_quality = 10,
+               heatmap_legend_param = legends_param)
+
+  return(h)
+}
+```
+
+```{r draw_meth_reg}
+fs <- 1.2
+h_met <- mean_methylation_fig(genes = CT_genes$external_gene_name,
+                              values_only = FALSE,
+                              row_fontsize = fs,
+                              column_fontsize = 5,
+                              na.omit = FALSE)
+
+h_DAC <- DAC_fig(CT_genes$external_gene_name, 
+                 multimapping = TRUE, 
+                 units = "log1p",
+                 row_fontsize = fs
+)
+
+h_list <- h_met + h_DAC 
+draw(h_list, main_heatmap = "Mean \nmethylation", ht_gap = unit(0.5, "cm"), 
+     merge_legend = TRUE)
+```
+
+```{r, eval = FALSE, echo = FALSE}
+pdf(file = paste0("figs/Methylation_classification.pdf"))
+draw(h_list, main_heatmap = "Mean \nmethylation", ht_gap = unit(0.5, "cm"), 
+     merge_legend = TRUE)
+dev.off()
+```
+
+## Proportion of methylation regulated
+
+```{r prop_meth}
+x <- enframe(table(CT_genes_types$Regulation), name = "Regulation") %>% 
+  left_join(enframe(prop.table(table(CT_genes_types$Regulation)), 
+                    name = "Regulation", value = "Percent"))
+met_prop_fig <- ggplot(x, aes(x = '', y = value, fill = Regulation)) +
+  geom_bar(stat="identity", width=1) +
+  coord_polar("y", start = 0) +
+  geom_text(aes(y = value/5 + c(0, cumsum(value)[-length(value)]), 
+                label = rev(paste0(Regulation, "\n", value, "/", nrow(CT_genes_types), "\n",
+                               round(Percent * 100), "%"))))+
+            #, size = 6
+  scale_fill_manual(values = c( "indianred2", "cyan4")) +
+  theme(axis.title.x = element_blank(),
+        axis.title.y = element_blank(),
+        legend.title = element_blank(),
+        panel.background = element_blank(),
+        panel.border = element_blank(),
+        panel.grid = element_blank(),
+        axis.ticks = element_blank(),
+        axis.text.x = element_blank(),
+        legend.position = "none")
+met_prop_fig
+```
+
+```{r, eval = F}
+pdf(file = paste0("figs/Met_proportions.pdf"))
+met_prop_fig
+dev.off()
+```
+
+## Enrichment on X chromosome
+
+Genes regulated by methylation are enriched on the X chromosome.
+
+```{r}
+CT_genes_types$chr <- factor(CT_genes_types$chr, levels = c(1:22, "X", "Y"))
+
+fig_X_enrichment <- ggplot(CT_genes_types, aes(x = chr, fill = Regulation)) +
+  geom_bar(stat = 'count') +
+  xlab("Chromosome") +
+  ylab("Gene number") +
+  scale_fill_manual(values = Regulation_colors) +
+  theme(legend.position = "bottom",
+        aspect.ratio = 1)
+fig_X_enrichment
+```
+
+```{r, eval = FALSE, echo=FALSE}
+pdf(file = paste0("figs/X_enrichment.pdf"), width = 7, height = 4)
+cowplot::plot_grid(met_prop_fig, fig_X_enrichment, labels = c('A', 'B'), 
+          rel_widths = c(1, 2),
+          label_size = 12)
+dev.off()
 ```
@@
 On the chromosome X, there is a clear enrichment of CT genes regulated by 
-methylation (`r table(CT_genes$X_linked, CT_genes$regulated_by_methylation)[2, 2]`/`r table(CT_genes$X_linked)[[2]]` 
-chrX genes or `r table(CT_genes$X_linked, CT_genes$regulated_by_methylation)[2, 2]`/`r table(CT_genes$regulated_by_methylation)[[2]]` 
-genes regulated by methylation).
-
-Let's check that with a statistical test, I here want to see if, on there is an 
-enrichment of genes regulated by methylation on the X chromosome. I need to do a
-Pearson Chi square test (to know if observed proportion differ from expected 
-proportion). It is a statistical method to determine if two categorical variables
-have a significant correlation between them. I can directly put a matrix 
-(my table) in the function
-
-```{r chisq}
-chisq.test(table(CT_genes$X_linked, CT_genes$regulated_by_methylation))
-```
-
-```{r X_meth_fig}
-CT_genes$chr_factor <- factor(CT_genes$chr,
-                       levels = c("1", "2", "3", "4", "5", "6", "7", "8", "9",
-                                  "10", "11", "12", "13", "14", "15", "16", "17", 
-                                  "18", "19", "20", "21", "22", "X", "Y"))
-
-CT_genes %>% 
-  mutate(regulated_by_methylation = ifelse(regulated_by_methylation,
-                                           "Regulated by methylation",
-                                           "Not regulated by methylation")) %>% 
-  ggplot(aes(x = chr_factor, fill = X_linked)) +
+methylation (`r table(CT_genes$X_linked, CT_genes$regulated_by_methylation)[2, 2]`
+/`r table(CT_genes$X_linked)[[2]]` chrX genes or 
+`r table(CT_genes$X_linked, CT_genes$regulated_by_methylation)[2, 2]`/
+`r table(CT_genes$regulated_by_methylation)[[2]]` genes regulated by 
+methylation).
+
+This has to be checked with stats
+
+```{r meth_stats}
+CT_genes_types <- CT_genes_types %>% 
+  mutate(group = case_when(chr == "X" ~ "X-linked",
+                           chr == "Y" ~ "Y-linked",
+                           !chr %in% c("X", "Y") ~ "Autosome"))
+
+x <- table(CT_genes_types$group, CT_genes_types$Regulation)
+x
+fisher.test(x[c("Autosome", "X-linked"),])
+fisher.test(x[c("Autosome", "Y-linked"),])
+
+chisq.test(x[c("Autosome", "X-linked"),])
+chisq.test(x[c("Autosome", "Y-linked"),])
+```
+
+Genes controlled by methylation are significantly enriched on the X
+
+## Put figs together
+
+Putting the figs on the same panel with a bit of fine tuning
+
+```{r meth_figs}
+x <- enframe(table(CT_genes_types$Regulation), name = "Regulation") %>% 
+  left_join(enframe(prop.table(table(CT_genes_types$Regulation)), 
+                    name = "Regulation", value = "Percent"))
+prop_fig <- ggplot(x, aes(x = '', y = value, fill = Regulation)) +
+  geom_bar(stat="identity", width = 1) +
+  coord_polar("y", start = 0) +
+  geom_text(aes(y = value/5 + c(0, cumsum(value)[-length(value)]),
+                label = rev(paste0(Regulation, "\n", value, "/", nrow(CT_genes_types), "\n",
+                               round(Percent * 100), "%"))), size = 1.2 )+
+  scale_fill_manual(values = Regulation_colors) +
+  theme(axis.title.x = element_blank(),
+        axis.title.y = element_blank(),
+        legend.title = element_blank(),
+        panel.background = element_blank(),
+        panel.border = element_blank(),
+        panel.grid = element_blank(),
+        axis.ticks = element_blank(),
+        axis.text.x = element_blank(),
+        legend.position = "none")
+
+fig_freq_by_chr <- ggplot(CT_genes_types, aes(x = chr, fill = Regulation)) +
   geom_bar(stat = 'count') +
-  scale_fill_manual(values = c("royalblue1", "deeppink")) +
-  facet_wrap(~ regulated_by_methylation, ncol = 2) +
+  xlab("Chromosome") +
+  ylab("Gene number") +
+  scale_fill_manual(values = Regulation_colors) +
+  theme_test()+
+  theme(legend.position = "top",
+        aspect.ratio = 1,
+        axis.text = element_text(size = 3),
+        axis.title = element_text(size = 5),
+        axis.ticks = element_line(linewidth = 0.2),
+        axis.ticks.length = unit(0.5, "mm"),
+        legend.text = element_text(size = 5),
+        legend.key.size = unit(2, "mm"),
+        legend.title = element_blank())
+
+right_fig <- cowplot::plot_grid(prop_fig, fig_freq_by_chr, labels = c('B', 'C'), 
+                                #hjust = -2, 
+                                vjust = 1,
+                                #scale = c(0.8, 1), 
+                                rel_heights = c(0.5, 1),
+                                ncol = 1)
+
+left_fig <- cowplot::plot_grid(grid.grabExpr(
+  draw(h_list, main_heatmap = "Mean \nmethylation", ht_gap = unit(0.3, "cm"), 
+       merge_legend = TRUE)), labels = c('A'))
+
+cowplot::plot_grid(left_fig, right_fig, 
+          rel_widths = c(2,1),
+          #scale = c(0.6, 0.5),
+          label_size = 10, nrow = 1, 
+          #align = "h", 
+          align = "vh")
+```
+
+```{r}
+pdf(file = paste0("figs/Methylation_panel.pdf"), width = 7, height = 4)
+cowplot::plot_grid(left_fig, right_fig, 
+          rel_widths = c(2,1),
+          scale = c(1, 0.8),
+          label_size = 10, nrow = 1,
+          align = "h")
+dev.off()
+```
+
+# Gene biotype
+
+Mainly protein-coding genes
+Most genes are mainly protein coding genes 
+(`r table(CT_genes$transcript_biotype)["protein_coding"]/dim(CT_genes)[1]*100`%).
+
+```{r biotype}
+prop.table(table(CT_genes$transcript_biotype))
+```
+
+Gene biotype / regulation
+
+```{r biotype_reg}
+prop.table(table(methylation = CT_genes$regulated_by_methylation,
+                 CT_genes$transcript_biotype))
+
+prop.table(table(CT_genes$transcript_biotype,
+                 methylation = CT_genes$regulated_by_methylation))
+```
+# CT genes in tumour : coexpression and link with global hypomethylation
+
+## Positive tumour proportion and expression level
+
+CT genes regulated by methylation are activated more frequently and at higher 
+levels
+
+```{r pos_tum_exprs}
+TCGA <- TCGA_TPM
+TCGA <- TCGA[, TCGA$shortLetterCode != "NT"]
+
+tumors <- unique(TCGA$project_id)
+
+thr_pos <- 1
+
+prop_of_pos_tum <- as_tibble(assay(TCGA), rownames = "ensembl_gene_id") %>% 
+  filter(ensembl_gene_id %in% CT_genes$ensembl_gene_id) %>% 
+  pivot_longer(names_to = "barcode", values_to = "TPM", -ensembl_gene_id) %>% 
+  mutate(activated = ifelse(TPM > thr_pos, "n_pos", "n_neg")) %>% 
+  group_by(ensembl_gene_id, activated) %>% 
+  dplyr::summarize(n = n()) %>% 
+  pivot_wider(names_from = activated, values_from = n) %>% 
+  mutate(proportion_of_positive_tum = n_pos * 100 / (n_pos + n_neg))
+
+mean_in_pos <- as_tibble(assay(TCGA), rownames = "ensembl_gene_id") %>% 
+  filter(ensembl_gene_id %in% CT_genes$ensembl_gene_id) %>%  
+  pivot_longer(names_to = "barcode", values_to = "TPM", -ensembl_gene_id) %>% 
+  mutate(activated = ifelse(TPM > thr_pos, "n_pos", "n_neg")) %>% 
+  filter(activated == "n_pos") %>% 
+  group_by(ensembl_gene_id) %>% 
+  summarise(mean_log = log1p(mean(TPM)),
+            mean_TPM = mean(TPM))
+
+mean_and_prop_TCGA <- prop_of_pos_tum %>% 
+  left_join(mean_in_pos) %>% 
+  right_join(CT_genes_types) %>% 
+  dplyr::select(gene, mean_log, proportion_of_positive_tum, Regulation, X_linked) %>% 
+  mutate(source = "TCGA tumor samples")
+
+# combine with CCLE data
+load("~/cluster/Packages/CTdata/eh_data/CCLE_data.rda")
+ccle <- CCLE_data
+tumors <- unique(ccle$primary_disease)
+
+prop_of_pos_tum <- as_tibble(assay(ccle), rownames = "ensembl_gene_id") %>% 
+  filter(ensembl_gene_id %in% CT_genes$ensembl_gene_id) %>%  
+  pivot_longer(names_to = "cell", values_to = "TPM", -ensembl_gene_id) %>% 
+  mutate(activated = ifelse(TPM > thr_pos, "n_pos", "n_neg")) %>% 
+  group_by(ensembl_gene_id, activated) %>% 
+  dplyr::summarize(n = n()) %>% 
+  pivot_wider(names_from = activated, values_from = n) %>% 
+  mutate(proportion_of_positive_tum = n_pos * 100 / (n_pos + n_neg))
+
+mean_in_pos <- as_tibble(assay(ccle), rownames = "ensembl_gene_id") %>% 
+  filter(ensembl_gene_id %in% CT_genes$ensembl_gene_id) %>%   
+  pivot_longer(names_to = "cell", values_to = "TPM", -ensembl_gene_id) %>% 
+  mutate(activated = ifelse(TPM > thr_pos, "n_pos", "n_neg")) %>% 
+  filter(activated == "n_pos") %>% 
+  group_by(ensembl_gene_id) %>% 
+  summarise(mean_log = log1p(mean(TPM)),
+            mean_TPM = mean(TPM))
+
+mean_and_prop_CCLE <- prop_of_pos_tum %>% 
+  left_join(mean_in_pos) %>% 
+  right_join(CT_genes_types) %>% 
+  dplyr::select(gene, mean_log, proportion_of_positive_tum, Regulation, X_linked) %>% 
+  mutate(source = "CCLE tumor cell lines")
+
+mean_and_prop <- rbind(mean_and_prop_TCGA, mean_and_prop_CCLE)
+
+labelled_genes <- mean_and_prop %>% 
+  filter(proportion_of_positive_tum > 22 | mean_log > 4)
+mean_and_prop$source <- factor(mean_and_prop$source, 
+                               levels = c("TCGA tumor samples", "CCLE tumor cell lines"))
+
+fig_mean_prop_in_tum <- ggplot(mean_and_prop,
+                               aes(x = mean_log, y = proportion_of_positive_tum,
+                                   shape = X_linked,
+                                   color = Regulation,
+                                   label = gene)) +
+  geom_point(size = 1) +
+  scale_colour_manual(values = Regulation_colors)+
+  ggrepel::geom_text_repel(
+    data = mean_and_prop %>% 
+      filter(proportion_of_positive_tum > 20 | 
+               (proportion_of_positive_tum > 5 & mean_log > 4)), 
+    aes(x = mean_log, y = proportion_of_positive_tum, 
+        label = gene), size = 1.5, color = "black", min.segment.length = 1,
+    max.overlaps = 30, direction = "both") +
+  xlab("Mean expression log(TPM + 1)\nin positive samples") +
+  ylab("Proportion of positive samples (%)") +
+  facet_wrap(~ source, ncol = 1) +
   theme_bw() +
-  xlab("Chromosome") +
-  ylab("Number of genes") +
-  theme(axis.text.x = element_text(angle = 90, hjust = 1),
-        legend.position = "none",
-        axis.title = element_text(size = 10, color = "gray10"))
-```
-
-There is indeed a significative link between regulation by methylation and being
-on the X chromosome. There is thus an enrichment of CT genes regulated by 
-methylation on the X chromosome (and inversely).
-
-## Tumour and methylation
-
-Using CTexploreR functions, we can explore all CT genes or focus on some 
-potential targets.
-
-### All CT
-
-For these heatmaps, the code comes from the function but has been copies to add 
-some annotations.
-
-```{r common_annot}
-chr_mat <- as.matrix(CT_genes$X_linked)
-chr_mat <- ifelse(chr_mat == TRUE, "X-linked", "Not X")
-rownames(chr_mat) <- CT_genes$external_gene_name
-row_ha_chr <- rowAnnotation(chr_factor = chr_mat,
-                            annotation_legend_param = legends_param,
-                            simple_anno_size = unit(0.5, "cm"),
-                            col = list(chr_factor = chr_colors),
-                            annotation_name_gp = gpar(fontsize = 8),
-                            annotation_name_side = "top")
-
-regulation_mat <- as.matrix(CT_genes$regulated_by_methylation)
-regulation_mat <- ifelse(regulation_mat == TRUE, "Methylation",
-                         "Not methylation")
-rownames(regulation_mat) <- CT_genes$external_gene_name
-
-row_ha_reg <- rowAnnotation(regulation = regulation_mat,
-                            annotation_legend_param = legends_param,
-                            simple_anno_size = unit(0.5, "cm"),
-                            col = list(regulation = meth_colors),
-                            annotation_name_gp = gpar(fontsize = 8),
-                            annotation_name_side = "top")
-
-left_annot <- c(row_ha_chr, row_ha_reg, gap = unit(1, "mm"))
-
-split <- data.frame(CT_genes$regulated_by_methylation, CT_genes$X_linked)
-```
-
-```{r TCGA_heatmap, message = FALSE}
-database <- TCGA_TPM
-database$tumor <- sub(pattern = 'TCGA-', x = database$project_id, '')
-database$type <- "Tumor"
-database$type[database$shortLetterCode == "NT"] <- "Peritumoral"
-database <- database[, order(database$tumor, database$type)]
-
-genes <- CT_genes$external_gene_name
-database <- database[rowData(database)$external_gene_name %in% genes, ]
-database <- database[match(genes, rowData(database)$external_gene_name), ]
-database <- database[, database$type == "Tumor"]
+  theme(legend.position = "right",
+        aspect.ratio = 0.8,
+        panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank(),
+        legend.key.size = unit(0.3, 'cm'),
+        legend.text = element_text(size = 6),
+        legend.title = element_text(size = 7),
+        axis.text = element_text(size = 5),
+        axis.ticks = element_line(size = 0.1),
+        axis.title = element_text(size = 6),
+        strip.text = element_text(size = 6))
+
+fig_mean_prop_in_tum
+```
+```{r, eval = FALSE, echo = FALSE}
+pdf(file = paste0("figs/Mean_prop_in_tumors.pdf"), width = 5, height = 4)
+fig_mean_prop_in_tum
+dev.off()
+```
+
+## Activation in TCGA, coexpression and global hypomethylation
+
+```{r AUC_score}
+keep <- rowSums(assay(TCGA)) > 0
+TCGA <- TCGA[keep, ]
+TCGA$project_id <- gsub("TCGA-", '', TCGA$project_id)
+
+cells_rankings <- suppressMessages(AUCell_buildRankings(assay(TCGA), 
+                                                        plotStats = TRUE, 
+                                                        verbose = FALSE))
+geneSetsMet <- list(methylation = CT_genes_types %>% 
+                      filter(Regulation == "Methylation") %>%  
+                      filter(ensembl_gene_id %in% rownames(TCGA)) %>% 
+                      pull(ensembl_gene_id),
+                    not_methylation = CT_genes_types %>% 
+                      filter(Regulation == "Not methylation") %>%  
+                      filter(ensembl_gene_id %in% rownames(TCGA)) %>% 
+                      pull(ensembl_gene_id))
+
+# Calculate enrichment scores
+cells_AUC_met <- AUCell_calcAUC(geneSetsMet, 
+                            cells_rankings, 
+                            aucMaxRank = ceiling(0.2 * nrow(cells_rankings)))
+
+TCGA$AUC_genes_regulated_by_met <- as.vector(assay(cells_AUC_met)["methylation", ])
+TCGA$AUC_genes_not_regulated_by_met <- as.vector(assay(cells_AUC_met)["not_methylation", ])
+```
+
+- Show One heatmap for CT-met with AUScores and Global hypo levels
+
+- Show another one with CT-notMet (low coexpression, no pattern, 
+uncorrelated with global hypo)
+
+Plotted on the heatmaps genes that have a TPM > 0 in at least 5% of tumor samples
+
+```{r TCGA_heatmap}
+legends_param_TCGA <- list(
+  labels_gp = gpar(col = "black", fontsize = 2),
+  title_gp = gpar(col = "black", fontsize = 3),
+  simple_anno_size = unit(0.1, "cm"),
+  row_names_gp = gpar(fontsize = 1),
+  annotation_name_side = "left",
+  border = FALSE,
+  border_gp = gpar(lwd = 0.1),
+  grid_width = unit(0.05, "cm"),
+  grid_height = unit(0.1, "cm"),
+  legend_height = unit(0.01, "cm"),
+  direction = "horizontal",
+  use_raster = TRUE,
+               raster_device = "CairoPNG",
+               raster_quality = 10)
+
+print_TCGA_heatmap_with_AUCscores <- function(genes = NULL,
+                                              tum_type = c("SKCM", "LUAD", 
+                                                           "LUSC", "COAD", 
+                                                           "BRCA", "HNSC"),
+                                              AUScores = "AUC_genes_met", 
+                                              scale_lims = NULL,
+                                              show_AUScore = TRUE,
+                                              fontsize = 4, 
+                                              clust_method = "ward.D",
+                                              split_rows_by = "regulation", 
+                                              h_width = 7,
+                                              h_height = 5,
+                                              clustRow = TRUE,
+                                              clustCol = FALSE){
+
+  tcga <- TCGA[, TCGA$project_id %in% tum_type & !is.na(TCGA$global_methylation)]
+  tcga$project_id <- factor(tcga$project_id , levels = tum_type)
+
+  # Rm genes expressed in less than 5% of tumors
+  keep <- (rowSums(assay(tcga) > 0) / dim(tcga)[2]) > 0.05
+  tcga <- tcga[keep,]
+
+  df_col <- data.frame("project_id" = tcga$project_id,
+                       "AUScore" = tcga$AUC_genes_regulated_by_met,
+                       "global_methylation" = tcga$global_methylation,
+                       "gender" = tcga$gender)
+
+  if (AUScores == "AUC_genes_not_met") {
+    df_col <- data.frame("project_id" = tcga$project_id,
+                         "AUScore" = tcga$AUC_genes_not_regulated_by_met,
+                         "global_methylation" = tcga$global_methylation,
+                         "gender" = tcga$gender)
+  }
+
+  rownames(df_col) <- rownames(colData(tcga))
+  df_col <- df_col[!is.na(df_col$global_methylation),]
+  df_col <- df_col[order(df_col$project_id, df_col$AUScore), ]
+  #  df_col <- df_col[order(df_col$project_id), ]
+  split_by <- factor(df_col$project_id)
 
   column_ha_tumor <- HeatmapAnnotation(
-  Tumor = database$tumor,
+    TCGA_tumor = df_col$project_id,
+    border = FALSE,
+    col = list(TCGA_tumor = TCGA_colors),
+    annotation_label = gpar(fontsize = 4),
+    annotation_name_gp = gpar(fontsize = 0),
+    simple_anno_size = unit(0.1, "cm"),
+    annotation_legend_param = legends_param_TCGA)
+
+  AUCscore_col <- colorRamp2(seq(0, quantile(tcga$AUC_genes_regulated_by_met, 0.9), length = 4),
+                             c("gray", "gold", "firebrick1", "firebrick4"))
+
+  column_ha_AUC_score <- HeatmapAnnotation(
+    AUCscore = df_col$AUScore,
+    border = FALSE,
+    col = list(AUCscore = AUCscore_col),
+    annotation_label = gpar(fontsize = 4),
+    annotation_name_gp = gpar(fontsize = 0),
+    simple_anno_size = unit(0.1, "cm"),
+    annotation_legend_param = legends_param_TCGA)
+
+    global_met_col <-  colorRamp2(
+    c(quantile(tcga$global_methylation, 0.1), quantile(tcga$global_methylation, 0.9)), 
+                               c("moccasin","dodgerblue4"))
+
+  column_ha_global_methylation <- HeatmapAnnotation(
+    "Global methylation level" = df_col$global_methylation,
+    border = FALSE,
+    col = list("Global methylation level"  = global_met_col),
+    annotation_label = gpar(fontsize = 4),
+    annotation_name_gp = gpar(fontsize = 0),
+    simple_anno_size = unit(0.1, "cm"),
+    annotation_legend_param = legends_param_TCGA)
+
+  column_ha_gender <- HeatmapAnnotation(
+    gender = df_col$gender,
+    border = FALSE,
+    #col = list(TCGA_tumor = TCGA_colors),
+    annotation_label = gpar(fontsize = 4),
+    annotation_name_gp = gpar(fontsize = 0),
+   simple_anno_size = unit(0.1, "cm"),
+    annotation_legend_param = legends_param_TCGA)
+
+  ## Use gene names instead of ENSEMBL IDs
+  mat <- SummarizedExperiment::assay(tcga)
+  rownames(mat) <- rowData(tcga)$external_gene_name
+
+  gene_type <- CT_genes_types %>% 
+    filter(gene %in% genes) %>% 
+    filter(gene %in%  rownames(mat))
+  mat <- log1p(mat)
+
+  rowSplit <- NULL
+  if (!is.null(split_rows_by)){
+    if (split_rows_by == "regulation") {
+      gene_type <- gene_type %>% arrange(Regulation)
+      rowSplit <- gene_type$Regulation
+    }
+
+    if (split_rows_by == "chr_met") {
+      gene_type <- gene_type %>% arrange(chr_met)
+      rowSplit <- gene_type$chr_met
+    }
+  }
+
+  Regulation <- as.matrix(gene_type$Regulation)
+  rownames(Regulation) <- gene_type$gene
+  row_ha_regulation <- rowAnnotation(Regulation = Regulation,
+                                     annotation_legend_param = legends_param_TCGA,
+                                     simple_anno_size = unit(0.2, "cm"),
+                                     col = list(Regulation = Regulation_colors),
+                                     annotation_name_gp = gpar(fontsize = 0))
+
+  chr_met <- as.matrix(gene_type$chr_met)
+  rownames(chr_met) <- gene_type$gene
+  row_ha_chr_met <- rowAnnotation(chr_met = chr_met,
+                                  annotation_legend_param = legends_param_TCGA,
+                                  simple_anno_size = unit(0.2, "cm"),
+                                  col = list(chr_met = chr_met_colors),
+                                  annotation_name_gp = gpar(fontsize = 0))
+
+  Chr <- as.matrix(gene_type$Chr)
+  rownames(Chr) <- gene_type$gene
+  row_ha_chr <- rowAnnotation(Chr = Chr,
+                              annotation_legend_param = legends_param_TCGA,
+                              simple_anno_size = unit(0.2, "cm"),
+                              col = list(Chr = Chr_colors),
+                              annotation_name_gp = gpar(fontsize = 0))
+
+  if (is.null(scale_lims)) {
+    scale_lims <- c(0, quantile(rowMaxs(mat), 0.8))
+  }
+
+  if (show_AUScore) {
+    topAnnot <- c(#column_ha_tumor,
+                  column_ha_AUC_score,
+                  column_ha_global_methylation,
+                  gap = unit(0.5, "mm"))
+  } else {
+    topAnnot <- c(column_ha_tumor,
+                  column_ha_global_methylation,
+                  gap = unit(0.5, "mm"))
+  }
+
+  ht_opt$TITLE_PADDING = unit(c(2, 2), "points")
+
+  h <- Heatmap(mat[gene_type$gene, rownames(df_col)],
+               name = "log(TPM +1)",
+               column_title = levels(df_col$project_id),
+               column_title_gp = gpar(col =c("COAD" = "darkorchid2", 
+                                             "HNSC" = "deeppink2",
+                                             "LUSC" = "seagreen3", 
+                                             "SKCM" = "red3"), 
+                                      fontsize = 3, face = "bold"),
+               clustering_method_rows = clust_method,
+               clustering_method_columns = clust_method,
+               column_split = split_by,
+               cluster_column_slices = FALSE,
+               row_gap = unit(c(1), "mm"),
+               column_gap = unit(c(1), "mm"),
+               col = colorRamp2(seq(scale_lims[1], scale_lims[2], length = 11),
+                                legend_colors),
+               cluster_rows = clustRow,
+               cluster_columns = clustCol,
+               show_row_names = TRUE,
+               show_column_names = FALSE,
+               show_row_dend = FALSE,
+               show_column_dend = FALSE,
+               column_names_gp = gpar(fontsize = 4),
+               column_names_centered = TRUE,
+               row_names_gp = gpar(fontsize = fontsize),
+               row_names_side = "left",
                border = TRUE,
-  col = list(Tumor = c("BRCA" = "midnightblue", "COAD" = "darkorchid2",
-                 "ESCA" = "gold", "HNSC" = "deeppink2",
-                 "LUAD" = "seagreen", "LUSC" = "seagreen3",
-                 "SKCM" = "red3")),
-  annotation_name_gp = gpar(fontsize = 8),
-  annotation_legend_param = legends_param)
-
-split_by <- factor(database$tumor)
-col_annot <- column_ha_tumor
-
-mat <- log1p(assay(database))
-rownames(mat) <- rowData(database)$external_gene_name
-
-Heatmap(mat[, , drop = FALSE],
-        name = "logTPM",
-        column_title = paste0("Expression in TCGA samples (all)"),
-        column_split = split_by,
-        row_split = split,
+               border_gp = gpar(lwd = 0.05),
+               column_names_side = c("top"),
+               column_names_rot = 90,
                row_title_gp = gpar(fontsize = 0),
-        col = colorRamp2(seq(0, max(mat), length = 11),
-                         legend_colors),
-        clustering_method_rows = "ward.D",
-        clustering_method_columns = "ward.D",
-        cluster_rows = TRUE,
-        show_column_names = FALSE,
-        cluster_columns = TRUE,
+               top_annotation = topAnnot,
+               row_split = rowSplit,
+               cluster_row_slices = FALSE,
+               heatmap_width = unit(h_width, "cm"),
+               heatmap_height = unit(h_height, "cm"),
+               use_raster = TRUE,
+               raster_device = "CairoPNG",
+               raster_quality = 10,
+               heatmap_legend_param = legends_param_TCGA,
+               show_heatmap_legend = TRUE)
+
+  h
+}
+
+met_genes <- CT_genes_types %>% 
+  filter(regulated_by_methylation) %>% 
+  pull(gene)
+
+tum_types <- c("SKCM", "LUSC", "HNSC", "COAD")
+
+h1 <- print_TCGA_heatmap_with_AUCscores(genes = met_genes, 
+                                        clustCol = TRUE,
+                                        clust_method = "ward.D",
+                                        h_width = 6,
+                                        h_height = 5,
+                                        fontsize = 1, 
+                                        scale_lims = c(0, 6),
+                                        # tum_type = c("SKCM", "LUAD", "COAD", "BRCA"),
+                                        tum_type = tum_types,
+                                        split_rows_by = NULL)
+draw(h1, column_title = "CT genes regulated by methylation",
+     column_title_gp = gpar(fontsize = 4), 
+     merge_legend = TRUE)
+
+not_met_genes <- CT_genes_types %>% 
+  filter(!regulated_by_methylation) %>% 
+  pull(gene)
+
+h2 <- print_TCGA_heatmap_with_AUCscores(genes = not_met_genes, 
+                                       clustCol = TRUE,
+                                       AUScores = "AUC_genes_not_met", 
+                                       show_AUScore = TRUE,
+                                       clust_method = "ward.D",
+                                       h_width = 6,
+                                       h_height = 2.5,
+                                       fontsize = 1, 
+                                       scale_lims = c(0, 6),
+                                       #  tum_type = c("SKCM", "LUAD", "COAD", "BRCA"),
+                                       tum_type = tum_types,
+                                       split_rows_by = NULL)
+draw(h2, column_title = "CT genes not regulated by methylation",
+     column_title_gp = gpar(fontsize = 4))
+
+```
+
+```{r, echo = FALSE, eval = FALSE}
+pdf(file = "figs/CT_met_AUC_hypo_TCGA.pdf", height = 2.3, width = 3)
+draw(h1, column_title = "CT genes regulated by methylation",
+     column_title_gp = gpar(fontsize = 5), show_heatmap_legend = FALSE,
+     show_annotation_legend = FALSE)
+dev.off()
+
+pdf(file = "figs/CT_not_met_AUC_hypo_TCGA.pdf", height = 2.3, width = 3)
+draw(h2, column_title = "CT genes not regulated by methylation",
+     column_title_gp = gpar(fontsize = 4), merge_legend = TRUE, 
+     heatmap_legend_side = "bottom")
+dev.off()
+```
+
+## AUC correlation with global hypomethylation
+
+```{r AUC_hypometh}
+tcga <- TCGA[, TCGA$project_id %in% tum_types & !is.na(TCGA$global_methylation)]
+tcga$project_id <- factor(tcga$project_id , levels = tum_types)
+
+f1 <- as_tibble(colData(TCGA), rownames = "barcode") %>% 
+  filter(project_id %in% tum_types) %>% 
+  mutate(project_id = factor(project_id, levels = tum_types,)) %>% 
+  ggplot(aes(x = global_methylation, y = AUC_genes_regulated_by_met, color = project_id)) +
+  geom_point(size = 0.3) +
+  facet_wrap(~ project_id, nrow = 1) +
+  scale_color_manual(values = c("COAD" = "darkorchid2", "HNSC" = "deeppink2",
+                                "LUSC" = "seagreen3", "SKCM" = "red3")) + 
+  geom_smooth(method = 'lm', color = "black", linewidth = 0.3, se = FALSE) +
+  xlab('Global methylation level') +
+  ylab("AUCscore CT genes\nregulated by methylation") +
+  theme(legend.position = "none",
+        aspect.ratio = 1,
+        panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank(),
+        axis.title.y = element_text(size = 8),
+        axis.title.x = element_text(size = 9),
+        axis.text = element_text(size = 4),
+        axis.ticks = element_line(size = 0.1)) +
+  ggpubr::stat_cor(method="pearson", size = 2.5, label.x = .4, 
+           label.y = 0.11, label.sep = "\n", color = "black") +
+  ylim(0, 0.12)
+
+f2 <- as_tibble(colData(TCGA), rownames = "barcode") %>% 
+  filter(project_id %in% tum_types) %>% 
+  mutate(project_id = factor(project_id, levels = tum_types,)) %>% 
+  ggplot(aes(x = global_methylation, y = AUC_genes_not_regulated_by_met, color = project_id)) +
+  geom_point(size = 0.3) +
+  facet_wrap(~ project_id, nrow = 1) +
+  scale_color_manual(values = c("COAD" = "darkorchid2", "HNSC" = "deeppink2",
+                                "LUSC" = "seagreen3", "SKCM" = "red3")) + 
+  geom_smooth(method = 'lm', color = "black", linewidth = 0.3, se = FALSE) +
+  xlab('Global methylation level') +
+  ylab("AUCscore CT genes\nnot regulated by methylation") +
+  theme(legend.position = "none",
+        aspect.ratio = 1,
+        panel.grid.major = element_blank(),
+        panel.grid.minor = element_blank(),
+        axis.title.y = element_text(size = 8),
+        axis.title.x = element_text(size = 9),
+        axis.text = element_text(size = 4),
+        axis.ticks = element_line(size = 0.1)) +
+  ggpubr::stat_cor(method="pearson", size = 2.5, label.x = .4, 
+           label.y = 0.11, label.sep = "\n", color = "black") +
+  ylim(0, 0.12)
+
+f1
+f2
+```
+
+```{r}
+pdf(file = paste0("figs/CT_met_AUCscore_hypo_TCGA.pdf"), height = 2.5)
+f1
+dev.off()
+
+pdf(file = paste0("figs/CT_not_met_AUCscore_hypo_TCGA.pdf"), height = 2.5)
+f2
+dev.off()
+```
+# CT expression in healthy cells (fetal and adult)
+
+First treating data
+
+```{r}
+################################################################################
+# Load testis scRNAseq data
+################################################################################
+
+testis_sce <- testis_sce[CT_genes$external_gene_name[
+  CT_genes$external_gene_name %in% rownames(testis_sce)],]
+
+# pool Sperm1 and Sperm2 types
+levels(testis_sce$type)[c(7,8)] <- "Sperm"
+germcells <-  c("SSC", "Spermatogonia", "Early_spermatocyte", 
+                "Late_spermatocyte", "Round_spermatid", 
+                "Elongated_spermatid", "Sperm")
+
+# add meiotic stage to the colData of testis_sce
+testis_sce$stage <- "pre-meiotic"
+testis_sce[, testis_sce$type %in% 
+             c("Early_spermatocyte", "Late_spermatocyte", 
+               "Round_spermatid")]$stage <- "meiotic"
+testis_sce[, testis_sce$type %in% 
+             c("Elongated_spermatid", "Sperm")]$stage <- "post-meiotic"
+testis_sce[, testis_sce$type %in% 
+             c("Macrophage", "Endothelial", "Myoid", "Sertoli", 
+               "Leydig")]$stage <- "somatic"
+testis_sce$stage <- factor(
+  testis_sce$stage, 
+  levels = c("pre-meiotic", "meiotic", "post-meiotic", "somatic"))
+
+# Keep genes detected in at least x % of germ cells 
+# ! Each cell type is present at different proportion 
+# (much more sperm cells than SSC cells for instance)
+# table(testis_sce$type)
+# => Calculate for each gene the proportion of expressing cells 
+# (> testis_logcount_thr) and keep genes expressed in at least x% of a cell type
+# instead of just keeping genes expressed in at least x% of cells.
+testis_percent_thr <- 0
+testis_logcount_thr <- 0
+
+percent_of_expressing_cells <- 
+  tibble(gene = rownames(testis_sce),
+         prop = rowSums(assay(testis_sce[, testis_sce$type %in% germcells[1]], 
+                              "logcounts") > testis_logcount_thr) / 
+           dim(testis_sce[, testis_sce$type %in% germcells[1]])[2])
+names(percent_of_expressing_cells) <- c("gene", germcells[1])
+
+for (cell in germcells[-1]) {
+  tmp <- 
+    tibble(gene = rownames(testis_sce),
+           prop = rowSums(assay(testis_sce[, testis_sce$type %in% cell], 
+                                           "logcounts") > testis_logcount_thr) / 
+             dim(testis_sce[, testis_sce$type %in% cell])[2])
+  names(tmp) <- c("gene", cell)
+  percent_of_expressing_cells <- left_join(percent_of_expressing_cells, tmp)
+}
+
+kept_genes <- rownames(testis_sce)[
+  rowSums(percent_of_expressing_cells[, -1] >= testis_percent_thr) > 0]
+testis_sce <- testis_sce[kept_genes,]
+
+################################################################################
+# Load FGC and oocytes scRNAseq data
+################################################################################
+
+selected_CT <- BiocGenerics::intersect(rownames(FGC_sce), rownames(testis_sce))
+selected_CT <- BiocGenerics::intersect(rownames(oocytes_sce), selected_CT)
+
+FGC_sce <- FGC_sce[selected_CT, ]
+testis_sce <- testis_sce[selected_CT, ]
+oocytes_sce <- oocytes_sce[selected_CT, ]
+
+legends_param_sc <- list(
+  labels_gp = gpar(col = "black", fontsize = 2),
+  title_gp = gpar(col = "black", fontsize = 3),
+  simple_anno_size = unit(0.15, "cm"),
+  row_names_gp = gpar(fontsize = 1),
+  annotation_name_side = "left",
+  border = FALSE,
+  border_gp = gpar(lwd = 0.01),
+  grid_width = unit(0.2, "cm"),
+  grid_height = unit(0.2, "cm"),
+  legend_height = unit(0.1, "cm"),
+  annotation_name_gp = gpar(fontsize = 0))
+
+```
+
+Kept `r length(selected_CT)` CT genes for the analysis:
+
+CT genes with logcounts >= `r testis_logcount_thr` in at least 
+`r testis_percent_thr * 100` % of any specific type of testis germ cells 
+(adult testis dataset) and present in FCG and oocytes datasets
+
+```{r healthy_cells_function}
+#######################################################################
+# Function to downsample cell types
+#######################################################################
+sample_cells <- function(barcodes, nmax){
+  sample(barcodes, size = min(nmax, length(barcodes)), replace = FALSE)
+}
+
+#######################################################################
+## Function to draw expression in FGC with specific annotations
+## and downasampling the number of cells from each cell type
+#######################################################################
+draw_FGC_heatmap <- function(genes, 
+                             split_by = "chr_met", 
+                             ncells_max = NULL,
+                             scale_lims = NULL,
+                             sex = c("F", "M"),
+                             clusterRow = FALSE, 
+                             font_size = 6,
+                             show_legend = TRUE,
+                             h_width = 5,
+                             h_height = 10,
+                             show_right_annot = TRUE){
+
+  fetal_sce <- FGC_sce
+
+  # Set the colors used in the final heatmap based on all samples (F + M), 
+  # even when representing only F or M samples on the heatmap
+  # => same scale of colors
+  mat <-  as.matrix(assay(fetal_sce, "logcounts")[genes, ])
+
+  if (is.null(scale_lims)) {
+    scale_lims <- c(0, quantile(rowMaxs(mat), 0.8))
+  }
+
+  heatmap_colors <- colorRamp2(
+    seq(scale_lims[1], scale_lims[2], length = 11),                 
+    legend_colors)
+
+  fetal_sce <- fetal_sce[, fetal_sce$sex %in% sex]
+
+  df_col_FCS <- data.frame(development_stage = fetal_sce$development_stage,
+                           sex = fetal_sce$sex,
+                           stage = fetal_sce$stage,
+                           donor_id = fetal_sce$donor_id,
+                           fetal_cell_type = fetal_sce$cell_type,
+                           type = fetal_sce$type)
+
+  rownames(df_col_FCS) <- colnames(fetal_sce)
+  df_col_FCS <- df_col_FCS[order(df_col_FCS$type),]
+
+  #Randomly pick ncells_max cells of each cell type
+  if (!is.null(ncells_max)) {
+    set.seed(123)
+    selected_cells <- sample_cells(
+      rownames(df_col_FCS[df_col_FCS$type == levels(df_col_FCS$type)[1],]), 
+      ncells_max)
+    for (type in levels(df_col_FCS$type)[-1]){
+      tmp <- sample_cells(
+        rownames(df_col_FCS[df_col_FCS$type == type,]), ncells_max)
+      selected_cells <- c(selected_cells, tmp)
+    }
+    df_col_FCS <- df_col_FCS[selected_cells,]
+  }
+
+  column_ha_type = HeatmapAnnotation(
+    Fetal_cell_type = df_col_FCS$type,
+    col = list(Fetal_cell_type = Fetal_cell_colors),   
+    border = TRUE, 
+    simple_anno_size = unit(0.15, "cm"), 
+    annotation_legend_param = legends_param_sc,
+    annotation_name_gp = gpar(fontsize = 0))
+
+  column_ha_development_stage = HeatmapAnnotation(
+    development_stage = df_col_FCS$development_stage,
+    border = TRUE, 
+    simple_anno_size = unit(0.15, "cm"), 
+    annotation_legend_param = legends_param_sc,
+    annotation_name_gp = gpar(fontsize = 0))
+
+  column_ha_sex = HeatmapAnnotation(
+    sex = df_col_FCS$sex,
+    col = list(sex = sex_colors),
+    border = TRUE, 
+    simple_anno_size = unit(0.15, "cm"), 
+    annotation_legend_param = legends_param_sc,
+    annotation_name_gp = gpar(fontsize = 0))
+
+  column_ha_donor_id = HeatmapAnnotation(
+    donor_id = df_col_FCS$donor_id,
+    border = TRUE, 
+    simple_anno_size = unit(0.15, "cm"), 
+    annotation_legend_param = legends_param_sc,
+    annotation_name_gp = gpar(fontsize = 0))
+
+  column_ha_stage = HeatmapAnnotation( 
+    stage = df_col_FCS$stage,
+    col = list(stage = stage_colors),
+    border = TRUE, 
+    simple_anno_size = unit(0.15, "cm"), 
+    annotation_legend_param = legends_param_sc,
+    annotation_name_gp = gpar(fontsize = 0))
+
+  gene_type <- CT_genes_types %>% 
+    filter(gene %in% genes)
+
+  if (split_by == "regulation") {
+    gene_type <- gene_type %>% arrange(Regulation)
+    rowSplit <- gene_type$Regulation
+  }
+  if (split_by == "testis_specificity") {
+    gene_type <- gene_type %>% arrange(Testis_specificity)
+    rowSplit <- gene_type$Testis_specificity
+  }
+  if (split_by == "chr_met") {
+    gene_type <- gene_type %>% arrange(chr_met)
+    rowSplit <- gene_type$chr_met
+  }
+  if (!clusterRow & split_by == "chr_met") {
+    gene_type$gene <- factor(gene_type$gene, levels = genes)
+    gene_type <- gene_type  %>%
+      arrange(gene)
+    gene_type$gene <- as.vector(gene_type$gene)
+    rowSplit <- gene_type$chr_met
+  }
+
+  Regulation <- as.matrix(gene_type$Regulation)
+  rownames(Regulation) <- gene_type$gene
+  row_ha_regulation <- rowAnnotation(
+    Regulation = Regulation,
+    annotation_legend_param = legends_param_sc,
+    border = FALSE,
+    simple_anno_size = unit(0.15, "cm"), 
+    col = list(Regulation = Regulation_colors),
+    annotation_name_gp = gpar(fontsize = 0))
+
+  chr_met <- as.matrix(gene_type$chr_met)
+  rownames(chr_met) <- gene_type$gene
+  row_ha_chr_met <- rowAnnotation(
+    chr_met = chr_met,
+    annotation_legend_param = legends_param_sc,
+    border = FALSE,
+    simple_anno_size = unit(0.15, "cm"), 
+    col = list(chr_met = chr_met_colors),
+    annotation_name_gp = gpar(fontsize = 0))
+
+  Chr <- as.matrix(gene_type$Chr)
+  rownames(Chr) <- gene_type$gene
+  row_ha_chr <- rowAnnotation(Chr = Chr,
+                              annotation_legend_param = legends_param_sc,
+                              border = FALSE,
+                              simple_anno_size = unit(0.15, "cm"), 
+                              col = list(Chr = Chr_colors),
+                              annotation_name_gp = gpar(fontsize = 0))
+
+  mat <- as.matrix(
+    assay(fetal_sce, "logcounts")[gene_type$gene, rownames(df_col_FCS)])
+
+  rAnnot <- c(row_ha_regulation, row_ha_chr, gap = unit(1, "mm"))
+
+  if (!show_right_annot) rAnnot <- NULL
+
+  title <- "Fetal germ cells"
+  if (all(sex == "F")) {
+    title <- "Female fetal germ cells"
+  }
+  if (all(sex == "M")) {
+    title <- "Male fetal germ cells"
+  }
+
+  if (is.null(scale_lims)) {
+    scale_lims <- c(0, quantile(rowMaxs(mat), 0.8))
+  }
+
+  heatmap_colors <- colorRamp2(
+    seq(scale_lims[1], scale_lims[2], length = 11),                 
+    legend_colors)
+
+  ht_FGCs <- Heatmap(mat,
+                     name = "Logcounts",
+                     column_title = title,
+                     column_split = df_col_FCS$type,
+                     cluster_row_slices = FALSE,
+                     cluster_column_slices = FALSE,
+                     col = heatmap_colors,
+                     border = TRUE,
+                     border_gp = gpar(lwd = 0.2),
+                     row_gap = unit(c(1), "mm"),
+                     column_gap = unit(c(0.5), "mm"),
                      show_column_dend = FALSE,
                      show_row_dend = FALSE,
-        row_names_gp = gpar(fontsize = 4),
-        heatmap_legend_param = legends_param,
-        top_annotation = col_annot,
-        left_annotation = left_annot)
-
-```
-
-```{r CCLE_heatmap}
-database <- CCLE_data
-database$type <- tolower(database$type)
-genes <- CT_genes$external_gene_name
-database <- database[rowData(database)$external_gene_name %in% genes, ]
-database <- database[match(genes, rowData(database)$external_gene_name), ]
-
-mat <- log1p(assay(database))
-rownames(mat) <- rowData(database)$external_gene_name
-
-df_col <- data.frame("cell_line" = colData(database)$cell_line_name,
-                     "type" = as.factor(colData(database)$type))
-rownames(df_col) <- rownames(colData(database))
-df_col <- df_col[order(df_col$type), ]
-
-column_ha_type <- HeatmapAnnotation(
-  type = df_col$type,
+                     row_names_gp = gpar(fontsize = font_size),
+                     row_names_side = "left",
+                     cluster_rows = clusterRow,
+                     cluster_columns = FALSE,
+                     show_column_names = FALSE,
+                     show_row_names = FALSE,
+                     row_split = rowSplit,
+                     top_annotation = c(column_ha_sex, 
+                                        column_ha_stage,
+                                        #column_ha_group, 
+                                        #column_ha_donor_id,
+                                        column_ha_type),  
+                     right_annotation = rAnnot,
+                     column_title_gp = gpar(fontsize = 5),
+                     row_title_gp = gpar(fontsize = 0),
+                     gap = unit(0.01, "mm"),
+                     use_raster = TRUE,
+                     raster_device = "CairoPNG",
+                     #heatmap_width = unit(h_width, "cm"),
+                     #heatmap_height = unit(h_height, "cm"),
+                     raster_quality = 10,
+                     show_heatmap_legend = show_legend,
+                     heatmap_legend_param = legends_param_sc)
+
+  return(ht_FGCs)
+}
+
+#######################################################################
+## Function to draw expression in adult testis cells with specific annotations
+## and downasampling the number of cells from each cell type
+#######################################################################
+draw_testis_heatmap <- function(genes, 
+                                split_by = "chr_met", 
+                                sex = c("F", "M"),
+                                ncells_max = NULL,
+                                font_size = 6,
+                                scale_lims = NULL,
+                                clusterRow = FALSE,
+                                show_legend = TRUE,
+                                h_width = 5,
+                                h_height = 10,
+                                clust_method = "ward.D"){
+
+  testis_adult_sce <- testis_sce[, testis_sce$stage != "somatic"]
+  testis_adult_sce$type <- factor(testis_adult_sce$type, levels = germcells)
+  testis_adult_sce$stage <- factor(testis_adult_sce$stage, 
+                                   levels = c("pre-meiotic", "meiotic", 
+                                              "post-meiotic"))
+  df_col_testis <- data.frame(type = testis_adult_sce$type,
+                              donor = testis_adult_sce$Donor,
+                              stage = testis_adult_sce$stage)
+
+  df_col_testis$sex <- "M"
+
+  rownames(df_col_testis) <- colnames(testis_adult_sce)
+  df_col_testis <- df_col_testis[order(df_col_testis$type),]
+
+  #Randomly pick ncells_max cells of each cell type
+  if (!is.null(ncells_max)) {
+    set.seed(123)
+    selected_cells <- sample_cells(
+      rownames(df_col_testis[df_col_testis$type == 
+                               levels(df_col_testis$type)[1],]), ncells_max)
+    for (type in levels(df_col_testis$type)[-1]){
+      tmp <- sample_cells(
+        rownames(df_col_testis[df_col_testis$type == type,]), ncells_max)
+      selected_cells <- c(selected_cells, tmp)
+    }
+    df_col_testis <- df_col_testis[selected_cells,]
+  }
+
+  column_ha_cell_type = HeatmapAnnotation(
+    Testis_cell_type = df_col_testis$type,
+    col = list(Testis_cell_type = testis_colors),      
+    annotation_legend_param = legends_param_sc,
     border = TRUE, 
-  annotation_name_gp = gpar(fontsize = 8),
-  annotation_legend_param = legends_param,
-  col = list(type = c("lung" = "seagreen3", "skin" = "red3",
-                 "bile_duct" = "mediumpurple1", "bladder" = "mistyrose2",
-                 "colorectal" = "plum", "lymphoma" = "steelblue1",
-                 "uterine" = "darkorange4", "myeloma" = "turquoise3", 
-                 "kidney" = "thistle4",
-                 "pancreatic" = "darkmagenta", "brain" = "palegreen2",
-                 "gastric" = "wheat3", "breast" = "midnightblue",
-                 "bone" = "sienna1", "head_and_neck" = "deeppink2",
-                 "ovarian" = "tan3", "sarcoma" = "lightcoral",
-                 "leukemia" = "steelblue4", "esophageal"= "khaki",
-                 "neuroblastoma" = "olivedrab1")))
-
-Heatmap(mat[, rownames(df_col), drop = FALSE],
-        name = "logTPM",
-        column_title = "Gene Expression in tumor cell lines (CCLE)",
-        column_split = factor(df_col$type),
-        row_split = split,
+    simple_anno_size = unit(0.15, "cm"), 
+    annotation_name_gp = gpar(fontsize = 0))
+
+  column_ha_stage = HeatmapAnnotation(
+    stage = df_col_testis$stage,
+    col = list(stage = stage_colors),
+    annotation_legend_param = legends_param_sc,
+    border = TRUE, 
+    simple_anno_size = unit(0.15, "cm"), 
+    annotation_name_gp = gpar(fontsize = 0))
+
+  column_ha_sex = HeatmapAnnotation(
+    sex = df_col_testis$sex,
+    col = list(sex = sex_colors),
+    annotation_legend_param = legends_param_sc,
+    border = TRUE, 
+    simple_anno_size = unit(0.15, "cm"), 
+    annotation_name_gp = gpar(fontsize = 0))
+
+  gene_type <- CT_genes_types %>% 
+    filter(gene %in% genes)
+
+  if (split_by == "regulation") {
+    gene_type <- gene_type %>% arrange(Regulation)
+    rowSplit <- gene_type$Regulation
+  }
+  if (split_by == "testis_specificity") {
+    gene_type <- gene_type %>% arrange(Testis_specificity)
+    rowSplit <- gene_type$Testis_specificity
+  }
+  if (split_by == "chr_met") {
+    gene_type <- gene_type %>% arrange(chr_met)
+    rowSplit <- gene_type$chr_met
+  }
+  if (!clusterRow & split_by == "chr_met") {
+    gene_type$gene <- factor(gene_type$gene, levels = genes)
+    gene_type <- gene_type  %>%
+      arrange(gene)
+    gene_type$gene <- as.vector(gene_type$gene)
+    rowSplit <- gene_type$chr_met
+  }
+
+  Regulation <- as.matrix(gene_type$Regulation)
+  rownames(Regulation) <- gene_type$gene
+  row_ha_regulation <- rowAnnotation(
+    Regulation = Regulation,
+    annotation_legend_param = legends_param_sc,
+    border = FALSE,
+    simple_anno_size = unit(0.15, "cm"), 
+    col = list(Regulation = Regulation_colors),
+    annotation_name_gp = gpar(fontsize = 0))
+
+  chr_met <- as.matrix(gene_type$chr_met)
+  rownames(chr_met) <- gene_type$gene
+  row_ha_chr_met <- rowAnnotation(
+    chr_met = chr_met,
+    annotation_legend_param = legends_param_sc,
+    border = FALSE,
+    simple_anno_size = unit(0.15, "cm"), 
+    col = list(chr_met = chr_met_colors),
+    annotation_name_gp = gpar(fontsize = 0))
+
+  Chr <- as.matrix(gene_type$Chr)
+  rownames(Chr) <- gene_type$gene
+  row_ha_chr <- rowAnnotation(Chr = Chr,
+                              annotation_legend_param = legends_param_sc,
+                              border = FALSE,
+                              simple_anno_size = unit(0.15, "cm"), 
+                              col = list(Chr = Chr_colors),
+                              annotation_name_gp = gpar(fontsize = 0))
+
+  mat <-  as.matrix(
+    assay(testis_adult_sce, "logcounts")[gene_type$gene, rownames(df_col_testis)])
+
+  if (is.null(scale_lims)) {
+    scale_lims <- c(0, quantile(rowMaxs(mat), 0.8))
+  }
+
+  heatmap_colors <- colorRamp2(
+    seq(scale_lims[1], scale_lims[2], length = 11),                 
+    legend_colors)
+
+  ht_testis <- Heatmap(mat,
+                       name = "logcounts",
+                       column_title = "Adult testis",
+                       column_split = df_col_testis$type,
+                       row_title = NULL,
+                       col = heatmap_colors,
+                       border = TRUE,
+                       border_gp = gpar(lwd = 0.2),
+                       row_gap = unit(c(0.5), "mm"),
+                       column_gap = unit(c(0.5), "mm"),
+                       show_column_dend = FALSE,
+                       show_row_dend = FALSE,
+                       row_names_gp = gpar(fontsize = font_size),
+                       row_names_side = "right",
+                       cluster_rows = clusterRow,
+                       clustering_method_rows = clust_method,
+                       row_dend_reorder = TRUE,
+                       cluster_columns = FALSE,
+                       show_column_names = FALSE,
+                       show_row_names = TRUE,
+                       row_split = rowSplit,
+                       cluster_row_slices = FALSE,
+                       cluster_column_slices = FALSE,
+                       top_annotation = c(column_ha_sex, 
+                                          column_ha_stage,
+                                          column_ha_cell_type),  
+                       right_annotation = c(row_ha_regulation, 
+                                            row_ha_chr, 
+                                            gap = unit(1, "mm")),
+                       column_title_gp = gpar(fontsize = 5),
                        row_title_gp = gpar(fontsize = 0),
-        col = colorRamp2(seq(0, max(mat), length = 11),
-                          legend_colors),
-        clustering_method_rows = "ward.D",
-        clustering_method_columns = "ward.D",
-        cluster_rows = TRUE,
+                       gap = unit(0.01, "mm"),
+                       use_raster = TRUE,
+                       raster_device = "CairoPNG",
+                       raster_quality = 10,
+                       #heatmap_width = unit(h_width, "cm"),
+                       #heatmap_height = unit(h_height, "cm"),
+                       show_heatmap_legend =  show_legend,
+                       heatmap_legend_param = legends_param_sc)
+
+  return(ht_testis)
+}
+
+######################################################################
+## Function to draw expression in adult oocytes cells with specific annotations
+## and downasampling the number of cells from each cell type
+#######################################################################
+draw_oocytes_heatmap <- function(genes, 
+                               split_by = "chr_met", 
+                               ncells_max = NULL,
+                               font_size = 6,
+                               scale_lims = NULL,
+                               clusterRow = TRUE,
+                               show_legend = TRUE,
+                               h_width = 5,
+                               h_height = 10,
+                               clust_method = "ward.D"){
+
+  adult_oocytes_sce <- oocytes_sce[, oocytes_sce$type %in% c("Growing oocytes", 
+                                                       #"Fully grown oocytes",  
+                                                       #"Metaphase I", 
+                                                       "Metaphase II")]
+
+  df_col_oocytes <- data.frame(type = adult_oocytes_sce$type,
+                             sex = adult_oocytes_sce$sex,
+                             stage = adult_oocytes_sce$stage)
+
+  rownames(df_col_oocytes) <- colnames(adult_oocytes_sce)
+  df_col_oocytes <- df_col_oocytes[order(df_col_oocytes$type),]
+
+  #Randomly pick ncells_max cells of each cell type
+  if (!is.null(ncells_max)) {
+    set.seed(123)
+    selected_cells <- sample_cells(
+      rownames(df_col_oocytes[df_col_oocytes$type == 
+                              levels(df_col_oocytes$type)[1],]), ncells_max)
+    for (type in levels(df_col_oocytes$type)[-1]){
+      tmp <- sample_cells(
+        rownames(df_col_oocytes[df_col_oocytes$type == type,]), ncells_max)
+      selected_cells <- c(selected_cells, tmp)
+    }
+    df_col_oocytes <- df_col_oocytes[selected_cells,]
+  }
+
+  column_ha_cell_type = HeatmapAnnotation(
+    oocytes_cell_type = df_col_oocytes$type,
+    col = list(Adult_cell_type = oocytes_colors),      
+    border = TRUE,
+    simple_anno_size = unit(0.15, "cm"),
+    annotation_name_gp = gpar(fontsize = 0),
+    annotation_legend_param = legends_param_sc)
+
+  column_ha_stage = HeatmapAnnotation(
+    stage = df_col_oocytes$stage,
+    col = list(stage = stage_colors),
+    border = TRUE, 
+    simple_anno_size = unit(0.15, "cm"), 
+    annotation_legend_param = legends_param_sc,
+    annotation_name_gp = gpar(fontsize = 0))
+
+  column_ha_sex = HeatmapAnnotation(
+    sex = df_col_oocytes$sex,
+    col = list(sex = sex_colors),
+    annotation_legend_param = legends_param_sc,
+    border = TRUE, 
+    simple_anno_size = unit(0.15, "cm"), 
+    annotation_name_gp = gpar(fontsize = 0))
+
+  gene_type <- CT_genes_types %>% 
+    filter(gene %in% genes)
+
+  if (split_by == "regulation") {
+    gene_type <- gene_type %>% arrange(Regulation)
+    rowSplit <- gene_type$Regulation
+  }
+  if (split_by == "testis_specificity") {
+    gene_type <- gene_type %>% arrange(Testis_specificity)
+    rowSplit <- gene_type$Testis_specificity
+  }
+  if (split_by == "chr_met") {
+    gene_type <- gene_type %>% arrange(chr_met)
+    rowSplit <- gene_type$chr_met
+  }
+  if (!clusterRow & split_by == "chr_met") {
+    gene_type$gene <- factor(gene_type$gene, levels = genes)
+    gene_type <- gene_type  %>%
+      arrange(gene)
+    gene_type$gene <- as.vector(gene_type$gene)
+    rowSplit <- gene_type$chr_met
+  }
+
+  Regulation <- as.matrix(gene_type$Regulation)
+  rownames(Regulation) <- gene_type$gene
+  row_ha_regulation <- rowAnnotation(
+    Regulation = Regulation,
+    annotation_legend_param = legends_param_sc,
+    border = FALSE,
+    simple_anno_size = unit(0.15, "cm"), 
+    col = list(Regulation = Regulation_colors),
+    annotation_name_gp = gpar(fontsize = 0))
+
+  chr_met <- as.matrix(gene_type$chr_met)
+  rownames(chr_met) <- gene_type$gene
+  row_ha_chr_met <- rowAnnotation(
+    chr_met = chr_met,
+    annotation_legend_param = legends_param_sc,
+    border = FALSE,
+    simple_anno_size = unit(0.15, "cm"), 
+    col = list(chr_met = chr_met_colors),
+    annotation_name_gp = gpar(fontsize = 0))
+
+  Chr <- as.matrix(gene_type$Chr)
+  rownames(Chr) <- gene_type$gene
+  row_ha_chr <- rowAnnotation(Chr = Chr,
+                              annotation_legend_param = legends_param_sc,
+                              border = FALSE,
+                              simple_anno_size = unit(0.15, "cm"), 
+                              col = list(Chr = Chr_colors),
+                              annotation_name_gp = gpar(fontsize = 0))
+
+  mat <-  as.matrix(
+    assay(adult_oocytes_sce, "logcounts")[gene_type$gene, rownames(df_col_oocytes)])
+
+  if (is.null(scale_lims)) {
+    scale_lims <- c(0, quantile(rowMaxs(mat), 0.8))
+  }
+
+  heatmap_colors <- colorRamp2(
+    seq(scale_lims[1], scale_lims[2], length = 11),                 
+    legend_colors)
+
+  ht_oocytes <- Heatmap(mat,
+                      name = "oocytes logcounts",
+                      column_title = "Adult oocytes",
+                      column_split = df_col_oocytes$type,
+                      col = heatmap_colors,
+                      border = TRUE,
+                      border_gp = gpar(lwd = 0.2),
+                      row_gap = unit(c(0.5), "mm"),
+                      column_gap = unit(c(0.5), "mm"),
+                      show_column_dend = FALSE,
                       show_row_dend = FALSE,
+                      row_names_gp = gpar(fontsize = font_size),
+                      row_names_side = "left",
+                      cluster_rows = clusterRow,
+                      clustering_method_rows = clust_method,
+                      row_dend_reorder = TRUE,
+                      cluster_columns = FALSE,
                       show_column_names = FALSE,
-        cluster_columns = TRUE,
-        show_column_dend = FALSE,
-        row_names_gp = gpar(fontsize = 4),
-        heatmap_legend_param = legends_param,
-        top_annotation = c(column_ha_type),
-        left_annotation = left_annot)
-```
-
-Next graph was removed from paper
-
-```{r all_meth}
-genes <- CT_genes$external_gene_name
-database <- mean_methylation_in_tissues[rownames(mean_methylation_in_tissues) %in% genes]
-
-mat <- na.omit(assay(database))
-clustering_option <- TRUE
-
-row_ha_chr_meth <- rowAnnotation(chr = chr_mat[rownames(mat),],
-                            annotation_legend_param = legends_param,
-                            simple_anno_size = unit(0.5, "cm"),
-                            col = list(chr = chr_colors),
-                            annotation_name_gp = gpar(fontsize = 8),
-                            annotation_name_side = "top")
-
-row_ha_reg_meth <- rowAnnotation(regulation = regulation_mat[rownames(mat),],
-                            annotation_legend_param = legends_param,
-                            simple_anno_size = unit(0.5, "cm"),
-                            col = list(regulation = meth_colors),
-                            annotation_name_gp = gpar(fontsize = 8),
-                            annotation_name_side = "top")
-
-left_annot_meth <- c(row_ha_chr_meth, row_ha_reg_meth, gap = unit(1, "mm"))
-
-split_meth <- data.frame(filter(CT_genes, external_gene_name %in% rownames(mat))$regulated_by_methylation, 
-                    filter(CT_genes, external_gene_name %in% rownames(mat))$X_linked)
-
-Heatmap(mat,
-        column_title = 'Promoter mean methylation level by tissue',
-        name = 'Meth',
-        col = colorRamp2(c(1:100),
-                         colorRampPalette(c("moccasin","dodgerblue4"))(100)),
-        na_col = "gray80",
-        cluster_rows = clustering_option,
-        cluster_columns = FALSE,
-        row_split = split_meth,
+                      show_row_names = TRUE,
+                      row_split = rowSplit,
+                      cluster_row_slices = FALSE,
+                      cluster_column_slices = FALSE,
+                      top_annotation = c(column_ha_sex, 
+                                         column_ha_stage,
+                                         column_ha_cell_type),  
+                      # right_annotation = c(row_ha_regulation, 
+                      #                      row_ha_chr, 
+                      #                      gap = unit(1, "mm")),
+                      column_title_gp = gpar(fontsize = 5),
                       row_title_gp = gpar(fontsize = 0),
-        show_row_names = TRUE,
-        show_heatmap_legend = TRUE,
-        show_row_dend = FALSE,
-        row_names_gp = gpar(fontsize = 3),
-        column_names_gp = gpar(fontsize = 8),
-        column_names_side = "bottom",
-        row_names_side = "right",
-        left_annotation = left_annot_meth)
-
-```
-
-### MAGE genes
-
-```{r MAGE_TCGA}
-MAGE_genes <- filter(CT_genes, family == "MAGE")$external_gene_name
-
-TCGA_expression(tumor = "all", 
-                genes = MAGE_genes,
-                units = "log_TPM")
-
-CCLE_expression(genes = MAGE_genes,
-                 type = c("lung", "skin", "bile_duct", "bladder", 
-                          "colorectal", "lymphoma", "uterine",
-                          "myeloma", "kidney", "pancreatic", "brain", 
-                          "gastric", "breast", "bone", "head_and_neck", 
-                          "ovarian", "sarcoma", "leukemia", "esophageal",
-                          "neuroblastoma"), units = "log_TPM")
-
-normal_tissues_mean_methylation(MAGE_genes, na.omit = TRUE)
-
-```
-
-## Single Cell analysis : timing of expression
-
-In CTexploreR, we there is single cell data from the testis, we thus can 
-analyse CT genes expression during spermatogenesis.
-
-There is only data for `r sum(CT_genes$external_gene_name %in% unique(rownames(testis_sce)))` of our `r nrow(CT_genes)` CT genes.
-
-NB : SSC, Spermatogonia and early spermatocytes are *premeiotic cells*. Late 
-spermatocytes (between both meiosis), round spermatid, elongated spermatid and 
-sperm are *postmeiotic cells*.
-
-```{r repartition_sc}
-genes_avail <- 
-  CT_genes[CT_genes$external_gene_name %in% unique(rownames(testis_sce)), ]
-CT_genes_avail <- left_join(genes_avail, data.frame(rowData(testis_sce)))
-table(CT_genes_avail$testis_cell_type)
-table(CT_genes_avail$testis_cell_type)/dim(CT_genes_avail)[1]*100
-```
-
-`r sum(table(CT_genes_avail$testis_cell_type)[c("Early_spermatocyte", "Spermatogonia", "SSC")])/dim(genes_avail)[1]*100`
-% of genes are mainly expressed pre-meioticly.
-
-```{r sc_heatmap, message = FALSE}
-germ_cells <- c("SSC", "Spermatogonia", "Early_spermatocyte",
-                "Late_spermatocyte","Round_spermatid", "Elongated_spermatid",
-                "Sperm1", "Sperm2")
-somatic_cells <- c("Macrophage", "Endothelial", "Myoid", "Sertoli", "Leydig")
-
-testis_sce_CT <- testis_sce[CT_genes_avail$external_gene_name, ]
-
-mat <- SingleCellExperiment::logcounts(testis_sce_CT)
-
-df_col <- data.frame(clusters = colData(testis_sce_CT)$clusters,
-                     type = colData(testis_sce_CT)$type,
-                     Donor = colData(testis_sce_CT)$Donor)
-rownames(df_col) <- colnames(testis_sce_CT)
-
-df_col <- df_col[order(df_col$type),]
-df_col$lineage <- "Germ cells"
-df_col$lineage[df_col$type %in% somatic_cells] <- "Somatic cells"
-
-column_ha_type = HeatmapAnnotation(
-  type = df_col$type,
-  border = TRUE,
-  col = list(type = c("SSC" = "floralwhite", "Spermatogonia" = "moccasin",
-                   "Early_spermatocyte" = "gold", 
-                   "Late_spermatocyte" = "orange",
-                   "Round_spermatid" = "red2", 
-                   "Elongated_spermatid" = "darkred",
-                   "Sperm1" = "violet", "Sperm2" = "purple", 
-                   "Sertoli" = "gray", 
-                   "Leydig" = "cadetblue2", "Myoid" = "springgreen3", 
-                   "Macrophage" = "gray10",
-                   "Endothelial" = "steelblue")),
-
-  annotation_name_gp = gpar(fontsize = 8),
-  annotation_legend_param = legends_param)
-
-column_ha_lineage = HeatmapAnnotation(
-  lineage = df_col$lineage,
-  border = TRUE,
-  col = list(lineage = c("Germ cells" = "salmon", "Somatic cells" = "cyan4")),
-  annotation_name_gp = gpar(fontsize = 8),
-  annotation_legend_param = legends_param)
-
-scale_lims <- c(0, quantile(rowMax(mat), 0.75))
-top_annot <- c(column_ha_lineage, column_ha_type)
-
-# Until here is what's in the function, hereunder is my addition/change in Heatmap()
-
-chr_mat <- as.matrix(CT_genes_avail$X_linked)
-chr_mat <- ifelse(chr_mat == TRUE, "X-linked", "Not X")
-rownames(chr_mat) <- CT_genes_avail$external_gene_name
-row_ha_chr <- rowAnnotation(chr = chr_mat,
-                            annotation_legend_param = legends_param,
-                            simple_anno_size = unit(0.5, "cm"),
-                            col = list(chr = chr_colors),
-                            annotation_name_gp = gpar(fontsize = 8),
-                            annotation_name_side = "top")
-
-regulation_mat <- as.matrix(CT_genes_avail$regulated_by_methylation)
-regulation_mat <- ifelse(regulation_mat == TRUE, "Methylation",
-                         "Not methylation")
-rownames(regulation_mat) <- CT_genes_avail$external_gene_name
-
-row_ha_reg <- rowAnnotation(regulation = regulation_mat,
-                            annotation_legend_param = legends_param,
-                            simple_anno_size = unit(0.5, "cm"),
-                            col = list(regulation = meth_colors),
-                            annotation_name_gp = gpar(fontsize = 8),
-                            annotation_name_side = "top")
-
-left_annot <- c(row_ha_chr, row_ha_reg, gap = unit(1, "mm"))
-
-split <- data.frame(CT_genes_avail$regulated_by_methylation, CT_genes_avail$X_linked)
-
-Heatmap(mat[CT_genes_avail$external_gene_name, rownames(df_col), drop = FALSE],
-        name = "logCounts",
-        column_title = "Expression in testis cells (scRNAseq)",
-        column_split = df_col$type,
-        row_split = split,
-        row_title_gp = gpar(fontsize = 0),
-        show_column_names = FALSE,
-        show_column_dend = FALSE,
-        clustering_method_rows = "ward.D",
-        clustering_method_columns = "ward.D",
-        cluster_rows = TRUE,
-        cluster_columns = FALSE,
-        show_row_dend = FALSE,
-        row_names_gp = gpar(fontsize = 4),
-        col = colorRamp2(seq(scale_lims[1], scale_lims[2], length = 11),                 
-                         legend_colors),
-        top_annotation = top_annot,
-        left_annotation = left_annot,
-        heatmap_legend_param = legends_param)
-
+                      gap = unit(0.01, "mm"),
+                      use_raster = TRUE,
+                      raster_device = "CairoPNG",
+                      raster_quality = 10,
+                      #heatmap_width = unit(h_width, "cm"),
+                      #heatmap_height = unit(h_height, "cm"),
+                      show_heatmap_legend = show_legend,
+                      heatmap_legend_param = legends_param_sc)
+
+  return(ht_oocytes)
+}
+```
+
+```{r draw_healthy_cells_exprs}
+# Manual adjustments to visualise
+
+X_met_new_order <- 
+  c("SPANXB1", "SPANXD", "SPANXC", "SPANXN3", "CXorf51B", "H2AB1", "DDX53", 
+    "TGIF2LX", "FTHL17","VCX3B", "VCX", "VCX3A", "VCX2", "FMR1NB",  "PAGE2", 
+    "PAGE5", "MAGEC2","GAGE2A","PAGE1", "SSX3", "FAM9A",  "CTAG2",  "MAGEA4", 
+    "PASD1","MAGEB2","CT55","MAGEA10",  "MAGEA3",  "MAGEA6","GAGE1", "MAGEA1", 
+    "CT45A10", "MAGEA11", "SSX1", "MAGEA9B","MAGEA9", "MAGEA8", "LUZP4",
+    "MAGEB1", "MAGEA12", "MAGEC1", "MAGEB16","GAGE12E",  "XAGE5", "GAGE13", 
+    "SSX5", "CT45A1", "SSX2B", "XAGE1B", "RHOXF2B", "DCAF8L2", "SSX2", "SAGE1", 
+    "CXorf49", "CTAG1A", "NXF2","CT47A6", "CT45A5",  "CTAG1B" )
+
+not_X_met_new_order <- 
+  c("SPMIP9", "FLJ36000", "TRIML1", "TRIML2", "ADAD1", "TUBA3C", "COX7B2", 
+    "DDX4", "BRDT", "DSCR8", "DMRTB1",  "DAZL","RBM46", "DCAF4L2",  "CCDC172",
+    "LINC02475", "TGIF2LY", "ASZ1", "RNF17", "PRDM9", "CTCFL", "LINC00470", 
+    "POTEC", "TMEM132D-AS1", "DPPA2", "LINC01413", "NLRP4", "LINC01030", 
+    "ZSCAN5B", "DPPA5", "LINC01518", "LINC00200", "RFPL4B", "KRT85", "POTEF", 
+    "GFY", "RBMY1E", "RBMY1A1")
+
+X_not_met_new_order <- c( "SPANXN4", "DUSP21", "ESX1", "MAGEB4")
+
+not_X_not_met_new_order <- 
+  c("SMCP","LINC01548","WFDC11", "LINC01467","LINC00408", "LINC01193", "ADAM29", "SPATA16",
+    "IZUMO3", "LIN28A","C1orf94", "NUTM1", "GOLGA8S", "LINC01206", "ADAM18", 
+    "ADAM2","SYCP1", "MEIOB", "PRDM7", "TSPY3", 
+    "TSPY10", "TSPY8", "TSPY2","DMRT1", "LINC01511", "TSPY4", "TSPY1")
+
+my_genes <- c(X_met_new_order, not_X_met_new_order, X_not_met_new_order, not_X_not_met_new_order)
+
+fs <- 2
+ncells <- 100
+scale_limits <- c(0, 4)
+ht_testis <- draw_testis_heatmap(my_genes,
+                    ncells_max = ncells,
+                    font_size = fs,
+                    scale_lims = scale_limits,
+                    clusterRow  = FALSE,
+                    split_by = "chr_met",
+                    show_legend = TRUE,
+                    clust_method = "ward.D2")
+
+female_FGC_ht <- draw_FGC_heatmap(my_genes, 
+                 ncells_max = ncells,
+                 sex = c("F"),
+                 clusterRow = FALSE, 
+                 font_size = fs,
+                 scale_lims = scale_limits, 
+                 show_legend = FALSE,
+                 show_right_annot = FALSE)
+
+male_FGC_ht <- draw_FGC_heatmap(my_genes, 
+                 ncells_max = ncells,
+                 sex = c("M"),
+                 clusterRow = FALSE, 
+                 font_size = fs,
+                 scale_lims = scale_limits, 
+                 show_legend = FALSE,
+                 show_right_annot = FALSE)
+
+ht_oocytes <- draw_oocytes_heatmap(my_genes, 
+                    ncells_max = ncells,
+                    font_size = fs,
+                    scale_lims = scale_limits, 
+                    clusterRow = FALSE,
+                    show_legend = FALSE,
+                    clust_method = "centroid")
+
+ht_list <- female_FGC_ht + ht_oocytes + male_FGC_ht + ht_testis
+draw(ht_list, main_heatmap = "logcounts", merge_legend = TRUE, ht_gap = unit(0.2, "cm"))
+
+```
+
+```{r, echo = FALSE, eval = FALSE}
+pdf(file = paste0("figs/CT_in_GermCells.pdf"), width = 6, height = 4)
+draw(ht_list, main_heatmap = "Logcounts", 
+     merge_legend = TRUE, 
+     ht_gap = unit(0.2, "cm"), 
+     legend_gap = unit(0.15, "cm"),
+     legend_grid_height = unit(0.08, "cm"),
+    legend_grid_width = unit(0.1, "cm"))
+dev.off()
 ```

### 2024-05-17 Paper_code-CT_analysis.Rmd
Vefified and corrected all figs

@@
 library(patchwork)
-library(BiocParallel)
-library(Biostrings)
 library(AUCell)
@@
   mutate(Regulation = case_when(regulated_by_methylation ~ "Methylation",
-                                !regulated_by_methylation ~ "Not methylation")) %>%
+                                !regulated_by_methylation ~ 
+                                  "Not methylation")) %>%
   mutate(Chr = case_when(X_linked ~ "X-linked",
@@
                              regulated_by_methylation & !X_linked ~ "not_X_met",
-                             !regulated_by_methylation & !X_linked ~ "not_X_not_met")) %>% 
-  mutate(specificity = case_when(GTEX_category == "testis_specific" ~ "Testis specific",
-                                        GTEX_category == "testis_preferential" ~ "Testis preferential",
-                                        GTEX_category == "lowly_expressed" ~ "Undetectable in GTEX")) %>% 
+                             !regulated_by_methylation & !X_linked ~ 
+                               "not_X_not_met")) %>% 
+  mutate(specificity = case_when(GTEX_category == "testis_specific" ~ 
+                                   "Testis specific",
+                                 GTEX_category == "testis_preferential" ~ 
+                                   "Testis preferential",
+                                 GTEX_category == "lowly_expressed" ~ 
+                                   "Undetectable in GTEX")) %>% 
   mutate(specificity = factor(specificity,
-                                     levels = c("Testis specific",  "Undetectable in GTEX", "Testis preferential"))) %>% 
+                              levels = c("Testis specific",  
+                                         "Undetectable in GTEX", 
+                                         "Testis preferential"))) %>% 
   mutate(chr_met = factor(chr_met, 
-                          levels = c("X_met", "not_X_met", "X_not_met", "not_X_not_met")))
+                          levels = c("X_met", "not_X_met", "X_not_met", 
+                                     "not_X_not_met")))
 
@@
 
+DAC_treatment_colors <- c("CTL" = "royalblue4", "5-aza" = "maroon3")
+
 Regulation_colors <- c("Methylation" = "indianred2", 
@@
 
-Chr_colors = c("X-linked" = "deeppink", "not X" = "lightgreen")
+Chr_colors = c("X-linked" = "deeppink3", "not X" = "lightgreen")
 
@@
 Fetal_cell_colors = c( "F_PGC" = "pink", 
-                       "F_GC" = "pink4", 
+                       "F_GC" = "pink3", 
                        "F_oogonia" = "palevioletred3",
@@
                        "M_GC" = "steelblue3",
-                       "M_pre_spermatogonia" = "darkblue")
-
-oocytes_colors <- c("Growing oocytes" = "paleturquoise3", 
+                       "M_pre_spermatogonia" = "royalblue3")
+
+oocytes_colors <- c("Growing oocytes" = "mediumpurple1",
                     # "Fully grown oocytes" ="lightgreen",
                      # "Metaphase I" = "mediumvioletred",
-                  "Metaphase II" = "mediumvioletred")
-
-sex_colors <- c("F" = "palevioletred1", "M"= "skyblue2")
-
-stage_colors <- c( "pre-meiotic" = "moccasin", 
+                    "Metaphase II" = "mediumpurple4")
+
+sex_colors <- c("F" = "palevioletred1", "M"= "skyblue1")
+
+stage_colors <- c( "pre-meiotic" = "peachpuff", 
                    "meiotic"= "lightsalmon",
@@
 Function to print GTEX expression with specific annotations.
+
 ```{r GTEX_function}
@@
 
-```{r meth_value_heatmap}
+```{r meth_heatmap_function}
 met_genes <- CT_genes %>% 
@@
 
-```{r DAC_heatmap}
+```{r DAC_heatmap_function}
 DAC_fig <- function(genes = gene_type$gene, multimapping = TRUE, 
@@
     treatment = df_col$treatment,
-    col = list(treatment = c("CTL" = "royalblue4", "5-aza" = "maroon3")),
+    col = list(treatment = DAC_treatment_colors),
     annotation_legend_param = legends_param,
@@
                  units = "log1p",
-                 row_fontsize = fs
-)
+                 row_fontsize = fs)
 
@@
   geom_text(aes(y = value/5 + c(0, cumsum(value)[-length(value)]), 
-                label = rev(paste0(Regulation, "\n", value, "/", nrow(CT_genes_types), "\n",
+                label = rev(paste0(Regulation, "\n", value, "/", 
+                                   sum(table(CT_genes_types$Regulation)), "\n",
                                round(Percent * 100), "%"))))+
@@
 
-fig_X_enrichment <- ggplot(CT_genes_types, aes(x = chr, fill = Regulation)) +
+fig_X_enrichment <- ggplot(CT_genes_types[
+  !is.na(CT_genes_types$regulated_by_methylation), ], 
+                           aes(x = chr, fill = Regulation)) +
   geom_bar(stat = 'count') +
@@
   geom_text(aes(y = value/5 + c(0, cumsum(value)[-length(value)]),
-                label = rev(paste0(Regulation, "\n", value, "/", nrow(CT_genes_types), "\n",
+                label = rev(paste0(Regulation, "\n", value, "/", 
+                                   sum(table(CT_genes_types$Regulation)), "\n",
                                round(Percent * 100), "%"))), size = 1.2 )+
@@
 
-fig_freq_by_chr <- ggplot(CT_genes_types, aes(x = chr, fill = Regulation)) +
+fig_freq_by_chr <- ggplot(CT_genes_types[
+  !is.na(CT_genes_types$regulated_by_methylation), ], 
+  aes(x = chr, fill = Regulation)) +
   geom_bar(stat = 'count') +
@@
   right_join(CT_genes_types) %>% 
-  dplyr::select(gene, mean_log, proportion_of_positive_tum, Regulation, X_linked) %>% 
+  dplyr::select(gene, mean_log, proportion_of_positive_tum, 
+                Regulation, X_linked) %>% 
   mutate(source = "TCGA tumor samples")
@@
 # combine with CCLE data
-load("~/cluster/Packages/CTdata/eh_data/CCLE_data.rda")
 ccle <- CCLE_data
@@
 
-fig_mean_prop_in_tum <- ggplot(mean_and_prop,
+fig_mean_prop_in_tum <- ggplot(mean_and_prop[!is.na(mean_and_prop$Regulation), ],
                                aes(x = mean_log, y = proportion_of_positive_tum,
@@
                column_title = levels(df_col$project_id),
-               column_title_gp = gpar(col =c("COAD" = "darkorchid2", 
-                                             "HNSC" = "deeppink2",
-                                             "LUSC" = "seagreen3", 
-                                             "SKCM" = "red3"), 
+               column_title_gp = gpar(col = TCGA_colors[tum_type], 
                                       fontsize = 3, face = "bold"),
@@
                                         scale_lims = c(0, 6),
-                                        # tum_type = c("SKCM", "LUAD", "COAD", "BRCA"),
                                         tum_type = tum_types,
@@
                                        scale_lims = c(0, 6),
-                                       #  tum_type = c("SKCM", "LUAD", "COAD", "BRCA"),
                                        tum_type = tum_types,
@@
   mutate(project_id = factor(project_id, levels = tum_types,)) %>% 
-  ggplot(aes(x = global_methylation, y = AUC_genes_regulated_by_met, color = project_id)) +
+  ggplot(aes(x = global_methylation, y = AUC_genes_regulated_by_met, 
+             color = project_id)) +
   geom_point(size = 0.3) +
   facet_wrap(~ project_id, nrow = 1) +
-  scale_color_manual(values = c("COAD" = "darkorchid2", "HNSC" = "deeppink2",
-                                "LUSC" = "seagreen3", "SKCM" = "red3")) + 
+  scale_color_manual(values = TCGA_colors) + 
   geom_smooth(method = 'lm', color = "black", linewidth = 0.3, se = FALSE) +
@@
   mutate(project_id = factor(project_id, levels = tum_types,)) %>% 
-  ggplot(aes(x = global_methylation, y = AUC_genes_not_regulated_by_met, color = project_id)) +
+  ggplot(aes(x = global_methylation, y = AUC_genes_not_regulated_by_met, 
+             color = project_id)) +
   geom_point(size = 0.3) +
   facet_wrap(~ project_id, nrow = 1) +
-  scale_color_manual(values = c("COAD" = "darkorchid2", "HNSC" = "deeppink2",
-                                "LUSC" = "seagreen3", "SKCM" = "red3")) + 
+  scale_color_manual(values = TCGA_colors) + 
   geom_smooth(method = 'lm', color = "black", linewidth = 0.3, se = FALSE) +
@@
 ```
+
 # CT expression in healthy cells (fetal and adult)
@@
 
-```{r}
+```{r, testis_data}
 ################################################################################
@@
 
-#######################################################################
+################################################################################
 ## Function to draw expression in adult testis cells with specific annotations
 ## and downasampling the number of cells from each cell type
-#######################################################################
+################################################################################
 draw_testis_heatmap <- function(genes, 
@@
 
-######################################################################
+################################################################################
 ## Function to draw expression in adult oocytes cells with specific annotations
 ## and downasampling the number of cells from each cell type
-#######################################################################
+################################################################################
 draw_oocytes_heatmap <- function(genes, 
@@
   column_ha_cell_type = HeatmapAnnotation(
-    oocytes_cell_type = df_col_oocytes$type,
-    col = list(Adult_cell_type = oocytes_colors),      
+    Oocytes_cell_type = df_col_oocytes$type,
+    col = list(Oocytes_cell_type = oocytes_colors),      
     border = TRUE,

### 2024-05-17 Paper_code-CT_analysis.Rmd
Added legend in TCGA heatmap

@@
 draw(h1, column_title = "CT genes regulated by methylation",
-     column_title_gp = gpar(fontsize = 5), show_heatmap_legend = FALSE,
-     show_annotation_legend = FALSE)
+     column_title_gp = gpar(fontsize = 5), show_heatmap_legend = TRUE,
+     show_annotation_legend = TRUE, heatmap_legend_side = "bottom",
+     merge_legend = TRUE)
 dev.off()

### 2024-05-22 Paper_code-CT_analysis.Rmd
Changed some figures

--- 2024-05-17 Paper_code-CT_analysis.Rmd
@@
 ```{r load_data, warning=FALSE, message=FALSE}
-# CCLE_data <- CTdata::CCLE_data()
-# TCGA_TPM <- CTdata::TCGA_TPM()
-# testis_sce <- CTdata::testis_sce()
-####
-
 load("../CTdata/eh_data/all_genes.rda")
@@
 
-```{r, eval = FALSE, echo=FALSE}
+```{r, eval = TRUE, echo=FALSE}
 pdf(file = paste0("figs/GTEX_Tumors.pdf"))
@@
 
-```{r, eval = FALSE, echo = FALSE}
+```{r, eval = TRUE, echo = FALSE}
 pdf(file = paste0("figs/Multimapping.pdf"))
@@
                border_gp = gpar(lwd = 0.5),
-               row_title_gp = gpar(col = c("indianred2", "cyan4"), fontsize = 6),
+               row_title_gp = gpar(col = c("indianred2", "cyan4"), fontsize = 8),
                row_split = factor(gene_type$Regulation,
@@
 
-```{r, eval = FALSE, echo = FALSE}
+```{r, eval = TRUE, echo = FALSE}
 pdf(file = paste0("figs/Methylation_classification.pdf"))
@@
 
-```{r, eval = FALSE, echo=FALSE}
+```{r, eval = TRUE, echo=FALSE}
 pdf(file = paste0("figs/X_enrichment.pdf"), width = 7, height = 4)
@@
 ```
+
 # CT genes in tumour : coexpression and link with global hypomethylation
@@
   ylab("Proportion of positive samples (%)") +
-  facet_wrap(~ source, ncol = 1) +
+  facet_wrap(~ source, ncol = 2) +
   theme_bw() +
@@
 ```
-```{r, eval = FALSE, echo = FALSE}
+```{r, eval = TRUE, echo = FALSE}
 pdf(file = paste0("figs/Mean_prop_in_tumors.pdf"), width = 5, height = 4)
@@
 
-```{r, echo = FALSE, eval = FALSE}
+```{r, echo = FALSE, eval = TRUE}
 pdf(file = "figs/CT_met_AUC_hypo_TCGA.pdf", height = 2.3, width = 3)
 draw(h1, column_title = "CT genes regulated by methylation",
-     column_title_gp = gpar(fontsize = 5), show_heatmap_legend = FALSE,
-     show_annotation_legend = FALSE)
+     column_title_gp = gpar(fontsize = 5), show_heatmap_legend = TRUE,
+     show_annotation_legend = TRUE, heatmap_legend_side = "bottom",
+     merge_legend = TRUE)
 dev.off()
@@
 
-```{r}
+```{r, echo = FALSE}
 pdf(file = paste0("figs/CT_met_AUCscore_hypo_TCGA.pdf"), height = 2.5)
@@
 
-```{r, testis_data}
-################################################################################
+```{r testis_data}
 # Load testis scRNAseq data
-################################################################################
 
@@
 
-################################################################################
-# Load FGC and oocytes scRNAseq data
-################################################################################
+## Load FGC and oocytes scRNAseq data
 
@@
 legends_param_sc <- list(
-  labels_gp = gpar(col = "black", fontsize = 2),
-  title_gp = gpar(col = "black", fontsize = 3),
+  labels_gp = gpar(col = "black", fontsize = 3),
+  title_gp = gpar(col = "black", fontsize = 4),
   simple_anno_size = unit(0.15, "cm"),
@@
 ```{r healthy_cells_function}
-#######################################################################
+
 # Function to downsample cell types
-#######################################################################
+
 sample_cells <- function(barcodes, nmax){
@@
 
-#######################################################################
 ## Function to draw expression in FGC with specific annotations
 ## and downasampling the number of cells from each cell type
-#######################################################################
+
 draw_FGC_heatmap <- function(genes, 
@@
                              h_height = 10,
-                             show_right_annot = TRUE){
+                             show_row_annot = TRUE, 
+                             row_title = TRUE){
 
@@
 
-  if (!show_right_annot) rAnnot <- NULL
+  if (!show_row_annot) rAnnot <- NULL
+
+  if (row_title) rtitle <- gpar(fontsize = 5, fontface = "bold") else rtitle <- gpar(fontsize = 0)
 
@@
                                         column_ha_type),  
-                     right_annotation = rAnnot,
+                     left_annotation = rAnnot,
                      column_title_gp = gpar(fontsize = 5),
-                     row_title_gp = gpar(fontsize = 0),
+                     row_title_gp = rtitle,
                      gap = unit(0.01, "mm"),
@@
                      show_heatmap_legend = show_legend,
-                     heatmap_legend_param = legends_param_sc)
+                     heatmap_legend_param = legends_param_sc,
+                     row_title_rot = 0)
 
@@
 
-################################################################################
 ## Function to draw expression in adult testis cells with specific annotations
 ## and downasampling the number of cells from each cell type
-################################################################################
+
 draw_testis_heatmap <- function(genes, 
@@
                                 h_height = 10,
-                                clust_method = "ward.D"){
+                                clust_method = "ward.D",
+                                show_row_annot = TRUE){
 
@@
                               annotation_name_gp = gpar(fontsize = 0))
+
+  rAnnot <- c(row_ha_regulation, row_ha_chr, gap = unit(1, "mm"))
+
+  if (!show_row_annot) rAnnot <- NULL
 
@@
                                           column_ha_cell_type),  
-                       right_annotation = c(row_ha_regulation, 
-                                            row_ha_chr, 
-                                            gap = unit(1, "mm")),
+                       right_annotation = rAnnot,
                        column_title_gp = gpar(fontsize = 5),
@@
 
-################################################################################
 ## Function to draw expression in adult oocytes cells with specific annotations
 ## and downasampling the number of cells from each cell type
-################################################################################
+
 draw_oocytes_heatmap <- function(genes, 
@@
                     show_legend = TRUE,
-                    clust_method = "ward.D2")
+                    clust_method = "ward.D2",
+                    show_row_annot = FALSE)
 
@@
                  show_legend = FALSE,
-                 show_right_annot = FALSE)
+                 show_row_annot = TRUE,
+                 row_title = TRUE)
 
@@
                  show_legend = FALSE,
-                 show_right_annot = FALSE)
+                 show_row_annot = FALSE,
+                 row_title = FALSE)
 
@@
 ht_list <- female_FGC_ht + ht_oocytes + male_FGC_ht + ht_testis
-draw(ht_list, main_heatmap = "logcounts", merge_legend = TRUE, ht_gap = unit(0.2, "cm"))
-
-```
-
-```{r, echo = FALSE, eval = FALSE}
+draw(ht_list, main_heatmap = "Logcounts", merge_legend = TRUE, 
+     ht_gap = unit(0.2, "cm"))
+
+```
+
+```{r, echo = FALSE, eval = TRUE}
 pdf(file = paste0("figs/CT_in_GermCells.pdf"), width = 6, height = 4)
@@
 
-## Gene function
+# Gene function
 

### 2024-05-23 Paper_code-CT_analysis.Rmd
Added some heatmaps

@@
 
+all_genes_types <- all_genes %>% 
+  dplyr::rename(gene = external_gene_name) %>% 
+  mutate(Regulation = case_when(regulated_by_methylation ~ "Methylation",
+                                !regulated_by_methylation ~ 
+                                  "Not methylation")) %>%
+  mutate(Chr = case_when(X_linked ~ "X-linked",
+                         !X_linked ~ "not X")) %>% 
+  mutate(chr_met = case_when(regulated_by_methylation & X_linked ~ "X_met",
+                             !regulated_by_methylation & X_linked ~ "X_not_met",
+                             regulated_by_methylation & !X_linked ~ "not_X_met",
+                             !regulated_by_methylation & !X_linked ~ 
+                               "not_X_not_met")) %>% 
+  mutate(specificity = case_when(GTEX_category == "testis_specific" ~ 
+                                   "Testis specific",
+                                 GTEX_category == "testis_preferential" ~ 
+                                   "Testis preferential",
+                                 GTEX_category == "lowly_expressed" ~ 
+                                   "Undetectable in GTEX")) %>% 
+  mutate(specificity = factor(specificity,
+                              levels = c("Testis specific",  
+                                         "Undetectable in GTEX", 
+                                         "Testis preferential"))) %>% 
+  mutate(chr_met = factor(chr_met, 
+                          levels = c("X_met", "not_X_met", "X_not_met", 
+                                     "not_X_not_met")))
+
 ```
@@
 
+# Genes that were selected with GTEx but leaky so removed (ubiquitary expression)
+all_genes %>% 
+  filter(GTEX_category %in% c("testis_specific", "testis_preferential")) %>% 
+  filter(TCGA_category == "leaky" | CCLE_category == "leaky") %>% 
+  dim()
+
 specific_to_none <- all_genes %>%
@@
 
-  gene_type <- CT_genes_types %>% 
+  gene_type <- all_genes_types %>% 
     filter(gene %in% genes)
@@
   rownames(mat) <- rowData(TCGA)$external_gene_name
-  mat <- mat[CT_genes$external_gene_name,]
+  mat <- mat[rownames(mat) %in% all_genes$external_gene_name,]
   mat <- log1p(mat)
 
-  gene_type <- CT_genes_types %>% 
+  gene_type <- all_genes_types %>% 
     filter(gene %in% genes)
@@
   rownames(mat) <- rowData(CCLE)$external_gene_name
-  mat <- mat[CT_genes$external_gene_name,]
+  mat <- mat[rownames(mat) %in% all_genes$external_gene_name, ]
   mat <- log1p(mat)
 
-  gene_type <- CT_genes_types %>% 
+  gene_type <- all_genes_types %>% 
     filter(gene %in% genes)
@@
 dev.off()
+```
+
+Counter examples
+
+```{r counter_examples}
+genes_not_testis <- all_genes %>% 
+  filter(GTEX_category != "lowly_expressed") %>% 
+  filter(testis_specificity == "not_testis_specific") %>% 
+  pull(external_gene_name) %>% 
+  sample(2)
+
+genes_not_TCGA_activated <- all_genes %>% 
+  filter(GTEX_category != "lowly_expressed") %>% 
+  filter(testis_specificity != "not_testis_specific") %>% 
+  filter(!TCGA_category %in% c("activated", "multimapping_issue")) %>% 
+  pull(external_gene_name) %>% 
+  sample(1)
+
+genes_not_ccle_activated <- all_genes %>% 
+  filter(GTEX_category != "lowly_expressed") %>% 
+  filter(testis_specificity != "not_testis_specific") %>% 
+  filter(CCLE_category %in% c("activated")) %>% 
+  pull(external_gene_name) %>% 
+  sample(1)
+
+my_genes_counter <- c(genes_not_testis, genes_not_TCGA_activated, 
+                      genes_not_ccle_activated)
+
+ht_TCGA_counter <- print_TCGA_heatmap(my_genes_counter, 
+                              fontsize = 4, 
+                              downsample_to_n = 50, 
+                              h_width = 3,
+                              h_height = 8,
+                              tum_type = c("TCGA-SKCM", "TCGA-LUAD",
+                                           "TCGA-HNSC"),
+                              clust_method = "ward.D",
+                              split_rows_by = "specificity")
+
+ht_CCLE_counter <- print_CCLE_heatmap(my_genes_counter, 
+                              downsample_to_n = 50, 
+                              tum_type = c("Skin", "Lung",
+                                           "Head_and_Neck"),
+                              h_width = 3,
+                              h_height = 8,
+                              split_rows_by = "specificity")
+
+ht_GTEX_counter <- print_GTEX_heatmap(my_genes_counter, 
+                              row_fontsize = 8,
+                              column_fontsize = 5,
+                              h_width = 5,
+                              h_height = 8,
+                             split_rows_by = "specificity")
+h_list_counter<- ht_GTEX_counter + ht_CCLE_counter + ht_TCGA_counter
+draw(h_list_counter, main_heatmap = "TCGA (logTPM)", ht_gap = unit(c(0.3, 0.01),
+                                                                   "cm"),
+     merge_legend = TRUE)
+
+my_genes_leaky <-  all_genes %>% 
+  filter(GTEX_category %in% c("testis_specific", "testis_preferential")) %>% 
+  filter(TCGA_category == "leaky" | CCLE_category == "leaky") %>% 
+   pull(external_gene_name)
+
+ht_TCGA_leaky <- print_TCGA_heatmap(my_genes_leaky, 
+                              fontsize = 4, 
+                              downsample_to_n = 50, 
+                              h_width = 3,
+                              h_height = 8,
+                              tum_type = c("TCGA-SKCM", "TCGA-LUAD",
+                                           "TCGA-HNSC"),
+                              clust_method = "ward.D",
+                              split_rows_by = "specificity")
+
+ht_CCLE_leaky <- print_CCLE_heatmap(my_genes_leaky, 
+                              downsample_to_n = 50, 
+                              tum_type = c("Skin", "Lung",
+                                           "Head_and_Neck"),
+                              h_width = 3,
+                              h_height = 8,
+                              split_rows_by = "specificity")
+
+ht_GTEX_leaky <- print_GTEX_heatmap(my_genes_leaky, 
+                              row_fontsize = 8,
+                              column_fontsize = 5,
+                              h_width = 5,
+                              h_height = 8,
+                             split_rows_by = "specificity")
+h_list_leaky<- ht_GTEX_leaky + ht_CCLE_leaky + ht_TCGA_leaky
+draw(h_list_leaky, main_heatmap = "TCGA (logTPM)", ht_gap = unit(c(0.3, 0.01), "cm"),
+     merge_legend = TRUE)
 ```

### 2024-05-29 Paper_code-CT_analysis.Rmd
Changed spermatid to post meiotic cells

@@
 
-```{r, eval = TRUE, echo=FALSE}
+```{r, eval = FALSE, echo=FALSE}
 pdf(file = paste0("figs/GTEX_Tumors.pdf"))
@@
 
-```{r, eval = TRUE, echo = FALSE}
+```{r, eval = FALSE, echo = FALSE}
 pdf(file = paste0("figs/Multimapping.pdf"))
@@
 
-```{r, eval = TRUE, echo = FALSE}
+```{r, eval = FALSE, echo = FALSE}
 pdf(file = paste0("figs/Methylation_classification.pdf"))
@@
 
-```{r, eval = TRUE, echo=FALSE}
+```{r, eval = FALSE, echo=FALSE}
 pdf(file = paste0("figs/X_enrichment.pdf"), width = 7, height = 4)
@@
 
-```{r}
+```{r, eval = FALSE, echo = FALSE}
 pdf(file = paste0("figs/Methylation_panel.pdf"), width = 7, height = 4)
@@
 ```
-```{r, eval = TRUE, echo = FALSE}
+```{r, eval = FALSE, echo = FALSE}
 pdf(file = paste0("figs/Mean_prop_in_tumors.pdf"), width = 5, height = 4)
@@
 
-```{r, echo = FALSE, eval = TRUE}
+```{r, echo = FALSE, eval = FALSE}
 pdf(file = "figs/CT_met_AUC_hypo_TCGA.pdf", height = 2.3, width = 3)
@@
 
-```{r, echo = FALSE}
+```{r, eval = FALSE, echo = FALSE}
 pdf(file = paste0("figs/CT_met_AUCscore_hypo_TCGA.pdf"), height = 2.5)
@@
 testis_sce[, testis_sce$type %in% 
-             c("Early_spermatocyte", "Late_spermatocyte", 
-               "Round_spermatid")]$stage <- "meiotic"
+             c("Early_spermatocyte", "Late_spermatocyte")]$stage <- "meiotic"
 testis_sce[, testis_sce$type %in% 
-             c("Elongated_spermatid", "Sperm")]$stage <- "post-meiotic"
+             c("Round_spermatid","Elongated_spermatid", 
+               "Sperm")]$stage <- "post-meiotic"
 testis_sce[, testis_sce$type %in% 
@@
 
-```{r, echo = FALSE, eval = TRUE}
+```{r, echo = FALSE, eval = FALSE}
 pdf(file = paste0("figs/CT_in_GermCells.pdf"), width = 6, height = 4)

### 2024-05-29 Paper_code-Database_Comparison.Rmd
Added some database comparison

@@
 
+Venn_all_database <- Venn(all_database_gene_list)
+Venn_all_database@IntersectionSets[["111111"]]
+
 common <- unique(c(core_ours@IntersectionSets[["11"]], 
@@
 
+The 3 genes found in all databases are 
+`r Venn_all_database@IntersectionSets[["111111"]]`.
+
 We have lost `r length(lost_list)` genes in total. Among them, 
@@
 new <- CT_genes %>% 
-  filter(!external_gene_name%in%common)
+  filter(external_gene_name %in% Venn_all_database@IntersectionSets[["100000"]])
 

### 2024-05-30 Paper_code-Database_Comparison.Rmd
Added small analyses for database comparison

@@
 
+```{r}
+all_genes %>%
+  filter(external_gene_name %in% CTdatabase_ours@IntersectionSets[["10"]]) %>% 
+  pull(regulated_by_methylation) %>% 
+  table()
+
+all_genes %>%
+  filter(external_gene_name %in% CTdatabase_ours@IntersectionSets[["10"]]) %>% 
+  pull(transcript_biotype) %>% 
+  table()
+table(CT_genes$transcript_biotype)
+#45 lncRNA in all CT_genes
+
+all_genes %>%
+  filter(external_gene_name %in% CTdatabase_ours@IntersectionSets[["10"]]) %>% 
+  pull(CpG_promoter) %>% 
+  table()
+table(CT_genes$CpG_promoter)
+
+all_genes %>%
+  filter(external_gene_name %in% CTdatabase_ours@IntersectionSets[["10"]])
+
+```
+
 ## CTexploreR VS omics databases
@@
 table(new$X_linked, new$regulated_by_methylation)
+table(new$CpG_promoter)
 TCGA_expression(tumor = "all", genes = new$external_gene_name, 
@@
                 units = "log_TPM")
+testis_expression(genes = new$external_gene_name, "germ_cells")
 ```

### 2024-06-05 Paper_code-Database_Comparison.Rmd
Adapted database fig

@@
 
-upset_text_size <- c(1.5, 1.5, 1.5, 1.5, 1.8, 2)
+upset_text_size <- c(1.5, 1.5, 1.5, 1.5, 2, 2)
  #c(intersection size title, intersection size tick labels, set size title, 
@@
 upset_all_database <- fromList(all_database_gene_list)
+
+upset_text_size_all <- c(1.5, 1.5, 1.5, 1.5, 2, 1)
+ #c(intersection size title, intersection size tick labels, set size title, 
+ # set size tick labels, set names, numbers above bars)
 upset(upset_all_database,
@@
       nintersects = 50,
-      text.scale = upset_text_size)
+      text.scale = upset_text_size_all)
 
@@
 
-```{r, eval = FALSE, echo = FALSE}
+```{r, eval = TRUE, echo = FALSE}
 pdf(file = paste0("figs/upset_all_lost.pdf"))
@@
       nsets = 6,
-      nintersects = 50)
+      nintersects = 50,
+      text.scale = upset_text_size_all)
 dev.off()

### 2024-06-11 Paper_code-CT_analysis.Rmd
Minor change

@@
   h <- Heatmap(mat[gene_type$gene, rownames(df_col), drop = FALSE],
-               name = "logTPM",
+               name = "logCounts",
                row_split = factor(gene_type$Regulation, 

### 2024-06-21 Paper_code-Database_Comparison.Rmd
Small verif in CTdatabase analysis
Updated with new tumors thresholds

@@
 
-### Burggeman's list
+### Bruggeman's list
 
@@
 
+CTdatabase_lost$external_gene_name[!CTdatabase_lost$external_gene_name %in% unlist(CTdatabase_lost_upset)]
+
 upset_CTdatabase <- fromList(CTdatabase_lost_upset)
@@
 
-```{r, eval = FALSE, echo = FALSE}
+```{r, eval = TRUE, echo = FALSE}
 pdf(file = paste0("figs/CTdatabase_comparison.pdf"))
@@
 % are not properly activated in tumors and/or cancer cell lines.
+
+There is one gene (SPANXN1) that is missing (in 165 from CTdatabase lost) but 
+that is not on the upset plot (164 on upset plot) as it was removed because of 
+the TSS.
 
@@
 
-```{r, eval = FALSE, echo = FALSE}
+```{r, eval = TRUE, echo = FALSE}
 pdf(file = paste0("figs/Jamin_comparison.pdf"))
@@
 
-```{r, eval = FALSE, echo = FALSE}
+```{r, eval = TRUE, echo = FALSE}
 pdf(file = paste0("figs/Wang_comparison.pdf"))
@@
 
-```{r, eval = FALSE, echo = FALSE}
+```{r, eval = TRUE, echo = FALSE}
 pdf(file = paste0("figs/Carter_comparison.pdf"))
@@
 
-```{r, eval = FALSE, echo = FALSE}
+```{r, eval = TRUE, echo = FALSE}
 pdf(file = paste0("figs/Bruggeman_comparison.pdf"))

### 2024-07-02 Paper_code-CT_analysis.Rmd
Updated with new tumors thresholds
Adapted all code to latest version of paper

--- 2024-05-29 Paper_code-CT_analysis.Rmd
@@
 ---
-title: "Code used to generate paper results : Database comparison and CT characteristics"
+title: "Code used to generate paper results : CT characteristics"
 author: "Julie Devis"
@@
 library(patchwork)
-library(AUCell)
+library(Vennerable)
 ```
@@
 load("../CTdata/eh_data/oocytes_sce.rda")
-
 load("../CTdata/eh_data/CT_genes.rda")
 
-CT_genes_types <- CT_genes %>% 
+load("~/cluster/DataSets/FGC/scWGBS/processed_data/scWGBS_FGC_SE.rda")  
+load("~/cluster/CBIOProjects/202406_CT_genes_paper/tmp_data/CT_promoter_regions_hg19_GR.rda")
+
+all_genes_clean <- all_genes %>% 
   dplyr::rename(gene = external_gene_name) %>% 
@@
                                "not_X_not_met")) %>% 
-  mutate(specificity = case_when(GTEX_category == "testis_specific" ~ 
+  mutate(GTEX_specificity = case_when(GTEX_category == "testis_specific" ~ 
                                         "Testis specific",
@@
                                         "Undetectable in GTEX")) %>% 
-  mutate(specificity = factor(specificity,
+  mutate(GTEX_specificity = factor(GTEX_specificity, 
                                      levels = c("Testis specific",  
@@
                                                 "Testis preferential"))) %>% 
-  mutate(chr_met = factor(chr_met, 
-                          levels = c("X_met", "not_X_met", "X_not_met", 
-                                     "not_X_not_met")))
-
-all_genes_types <- all_genes %>% 
-  dplyr::rename(gene = external_gene_name) %>% 
-  mutate(Regulation = case_when(regulated_by_methylation ~ "Methylation",
-                                !regulated_by_methylation ~ 
-                                  "Not methylation")) %>%
-  mutate(Chr = case_when(X_linked ~ "X-linked",
-                         !X_linked ~ "not X")) %>% 
-  mutate(chr_met = case_when(regulated_by_methylation & X_linked ~ "X_met",
-                             !regulated_by_methylation & X_linked ~ "X_not_met",
-                             regulated_by_methylation & !X_linked ~ "not_X_met",
-                             !regulated_by_methylation & !X_linked ~ 
-                               "not_X_not_met")) %>% 
-  mutate(specificity = case_when(GTEX_category == "testis_specific" ~ 
-                                   "Testis specific",
-                                 GTEX_category == "testis_preferential" ~ 
-                                   "Testis preferential",
-                                 GTEX_category == "lowly_expressed" ~ 
-                                   "Undetectable in GTEX")) %>% 
-  mutate(specificity = factor(specificity,
+  mutate(testis_specificity = case_when(testis_specificity == 
+                                          "testis_specific" ~ "Testis specific",
+                                        testis_specificity == 
+                                          "testis_preferential" ~ 
+                                          "Testis preferential")) %>%
+  mutate(testis_specificity = factor(testis_specificity, 
                                      levels = c("Testis specific", 
-                                         "Undetectable in GTEX", 
                                                 "Testis preferential"))) %>% 
@@
                           levels = c("X_met", "not_X_met", "X_not_met", 
-                                     "not_X_not_met")))
-
+                                     "not_X_not_met"))) %>% 
+  mutate(met_3_groups = case_when(regulated_by_methylation & X_linked ~ 
+                                    "Methylation X-linked",
+                                   regulated_by_methylation & !X_linked ~ 
+                                    "Methylation not X-linked",
+                             !regulated_by_methylation ~ "Not Methylation")) %>% 
+  mutate(met_3_groups = factor(met_3_groups, 
+                          levels = c("Methylation X-linked", 
+                                     "Methylation not X-linked", 
+                                     "Not Methylation")))
+
+CT_genes <- all_genes %>%
+  filter(CT_gene_type == "CT_gene")
+CT_genes_types <- all_genes_clean %>% 
+  filter(CT_gene_type == "CT_gene")
+
+CTP_genes <- all_genes %>%
+  filter(CT_gene_type == "CTP_gene")
+CTP_genes_types <- all_genes_clean %>% 
+  filter(CT_gene_type == "CTP_gene")
 ```
@@
 Chr_colors = c("X-linked" = "deeppink3", "not X" = "lightgreen")
+Chr_colors_with_Y = c("X-linked" = "deeppink3", "Autosome" = "lightgreen", 
+                      "Y-linked" = "black")
+
+GTEX_specificity_colors = c("Testis specific" = "mediumpurple4",
+                            "Testis preferential" = "goldenrod3",
+                            "Undetectable in GTEX" = "gray")
 
@@
 
+Fetal_cellType_colors <- c( 
+"F_GC" = "pink3",
+  "F_oogonia" = "palevioletred3",
+  "F_oocyte" = "mediumorchid4",
+  "M_GC" = "steelblue3",
+  "M_pre_spermatogonia" = "royalblue3",
+  "F_soma" = "gray",
+  "M_soma" = "gray")
+
+Fetal_time_colors <- c("6W" = "aliceblue",
+                      "7W" = "cyan",
+                      "8W" = "cyan4",
+                      "9W" = "skyblue1",
+                      "10W" = "dodgerblue",
+                      "17W" = "blue",
+                      "21W" = "lightblue4",
+                      "24W" = "black")
+
 oocytes_colors <- c("Growing oocytes" = "mediumpurple1",
@@
 
-table(all_genes$CT_gene_type) #Before and after TSS verif
+table(all_genes$IGV_backbone)
+table(all_genes$CT_gene_type)
 ```
@@
                                h_height = 10,
-                               split_rows_by = "regulation") {
+                               split_rows_by = "regulation",
+                               clust_method = "ward.D",
+                               clustRow = FALSE) {
 
@@
 
-  gene_type <- all_genes_types %>% 
-    filter(gene %in% genes)
+  gene_type <- all_genes_clean %>% 
+    filter(gene %in% genes) %>% 
+    mutate(gene = factor(gene, levels = genes)) %>% 
+    arrange(gene)
 
     if (split_rows_by == "regulation") {
-    gene_type <- gene_type %>% arrange(Regulation)
+        if (clustRow) gene_type <- gene_type %>% 
+            arrange(Regulation)
       rowSplit <- gene_type$Regulation
@@
     if (split_rows_by == "testis_specificity") {
-    gene_type <- gene_type %>% arrange(Testis_specificity)
-    rowSplit <- gene_type$Testis_specificity
+        if (clustRow) gene_type <- gene_type %>% 
+            arrange(testis_specificity)
+      rowSplit <- gene_type$testis_specificity
     }
     if (split_rows_by == "chr_met") {
-    gene_type <- gene_type %>% arrange(chr_met)
+      if (clustRow) gene_type <- gene_type %>% 
+          arrange(chr_met)
       rowSplit <- gene_type$chr_met
     }
-  if (split_rows_by == "specificity") {
-    gene_type <- gene_type %>% arrange(specificity)
-    rowSplit <- gene_type$specificity
-  }
-
-  if (split_rows_by == "GTEX_category") {
-    GTEX_category <- gene_type %>% arrange(GTEX_category)
-    rowSplit <- gene_type$GTEX_category
+    if (split_rows_by == "GTEX_specificity") {
+      if (clustRow) GTEX_category <- gene_type %>% 
+          arrange(GTEX_specificity)
+      rowSplit <- gene_type$GTEX_specificity
     }
@@
 
-  specificity <- as.matrix(gene_type$specificity)
-  rownames(specificity) <- gene_type$gene
-  row_ha_ts <- rowAnnotation(specificity = specificity,
+  GTEX_specificity <- as.matrix(gene_type$GTEX_specificity)
+  rownames(GTEX_specificity) <- gene_type$gene
+  row_ha_GTEX <- rowAnnotation(specificity = GTEX_specificity,
                              annotation_legend_param = legends_param,
@@
                              col = list(specificity = 
+                                          GTEX_specificity_colors),
+                             annotation_name_gp = gpar(fontsize = 0))
+
+  testis_specificity <- as.matrix(gene_type$testis_specificity)
+  rownames(testis_specificity) <- gene_type$gene
+  row_ha_TS <- rowAnnotation(testis_specificity = testis_specificity,
+                             annotation_legend_param = legends_param,
+                             simple_anno_size = unit(0.2, "cm"),
+                             col = list(testis_specificity = 
                                           Testis_specificity_colors),
@@
 
-  h <- Heatmap(mat[gene_type$gene, ],
+  h <- Heatmap(mat[as.vector(gene_type$gene), ],
                name = "log(TPM +1)",
@@
                cluster_row_slices = FALSE,
-               row_title_gp = gpar(fontsize = 4, face = "bold"),
+               row_title_gp = gpar(fontsize = 6, face = "bold"),
                border = TRUE,
@@
                col = heatmap_colors,
-               cluster_rows = TRUE,
+               cluster_rows = clustRow,
                cluster_columns = TRUE,
@@
                height = unit(h_height, "cm"),
-               clustering_method_rows = "ward.D",
+               clustering_method_rows = clust_method,
                use_raster = TRUE,
@@
                                h_height = 10,
-                               clustRow = TRUE,
+                               show_annotation_legend = TRUE,
+                               show_top_annotations = TRUE,
+                               show_row_names = FALSE,
+                               clustRow = FALSE,
                                clustCol = TRUE) {
@@
   rownames(mat) <- rowData(TCGA)$external_gene_name
-  mat <- mat[rownames(mat) %in% all_genes$external_gene_name,]
   mat <- log1p(mat)
 
-  gene_type <- all_genes_types %>% 
-    filter(gene %in% genes)
+  gene_type <- all_genes_clean %>% 
+    filter(gene %in% genes) %>% 
+    mutate(gene = factor(gene, levels = genes)) %>% 
+    arrange(gene)
 
@@
     if (split_rows_by == "regulation") {
-    gene_type <- gene_type %>% arrange(Regulation)
+      if (clustRow) gene_type <- gene_type %>% 
+          arrange(Regulation)
       rowSplit <- gene_type$Regulation
@@
     if (split_rows_by == "testis_specificity") {
-    gene_type <- gene_type %>% arrange(Testis_specificity)
+      if (clustRow) gene_type <- gene_type %>% 
+          arrange(testis_specificity)
       rowSplit <- gene_type$testis_specificity
     }
-  if (split_rows_by == "specificity") {
-    gene_type <- gene_type %>% arrange(specificity)
-    rowSplit <- gene_type$specificity
+    if (split_rows_by == "GTEX_specificity") {
+      if (clustRow) gene_type <- gene_type %>% 
+          arrange(GTEX_specificity)
+      rowSplit <- gene_type$GTEX_specificity
     }
@@
     if (split_rows_by == "chr_met") {
-    gene_type <- gene_type %>% arrange(chr_met)
+      if (clustRow) gene_type <- gene_type %>% 
+          arrange(chr_met)
       rowSplit <- gene_type$chr_met
-  }
-
-  if (split_rows_by == "GTEX_category") {
-    GTEX_category <- gene_type %>% arrange(GTEX_category)
-    rowSplit <- gene_type$GTEX_category
     }
@@
 
-  specificity <- as.matrix(gene_type$specificity)
-  rownames(specificity) <- gene_type$gene
-  row_ha_ts <- rowAnnotation(specificity = specificity,
+  GTEX_specificity <- as.matrix(gene_type$GTEX_specificity)
+  rownames(GTEX_specificity) <- gene_type$gene
+  row_ha_GTEX <- rowAnnotation(GTEX_specificity = GTEX_specificity,
                              annotation_legend_param = legends_param,
                              simple_anno_size = unit(0.2, "cm"),
-                             col = list(specificity = 
+                             col = list(GTEX_specificity = 
+                                          GTEX_specificity_colors),
+                             annotation_name_gp = gpar(fontsize = 0))
+
+  testis_specificity <- as.matrix(gene_type$testis_specificity)
+  rownames(testis_specificity) <- gene_type$gene
+  row_ha_TS <- rowAnnotation(specificity = testis_specificity,
+                             annotation_legend_param = legends_param,
+                             simple_anno_size = unit(0.2, "cm"),
+                             col = list(testis_specificity = 
                                           Testis_specificity_colors),
                              annotation_name_gp = gpar(fontsize = 0))
+
+  rAnnot <- NULL
+  if(show_annotation_legend){
+    if (split_rows_by == "regulation") {
+  rAnnot <- c(row_ha_regulation, gap = unit(1, "mm"))
+    }
+    if (split_rows_by == "testis_specificity") {
+  rAnnot <- c(row_ha_TS, gap = unit(1, "mm"))
+    }
+    if (split_rows_by == "GTEX_specificity") {
+  rAnnot <- c(row_ha_GTEX, gap = unit(1, "mm"))
+    }
+
+    if (split_rows_by == "chr_met") {
+  rAnnot <- c(row_ha_TS, gap = unit(1, "mm"))
+    }
+  }
 
@@
 
-  Heatmap(mat[gene_type$gene, ],
+  tAnnot <- NULL
+  if (show_top_annotations) tAnnot <- column_ha_tumor
+
+  Heatmap(mat[as.vector(gene_type$gene), ],
           name = "TCGA (logTPM)",
@@
           cluster_columns = clustCol,
-          show_row_names = FALSE,
+          show_row_names = show_row_names,
           show_column_names = FALSE,
@@
           column_names_centered = TRUE,
-          row_names_gp = gpar(fontsize = 0),
+          row_names_gp = gpar(fontsize = fontsize),
           row_names_side = "left",
@@
           row_title_gp = gpar(fontsize = 0),
-          top_annotation = column_ha_tumor,
-          right_annotation = c(row_ha_ts,
-                               #row_ha_regulation,
-                               #row_ha_chr,
-                               gap = unit(1, "mm")),
+          top_annotation = tAnnot,
+          right_annotation = rAnnot,
           row_split = rowSplit,
@@
 }
-
 ```
@@
                                split_rows_by = "regulation", 
-                               clustRow = TRUE,
+                               show_annotation_legend = TRUE,
+                               clustRow = FALSE,
                                clustCol = TRUE){
@@
   rownames(mat) <- rowData(CCLE)$external_gene_name
-  mat <- mat[rownames(mat) %in% all_genes$external_gene_name, ]
   mat <- log1p(mat)
 
-  gene_type <- all_genes_types %>% 
-    filter(gene %in% genes)
-
+  gene_type <- all_genes_clean %>% 
+    filter(gene %in% genes) %>% 
+    mutate(gene = factor(gene, levels = genes)) %>% 
+    arrange(gene)
+
+ rowSplit <- NULL
+  if(!is.null(split_rows_by)){
     if (split_rows_by == "regulation") {
-    gene_type <- gene_type %>% arrange(Regulation)
+      if (clustRow) gene_type <- gene_type %>% 
+          arrange(Regulation)
       rowSplit <- gene_type$Regulation
@@
     if (split_rows_by == "testis_specificity") {
-    gene_type <- gene_type %>% arrange(testis_specificity)
+      if (clustRow) gene_type <- gene_type %>% 
+          arrange(testis_specificity)
       rowSplit <- gene_type$testis_specificity
     }
-  if (split_rows_by == "specificity") {
-    gene_type <- gene_type %>% arrange(specificity)
-    rowSplit <- gene_type$specificity
-  }
+    if (split_rows_by == "GTEX_specificity") {
+      if (clustRow) gene_type <- gene_type %>% 
+          arrange(GTEX_specificity)
+      rowSplit <- gene_type$GTEX_specificity
+    }
+
     if (split_rows_by == "chr_met") {
-    gene_type <- gene_type %>% arrange(chr_met)
+      if (clustRow) gene_type <- gene_type %>% 
+          arrange(chr_met)
       rowSplit <- gene_type$chr_met
     }
-
-   if (split_rows_by == "GTEX_category") {
-    GTEX_category <- gene_type %>% arrange(GTEX_category)
-    rowSplit <- gene_type$GTEX_category
   }
@@
 
-  specificity <- as.matrix(gene_type$specificity)
-  rownames(specificity) <- gene_type$gene
-  row_ha_ts <- rowAnnotation(specificity = specificity,
+  GTEX_specificity <- as.matrix(gene_type$GTEX_specificity)
+  rownames(GTEX_specificity) <- gene_type$gene
+  row_ha_GTEX <- rowAnnotation(GTEX_specificity = GTEX_specificity,
                              annotation_legend_param = legends_param,
                              simple_anno_size = unit(0.2, "cm"),
-                                           col = list(specificity = 
+                             col = list(GTEX_specificity = 
+                                          GTEX_specificity_colors),
+                             annotation_name_gp = gpar(fontsize = 0))
+
+  testis_specificity <- as.matrix(gene_type$testis_specificity)
+  rownames(testis_specificity) <- gene_type$gene
+  row_ha_TS <- rowAnnotation(specificity = testis_specificity,
+                             annotation_legend_param = legends_param,
+                             simple_anno_size = unit(0.2, "cm"),
+                             col = list(testis_specificity = 
                                           Testis_specificity_colors),
@@
 
-  # heatmap_colors <- colorRamp2(
-  #   seq(0, quantile(rowMax(mat), 0.8), length = 11),                 
-  #   legend_colors)
+  rAnnot <- NULL
+  if(show_annotation_legend){
+    if (split_rows_by == "regulation") {
+  rAnnot <- c(row_ha_regulation, gap = unit(1, "mm"))
+    }
+    if (split_rows_by == "testis_specificity") {
+  rAnnot <- c(row_ha_TS, gap = unit(1, "mm"))
+    }
+    if (split_rows_by == "GTEX_specificity") {
+  rAnnot <- c(row_ha_GTEX, gap = unit(1, "mm"))
+    }
+
+    if (split_rows_by == "chr_met") {
+  rAnnot <- c(row_ha_TS, gap = unit(1, "mm"))
+    }
+  }
 
@@
 
-  Heatmap(mat[gene_type$gene, ],
+  Heatmap(mat[as.vector(gene_type$gene), ],
           name = "CCLE (logTPM)",
@@
           top_annotation = column_ha_tumor,
-          # right_annotation = c(row_ha_ts,
-          #                      row_ha_regulation,
-          #                      #row_ha_chr,
-          #                      gap = unit(1, "mm")),
+          right_annotation = rAnnot,
           row_split = rowSplit,
@@
 ```{r selection_heatmap}
-my_genes <- CT_genes %>% 
+my_genes <- all_genes %>% 
+  filter(CT_gene_type != "other") %>% 
   pull(external_gene_name)
 
+# Run the heatmap with custRow parameters to define the correct order
+# This order will then be used for GTEX and CCLE heatmaps
 ht_TCGA <- print_TCGA_heatmap(my_genes, 
@@
                               clust_method = "ward.D",
-                              split_rows_by = "specificity")
+                              clustRow = TRUE,
+                              split_rows_by = "testis_specificity")
+indices <- row_order(ht_TCGA)
+gene_order <- all_genes_clean %>%
+  filter(gene %in% my_genes) %>%
+  arrange(testis_specificity) %>%
+  pull(gene)
+
+TS_order <- gene_order[indices$`Testis specific`]
+TP_order <- gene_order[indices$`Testis preferential`]
+my_genes <- c(TS_order, TP_order)
+
+ht_TCGA <- print_TCGA_heatmap(my_genes, 
+                              fontsize = 4, 
+                              downsample_to_n = 50, 
+                              h_width = 3,
+                              h_height = 8,
+                              tum_type = c("TCGA-SKCM", "TCGA-LUAD", 
+                                           #"TCGA-ESCA", "TCGA-LUSC", 
+                                           #"TCGA-COAD",
+                                           #"TCGA-BRCA",
+                                           "TCGA-HNSC"),
+                              clust_method = "ward.D",
+                              show_annotation_legend = FALSE,
+                              clustRow = FALSE,
+                              split_rows_by = "testis_specificity")
 
@@
                               #"Leukemia", "Lymphoma"),
+                              clustRow = FALSE,
+                              show_annotation_legend = FALSE,
                               h_width = 3,
                               h_height = 8,
-                              split_rows_by = "specificity")
+                              split_rows_by = "testis_specificity")
 
@@
                               h_height = 8,
-                             split_rows_by = "specificity")
+                              clustRow = FALSE,
+                              split_rows_by = "testis_specificity")
+
 h_list <- ht_GTEX + ht_CCLE + ht_TCGA
-draw(h_list, main_heatmap = "TCGA (logTPM)", ht_gap = unit(c(0.3, 0.01), "cm"),
+draw(h_list, main_heatmap = "log(TPM +1)", ht_gap = unit(c(0.3, 0.01), "cm"),
      merge_legend = TRUE)
@@
 pdf(file = paste0("figs/GTEX_Tumors.pdf"))
-draw(h_list, main_heatmap = "TCGA (logTPM)", 
+draw(h_list, main_heatmap = "log(TPM +1)", 
      ht_gap = unit(c(0.3, 0.01), "cm"),
@@
 
-Counter examples
-
-```{r counter_examples}
-genes_not_testis <- all_genes %>% 
-  filter(GTEX_category != "lowly_expressed") %>% 
-  filter(testis_specificity == "not_testis_specific") %>% 
-  pull(external_gene_name) %>% 
-  sample(2)
-
-genes_not_TCGA_activated <- all_genes %>% 
-  filter(GTEX_category != "lowly_expressed") %>% 
-  filter(testis_specificity != "not_testis_specific") %>% 
-  filter(!TCGA_category %in% c("activated", "multimapping_issue")) %>% 
-  pull(external_gene_name) %>% 
-  sample(1)
-
-genes_not_ccle_activated <- all_genes %>% 
-  filter(GTEX_category != "lowly_expressed") %>% 
-  filter(testis_specificity != "not_testis_specific") %>% 
-  filter(CCLE_category %in% c("activated")) %>% 
-  pull(external_gene_name) %>% 
-  sample(1)
-
-my_genes_counter <- c(genes_not_testis, genes_not_TCGA_activated, 
-                      genes_not_ccle_activated)
-
-ht_TCGA_counter <- print_TCGA_heatmap(my_genes_counter, 
-                              fontsize = 4, 
-                              downsample_to_n = 50, 
-                              h_width = 3,
-                              h_height = 8,
-                              tum_type = c("TCGA-SKCM", "TCGA-LUAD",
-                                           "TCGA-HNSC"),
-                              clust_method = "ward.D",
-                              split_rows_by = "specificity")
-
-ht_CCLE_counter <- print_CCLE_heatmap(my_genes_counter, 
-                              downsample_to_n = 50, 
-                              tum_type = c("Skin", "Lung",
-                                           "Head_and_Neck"),
-                              h_width = 3,
-                              h_height = 8,
-                              split_rows_by = "specificity")
-
-ht_GTEX_counter <- print_GTEX_heatmap(my_genes_counter, 
-                              row_fontsize = 8,
-                              column_fontsize = 5,
-                              h_width = 5,
-                              h_height = 8,
-                             split_rows_by = "specificity")
-h_list_counter<- ht_GTEX_counter + ht_CCLE_counter + ht_TCGA_counter
-draw(h_list_counter, main_heatmap = "TCGA (logTPM)", ht_gap = unit(c(0.3, 0.01),
-                                                                   "cm"),
-     merge_legend = TRUE)
-
-my_genes_leaky <-  all_genes %>% 
-  filter(GTEX_category %in% c("testis_specific", "testis_preferential")) %>% 
-  filter(TCGA_category == "leaky" | CCLE_category == "leaky") %>% 
-   pull(external_gene_name)
-
-ht_TCGA_leaky <- print_TCGA_heatmap(my_genes_leaky, 
-                              fontsize = 4, 
-                              downsample_to_n = 50, 
-                              h_width = 3,
-                              h_height = 8,
-                              tum_type = c("TCGA-SKCM", "TCGA-LUAD",
-                                           "TCGA-HNSC"),
-                              clust_method = "ward.D",
-                              split_rows_by = "specificity")
-
-ht_CCLE_leaky <- print_CCLE_heatmap(my_genes_leaky, 
-                              downsample_to_n = 50, 
-                              tum_type = c("Skin", "Lung",
-                                           "Head_and_Neck"),
-                              h_width = 3,
-                              h_height = 8,
-                              split_rows_by = "specificity")
-
-ht_GTEX_leaky <- print_GTEX_heatmap(my_genes_leaky, 
-                              row_fontsize = 8,
-                              column_fontsize = 5,
-                              h_width = 5,
-                              h_height = 8,
-                             split_rows_by = "specificity")
-h_list_leaky<- ht_GTEX_leaky + ht_CCLE_leaky + ht_TCGA_leaky
-draw(h_list_leaky, main_heatmap = "TCGA (logTPM)", ht_gap = unit(c(0.3, 0.01), "cm"),
-     merge_legend = TRUE)
-```
-
 ## Multimapping Fig
@@
 ```{r multimapping}
-my_genes <- CT_genes %>% 
+my_genes <- all_genes_clean %>% 
   filter(GTEX_category == 'lowly_expressed')  %>% 
-  filter(testis_specificity == "testis_specific")
+  filter(CT_gene_type != 'other')  %>% 
+    arrange(testis_specificity)
+rowSplit <- my_genes$testis_specificity
+
+testis_specificity <- as.matrix(my_genes$testis_specificity)
+rownames(testis_specificity) <- my_genes$gene
+row_ha_TS <- rowAnnotation(testis_specificity = testis_specificity,
+                           annotation_legend_param = legends_param,
+                           simple_anno_size = unit(0.2, "cm"),
+                           col = list(testis_specificity = 
+                                        Testis_specificity_colors),
+                           annotation_name_gp = gpar(fontsize = 0))
 
@@
 rownames(without_MP) <- rownames(with_MP) <- 
-  rowData(normal_tissues_multimapping_data)[my_genes$ensembl_gene_id,]$external_gene_name
-
+  rowData(normal_tissues_multimapping_data)[my_genes$ensembl_gene_id, 
+                                            ]$external_gene_name
 # Fix same range of colors for both heatmaps
@@
 
-# Or same range of colors than GTEX heatmap
-# heatmap_colors <- colorRamp2(
-#   seq(0, 6, length = 11), legend_colors)
-
-colnames(without_MP) <- gsub("_", x = str_to_sentence(colnames(without_MP)), ' ')
+colnames(without_MP) <- gsub("_", x = str_to_sentence(colnames(without_MP)),
+                             ' ')
 colnames(with_MP) <- gsub("_", x = str_to_sentence(colnames(with_MP)), ' ')
@@
   c("Testis", colnames(without_MP)[- which(colnames(without_MP) == "Testis")])
-h_without_MP <- Heatmap(without_MP[my_genes$external_gene_name, ordered_tissues],
+h_without_MP <- Heatmap(without_MP[my_genes$gene, ordered_tissues],
                         name = "no MP",
-                        column_title = "Multimapped reads are excluded",
+                        column_title = "Multimapped reads excluded",
                         column_title_gp = gpar(fontsize = 10, face = "bold"),
+                        row_title_gp = gpar(fontsize = 0, face = "bold", 
+                                            angle  = 90),
+                        row_split = rowSplit,
                         border = TRUE,
@@
                         col = heatmap_colors,
-                        cluster_rows = TRUE,
+                        cluster_rows = FALSE,
                         cluster_columns = FALSE,
@@
 
-h_with_MP <- Heatmap(with_MP[my_genes$external_gene_name, ordered_tissues],
+h_with_MP <- Heatmap(with_MP[my_genes$gene, ordered_tissues],
                      name = "log(TPM)",
-                     column_title = "Multimapped reads are counted",
+                     column_title = "Multimapped reads counted",
                      column_title_gp = gpar(fontsize = 10, face = "bold"),
+                     row_title_gp = gpar(fontsize = 0, face = "bold"),
+                     row_split = rowSplit,
                      border = TRUE,
@@
                      clustering_method_rows = "ward.D",
+                     right_annotation = c(row_ha_TS, gap = unit(1, "mm")),
                      use_raster = TRUE,
@@
 h_list <- h_without_MP + h_with_MP
-draw(h_list, main_heatmap = "log(TPM)", ht_gap = unit(0.3, "cm"),
+draw(h_list, main_heatmap = "no MP", ht_gap = unit(0.3, "cm"),
                  height = unit(5, "cm"))
@@
 pdf(file = paste0("figs/Multimapping.pdf"))
-draw(h_list, main_heatmap = "log(TPM)", ht_gap = unit(0.3, "cm"),
+draw(h_list, main_heatmap = "no MP", ht_gap = unit(0.3, "cm"),
                  height = unit(5, "cm"))
@@
 
-# CT_genes specific
-
-From now on, we'll focus on CT genes that are specific
-
-```{r}
-CT_genes <- CT_genes %>%
-  filter(CT_gene_type == "CT_gene") 
-CT_genes_types <- CT_genes_types %>% 
-  filter(CT_gene_type == "CT_gene")
-```
-
 # Methylation analysis
 
-## Regulation by methylation
+## Regulation by methylation : WGBS + DAC heatmap
 
@@
   pull(external_gene_name)
+
 met_value_number_per_gene <- tibble(
@@
 n_met_values <- ncol(mean_methylation_in_tissues)
+
 CT_met_with_WGBS <- met_value_number_per_gene %>%
@@
   filter(regulated_by_methylation) %>% 
-  filter(NA_met_values_number == n_met_values )
+  filter(NA_met_values_number == n_met_values)
 
@@
   filter(!regulated_by_methylation) %>% 
-  filter(NA_met_values_number == n_met_values )
+  filter(NA_met_values_number == n_met_values)
 
@@
                              gene %in% CT_not_met_with_WGBS$gene ~ "group3",
-                             gene %in% CT_not_met_without_WGBS$gene ~ "group4")) %>% 
+                             gene %in% CT_not_met_without_WGBS$gene ~ 
+                               "group4")) %>% 
     left_join(CT_genes %>% dplyr::rename(gene = external_gene_name) %>% 
@@
                border_gp = gpar(lwd = 0.5),
-               row_title_gp = gpar(col = c("indianred2", "cyan4"), fontsize = 8),
+               row_title_gp = gpar(col = Regulation_colors, fontsize = 6),
                row_split = factor(gene_type$Regulation,
@@
                col = colorRamp2(seq_len(100),
-                                colorRampPalette(c("moccasin", "dodgerblue4"))(100)),
+                                colorRampPalette(c("moccasin", 
+                                                   "dodgerblue4"))(100)),
                na_col = "gray80",
@@
 ```{r DAC_heatmap_function}
-DAC_fig <- function(genes = gene_type$gene, multimapping = TRUE, 
+DAC_fig <- function(genes = gene_type$gene, 
+                    multimapping = TRUE, 
                     row_fontsize = 2,
@@
 
-  specificity <- as.matrix(gene_type$specificity)
-  rownames(specificity) <- gene_type$gene
-  row_ha_ts <- rowAnnotation(specificity = specificity,
+  testis_specificity <- as.matrix(gene_type$testis_specificity)
+  rownames(testis_specificity) <- gene_type$gene
+  row_ha_TS <- rowAnnotation(testis_specificity = testis_specificity,
                              annotation_legend_param = legends_param,
                              simple_anno_size = unit(0.2, "cm"),
-                             col = list(specificity = 
+                             col = list(testis_specificity = 
                                           Testis_specificity_colors),
@@
                border_gp = gpar(lwd = 0.5),
-               column_title = "Expression in cells treated with 5-Aza",
+               column_title = "DAC induction in cell lines",
                column_title_gp = gpar(fontsize = 6),
@@
                #                      gap = unit(1, "mm")),
-               heatmap_width = unit(6, "cm"),
+               heatmap_width = unit(3.5, "cm"),
                heatmap_height = unit(10, "cm"),
@@
 ```{r draw_meth_reg}
-fs <- 1.2
-h_met <- mean_methylation_fig(genes = CT_genes$external_gene_name,
+fs <- 1.8
+my_genes <- CT_genes %>% 
+  filter(!is.na(DAC_induced))
+h_met <- mean_methylation_fig(genes = my_genes$external_gene_name,
                               values_only = FALSE,
@@
 
-h_DAC <- DAC_fig(CT_genes$external_gene_name, 
+h_DAC <- DAC_fig(my_genes$external_gene_name, 
                  multimapping = TRUE, 
@@
 ```{r prop_meth}
-x <- enframe(table(CT_genes_types$Regulation), name = "Regulation") %>% 
+x <- enframe(table(CT_genes_types %>% 
+  filter(gene %in% CT_genes$external_gene_name) %>% 
+  pull(Regulation)), name = "Regulation") %>% 
   left_join(enframe(prop.table(table(CT_genes_types$Regulation)), 
                     name = "Regulation", value = "Percent"))
+
 met_prop_fig <- ggplot(x, aes(x = '', y = value, fill = Regulation)) +
@@
   coord_polar("y", start = 0) +
-  geom_text(aes(y = value/5 + c(0, cumsum(value)[-length(value)]), 
+  geom_text(aes(y = value/5.5 + c(0, cumsum(value)[-length(value)]), 
                 label = rev(paste0(Regulation, "\n", value, "/", 
-                                   sum(table(CT_genes_types$Regulation)), "\n",
-                               round(Percent * 100), "%"))))+
-            #, size = 6
-  scale_fill_manual(values = c( "indianred2", "cyan4")) +
+                                   nrow(CT_genes_types %>%
+                                          filter(!is.na(Regulation))), "\n",
+                               round(Percent * 100), "%")))) +
+  scale_fill_manual(values = Regulation_colors) +
   theme(axis.title.x = element_blank(),
@@
 ```{r, eval = F}
-pdf(file = paste0("figs/Met_proportions.pdf"))
+pdf(file = paste0("figs/Met_proportions.pdf"), width = 4, height = 4)
 met_prop_fig
@@
 
-fig_X_enrichment <- ggplot(CT_genes_types[
-  !is.na(CT_genes_types$regulated_by_methylation), ], 
+fig_X_enrichment <- ggplot(CT_genes_types %>% 
+                             filter(!is.na(regulated_by_methylation)), 
                            aes(x = chr, fill = Regulation)) +
@@
   scale_fill_manual(values = Regulation_colors) +
-  theme(legend.position = "bottom",
-        aspect.ratio = 1)
+  theme_bw() +
+  theme(legend.position = "inside",
+        legend.position.inside = c(0.2, 0.85),
+        #aspect.ratio = 1.5,
+        panel.grid = element_blank())
+
 fig_X_enrichment
@@
 ```{r, eval = FALSE, echo=FALSE}
-pdf(file = paste0("figs/X_enrichment.pdf"), width = 7, height = 4)
-cowplot::plot_grid(met_prop_fig, fig_X_enrichment, labels = c('A', 'B'), 
-          rel_widths = c(1, 2),
-          label_size = 12)
+pdf(file = paste0("figs/X_enrichment.pdf"), width = 4, height = 6)
+fig_X_enrichment
 dev.off()
@@
 
-This has to be checked with stats
-
-```{r meth_stats}
-CT_genes_types <- CT_genes_types %>% 
-  mutate(group = case_when(chr == "X" ~ "X-linked",
+## Stats
+
+### Testis-specific and methylation
+
+CT genes are enriched among testis-specific genes regulated by methylation.
+Proportion of Testis-specific genes regulated by methylation
+
+```{r testis_spec_meth}
+all_genes_chr <- all_genes %>% 
+  mutate(chr = case_when(chr == "X" ~ "X-linked",
                            chr == "Y" ~ "Y-linked",
@@
 
-x <- table(CT_genes_types$group, CT_genes_types$Regulation)
+x <- table(met = all_genes_chr$regulated_by_methylation, 
+           testis_specificity = all_genes_chr$testis_specificity, 
+           useNA = "ifany")
 x
-fisher.test(x[c("Autosome", "X-linked"),])
-fisher.test(x[c("Autosome", "Y-linked"),])
-
-chisq.test(x[c("Autosome", "X-linked"),])
-chisq.test(x[c("Autosome", "Y-linked"),])
-```
-
-Genes controlled by methylation are significantly enriched on the X
+
+chisq.test(x[c("FALSE", "TRUE"), c("not_testis_specific", "testis_specific")])
+chisq.test(x[c("FALSE", "TRUE"), c("not_testis_specific",
+                                   "testis_specific")])$p.value
+```
+- `r x["TRUE", "testis_specific"]`/`r x["TRUE", "testis_specific"] + x["FALSE", "testis_specific"]`
+(`r x["TRUE", "testis_specific"]/(x["TRUE", "testis_specific"] + x["FALSE", "testis_specific"])*100`
+%) of Testis-specific are regulated by methylation
+
+=> Significant enrichment of genes regulated by methylation among testis-specific genes 
+(pval = `r chisq.test(x[c("FALSE", "TRUE"), c("not_testis_specific", "testis_specific")])$p.value`)
+
+Proportion of CT genes among Testis-specific genes regulated or not by methylation
+
+```{r CT_meth}
+TS_genes <- all_genes_chr %>% 
+  filter(testis_specificity == "testis_specific")
+x <- table(TS_genes = TS_genes$CT_gene_type, regulated_by_met = TS_genes$regulated_by_methylation)
+x
+chisq.test(x)
+chisq.test(x)$p.value
+```
+
+- `r x["CT_gene", "TRUE"]` / `r x["CT_gene", "TRUE"] + x["other", "TRUE"]`
+(`r x["CT_gene", "TRUE"]/(x["CT_gene", "TRUE"] + x["other", "TRUE"]) * 100`%)
+of TS genes regulated by methylation are CT genes
+
+- `r x["CT_gene", "FALSE"]` / `r x["CT_gene", "FALSE"] + x["other", "FALSE"]`
+(`r x["CT_gene", "FALSE"]/(x["CT_gene", "FALSE"] + x["other", "FALSE"]) * 100`%)
+of TS genes not regulated by methylation are CT genes
+
+=> Strong enrichment in CT genes among Testis-specific genes regulated by 
+methylation (`chisq.test(x)$p.value`)
+
+```{r}
+fig_TS_CT_genes <- TS_genes %>% 
+  filter(!is.na(regulated_by_methylation)) %>% 
+  mutate(Regulation = case_when(regulated_by_methylation ~ "Met",
+                                !regulated_by_methylation ~ "Not Met")) %>% 
+  mutate(Regulation = factor(Regulation, levels = c("Not Met", "Met"))) %>% 
+  mutate(Group = case_when(CT_gene_type == "CT_gene" ~ "CT gene",
+                             CT_gene_type != "CT_gene" ~ "Not CT gene")) %>% 
+  group_by(Regulation, Group) %>% 
+  dplyr::count() %>% 
+  mutate(testis_specificity = "Testis-specific genes") %>% 
+  ggplot(aes(x = Regulation, y = n, fill = Group)) +
+  geom_col() +
+  scale_fill_manual(values = c("seagreen2", "gray50")) +
+  xlab('') +
+  ylab("Number of genes") +
+  facet_wrap(~ testis_specificity) +
+  theme_bw() +
+  theme(legend.position = "right",
+        plot.title = element_text(hjust = 0.5),
+        panel.grid = element_blank(),
+        strip.text = element_text(size = 10),
+        axis.text.x = element_text(size = 10),
+        axis.title.y = element_text(size = 10))        
+fig_TS_CT_genes
+
+```
+
+### Chromosomal location
+
+CT chromosomal location
+
+```{r}
+x <- table(gene = all_genes_chr$CT_gene_type, chr = all_genes_chr$chr)
+x
+```
+
+- `r x["CT_gene", "X-linked"]` / `r sum(x["CT_gene", ])`
+(`r x["CT_gene", "X-linked"]/sum(x["CT_gene", ]) * 100`%) CT genes are X-linked
+
+```{r}
+CT_genes_chr <- all_genes_chr %>% 
+  filter(CT_gene_type == "CT_gene")
+
+x <- table(chr = CT_genes_chr$chr, 
+           regulated_by_met = CT_genes_chr$regulated_by_methylation)
+x
+```
+
+- `r x["X-linked", "TRUE"]` / `r sum(x["X-linked", ])`
+(`r x["X-linked", "TRUE"]/sum(x["X-linked", ]) * 100`%)
+of X-linked genes are regulated by methylation
+
+```{r}
+chisq.test(x[c("Autosome", "X-linked"), c("FALSE", "TRUE")])$p.value
+```
+
+All genes Chromosomal location
+
+```{r}
+x <- table(chr = all_genes_chr$chr, 
+           testis_specificity = all_genes_chr$testis_specificity)
+x
+
+chisq.test(x[c("Autosome", "X-linked"), c("not_testis_specific", 
+                                          "testis_specific")])
+
+chisq.test(x[c("Autosome", "X-linked"), c("not_testis_specific", 
+                                          "testis_specific")])$p.value
+
+chisq.test(x[c("Autosome", "Y-linked"), c("not_testis_specific",
+                                          "testis_specific")])$p.value
+```
+
+- Testis-specific genes are enriched on the X (pval = 
+`r chisq.test(x[c("Autosome", "X-linked"), c("not_testis_specific", "testis_specific")])$p.value`)
+
+- Testis-specific genes are also enriched on the Y (pval = 
+`r chisq.test(x[c("Autosome", "Y-linked"), c("not_testis_specific", "testis_specific")])$p.value`)
+
+- But testis-preferential genes are not enriched on the X (pval 
+`r chisq.test(x[c("Autosome", "X-linked"), c("not_testis_specific", "testis_preferential")])$p.value`)
+
+- and testis-preferential genes are not enriched on the X (pval 
+`r chisq.test(x[c("Autosome", "Y-linked"), c("not_testis_specific", "testis_preferential")])$p.value`)
+
+```{r}
+chisq.test(x[c("Autosome", "X-linked"), c("not_testis_specific", "testis_preferential")])
+
+chisq.test(x[c("Autosome", "Y-linked"), c("not_testis_specific", "testis_preferential")])
+```
+
+### Chromosomal location, testis specificity and regulation by methylation
+
+Proportion of X-linked genes among Testis-specific genes regulated or not by 
+methylation.
+
+```{r}
+TS_genes_chr <- all_genes_chr %>% 
+  filter(testis_specificity == "testis_specific")
+x <- table(chr = TS_genes_chr$chr, 
+           TS_genes_regulated_by_met = TS_genes_chr$regulated_by_methylation, 
+           useNA = "ifany")
+x
+```
+
+- `r x["X-linked", "TRUE"]` / `r sum(x[, "TRUE"])`
+(`r x["X-linked", "TRUE"]/sum(x[, "TRUE"]) * 100`%) of testis-specific genes 
+regulated by methylation are X-linked
+
+- `r x["X-linked", "FALSE"]` / `r sum(x[, "FALSE"])`
+(`r x["X-linked", "FALSE"]/sum(x[, "FALSE"]) * 100`%) of testis-specific genes 
+ not regulated by methylatoin are X-linked
+
+```{r}
+chisq.test(x[c("Autosome", "X-linked"), c("FALSE", "TRUE")])
+chisq.test(x[c("Autosome", "X-linked"), c("FALSE", "TRUE")])$p.value
+```
+
+```{r}
+fig_TS_X <- TS_genes_chr %>% 
+  filter(!is.na(regulated_by_methylation)) %>% 
+  mutate(chr = case_when(chr == "X-linked" ~ "chrX",
+                         chr == "Y-linked" ~ "chrY",
+                         !chr %in% c("X-linked", "Y-linked") ~ "Autosome")) %>%
+  mutate(Regulation = case_when(
+    regulated_by_methylation ~ 
+      "Testis-specific genes\nregulated by methylation",
+    !regulated_by_methylation ~ 
+      "Testis-specific genes\nNOT regulated by methylation")) %>% 
+  mutate(Regulation = 
+           factor(Regulation, 
+                  levels = 
+                    c("Testis-specific genes\nNOT regulated by methylation", 
+                      "Testis-specific genes\nregulated by methylation"))) %>% 
+  mutate(Group = case_when(CT_gene_type == "CT_gene" ~ "CT gene",
+                           CT_gene_type != "CT_gene" ~ "Not CT gene")) %>% 
+  group_by(Regulation, chr, Group) %>% 
+  dplyr::count() %>% 
+  ggplot(aes(x = chr, y = n, fill = Group)) +
+  geom_col() + #width = 0.5) +
+  scale_fill_manual(values = c("seagreen2", "gray50")) +
+  facet_wrap(~ Regulation, scale = "free_y") +
+  xlab('') +
+  ylab("") +
+  theme_bw() +
+  theme(legend.position = "right",
+        plot.title = element_text(hjust = 0.5),
+        #title = element_text(size = 10, face = "bold",
+        #aspect.ratio = 1.5,
+        panel.grid = element_blank(),
+        strip.text = element_text(size = 9),
+        axis.text.x = element_text(size = 10),
+        axis.title.y = element_text(size = 12))
+
+fig_TS_X
+```
 
 ## Put figs together
+
+```{r}
+fig <- cowplot::plot_grid(fig_TS_CT_genes + theme(legend.position = "none"), 
+                          fig_TS_X + theme(legend.position = "none"),
+           rel_widths = c(1,2))
+
+legend <- cowplot::get_legend(fig_TS_CT_genes)
+
+cowplot::plot_grid(fig, legend, align = "v", rel_widths = c(1, 0.2))
+```
+
+```{r}
+pdf(file = paste0("figs/Testis_specific.pdf"), width = 8, height = 4)
+cowplot::plot_grid(fig, legend, 
+                   rel_widths = c(1, 0.2))
+
+dev.off()
+```
 
@@
                     name = "Regulation", value = "Percent"))
+
 prop_fig <- ggplot(x, aes(x = '', y = value, fill = Regulation)) +
@@
 
-right_fig <- cowplot::plot_grid(prop_fig, fig_freq_by_chr, labels = c('B', 'C'), 
+right_fig <- cowplot::plot_grid(prop_fig, fig_freq_by_chr, fig,
+                                labels = c('B', 'C', 'D'), 
                                 #hjust = -2, 
@@
 ```
+# X-linked are more testis-specific than autosomal genes
+
+```{r}
+TS_and_TP_genes_chr <- all_genes_chr %>% 
+  filter(CT_gene_type %in% c("CT_gene", "CTP_gene"))
+table(chr = TS_and_TP_genes_chr$chr, 
+      TS_and_TP_genes = TS_and_TP_genes_chr$testis_specificity)
+```
 
@@
 
-```{r biotype}
-prop.table(table(CT_genes$transcript_biotype))
+```{r}
+table(Gene = CT_genes_chr$chr, CT_gene_biotype = CT_genes_chr$transcript_biotype)
+```
+```{r}
+table(Gene = TS_genes_chr$chr, Testis_specific_genes = TS_genes_chr$transcript_biotype)
 ```
@@
 
-```{r biotype_reg}
-prop.table(table(methylation = CT_genes$regulated_by_methylation,
-                 CT_genes$transcript_biotype))
-
-prop.table(table(CT_genes$transcript_biotype,
-                 methylation = CT_genes$regulated_by_methylation))
+```{r}
+table(CT_genes_chr$transcript_biotype,
+                 methylation = CT_genes_chr$regulated_by_methylation)
 ```
@@
   pivot_longer(names_to = "barcode", values_to = "TPM", -ensembl_gene_id) %>% 
-  mutate(activated = ifelse(TPM > thr_pos, "n_pos", "n_neg")) %>% 
+  mutate(activated = ifelse(TPM >= thr_pos, "n_pos", "n_neg")) %>% 
   group_by(ensembl_gene_id, activated) %>% 
@@
   pivot_longer(names_to = "barcode", values_to = "TPM", -ensembl_gene_id) %>% 
-  mutate(activated = ifelse(TPM > thr_pos, "n_pos", "n_neg")) %>% 
+  mutate(activated = ifelse(TPM >= thr_pos, "n_pos", "n_neg")) %>% 
   filter(activated == "n_pos") %>% 
@@
   right_join(CT_genes_types) %>% 
-  dplyr::select(gene, mean_log, proportion_of_positive_tum, 
-                Regulation, X_linked) %>% 
+  dplyr::select(gene, mean_log, proportion_of_positive_tum, Regulation, 
+                X_linked) %>% 
   mutate(source = "TCGA tumor samples")
@@
 # combine with CCLE data
+load("~/cluster/Packages/CTdata/eh_data/CCLE_data.rda")
 ccle <- CCLE_data
@@
   pivot_longer(names_to = "cell", values_to = "TPM", -ensembl_gene_id) %>% 
-  mutate(activated = ifelse(TPM > thr_pos, "n_pos", "n_neg")) %>% 
+  mutate(activated = ifelse(TPM >= thr_pos, "n_pos", "n_neg")) %>% 
   group_by(ensembl_gene_id, activated) %>% 
@@
   pivot_longer(names_to = "cell", values_to = "TPM", -ensembl_gene_id) %>% 
-  mutate(activated = ifelse(TPM > thr_pos, "n_pos", "n_neg")) %>% 
+  mutate(activated = ifelse(TPM >= thr_pos, "n_pos", "n_neg")) %>% 
   filter(activated == "n_pos") %>% 
@@
   right_join(CT_genes_types) %>% 
-  dplyr::select(gene, mean_log, proportion_of_positive_tum, Regulation, X_linked) %>% 
+  dplyr::select(gene, mean_log, proportion_of_positive_tum, Regulation, 
+                X_linked) %>% 
   mutate(source = "CCLE tumor cell lines")
 
-mean_and_prop <- rbind(mean_and_prop_TCGA, mean_and_prop_CCLE)
-
-labelled_genes <- mean_and_prop %>% 
-  filter(proportion_of_positive_tum > 22 | mean_log > 4)
-mean_and_prop$source <- factor(mean_and_prop$source, 
-                               levels = c("TCGA tumor samples", "CCLE tumor cell lines"))
-
-fig_mean_prop_in_tum <- ggplot(mean_and_prop[!is.na(mean_and_prop$Regulation), ],
+mean_and_prop <- rbind(mean_and_prop_TCGA, mean_and_prop_CCLE) %>% 
+  filter(!is.na(Regulation))
+
+fig_mean_prop_in_tum_horizontal <- 
+  ggplot(mean_and_prop,
          aes(x = mean_log, y = proportion_of_positive_tum,
@@
       filter(proportion_of_positive_tum > 20 | 
-               (proportion_of_positive_tum > 5 & mean_log > 4)), 
+               (proportion_of_positive_tum > 10 & mean_log > 3.5)), 
     aes(x = mean_log, y = proportion_of_positive_tum, 
-        label = gene), size = 1.5, color = "black", min.segment.length = 1,
+        label = gene), size = 1.5, color = "black", min.segment.length = 4,
     max.overlaps = 30, direction = "both") +
-  xlab("Mean expression log(TPM + 1)\nin positive samples") +
-  ylab("Proportion of positive samples (%)") +
+  xlab("Mean expression log(TPM + 1) in positive tumors") +
+  ylab("Proportion of positive tumors (%)") +
   facet_wrap(~ source, ncol = 2) +
@@
         legend.text = element_text(size = 6),
-        legend.title = element_text(size = 7),
-        axis.text = element_text(size = 5),
+        legend.title = element_text(size = 8),
+        axis.text = element_text(size = 7),
         axis.ticks = element_line(size = 0.1),
-        axis.title = element_text(size = 6),
-        strip.text = element_text(size = 6))
-
-fig_mean_prop_in_tum
+        axis.title = element_text(size = 8),
+        strip.text = element_text(size = 8)) +
+  ylim(0,42)
+
+fig_mean_prop_in_tum_horizontal
 ```
 ```{r, eval = FALSE, echo = FALSE}
-pdf(file = paste0("figs/Mean_prop_in_tumors.pdf"), width = 5, height = 4)
-fig_mean_prop_in_tum
+pdf(file = paste0("figs/Mean_prop_in_tumors.pdf"), width = 6, height = 3)
+fig_mean_prop_in_tum_horizontal
 dev.off()
@@
 
-## Activation in TCGA, coexpression and global hypomethylation
-
-```{r AUC_score}
-keep <- rowSums(assay(TCGA)) > 0
-TCGA <- TCGA[keep, ]
-TCGA$project_id <- gsub("TCGA-", '', TCGA$project_id)
-
-cells_rankings <- suppressMessages(AUCell_buildRankings(assay(TCGA), 
-                                                        plotStats = TRUE, 
-                                                        verbose = FALSE))
-geneSetsMet <- list(methylation = CT_genes_types %>% 
-                      filter(Regulation == "Methylation") %>%  
-                      filter(ensembl_gene_id %in% rownames(TCGA)) %>% 
-                      pull(ensembl_gene_id),
-                    not_methylation = CT_genes_types %>% 
-                      filter(Regulation == "Not methylation") %>%  
-                      filter(ensembl_gene_id %in% rownames(TCGA)) %>% 
-                      pull(ensembl_gene_id))
-
-# Calculate enrichment scores
-cells_AUC_met <- AUCell_calcAUC(geneSetsMet, 
-                            cells_rankings, 
-                            aucMaxRank = ceiling(0.2 * nrow(cells_rankings)))
-
-TCGA$AUC_genes_regulated_by_met <- as.vector(assay(cells_AUC_met)["methylation", ])
-TCGA$AUC_genes_not_regulated_by_met <- as.vector(assay(cells_AUC_met)["not_methylation", ])
-```
-
-- Show One heatmap for CT-met with AUScores and Global hypo levels
-
-- Show another one with CT-notMet (low coexpression, no pattern, 
-uncorrelated with global hypo)
-
-Plotted on the heatmaps genes that have a TPM > 0 in at least 5% of tumor samples
-
-```{r TCGA_heatmap}
-legends_param_TCGA <- list(
-  labels_gp = gpar(col = "black", fontsize = 2),
-  title_gp = gpar(col = "black", fontsize = 3),
-  simple_anno_size = unit(0.1, "cm"),
-  row_names_gp = gpar(fontsize = 1),
-  annotation_name_side = "left",
-  border = FALSE,
-  border_gp = gpar(lwd = 0.1),
-  grid_width = unit(0.05, "cm"),
-  grid_height = unit(0.1, "cm"),
-  legend_height = unit(0.01, "cm"),
-  direction = "horizontal",
-  use_raster = TRUE,
-               raster_device = "CairoPNG",
-               raster_quality = 10)
-
-print_TCGA_heatmap_with_AUCscores <- function(genes = NULL,
-                                              tum_type = c("SKCM", "LUAD", 
-                                                           "LUSC", "COAD", 
-                                                           "BRCA", "HNSC"),
-                                              AUScores = "AUC_genes_met", 
-                                              scale_lims = NULL,
-                                              show_AUScore = TRUE,
+## Activation in TCGA
+
+```{r}
+# slightly modified the function 
+print_CCLE_heatmap <- function(genes = NULL,
+                               tum_type = c("Skin", "Lung", 
+                                            "Colorectal", 
+                                            "Breast", 
+                                            "Head_and_Neck",
+                                            "Esophageal"),
+                               h_width = 5,
+                               h_height = 10,
+                               downsample_to_n = 450,
                                fontsize = 4, 
@@
                                split_rows_by = "regulation", 
-                                              h_width = 7,
-                                              h_height = 5,
-                                              clustRow = TRUE,
-                                              clustCol = FALSE){
-
-  tcga <- TCGA[, TCGA$project_id %in% tum_type & !is.na(TCGA$global_methylation)]
-  tcga$project_id <- factor(tcga$project_id , levels = tum_type)
-
-  # Rm genes expressed in less than 5% of tumors
-  keep <- (rowSums(assay(tcga) > 0) / dim(tcga)[2]) > 0.05
-  tcga <- tcga[keep,]
-
-  df_col <- data.frame("project_id" = tcga$project_id,
-                       "AUScore" = tcga$AUC_genes_regulated_by_met,
-                       "global_methylation" = tcga$global_methylation,
-                       "gender" = tcga$gender)
-
-  if (AUScores == "AUC_genes_not_met") {
-    df_col <- data.frame("project_id" = tcga$project_id,
-                         "AUScore" = tcga$AUC_genes_not_regulated_by_met,
-                         "global_methylation" = tcga$global_methylation,
-                         "gender" = tcga$gender)
-  }
-
-  rownames(df_col) <- rownames(colData(tcga))
-  df_col <- df_col[!is.na(df_col$global_methylation),]
-  df_col <- df_col[order(df_col$project_id, df_col$AUScore), ]
-  #  df_col <- df_col[order(df_col$project_id), ]
-  split_by <- factor(df_col$project_id)
+                               show_annotation_legend = TRUE,
+                               clustRow = FALSE,
+                               clustCol = TRUE){
+
+  ## downsample tumors to have the same number of tumors in each type
+  ccle <- CCLE_data
+  CCLE <- ccle[, ccle$type == tum_type[1]]
+  CCLE <- CCLE[, 1:min(ncol(CCLE), downsample_to_n)]
+
+  for (tum in tum_type[-1]){
+    tmp <- ccle[, ccle$type == tum]
+    tmp <- tmp[, 1:min(ncol(tmp), downsample_to_n)]
+    CCLE <- cbind(CCLE, tmp)
+  }
+
+  split_by <- factor(CCLE$type, levels = tum_type)
 
   column_ha_tumor <- HeatmapAnnotation(
-    TCGA_tumor = df_col$project_id,
-    border = FALSE,
-    col = list(TCGA_tumor = TCGA_colors),
+    CCLE_cell_type = CCLE$type,
+   # border = TRUE,
+    col = list(CCLE_cell_type = CCLE_colors),
     annotation_label = gpar(fontsize = 4),
@@
     simple_anno_size = unit(0.1, "cm"),
-    annotation_legend_param = legends_param_TCGA)
-
-  AUCscore_col <- colorRamp2(seq(0, quantile(tcga$AUC_genes_regulated_by_met, 0.9), length = 4),
-                             c("gray", "gold", "firebrick1", "firebrick4"))
-
-  column_ha_AUC_score <- HeatmapAnnotation(
-    AUCscore = df_col$AUScore,
-    border = FALSE,
-    col = list(AUCscore = AUCscore_col),
-    annotation_label = gpar(fontsize = 4),
-    annotation_name_gp = gpar(fontsize = 0),
-    simple_anno_size = unit(0.1, "cm"),
-    annotation_legend_param = legends_param_TCGA)
-
-    global_met_col <-  colorRamp2(
-    c(quantile(tcga$global_methylation, 0.1), quantile(tcga$global_methylation, 0.9)), 
-                               c("moccasin","dodgerblue4"))
-
-  column_ha_global_methylation <- HeatmapAnnotation(
-    "Global methylation level" = df_col$global_methylation,
-    border = FALSE,
-    col = list("Global methylation level"  = global_met_col),
-    annotation_label = gpar(fontsize = 4),
-    annotation_name_gp = gpar(fontsize = 0),
-    simple_anno_size = unit(0.1, "cm"),
-    annotation_legend_param = legends_param_TCGA)
-
-  column_ha_gender <- HeatmapAnnotation(
-    gender = df_col$gender,
-    border = FALSE,
-    #col = list(TCGA_tumor = TCGA_colors),
-    annotation_label = gpar(fontsize = 4),
-    annotation_name_gp = gpar(fontsize = 0),
-   simple_anno_size = unit(0.1, "cm"),
-    annotation_legend_param = legends_param_TCGA)
+    annotation_legend_param = legends_param)
 
   ## Use gene names instead of ENSEMBL IDs
-  mat <- SummarizedExperiment::assay(tcga)
-  rownames(mat) <- rowData(tcga)$external_gene_name
-
-  gene_type <- CT_genes_types %>% 
+  mat <- SummarizedExperiment::assay(CCLE)
+  rownames(mat) <- rowData(CCLE)$external_gene_name
+  mat <- log1p(mat)
+
+  gene_type <- all_genes_clean %>% 
     filter(gene %in% genes) %>% 
-    filter(gene %in%  rownames(mat))
-  mat <- log1p(mat)
+    mutate(gene = factor(gene, levels = genes)) %>% 
+    arrange(gene)
 
  rowSplit <- NULL
-  if (!is.null(split_rows_by)){
+  if(!is.null(split_rows_by)){
     if (split_rows_by == "regulation") {
-      gene_type <- gene_type %>% arrange(Regulation)
+      if (clustRow) gene_type <- gene_type %>% arrange(Regulation)
       rowSplit <- gene_type$Regulation
     }
+    if (split_rows_by == "testis_specificity") {
+      if (clustRow) gene_type <- gene_type %>% arrange(testis_specificity)
+      rowSplit <- gene_type$testis_specificity
+    }
+    if (split_rows_by == "GTEX_specificity") {
+      if (clustRow) gene_type <- gene_type %>% arrange(GTEX_specificity)
+      rowSplit <- gene_type$GTEX_specificity
+    }
 
     if (split_rows_by == "chr_met") {
-      gene_type <- gene_type %>% arrange(chr_met)
+      if (clustRow) gene_type <- gene_type %>% arrange(chr_met)
       rowSplit <- gene_type$chr_met
@@
   row_ha_regulation <- rowAnnotation(Regulation = Regulation,
-                                     annotation_legend_param = legends_param_TCGA,
+                                     annotation_legend_param = legends_param,
                                      simple_anno_size = unit(0.2, "cm"),
@@
   row_ha_chr_met <- rowAnnotation(chr_met = chr_met,
-                                  annotation_legend_param = legends_param_TCGA,
+                                  annotation_legend_param = legends_param,
                                   simple_anno_size = unit(0.2, "cm"),
@@
   row_ha_chr <- rowAnnotation(Chr = Chr,
-                              annotation_legend_param = legends_param_TCGA,
+                              annotation_legend_param = legends_param,
                               simple_anno_size = unit(0.2, "cm"),
@@
 
-  if (is.null(scale_lims)) {
-    scale_lims <- c(0, quantile(rowMaxs(mat), 0.8))
-  }
-
-  if (show_AUScore) {
-    topAnnot <- c(#column_ha_tumor,
-                  column_ha_AUC_score,
-                  column_ha_global_methylation,
-                  gap = unit(0.5, "mm"))
-  } else {
-    topAnnot <- c(column_ha_tumor,
-                  column_ha_global_methylation,
-                  gap = unit(0.5, "mm"))
-  }
-
-  ht_opt$TITLE_PADDING = unit(c(2, 2), "points")
-
-  h <- Heatmap(mat[gene_type$gene, rownames(df_col)],
-               name = "log(TPM +1)",
-               column_title = levels(df_col$project_id),
-               column_title_gp = gpar(col = TCGA_colors[tum_type], 
-                                      fontsize = 3, face = "bold"),
+  GTEX_specificity <- as.matrix(gene_type$GTEX_specificity)
+  rownames(GTEX_specificity) <- gene_type$gene
+  row_ha_GTEX <- rowAnnotation(GTEX_specificity = GTEX_specificity,
+                             annotation_legend_param = legends_param,
+                             simple_anno_size = unit(0.2, "cm"),
+                             col = list(GTEX_specificity = 
+                                          GTEX_specificity_colors),
+                             annotation_name_gp = gpar(fontsize = 0))
+
+  testis_specificity <- as.matrix(gene_type$testis_specificity)
+  rownames(testis_specificity) <- gene_type$gene
+  row_ha_TS <- rowAnnotation(specificity = testis_specificity,
+                             annotation_legend_param = legends_param,
+                             simple_anno_size = unit(0.2, "cm"),
+                             col = list(testis_specificity = 
+                                          Testis_specificity_colors),
+                             annotation_name_gp = gpar(fontsize = 0))
+
+  rAnnot <- NULL
+  if(show_annotation_legend){
+    if (split_rows_by == "regulation") {
+  rAnnot <- c(row_ha_regulation, gap = unit(1, "mm"))
+    }
+    if (split_rows_by == "testis_specificity") {
+  rAnnot <- c(row_ha_TS, gap = unit(1, "mm"))
+    }
+    if (split_rows_by == "GTEX_specificity") {
+  rAnnot <- c(row_ha_GTEX, gap = unit(1, "mm"))
+    }
+
+    if (split_rows_by == "chr_met") {
+  rAnnot <- c(row_ha_TS, gap = unit(1, "mm"))
+    }
+  }
+
+  # heatmap_colors <- colorRamp2(
+  #   seq(0, quantile(rowMax(mat), 0.8), length = 11),                 
+  #   legend_colors)
+
+  heatmap_colors <- colorRamp2(
+    seq(0, 5, length = 11),                 
+    legend_colors)
+
+  Heatmap(mat[as.vector(gene_type$gene), ],
+          name = "CCLE (logTPM)",
+          #column_title = paste0("CCLE tumor cell lines"),
+          column_title_gp = gpar(fontsize = 2, face = "bold"),
           clustering_method_rows = clust_method,
@@
           column_split = split_by,
-               cluster_column_slices = FALSE,
-               row_gap = unit(c(1), "mm"),
-               column_gap = unit(c(1), "mm"),
-               col = colorRamp2(seq(scale_lims[1], scale_lims[2], length = 11),
-                                legend_colors),
+          row_gap = unit(c(0.5), "mm"),
+          col = heatmap_colors,
           cluster_rows = clustRow,
           cluster_columns = clustCol,
+          cluster_column_slices = FALSE,
           show_row_names = TRUE,
@@
           show_column_dend = FALSE,
-               column_names_gp = gpar(fontsize = 4),
+          column_names_gp = gpar(fontsize = 6),
           column_names_centered = TRUE,
@@
           border = TRUE,
-               border_gp = gpar(lwd = 0.05),
+          border_gp = gpar(lwd = 0.1),
           column_names_side = c("top"),
@@
           row_title_gp = gpar(fontsize = 0),
-               top_annotation = topAnnot,
+          #top_annotation = column_ha_tumor,
+          right_annotation = rAnnot,
           row_split = rowSplit,
-               cluster_row_slices = FALSE,
-               heatmap_width = unit(h_width, "cm"),
-               heatmap_height = unit(h_height, "cm"),
+          width = unit(h_width, "cm"),
+          height = unit(h_height, "cm"),
           use_raster = TRUE,
@@
           raster_quality = 10,
-               heatmap_legend_param = legends_param_TCGA,
+          heatmap_legend_param = legends_param,
           show_heatmap_legend = TRUE)
-
-  h
-}
-
+}
+```
+
+```{r}
 met_genes <- CT_genes_types %>% 
@@
 
-tum_types <- c("SKCM", "LUSC", "HNSC", "COAD")
-
-h1 <- print_TCGA_heatmap_with_AUCscores(genes = met_genes, 
-                                        clustCol = TRUE,
+h1 <- print_CCLE_heatmap(met_genes, 
+                         downsample_to_n = 100, 
+                         h_width = 2.5,
+                         h_height = 2.5,
+                         fontsize = .5, 
+                         tum_type = c("Skin", "Lung", 
+                                      "Brain", "Breast","Lymphoma", 
+                                      "Leukemia", "Colorectal"),
                    clust_method = "ward.D",
-                                        h_width = 6,
-                                        h_height = 5,
-                                        fontsize = 1, 
-                                        scale_lims = c(0, 6),
-                                        tum_type = tum_types,
-                                        split_rows_by = NULL)
+                   clustRow = TRUE,
+                  # show_row_names = TRUE,
+                  # show_top_annotations = FALSE,
+                   show_annotation_legend = FALSE,
+                   split_rows_by = "regulation")
 draw(h1, column_title = "CT genes regulated by methylation",
-     column_title_gp = gpar(fontsize = 4), 
-     merge_legend = TRUE)
+     column_title_gp = gpar(fontsize = 2.5), show_heatmap_legend = FALSE,
+     show_annotation_legend = FALSE)
 
@@
 
-h2 <- print_TCGA_heatmap_with_AUCscores(genes = not_met_genes, 
-                                       clustCol = TRUE,
-                                       AUScores = "AUC_genes_not_met", 
-                                       show_AUScore = TRUE,
+h2 <- print_CCLE_heatmap(not_met_genes, 
+                         fontsize = 0.5, 
+                         downsample_to_n = 100, 
+                         h_width = 2.5,
+                         h_height = 1,
+                         tum_type = c("Skin", "Lung", 
+                                      "Brain", "Breast","Lymphoma", 
+                                      "Leukemia", "Colorectal"),
                          clust_method = "ward.D",
-                                       h_width = 6,
-                                       h_height = 2.5,
-                                       fontsize = 1, 
-                                       scale_lims = c(0, 6),
-                                       tum_type = tum_types,
-                                       split_rows_by = NULL)
+                         clustRow = TRUE,
+                         show_annotation_legend = F,
+                         split_rows_by = "regulation")
+
 draw(h2, column_title = "CT genes not regulated by methylation",
-     column_title_gp = gpar(fontsize = 4))
-
-```
-
-```{r, echo = FALSE, eval = FALSE}
-pdf(file = "figs/CT_met_AUC_hypo_TCGA.pdf", height = 2.3, width = 3)
+     column_title_gp = gpar(fontsize = 2.5), show_heatmap_legend = TRUE,
+     heatmap_legend_side = "bottom",
+     show_annotation_legend = FALSE)
+
+```
+
+```{r, eval = FALSE, echo = FALSE}
+pdf(file = paste0("figs/CT_met_CCLE.pdf"), width = 1.3, height = 1.5)
 draw(h1, column_title = "CT genes regulated by methylation",
-     column_title_gp = gpar(fontsize = 5), show_heatmap_legend = TRUE,
-     show_annotation_legend = TRUE, heatmap_legend_side = "bottom",
-     merge_legend = TRUE)
+     column_title_gp = gpar(fontsize = 2.5), show_heatmap_legend = FALSE,
+     show_annotation_legend = FALSE)
 dev.off()
 
-pdf(file = "figs/CT_not_met_AUC_hypo_TCGA.pdf", height = 2.3, width = 3)
+pdf(file = "figs/CT_not_met_CCLE.pdf", height = 2, width = 1.5)
 draw(h2, column_title = "CT genes not regulated by methylation",
-     column_title_gp = gpar(fontsize = 4), merge_legend = TRUE, 
-     heatmap_legend_side = "bottom")
+     column_title_gp = gpar(fontsize = 2.5), show_heatmap_legend = TRUE,
+     heatmap_legend_side = "bottom",
+     show_annotation_legend = FALSE)
 dev.off()
@@
 
-## AUC correlation with global hypomethylation
-
-```{r AUC_hypometh}
-tcga <- TCGA[, TCGA$project_id %in% tum_types & !is.na(TCGA$global_methylation)]
-tcga$project_id <- factor(tcga$project_id , levels = tum_types)
-
-f1 <- as_tibble(colData(TCGA), rownames = "barcode") %>% 
-  filter(project_id %in% tum_types) %>% 
-  mutate(project_id = factor(project_id, levels = tum_types,)) %>% 
-  ggplot(aes(x = global_methylation, y = AUC_genes_regulated_by_met, 
-             color = project_id)) +
-  geom_point(size = 0.3) +
-  facet_wrap(~ project_id, nrow = 1) +
-  scale_color_manual(values = TCGA_colors) + 
-  geom_smooth(method = 'lm', color = "black", linewidth = 0.3, se = FALSE) +
-  xlab('Global methylation level') +
-  ylab("AUCscore CT genes\nregulated by methylation") +
-  theme(legend.position = "none",
-        aspect.ratio = 1,
-        panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank(),
-        axis.title.y = element_text(size = 8),
-        axis.title.x = element_text(size = 9),
-        axis.text = element_text(size = 4),
-        axis.ticks = element_line(size = 0.1)) +
-  ggpubr::stat_cor(method="pearson", size = 2.5, label.x = .4, 
-           label.y = 0.11, label.sep = "\n", color = "black") +
-  ylim(0, 0.12)
-
-f2 <- as_tibble(colData(TCGA), rownames = "barcode") %>% 
-  filter(project_id %in% tum_types) %>% 
-  mutate(project_id = factor(project_id, levels = tum_types,)) %>% 
-  ggplot(aes(x = global_methylation, y = AUC_genes_not_regulated_by_met, 
-             color = project_id)) +
-  geom_point(size = 0.3) +
-  facet_wrap(~ project_id, nrow = 1) +
-  scale_color_manual(values = TCGA_colors) + 
-  geom_smooth(method = 'lm', color = "black", linewidth = 0.3, se = FALSE) +
-  xlab('Global methylation level') +
-  ylab("AUCscore CT genes\nnot regulated by methylation") +
-  theme(legend.position = "none",
-        aspect.ratio = 1,
-        panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank(),
-        axis.title.y = element_text(size = 8),
-        axis.title.x = element_text(size = 9),
-        axis.text = element_text(size = 4),
-        axis.ticks = element_line(size = 0.1)) +
-  ggpubr::stat_cor(method="pearson", size = 2.5, label.x = .4, 
-           label.y = 0.11, label.sep = "\n", color = "black") +
-  ylim(0, 0.12)
-
-f1
-f2
+# Oncogenes and Tumor suppressor
+
+```{r}
+onco_TS_list <- list("Tumor suppressors" = CT_genes %>% 
+                       filter(tumor_suppressor == "tumor_suppressor") %>%
+                       pull(external_gene_name),
+                     Oncogenes = CT_genes %>% 
+                       filter(oncogene == "oncogene") %>%
+                       pull(external_gene_name))
+onco_TS_list
+
+gp <- VennThemes(compute.Venn(Venn(onco_TS_list)))
+gp[["Face"]][["10"]]$fill <-  "lightsteelblue1"
+gp[["Face"]][["01"]]$fill <-  "mediumaquamarine"
+gp[["Face"]][["11"]]$fill <-  "gray80"
+gp[["Set"]][["Set2"]]$col <-  "darkseagreen4"
+gp[["Set"]][["Set1"]]$col <-  "lightsteelblue4"
+gp[["SetText"]][["Set1"]]$col <-  "steelblue"
+gp[["SetText"]][["Set2"]]$col <-  "paleturquoise4"
+gp[["FaceText"]][["10"]]$col <-  "black"
+plot(Vennerable::Venn(onco_TS_list), gp = gp)
 ```
@@
 ```{r, eval = FALSE, echo = FALSE}
-pdf(file = paste0("figs/CT_met_AUCscore_hypo_TCGA.pdf"), height = 2.5)
-f1
+pdf(file = paste0("figs/Oncogenic_CT_Venn.pdf"))
+plot(Venn(onco_TS_list), gp = gp)
 dev.off()
-
-pdf(file = paste0("figs/CT_not_met_AUCscore_hypo_TCGA.pdf"), height = 2.5)
-f2
-dev.off()
 ```
@@
 
+## Heatmap
+
 First treating data
 
-```{r testis_data}
-# Load testis scRNAseq data
+```{r}
+# Missing genes are probably just not detected
+# Add zero counts to  missing genes in all datasets
 
@@
   CT_genes$external_gene_name %in% rownames(testis_sce)],]
+
+# add missing CT genes (not in the downloaded dataset, as not detected 
+# (below detection) => were removed)
+# Add zero count values for these genes
+rowData(testis_sce) <- NULL
+missing_CT <- CT_genes$external_gene_name[!CT_genes$external_gene_name %in% 
+                                            rownames(testis_sce)]
+missing_CT
+sce_missing_CT <- testis_sce[1:length(missing_CT),]
+rownames(sce_missing_CT) <- missing_CT
+count_mat_missing_CT <- matrix(nrow = length(missing_CT), 
+                               ncol = ncol(testis_sce), data = 0)
+rownames(count_mat_missing_CT) <- missing_CT
+colnames(count_mat_missing_CT) <- colnames(testis_sce)
+assay(sce_missing_CT,1) <- count_mat_missing_CT
+assay(sce_missing_CT,2) <- count_mat_missing_CT
+testis_sce <- rbind(testis_sce, sce_missing_CT)
+
+# add missing CT genes (not in the downloaded dataset, as not detected 
+# (below detection) => were removed)
+# Add zero count values for these genes
+rowData(FGC_sce) <- NULL
+
+FGC_sce <- FGC_sce[CT_genes$external_gene_name[
+  CT_genes$external_gene_name %in% rownames(FGC_sce)],]
+
+missing_CT <- CT_genes$external_gene_name[!CT_genes$external_gene_name %in%
+                                            rownames(FGC_sce)]
+sce_missing_CT <- FGC_sce[1:length(missing_CT),]
+rownames(sce_missing_CT) <- missing_CT
+count_mat_missing_CT <- matrix(nrow = length(missing_CT), 
+                               ncol = ncol(FGC_sce), data = 0)
+rownames(count_mat_missing_CT) <- missing_CT
+colnames(count_mat_missing_CT) <- colnames(FGC_sce)
+assay(sce_missing_CT,1) <- count_mat_missing_CT
+assay(sce_missing_CT,2) <- count_mat_missing_CT
+FGC_sce <- rbind(FGC_sce, sce_missing_CT)
+
+# add missing CT genes (not in the downloaded dataset, as not detected (below 
+# detection) => were removed)
+# Add zero count values for these genes
+rowData(oocytes_sce) <- NULL
+oocytes_sce <- oocytes_sce[CT_genes$external_gene_name[
+  CT_genes$external_gene_name %in% rownames(oocytes_sce)],]
+missing_CT <- CT_genes$external_gene_name[!CT_genes$external_gene_name %in% 
+                                            rownames(oocytes_sce)]
+sce_missing_CT <- oocytes_sce[1:length(missing_CT),]
+rownames(sce_missing_CT) <- missing_CT
+count_mat_missing_CT <- matrix(nrow = length(missing_CT), 
+                               ncol = ncol(oocytes_sce), data = 0)
+rownames(count_mat_missing_CT) <- missing_CT
+colnames(count_mat_missing_CT) <- colnames(oocytes_sce)
+assay(sce_missing_CT,1) <- count_mat_missing_CT
+assay(sce_missing_CT,2) <- count_mat_missing_CT
+oocytes_sce <- rbind(oocytes_sce, sce_missing_CT)
 
@@
 testis_sce[, testis_sce$type %in% 
-             c("Round_spermatid","Elongated_spermatid", 
-               "Sperm")]$stage <- "post-meiotic"
+             c("Round_spermatid", "Elongated_spermatid", "Sperm")]$stage <- 
+  "post-meiotic"
 testis_sce[, testis_sce$type %in% 
@@
 
-# Keep genes detected in at least x % of germ cells 
-# ! Each cell type is present at different proportion 
-# (much more sperm cells than SSC cells for instance)
-# table(testis_sce$type)
-# => Calculate for each gene the proportion of expressing cells 
-# (> testis_logcount_thr) and keep genes expressed in at least x% of a cell type
-# instead of just keeping genes expressed in at least x% of cells.
-testis_percent_thr <- 0
-testis_logcount_thr <- 0
-
-percent_of_expressing_cells <- 
-  tibble(gene = rownames(testis_sce),
-         prop = rowSums(assay(testis_sce[, testis_sce$type %in% germcells[1]], 
-                              "logcounts") > testis_logcount_thr) / 
-           dim(testis_sce[, testis_sce$type %in% germcells[1]])[2])
-names(percent_of_expressing_cells) <- c("gene", germcells[1])
-
-for (cell in germcells[-1]) {
-  tmp <- 
-    tibble(gene = rownames(testis_sce),
-           prop = rowSums(assay(testis_sce[, testis_sce$type %in% cell], 
-                                           "logcounts") > testis_logcount_thr) / 
-             dim(testis_sce[, testis_sce$type %in% cell])[2])
-  names(tmp) <- c("gene", cell)
-  percent_of_expressing_cells <- left_join(percent_of_expressing_cells, tmp)
-}
-
-kept_genes <- rownames(testis_sce)[
-  rowSums(percent_of_expressing_cells[, -1] >= testis_percent_thr) > 0]
-testis_sce <- testis_sce[kept_genes,]
-
-## Load FGC and oocytes scRNAseq data
-
-selected_CT <- BiocGenerics::intersect(rownames(FGC_sce), rownames(testis_sce))
-selected_CT <- BiocGenerics::intersect(rownames(oocytes_sce), selected_CT)
-
-FGC_sce <- FGC_sce[selected_CT, ]
-testis_sce <- testis_sce[selected_CT, ]
-oocytes_sce <- oocytes_sce[selected_CT, ]
-
-legends_param_sc <- list(
-  labels_gp = gpar(col = "black", fontsize = 3),
-  title_gp = gpar(col = "black", fontsize = 4),
+# Remove genes with rowcounts = 0 in all datasets
+# !!! Few genes are not detetected in testis_Sce but in FGC or oocytes
+not_in_testis_sce <- rownames(testis_sce)[rowSums(assay(testis_sce)) == 0]
+not_in_FGC_sce <- rownames(FGC_sce)[rowSums(assay(FGC_sce)) == 0]
+not_in_oocytes_sce <- rownames(oocytes_sce)[rowSums(assay(oocytes_sce)) == 0]
+not_in_testis_FGC <- BiocGenerics::intersect(not_in_testis_sce, not_in_FGC_sce)
+not_in_testis_FGC_oocytes <- BiocGenerics::intersect(not_in_testis_FGC, 
+                                                     not_in_oocytes_sce)
+
+kept_genes <- CT_genes %>% 
+  filter(!is.na(regulated_by_methylation)) %>% 
+  filter(!external_gene_name %in% not_in_testis_FGC_oocytes) %>% 
+  pull(external_gene_name)
+
+never <- CT_genes %>% 
+  filter(!is.na(regulated_by_methylation)) %>% 
+  filter(external_gene_name %in% not_in_testis_FGC_oocytes) %>% 
+  pull(external_gene_name)
+
+legends_param <- list(
+  labels_gp = gpar(col = "black", fontsize = 2),
+  title_gp = gpar(col = "black", fontsize = 3),
   simple_anno_size = unit(0.15, "cm"),
@@
 
-```
-
-Kept `r length(selected_CT)` CT genes for the analysis:
-
-CT genes with logcounts >= `r testis_logcount_thr` in at least 
-`r testis_percent_thr * 100` % of any specific type of testis germ cells 
-(adult testis dataset) and present in FCG and oocytes datasets
+# To label Y linked genes on the heatmaps
+CT_genes_types <- CT_genes_types %>% 
+  mutate(chromosome = case_when(chr == "X" ~ "X-linked",
+                                chr == "Y" ~ "Y-linked",
+                                !chr %in% c("X", "Y") ~ "Autosome"))
+table(CT_genes_types$chromosome)
+```
+Kept `r length(kept_genes)` CT genes for the analysis:
+
+CT genes with logcounts >= 0 in at least  one of the dataset
 
@@
                              ncells_max = NULL,
+                             cell_type = NULL,
                              scale_lims = NULL,
@@
                              h_height = 10,
-                             show_row_annot = TRUE, 
-                             row_title = TRUE){
+                             show_right_annot = TRUE){
 
   fetal_sce <- FGC_sce
+
+  if (!is.null(cell_type)) fetal_sce <- fetal_sce[,FGC_sce$type %in% cell_type]
 
@@
     simple_anno_size = unit(0.15, "cm"), 
-    annotation_legend_param = legends_param_sc,
+    annotation_legend_param = legends_param,
     annotation_name_gp = gpar(fontsize = 0))
@@
   }
+  if (split_by == "met_3_groups") {
+    gene_type <- gene_type %>% arrange(met_3_groups)
+    rowSplit <- gene_type$met_3_groups
+  }
+   if (!clusterRow & split_by == "met_3_groups") {
+    gene_type$gene <- factor(gene_type$gene, levels = genes)
+    gene_type <- gene_type  %>%
+      arrange(gene)
+    gene_type$gene <- as.vector(gene_type$gene)
+    rowSplit <- gene_type$met_3_groups
+   }
 
@@
     Regulation = Regulation,
-    annotation_legend_param = legends_param_sc,
+    annotation_legend_param = legends_param,
     border = FALSE,
@@
     chr_met = chr_met,
-    annotation_legend_param = legends_param_sc,
+    annotation_legend_param = legends_param,
     border = FALSE,
@@
 
-  Chr <- as.matrix(gene_type$Chr)
+  Chr <- as.matrix(gene_type$chromosome)
   rownames(Chr) <- gene_type$gene
   row_ha_chr <- rowAnnotation(Chr = Chr,
-                              annotation_legend_param = legends_param_sc,
+                              annotation_legend_param = legends_param,
                               border = FALSE,
                               simple_anno_size = unit(0.15, "cm"), 
-                              col = list(Chr = Chr_colors),
+                              col = list(Chr = Chr_colors_with_Y),
                               annotation_name_gp = gpar(fontsize = 0))
@@
 
-  if (!show_row_annot) rAnnot <- NULL
-
-  if (row_title) rtitle <- gpar(fontsize = 5, fontface = "bold") else rtitle <- gpar(fontsize = 0)
+  if (!show_right_annot) rAnnot <- NULL
 
@@
                      show_column_names = FALSE,
-                     show_row_names = FALSE,
+                     show_row_names = TRUE,
                      row_split = rowSplit,
@@
                                         column_ha_type),  
-                     left_annotation = rAnnot,
+                     right_annotation = rAnnot,
                      column_title_gp = gpar(fontsize = 5),
-                     row_title_gp = rtitle,
+                     row_title_gp = gpar(fontsize = 0),
                      gap = unit(0.01, "mm"),
@@
                      show_heatmap_legend = show_legend,
-                     heatmap_legend_param = legends_param_sc,
-                     row_title_rot = 0)
+                     heatmap_legend_param = legends_param)
 
@@
                                 h_height = 10,
-                                clust_method = "ward.D",
-                                show_row_annot = TRUE){
+                                clust_method = "ward.D"){
 
@@
     col = list(Testis_cell_type = testis_colors),      
-    annotation_legend_param = legends_param_sc,
+    annotation_legend_param = legends_param,
     border = TRUE, 
@@
     col = list(stage = stage_colors),
-    annotation_legend_param = legends_param_sc,
+    annotation_legend_param = legends_param,
     border = TRUE, 
@@
     col = list(sex = sex_colors),
-    annotation_legend_param = legends_param_sc,
+    annotation_legend_param = legends_param,
     border = TRUE, 
@@
 
+  if (split_by == "met_3_groups") {
+    gene_type <- gene_type %>% arrange(met_3_groups)
+    rowSplit <- gene_type$met_3_groups
+  }
+  if (!clusterRow & split_by == "met_3_groups") {
+    gene_type$gene <- factor(gene_type$gene, levels = genes)
+    gene_type <- gene_type  %>%
+      arrange(gene)
+    gene_type$gene <- as.vector(gene_type$gene)
+    rowSplit <- gene_type$met_3_groups
+  }
+
   Regulation <- as.matrix(gene_type$Regulation)
@@
 
-  Chr <- as.matrix(gene_type$Chr)
+  Chr <- as.matrix(gene_type$chromosome)
   rownames(Chr) <- gene_type$gene
   row_ha_chr <- rowAnnotation(Chr = Chr,
-                              annotation_legend_param = legends_param_sc,
+                              annotation_legend_param = legends_param,
                               border = FALSE,
                               simple_anno_size = unit(0.15, "cm"), 
-                              col = list(Chr = Chr_colors),
-                              annotation_name_gp = gpar(fontsize = 0))
-
-  rAnnot <- c(row_ha_regulation, row_ha_chr, gap = unit(1, "mm"))
-
-  if (!show_row_annot) rAnnot <- NULL
+                              col = list(Chr = Chr_colors_with_Y),
+                              annotation_name_gp = gpar(fontsize = 0))
 
@@
                        show_column_names = FALSE,
-                       show_row_names = TRUE,
+                       show_row_names = FALSE,
                        row_split = rowSplit,
@@
                                           column_ha_cell_type),  
-                       right_annotation = rAnnot,
+                       right_annotation = c(row_ha_regulation, 
+                                            row_ha_chr, 
+                                            gap = unit(1, "mm")),
                        column_title_gp = gpar(fontsize = 5),
@@
                        show_heatmap_legend =  show_legend,
-                       heatmap_legend_param = legends_param_sc)
+                       heatmap_legend_param = legends_param)
 
@@
     annotation_name_gp = gpar(fontsize = 0),
-    annotation_legend_param = legends_param_sc)
+    annotation_legend_param = legends_param)
 
@@
   }
+  if (split_by == "met_3_groups") {
+    gene_type <- gene_type %>% arrange(met_3_groups)
+    rowSplit <- gene_type$met_3_groups
+  }
+  if (!clusterRow & split_by == "met_3_groups") {
+    gene_type$gene <- factor(gene_type$gene, levels = genes)
+    gene_type <- gene_type  %>%
+      arrange(gene)
+    gene_type$gene <- as.vector(gene_type$gene)
+    rowSplit <- gene_type$met_3_groups
+  }
 
@@
                         show_heatmap_legend = show_legend,
-                      heatmap_legend_param = legends_param_sc)
+                        heatmap_legend_param = legends_param)
 
@@
 # Manual adjustments to visualise
-
 X_met_new_order <- 
-  c("SPANXB1", "SPANXD", "SPANXC", "SPANXN3", "CXorf51B", "H2AB1", "DDX53", 
+  c("SPANXB1", "SPANXD", "SPANXC",  "CXorf51B", "DDX53", 
     "TGIF2LX", "FTHL17","VCX3B", "VCX", "VCX3A", "VCX2", "FMR1NB",  "PAGE2", 
-    "PAGE5", "MAGEC2","GAGE2A","PAGE1", "SSX3", "FAM9A",  "CTAG2",  "MAGEA4", 
+    "PAGE5", "MAGEB6", "MAGEC2","GAGE2A","PAGE1",  "CTAG2",  "MAGEA4", 
     "PASD1","MAGEB2","CT55","MAGEA10",  "MAGEA3",  "MAGEA6","GAGE1", "MAGEA1", 
-    "CT45A10", "MAGEA11", "SSX1", "MAGEA9B","MAGEA9", "MAGEA8", "LUZP4",
+    "CT45A10", "MAGEA11", "TEX13C", "SSX1", "MAGEA9B","MAGEA9", "MAGEA8", 
     "MAGEB1", "MAGEA12", "MAGEC1", "MAGEB16","GAGE12E",  "XAGE5", "GAGE13", 
-    "SSX5", "CT45A1", "SSX2B", "XAGE1B", "RHOXF2B", "DCAF8L2", "SSX2", "SAGE1", 
-    "CXorf49", "CTAG1A", "NXF2","CT47A6", "CT45A5",  "CTAG1B" )
+    "SSX5",  "SSX2B", "XAGE1B",  "DCAF8L2", "SSX2", "SAGE1", 
+    "CXorf49", "CTAG1A", "NXF2",  "CTAG1B", "XAGE1A", 
+    "CXorf49B","PPP4R3C", "CT47A12",
+    "CT47A11", "CT45A1","CT45A5","CT45A7", "MAGEA4-AS1", "MAGEA2",
+      "CENPVL2", "CENPVL1", "CT47A1",  "CT45A3",  "CT45A6", "CT45A8", "CT45A9")
 
 not_X_met_new_order <- 
-  c("SPMIP9", "FLJ36000", "TRIML1", "TRIML2", "ADAD1", "TUBA3C", "COX7B2", 
-    "DDX4", "BRDT", "DSCR8", "DMRTB1",  "DAZL","RBM46", "DCAF4L2",  "CCDC172",
-    "LINC02475", "TGIF2LY", "ASZ1", "RNF17", "PRDM9", "CTCFL", "LINC00470", 
-    "POTEC", "TMEM132D-AS1", "DPPA2", "LINC01413", "NLRP4", "LINC01030", 
-    "ZSCAN5B", "DPPA5", "LINC01518", "LINC00200", "RFPL4B", "KRT85", "POTEF", 
-    "GFY", "RBMY1E", "RBMY1A1")
-
-X_not_met_new_order <- c( "SPANXN4", "DUSP21", "ESX1", "MAGEB4")
-
-not_X_not_met_new_order <- 
-  c("SMCP","LINC01548","WFDC11", "LINC01467","LINC00408", "LINC01193", "ADAM29", "SPATA16",
-    "IZUMO3", "LIN28A","C1orf94", "NUTM1", "GOLGA8S", "LINC01206", "ADAM18", 
-    "ADAM2","SYCP1", "MEIOB", "PRDM7", "TSPY3", 
-    "TSPY10", "TSPY8", "TSPY2","DMRT1", "LINC01511", "TSPY4", "TSPY1")
-
-my_genes <- c(X_met_new_order, not_X_met_new_order, X_not_met_new_order, not_X_not_met_new_order)
+  c( "FLJ36000", "TRIML2", "LINC01413", "NLRP4", "SUN3",
+    "LINC02475",  "TUBA3C", "COX7B2", 
+     "BRDT", "DSCR8",  "RBM46", "RNF17", "DCAF4L2",
+     "POTEC","CTCFL", "PRDM9",  "LINC00470", 
+    "TMEM132D-AS1", "DPPA2", 
+     "LINC01518", "LINC00200", "RFPL4B", "POTEF", 
+      "GFY", "SLC7A11-AS1",
+    "KAZN-AS1", "LINC02377",    "LINC02241",   "SPATA31C2",   
+    "FAM230C", "GOLGA6L7", "TEX48", "LINC02663", "LINC02864",
+    "RBMY1E", "VCY", "VCY1B", "RBMY1A1","RBMY1B"   )
+
+not_met_new_order <- 
+  c("CHCT1", "RAD21L1", "CSTL1", "WDR64", "CBY2", "GARIN2","LINC01193", 
+    "FAM217A", "LIN28A", "LINC01249", 
+    "CDRT15","C1orf94", "LINC01206", "SYCP1", "MEIOB", "MAJIN",
+    "PRDM7",  "LINC01098", "ESX1", "DMRT1", "LINC01511", "ZNF729", 
+    "NFIA-AS1", "THORLNC", "LINC02228", "SEC61G-DT", "ZNF722",  "ZNF723",
+    "LRRC52-AS1",
+
+     "LINC03004", 
+    "GTF2I-AS1", "LINC02750",  "TSPY4", "TSPY1", "TSPY9", "TSPY3", "TSPY2")
+
+my_genes <- c(X_met_new_order, not_X_met_new_order, not_met_new_order)
 
 fs <- 2
+scale_limits <- c(0, 4)
 ncells <- 100
-scale_limits <- c(0, 4)
 ht_testis <- draw_testis_heatmap(my_genes,
@@
                                  clusterRow  = FALSE,
-                    split_by = "chr_met",
+                                 split_by = "met_3_groups",
                                  show_legend = TRUE,
-                    clust_method = "ward.D2",
-                    show_row_annot = FALSE)
+                                 clust_method = "ward.D2")
 
@@
                                   show_legend = FALSE,
-                 show_row_annot = TRUE,
-                 row_title = TRUE)
+                                  show_right_annot = FALSE)
 
@@
                                 show_legend = FALSE,
-                 show_row_annot = FALSE,
-                 row_title = FALSE)
+                                show_right_annot = FALSE)
 
@@
 ht_list <- female_FGC_ht + ht_oocytes + male_FGC_ht + ht_testis
-draw(ht_list, main_heatmap = "Logcounts", merge_legend = TRUE, 
+draw(ht_list, main_heatmap = "logcounts", merge_legend = TRUE, 
      ht_gap = unit(0.2, "cm"))
@@
 pdf(file = paste0("figs/CT_in_GermCells.pdf"), width = 6, height = 4)
-draw(ht_list, main_heatmap = "Logcounts", 
+draw(ht_list, main_heatmap = "logcounts", 
      merge_legend = TRUE, 
@@
 
+## Numbers
+
+```{r}
+CT_genes_not_Y <- CT_genes %>% 
+  filter(chr != "Y") %>% 
+  filter(!external_gene_name %in% never) %>% 
+  pull(external_gene_name)
+
+dds_FCE <- scuttle::summarizeAssayByGroup(FGC_sce, 
+                            ids = colData(FGC_sce)[, "type"],
+                            statistics = "prop.detected")
+dds_oocytes <- scuttle::summarizeAssayByGroup(oocytes_sce, 
+                            ids = colData(oocytes_sce)[, "type"],
+                            statistics = "prop.detected")
+testis_Germ_cells <- testis_sce[, testis_sce$type %in% germcells]
+dds_testis <- scuttle::summarizeAssayByGroup(testis_Germ_cells, 
+                            ids = colData(testis_Germ_cells)[, "type"],
+                            statistics = "prop.detected")
+
+Male_cells <- cbind(assay(dds_testis[CT_genes_not_Y, ]) > 0.1, 
+      assay(dds_FCE[CT_genes_not_Y, c("M_PGC", "M_GC", 
+                                      "M_pre_spermatogonia")]) > 0.1)
+
+Female_cells <- cbind(assay(dds_FCE[CT_genes_not_Y, 
+                                    c("F_PGC", "F_GC", 
+                                      "F_oogonia", "F_oocyte")]) > 0.1, 
+      assay(dds_oocytes[CT_genes_not_Y, c("Growing oocytes", 
+                                          "Metaphase II")]) > 0.1)
+
+prop_CT_genes_by_gender <- tibble(gene = CT_genes_not_Y, 
+                                  in_female = rowSums(Female_cells) > 0,
+       in_male = rowSums(Male_cells) > 0) %>% 
+  summarize(Female = sum(in_female)/length(CT_genes_not_Y),
+            Male = sum(in_male)/length(CT_genes_not_Y)) %>% 
+  mutate(gene = "CT gene") %>% 
+  pivot_longer(names_to = "Germ cell", 
+               values_to = "Proportion of CT genes detected in germ cells",
+               -gene)
+
+A <- prop_CT_genes_by_gender %>% 
+  ggplot(aes(x = `Germ cell`, 
+             y = `Proportion of CT genes detected in germ cells`, 
+             fill = `Germ cell`)) +
+  geom_col() +
+  scale_fill_manual(values = c("Female" = "deeppink", "Male" = "steelblue")) +
+  theme_bw() +
+  theme(aspect.ratio = 1.8) +
+  xlab('') +
+  theme(legend.position = "none")
+
+prop_CT_genes_by_gender_by_group <- 
+  tibble(gene = CT_genes_not_Y, in_female = rowSums(Female_cells) > 0,
+in_male = rowSums(Male_cells) > 0) %>% 
+  left_join(CT_genes %>% 
+              dplyr::rename(gene = external_gene_name) %>% 
+              dplyr::select(gene, X_linked, regulated_by_methylation)) %>% 
+  mutate(group = case_when(X_linked & regulated_by_methylation ~ "met\nX-linked",
+                           !X_linked & regulated_by_methylation ~ "met\nAutosome",
+                           !regulated_by_methylation ~ "Not\nmet")) %>% 
+  mutate(group = factor(group, levels = c("met\nX-linked", "met\nAutosome", 
+                                          "Not\nmet"))) %>% 
+  group_by(group) %>% 
+  summarize(Female = sum(in_female), Male = sum(in_male)) %>% 
+  pivot_longer(names_to = "Germ cell", 
+               values_to = "Number of genes detected", -group) %>% 
+  filter(!is.na(group))
+
+B <-  prop_CT_genes_by_gender_by_group %>%  
+  ggplot(aes(x = group, y = `Number of genes detected`, fill = `Germ cell`)) +
+  geom_col() +
+   scale_fill_manual(values = c("Female" = "deeppink", "Male" = "steelblue")) +
+  theme_bw() +
+  theme(aspect.ratio = 1.8) +
+  xlab('')
+
+x <- tibble(gene = CT_genes_not_Y, in_female = rowSums(Female_cells) > 0,
+in_male = rowSums(Male_cells) > 0) %>% 
+  left_join(CT_genes %>% dplyr::rename(gene = external_gene_name) %>% 
+              dplyr::select(gene, X_linked, regulated_by_methylation)) %>% 
+  mutate(group = case_when(X_linked & regulated_by_methylation ~ "met\nX-linked",
+                           !X_linked & regulated_by_methylation ~ "met\nAutosome",
+                           !regulated_by_methylation ~ "Not\nmet")) %>% 
+  mutate(group = factor(group, 
+                        levels = c("met\nX-linked", "met\nAutosome", 
+                                   "Not\nmet"))) %>% 
+  group_by(group) %>% 
+  summarize(Female = sum(in_female), Male = sum(in_male)) %>% 
+  filter(!is.na(group))
+x
+chisq.test(x[,-1])
+
+fig <- cowplot::plot_grid(A + theme(legend.position = "none"), 
+                          B + theme(legend.position = "none"),
+           rel_heights = c(1, 1))
+
+legend <- cowplot::get_legend(B)
+cowplot::plot_grid(fig, legend, align = "v", rel_widths = c(1, 0.2))
+```
+
+```{r, echo = FALSE, eval = FALSE}
+pdf(file = paste0("figs/Prop_Female_Male.pdf"), width = 6.8, height = 4)
+cowplot::plot_grid(fig, legend,
+                rel_widths = c(2, 1))
+dev.off()
+```
+
+# CT methylation in fetal germ cells
+
+```{r}
+scWGBS_FGC_SE$cellType <- factor(scWGBS_FGC_SE$cellType, 
+                          levels = c("F_FGC", "F_meiotic_FGC","F_oogenesis_FGC",
+                                     "M_mitotic_FGC", "M_mitotic_arrested_FGC",
+                                     "F_soma","M_soma"))
+
+# Changed cellType names for coherence with expression heatmap
+scWGBS_FGC_SE$cellType <- as.vector(scWGBS_FGC_SE$cellType)
+scWGBS_FGC_SE$cellType[scWGBS_FGC_SE$cellType == "F_FGC"] <- "F_GC"
+scWGBS_FGC_SE$cellType[scWGBS_FGC_SE$cellType == "F_meiotic_FGC"] <- "F_oogonia"
+scWGBS_FGC_SE$cellType[scWGBS_FGC_SE$cellType == "F_oogenesis_FGC"] <- "F_oocyte"
+scWGBS_FGC_SE$cellType[scWGBS_FGC_SE$cellType == "M_mitotic_FGC"] <- "M_GC"
+scWGBS_FGC_SE$cellType[scWGBS_FGC_SE$cellType == "M_mitotic_arrested_FGC"] <- "M_pre_spermatogonia"
+
+scWGBS_FGC_SE$cellType <- factor(scWGBS_FGC_SE$cellType, 
+                          levels = c("F_GC", "F_oogonia","F_oocyte",
+                                     "M_GC", "M_pre_spermatogonia",
+                                     "F_soma","M_soma"))
+```
+
+## Mean meth values heatmap
+
+```{r, warning = FALSE, message = FALSE}
+nt_up <- 1000
+nt_down <- 500
+
+CT_promoter_regions <- as_tibble(CT_promoter_regions_hg19_GR) %>% 
+  dplyr::select(external_gene_name, chr, TSS_liftover, strand) %>%
+  mutate(start = case_when(strand == "+" ~ TSS_liftover - nt_up,
+                           strand == "-" ~ TSS_liftover - nt_down)) %>%
+  mutate(stop = case_when(strand == "+" ~ TSS_liftover + nt_down,
+                          strand == "-" ~ TSS_liftover + nt_up)) %>%
+  mutate(chromosome_name = paste0("chr", chr))
+
+CT_prom_GR <- GenomicRanges::makeGRangesFromDataFrame(CT_promoter_regions,
+                                                      keep.extra.columns = TRUE,
+                                                      seqnames.field = "chromosome_name",
+                                                      start.field = "start",
+                                                      end.field = "stop")
+
+tested_gene_GR <- subsetByOverlaps(scWGBS_FGC_SE, CT_promoter_regions_hg19_GR[1])
+mat <- as.matrix(assay(tested_gene_GR))
+met_means <- enframe(colMeans(mat, na.rm = TRUE))
+names(met_means) <- c("sample", CT_promoter_regions_hg19_GR[1]$external_gene_name)
+
+for (i in 2:length(CT_promoter_regions_hg19_GR)){
+    tested_gene_GR <- subsetByOverlaps(scWGBS_FGC_SE, CT_promoter_regions_hg19_GR[i])
+
+      mat <- as.matrix(assay(tested_gene_GR))
+
+      tmp <- enframe(colMeans(mat, na.rm = TRUE))
+      names(tmp) <- c("sample", CT_promoter_regions_hg19_GR[i]$external_gene_name)
+      met_means <- left_join(met_means, tmp)
+  }
+
+met_means <- met_means %>% pivot_longer(names_to = "gene", values_to = "met", -sample) %>% 
+      pivot_wider(names_from = sample, values_from = met)
+```
+
+Methylation in promoter regions defined as `r, nt_up` nt upstream TSS 
+and `r, nt_down` nt downstream TSS
+
+```{r}
+# add NA when missing CT genes
+
+missing_CT <- CT_genes$external_gene_name[!CT_genes$external_gene_name %in% met_means$gene]
+met_means_missing_CT <- met_means[1:length(missing_CT),]
+met_means_missing_CT$gene <- missing_CT
+met_means_missing_CT[, 2:ncol(met_means_missing_CT)] <- NA
+met_means <- rbind(met_means, met_means_missing_CT)
+
+plot_mean_methylation_heatmap <- function(genes, 
+                                          clusterRows = F,
+                                          fs = 8){
+
+  df_col <- colData(scWGBS_FGC_SE)
+  df_col <- df_col[order(df_col$cellType),]
+
+  column_ha_time = HeatmapAnnotation(
+    time = df_col$time,
+    col = list(time = Fetal_time_colors),
+    border = T, simple_anno_size = unit(0.4, "cm"))
+
+  column_ha_sex = HeatmapAnnotation(
+    sex = df_col$sex,
+    col = list(sex = sex_colors),
+    border = T, simple_anno_size = unit(0.4, "cm"))
+
+  column_ha_cellType = HeatmapAnnotation(
+    cellType = df_col$cellType,
+    col = list(cellType = Fetal_cellType_colors),
+    border = T, simple_anno_size = unit(0.4, "cm"))
+
+  genes <- genes[genes %in% met_means$gene]
+  mat <- as.matrix(met_means[,-1])
+  rownames(mat)<- met_means$gene
+  mat <- mat[genes, df_col$sample]
+
+  # Rm cells with no methylation values
+  # mat <- mat[rowSums(!is.na(mat)) > 0, ,drop = F]
+  # df_col <- df_col[rownames(mat),]
+
+  h <- Heatmap(mat,
+               column_title = "Mean methylation in promoters",
+               name = 'Met',
+               column_split = df_col$group,
+               row_gap = unit(c(2), "mm"),
+               row_title_gp = gpar(fontsize = 8, fontface = "bold"),
+               column_title_gp = gpar(fontsize = 8),
+               row_title_rot = 0,
+               col = colorRamp2(c(1:100),colorRampPalette(
+                 c("moccasin","dodgerblue4"))(100)),
+               na_col = "gray80",
+               clustering_method_rows = "ward.D",
+               cluster_rows = clusterRows,
+               cluster_columns = FALSE,
+               show_row_names = TRUE,
+               show_column_names = TRUE,
+               show_heatmap_legend = TRUE,
+               show_row_dend = FALSE,
+               top_annotation = c(column_ha_cellType,
+                                  column_ha_time,
+                                  column_ha_sex,
+                                  gap = unit(0.1, "mm")),
+               # right_annotation = c(row_ha_week,  row_ha_sex,
+               #                      row_ha_type,
+               #                      gap = unit(1, "mm")),
+               row_names_gp = gpar(fontsize = fs),
+               column_names_gp = gpar(fontsize = 8),
+               column_names_side = "bottom",
+              # width = unit(0.2, "npc"),
+               row_names_side = "left")
+
+  print(h) 
+}
+```
+
+```{r}
+X_met_new_order <- 
+  c("PASD1","TEX13C",  "SSX1", 
+    "VCX3B", "VCX3A", "VCX2", "FMR1NB", "MAGEB6", "MAGEC2","GAGE2A","PAGE1",
+    "MAGEA1", 
+    "CT45A10", "CT45A1","SSX5",  "SAGE1", "PPP4R3C","MAGEA12", "MAGEC1", "MAGEB16","MAGEA3",
+    "MAGEA6","GAGE1", "MAGEA9B","MAGEA9", "GAGE12E", "GAGE13", "MAGEA2", 
+    "CXorf49", "VCX", 
+    "PAGE2", 
+    "PAGE5","CTAG2","MAGEA8", 
+    "MAGEB1", 
+    "MAGEA4", 
+    "TGIF2LX", "FTHL17",
+    "MAGEB2","CT55","MAGEA10",    "MAGEA11",  
+    "XAGE5", 
+    "SPANXD",
+    "DCAF8L2", "SPANXC", 
+    "MAGEA4-AS1",  "DDX53",
+    "CENPVL1", "SPANXB1","CXorf51B", "SSX2B",
+    "XAGE1B",  "SSX2", "CTAG1A", "NXF2",  "CTAG1B", "XAGE1A",
+    "CXorf49B", "CT47A12",
+    "CT47A11","CT45A5","CT45A7", "CENPVL2","CT47A1",
+    "CT45A3",  "CT45A6",  "CT45A8",  "CT45A9" )
+
+not_X_met_new_order <- 
+  c( "LINC01518", "POTEF", "KAZN-AS1","GOLGA6L7", "TUBA3C", "COX7B2",  "RNF17", 
+     "DCAF4L2","DPPA2", "LINC02663",
+     "LINC00200", "RFPL4B", 
+     "LINC01413", "NLRP4", "SUN3",
+     "LINC02475",  
+     "BRDT", "DSCR8",  "RBM46", 
+     "POTEC","CTCFL", "PRDM9",  "LINC00470", 
+     "TMEM132D-AS1", 
+     "SLC7A11-AS1",
+     "LINC02377",    "LINC02241",   "SPATA31C2",   
+     "TEX48",  "TRIML2", "LINC02864", "GFY",
+"FLJ36000", "FAM230C", "RBMY1E", "VCY", "VCY1B", "RBMY1A1","RBMY1B"   )
+
+not_met_new_order <- 
+  c("LINC01098",  "LINC01511", "ZNF722", "LRRC52-AS1", 
+    "CSTL1", "WDR64", "CBY2", "GARIN2","LINC01193", 
+    "FAM217A",  "LINC01249", 
+    "CDRT15", "LINC01206", "SYCP1", "MEIOB", "MAJIN",
+    "PRDM7",  
+    "NFIA-AS1",  "LINC02228",  "ZNF723",
+    "LINC03004", 
+    "GTF2I-AS1", "LINC02750",  "CHCT1", "RAD21L1", "LIN28A","C1orf94", 
+    "ESX1", "DMRT1", "ZNF729", "THORLNC","SEC61G-DT", "TSPY4", "TSPY1", 
+    "TSPY9", "TSPY3", "TSPY2")
+
+my_genes <- c(X_met_new_order, not_X_met_new_order, not_met_new_order)
+```
+
+```{r}
+# slightly modified the functions to create the final heatmap
+draw_FGC_heatmap <- function(genes, 
+                             split_by = "chr_met", 
+                             ncells_max = NULL,
+                             cell_type = NULL,
+                             scale_lims = NULL,
+                             sex = c("F", "M"),
+                             clusterRow = FALSE, 
+                             font_size = 6,
+                             show_legend = TRUE,
+                             h_width = 5,
+                             h_height = 10,
+                             show_right_annot = TRUE){
+
+  fetal_sce <- FGC_sce
+
+  if (!is.null(cell_type)) fetal_sce <- fetal_sce[,FGC_sce$type %in% cell_type]
+
+  # Set the colors used in the final heatmap based on all samples (F + M), 
+  # even when representing only F or M samples on the heatmap
+  # => same scale of colors
+  mat <-  as.matrix(assay(fetal_sce, "logcounts")[genes, ])
+
+  if (is.null(scale_lims)) {
+    scale_lims <- c(0, quantile(rowMaxs(mat), 0.8))
+  }
+
+  heatmap_colors <- colorRamp2(
+    seq(scale_lims[1], scale_lims[2], length = 11),                 
+    legend_colors)
+
+  fetal_sce <- fetal_sce[, fetal_sce$sex %in% sex]
+
+  df_col_FCS <- data.frame(development_stage = fetal_sce$development_stage,
+                           sex = fetal_sce$sex,
+                           stage = fetal_sce$stage,
+                           donor_id = fetal_sce$donor_id,
+                           fetal_cell_type = fetal_sce$cell_type,
+                           type = fetal_sce$type)
+
+  rownames(df_col_FCS) <- colnames(fetal_sce)
+  df_col_FCS <- df_col_FCS[order(df_col_FCS$type),]
+
+  #Randomly pick ncells_max cells of each cell type
+  if (!is.null(ncells_max)) {
+    set.seed(123)
+    selected_cells <- sample_cells(
+      rownames(df_col_FCS[df_col_FCS$type == levels(df_col_FCS$type)[1],]), 
+      ncells_max)
+    for (type in levels(df_col_FCS$type)[-1]){
+      tmp <- sample_cells(
+        rownames(df_col_FCS[df_col_FCS$type == type,]), ncells_max)
+      selected_cells <- c(selected_cells, tmp)
+    }
+    df_col_FCS <- df_col_FCS[selected_cells,]
+  }
+
+  column_ha_type = HeatmapAnnotation(
+    Fetal_cell_type = df_col_FCS$type,
+    col = list(Fetal_cell_type = Fetal_cell_colors),   
+    border = FALSE, 
+    simple_anno_size = unit(0.15, "cm"), 
+    annotation_legend_param = legends_param,
+    annotation_name_gp = gpar(fontsize = 0),
+    show_legend = FALSE)
+
+  column_ha_development_stage = HeatmapAnnotation(
+    development_stage = df_col_FCS$development_stage,
+    border = FALSE, 
+    simple_anno_size = unit(0.15, "cm"), 
+    annotation_legend_param = legends_param,
+    annotation_name_gp = gpar(fontsize = 0))
+
+  column_ha_sex = HeatmapAnnotation(
+    sex = df_col_FCS$sex,
+    col = list(sex = sex_colors),
+    border = FALSE, 
+    simple_anno_size = unit(0.15, "cm"), 
+    annotation_legend_param = legends_param,
+    annotation_name_gp = gpar(fontsize = 0))
+
+  column_ha_donor_id = HeatmapAnnotation(
+    donor_id = df_col_FCS$donor_id,
+    border = FALSE, 
+    simple_anno_size = unit(0.15, "cm"), 
+    annotation_legend_param = legends_param,
+    annotation_name_gp = gpar(fontsize = 0))
+
+  column_ha_stage = HeatmapAnnotation( 
+    stage = df_col_FCS$stage,
+    col = list(stage = stage_colors),
+    border = FALSE, 
+    simple_anno_size = unit(0.15, "cm"), 
+    annotation_legend_param = legends_param,
+    annotation_name_gp = gpar(fontsize = 0))
+
+  gene_type <- CT_genes_types %>% 
+    filter(gene %in% genes)
+
+  if (split_by == "regulation") {
+    gene_type <- gene_type %>% arrange(Regulation)
+    rowSplit <- gene_type$Regulation
+  }
+  if (split_by == "testis_specificity") {
+    gene_type <- gene_type %>% arrange(Testis_specificity)
+    rowSplit <- gene_type$Testis_specificity
+  }
+  if (split_by == "chr_met") {
+    gene_type <- gene_type %>% arrange(chr_met)
+    rowSplit <- gene_type$chr_met
+  }
+  if (!clusterRow & split_by == "chr_met") {
+    gene_type$gene <- factor(gene_type$gene, levels = genes)
+    gene_type <- gene_type  %>%
+      arrange(gene)
+    gene_type$gene <- as.vector(gene_type$gene)
+    rowSplit <- gene_type$chr_met
+  }
+  if (split_by == "met_3_groups") {
+    gene_type <- gene_type %>% arrange(met_3_groups)
+    rowSplit <- gene_type$met_3_groups
+  }
+   if (!clusterRow & split_by == "met_3_groups") {
+    gene_type$gene <- factor(gene_type$gene, levels = genes)
+    gene_type <- gene_type  %>%
+      arrange(gene)
+    gene_type$gene <- as.vector(gene_type$gene)
+    rowSplit <- gene_type$met_3_groups
+   }
+
+  Regulation <- as.matrix(gene_type$Regulation)
+  rownames(Regulation) <- gene_type$gene
+  row_ha_regulation <- rowAnnotation(
+    Regulation = Regulation,
+    annotation_legend_param = legends_param,
+    border = FALSE,
+    simple_anno_size = unit(0.15, "cm"), 
+    col = list(Regulation = Regulation_colors),
+    annotation_name_gp = gpar(fontsize = 0))
+
+  chr_met <- as.matrix(gene_type$chr_met)
+  rownames(chr_met) <- gene_type$gene
+  row_ha_chr_met <- rowAnnotation(
+    chr_met = chr_met,
+    annotation_legend_param = legends_param,
+    border = FALSE,
+    simple_anno_size = unit(0.15, "cm"), 
+    col = list(chr_met = chr_met_colors),
+    annotation_name_gp = gpar(fontsize = 0))
+
+  Chr <- as.matrix(gene_type$chromosome)
+  rownames(Chr) <- gene_type$gene
+  row_ha_chr <- rowAnnotation(Chr = Chr,
+                              annotation_legend_param = legends_param,
+                              border = FALSE,
+                              simple_anno_size = unit(0.15, "cm"), 
+                              col = list(Chr = Chr_colors_with_Y),
+                              annotation_name_gp = gpar(fontsize = 0))
+
+  mat <- as.matrix(
+    assay(fetal_sce, "logcounts")[gene_type$gene, rownames(df_col_FCS)])
+
+  rAnnot <- c(row_ha_regulation, row_ha_chr, gap = unit(1, "mm"))
+
+  if (!show_right_annot) rAnnot <- NULL
+
+  title <- "Expression"
+
+  if (is.null(scale_lims)) {
+    scale_lims <- c(0, quantile(rowMaxs(mat), 0.8))
+  }
+
+  heatmap_colors <- colorRamp2(
+    seq(scale_lims[1], scale_lims[2], length = 11),                 
+    legend_colors)
+
+  ht_FGCs <- Heatmap(mat,
+                     name = "Logcounts",
+                     column_title = title,
+                     column_split = df_col_FCS$sex,
+                     cluster_row_slices = FALSE,
+                     cluster_column_slices = FALSE,
+                     row_title_gp = gpar(fontsize = 3),
+                     row_title_rot = 90,
+                     col = heatmap_colors,
+                     border = TRUE,
+                     border_gp = gpar(lwd = 0.2),
+                     row_gap = unit(c(0.4), "mm"),
+                     column_gap = unit(c(0.4), "mm"),
+                     show_column_dend = FALSE,
+                     show_row_dend = FALSE,
+                     row_names_gp = gpar(fontsize = font_size),
+                     row_names_side = "left",
+                     cluster_rows = clusterRow,
+                     cluster_columns = FALSE,
+                     show_column_names = FALSE,
+                     show_row_names = TRUE,
+                     row_split = rowSplit,
+                     top_annotation = c(column_ha_sex, 
+                                        #column_ha_stage,
+                                        #column_ha_group, 
+                                        #column_ha_donor_id,
+                                        column_ha_type,
+                                        gap = unit(0.1, "mm")), 
+                     right_annotation = rAnnot,
+                     column_title_gp = gpar(fontsize = 4),
+                     gap = unit(0.01, "mm"),
+                     use_raster = TRUE,
+                     raster_device = "CairoPNG",
+                     heatmap_width = unit(h_width, "cm"),
+                     heatmap_height = unit(h_height, "cm"),
+                     raster_quality = 10,
+                     show_heatmap_legend = show_legend,
+                     heatmap_legend_param = legends_param)
+
+  return(ht_FGCs)
+}
+```
+
+```{r}
+# Modify the function to keep the initial order of genes, 
+# grouping by the met_3groups column
+plot_mean_methylation_heatmap_met_3_groups <- function(genes,
+                                                       fs = 8,
+                                                       h_width = 3,
+                                                       h_height = 8){
+
+  df_col <- colData(scWGBS_FGC_SE)
+  df_col <- df_col[order(df_col$cellType),]
+
+  column_ha_time = HeatmapAnnotation(
+    time = df_col$time,
+    col = list(time = Fetal_time_colors),   
+    border = FALSE, 
+    simple_anno_size = unit(0.15, "cm"), 
+    annotation_legend_param = legends_param,
+    annotation_name_gp = gpar(fontsize = 0))
+
+  column_ha_sex = HeatmapAnnotation(
+    sex = df_col$sex,
+    col = list(sex = sex_colors),   
+   border = FALSE, 
+    simple_anno_size = unit(0.15, "cm"), 
+    annotation_legend_param = legends_param,
+    annotation_name_gp = gpar(fontsize = 0))
+
+  column_ha_cellType = HeatmapAnnotation(
+    cellType = df_col$cellType,
+    col = list(cellType = Fetal_cellType_colors),
+      border = FALSE, 
+    simple_anno_size = unit(0.15, "cm"), 
+    annotation_legend_param = legends_param,
+    annotation_name_gp = gpar(fontsize = 0))
+
+  # add NA values to missing genes
+  missing_genes <- CT_genes_types %>%
+    #filter(gene %in% genes) %>%
+    filter(!gene %in% met_means$gene)
+  mat <- as.matrix(met_means[,-1])
+  rownames(mat)<- met_means$gene
+  mat_missing_values <- matrix(NA, nrow(missing_genes), ncol(mat))
+  rownames(mat_missing_values) <- missing_genes$gene
+  colnames(mat_missing_values) <- colnames(mat)
+  mat <- rbind(mat, mat_missing_values)
+
+  #genes <- genes[genes %in% met_means$gene]
+  mat <- mat[genes, df_col$sample]
+
+  gene_type <- CT_genes_types %>% 
+    filter(gene %in% genes)
+
+  # split_by "met_3_groups"
+    gene_type$gene <- factor(gene_type$gene, levels = genes)
+    gene_type <- gene_type  %>%
+      arrange(gene)
+    gene_type$gene <- as.vector(gene_type$gene)
+    rowSplit <- gene_type$met_3_groups
+
+  mat <- mat[genes, df_col$sample]
+
+  # Rm cells with no methylation values
+  # mat <- mat[rowSums(!is.na(mat)) > 0, ,drop = F]
+  # df_col <- df_col[rownames(mat),]
+
+  h <- Heatmap(mat,
+               column_title = "Mean methylation",
+               name = 'Met',
+               column_split = df_col$sex_type,
+               row_split = rowSplit,
+               cluster_row_slices = FALSE,
+               border_gp = gpar(lwd = 0.2),
+               row_gap = unit(c(0.4), "mm"),
+               column_gap = unit(c(0.4), "mm"),
+               row_title_gp = gpar(fontsize = 5),
+               row_title_rot = 90,
+               column_title_gp = gpar(fontsize = 4),
+               col = colorRamp2(c(1:100),colorRampPalette(
+                 c("moccasin","dodgerblue4"))(100)),
+               na_col = "gray80",
+               clustering_method_rows = "ward.D",
+               cluster_rows = FALSE,
+               cluster_columns = FALSE,
+               show_row_names = TRUE,
+               show_column_names = FALSE,
+               show_heatmap_legend = TRUE,
+               show_row_dend = FALSE,
+               top_annotation = c(#column_ha_time, 
+                                  column_ha_sex,
+                                  column_ha_cellType,
+                                  gap = unit(0.1, "mm")), 
+               row_names_gp = gpar(fontsize = fs),
+               column_names_gp = gpar(fontsize = 3),
+               column_names_side = "top",
+               gap = unit(0.01, "mm"),
+               use_raster = TRUE,
+               raster_device = "CairoPNG",
+               heatmap_width = unit(h_width, "cm"),
+               heatmap_height = unit(h_height, "cm"),
+               raster_quality = 20,
+               row_names_side = "left",
+               heatmap_legend_param = legends_param)
+
+  print(h) 
+}
+```
+
+```{r}
+legends_param <- list(
+  labels_gp = gpar(col = "black", fontsize = 2),
+  title_gp = gpar(col = "black", fontsize = 3),
+  simple_anno_size = unit(0.15, "cm"),
+  row_names_gp = gpar(fontsize = 1),
+  annotation_name_side = "left",
+  border = FALSE,
+  border_gp = gpar(lwd = 0.01),
+  grid_width = unit(0.1, "cm"),
+  grid_height = unit(0.2, "cm"),
+  legend_height = unit(0.1, "cm"),
+  annotation_name_gp = gpar(fontsize = 0))
+
+FGC_ht <- draw_FGC_heatmap(my_genes, 
+                           ncells_max = 100,
+                           split_by = "met_3_groups",
+                           cell_type = c("F_GC", "F_oogonia", "F_oocyte", 
+                                         "M_GC", "M_pre_spermatogonia"),
+                           #sex = c("F"),
+                           clusterRow = FALSE, 
+                           font_size = 1,
+                           scale_lims = scale_limits, 
+                           show_legend = TRUE,
+                           h_width = 2,
+                           h_height = 6,
+                           show_right_annot = FALSE)
+
+fig <- plot_mean_methylation_heatmap_met_3_groups(my_genes, 
+                                                  h_width = 3,
+                                                  h_height = 6,
+                                                  fs  = 2)
+
+ht_list <- FGC_ht + fig
+
+draw(ht_list,  
+     merge_legend = TRUE)
+
+```
+```{r, echo = FALSE, eval = FALSE}
+pdf(file = paste0("figs/Exp_methylation_in_FGC.pdf"), width = 5, height = 3)
+draw(ht_list,
+     merge_legend = TRUE)
+dev.off()
+```
+
+## Gene by gene
+
+Change to hg19 coordinates + heatmap
+
+```{r}
+promoter_methylation_by_gene <- function(gene, 
+                                         nt_up = 1000, 
+                                         nt_down = 500,
+                                         cellType = NULL,
+                                         sex = NULL,
+                                         h_width = 5,
+                                         h_height = 5,
+                                         fs = 8,
+                                         col_fs = 4,
+                                         title = NULL){
+
+  x <- CT_promoter_regions_hg19_GR[
+    CT_promoter_regions_hg19_GR$external_gene_name == gene, ]
+  tested_gene_GR <- subsetByOverlaps(scWGBS_FGC_SE, x)
+  CpGpos <- tested_gene_GR@rowRanges@ranges@start
+  TSS <- x$TSS_liftover
+  if(as.vector(x@strand) == '+'){
+    CpGpositions <- CpGpos - x$TSS_liftover
+  }
+
+  if(as.vector(x@strand) == '-'){
+    CpGpositions <- x$TSS_liftover  - CpGpos 
+  }
+  df_col <- colData(tested_gene_GR)
+
+  if (!is.null(cellType)) df_col <- df_col[df_col$cellType %in% cellType, ]
+  if (!is.null(sex)) df_col <- df_col[df_col$sex %in% sex, ]
+
+  df_col <- df_col[order(df_col$cellType),] 
+  levels(df_col$cellType) <- gsub("^\\w_", '', levels(df_col$cellType))
+
+  y <- as_tibble(assay(tested_gene_GR)[, df_col$sample]) %>% 
+    mutate(CpG = CpGpositions) %>% dplyr::select(CpG, everything()) %>% 
+    filter(CpG > -nt_up & CpG < nt_down) %>% 
+    arrange(CpG) %>% 
+    pivot_longer(names_to = "sample", values_to = "met", -CpG) %>% 
+    pivot_wider(names_from = CpG, values_from = met)
+
+  mat <- as.matrix(y[, -1])
+  rownames(mat) <- y$sample
+
+  # Rm cells with no methylation values
+  mat <- mat[rowSums(!is.na(mat)) > 5, ,drop = F]
+  df_col <- df_col[rownames(mat),]
+
+  sex <- as.matrix(df_col$sex)
+  rownames(sex) <- df_col$sex
+  row_ha_sex <- rowAnnotation(sex = sex,
+                              na_col = "gray80",
+                              annotation_legend_param = legends_param,
+                              simple_anno_size = unit(0.2, "cm"),
+                              col = list(sex = sex_colors),
+                              annotation_name_gp = gpar(fontsize = 0))
+
+  time <- as.matrix(df_col$time)
+  rownames(time) <- df_col$time
+  row_ha_time <- rowAnnotation(time = time,
+                               na_col = "gray80",
+                               annotation_legend_param = legends_param,
+                               simple_anno_size = unit(0.2, "cm"),
+                               col = list(time = Fetal_time_colors),
+                               annotation_name_gp = gpar(fontsize = 0))
+
+  group <- as.matrix(df_col$group)
+  rownames(group) <- df_col$group
+  row_ha_group <- rowAnnotation(group = group,
+                                na_col = "gray80",
+                                annotation_legend_param = legends_param,
+                                simple_anno_size = unit(0.2, "cm"),
+                                annotation_name_gp = gpar(fontsize = 0))
+
+  cellType <- as.matrix(df_col$cellType)
+  rownames(cellType) <- df_col$cellType
+  row_ha_cellType <- rowAnnotation(cellType = cellType,
+                                   na_col = "gray80",
+                                   annotation_legend_param = legends_param,
+                                   simple_anno_size = unit(0.2, "cm"),
+                                   col = list(sex = Fetal_cellType_colors),
+                                   annotation_name_gp = gpar(fontsize = 0))
+
+   if (is.null(title)) title <- paste0('Promoter methylation of ', 
+                                       gene, "\nTSS: ", TSS, " (hg37)")
+  ht_opt$ROW_ANNO_PADDING = unit(c(0.2, 0.2), "points")
+  h <-  Heatmap(mat,
+                column_title = title,
+                name = 'Met',
+                row_split = df_col$cellType,
+                cluster_row_slices = FALSE,
+                row_gap = unit(c(2), "mm"),
+                row_title_gp = gpar(fontsize = 5, fontface = "bold"),
+                column_title_gp = gpar(fontsize = 6),
+                row_title_rot = 90,
+                col = colorRamp2(c(1:100),colorRampPalette(
+                  c("moccasin","dodgerblue4"))(100)),
+                na_col = "gray80",
+                clustering_method_rows = "ward.D",
+                cluster_rows = FALSE,
+                cluster_columns = FALSE,
+                show_row_names = TRUE,
+                show_column_names = TRUE,
+                show_heatmap_legend = TRUE,
+                show_row_dend = FALSE,
+                row_names_gp = gpar(fontsize = fs),
+                column_names_gp = gpar(fontsize = col_fs),
+                column_names_side = "bottom",
+                heatmap_legend_param = legends_param,
+                width = unit(h_width, "cm"),
+                height = unit(h_height, "cm"),
+                row_names_side = "left")
+
+  print(h) 
+}
+```
+
+### PAGE1
+
+```{r}
+gene <- "PAGE1"
+
+A <- cowplot::plot_grid(grid.grabExpr(
+  promoter_methylation_by_gene(gene = gene,
+                                   sex = "F",
+                                   h_width = 2,
+                                   h_height = 5,
+                                   fs  = 0,
+                                   col_fs = 2,
+                                   title = paste0("Female fetal germ cells\n(", gene, ')'))))
+
+B <- cowplot::plot_grid(grid.grabExpr(
+  promoter_methylation_by_gene(gene = gene,
+                                   sex = "M",
+                                   h_width = 2,
+                                   h_height = 5,
+                                   fs  = 0,
+                                   col_fs = 2,
+                                    title = paste0("Male fetal germ cells\n(", gene, ')'))))
+
+fig <- cowplot::plot_grid(A, B, 
+                   # rel_widths = c(2,1),
+                   # scale = c(1, 0.8),
+                   # label_size = 10, 
+                   nrow = 1,
+                   align = "h")
+fig
+```
+
+```{r, eval = FALSE, echo=FALSE}
+pdf(file = "figs/PAGE1_fetal_meth.pdf", height = 2.5, width = 5)
+fig
+dev.off()
+```
+
+### GAGE2A
+
+```{r}
+gene <- "GAGE2A"
+
+A <- cowplot::plot_grid(grid.grabExpr(
+  promoter_methylation_by_gene(gene = gene,
+                                   sex = "F",
+                                   h_width = 2,
+                                   h_height = 5,
+                                   nt_up = 100,
+                                   fs  = 0,
+                                   col_fs = 2,
+                                   title = paste0("Female fetal germ cells\n(", gene, ')'))))
+
+B <- cowplot::plot_grid(grid.grabExpr(
+  promoter_methylation_by_gene(gene = gene,
+                                   sex = "M",
+                                   h_width = 2,
+                                   h_height = 5,
+                               nt_up = 100,
+                                   fs  = 0,
+                                   col_fs = 2,
+                                    title = paste0("Male fetal germ cells\n(", gene, ')'))))
+
+fig <- cowplot::plot_grid(A, B, 
+                   # rel_widths = c(2,1),
+                   # scale = c(1, 0.8),
+                   # label_size = 10, 
+                   nrow = 1,
+                   align = "h")
+
+fig
+```
+```{r, eval = FALSE, echo=FALSE}
+pdf(file = "figs/GAGE2A_fetal_meth.pdf", height = 2.5, width = 5)
+fig
+dev.off()
+```
+
 # Gene function
 
-All CT genes function
-
-```{r GO_ORA}
-msigdbr(species = "Homo sapiens" , category = "C5") %>% 
-  filter(gene_symbol %in% CT_genes$external_gene_name) %>% 
-  pull(gene_symbol) %>% 
-  unique() %>% 
-  length()
-
-msigdbr(species = "Homo sapiens" , category = "H") %>% 
-  filter(gene_symbol %in% CT_genes$external_gene_name) %>% 
-  pull(gene_symbol) %>% 
-  unique() %>% 
-  length()
-
-go_ora <- enrichGO(gene = CT_genes$ensembl_gene_id,
-                   keyType = "ENSEMBL",
-                   OrgDb = org.Hs.eg.db,
-                   ont = "all",
-                   readable = TRUE)
-as_tibble(go_ora)
-
-as_tibble(go_ora) %>% 
-  arrange(desc(Count)) %>% 
-  head(12) %>% 
-  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/177,
-                           ONTOLOGY == "CC"~ Count/197,
-                           ONTOLOGY == "MF"~ Count/186)) %>% 
-  ggplot(aes(x = Ratio, y = Description, fill = Description)) +
-  geom_col() +
-  theme_bw() +
-  ylab("GO term") +
-  xlab("Gene Ratio") +
-  theme(axis.text.y = element_blank(),
-        legend.position = "none",
-        axis.ticks.y = element_blank(),
-        axis.title = element_text(size = 10, color = "gray10"))+ 
-  geom_text(aes(0, y = Description, label = Description),
-            hjust = 0,
-            nudge_x = 0.005,
-            colour = "floralwhite",
-            size = 4)
-
-ora_to_plot <- as_tibble(simplify(go_ora))
-
-ora_to_plot <- ora_to_plot %>% 
-  arrange(desc(Count)) %>% 
-  head(9) %>% 
-  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/102,
-                           ONTOLOGY == "MF"~ Count/186))
-
-ora_to_plot %>% 
-  ggplot(aes(x = Ratio, y = Description, fill = Description)) +
-  geom_col() +
-  theme_bw() +
-  ylab("GO term") +
-  xlab("Gene Ratio") +
-  theme(axis.text.y = element_blank(),
-        legend.position = "none",
-        axis.ticks.y = element_blank(),
-        axis.title = element_text(size = 10, color = "gray10"))+ 
-  geom_text(aes(0, y = Description, label = Description),
-            hjust = 0,
-            nudge_x = 0.005,
-            colour = "floralwhite",
-            size = 3.7)
-```
-
-As we can see here, most of genes are indeed linked to functions from reproduction.
-I represented here the 12 categories with the most genes, all from biological 
-processes. However, they are enriched in `r nrow(as_tibble(go_ora))` different GO terms.
-
-Is there a difference between meth reg or not ?
-
-```{r GO_ORA_meth}
-go_ora_meth <- enrichGO(gene = 
-                          filter(CT_genes, regulated_by_methylation)$ensembl_gene_id,
-                        keyType = "ENSEMBL",
-                        OrgDb = org.Hs.eg.db,
-                        ont = "all",
-                        readable = TRUE)
-go_ora_meth <- simplify(go_ora_meth)
-as_tibble(go_ora_meth)
-
-go_ora_not_meth <- enrichGO(gene = 
-                          filter(CT_genes, !regulated_by_methylation)$ensembl_gene_id,
-                        keyType = "ENSEMBL",
-                        OrgDb = org.Hs.eg.db,
-                        ont = "all",
-                        readable = TRUE)
-go_ora_not_meth <- simplify(go_ora_not_meth)
-as_tibble(go_ora_not_meth)
-
-go_ora_meth_plot <- as_tibble(go_ora_meth) %>% 
-  arrange(desc(Count)) %>% 
-  head(10) %>% 
-  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/73,
-                           ONTOLOGY == "MF"~ Count/80,
-                           ONTOLOGY == "CC"~ Count/78)) %>%  
-  mutate(regulation = "Methylation")
-
-go_ora_not_meth_plot <- as_tibble(go_ora_not_meth) %>% 
-  arrange(desc(Count)) %>% 
-  head(8) %>% 
-  mutate(Ratio = case_when(ONTOLOGY == "BP"~ Count/29,
-                           ONTOLOGY == "MF"~ Count/30))%>%  
-  mutate(regulation = "Not methylation")
-
-rbind(go_ora_meth_plot, go_ora_not_meth_plot) %>% 
-  ggplot(aes(x = Ratio, y = Description, fill = Description)) +
-  geom_col() +
-  theme_bw() +
-  ylab("GO term") +
-  xlab("Gene Ratio") +
-  facet_wrap(~ regulation) + 
-  theme(legend.position = "none",
-        axis.ticks.y = element_blank(),
-        axis.text.y = element_text(size = 6, color = "gray10"),
-        axis.title.x = element_text(size = 10, color = "gray10"),
-        axis.title.y = element_blank())+ 
-  geom_text(aes(0, y = Description, label = ID),
-            hjust = 0,
-            nudge_x = 0.005,
-            colour = "floralwhite",
-            size = 2)
-
-```
-
-We've also added a column tumor suppressor and oncogene in CTexploreR. 
-These information come from Cancermine.
-
-```{r C_TS}
-table(CT_genes$oncogene) 
-table(CT_genes$tumor_suppressor)
-
-table(CT_genes$oncogene, CT_genes$tumor_suppressor) 
+```{r, warning=FALSE}
+gmt_file <- msigdbr(species = "Homo sapiens", category = "C5", 
+                    subcategory = "BP") %>%
+  dplyr::select(gs_name, entrez_gene) %>%
+  dplyr::rename(ont = gs_name, gene = entrez_gene)
+
+padj_thr <- 0.05
+maxGSSize <- 200
+
+all_genes <- all_genes %>% 
+  mutate(ENTREZID = mapIds(org.Hs.eg.db, external_gene_name , "ENTREZID", 
+                           "SYMBOL") %>% unname()) %>% 
+  filter(!is.na(ENTREZID))
+
+CT_met_X <- all_genes %>% 
+  filter(CT_gene_type == "CT_gene") %>% 
+  filter(regulated_by_methylation, X_linked)
+
+egmt <- enricher(gene          = CT_met_X$ENTREZID,
+                 universe      = all_genes$ENTREZID,
+                 pAdjustMethod = "BH",
+                 minGSSize = 10,
+                 maxGSSize = maxGSSize,
+                 pvalueCutoff  = 1,
+                 qvalueCutoff = 1,
+                 TERM2GENE = gmt_file)
+
+egmt_tbl <- as_tibble(egmt) %>% filter(p.adjust < 0.05)
+
+IDs <- egmt_tbl[, which(colnames(egmt_tbl) == "geneID")]
+tested_ID <- strsplit(IDs$geneID, "/")
+egmt_tbl$gene_names <- sapply(tested_ID, function(ids){
+  paste(mapIds(org.Hs.eg.db, ids, "SYMBOL", "ENTREZID"), collapse = '/')})
+
+egmt_tbl_CT_met_X <- egmt_tbl
+
+CT_not_Xmet <- all_genes %>% 
+  filter(CT_gene_type == "CT_gene") %>% 
+  filter(!external_gene_name %in% CT_met_X$external_gene_name)
+
+egmt <- enricher(gene          = CT_not_Xmet$ENTREZID,
+                 universe      = all_genes$ENTREZID,
+                 pAdjustMethod = "BH",
+                 minGSSize = 10,
+                 maxGSSize = maxGSSize,
+                 pvalueCutoff  = 1,
+                 qvalueCutoff = 1,
+                 TERM2GENE     = gmt_file)
+
+egmt_tbl <- as_tibble(egmt) %>% filter(p.adjust < 0.05)
+
+IDs <- egmt_tbl[, which(colnames(egmt_tbl) == "geneID")]
+tested_ID <- strsplit(IDs$geneID, "/")
+egmt_tbl$gene_names <- sapply(tested_ID, function(ids){
+  paste(mapIds(org.Hs.eg.db, ids, "SYMBOL", "ENTREZID"), collapse = '/')})
+
+egmt_tbl_CT_not_Xmet <- egmt_tbl
+
+GO_2_groups <- rbind(as_tibble(egmt_tbl_CT_met_X) %>% 
+                       filter(p.adjust < padj_thr) %>% 
+                       dplyr::select(Description, Count, p.adjust) %>% 
+                       mutate(Group = "Methylation\nX-linked"),
+                     as_tibble(egmt_tbl_CT_not_Xmet) %>% 
+                       filter(p.adjust < padj_thr) %>% 
+                       dplyr::select(Description, Count, p.adjust) %>% 
+                       mutate(Group = "Other\nCT genes")) %>% 
+  dplyr::select(-Count) %>% 
+  pivot_wider(names_from = Group, values_from = p.adjust) %>% 
+  pivot_longer(names_to = "Group", values_to = "padjusted", -Description) %>% 
+  left_join(rbind(as_tibble(egmt_tbl_CT_met_X) %>% 
+                    filter(p.adjust < padj_thr) %>% 
+                    dplyr::select(Description, Count) %>% 
+                    mutate(Group = "Methylation\nX-linked"),
+                  as_tibble(egmt_tbl_CT_not_Xmet) %>% 
+                    filter(p.adjust < padj_thr) %>% 
+                    dplyr::select(Description, Count) %>% 
+                    mutate(Group = "Other\nCT genes"))) %>% 
+  mutate(group = factor(Group, levels = c("Methylation\nX-linked", 
+                                          "Other\nCT genes"))) %>% 
+  mutate(Description = gsub("GOBP_", x = Description, replace = ''))
+
+sorted_GO <- GO_2_groups %>% 
+  arrange(group) %>%  
+  pull(Description) %>% unique()
+
+GO_2_groups <- GO_2_groups %>% 
+  arrange(group) %>% 
+  filter(Count >= 1)
+
+GO_2_groups$Description <- 
+  factor(GO_2_groups$Description, levels = rev(unique(sorted_GO)))
+
+# redundant GO terms (or unrelevant)
+redundant <- c("NCRNA_3_END_PROCESSING",
+               "RNA_3_END_PROCESSING",
+               "SNRNA_METABOLIC_PROCESS",
+               "MALE_MEIOTIC_NUCLEAR_DIVISION",
+               "DNA_METHYLATION",
+               "DNA_METHYLATION_OR_DEMETHYLATION",
+               "REGULATION_OF_HISTONE_H3_K4_METHYLATION",
+               "REGULATION_OF_BIOLOGICAL_PROCESS_INVOLVED_IN_SYMBIOTIC_INTERACTION",
+               "NEGATIVE_REGULATION_OF_MRNA_CATABOLIC_PROCESS",
+               "NEGATIVE_REGULATION_OF_RNA_CATABOLIC_PROCESS",
+               "REGULATION_OF_HISTONE_METHYLATION",
+               "NEGATIVE_REGULATION_OF_MRNA_METABOLIC_PROCESS",
+               "HOMOLOGOUS_RECOMBINATION",
+               "CHROMOSOME_ORGANIZATION_INVOLVED_IN_MEIOTIC_CELL_CYCLE",
+               "HOMOLOGOUS_CHROMOSOME_SEGREGATION",
+               "POSITIVE_REGULATION_OF_MEIOTIC_NUCLEAR_DIVISION",
+               "REGULATION_OF_MEIOTIC_NUCLEAR_DIVISION",
+               "POSITIVE_REGULATION_OF_NUCLEAR_DIVISION",
+               "CELL_CELL_RECOGNITION", "POSITIVE_REGULATION_OF_REPRODUCTIVE_PROCESS",
+               "REGULATION_OF_MEIOTIC_CELL_CYCLE",
+               "REGULATION_OF_VIRAL_ENTRY_INTO_HOST_CELL",
+               "MALE_MEIOSIS_I",
+               "POSITIVE_REGULATION_OF_MEIOTIC_CELL_CYCLE",
+               "HISTONE_H3_K4_TRIMETHYLATION")
+
+GO_2_groups_not_redundant <- GO_2_groups %>% 
+  arrange(group) %>% 
+  filter(Count > 1) %>% 
+  filter(!Description %in% redundant)
+
+GO_2_groups_not_redundant$Description <- 
+  factor(GO_2_groups_not_redundant$Description, 
+         levels = rev(unique(GO_2_groups_not_redundant$Description)))
+
+GO_terms_fig <- GO_2_groups_not_redundant %>% 
+  arrange(Description) %>% 
+  ggplot(aes(x = group, y = Description, color = padjusted, size = Count)) +
+  geom_point() +
+  scale_color_continuous(low = "red", high = "blue", 
+                         guide = guide_colorbar(reverse = TRUE))+
+  #  na.value = "white") +
+  scale_size_continuous(limits = c(1, max(GO_2_groups_not_redundant$Count, na.rm = TRUE)), range = c(0, 3))+
+  theme(axis.text.y = element_text(size = 5),
+        axis.text.x = element_text(size = 6, face = "bold", vjust = 0.5),
+        axis.ticks = element_line(linewidth = 0.1),
+        legend.key.size = unit(4, "mm"),
+        aspect.ratio = 1,
+        legend.text = element_text(size = 7),
+        legend.title = element_text(size = 7)) +
+  xlab('') +
+  ylab('')
+GO_terms_fig
+```
+
+```{r, eval = FALSE, echo=FALSE}
+pdf(file = paste0("figs/GO_terms_2_groups_rm_redundancy.pdf"), height = 3)
+GO_terms_fig
+dev.off()
 ```

### 2024-10-03 Paper_code-Database_Comparison.Rmd
Update sizes

@@
 
-upset_text_size <- c(1.5, 1.5, 1.5, 1.5, 2, 2)
+upset_text_size <- c(2, 2, 2, 2, 3.5, 3.5)
  #c(intersection size title, intersection size tick labels, set size title, 
@@
 gp[["SetText"]][["Set1"]]$col <-  "darkseagreen4"
+gp[["FaceText"]][["11"]]$fontsize <- 30
+gp[["FaceText"]][["10"]]$fontsize <- 30
+gp[["FaceText"]][["01"]]$fontsize <- 30
+gp[["SetText"]][["Set2"]]$fontsize <-  30
+gp[["SetText"]][["Set1"]]$fontsize <- 30
 plot(CTdatabase_ours, gp = gp)
@@
 upset(upset_CTdatabase,
-      text.scale = upset_text_size)
+      text.scale = upset_text_size, 
+      mb.ratio = c(0.6, 0.4))
 ```
@@
 ```{r, eval = TRUE, echo = FALSE}
-pdf(file = paste0("figs/CTdatabase_comparison.pdf"))
+pdf(file = paste0("figs/CTdatabase_comparison.pdf"),
+    paper = "a4r")
 plot(CTdatabase_ours, gp = gp)
 upset(upset_CTdatabase,
-      text.scale = upset_text_size)
+      text.scale = upset_text_size,
+      mb.ratio = c(0.6, 0.4))
 dev.off()
@@
 
-upset_text_size_all <- c(1.5, 1.5, 1.5, 1.5, 2, 1)
+upset_text_size_all <- c(2, 2, 2, 2, 3, 2)
  #c(intersection size title, intersection size tick labels, set size title, 
@@
       nintersects = 50,
-      text.scale = upset_text_size_all)
+      text.scale = upset_text_size_all,
+      mb.ratio = c(0.6, 0.4))
 

### 2024-10-04 Paper_code-Database_Comparison.Rmd
Changes in size figure

@@
 
-```{r, eval = TRUE, echo = FALSE}
-pdf(file = paste0("figs/CTdatabase_comparison.pdf"),
-    paper = "a4r")
-plot(CTdatabase_ours, gp = gp)
+```{r, eval = FALSE, echo = FALSE}
+gp_no_label <- gp
+gp_no_label[["SetText"]][["Set2"]]$fontsize <-  0
+gp_no_label[["SetText"]][["Set1"]]$fontsize <- 0
+gp_no_label[["FaceText"]][["11"]]$fontsize <- 40
+gp_no_label[["FaceText"]][["10"]]$fontsize <- 40
+gp_no_label[["FaceText"]][["01"]]$fontsize <- 40
+pdf(file = paste0("figs/CTdatabase_comparison_venn.pdf"),
+    height = 9, width = 5)
+plot(CTdatabase_ours, gp = gp_no_label)
+dev.off()
+
+upset_text_size_save <- c(3, 2, 3, 2, 4, 4)
+ #c(intersection size title, intersection size tick labels, set size title, 
+ # set size tick labels, set names, numbers above bars)
+
+pdf(file = paste0("figs/CTdatabase_comparison_upset.pdf"),
+    height = 7, width = 8)
 upset(upset_CTdatabase,
-      text.scale = upset_text_size,
+      text.scale = upset_text_size_save,
       mb.ratio = c(0.6, 0.4))
@@
 
-```{r, eval = TRUE, echo = FALSE}
+```{r, eval = FALSE, echo = FALSE}
 pdf(file = paste0("figs/Jamin_comparison.pdf"))
@@
 
-```{r, eval = TRUE, echo = FALSE}
+```{r, eval = FALSE, echo = FALSE}
 pdf(file = paste0("figs/Wang_comparison.pdf"))
@@
 
-```{r, eval = TRUE, echo = FALSE}
+```{r, eval = FALSE, echo = FALSE}
 pdf(file = paste0("figs/Carter_comparison.pdf"))
@@
 
-```{r, eval = TRUE, echo = FALSE}
+```{r, eval = FALSE, echo = FALSE}
 pdf(file = paste0("figs/Bruggeman_comparison.pdf"))
@@
 
-```{r, eval = TRUE, echo = FALSE}
+```{r, eval = FALSE, echo = FALSE}
 pdf(file = paste0("figs/upset_all_lost.pdf"))
@@
 
-pdf(file = paste0("figs/upset_all_databases.pdf"))
+upset_text_size_all_save <- c(3, 2, 3, 2, 4, 0)
+ #c(intersection size title, intersection size tick labels, set size title, 
+ # set size tick labels, set names, numbers above bars)
+
+pdf(file = paste0("figs/upset_all_databases.pdf"),
+    width = 15, height = 7)
 upset(upset_all_database,
@@
       nintersects = 50,
-      text.scale = upset_text_size_all)
+      text.scale = upset_text_size_all_save,
+      mb.ratio = c(0.55, 0.45))
 dev.off()

### 2024-10-18 Enrichment_unduplicated_family.Rmd
Added enrichment analysis removing duplicates in family

---
title: "Is there still an enrichment of CT genes on the X chromosome if we take duplicated genes in family into account ? "
author: "Julie Devis"
date: "2024-10-18"
output: 
  html_document:
    theme: cosmo
    df_print : paged
    highlight : default
    toc: yes
    number_sections : TRUE
    toc_depth : 4
---

```{r, message = FALSE}
.libPaths("/storage/research/dduv/cbio-lg/rlib/Bioc320", include.site = FALSE)
library(CTexploreR)
library(tidyverse)
```

We've shown that there is an enrichment of CT genes on the X chromosome, and that 
this enrichment was only seen for genes that are MethDep. However, we wonder if
this would still be seen if we take into account the fact that many CT genes are
duplication of each other as they're from the same families.

```{r}
CT_genes <- CT_genes %>% 
  filter(CT_gene_type == "CT_gene")
table(CT_genes$family)
sum(table(CT_genes$family))
table(CT_genes$family, CT_genes$chr)
table(CT_genes$family, CT_genes$regulated_by_methylation)

```

There are 10 families containing 58 genes.

All members of a family are on the same chromosome (only X and Y) and the same 
regulated by methylation status. I can thus only keep one member of each family
and it will be representative of it all. It'll thus remove 46 genes from my 
list (58 - 10 which is 1 of each family kept)

```{r}
CT_no_duplicated_fam <- CT_genes %>% 
  filter(!duplicated(family) | is.na(family)) %>% 
  mutate(Regulation = case_when(regulated_by_methylation ~ "Methylation",
                                !regulated_by_methylation ~ "Not methylation")) %>%
  mutate(Chr = case_when(X_linked ~ "X-linked",
                         !X_linked ~ "not X")) 
dim(CT_no_duplicated_fam)

CT_no_duplicated_fam$chr <- factor(CT_no_duplicated_fam$chr, levels = c(1:22, "X", "Y"))

fig_X_enrichment <- ggplot(CT_no_duplicated_fam %>% 
                             filter(!is.na(regulated_by_methylation)), 
                           aes(x = chr, fill = Regulation)) +
  geom_bar(stat = 'count') +
  xlab("Chromosome") +
  ylab("Gene number") +
  scale_fill_manual(values = c("Methylation" = "indianred2",
                               "Not methylation"= "cyan4"))+
  theme_bw() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.2, 0.85),
        #aspect.ratio = 1.5,
        panel.grid = element_blank())

fig_X_enrichment

table(CT_no_duplicated_fam$chr)

CT_no_duplicated_fam_rep <- CT_no_duplicated_fam %>% 
  filter(!is.na(regulated_by_methylation)) %>% 
  group_by(Regulation, Chr) %>%
  dplyr::count() %>% 
  pivot_wider(names_from = Regulation, values_from = n)

CT_no_duplicated_fam_rep

chisq.test(CT_no_duplicated_fam_rep[,-1])

```

We can still see that there is more genes on the X chromosome than on others.
However, in comparison with the analysis not removing duplicated families, 
the MethDep repartition between X and not X is different, as there is now more 
genes that are not X (37 out of the initial 39) than those that are X-linked 
(27 out of the 69). However, pvalue is still significative.

Could also be done by autosomes/X

```{r}
CT_genes_X_linked <- CT_no_duplicated_fam %>% 
  mutate(chr = case_when(chr == "X" ~ "X-linked",
                           chr == "Y" ~ "Y-linked",
                           !chr %in% c("X", "Y") ~ "Autosome")) %>% 
  filter(!is.na(regulated_by_methylation)) %>% 
  group_by(Regulation, chr) %>%
  dplyr::count() %>% 
  pivot_wider(names_from = Regulation, values_from = n)
CT_genes_X_linked
```

What does it give if we normalise according to the number of gene per chromosome ?

```{r}
n_genes_per_chr <- all_genes %>% 
  filter(chr != "MT") %>% 
  group_by(chr) %>% 
  summarize(n = n()) %>% 
  mutate(chr = factor(chr, levels = c(1:22, "X", "Y"))) %>% 
  arrange(chr)

n_genes <- n_genes_per_chr %>% 
  filter(!chr %in% "MT") %>% 
  pull(n) %>% 
  sum()
n_genes_autosome <- n_genes_per_chr %>% 
  filter(chr %in% 1:22) %>% 
  pull(n) %>% 
  sum()
n_genes_X <- n_genes_per_chr %>% 
  filter(chr %in% "X") %>% 
  pull(n) 
n_genes_Y <- n_genes_per_chr %>% 
  filter(chr %in% "Y") %>% 
  pull(n)

n_genes_per_chr_autosome_X_Y <- tibble("chr" = c("Autosome", "X-linked", "Y-linked"), 
       n = c(n_genes_autosome, n_genes_X, n_genes_Y))

ggplot(n_genes_per_chr, aes(x = chr, y = n)) +
  geom_col() +
  xlab("Chromosome") +
  ylab("Gene number") +
  scale_fill_manual(values = c("Methylation" = "indianred2",
                               "Not methylation"= "cyan4")) +
  theme_bw() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.2, 0.85),
        #aspect.ratio = 1.5,
        panel.grid = element_blank()) +
  ggtitle("Total number of genes per chromosome")

n_genes_per_chr_autosome_X_Y <- n_genes_per_chr_autosome_X_Y %>% 
  mutate(proba = n / n_genes)
n_genes_per_chr_autosome_X_Y

CT_genes_Met <- colSums(CT_genes_X_linked[, -1])["Methylation"]
CT_genes_not_Met <- colSums(CT_genes_X_linked[, -1])["Not methylation"]
CT_genes_X_Y_observed_expected <- CT_genes_X_linked %>% 
  left_join(n_genes_per_chr_autosome_X_Y %>% 
              dplyr::select(-n)) %>% 
  mutate(Expected_MET_if_random = (CT_genes_Met * proba),
         Expected_NOT_MET_if_random = (CT_genes_not_Met * proba)) %>% 
  dplyr::select(chr, Methylation, Expected_MET_if_random, `Not methylation`, Expected_NOT_MET_if_random)
CT_genes_X_Y_observed_expected

chisq.test(CT_genes_X_Y_observed_expected[
  CT_genes_X_Y_observed_expected$chr %in% c("Autosome", "X-linked"), c("Methylation", "Expected_MET_if_random")]) 
```

CT genes regulated by met are enriched on X when taking into account the number 
of genes per chromosome, verified statistically.

### 2025-03-25 1.CT_selection_and_methylation.Rmd
Added code used to generate the paper

---
title: "CT genes analysis"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: false
---

```{r, echo = FALSE, Results = 'asis'}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
library("tidyverse")
library("CTexploreR")
library("ComplexHeatmap")
library("circlize")
library("SummarizedExperiment")
library("flextable")

set_flextable_defaults(
font.family = "Arial", font.size = 10,
border.color = "black",
padding = 6,
background.color = "#EFEFEF",
big.mark = "")
```

#

```{r}
legends_param <- list(
  labels_gp = gpar(col = "black", fontsize = 8),
  title_gp = gpar(col = "black", fontsize = 8, fontface = "bold"),
  simple_anno_size = unit(0.2, "cm"),
  row_names_gp = gpar(fontsize = 5),
  annotation_name_side = "left",
  border = FALSE,
  border_gp = gpar(lwd = 0.2),
  grid_width = unit(0.2, "cm"),
  grid_height = unit(0.05, "cm"),
  legend_height = unit(1, "cm"),
  use_raster = TRUE,
               raster_device = "CairoPNG",
               raster_quality = 10)

legend_colors <- c("#5E4FA2", "#3288BD", "#66C2A5", "#ABDDA4", "#E6F598",
"#FFFFBF", "#FEE08B", "#FDAE61", "#F46D43", "#D53E4F",
"#9E0142")

TCGA_colors <- c(
"BRCA" = "midnightblue", "COAD" = "darkorchid2",
"ESCA" = "gold", "HNSC" = "deeppink2",
"LUAD" = "seagreen", "LUSC" = "seagreen3",
"SKCM" = "red3")

CCLE_colors <- c(
"Lung" = "seagreen3", "Skin" = "red3",
"Bile_Duct" = "mediumpurple1", "Bladder" = "mistyrose2",
"Colorectal" = "plum", "Lymphoma" = "steelblue1",
"Uterine" = "darkorange4", "Myeloma" = "turquoise3",
"Kidney" = "thistle4",
"Pancreatic" = "darkmagenta", "Brain" = "palegreen2",
"Gastric" = "wheat3", "Breast" = "midnightblue",
"Bone" = "sienna1", "Head_and_Neck" = "deeppink2",
"Ovarian" = "tan3", "Sarcoma" = "lightcoral",
"Leukemia" = "steelblue4", "Esophageal" = "khaki",
"Neuroblastoma" = "olivedrab1")

DAC_colors <- c(
"B2-1" = "olivedrab2", "HCT116" = "lightcoral",
"HEK293T" = "seagreen3", "HMLER" = "mediumpurple1",
"IMR5-75" = "deeppink2", "NCH1681" = "steelblue2",
"NCH612" = "red3", "TS603" = "darkmagenta")

DAC_treatment_colors <- c("CTL" = "royalblue4", "5-aza" = "maroon3")

Regulation_colors <- c("Methylation" = "indianred2",
"Not methylation"= "cyan4")

MethDep_colors <- c("MethDep" = "indianred2",
"Non-MethDep"= "cyan4")

MethDep_full_name_colors <- c("MethDep CG gene" = "indianred2",
"Non-MethDep CG gene"= "cyan4")

Testis_specificity_colors = c("testis_specific" = "mediumpurple2",
"testis_preferential" = "goldenrod1",
"Undetectable in GTEX" = "gray")

```

# CT selection

## CT genes and CTP genes

```{r}
GTEX_data <- CTdata::GTEX_data()

#Function to print GTEX expression with specific annotations
print_GTEX_heatmap <- function(genes = NULL, units = "log_TPM", 
                               row_fontsize = 2,
                               column_fontsize = 6, 
                               h_width = 3,
                               h_height = 10,
                               clust_method = "ward.D",
                               clustRow = FALSE) {

  suppressMessages({
    database <- GTEX_data
  })

  database <- database[rowData(database)$external_gene_name %in% genes, ]

  mat <- assay(database)
  rownames(mat) <- rowData(database)$external_gene_name

  if (units == "log_TPM") mat <- log1p(mat)

  gene_type <- CT_genes %>% 
    filter(external_gene_name %in% genes) %>% 
    mutate(gene = factor(external_gene_name, levels = genes)) %>% 
    arrange(gene)

   rowSplit <- gene_type$testis_specificity
   rowSplit[rowSplit == "testis_preferential"] <- "CG genes testis-preferential"
   rowSplit[rowSplit == "testis_specific"] <- "CG genes testis-specific"
   rowSplit <- factor(rowSplit, levels = c("CG genes testis-specific", "CG genes testis-preferential"))

  heatmap_colors <- colorRamp2(
    seq(0, 5, length = 11),                 
    legend_colors)

  h <- Heatmap(mat[as.vector(gene_type$gene), ],
               name = "log(TPM +1)",
               column_title = "Normal tissues (GTEx)",
               column_title_gp = gpar(fontsize = 8, face = "bold"),
               row_split = rowSplit,
               cluster_row_slices = FALSE,
               row_title_gp = gpar(fontsize = 8, face = "bold"),
               border = TRUE,
               border_gp = gpar(lwd = 0.5),
               col = heatmap_colors,
               cluster_rows = clustRow,
               cluster_columns = TRUE,
               show_row_dend = FALSE,
               show_column_dend = FALSE,
               row_names_gp = gpar(fontsize = row_fontsize),
               column_names_gp = gpar(fontsize = column_fontsize),
               column_names_side = "top",
               row_names_side = "left",
               width = unit(h_width, "cm"),
               height = unit(h_height, "cm"),
               clustering_method_rows = clust_method,
               use_raster = TRUE,
               raster_device = "CairoPNG",
               raster_quality = 10,
               heatmap_legend_param = legends_param)

  return(h)
}
```

```{r}
TCGA_TPM <- CTdata::TCGA_TPM()

# Function to print TCGA expression with specific annotations
# and to downsample the number of tumor samples per tumor type 
print_TCGA_heatmap <- function(genes = NULL,
                               tum_type = c("TCGA-SKCM", "TCGA-LUAD", 
                                            "TCGA-LUSC", "TCGA-COAD", 
                                            "TCGA-BRCA", "TCGA-HNSC",
                                            "TCGA-ESCA"),
                               downsample_to_n = 450,
                               fontsize = 4, 
                               clust_method = "ward.D",
                               h_width = 5, 
                               h_height = 10,
                               show_annotation_legend = TRUE,
                               show_top_annotations = TRUE,
                               show_row_names = FALSE,
                               clustRow = FALSE,
                               clustCol = TRUE) {

  tcga <- TCGA_TPM
  tcga <- tcga[,tcga$shortLetterCode != "NT"]

  ## downsample tumors to have the same number of tumors in each type
  TCGA <- tcga[, tcga$project_id == tum_type[1]]
  TCGA <- TCGA[, 1:min(ncol(TCGA), downsample_to_n)]

  for (tum in tum_type[-1]){
    tmp <- tcga[, tcga$project_id == tum]
    tmp <- tmp[, 1:min(ncol(tmp), downsample_to_n)]
    TCGA <- cbind(TCGA, tmp)
  }

  TCGA$project_id <- gsub(pattern = "TCGA-", x = TCGA$project_id, replacement = '')
  split_by <- factor(TCGA$project_id)

  column_ha_tumor <- HeatmapAnnotation(
    TCGA_tumor = TCGA$project_id,
    border = TRUE,
    col = list(TCGA_tumor = TCGA_colors),
    annotation_label = gpar(fontsize = 4),
    annotation_name_gp = gpar(fontsize = 0),
    simple_anno_size = unit(0.3, "cm"),
    annotation_legend_param = legends_param)

  ## Use gene names instead of ENSEMBL IDs
  mat <- SummarizedExperiment::assay(TCGA)
  rownames(mat) <- rowData(TCGA)$external_gene_name
  mat <- log1p(mat)

  gene_type <- CT_genes %>% 
    filter(external_gene_name %in% genes) %>% 
    mutate(gene = factor(external_gene_name, levels = genes)) %>% 
    arrange(gene)

   rowSplit <- gene_type$testis_specificity
   rowSplit[rowSplit == "testis_preferential"] <- "CG genes testis-preferential"
   rowSplit[rowSplit == "testis_specific"] <- "CG genes testis-specific"
   rowSplit <- factor(rowSplit, levels = c("CG genes testis-specific", "CG genes testis-preferential"))

  # same range as in GTEX
  heatmap_colors <- colorRamp2(
    seq(0, 5, length = 11),                 
    legend_colors)

  tAnnot <- NULL
  if (show_top_annotations) tAnnot <- column_ha_tumor

  Heatmap(mat[as.vector(gene_type$gene), ],
          name = "TCGA (logTPM)",
          column_title = "TCGA tumor samples",
          column_title_gp = gpar(fontsize = 8, face = "bold"),
          clustering_method_rows = clust_method,
          clustering_method_columns = clust_method,
          column_split = split_by,
          row_gap = unit(c(1), "mm"),
          col = heatmap_colors,
          cluster_rows = clustRow,
          cluster_columns = clustCol,
          show_row_names = show_row_names,
          show_column_names = FALSE,
          show_row_dend = FALSE,
          show_column_dend = FALSE,
          column_names_gp = gpar(fontsize = 6),
          column_names_centered = TRUE,
          row_names_gp = gpar(fontsize = fontsize),
          row_names_side = "left",
          border = TRUE,
          border_gp = gpar(lwd = 0.5),
          column_names_side = c("top"),
          column_names_rot = 90,
          row_title_gp = gpar(fontsize = 0),
          top_annotation = tAnnot,
          row_split = rowSplit,
          cluster_row_slices = FALSE,
          width = unit(h_width, "cm"),
          height = unit(h_height, "cm"),
          use_raster = TRUE,
          raster_device = "CairoPNG",
          raster_quality = 10,
          heatmap_legend_param = legends_param,
          show_heatmap_legend = FALSE)
}
```

```{r}
CCLE_data <- CTdata::CCLE_data()

# Function to print CCLE expression with specific annotations
# Allows to downsample the number of tumor samples pet tumor type to show
print_CCLE_heatmap <- function(genes = NULL,
                               Cellosaurus_NCIt_disease = 
                                 c("Melanoma", "Lung adenocarcinoma"),
                               h_width = 5,
                               h_height = 10,
                               downsample_to_n = 450,
                               fontsize = 4, 
                               clust_method = "ward.D",
                               split_rows_by = "regulation", 
                               show_annotation_legend = TRUE,
                               clustRow = FALSE,
                               clustCol = TRUE){

  ## downsample tumors to have the same number of tumors in each type
  ccle <- CCLE_data
  CCLE <- ccle[, !is.na(ccle$Cellosaurus_NCIt_disease) & 
                 ccle$Cellosaurus_NCIt_disease == Cellosaurus_NCIt_disease[1]]
  CCLE <- CCLE[, 1:min(ncol(CCLE), downsample_to_n)]

  for (tum in Cellosaurus_NCIt_disease[-1]){
    tmp <- ccle[, !is.na(ccle$Cellosaurus_NCIt_disease) & 
                 ccle$Cellosaurus_NCIt_disease == tum]
    tmp <- tmp[, 1:min(ncol(tmp), downsample_to_n)]
    CCLE <- cbind(CCLE, tmp)
  }

  split_by <- factor(CCLE$Cellosaurus_NCIt_disease)

  column_ha_tumor <- HeatmapAnnotation(
    CCLE_cell_type = CCLE$Cellosaurus_NCIt_disease,
    border = TRUE,
    col = list(CCLE_cell_type = c(
      "Lung adenocarcinoma" = "seagreen", "Melanoma" = "red3")),
      annotation_label = gpar(fontsize = 4),
      annotation_name_gp = gpar(fontsize = 0),
      simple_anno_size = unit(0.3, "cm"),
      annotation_legend_param = legends_param)

  ## Use gene names instead of ENSEMBL IDs
  mat <- SummarizedExperiment::assay(CCLE)
  rownames(mat) <- rowData(CCLE)$external_gene_name
  mat <- log1p(mat)

   gene_type <- CT_genes %>% 
    filter(external_gene_name %in% genes) %>% 
    mutate(gene = factor(external_gene_name, levels = genes)) %>% 
    arrange(gene)

   rowSplit <- gene_type$testis_specificity
   rowSplit[rowSplit == "testis_preferential"] <- "CG genes testis-preferential"
   rowSplit[rowSplit == "testis_specific"] <- "CG genes testis-specific"
   rowSplit <- factor(rowSplit, levels = c("CG genes testis-specific", "CG genes testis-preferential"))

    # Same range as GTEX and TCGA
  heatmap_colors <- colorRamp2(
    seq(0, 5, length = 11),                 
    legend_colors)

  Heatmap(mat[as.vector(gene_type$gene), ],
          name = "CCLE (logTPM)",
          column_title = paste0("CCLE tumor cell lines"),
          column_title_gp = gpar(fontsize = 8, face = "bold"),
          clustering_method_rows = clust_method,
          clustering_method_columns = clust_method,
          column_split = split_by,
          row_gap = unit(c(1), "mm"),
          col = heatmap_colors,
          cluster_rows = clustRow,
          cluster_columns = clustCol,
          show_row_names = TRUE,
          show_column_names = FALSE,
          show_row_dend = FALSE,
          show_column_dend = FALSE,
          column_names_gp = gpar(fontsize = 6),
          column_names_centered = TRUE,
          row_names_gp = gpar(fontsize = fontsize),
          row_names_side = "left",
          border = TRUE,
          border_gp = gpar(lwd = 0.5),
          column_names_side = c("top"),
          column_names_rot = 90,
          row_title_gp = gpar(fontsize = 0),
          top_annotation = column_ha_tumor,
          #right_annotation = rAnnot,
          row_split = rowSplit,
          width = unit(h_width, "cm"),
          height = unit(h_width, "cm"),
          use_raster = TRUE,
          raster_device = "CairoPNG",
          raster_quality = 10,
          heatmap_legend_param = legends_param,
          show_heatmap_legend = FALSE)
}

```

Top panel = Testis specific genes

Bottom panel = Testis preferential genes

```{r}
my_genes <- CT_genes %>% 
  pull(external_gene_name)

# Run the heatmap with custRow parameters to define the correct order
# This order will then be used for GTEX and CCLE heatmaps
ht_TCGA <- print_TCGA_heatmap(my_genes, 
                              fontsize = 0, 
                              downsample_to_n = 50, 
                              h_width = 3,
                              h_height = 8,
                              tum_type = c("TCGA-SKCM", "TCGA-LUAD"),
                                           #"TCGA-ESCA", "TCGA-LUSC", 
                                           #"TCGA-COAD",
                                           #"TCGA-BRCA",
                                           #"TCGA-HNSC"),
                              clust_method = "ward.D",
                              clustRow = TRUE)

indices <- row_order(ht_TCGA)
gene_order <- CT_genes %>%
  filter(external_gene_name %in% my_genes) %>%
  mutate(testis_specificity = factor(testis_specificity, levels = c("testis_specific", "testis_preferential"))) %>% 
  arrange(testis_specificity) %>%
  pull(external_gene_name)

TS_order <- gene_order[indices$`CG genes testis-specific`]
TP_order <- gene_order[indices$`CG genes testis-preferential`]
my_genes <- c(TS_order, TP_order)

ht_TCGA <- print_TCGA_heatmap(my_genes, 
                              fontsize = 4, 
                              downsample_to_n = 50, 
                              h_width = 3,
                              h_height = 8,
                              tum_type = c("TCGA-SKCM", "TCGA-LUAD"), 
                                           #"TCGA-ESCA", "TCGA-LUSC", 
                                           #"TCGA-COAD",
                                           #"TCGA-BRCA",
                                           #"TCGA-HNSC"),
                              clust_method = "ward.D",
                              show_annotation_legend = FALSE,
                              clustRow = FALSE)

ht_CCLE <- print_CCLE_heatmap(my_genes, 
                              downsample_to_n = 50, 
                               Cellosaurus_NCIt_disease = 
                                c("Melanoma", "Lung adenocarcinoma"),
                              clustRow = FALSE,
                              show_annotation_legend = FALSE,
                              h_width = 3,
                              h_height = 8,
                              split_rows_by = "testis_specificity")

ht_GTEX <- print_GTEX_heatmap(my_genes, 
                              row_fontsize = 0,
                              column_fontsize = 7,
                              h_width = 8,
                              h_height = 8,
                              clustRow = FALSE)

h_list <- ht_GTEX + ht_CCLE + ht_TCGA
fig <- draw(h_list, main_heatmap = "log(TPM +1)", ht_gap = unit(c(0.3, 0.03), "cm"),
     merge_legend = TRUE)
```

```{r}
pdf(file = paste0("../figs/CT_genes_selection.pdf"), width = 8, height = 5)
fig
dev.off()
```

## Figure Multimapping

```{r}
normal_tissues_multimapping_data <- CTdata::normal_tissues_multimapping_data()

my_genes <- CT_genes %>% 
  filter(GTEX_category == 'lowly_expressed') 
my_genes$testis_specificity <- factor(my_genes$testis_specificity, levels = c("testis_specific", "testis_preferential"))
my_genes <- my_genes %>% arrange(testis_specificity)
rowSplit <- my_genes$testis_specificity

testis_specificity <- as.matrix(my_genes$testis_specificity)
rownames(testis_specificity) <- my_genes$gene
row_ha_TS <- rowAnnotation(testis_specificity = testis_specificity,
                           annotation_legend_param = legends_param,
                           simple_anno_size = unit(0.2, "cm"),
                           col = list(testis_specificity = Testis_specificity_colors),
                           annotation_name_gp = gpar(fontsize = 0))

without_MP <- log1p(
  assay(normal_tissues_multimapping_data[my_genes$ensembl_gene_id,], 
    "TPM_no_multimapping"))
with_MP <-  log1p(
  assay(normal_tissues_multimapping_data[my_genes$ensembl_gene_id,], 
    "TPM_with_multimapping"))
rownames(without_MP) <- rownames(with_MP) <- 
  rowData(normal_tissues_multimapping_data)[my_genes$ensembl_gene_id,]$external_gene_name
# Fix same range of colors for both heatmaps
full_mat <- cbind(without_MP, with_MP)
heatmap_colors <- colorRamp2(
  seq(0, quantile(rowMax(full_mat), 0.8), length = 11), legend_colors)

# Or same range of colors than GTEX heatmap
# heatmap_colors <- colorRamp2(
#   seq(0, 6, length = 11), legend_colors)

colnames(without_MP) <- gsub("_", x = str_to_sentence(colnames(without_MP)), ' ')
colnames(with_MP) <- gsub("_", x = str_to_sentence(colnames(with_MP)), ' ')
ordered_tissues <- 
  c("Testis", colnames(without_MP)[- which(colnames(without_MP) == "Testis")])
h_without_MP <- Heatmap(without_MP[my_genes$external_gene_name, ordered_tissues],
                        name = "no MP",
                        column_title = "Multimapped reads excluded",
                        column_title_gp = gpar(fontsize = 10, face = "bold"),
                        row_title_gp = gpar(fontsize = 0, face = "bold", angle  = 90),
                        row_split = rowSplit,
                        border = TRUE,
                        border_gp = gpar(lwd = 0.5),
                        col = heatmap_colors,
                        cluster_rows = FALSE,
                        cluster_columns = FALSE,
                        show_row_dend = FALSE,
                        show_column_dend = FALSE,
                        row_names_gp = gpar(fontsize = 4),
                        column_names_gp = gpar(fontsize = 6),
                        column_names_side = "top",
                        row_names_side = "left",
                        heatmap_width = unit(6, "cm"),
                        heatmap_height = unit(12, "cm"),
                        clustering_method_rows = "ward.D",
                        use_raster = TRUE,
                        raster_device = "CairoPNG",
                        raster_quality = 10,
                        show_heatmap_legend = FALSE,
                        heatmap_legend_param = legends_param)

h_with_MP <- Heatmap(with_MP[my_genes$external_gene_name, ordered_tissues],
                     name = "log(TPM)",
                     column_title = "Multimapped reads counted",
                     column_title_gp = gpar(fontsize = 10, face = "bold"),
                     row_title_gp = gpar(fontsize = 0, face = "bold"),
                     row_split = rowSplit,
                     border = TRUE,
                     border_gp = gpar(lwd = 0.5),
                     col = heatmap_colors,
                     cluster_rows = TRUE,
                     cluster_columns = FALSE,
                     show_row_dend = FALSE,
                     show_column_dend = FALSE,
                     row_names_gp = gpar(fontsize = 4),
                     column_names_gp = gpar(fontsize = 6),
                     column_names_side = "top",
                     row_names_side = "left",
                     heatmap_width = unit(6, "cm"),
                     heatmap_height = unit(12, "cm"),
                     clustering_method_rows = "ward.D",
                     right_annotation = c(row_ha_TS, gap = unit(1, "mm")),
                     use_raster = TRUE,
                     raster_device = "CairoPNG",
                     raster_quality = 10,
                     heatmap_legend_param = legends_param)

h_list <- h_without_MP + h_with_MP
draw(h_list, main_heatmap = "no MP", ht_gap = unit(0.3, "cm"),
                 height = unit(5, "cm"))
```

```{r}
pdf(file = paste0("../figs/FigS1_multimapping.pdf"))
draw(h_list, main_heatmap = "no MP", ht_gap = unit(0.3, "cm"),
                 height = unit(5, "cm"))
dev.off()
```

# Methylation analysis

```{r}
# Work only on testis_specific genes!
CT_genes <- CT_genes %>% filter(testis_specificity == "testis_specific")
```

## Fig3B

```{r}
mean_methylation_in_tissues <- CTdata::mean_methylation_in_tissues()

# Remove genes without enough methylation values for the figure
met_value_number_per_gene <- tibble(
  gene = CT_genes$external_gene_name,
  regulated_by_methylation = CT_genes$regulated_by_methylation,
  NA_met_values_number = rowSums(
    is.na(assay(mean_methylation_in_tissues[CT_genes$external_gene_name, ]))))
my_genes <- met_value_number_per_gene %>% 
  arrange(desc(NA_met_values_number)) %>% 
  filter(NA_met_values_number < 5) %>% 
  pull(gene)

tissue_order <- c("adipose", "colon", "esophagus", "heart", "intestine",  "lung", 
                  "muscle", "pancreas", "skin", "stomach", 
                  "thyroid", "testis", "placenta", "sperm" )

database <- mean_methylation_in_tissues
rowData(database)$external_gene_name <- rownames(database)
database <- database[my_genes, ]
colnames(database) = str_to_sentence(colnames(database))
x <- as_tibble(assay(database), rownames = "gene") %>% 
  pivot_longer(names_to = "tissue", values_to = "met", -gene) %>% 
  left_join(CT_genes %>% dplyr::rename(gene = "external_gene_name")) %>% 
  filter(CT_gene_type == "CT_gene") %>% 
  filter(!is.na(regulated_by_methylation)) %>% 
  mutate(tissue = factor(tissue, levels = str_to_sentence(tissue_order))) %>% 
  mutate(group = case_when(regulated_by_methylation  ~"MethDep CG gene", 
                           !regulated_by_methylation ~ "Non-MethDep CG gene"))

library(ggh4x)
strip <- strip_themed(background_x = elem_list_rect(fill = MethDep_colors))

figA <- x %>% 
  ggplot(aes(x = tissue, y = met, color = group)) +
  geom_jitter(size = 0.5, width = 0.11) +
  scale_color_manual(values = MethDep_full_name_colors) +
  #geom_boxplot(alpha = 0) +
  #geom_violin(alpha = 0) +
  #facet_wrap(~ group, ncol = 2) +
  facet_grid2( ~ group ,  strip = strip) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 18),
        axis.text.y = element_text(size = 16),
        aspect.ratio = 0.8,
        axis.title.y = element_text(size = 18),
        strip.text = element_text(size = 18, color = "white"),
        legend.position = "none") +
  ylab("Promoter mean methylation") +
  xlab("")

figA

pdf(file = paste0("../figs/Fig3B_prom_methylation_in_tissues.pdf"))
figA
dev.off()

```

## Proportion of CT genes regulated by methylation

NB: One CT gene could not be classified as regulated or not by methylation

=> proportion based on 145/146 CT genes

```{r}
x <- enframe(table(CT_genes %>% 
  pull(regulated_by_methylation)), name = "Regulation") %>% 
  left_join(enframe(prop.table(table(CT_genes$regulated_by_methylation)), 
                    name = "Regulation", value = "Percent")) %>% 
  mutate(group = case_when(Regulation == TRUE ~ "MethDep", 
                           Regulation == FALSE ~ "Non\nMethDep"))

my_colors <- MethDep_colors
names(my_colors) <- c("MethDep","Non\nMethDep")
met_prop_fig <- ggplot(x, aes(x = '', y = value, fill = group)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start = 0) +
  geom_text(aes(y = value/5.5 + c(0, cumsum(value)[-length(value)]), 
                label = rev(paste0(group, "\n", 
                               round(Percent * 100), "%"))), 
            size = 8, 
            color = "white",
            face = "bold") +
            #, size = 6
  scale_fill_manual(values = my_colors) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.title = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid = element_blank(),
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        legend.position = "none")

met_prop_fig

pdf(file = paste0("../figs/Fig3A_proportion_of_CT_methylated.pdf"))
met_prop_fig
dev.off()
```

## Fig 3C: DAC heatmap

```{r}
DAC_treated_cells_multimapping <- CTdata::DAC_treated_cells_multimapping()
DAC_treated_cells <- CTdata::DAC_treated_cells()

DAC_fig <- function(genes = gene_type$gene, 
                    multimapping = TRUE, 
                    row_fontsize = 2,
                    units = "log1p",
                    values_only = FALSE, 
                    scale_lims = NULL) {

  if (multimapping) {
    database <- CTdata::DAC_treated_cells_multimapping()
    #database <- DAC_treated_cells_multimapping

  } else {
    database <- CTdata::DAC_treated_cells()
  }

  database$treatment[database$treatment == "DAC"] <- "5-aza"
  database$treatment <- factor(database$treatment, levels = c("CTL", "5-aza"))
  gene_type <- CT_genes %>% 
    filter(external_gene_name %in% genes)

  # sort genes regulated by met differently than not regulated by met
  gene_type_met <- gene_type %>%
    filter(regulated_by_methylation)

  gene_type_not_met <- gene_type %>%
    filter(!regulated_by_methylation) %>%
    arrange(germline_methylation, somatic_met_level)

  gene_type <- rbind(gene_type_met, gene_type_not_met)
  rownames(database) <- rowData(database)$external_gene_name

  gene_type <- gene_type %>% 
    mutate(regulation = case_when(regulated_by_methylation == TRUE ~ "Methylation",
                                  regulated_by_methylation == FALSE ~ "Not Methylation")) %>% 
    mutate(regulation = factor(regulation, 
                                  levels = c("Methylation", "Not Methylation")))

  mat <- assay(database, units)

  df_col <- data.frame("cell" = colData(database)$cell,
                       "treatment" = colData(database)$treatment)
  rownames(df_col) <- rownames(colData(database))
  df_col <- df_col[order(df_col$cell, df_col$treatment), ]

  column_ha_cell <- HeatmapAnnotation(
    cell = df_col$cell,
    border = FALSE, 
    #border_gp = gpar(lwd = 0.1),
    simple_anno_size = unit(0.2, "cm"),
    annotation_legend_param = legends_param,
    col = list("cell" = DAC_colors),
    annotation_name_gp = gpar(fontsize = 0))

  column_ha_treatment <- HeatmapAnnotation(
    treatment = df_col$treatment,
    col = list(treatment = c("CTL" = "dodgerblue3", "5-aza" = "firebrick1")),
    annotation_legend_param = legends_param,
    border = FALSE, 
    simple_anno_size = unit(0.2, "cm"),
    annotation_name_gp = gpar(fontsize = 0))

  Regulation <- as.matrix(gene_type$regulated_by_methylation)
  rownames(Regulation) <- gene_type$gene
  row_ha_regulation <- rowAnnotation(Regulation = Regulation,
                                     annotation_legend_param = legends_param,
                                     simple_anno_size = unit(0.2, "cm"),
                                     col = list(Regulation = Regulation_colors),
                                     annotation_name_gp = gpar(fontsize = 0))

  if (is.null(scale_lims)) scale_lims <- c(0, quantile(rowMaxs(mat), 0.75))

  h <- Heatmap(mat[gene_type$external_gene_name, rownames(df_col), drop = FALSE],
               name = "logCounts",
               row_split = gene_type$regulation,
               left_annotation = rowAnnotation(
                       regulation = anno_block(
                         gp = gpar(fill = c(my_colors[1], my_colors[2]),
                                   col = "black", lwd = 0.1),
                         width = unit(c(0.5), "cm"),
                         labels = c("MethDep",  
                                    "Non\nMethDep"), 
                         labels_gp = gpar(col = "white", fontsize = 5, face = "bold"))),
               row_title_gp = gpar(col = my_colors, fontsize = 0),
               border = TRUE,
               border_gp = gpar(lwd = 0.2),
               column_title = "DAC induction in cell lines",
               column_title_gp = gpar(fontsize = 6),
               column_split = factor(df_col$cell),
               cluster_row_slices = FALSE,
               col = colorRamp2(seq(scale_lims[1], scale_lims[2], length = 11),
                                legend_colors),
               cluster_rows = TRUE,
               show_row_dend = FALSE,
               show_row_names = FALSE,
               clustering_method_rows = "ward.D",
               show_column_names = FALSE,
               cluster_columns = FALSE,
               row_names_gp = gpar(fontsize = row_fontsize),
               top_annotation = c(column_ha_cell, column_ha_treatment),
               heatmap_width = unit(4, "cm"),
               heatmap_height = unit(4, "cm"),
               use_raster = TRUE,
               raster_device = "CairoPNG",
               raster_quality = 10,
               heatmap_legend_param = legends_param)

  return(h)
}
```

```{r}
legends_param <- list(
  labels_gp = gpar(col = "black", fontsize = 4),
  title_gp = gpar(col = "black", fontsize = 4, fontface = "bold"),
  simple_anno_size = unit(0.2, "cm"),
  row_names_gp = gpar(fontsize = 5),
  annotation_name_side = "left",
  border = FALSE,
  border_gp = gpar(lwd = 0.2),
  grid_width = unit(0.2, "cm"),
  grid_height = unit(0.05, "cm"),
  legend_height = unit(1, "cm"),
  use_raster = TRUE,
               raster_device = "CairoPNG",
               raster_quality = 10)

fs <- 1.8
my_genes <- CT_genes %>% 
  filter(!is.na(DAC_induced))

h_DAC <- DAC_fig(my_genes$external_gene_name, 
                 multimapping = TRUE, 
                 units = "log1p",
                 row_fontsize = fs)

h_DAC             
```

```{r}
pdf(file = paste0("../figs/Fig3C_methylation_classification.pdf"))
h_DAC
dev.off()
```

# CT genes regulated by met are enriched on X

```{r}
CT_genes$chr <- factor(CT_genes$chr, levels = c(1:22, "X", "Y"))

fig_X_enrichment <- ggplot(CT_genes %>% 
                             filter(!is.na(regulated_by_methylation)) %>% 
  mutate(group = case_when(regulated_by_methylation == TRUE ~"MethDep", 
                           regulated_by_methylation == FALSE ~ "Non-MethDep")), 
                           aes(x = chr, fill = group)) +
  geom_bar(stat = 'count') +
  xlab("Chromosome") +
  ylab("Gene number") +
  scale_fill_manual(values = MethDep_colors)+
  theme_bw() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.2, 0.85),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        #aspect.ratio = .5,
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 12),
        panel.grid = element_blank())

fig_X_enrichment
```

```{r}
pdf(file = paste0("../figs/Fig4A_enrichment_on_X.pdf"), height = 4, width = 6)
fig_X_enrichment
dev.off()
```

```{r}
CT_genes_X_linked <- CT_genes %>% 
  mutate(Regulation = case_when(regulated_by_methylation == TRUE ~ "Methylation",
                                regulated_by_methylation == FALSE ~ "Not Methylation")) %>% 
  mutate(Chr = case_when(X_linked ~ "X-linked",
                         !X_linked ~ "not X")) %>% 
  filter(!is.na(regulated_by_methylation)) %>% 
  group_by(Regulation, Chr) %>%
  dplyr::count() %>% 
  pivot_wider(names_from = Regulation, values_from = n)

flextable(CT_genes_X_linked, 
                     cwidth = 1.5,
                     cheight = 0.25) %>%
  set_caption(
    as_paragraph(
    as_chunk("CT genes", props = fp_text_default(font.family = "Arial",
                                                 font.size = 10))
    ), word_stylename = "Table Caption",
    align_with_table = TRUE
  )

chisq.test(CT_genes_X_linked[,-1], correct = FALSE) 
```

idem but just comparing CT-X-linked to CT-autosomal

```{r}

CT_genes_X_linked <- CT_genes %>% 
  mutate(Regulation = case_when(regulated_by_methylation == TRUE ~ "Methylation",
                                regulated_by_methylation == FALSE ~ "Not Methylation")) %>% 
  mutate(Chr = case_when(chr == "X" ~ "X-linked",
                           chr == "Y" ~ "Y-linked",
                           !chr %in% c("X", "Y") ~ "Autosome")) %>% 
  filter(!is.na(regulated_by_methylation)) %>% 
  group_by(Regulation, Chr) %>%
  dplyr::count() %>% 
  pivot_wider(names_from = Regulation, values_from = n)

flextable(CT_genes_X_linked, 
                     cwidth = 1.5,
                     cheight = 0.25) %>%
  set_caption(
    as_paragraph(
    as_chunk("CT genes", props = fp_text_default(font.family = "Arial",
                                                 font.size = 10))
    ), word_stylename = "Table Caption",
    align_with_table = TRUE
  )

chisq.test(CT_genes_X_linked[1:2, c("Methylation", "Not Methylation")], correct = FALSE) 
```

### Idem but normalised by number of gene per chromosome

```{r}
n_genes_per_chr <- all_genes %>% filter(chr != "MT") %>% 
  group_by(chr) %>% 
  summarize(n = n()) %>% 
  mutate(Chr = factor(chr, levels = c(1:22, "X", "Y"))) %>% 
  arrange(Chr)

n_genes <- n_genes_per_chr %>% 
  filter(!chr %in% "MT") %>% 
  pull(n) %>% sum()
n_genes_autosome <- n_genes_per_chr %>% 
  filter(chr %in% 1:22) %>% 
  pull(n) %>% sum()
n_genes_X <- n_genes_per_chr %>% 
  filter(chr %in% "X") %>% 
  pull(n) 
n_genes_Y <- n_genes_per_chr %>% 
  filter(chr %in% "Y") %>% 
  pull(n)

n_genes_per_chr_autosome_X_Y <- tibble("Chr" = c("Autosome", "X-linked", "Y-linked"), 
       n = c(n_genes_autosome, n_genes_X, n_genes_Y))

ggplot(n_genes_per_chr, aes(x = chr, y = n)) +
  geom_col() +
  xlab("Chromosome") +
  ylab("Gene number") +
  scale_fill_manual(values = Regulation_colors) +
  theme_bw() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.2, 0.85),
        #aspect.ratio = 1.5,
        panel.grid = element_blank()) +
  ggtitle("Total number of genes per chromosome")

n_genes_per_chr_autosome_X_Y <- n_genes_per_chr_autosome_X_Y %>% 
  mutate(proba = n / n_genes)
DT::datatable(n_genes_per_chr_autosome_X_Y)

CT_genes_Met <- colSums(CT_genes_X_linked[, -1])["Methylation"]
CT_genes_not_Met <- colSums(CT_genes_X_linked[, -1])["Not Methylation"]

CT_genes_X_Y_observed_expected <- CT_genes_X_linked %>% 
  left_join(n_genes_per_chr_autosome_X_Y %>% dplyr::select(-n)) %>% 
  mutate(Expected_MET_if_random = (CT_genes_Met * proba),
         Expected_NOT_MET_if_random = (CT_genes_not_Met * proba)) %>% 
  dplyr::select(Chr, Methylation, Expected_MET_if_random, `Not Methylation`, Expected_NOT_MET_if_random)

flextable(CT_genes_X_Y_observed_expected, 
                     cwidth = 1.5,
                     cheight = 0.25) %>%
  set_caption(
    as_paragraph(
    as_chunk("CT genes", props = fp_text_default(font.family = "Arial",
                                                 font.size = 10))
    ), word_stylename = "Table Caption",
    align_with_table = TRUE
  )
```

**=> CT genes regulated by met are enriched on X when taking into account the number of genes per chromosome:**

```{r}
chisq.test(CT_genes_X_Y_observed_expected[
  CT_genes_X_Y_observed_expected$Chr %in% c("Autosome", "X-linked"), c("Methylation", "Expected_MET_if_random")], correct = FALSE) 
```

pvalue = `r chisq.test(CT_genes_X_Y_observed_expected[CT_genes_X_Y_observed_expected$Chr %in% c("Autosome", "X-linked"), c("Methylation", "Expected_MET_if_random")], correct = FALSE)$p.value`

```{r}
fisher.test(CT_genes_X_Y_observed_expected[
  CT_genes_X_Y_observed_expected$Chr %in% c("Autosome", "X-linked"), c("Methylation", "Expected_MET_if_random")]) 
```

pvalue = `r fisher.test(CT_genes_X_Y_observed_expected[CT_genes_X_Y_observed_expected$Chr %in% c("Autosome", "X-linked"), c("Methylation", "Expected_MET_if_random")])$p.value`

**=> But CT genes NOT regulated by met are NOT enriched on X when taking into account the number of genes per chromosome:**

```{r}
chisq.test(CT_genes_X_Y_observed_expected[
  CT_genes_X_Y_observed_expected$Chr %in% c("Autosome", "X-linked"), c("Not Methylation", "Expected_NOT_MET_if_random")], correct = FALSE) 
```

```{r}
fisher.test(CT_genes_X_Y_observed_expected[
  CT_genes_X_Y_observed_expected$Chr %in% c("Autosome", "X-linked"), c("Not Methylation", "Expected_NOT_MET_if_random")])
```

## CT genes are enriched among Testis-specific genes regulated by methylation

```{r}
all_genes_chr <- all_genes %>% 
  mutate(chr = case_when(chr == "X" ~ "X-linked",
                           chr == "Y" ~ "Y-linked",
                           !chr %in% c("X", "Y") ~ "Autosome"))

TS_genes <- all_genes_chr %>% 
  filter(testis_specificity == "testis_specific")
```

`r nrow(TS_genes %>% filter(!is.na(regulated_by_methylation)))` Testis-specific genes 
(removing the 17 genes from the `r nrow(TS_genes)` genes that could not be 
classified as regulated or not by methylation)

```{r}
prop_CT_among_TS <- TS_genes %>% 
  filter(!is.na(regulated_by_methylation)) %>% 
  mutate(Regulation = case_when(regulated_by_methylation ~ "Methylation",
                                !regulated_by_methylation ~ "Not Methylation")) %>% 
 # mutate(Regulation = factor(Regulation, levels = c("Not Methylation", "Methylation"))) %>% 
  mutate(`Testis-specific genes` = case_when(CT_gene_type == "CT_gene" ~ "CT gene",
                             CT_gene_type != "CT_gene" ~ "other")) %>% 
  group_by(Regulation, `Testis-specific genes`) %>% 
  dplyr::count() %>% 
  pivot_wider(names_from = "Regulation", values_from = "n")

flextable(prop_CT_among_TS, 
                     cwidth = 1.5,
                     cheight = 0.25)

chisq.test(prop_CT_among_TS[,-1], correct = FALSE) 
```

**pvalue = `r chisq.test(prop_CT_among_TS[,-1], correct = FALSE)$p.value`**

## Testis-specific genes (including CT genes) regulated by methylation are enriched on X

```{r}
TS_genes_X <- TS_genes %>% 
  mutate(Regulation = case_when(regulated_by_methylation ~ "Methylation",
                                !regulated_by_methylation ~ "Not methylation")) %>% 
  filter(!is.na(Regulation)) %>% 
  #filter(chr != "Y-linked") %>% 
  group_by(chr, Regulation) %>% 
  dplyr::count() %>% 
  pivot_wider(names_from = Regulation, values_from = "n")

flextable(TS_genes_X, 
                     cwidth = 1.5,
                     cheight = 0.25) %>%
  set_caption(
    as_paragraph(
    as_chunk("Testis-specific genes", props = fp_text_default(font.family = "Arial",
                                                 font.size = 10))
    ), word_stylename = "Table Caption",
    align_with_table = TRUE
  )

TS_genes_X_Y <- TS_genes %>% 
  mutate(Regulation = case_when(regulated_by_methylation ~ "Methylation",
                                !regulated_by_methylation ~ "Not methylation")) %>% 
  filter(!is.na(Regulation)) %>% 
  #filter(chr != "Y-linked") %>% 
  group_by(chr, Regulation) %>% 
  dplyr::count() %>% 
  pivot_wider(names_from = Regulation, values_from = "n")

chisq.test(TS_genes_X_Y[TS_genes_X_Y$chr %in% c("Autosome", "X-linked"), c("Methylation", "Not methylation")], correct = FALSE)  
```

pvalue = `r chisq.test(TS_genes_X[TS_genes_X_Y$chr %in% c("Autosome", "X-linked"),c("Methylation", "Not methylation")], correct = FALSE)$p.value`

**Idem but Normalised by number of gene per chromosome:**

```{r}
TS_genes_Met <- colSums(TS_genes_X_Y[, -1])["Methylation"]
TS_genes_not_Met <- colSums(TS_genes_X_Y[, -1])["Not methylation"]

TS_genes_X_Y_observed_expected <- TS_genes_X_Y %>% 
  left_join(n_genes_per_chr_autosome_X_Y %>% dplyr::select(-n) %>% dplyr::rename(chr = Chr)) %>% 
  mutate(Expected_MET_if_random = (TS_genes_Met * proba),
         Expected_NOT_MET_if_random = (TS_genes_not_Met * proba)) %>% 
  dplyr::select(chr, Methylation, Expected_MET_if_random, `Not methylation`, Expected_NOT_MET_if_random)

flextable(TS_genes_X_Y_observed_expected, 
                     cwidth = 1.5,
                     cheight = 0.25) %>%
  set_caption(
    as_paragraph(
    as_chunk("Testis-specific genes", props = fp_text_default(font.family = "Arial",
                                                 font.size = 10))
    ), word_stylename = "Table Caption",
    align_with_table = TRUE
  )
```

**=> Testis-specific genes regulated by methylation are significantly enriched on X when taking into account the number of genes per chromosome:**

```{r}
chisq.test(TS_genes_X_Y_observed_expected[
  TS_genes_X_Y_observed_expected$chr %in% c("Autosome", "X-linked"), 
  c("Methylation", "Expected_MET_if_random")], correct = FALSE) 
```

pvalue = `r chisq.test(TS_genes_X_Y_observed_expected[TS_genes_X_Y_observed_expected$chr %in% c("Autosome", "X-linked"), c("Methylation", "Expected_MET_if_random")], correct = FALSE)$p.value`)

**=> But testis-specific genes NOT regulated by methylation are not enriched on X when taking into account the number of genes per chromosome:**

```{r}
chisq.test(TS_genes_X_Y_observed_expected[
  TS_genes_X_Y_observed_expected$chr %in% c("Autosome", "X-linked"), 
  c("Not methylation", "Expected_NOT_MET_if_random")], correct = FALSE) 
```

pvalue = `r chisq.test(TS_genes_X_Y_observed_expected[TS_genes_X_Y_observed_expected$chr %in% c("Autosome", "X-linked"), c("Not methylation", "Expected_NOT_MET_if_random")], correct = FALSE)$p.value`)

## Testis-specific genes (other than CT genes) regulated by methylation are also enriched on X

```{r}
TS_other_X <- TS_genes %>% 
  mutate(Regulation = case_when(regulated_by_methylation ~ "Methylation",
                                !regulated_by_methylation ~ "Not methylation")) %>% 
 filter(!is.na(Regulation)) %>% 
  filter(CT_gene_type == "other") %>% 
  #filter(chr != "Y-linked") %>% 
  group_by(chr, Regulation) %>% 
  dplyr::count() %>% 
  pivot_wider(names_from = Regulation, values_from = "n")

TS_genes_other_Met <- colSums(TS_other_X[, -1])["Methylation"]
TS_genes_other_not_Met <- colSums(TS_other_X[, -1])["Not methylation"]

TS_other_X <- TS_other_X %>% 
  left_join(n_genes_per_chr_autosome_X_Y %>% dplyr::select(-n) %>% dplyr::rename(chr = Chr)) %>% 
  mutate(Expected_MET_if_random = (TS_genes_other_Met * proba),
         Expected_NOT_MET_if_random = (TS_genes_other_not_Met * proba)) %>% 
  dplyr::select(chr, Methylation, Expected_MET_if_random, `Not methylation`, Expected_NOT_MET_if_random)

flextable(TS_other_X, 
                     cwidth = 1.5,
                     cheight = 0.25) %>%
  set_caption(
    as_paragraph(
    as_chunk("Testis-specific genes (other than CT genes)", props = fp_text_default(font.family = "Arial",
                                                 font.size = 10))
    ), word_stylename = "Table Caption",
    align_with_table = TRUE
  )
```

**=> Testis-specific genes (other than CT genes) regulated by methylation are also enriched on X:**

```{r}
chisq.test(TS_other_X[TS_other_X$chr %in% c("Autosome", "X-linked"),
                      c("Methylation", "Expected_MET_if_random")], correct = FALSE)  
```

pvalue = `r chisq.test(TS_other_X[TS_other_X$chr %in% c("Autosome", "X-linked"), c("Methylation", "Expected_MET_if_random")], correct = FALSE)$p.value`

```{r}
fisher.test(TS_other_X[TS_other_X$chr %in% c("Autosome", "X-linked"),
                      c("Methylation", "Expected_MET_if_random")])  
```

pvalue = `r fisher.test(TS_other_X[TS_other_X$chr %in% c("Autosome", "X-linked"), c("Methylation", "Expected_MET_if_random")])$p.value`

**=> Testis-specific genes (other than CT genes) NOT regulated by methylation are NOT enriched on X:**

```{r}
chisq.test(TS_other_X[TS_other_X$chr %in% c("Autosome", "X-linked"),
                      c("Not methylation", "Expected_NOT_MET_if_random")], correct = FALSE)   
```

# session info

```{r}
sessionInfo()
```

### 2025-03-25 3.CT_in_tumors.Rmd
Added code used to generate the paper

---
title: "CT genes in tumors"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: false
---

```{r, echo = FALSE, Results = 'asis'}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
library("tidyverse")
library("CTexploreR")
library("ComplexHeatmap")
library("circlize")
library("SummarizedExperiment")
```

#

```{r}
load("~/cluster/Packages/CTdata/eh_data/all_genes.rda")

CT_genes_all <- all_genes

CT_genes_types_all <- all_genes %>% 
  dplyr::rename(gene = external_gene_name) %>% 
  mutate(Regulation = case_when(regulated_by_methylation ~ "Methylation",
                                !regulated_by_methylation ~ "Not methylation")) %>%
  mutate(Chr = case_when(X_linked ~ "X-linked",
                         !X_linked ~ "not X")) %>% 
  mutate(chr_met = case_when(regulated_by_methylation & X_linked ~ "X_met",
                             !regulated_by_methylation & X_linked ~ "X_not_met",
                             regulated_by_methylation & !X_linked ~ "not_X_met",
                             !regulated_by_methylation & !X_linked ~ "not_X_not_met")) %>% 
  mutate(GTEX_specificity = case_when(GTEX_category == "testis_specific" ~ "Testis specific",
                                        GTEX_category == "testis_preferential" ~ "Testis preferential",
                                        GTEX_category == "lowly_expressed" ~ "Undetectable in GTEX")) %>% 
  mutate(GTEX_specificity = factor(GTEX_specificity, 
                                     levels = c("Testis specific",  "Undetectable in GTEX", "Testis preferential"))) %>% 
  mutate(testis_specificity = case_when(testis_specificity == "testis_specific" ~ "Testis specific",
                                        testis_specificity == "testis_preferential" ~ "Testis preferential")) %>%
  mutate(testis_specificity = factor(testis_specificity, levels = c("Testis specific", "Testis preferential"))) %>% 
  mutate(chr_met = factor(chr_met, 
                          levels = c("X_met", "not_X_met", "X_not_met", "not_X_not_met"))) %>% 
  mutate(met_3_groups = case_when(regulated_by_methylation & X_linked ~ "Methylation X-linked",
                                   regulated_by_methylation & !X_linked ~ "Methylation not X-linked",
                             !regulated_by_methylation ~ "Not Methylation")) %>% 
  mutate(met_3_groups = factor(met_3_groups, 
                          levels = c("Methylation X-linked", "Methylation not X-linked", "Not Methylation")))

CT_genes <- CT_genes_all %>%
  filter(CT_gene_type == "CT_gene")
CT_genes_types <- CT_genes_types_all %>% 
  filter(CT_gene_type == "CT_gene")

CTP_genes <- CT_genes_all %>%
  filter(CT_gene_type == "CTP_gene")
CTP_genes_types <- CT_genes_types_all %>% 
  filter(CT_gene_type == "CTP_gene")
```

```{r}
legends_param <- list(
  labels_gp = gpar(col = "black", fontsize = 4),
  title_gp = gpar(col = "black", fontsize = 5, fontface = "bold"),
  simple_anno_size = unit(0.2, "cm"),
  row_names_gp = gpar(fontsize = 3),
  annotation_name_side = "left",
  border = FALSE,
  border_gp = gpar(lwd = 0.2),
  grid_width = unit(0.2, "cm"),
  grid_height = unit(0.05, "cm"),
  legend_height = unit(1, "cm"),
  use_raster = TRUE,
               raster_device = "CairoPNG",
               raster_quality = 10)

legend_colors <- c("#5E4FA2", "#3288BD", "#66C2A5", "#ABDDA4", "#E6F598",
"#FFFFBF", "#FEE08B", "#FDAE61", "#F46D43", "#D53E4F",
"#9E0142")

TCGA_colors <- c(
"BRCA" = "midnightblue", "COAD" = "darkorchid2",
"ESCA" = "gold", "HNSC" = "deeppink2",
"LUAD" = "seagreen", "LUSC" = "seagreen3",
"SKCM" = "red3")

CCLE_colors <- c(
"Lung" = "seagreen3", "Skin" = "red3",
"Bile_Duct" = "mediumpurple1", "Bladder" = "mistyrose2",
"Colorectal" = "plum", "Lymphoma" = "steelblue1",
"Uterine" = "darkorange4", "Myeloma" = "turquoise3",
"Kidney" = "thistle4",
"Pancreatic" = "darkmagenta", "Brain" = "palegreen2",
"Gastric" = "wheat3", "Breast" = "midnightblue",
"Bone" = "sienna1", "Head_and_Neck" = "deeppink2",
"Ovarian" = "tan3", "Sarcoma" = "lightcoral",
"Leukemia" = "steelblue4", "Esophageal" = "khaki",
"Neuroblastoma" = "olivedrab1")

DAC_colors <- c(
"B2-1" = "olivedrab2", "HCT116" = "lightcoral",
"HEK293T" = "seagreen3", "HMLER" = "mediumpurple1",
"IMR5-75" = "deeppink2", "NCH1681" = "steelblue2",
"NCH612" = "red3", "TS603" = "darkmagenta")

DAC_treatment_colors <- c("CTL" = "royalblue4", "5-aza" = "maroon3")

Regulation_colors <- c("Methylation" = "indianred2",
"Not methylation"= "cyan4")

MethDep_colors <- c("MethDep" = "indianred2",
"Non-MethDep"= "cyan4")

chr_met_colors = c("X_met" = "mediumorchid2",
"X_not_met" = "mediumseagreen",
"not_X_met"= "darkorange1",
"not_X_not_met" = "dodgerblue1")

Chr_colors = c("X-linked" = "deeppink3", "not X" = "lightgreen")
Chr_colors_with_Y = c("X-linked" = "deeppink3", "Autosome" = "lightgreen", "Y-linked" = "black")

GTEX_specificity_colors = c("Testis specific" = "green",
                              "Testis preferential" = "purple",
                              "Undetectable in GTEX" = "gray")

Testis_specificity_colors = c("Testis specific" = "mediumpurple2",
"Testis preferential" = "goldenrod1",
"Undetectable in GTEX" = "gray")

# Testis_specificity_colors = c("Testis specific" = "green",
#                               "Testis preferential" = "purple")

testis_colors <- c(
"SSC" = "floralwhite", "Spermatogonia" = "moccasin",
"Early_spermatocyte" = "gold",
"Late_spermatocyte" = "orange",
"Round_spermatid" = "red2",
"Elongated_spermatid" = "darkred", "Sperm" = "purple",
"Sertoli" = "gray",
"Leydig" = "cadetblue2", "Myoid" = "springgreen3",
"Macrophage" = "gray10",
"Endothelial" = "steelblue")

Fetal_cell_colors = c( "F_PGC" = "pink",
"F_GC" = "pink3",
"F_oogonia" = "palevioletred3",
"F_oocyte" = "mediumorchid4",
"M_PGC" = "lightblue1",
"M_GC" = "steelblue3",
"M_pre_spermatogonia" = "royalblue3")

oocytes_colors <- c("Growing oocytes" = "mediumpurple1",
# "Fully grown oocytes" ="lightgreen",
# "Metaphase I" = "mediumvioletred",
"Metaphase II" = "mediumpurple4")

sex_colors <- c("F" = "palevioletred1", "M"= "skyblue1")

stage_colors <- c( "pre-meiotic" = "peachpuff",
"meiotic"= "lightsalmon",
"post-meiotic" = "maroon4",
"somatic" = "gray")

CT_genes <- CT_genes %>% filter(CT_gene_type == "CT_gene")
```

# Activation frequencies / mean expression

```{r}
####################################################################
### TCGA
####################################################################
load("~/cluster/Packages/CTdata/eh_data/TCGA_TPM.rda")
TCGA <- TCGA_TPM
TCGA <- TCGA[, TCGA$shortLetterCode != "NT"]
TCGA_Subtype <- c("TCGA-SKCM", "TCGA-LUAD", "TCGA-HNSC")
TCGA <- TCGA[, TCGA$project_id %in% TCGA_Subtype]

tumors <- unique(TCGA$project_id)

thr_pos <- 1

prop_of_pos_tum <- as_tibble(assay(TCGA), rownames = "ensembl_gene_id") %>% 
  filter(ensembl_gene_id %in% CT_genes$ensembl_gene_id) %>% 
  pivot_longer(names_to = "barcode", values_to = "TPM", -ensembl_gene_id) %>% 
  mutate(activated = ifelse(TPM >= thr_pos, "n_pos", "n_neg")) %>% 
  left_join(as_tibble(colData(TCGA), rownames = "barcode") %>% 
              dplyr::select(barcode, project_id)) %>% 
  group_by(ensembl_gene_id, activated, project_id) %>% 
  dplyr::summarize(n = n()) %>% 
  pivot_wider(names_from = activated, values_from = n) %>% 
  mutate(proportion_of_positive_tum = n_pos * 100 / (n_pos + n_neg))

mean_in_pos <- as_tibble(assay(TCGA), rownames = "ensembl_gene_id") %>% 
  filter(ensembl_gene_id %in% CT_genes$ensembl_gene_id) %>%  
  pivot_longer(names_to = "barcode", values_to = "TPM", -ensembl_gene_id)   %>% 
  mutate(activated = ifelse(TPM >= thr_pos, "activated", "not_activated")) %>% 
  filter(activated == "activated") %>% 
  left_join(as_tibble(colData(TCGA), rownames = "barcode") %>% 
              dplyr::select(barcode, project_id)) %>% 
  group_by(ensembl_gene_id, project_id) %>% 
  summarise(mean_log = log1p(mean(TPM)),
            mean_TPM = mean(TPM))

mean_and_prop_TCGA <- prop_of_pos_tum %>% 
  left_join(mean_in_pos) %>% 
  right_join(CT_genes) %>% 
  dplyr::select(ensembl_gene_id, project_id, mean_log, proportion_of_positive_tum, regulated_by_methylation, X_linked) %>% 
  mutate(source = "TCGA tumor samples")

####################################################################
### CCLE data
####################################################################
load("~/cluster/Packages/CTdata/eh_data/CCLE_data.rda")
CCLE <- CCLE_data
squamous_cell_carcinomas <- unique(grep("Squamous Cell Carcinoma", 
                                        x = CCLE$Subtype, value = TRUE))

CCLE_Subtype <- c("Melanoma", 
                 "Non-Small Cell Lung Cancer (NSCLC), Adenocarcinoma", 
                 squamous_cell_carcinomas)

CCLE <- CCLE[, !is.na(CCLE$Subtype) & CCLE$Subtype %in% CCLE_Subtype]
CCLE$type <- NA
CCLE$type[CCLE$Subtype == "Melanoma" & CCLE$primary_disease == "Skin Cancer"] <- "CCLE-SKCM"
CCLE$type[CCLE$Subtype == "Non-Small Cell Lung Cancer (NSCLC), Adenocarcinoma" & 
            CCLE$primary_disease == "Lung Cancer"] <- "CCLE-LUAD"
CCLE$type[CCLE$Subtype == "Non-Small Cell Lung Cancer (NSCLC), Squamous Cell Carcinoma" & 
            CCLE$primary_disease == "Lung Cancer"] <- "CCLE-LUSC"
CCLE$type[CCLE$Subtype %in% squamous_cell_carcinomas & 
            CCLE$primary_disease == "Head and Neck Cancer"] <- "CCLE-HNSC"

CCLE <- CCLE[, !is.na(CCLE$type) & CCLE$type != "CCLE-LUSC"]
CCLE$type <- factor(CCLE$type, levels = c("CCLE-SKCM", 
                                          "CCLE-LUAD", 
                                          #"CCLE-LUSC", 
                                          "CCLE-HNSC"))

prop_of_pos_tum <- as_tibble(assay(CCLE), rownames = "ensembl_gene_id") %>% 
    filter(ensembl_gene_id %in% CT_genes$ensembl_gene_id) %>%  
    pivot_longer(names_to = "cell", values_to = "TPM", -ensembl_gene_id) %>% 
    mutate(activated = ifelse(TPM >= thr_pos, "n_pos", "n_neg")) %>% 
    left_join(as_tibble(colData(CCLE), rownames = "cell") %>% 
                dplyr::select(cell, Subtype, lineage_subtype, type)) %>% 
    group_by(ensembl_gene_id, activated, type) %>% 
    dplyr::summarize(n = n()) %>% 
    pivot_wider(names_from = activated, values_from = n, values_fill = 0) %>% 
    mutate(proportion_of_positive_tum = n_pos * 100 / (n_pos + n_neg))

mean_in_pos <- as_tibble(assay(CCLE), rownames = "ensembl_gene_id") %>% 
  filter(ensembl_gene_id %in% CT_genes$ensembl_gene_id) %>%   
  pivot_longer(names_to = "cell", values_to = "TPM", -ensembl_gene_id)  %>% 
  mutate(activated = ifelse(TPM >= thr_pos, "activated", "not_activated")) %>% 
  filter(activated == "activated") %>% 
  left_join(as_tibble(colData(CCLE), rownames = "cell") %>% 
              dplyr::select(cell, Subtype, type)) %>% 
  group_by(ensembl_gene_id, type) %>% 
  summarise(mean_log = log1p(mean(TPM)),
            mean_TPM = mean(TPM))

mean_and_prop_CCLE <- prop_of_pos_tum %>% 
  left_join(mean_in_pos) %>% 
  right_join(CT_genes) %>% 
  dplyr::select(ensembl_gene_id, type, mean_log, proportion_of_positive_tum, regulated_by_methylation, X_linked) %>% 
  mutate(source = "CCLE tumor cell lines")

####################################################################
### combine TCGA and CCLE data
####################################################################

bilan <- rbind(mean_and_prop_CCLE, 
               mean_and_prop_TCGA %>% 
                 dplyr::rename(type = project_id))

bilan <- bilan %>% 
  mutate(type = factor(type, levels = c("CCLE-SKCM", 
                                        "CCLE-LUAD", 
                                        #"CCLE-LUSC", 
                                        "CCLE-HNSC",
                                        "TCGA-SKCM", 
                                        "TCGA-LUAD", 
                                        #"TCGA-LUSC", 
                                        "TCGA-HNSC"))) %>% 
  dplyr::rename("Positive tumors (%)" = proportion_of_positive_tum,
                'Mean expression\nin positive tumors' = mean_log)

library(ggh4x)

strip <- strip_themed(background_x = elem_list_rect(fill = c(rep("gray95", 3), rep("gray80", 3))))

fig <- bilan %>% 
  mutate(type = gsub("-", x = type, replace = "\n")) %>% 
  mutate(type = factor(type, 
                       levels =  c("CCLE\nSKCM", 
                                   "CCLE\nLUAD", 
                                    #"CCLE\nLUSC", 
                                   "CCLE\nHNSC",
                                   "TCGA\nSKCM", 
                                   "TCGA\nLUAD", 
                                    #"TCGA\nLUSC", 
                                   "TCGA\nHNSC" ))) %>% 
  mutate("CG gene" = case_when(regulated_by_methylation  ~ "MethDep",
                               !regulated_by_methylation ~ "Non-MethDep")) %>% 
  pivot_longer(names_to = "metric", values_to = "value", 
               -c(ensembl_gene_id,  `CG gene`, type, regulated_by_methylation, X_linked, source)) %>% 
  filter(!is.na(`CG gene`)) %>% 
  ggplot(aes(x = `CG gene`, y = value, color = `CG gene`)) +
  geom_jitter(size = 0.2, alpha = .8) +
  scale_colour_manual(values = MethDep_colors) +
  geom_boxplot(alpha = 0, color = "black", linewidth = 0.4) +
  facet_grid2(metric ~ type, scales = "free", switch = "y", strip = strip) +
  xlab('') +
  theme_bw() +
  theme(#aspect.ratio = 0.8,
        legend.position = "right",
        axis.position = "right",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.key.size = unit(0.5, 'cm'),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 12),
        axis.text.x = element_text(size = 0),
        axis.text.y = element_text(size = 10),
        axis.ticks.x = element_line(size = 0),
        #axis.ticks.y = element_line(size = 0),
        axis.title = element_text(size = 0),
        strip.text.x = element_text(size = 10),
        strip.text.y = element_text(size = 12),
        strip.background.y =  element_blank()) +
  scale_y_continuous(position = "right")

pdf(file = "../figs/Fig5B_CG_in_CCLE_TCGA.pdf", width = 6, height = 4)
fig
dev.off()

fig
```

## t-tests

```{r}
tumors <- levels(bilan$type)
tum <- tumors[1]
prop_of_positive_tum_pval <- t.test(bilan %>% 
                pivot_longer(names_to = "metric", values_to = "value", 
                             -c(ensembl_gene_id,  type, regulated_by_methylation, X_linked, source)) %>% 
                filter(!is.na(regulated_by_methylation)) %>% 
                filter(type == tum, metric == "Positive tumors (%)") %>% 
                filter(regulated_by_methylation) %>% 
                pull(value),
              bilan %>% 
                pivot_longer(names_to = "metric", values_to = "value", 
                             -c(ensembl_gene_id,  type, regulated_by_methylation, X_linked, source)) %>% 
                filter(!is.na(regulated_by_methylation)) %>% 
                filter(type == tum, metric == "Positive tumors (%)") %>% 
                filter(regulated_by_methylation) %>% 
                pull(value), alternative = "greater")$p.value

mean_log_pval = t.test(bilan %>% 
                pivot_longer(names_to = "metric", values_to = "value", 
                             -c(ensembl_gene_id,  type, regulated_by_methylation, X_linked, source)) %>% 
                filter(!is.na(regulated_by_methylation)) %>% 
                filter(type == tum, metric == "Mean expression\nin positive tumors") %>% 
                filter(regulated_by_methylation) %>% 
                pull(value),
              bilan %>% 
                pivot_longer(names_to = "metric", values_to = "value", 
                             -c(ensembl_gene_id,  type, regulated_by_methylation, X_linked, source)) %>% 
                filter(!is.na(regulated_by_methylation)) %>% 
                filter(type == tum, metric == "Mean expression\nin positive tumors") %>% 
                filter(!regulated_by_methylation) %>% 
                pull(value), alternative = "greater")$p.value

t_tests <- tibble(Tum = tum, prop_of_positive_tum_pval = prop_of_positive_tum_pval, mean_log_pval = mean_log_pval)

for (tum in tumors[-1]){

  tmp1 = t.test(bilan %>% 
                pivot_longer(names_to = "metric", values_to = "value", 
                             -c(ensembl_gene_id,  type, regulated_by_methylation, X_linked, source)) %>% 
                filter(!is.na(regulated_by_methylation)) %>% 
                filter(type == tum, metric == "Positive tumors (%)") %>% 
                filter(regulated_by_methylation) %>% 
                pull(value),
              bilan %>% 
                pivot_longer(names_to = "metric", values_to = "value", 
                             -c(ensembl_gene_id,  type, regulated_by_methylation, X_linked, source)) %>% 
                filter(!is.na(regulated_by_methylation)) %>% 
                filter(type == tum, metric == "Positive tumors (%)") %>% 
                filter(!regulated_by_methylation) %>% 
                pull(value), alternative = "greater")$p.value

  tmp2 = t.test(bilan %>% 
                pivot_longer(names_to = "metric", values_to = "value", 
                             -c(ensembl_gene_id,  type, regulated_by_methylation, X_linked, source)) %>% 
                filter(!is.na(regulated_by_methylation)) %>% 
                filter(type == tum, metric == "Mean expression\nin positive tumors") %>% 
                filter(regulated_by_methylation) %>% 
                pull(value),
              bilan %>% 
                pivot_longer(names_to = "metric", values_to = "value", 
                             -c(ensembl_gene_id,  type, regulated_by_methylation, X_linked, source)) %>% 
                filter(!is.na(regulated_by_methylation)) %>% 
                filter(type == tum, metric == "Mean expression\nin positive tumors") %>% 
                filter(!regulated_by_methylation) %>% 
                pull(value), alternative = "greater")$p.value

  tmp <- tibble(Tum = tum, prop_of_positive_tum_pval = tmp1,  mean_log_pval = tmp2)
  t_tests <- rbind(t_tests, tmp)
}

t_tests$prop_of_positive_tum_padjusted <- p.adjust(t_tests$prop_of_positive_tum_pval, method = "BH")
t_tests$Mean_expression_padjusted <- p.adjust(t_tests$mean_log_pval, method = "BH")
DT::datatable(t_tests %>% dplyr::select(Tum, prop_of_positive_tum_padjusted, Mean_expression_padjusted))
```

=> Always significant (except the `proportion of positive tumors` in TCGA-LUAD)

```{r}
strip <- strip_themed(background_x = elem_list_rect(fill = c("gray95", "gray80")))

bilan_all_types <- bilan %>% 
         filter(!is.na(regulated_by_methylation)) %>% 
         mutate("CG gene" = case_when(regulated_by_methylation ~ "MethDep",
                               !regulated_by_methylation ~ "Non-MethDep")) %>% 
       filter(!is.na(regulated_by_methylation)) %>% 
  group_by(ensembl_gene_id, source) %>% 
  summarize(`Mean expression\nin positive tumors` = mean(`Mean expression\nin positive tumors`, na.rm = TRUE),
            `Positive tumors (%)` = mean(`Positive tumors (%)`, na.rm = TRUE)) %>% 
  left_join(bilan %>% filter(!is.na(regulated_by_methylation)) %>% 
         mutate("CG gene" = case_when(regulated_by_methylation ~ "MethDep",
                               !regulated_by_methylation ~ "Non-MethDep")) %>% 
           dplyr::select(ensembl_gene_id, "CG gene", "X_linked") %>% unique())
```

```{r}

selected_genes_in_TCGA <- c("MAGEA3", "MAGEA6", "MAGEA4", "MAGEA1", "MAGEA12", 
                            "MAGEC2", "CTAG2", "CBY2", "PAGE2", "PAGE5, GAGE2A", 
                            "SPANXB1", "PAGE1", "LINC01518", "BRDT", #"CT45A10", 
                            "SSX1", "MAGEC2", "MAGEB2", "MAGEA10")

selected_genes_in_CCLE <- c("MAGEA3", "MAGEA6", "MAGEA4", #"MAGEA1",
                            "TRIML2", "LINC02864", 
                            "MAGEA12",
                            "MAGEA2", "SPANXB1","LIN1518",
                            "XAGE1A", "XAGE1B", "CTAG2", "CT45A1", "GAGE2A", 
                            "PAGE2", "DSCR8", "MAGEB2", #"VCY", "SSX2B", "TUBA3C",
                            "GAGE12E")

fig <- ggplot(bilan_all_types %>% left_join(CT_genes %>% 
                                              dplyr::select(ensembl_gene_id, external_gene_name) %>% 
                                              dplyr::rename(gene = external_gene_name)) %>% 
                mutate(source = factor(source, 
                       levels =  c("TCGA tumor samples", "CCLE tumor cell lines"))) %>% 
         mutate(Location = case_when(X_linked ~ "X-linked",
                                       !X_linked ~ "Not X")),
         aes(x = `Mean expression\nin positive tumors`, y = `Positive tumors (%)`,
             label = gene, colour = `CG gene`, shape = Location)) +
  geom_point(size = 2) +
  scale_colour_manual(values = MethDep_colors) +
  facet_wrap2(~ source, ncol = 2, strip = strip) +
  ggrepel::geom_text_repel(
    data = bilan_all_types %>% 
      left_join(CT_genes %>% 
                  dplyr::select(ensembl_gene_id, external_gene_name) %>% 
                  dplyr::rename(gene = external_gene_name)) %>% 
      mutate(Location = case_when(X_linked ~ "X-linked",
                                  !X_linked ~ "Not X")) %>%
      filter(source == "TCGA tumor samples" & gene %in% selected_genes_in_TCGA) ,
    aes(x = `Mean expression\nin positive tumors`, y = `Positive tumors (%)`,
        label = gene), size = 2.5, color = "black", min.segment.length = 2,
    max.overlaps = 30) + #, direction = "x") +
  ggrepel::geom_text_repel(
    data = bilan_all_types %>% 
      left_join(CT_genes %>% 
                  dplyr::select(ensembl_gene_id, external_gene_name) %>% 
                  dplyr::rename(gene = external_gene_name)) %>% 
      mutate(Location = case_when(X_linked ~ "X-linked",
                                  !X_linked ~ "Not X")) %>%
      filter(source == "CCLE tumor cell lines" & gene %in% selected_genes_in_CCLE) ,
    aes(x = `Mean expression\nin positive tumors`, y = `Positive tumors (%)`,
        label = gene), size = 2.5, color = "black", min.segment.length = 2,
    max.overlaps = 30) + #, direction = "x") +
  xlab("Mean expression log(TPM + 1) in positive tumors") +
  ylab("Proportion of positive tumors (%)") +

  theme_bw() +
  theme(legend.position = "right",
        #aspect.ratio = 0.8,
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        #legend.key.size = unit(0.3, 'cm'),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        axis.ticks = element_line(size = 0.1),
        axis.title = element_text(size = 12),
        strip.text = element_text(size = 12))

pdf(file = "../figs/Fig5C_CG_in_CCLE_TCGA_by_chr.pdf", width = 7, height = 3)
fig
dev.off()

fig

CCLE_data <- CCLE
TCGA_data <- TCGA
```

# CT-met cluster together

```{r}
legends_param <- list(
  labels_gp = gpar(col = "black", fontsize = 6),
  title_gp = gpar(col = "black", fontsize = 6),
  simple_anno_size = unit(0.8, "cm"),
  border_gp = gpar(lwd = 0.01),
  row_names_gp = gpar(fontsize = 1),
  annotation_name_side = "none")

tum_heatmap <- function(genes = NULL,
                               CCLE_Subtype = c("CCLE-SKCM", "CCLE-LUAD", 
                                            #"CCLE-LUSC",  
                                            "CCLE-HNSC"),
                               TCGA_Subtype = c("TCGA-SKCM", "TCGA-LUAD", 
                                            #"TCGA-LUSC",  
                                            "TCGA-HNSC"),
                               h_width = 5,
                               h_height = 10,
                               downsample_to_n = 450,
                               fontsize = 4, 
                               clust_method = "ward.D",
                               split_rows_by = "regulation", 
                               show_annotation_legend = TRUE,
                               clustRow = FALSE,
                               clustCol = TRUE){
  # CCLE data
  ## downsample tumors to have the same number of tumors in each type
  CCLE <- CCLE_data[, CCLE_data$type == CCLE_Subtype[1]]
  CCLE <- CCLE[, 1:min(ncol(CCLE), downsample_to_n)]

  for (tum in CCLE_Subtype[-1]){
    tmp <- CCLE_data[, CCLE_data$type == tum]
    tmp <- tmp[, 1:min(ncol(tmp), downsample_to_n)]
    CCLE <- cbind(CCLE, tmp)
  }

  CCLE$type <- gsub("-", x = CCLE$type, replacement = "\n")
  CCLE$type <- factor(CCLE$type, levels = c("CCLE\nSKCM", "CCLE\nLUAD", #"CCLE\nLUSC", 
                                            "CCLE\nHNSC"))

  colData(CCLE) <- colData(CCLE)[, c(14, 30)]

  # TCGA data
  TCGA <- TCGA_data

  ## downsample tumors to have the same number of tumors in each type
  TCGA <- TCGA_data[, TCGA_data$project_id == TCGA_Subtype[1]]
  TCGA <- TCGA[, 1:min(ncol(TCGA), downsample_to_n)]

  for (tum in TCGA_Subtype[-1]){
    tmp <- TCGA_data[, TCGA_data$project_id == tum]
    tmp <- tmp[, 1:min(ncol(tmp), downsample_to_n)]
    TCGA <- cbind(TCGA, tmp)
  }

  TCGA$type <- TCGA$project_id
  TCGA$type <- gsub("-", x = TCGA$type, replacement = "\n")
  TCGA$type <- factor(TCGA$type, levels = c("TCGA\nSKCM", "TCGA\nLUAD", #"TCGA\nLUSC", 
                                            "TCGA\nHNSC"))

  #names(colData(TCGA))
  colData(TCGA) <- colData(TCGA)[, c(56, 66)]
  #TCGA$project_id <- gsub(pattern = "TCGA-", x = TCGA$project_id, replacement = '')
  my_genes <- genes[genes %in% c(rowData(CCLE)$external_gene_name, rowData(TCGA)$external_gene_name)]
  my_ensembl <- CT_genes %>% filter(external_gene_name %in% my_genes) %>% pull(ensembl_gene_id)

  print(length(my_genes))

  names(colData(TCGA)) <- c("Subtype", "type")
  tum_SE <- cbind(CCLE[my_ensembl,], TCGA[my_ensembl,])

  df_col <- tibble(sample = colnames(tum_SE), type = tum_SE$type)

  df_col <- df_col %>% arrange(type)

  tum_SE <- tum_SE[, df_col$sample]
  split_by <- tum_SE$type

  tum_col <- c(rep("gray95", 4), rep("gray80", 4))
  names(tum_col) <- levels(tum_SE$type)

  ## Use gene names instead of ENSEMBL IDs
  mat <- SummarizedExperiment::assay(tum_SE)
  rownames(mat) <- rowData(tum_SE)$external_gene_name
  mat <- log1p(mat)

  gene_type <- CT_genes %>% 
    filter(external_gene_name %in% genes) %>% 
    mutate(gene = factor(external_gene_name, levels = genes)) %>% 
    mutate(Regulation = ifelse(regulated_by_methylation, "MethDep", "Non\nMethDep")) %>% 
     mutate(Regulation = factor(Regulation, levels = c("MethDep", "Non\nMethDep"))) %>% 
    arrange(gene)

 if (clustRow) gene_type <- gene_type %>% arrange(Regulation)
 rowSplit <- gene_type$Regulation

  heatmap_colors <- colorRamp2(
    seq(0, 5, length = 11),                 
    legend_colors)

  Heatmap(mat[as.vector(gene_type$gene), ],
          top_annotation = HeatmapAnnotation(
            type = anno_block(gp = gpar(fill = c(rep("gray95", 3), rep("gray80", 3)), lwd = 0.2),
                              height = unit(c(0.7), "cm"),
                              labels = levels(df_col$type), 
                              labels_gp = gpar(col = "black", fontsize = 8))),
          left_annotation = rowAnnotation(
            met = anno_block(gp = gpar(fill = MethDep_colors, lwd = 0.2),
                             width = unit(c(0.8), "cm"),
                             labels = c("MethDep", "Non\nMethDep"), 
                             labels_gp = gpar(col = "white", fontsize = 10))),
          name = "logTPM",
          #column_title = paste0("CCLE tumor cell lines"),
          #column_title_gp = gpar(fontsize = 7, face = "bold"),
          clustering_method_rows = clust_method,
          clustering_method_columns = clust_method,
          column_split = split_by,
          row_gap = unit(c(1), "mm"),
          col = heatmap_colors,
          cluster_rows = clustRow,
          cluster_columns = clustCol,
          cluster_column_slices = FALSE,
          cluster_row_slices = FALSE,
          show_row_names = FALSE,
          show_column_names = FALSE,
          show_row_dend = FALSE,
          show_column_dend = FALSE,
          column_names_gp = gpar(fontsize = 6),
          column_names_centered = TRUE,
          row_names_gp = gpar(fontsize = fontsize),
          row_names_side = "left",
          border = TRUE,
          border_gp = gpar(lwd = 0.5),
          column_names_side = c("top"),
          column_names_rot = 90,
          row_title_gp = gpar(fontsize = 0),
          column_title_gp = gpar(fontsize = 0),
          #top_annotation = column_ha_tumor,
          #right_annotation = rAnnot,
          #left_annotation = rAnnot,
          row_split = rowSplit,
          width = unit(h_width, "cm"),
          height = unit(h_height, "cm"),
          use_raster = TRUE,
          raster_device = "CairoPNG",
          raster_quality = 10,
          heatmap_legend_param = legends_param,
          show_heatmap_legend = TRUE)
}
```

```{r}
met_genes <- CT_genes %>% 
  filter(regulated_by_methylation) %>% 
  pull(external_gene_name)

not_met_genes <- CT_genes %>% 
  filter(!regulated_by_methylation) %>% 
  pull(external_gene_name)

h <- tum_heatmap(c(met_genes, not_met_genes), 
                         downsample_to_n = 50, 
                         h_width = 7,
                         h_height = 6,
                         fontsize = 0, 
                   clust_method = "ward.D",
                   clustRow = TRUE,
                   #show_row_names = TRUE,
                   show_annotation_legend = TRUE,
                   split_rows_by = "regulation")
h
pdf(file = "../figs/Fig5A_CG_clustering_in_tumors.pdf")#, width = 7, height = 3)
h
dev.off()

```

# Activation score distribution

- Activated when TPM >= 1

- Activation score :

    number of methDep gene activated / number of methDep gene

    number of Non-methDep gene activated / number of Non-methDep gene

```{r}
### in CCLE
CCLE <- CCLE_data

keep <- rowSums(assay(CCLE)) > 0
CCLE <- CCLE[keep, ]

met_genes <- CT_genes %>% 
  filter(ensembl_gene_id %in% rownames(CCLE)) %>% 
  filter(regulated_by_methylation) %>% 
  pull(ensembl_gene_id)

not_met_genes <- CT_genes %>% 
  filter(ensembl_gene_id %in% rownames(CCLE)) %>% 
  filter(!regulated_by_methylation) %>% 
  pull(ensembl_gene_id)

CCLE <- CCLE[c(met_genes,not_met_genes), ]
binary_mat_methDep <- assay(CCLE) >= 1
methDep_score <- colSums(binary_mat_methDep) / length(met_genes)
CCLE$methDep_score <- methDep_score

binary_mat_NonMethDep <- assay(CCLE[not_met_genes,]) >= 1
NonMethDep_score <- colSums(binary_mat_NonMethDep) / length(not_met_genes)
CCLE$NonMethDep_score <- NonMethDep_score

### in TCGA
TCGA <- TCGA_data

keep <- rowSums(assay(TCGA)) > 0
TCGA <- TCGA[keep, ]

met_genes <- CT_genes %>% 
  filter(ensembl_gene_id %in% rownames(TCGA)) %>% 
  filter(regulated_by_methylation) %>% 
  pull(ensembl_gene_id)

not_met_genes <- CT_genes %>% 
  filter(ensembl_gene_id %in% rownames(TCGA)) %>% 
  filter(!regulated_by_methylation) %>% 
  pull(ensembl_gene_id)
TCGA <- TCGA[c(met_genes, not_met_genes), ]

binary_mat_methDep <- assay(TCGA[met_genes,]) >= 1
methDep_score <- colSums(binary_mat_methDep) / length(met_genes)
TCGA$methDep_score <- methDep_score

binary_mat_NonMethDep <- assay(TCGA[not_met_genes,]) >= 1
NonMethDep_score <- colSums(binary_mat_NonMethDep) / length(not_met_genes)
TCGA$NonMethDep_score <- NonMethDep_score

bilan <- rbind(as_tibble(colData(CCLE)) %>%
        mutate(source = "CCLE") %>%
        dplyr::select(source, methDep_score, NonMethDep_score),
      as_tibble(colData(TCGA)) %>%
        mutate(source = "TCGA") %>%
        dplyr::select(source, methDep_score, NonMethDep_score)) %>% 
  pivot_longer(names_to = "methDependency", values_to = "score", -source) %>% 
  mutate("methDependency" = ifelse(methDependency == "methDep_score", "MethDep", "Non-MethDep"))

bilan <- bilan %>% 
  mutate(group = case_when(score == 0 ~ "0",
                           score >= 0 & score < 0.05 ~ "<5",
                           score >= 0.05 & score < 0.10 ~ "5-10",
                           score >= 0.10 & score < 0.15 ~ "10-15",
                           score >= 0.15 & score < 0.20 ~ "15-20",
                           score >= 0.20 ~ ">20")) %>% 
  mutate(group = factor(group, levels = c("0", "<5", "5-10", "10-15","15-20", ">20")))
strip <- strip_themed(background_x = elem_list_rect(fill = c("gray95","gray80")))
fig <- bilan %>% 
  group_by(source, methDependency, group) %>% 
  summarize(n = n())  %>% #filter(methDependency == "MethDep") %>% 
  ggplot(aes(x = group, y = n, group = methDependency, fill = methDependency)) +
  geom_col(stat = "identity", width = 0.6, position = position_dodge(0.6))+
  scale_fill_manual(values = MethDep_colors)+
  xlab("Activation scores") +
  ylab("Number of tumors") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        #aspect.ratio = 1, 
        legend.position = c(0.65, 0.95),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 0),
        legend.key.size = unit(0.2, 'cm'),
        legend.background = element_blank()) +
  facet_wrap2(~source, ncol = 1, scales = "free_y", strip = strip)
fig
pdf(file = "../figs/Fig5D_ActivationScores.pdf", width = 3, height = 4)
fig
dev.off()
```

### 2025-03-25 4.CT_in_germ_cells.Rmd
Added code used to generate the paper

---
title: "CT genes analysis in germ cells"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 4
    toc_float: false
---

```{r, echo = FALSE, Results = 'asis'}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
library("tidyverse")
library("CTexploreR")
library("ComplexHeatmap")
library("circlize")
library("SummarizedExperiment")
library("flextable")

set_flextable_defaults(
font.family = "Arial", font.size = 10,
border.color = "black",
padding = 6,
background.color = "#EFEFEF",
big.mark = "")
```

#

```{r}
all_genes_types <- all_genes %>% 
  dplyr::rename(gene = external_gene_name) %>% 
  mutate(Regulation = case_when(regulated_by_methylation ~ "Methylation",
                                !regulated_by_methylation ~ "Not methylation")) %>%
  mutate(Chr = case_when(X_linked ~ "X-linked",
                         !X_linked ~ "not X")) %>%  
  mutate(chromosome = case_when(chr == "X" ~ "X-linked",
                                chr == "Y" ~ "Y-linked",
                                !chr %in% c("X", "Y") ~ "Autosome")) %>% 
  mutate(chr_met = case_when(regulated_by_methylation & X_linked ~ "X_met",
                             !regulated_by_methylation & X_linked ~ "X_not_met",
                             regulated_by_methylation & !X_linked ~ "not_X_met",
                             !regulated_by_methylation & !X_linked ~ "not_X_not_met")) %>% 
  mutate(GTEX_specificity = case_when(GTEX_category == "testis_specific" ~ "Testis specific",
                                        GTEX_category == "testis_preferential" ~ "Testis preferential",
                                        GTEX_category == "lowly_expressed" ~ "Undetectable in GTEX")) %>% 
  mutate(testis_specificity = case_when(testis_specificity == "testis_specific" ~ "Testis specific",
                                        testis_specificity == "testis_preferential" ~ "Testis preferential")) %>%
  mutate(testis_specificity = factor(testis_specificity, levels = c("Testis specific", "Testis preferential"))) %>% 
  mutate(chr_met = factor(chr_met, 
                          levels = c("X_met", "not_X_met", "X_not_met", "not_X_not_met"))) %>% 
  mutate(met_3_groups = case_when(regulated_by_methylation & X_linked ~ "Methylation X-linked",
                                   regulated_by_methylation & !X_linked ~ "Methylation not X-linked",
                             !regulated_by_methylation ~ "Not Methylation")) %>% 
  mutate(met_3_groups = factor(met_3_groups, 
                          levels = c("Methylation X-linked", "Methylation not X-linked", "Not Methylation")))

CT_genes_types <- all_genes_types %>%
  filter(CT_gene_type == "CT_gene")

CT_genes <- CT_genes %>% 
  filter(CT_gene_type == "CT_gene")
```

```{r}
legends_param <- list(
  labels_gp = gpar(col = "black", fontsize = 4),
  title_gp = gpar(col = "black", fontsize = 5, fontface = "bold"),
  simple_anno_size = unit(0.2, "cm"),
  row_names_gp = gpar(fontsize = 3),
  annotation_name_side = "left",
  border = FALSE,
  border_gp = gpar(lwd = 0.2),
  grid_width = unit(0.2, "cm"),
  grid_height = unit(0.05, "cm"),
  legend_height = unit(1, "cm"),
  use_raster = TRUE,
               raster_device = "CairoPNG",
               raster_quality = 10)

legend_colors <- c("#5E4FA2", "#3288BD", "#66C2A5", "#ABDDA4", "#E6F598",
"#FFFFBF", "#FEE08B", "#FDAE61", "#F46D43", "#D53E4F",
"#9E0142")

TCGA_colors <- c(
"BRCA" = "midnightblue", "COAD" = "darkorchid2",
"ESCA" = "gold", "HNSC" = "deeppink2",
"LUAD" = "seagreen", "LUSC" = "seagreen3",
"SKCM" = "red3")

CCLE_colors <- c(
"Lung" = "seagreen3", "Skin" = "red3",
"Bile_Duct" = "mediumpurple1", "Bladder" = "mistyrose2",
"Colorectal" = "plum", "Lymphoma" = "steelblue1",
"Uterine" = "darkorange4", "Myeloma" = "turquoise3",
"Kidney" = "thistle4",
"Pancreatic" = "darkmagenta", "Brain" = "palegreen2",
"Gastric" = "wheat3", "Breast" = "midnightblue",
"Bone" = "sienna1", "Head_and_Neck" = "deeppink2",
"Ovarian" = "tan3", "Sarcoma" = "lightcoral",
"Leukemia" = "steelblue4", "Esophageal" = "khaki",
"Neuroblastoma" = "olivedrab1")

DAC_colors <- c(
"B2-1" = "olivedrab2", "HCT116" = "lightcoral",
"HEK293T" = "seagreen3", "HMLER" = "mediumpurple1",
"IMR5-75" = "deeppink2", "NCH1681" = "steelblue2",
"NCH612" = "red3", "TS603" = "darkmagenta")

DAC_treatment_colors <- c("CTL" = "royalblue4", "5-aza" = "maroon3")

Regulation_colors <- c("Methylation" = "indianred2",
"Not methylation"= "cyan4")

chr_met_colors = c("X_met" = "mediumorchid2",
"X_not_met" = "mediumseagreen",
"not_X_met"= "darkorange1",
"not_X_not_met" = "dodgerblue1")

Chr_colors = c("X-linked" = "deeppink3", "not X" = "lightgreen")
Chr_colors_with_Y = c("X-linked" = "deeppink3", "Autosome" = "lightgreen", "Y-linked" = "black")

GTEX_specificity_colors = c("Testis specific" = "green",
                              "Testis preferential" = "purple",
                              "Undetectable in GTEX" = "gray")

Testis_specificity_colors = c("Testis specific" = "mediumpurple2",
"Testis preferential" = "goldenrod1",
"Undetectable in GTEX" = "gray")

# Testis_specificity_colors = c("Testis specific" = "green",
#                               "Testis preferential" = "purple")

testis_colors <- c(
  "Spermatogonial stem cells" = "floralwhite", "Spermatogonia" = "moccasin",
  "Early spermatocyte" = "gold",
  "Late spermatocyte" = "orange",
  "Round spermatid" = "red2",
  "Elongated spermatid" = "darkred", "Sperm" = "purple")

Fetal_cell_colors = c("Primordial germ Cells" = "lightblue1",
                      "Primordial germ Cells (F)" = "pink",
                      "Fetal germ cells (F)" = "pink3",
                      "Oogonia" = "palevioletred3",
                      "Oocytes" = "mediumorchid4",
                      "Primordial germ Cells" = "lightblue1",
                      "Primordial germ Cells (M)" = "lightblue1",
                      "Fetal germ cells" = "steelblue3",
                      "Fetal germ cells (M)" = "steelblue3",
                      "Pre-spermatogonia" = "blue",
                      "soma (F)" = "gray",
                      "soma (M)" = "gray30")

Fetal_cellType_colors = c(
                      "Mitotic FGC (F)" = "pink3",
                      "Meiotic FGC (F)" = "palevioletred3",
                      "Oogenesis FGC (F)" = "mediumorchid4",
                      "Mitotic FGC (M)" = "steelblue3",
                      "Mitotic arrested (M)" = "blue",
                      "soma (F)" = "gray",
                      "soma (M)" = "gray30")

oocytes_colors <- c("Growing oocytes" = "mediumpurple1",
"Fully grown oocytes" ="lightgreen",
"Metaphase I" = "mediumvioletred",
"Metaphase II" = "mediumpurple4")

sex_colors <- c("F" = "deeppink", "M"= "steelblue")

stage_colors <- c( "pre-meiotic" = "peachpuff",
"meiotic"= "lightsalmon",
"post-meiotic" = "maroon4",
"somatic" = "gray")

Stage_colors <- c( "Fetal" = "lightgreen",
"Adult"= "steelblue")

MethDep_colors <- c("MethDep" = "indianred2",
"Non-MethDep"= "cyan4")

CT_group_colors = c("X-linked MethDep" = "indianred2", 
                    "Not X MethDep" = "darksalmon",
                  "Non-MethDep" =   "cyan4")
```

# Fig6A and 7A: Heatmaps

```{r}
testis_sce <- CTdata::testis_sce()

testis_sce <- testis_sce[CT_genes$external_gene_name[
  CT_genes$external_gene_name %in% rownames(testis_sce)],]

missing <- CT_genes$external_gene_name[
  !CT_genes$external_gene_name %in% rownames(testis_sce)]
```

- `r length(missing)` genes not detected at all in adult testis scRNAseq:

`r missing`

```{r}
# pool Sperm1 and Sperm2 types
levels(testis_sce$type)[c(7,8)] <- "Sperm"
germcells <-  c("SSC", "Spermatogonia", "Early_spermatocyte", 
                "Late_spermatocyte", "Round_spermatid", 
                "Elongated_spermatid", "Sperm")

testis_sce <- testis_sce[, testis_sce$type %in% germcells]
testis_sce$type <- factor(testis_sce$type, levels = germcells)

# Keep genes detected in at least x % of germ cells 
# ! Each cell type is present at different proportion 
# (much more sperm cells than SSC cells for instance)
# table(testis_sce$type)
# => Calculate for each gene the proportion of expressing cells 
# (> testis_logcount_thr) and keep genes expressed in at least x% of a cell type
# instead of just keeping genes expressed in at least x% of cells.
testis_percent_thr <- 0.05
testis_logcount_thr <- 0

percent_of_expressing_cells <- 
  tibble(gene = rownames(testis_sce),
         prop = rowSums(assay(testis_sce[, testis_sce$type %in% germcells[1]], 
                              "logcounts") > testis_logcount_thr) / 
           dim(testis_sce[, testis_sce$type %in% germcells[1]])[2])
names(percent_of_expressing_cells) <- c("gene", germcells[1])

for (cell in germcells[-1]) {
  tmp <- 
    tibble(gene = rownames(testis_sce),
           prop = rowSums(assay(testis_sce[, testis_sce$type %in% cell], 
                                           "logcounts") > testis_logcount_thr) / 
             dim(testis_sce[, testis_sce$type %in% cell])[2])
  names(tmp) <- c("gene", cell)
  percent_of_expressing_cells <- left_join(percent_of_expressing_cells, tmp)
}

kept_genes <- rownames(testis_sce)[rowSums(percent_of_expressing_cells[,-1] > testis_percent_thr) > 0]
```

- `r length(kept_genes)` genes are detected in at least `r testis_percent_thr * 100`% of a specific testis cell type

- Genes removed (detected in less than `r testis_percent_thr * 100`% of cells):

`r rownames(testis_sce)[!rownames(testis_sce) %in% kept_genes]`

```{r}
testis_sce <- testis_sce[kept_genes,]
```

```{r}
FGC_sce <- CTdata::FGC_sce()
FGC_sce <- FGC_sce[, FGC_sce$low_ncounts == "False"]
FGC_sce <- FGC_sce[, FGC_sce$high_mito == "False"]
# add missing CT genes (not in the downloaded dataset, as not detected (below detection) => were removed)
# Add zero count values for these genes
rowData(FGC_sce) <- NULL

levels(FGC_sce$type) <- 
    c("Primordial germ Cells (F)", "Fetal germ cells (F)", "Oogonia", "Oocytes",
      "Primordial germ Cells (M)", "Fetal germ cells (M)", "Pre-spermatogonia")

FGC_sce <- FGC_sce[rownames(testis_sce)[rownames(testis_sce) %in% rownames(FGC_sce)],]
```

Genes detected in adult testis scRNAseq but not in fetal gonads scRNAseq:

```{r}
missing_CT <- rownames(testis_sce)[!rownames(testis_sce) %in% rownames(FGC_sce)]
missing_CT
```

```{r}
if (length(missing_CT) > 0) {
  print("added zero values to these genes in fetal scRNAseq")
  print(missing_CT)
  sce_missing_CT <- FGC_sce[1:length(missing_CT),]
  rownames(sce_missing_CT) <- missing_CT
  count_mat_missing_CT <- matrix(nrow = length(missing_CT), ncol = ncol(FGC_sce), data = 0)
  rownames(count_mat_missing_CT) <- missing_CT
  colnames(count_mat_missing_CT) <- colnames(FGC_sce)
  assay(sce_missing_CT,1) <- count_mat_missing_CT
  assay(sce_missing_CT,2) <- count_mat_missing_CT
  FGC_sce <- rbind(FGC_sce, sce_missing_CT)
}
```

- Genes not detected in adult oocytes dataset:

```{r}
oocytes_sce <- CTdata::oocytes_sce()

# add missing CT genes (not in the downloaded dataset, as not detected (below detection) => were removed)
# Add zero count values for these genes
rowData(oocytes_sce) <- NULL
oocytes_sce <- oocytes_sce[, oocytes_sce$type %in% c("Growing oocytes",  
  "Fully grown oocytes",  
  "Metaphase I", 
  "Metaphase II")]
oocytes_sce$type <- factor(oocytes_sce$type, levels = c("Growing oocytes",  
                                                        "Fully grown oocytes",  
                                                        "Metaphase I", 
                                                        "Metaphase II"))

oocytes_sce <- oocytes_sce[rownames(testis_sce)[
  rownames(testis_sce) %in% rownames(oocytes_sce)], ]

missing_CT <- rownames(testis_sce)[
  !rownames(testis_sce) %in% rownames(oocytes_sce)]

```

`r missing_CT`

```{r}
if (length(missing_CT) > 0) {
  print("added zero values to these genes in oocytes scRNAseq")
  print(missing_CT)
  sce_missing_CT <- oocytes_sce[1:length(missing_CT),]
rownames(sce_missing_CT) <- missing_CT
count_mat_missing_CT <- matrix(nrow = length(missing_CT), ncol = ncol(oocytes_sce), data = 0)
rownames(count_mat_missing_CT) <- missing_CT
colnames(count_mat_missing_CT) <- colnames(oocytes_sce)
assay(sce_missing_CT,1) <- count_mat_missing_CT
assay(sce_missing_CT,2) <- count_mat_missing_CT
oocytes_sce <- rbind(oocytes_sce, sce_missing_CT)
}
```

```{r }
legends_param <- list(
  labels_gp = gpar(col = "black", fontsize = 5),
  title_gp = gpar(col = "black", fontsize = 5),
  simple_anno_size = unit(0.2,"cm"),
  row_names_gp = gpar(fontsize = 1),
  annotation_name_side = "left",
  border = FALSE,
  border_gp = gpar(lwd = 0.01),
  grid_width = unit(0.2, "cm"),
  grid_height = unit(0.2, "cm"),
  legend_height = unit(0.2, "cm"),
  annotation_name_gp = gpar(fontsize = 0))

```

Kept `r nrow(testis_sce)` CT genes for the analysis.

```{r}
#######################################################################
# Function to downsample cell types
#######################################################################
sample_cells <- function(barcodes, nmax){
  sample(barcodes, size = min(nmax, length(barcodes)), replace = FALSE)
}
```

```{r}
#######################################################################
## Function to draw expression in FGC with specific annotations
## and down-sampling the number of cells from each cell type
#######################################################################
draw_FGC_heatmap <- function(genes, 
                             ncells_max = NULL,
                             scale_lims = NULL,
                             title_fs = 6,
                             sex = c("F", "M"),
                             clusterRow = FALSE, 
                             font_size = 6,
                             show_legend = TRUE,
                             h_width = 5,
                             h_height = 10,
                             column_name_rot = 90){

  fetal_sce <- FGC_sce
  fetal_sce$Stage <- "Fetal"

  # Set the colors used in the final heatmap based on all samples (F + M), 
  # even when representing only F or M samples on the heatmap
  # => same scale of colors
  mat <-  as.matrix(assay(fetal_sce, "logcounts")[genes, ])

  if (is.null(scale_lims)) {
    scale_lims <- c(0, quantile(rowMaxs(mat), 0.8))
  }

  heatmap_colors <- colorRamp2(
    seq(scale_lims[1], scale_lims[2], length = 11),                 
    legend_colors)

  fetal_sce <- fetal_sce[, fetal_sce$sex %in% sex]

  df_col_FCS <- data.frame(development_stage = fetal_sce$development_stage,
                           sex = fetal_sce$sex,
                           stage = fetal_sce$stage,
                           donor_id = fetal_sce$donor_id,
                           fetal_cell_type = fetal_sce$cell_type,
                           fetal_celltype = fetal_sce$celltype,
                           type = fetal_sce$type,
                           time = fetal_sce$PCW,
                           Stage = fetal_sce$Stage)

  rownames(df_col_FCS) <- colnames(fetal_sce)
  df_col_FCS$type <- factor(df_col_FCS$type)
  if (all(sex == "M")) levels(df_col_FCS$type) <- c("PGC", "FGC", "Pre-\nSpg")
  if (all(sex == "F")) levels(df_col_FCS$type) <- c("PGC", "FGC", "Oogonia", "Oocytes")

  #Randomly pick ncells_max cells of each cell type
  if (!is.null(ncells_max)) {
    set.seed(123)
    selected_cells <- sample_cells(
      rownames(df_col_FCS[df_col_FCS$type == levels(df_col_FCS$type)[1],]), 
      ncells_max)
    for (type in levels(df_col_FCS$type)[-1]){
      tmp <- sample_cells(
        rownames(df_col_FCS[df_col_FCS$type == type,]), ncells_max)
      selected_cells <- c(selected_cells, tmp)
    }
    df_col_FCS <- df_col_FCS[selected_cells,]
  }

  df_col_FCS <- df_col_FCS[order(df_col_FCS$time),]
  df_col_FCS <- df_col_FCS[order(df_col_FCS$type),]

  gene_type <- CT_genes_types %>% 
    filter(gene %in% genes)

  gene_type <- gene_type %>% arrange(met_3_groups)
  rowSplit <- gene_type$met_3_groups

   if (!clusterRow) {
    gene_type$gene <- factor(gene_type$gene, levels = genes)
    gene_type <- gene_type  %>%
      arrange(gene)
    gene_type$gene <- as.vector(gene_type$gene)
    rowSplit <- gene_type$met_3_groups
   }

  mat <- as.matrix(
    assay(fetal_sce, "logcounts")[gene_type$gene, rownames(df_col_FCS)])

  title <- "Fetal germ cells"
  if (all(sex == "F")) {
    title <- "Female fetal germ cells"
  }
  if (all(sex == "M")) {
    title <- "Male fetal germ cells"
  }

  if (is.null(scale_lims)) {
    scale_lims <- c(0, quantile(rowMaxs(mat), 0.8))
  }

  heatmap_colors <- colorRamp2(
    seq(scale_lims[1], scale_lims[2], length = 11),                 
    legend_colors)

  ht_opt$TITLE_PADDING = unit(c(2.5, 5), "points")

  ht_FGCs <- Heatmap(mat,
                     name = "Logcounts",
                     left_annotation = rowAnnotation(
                       met = anno_block(
                         gp = gpar(fill = c("indianred2", "darksalmon", MethDep_colors[2]),
                                   col = "black", lwd = 0.2),
                         width = unit(c(0.5), "cm"),
                         labels = c("X-linked\nMethDep", 
                                    "Not X\nMethDep", 
                                    "Non\nMethDep"), 
                         labels_gp = gpar(col = "white", fontsize = 5))),
                     #column_title = title,
                     column_split = df_col_FCS$type,
                     cluster_row_slices = FALSE,
                     cluster_column_slices = FALSE,
                     col = heatmap_colors,
                     border = TRUE,
                     border_gp = gpar(lwd = 0.2),
                     row_gap = unit(c(2), "mm"),
                     column_gap = unit(c(0.5), "mm"),
                     show_column_dend = FALSE,
                     show_row_dend = FALSE,
                     row_names_gp = gpar(fontsize = font_size),
                     row_names_side = "left",
                     show_column_names = FALSE,
                     cluster_rows = clusterRow,
                     cluster_columns = FALSE,
                     column_title_rot = column_name_rot,
                     column_title_gp = gpar(fontsize = title_fs),
                     show_row_names = TRUE,
                     row_split = rowSplit,
                     row_title_gp = gpar(fontsize = 4),
                     row_title_rot = 90,
                     gap = unit(0.01, "mm"),
                     use_raster = TRUE,
                     raster_device = "CairoPNG",
                     width = unit(h_width, "cm"),
                     height = unit(h_height, "cm"),
                     raster_quality = 10,
                     show_heatmap_legend = show_legend,
                     heatmap_legend_param = legends_param)

  return(ht_FGCs)
}

```

```{r}
#######################################################################
## Function to draw expression in adult testis cells with specific annotations
## and downasampling the number of cells from each cell type
#######################################################################
draw_testis_heatmap <- function(genes, 
                                gene_to_label = NULL,
                                label_fonsize = 6,
                                sex = c("F", "M"),
                                ncells_max = NULL,
                                font_size = 6,
                                title_fs = 6,
                                scale_lims = NULL,
                                clusterRow = FALSE,
                                show_legend = TRUE,
                                h_width = 5,
                                h_height = 10,
                                clust_method = "ward.D"){

 # testis_adult_sce <- testis_sce[, testis_sce$stage != "somatic"]
  testis_adult_sce <- testis_sce
  testis_adult_sce$Stage <- "Adult"
  testis_adult_sce$type <- factor(testis_adult_sce$type, levels = germcells)

  df_col_testis <- data.frame(type = testis_adult_sce$type,
                              donor = testis_adult_sce$Donor,
                              Stage = testis_adult_sce$Stage)
  levels(df_col_testis$type) <- c("Spg\nstem", "Spg", "Spc\nearly", "Spc\nlate",
                                  "Spd\nround", "Spd\nelong", "Sperm")

  df_col_testis$sex <- "M"

  rownames(df_col_testis) <- colnames(testis_adult_sce)
  df_col_testis <- df_col_testis[order(df_col_testis$type),]

  #Randomly pick ncells_max cells of each cell type
  if (!is.null(ncells_max)) {
    set.seed(123)
    selected_cells <- sample_cells(
      rownames(df_col_testis[df_col_testis$type == 
                               levels(df_col_testis$type)[1],]), ncells_max)
    for (type in levels(df_col_testis$type)[-1]){
      tmp <- sample_cells(
        rownames(df_col_testis[df_col_testis$type == type,]), ncells_max)
      selected_cells <- c(selected_cells, tmp)
    }
    df_col_testis <- df_col_testis[selected_cells,]
  }

  gene_type <- CT_genes_types %>% 
    filter(gene %in% genes)

  gene_type <- gene_type %>% arrange(met_3_groups)
  rowSplit <- gene_type$met_3_groups

  if (!clusterRow) {
    gene_type$gene <- factor(gene_type$gene, levels = genes)
    gene_type <- gene_type  %>%
      arrange(gene)
    gene_type$gene <- as.vector(gene_type$gene)
    rowSplit <- gene_type$met_3_groups
  }

    mat <-  as.matrix(
    assay(testis_adult_sce, "logcounts")[gene_type$gene, rownames(df_col_testis)])

  if (is.null(scale_lims)) {
    scale_lims <- c(0, quantile(rowMaxs(mat), 0.8))
  }

  heatmap_colors <- colorRamp2(
    seq(scale_lims[1], scale_lims[2], length = 11),                 
    legend_colors)

   rAnnot <- NULL

  ha <- NULL
  if (!is.null(gene_to_label)){
    ind_to_label <- which(rownames(mat) %in% gene_to_label)
    ha <- rowAnnotation(foo = anno_mark(at = ind_to_label, 
                                        labels = rownames(mat[ind_to_label,]),
                                        labels_gp = gpar(fontsize = label_fonsize)))
  }

  ht_opt$TITLE_PADDING = unit(c(2.5, 5), "points")
  ht_testis <- Heatmap(mat,
                       name = "logcounts",
                       column_split = df_col_testis$type,
                       column_title_rot = 90,
                       column_title_gp = gpar(fontsize = title_fs),
                       row_title = NULL,
                       col = heatmap_colors,
                       border = TRUE,
                       border_gp = gpar(lwd = 0.2),
                       row_gap = unit(c(2), "mm"),
                       column_gap = unit(c(0.5), "mm"),
                       show_column_dend = FALSE,
                       show_row_dend = FALSE,
                       row_names_gp = gpar(fontsize = font_size),
                       row_names_side = "right",
                       cluster_rows = clusterRow,
                       clustering_method_rows = clust_method,
                       row_dend_reorder = TRUE,
                       cluster_columns = FALSE,
                       show_column_names = FALSE,
                       show_row_names = FALSE,
                       row_split = rowSplit,
                       cluster_row_slices = FALSE,
                       cluster_column_slices = FALSE,
                       right_annotation = ha,
                       row_title_gp = gpar(fontsize = 8),
                       gap = unit(0.01, "mm"),
                       use_raster = TRUE,
                       raster_device = "CairoPNG",
                       raster_quality = 10,
                       width = unit(h_width, "cm"),
                       height = unit(h_height, "cm"),
                       show_heatmap_legend =  show_legend,
                       heatmap_legend_param = legends_param)

  return(ht_testis)
}
```

```{r}
#######################################################################
## Function to draw expression in adult oocytes cells with specific annotations
## and downasampling the number of cells from each cell type
#######################################################################
draw_oocytes_heatmap <- function(genes, 
                                 gene_to_label = NULL,
                                 ncells_max = NULL,
                                 font_size = 6,
                                 label_fonsize = 6, 
                                 scale_lims = NULL,
                                 clusterRow = TRUE,
                                 show_legend = TRUE,
                                 title_fs = 6,
                                 h_width = 5,
                                 h_height = 10,
                                 column_name_rot = 90,
                                clust_method = "ward.D",
                                show_right_annot = TRUE,
                                show_top_annot = TRUE,
                                top_annot = "Stage"){

  adult_oocytes_sce <- oocytes_sce[, oocytes_sce$type %in% c("Growing oocytes", 
                                                             # "Fully grown oocytes",  
                                                             # "Metaphase I", 
                                                             "Metaphase II")]

  adult_oocytes_sce$Stage <- "Adult"
  levels(adult_oocytes_sce$type) <- c("Growing\noocyte", 
                                      "Fully grown\noocyte",  
                                      "MI\noocyte", 
                                      "MII\noocyte")

  df_col_oocytes <- data.frame(type = adult_oocytes_sce$type,
                               sex = adult_oocytes_sce$sex,
                               stage = adult_oocytes_sce$stage,
                               Stage = adult_oocytes_sce$Stage)

  rownames(df_col_oocytes) <- colnames(adult_oocytes_sce)
  df_col_oocytes <- df_col_oocytes[order(df_col_oocytes$type),]

  #Randomly pick ncells_max cells of each cell type
  if (!is.null(ncells_max)) {
    set.seed(123)
    selected_cells <- sample_cells(
      rownames(df_col_oocytes[df_col_oocytes$type == 
                                levels(df_col_oocytes$type)[1],]), ncells_max)
    for (type in levels(df_col_oocytes$type)[-1]){
      tmp <- sample_cells(
        rownames(df_col_oocytes[df_col_oocytes$type == type,]), ncells_max)
      selected_cells <- c(selected_cells, tmp)
    }
    df_col_oocytes <- df_col_oocytes[selected_cells,]
  }

  gene_type <- CT_genes_types %>% 
    filter(gene %in% genes)

  gene_type <- gene_type %>% arrange(met_3_groups)
  rowSplit <- gene_type$met_3_groups

  if (!clusterRow) {
    gene_type$gene <- factor(gene_type$gene, levels = genes)
    gene_type <- gene_type  %>%
      arrange(gene)
    gene_type$gene <- as.vector(gene_type$gene)
    rowSplit <- gene_type$met_3_groups
  }

  mat <-  as.matrix(
    assay(adult_oocytes_sce, "logcounts")[gene_type$gene, rownames(df_col_oocytes)])

  if (is.null(scale_lims)) {
    scale_lims <- c(0, quantile(rowMaxs(mat), 0.8))
  }

  heatmap_colors <- colorRamp2(
    seq(scale_lims[1], scale_lims[2], length = 11),                 
    legend_colors)

  ha <- NULL
  if (!is.null(gene_to_label)){
    ind_to_label <- which(rownames(mat) %in% gene_to_label)
    ha <- rowAnnotation(foo = anno_mark(at = ind_to_label, 
                                        labels = rownames(mat[ind_to_label,]),
                                        labels_gp = gpar(fontsize = label_fonsize)))
  }

  ht_oocytes <- Heatmap(mat,
                        name = "logcounts",
                        column_split = df_col_oocytes$type,
                        column_title_rot =  column_name_rot,
                        column_title_gp = gpar(fontsize = title_fs),
                        col = heatmap_colors,
                        border = TRUE,
                        border_gp = gpar(lwd = 0.2),
                        row_gap = unit(c(2), "mm"),
                        column_gap = unit(c(0.5), "mm"),
                        show_column_dend = FALSE,
                        show_row_dend = FALSE,
                        row_names_gp = gpar(fontsize = font_size),
                        row_names_side = "left",
                        cluster_rows = clusterRow,
                        clustering_method_rows = clust_method,
                        row_dend_reorder = TRUE,
                        cluster_columns = FALSE,
                        show_column_names = FALSE,
                        show_row_names = TRUE,
                        row_split = rowSplit,
                        cluster_row_slices = FALSE,
                        cluster_column_slices = FALSE,
                        right_annotation = ha,
                        row_title_gp = gpar(fontsize = 0),
                        gap = unit(0.01, "mm"),
                        use_raster = TRUE,
                        raster_device = "CairoPNG",
                        raster_quality = 10,
                        width = unit(h_width, "cm"),
                        height = unit(h_height, "cm"),
                        show_heatmap_legend = show_legend,
                        heatmap_legend_param = legends_param)

  return(ht_oocytes)
}
```

```{r}
# Manual adjustements

X_met_order <- c(
  "SPANXB1", "SPANXD", "SPANXC", "DDX53", "TGIF2LX", "CXorf51B",
  "VCX3B", "VCX", "VCX3A", "VCX2", "FMR1NB",  "PAGE1",
  "PAGE2", "PAGE5", "MAGEA4",  "MAGEC2", "GAGE2A",
  "MAGEB2", "GAGE1", "CTAG2", "PASD1", "MAGEA3", "MAGEA6", "CT45A10",
  "CT55", "MAGEA1", "SSX1", "MAGEA10", "MAGEA11", "MAGEA9B", "MAGEB6",
  "MAGEC1", "MAGEA12", "MAGEB1", "MAGEB16",  "FTHL17",
  "MAGEA8", "TEX13C", "GAGE13", "GAGE12E", "XAGE5", "SAGE1", 
  "CT45A1", "SSX5", "SSX2B", "SSX2", "XAGE1A", "XAGE1B", "DCAF8L2")

not_X_met_order <- c(
  "FLJ36000", "SUN3", "LINC02475", "TUBA3C", "DSCR8", "BRDT", "COX7B2", "RBM46", 
  "DCAF4L2", "CTCFL", "RNF17", "POTEC", "NLRP4","RFPL4B","LINC00470", 
  "TMEM132D-AS1", "PRDM9", "LINC01413", "DPPA2", "TRIML2", "LINC01518", 
  "LINC00200", "POTEF")

not_met_order <- c( 
  "LINC01193", "CSTL1", "GARIN2", "WDR64","CBY2", "FAM217A", "CDRT15", "C1orf94",
  "LINC01249", "RAD21L1", "CHCT1", "SYCP1","LINC01206","LIN28A","MEIOB", "MAJIN", 
  "PRDM7","DMRT1", "ESX1", "TSPY2", "LINC01511", "ZNF729", "TSPY3", "LINC01098")

my_genes <- c(X_met_order, not_X_met_order, not_met_order)

```

## Fig 6A: Male germcells expression (fetal + adult)

```{r}
fs <- 0
ncells <- 50
set.seed(123)
scale_limits <- c(0, 5)
label_fs <- 6

ht_testis <- draw_testis_heatmap(my_genes, 
                                 gene_to_label = c("SPANXB1", "VCX3B", "MAGEC2", 
                                                   "CTAG2", "MAGEA3", "SSX1", 
                                                   "DSCR8", "CTCFL", "DPPA2", 
                                                   "CBY2", "SYCP1", "DMRT1"),
                                 ncells_max = ncells,
                                 font_size = fs,
                                 scale_lims = scale_limits, 
                                 title_fs = 6,
                                 h_width = 5,
                                 h_height = 6,
                                 clusterRow  = FALSE,
                                 show_legend = TRUE,
                                 label_fonsize = label_fs, 
                                 clust_method = "centroid")

male_FGC_ht <- draw_FGC_heatmap(my_genes, 
                                ncells_max = ncells,
                                sex = c("M"),
                                clusterRow = FALSE, 
                                h_width = 2,
                                h_height = 6,
                                font_size = fs,
                                title_fs = 6,
                                scale_lims = scale_limits, 
                                show_legend = FALSE)

ht_list <- male_FGC_ht + ht_testis

draw(ht_list, main_heatmap = "logcounts",
     merge_legend = TRUE)

pdf(file = paste0("../figs/Fig6A_Expression_in_MALE_FGC_and_adult_testis.pdf"))
draw(ht_list, main_heatmap = "logcounts",
     merge_legend = TRUE, ht_gap = unit(0.1, "mm"))
dev.off()
```

## Fig 7A: Female germcells Expression (fetal + adult)

```{r}
fs <- 0
ncells <- 50
set.seed(123)

to_label <- c("DDX53", "TGIF2LX", #"VCX", 
              "PAGE2", "PAGE5", 
              "MAGEB2", "MAGEB1", "MAGEC1", "FTHL17", #"SAGE1",
              "DSCR8", "BRDT", "RBM46", "DCAF4L2", "CTCFL", "RNF17", 
              "NLRP4", "LINC01413", "FAM217A", "LIN28A", "MEIOB", "MAJIN", "LINC01511")
to_label[!to_label%in% CT_genes$external_gene_name]
ht_oocytes <- draw_oocytes_heatmap(my_genes, 
                                   gene_to_label = to_label,
                                   label_fonsize = label_fs, 
                                 ncells_max = ncells,
                                 font_size = fs,
                                 scale_lims = scale_limits, 
                                 title_fs = 6,
                                 h_width = 2,
                                 h_height = 6,
                                 column_name_rot = 0,
                                 clusterRow  = FALSE,
                                 show_legend = TRUE,
                                 clust_method = "centroid",
                                 show_right_annot = FALSE,
                                 show_top_annot = FALSE,
                                 top_annot = "Stage")

female_FGC_ht <- draw_FGC_heatmap(my_genes, 
                                ncells_max = ncells,
                                sex = c("F"),
                                clusterRow = FALSE, 
                                h_width = 4,
                                h_height = 6,
                                column_name_rot = 0,
                                font_size = fs,
                                title_fs = 6,
                                scale_lims = scale_limits, 
                                show_legend = FALSE)

ht_list <- female_FGC_ht + ht_oocytes

draw(ht_list, main_heatmap = "logcounts",
    merge_legend = TRUE)

pdf(file = paste0("../figs/Fig7A_Expression_in_FEMALE_germcells.pdf"))
#, width = 6, height = 4)
draw(ht_list, main_heatmap = "logcounts",
    merge_legend = TRUE, ht_gap = unit(0.1, "mm"))
dev.off()
```

# Fig 6B: Repression of X-linked CG genes coincides with MSCI

- Calculated mean expression in each cell type

- Removed genes never expressed in pre-meiotic germ cells

- Calculated mean expression by chr

    - for Autosomial genes (not CG genes), 
    - X-linked (not CG genes), 
    - X-linked CG genes

```{r}
testis_sce <- CTdata::testis_sce()

levels(testis_sce$type)[c(7,8)] <- "Sperm"
levels(testis_sce$type) <- c("Spermatogonial\nstem cells", "Spermatogonia",
                             "Early\nspermatocytes", "Late\nspermatocytes",
                             "Round\nspermatids", "Elongated\nspermatids",
                             "Sperm", "Macrophage", "Endothelial", "Myoid",
                             "Sertoli", "Leydig" )
somatic <- c("Macrophage", "Endothelial", "Myoid", "Sertoli", "Leydig")
testis_sce <- testis_sce[, !testis_sce$type %in% somatic]
testis_sce$type <- droplevels(testis_sce$type)

# Mean expression level per gene and per cell type
mean_gene_counts_by_type <- as_tibble(assay(testis_sce, "logcounts"), 
                                      rownames = "gene") %>% 
  pivot_longer(names_to = "cell", values_to = "count", -gene) %>% 
  left_join(as_tibble(colData(testis_sce), rownames = "cell") %>% 
              dplyr::select(cell, type)) %>% 
  group_by(gene, type) %>% 
  summarise(mean_count = mean(count)) %>% 
  left_join(all_genes %>% 
              dplyr::select(external_gene_name, chr) %>% 
              dplyr::rename(gene = external_gene_name))

pre_meiotic <- c("Spermatogonial\nstem cells", "Spermatogonia", 
                 "Early\nspermatocytes")

# To remove later on genes never detected in premeiotic cells
genes_never_expressed_in_premeiotic <- mean_gene_counts_by_type %>% 
  filter(type %in% pre_meiotic) %>% 
  group_by(gene) %>% 
  mutate(mean_count_in_premeiotic = mean(mean_count)) %>% 
  filter(mean_count_in_premeiotic == 0) %>% 
  pull(gene)

Xmet <- CT_genes %>% 
  filter(regulated_by_methylation) %>% 
  filter(X_linked) %>% 
  pull(external_gene_name)

# Mean expression level per chromosome and per cell type
# Removed CT genes MethDep from the genes
# Removed genes never_expressed_in_premeiotic_cells
mean_gene_counts_by_chromosome_and_by_type <- mean_gene_counts_by_type %>% 
  filter(!gene %in% Xmet) %>% 
  filter(!gene %in% genes_never_expressed_in_premeiotic) %>% 
  filter(chr %in% c(1:22, "X")) %>% 
  mutate(group = case_when(chr %in% 1:22 ~ "Autosomal genes",
                           chr == "X" ~ "X-linked genes")) %>% 
  group_by(group, type) %>% 
  summarize(gene_expression_mean = mean(mean_count)) %>% 
  mutate(type = factor(type, levels = levels(mean_gene_counts_by_type$type)))

# Mean expression level of CTgenes MethDep per cell type
# Removed genes never_expressed_in_premeiotic_cells
mean_CT_counts_by_chromosome_and_by_type <- mean_gene_counts_by_type %>% 
  filter(gene %in% Xmet) %>% 
  filter(!gene %in% genes_never_expressed_in_premeiotic) %>% 
  mutate(group = 'X-linked CG MethDep') %>% 
  group_by(group, type) %>% 
  summarize(gene_expression_mean = mean(mean_count)) %>% 
  mutate(type = factor(type, levels = levels(mean_gene_counts_by_type$type)))

levels(mean_gene_counts_by_chromosome_and_by_type$type) <- 
  c("Spg stem", "Spg", "Spc early", "Spc late", "Spd round", "Spd elong", "Sperm")
levels(mean_CT_counts_by_chromosome_and_by_type$type) <- 
  c("Spg stem", "Spg", "Spc early", "Spc late", "Spd round", "Spd elong", "Sperm")

fig <- rbind(mean_gene_counts_by_chromosome_and_by_type, 
             mean_CT_counts_by_chromosome_and_by_type) %>% 
  ggplot(aes(x = type, y = gene_expression_mean, group = group)) + #, color = group)) +
  geom_vline(xintercept = 3, color = "red", linetype = "dashed")+
  geom_point() + geom_line() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
        axis.title.x = element_text(size = 0),
        legend.position = "none", 
        panel.grid = element_blank(),
        aspect.ratio = 0.8,
        strip.text.x = element_text(size = 7)) +
  xlab("Spermatogenesis stage") +
  ylab("Gene mean expression") +
  facet_wrap(~ group, scales = "free_y", ncol = 1)

fig

pdf(file = paste0("../figs/Fig6B_Repression_of_MethDep_and_MSCI.pdf"), width = 2.5, height = 4)
fig
dev.off()
```

```{r}
# Number of genes in each group:
x <- mean_gene_counts_by_type %>% 
  filter(!gene %in% Xmet) %>% 
  filter(!gene %in% genes_never_expressed_in_premeiotic) %>% 
  filter(chr %in% c(1:22, "X")) %>% 
  mutate(group = case_when(chr %in% 1:22 ~ "Autosomal genes",
                                chr == "X" ~ "X-linked genes")) %>% 
  dplyr::select(gene, group) %>% unique()

y <- mean_gene_counts_by_type %>% 
  filter(gene %in% Xmet) %>% 
  mutate(group = 'X-linked CG MethDep') %>% 
  dplyr::select(gene, group) %>% unique()

n_gene_per_group <- rbind(x, y)

flextable(enframe(table(n_gene_per_group$group), 
                  name = "Group", value = "number of gene"), 
                     cwidth = 1.5,
                     cheight = 0.25) %>%
  set_caption(as_paragraph(
    as_chunk("Number of analysed genes", 
             props = fp_text_default(font.family = "Arial",
                                                 font.size = 10))), 
    word_stylename = "Table Caption",
    align_with_table = TRUE
  )
```

# Fig 6D: CG expression and methylation in Male FGC

## Expression

Proportions of positive cells

=> proportion of count values in each group that are non-zero

```{r}
my_genes <- CT_genes$external_gene_name[CT_genes$external_gene_name %in% rownames(FGC_sce)]

dds <- scuttle::summarizeAssayByGroup(FGC_sce, 
                                      assay.type = "logcounts",
                            ids = FGC_sce$type,
                            statistics = "prop.detected")

dds$type <- dds$ids
rownames(colData(dds)) <- dds$ids
colnames(dds) <- dds$ids
dds$sex <- c(rep("F", 4), rep("M", 3))

bilan_exp <- as_tibble(assay(dds[my_genes, ]), rownames = "gene") %>% 
  pivot_longer(names_to = "type_sex", values_to = "proportion_of_pos", -gene) %>% 
  left_join(CT_genes %>% 
              dplyr::select(external_gene_name, regulated_by_methylation, X_linked) %>% 
              dplyr::rename(gene = external_gene_name)) %>% 
  filter(!is.na(regulated_by_methylation)) %>% 
  mutate(CT_group = case_when(regulated_by_methylation & X_linked ~ "X-linked MethDep",
                              regulated_by_methylation & !X_linked ~ "Not X MethDep",
                              !regulated_by_methylation ~ "Non-MethDep")) %>% 
   mutate(CT_group = factor(CT_group, 
                            levels = c("X-linked MethDep", 
                                       "Not X MethDep",
                                       "Non-MethDep"))) %>% 
  mutate(type_sex = factor(type_sex, levels = levels(FGC_sce$type) )) %>% 
  left_join(as_tibble(colData(dds)) %>% dplyr::rename(type_sex = type))

levels(bilan_exp$type_sex) <- c("PGC", "FGC", "Oogonia", "Oocyte", "PGC", "FGC", "Pre-SpG")
bilan_exp$sex <- factor(bilan_exp$sex)
levels(bilan_exp$sex) <- c("Female FGCs", "Male FGCs" )

library(ggh4x)
strip <- strip_themed(background_y = elem_list_rect(fill = CT_group_colors))

my_colors <- CT_group_colors
names(my_colors) <- c("X-linked\nMethDep", "Not X\nMethDep","Non-MethDep")

fig_expression_in_Male_FGC <- bilan_exp %>% 
  filter(sex == "Male FGCs") %>% 
  mutate(sex = "scRNAseq\nMale gonad") %>% 
  filter(CT_group != "Non-MethDep") %>% 
  mutate(CT_group = case_when(CT_group == "X-linked MethDep" ~ "X-linked\nMethDep",
                              CT_group == "Not X MethDep" ~ "Not X\nMethDep")) %>%
  mutate(CT_group = factor(CT_group, levels =  c("X-linked\nMethDep", "Not X\nMethDep"))) %>%
  ggplot(aes(x = type_sex, y = proportion_of_pos)) +
  scale_color_manual(values = CT_group_colors) +
  geom_boxplot(data = bilan_exp %>% 
                 filter(sex == "Male FGCs") %>% 
                 mutate(sex = "scRNAseq\nMale gonad") %>% 
                 filter(CT_group != "Non-MethDep") %>% 
                 mutate(CT_group = case_when(CT_group == "X-linked MethDep" ~ "X-linked\nMethDep",
                                             CT_group == "Not X MethDep" ~ "Not X\nMethDep")) %>%
                 mutate(CT_group = factor(CT_group, levels =  c("X-linked\nMethDep", "Not X\nMethDep"))) ,
               aes(x = type_sex, y = proportion_of_pos, fill = CT_group), 
               alpha = 0.5,
               outliers = F, 
               varwidth = TRUE,
               linewidth = 0.3) +
  scale_fill_manual(values = my_colors) +
  facet_grid2(CT_group ~ sex, drop = T, 
              switch = "y", 
              scales = "free_x", 
              strip = strip) +#, space = "free") +#, switch = "y") +
  theme_bw() +
  theme(aspect.ratio = 1.5,
        axis.text.x = element_text(angle = 0, 
                                   size = 8,
                                   vjust = 0.5,  
                                   hjust = 0.5,
                                   face = "bold"),
        axis.title.y = element_text(size = 12),
        axis.title.x = element_text(size = 0),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.position = "none",
        strip.text.x = element_text(size = 10),
        strip.text.y = element_text(size = 10, color = "white", face = "bold"),
        strip.placement.y = "inside",
        strip.background.y = element_rect(fill=CT_group_colors[2]),
        strip.background.x = element_blank() ) +
  ylab("Proportions of positive cells") +
  xlab("Fetal cell stages")

bilan_exp %>% 
  filter(sex == "Male FGCs") %>% 
  mutate(sex = "scRNAseq\nMale gonad") %>% 
  filter(CT_group != "Non-MethDep") %>% 
  mutate(CT_group = case_when(CT_group == "X-linked MethDep" ~ "X-linked\nMethDep",
                              CT_group == "Not X MethDep" ~ "Not X\nMethDep")) %>%
  mutate(CT_group = factor(CT_group, levels =  c("X-linked\nMethDep", "Not X\nMethDep"))) %>% 
  dplyr::select(gene, CT_group) %>% 
  unique() %>% 
  pull(CT_group) %>% 
  table()
```

## Methylation

```{r}
mean_methylation_in_FGC <- CTdata::mean_methylation_in_FGC()

bilan <- as_tibble(assay(mean_methylation_in_FGC), rownames = "gene") %>% 
  filter(gene %in% CT_genes$external_gene_name) %>% 
  pivot_longer(names_to = "sample", values_to = "met", -gene) %>% 
  left_join(CT_genes %>% 
              dplyr::select(external_gene_name, regulated_by_methylation, X_linked) %>% 
              dplyr::rename(gene = external_gene_name)) %>% 
  filter(!is.na(regulated_by_methylation)) %>% 
  mutate(CT_group = case_when(regulated_by_methylation & X_linked ~ "X-linked MethDep",
                              regulated_by_methylation & !X_linked ~ "Not X MethDep",
                              !regulated_by_methylation ~ "Non-MethDep")) %>% 
  mutate(CT_group = factor(CT_group, 
                           levels = c("X-linked MethDep", 
                                      "Not X MethDep", 
                                      "Non-MethDep"))) %>% 
  left_join(as_tibble(colData(mean_methylation_in_FGC)) %>% 
              dplyr::select(sample, group, type, time_week, sex, sex_time, sex_type, type) %>% 
              dplyr::rename(FGC_group = group)) %>% 
  #filter(gene == "DCAF8L2") %>% 
  group_by(CT_group, gene, type, time_week, sex) %>%
  summarise(mean_met = mean(met, na.rm = TRUE),
            met_sd = sd(met, na.rm = TRUE),
            median_met = median(met, na.rm = TRUE)) %>%
  mutate(type_sex = case_when(type == "FGC" & sex == "F" ~ "Female FGC",
                              type == "FGC" & sex == "M" ~ "Male FGC",
                              type == "Soma"& sex == "F" ~ "Somatic cells",
                              type == "Soma"& sex == "M" ~ "Somatic cells")) %>% 
  mutate(sex = case_when(sex == "F" ~ "Female",
                         sex == "M" ~ "Male")) %>% 
  mutate(Group = paste0(CT_group, type)) %>% 
  mutate(Group = factor(Group, levels = c("X-linked MethDepFGC",
                                          "X-linked MethDepSoma",
                                          "Not X MethDepFGC",
                                          "Not X MethDepSoma", 
                                          "Non-MethDepFGC", 
                                          "Non-MethDepSoma"))) %>%
  mutate(Type = case_when(type == "Soma" ~ "Somatic cells",
                          type == "FGC" ~ "FGC"))

library(ggh4x)
strip <- strip_themed(background_y = elem_list_rect(fill = CT_group_colors))

fig_met_in_Male_FGC <- bilan %>% 
  filter(sex == "Male") %>%  
  mutate(type_sex = case_when(type_sex == "Male FGC" ~"scBSseq\nMale gonad",
                              type_sex == "Somatic cells" ~"Soma")) %>% 
  filter(CT_group != "Non-MethDep") %>% 
  mutate(CT_group = case_when(CT_group == "X-linked MethDep" ~ "X-linked\nMethDep",
                           CT_group == "Not X MethDep" ~ "Not X\nMethDep")) %>%
  mutate(CT_group = factor(CT_group, levels =  c("X-linked\nMethDep", "Not X\nMethDep"))) %>%
  mutate(time_week = factor(time_week, 
                            levels = c("6", "7", "8", "9", "10", "17", "21", "24"))) %>% 
  ggplot(aes(x = time_week, y = mean_met, color = CT_group, group = time_week)) +
  #geom_jitter(size = .2, alpha = 1) + 
  geom_boxplot(data = bilan %>% 
  filter(sex == "Male") %>%  
  mutate(type_sex = case_when(type_sex == "Male FGC" ~"scBSseq\nMale gonad",
                              type_sex == "Somatic cells" ~"Soma")) %>% 
  filter(CT_group != "Non-MethDep") %>% 
  mutate(CT_group = case_when(CT_group == "X-linked MethDep" ~ "X-linked\nMethDep",
                           CT_group == "Not X MethDep" ~ "Not X\nMethDep")) %>%
  mutate(CT_group = factor(CT_group, levels =  c("X-linked\nMethDep", "Not X\nMethDep"))) %>%
  mutate(time_week = factor(time_week, 
                            levels = c("6", "7", "8", "9", "10", "17", "21", "24"))),
  aes(x = time_week, y = mean_met, color = CT_group, group = time_week, fill = CT_group),
               linewidth = 0.3,
               alpha = 0.5, color = "black",
               outliers = F) +
  scale_color_manual(values = CT_group_colors) +
  scale_fill_manual(values = my_colors) +
  facet_grid2(CT_group ~ type_sex,  
              scales = "free_x", 
              space = "free",
             drop = TRUE, 
             #strip = strip,
             switch = "y"
             ) +
  theme_bw() +
  theme(#aspect.ratio = 2,
        axis.text.x = element_text(angle = 0, 
                                   size = 8,
                                   vjust = 0.5,  
                                   hjust = 0.5,
                                   face = "bold"),
        axis.title.y = element_text(size = 12),
        axis.title.x = element_text(size = 0),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.position = "none",
        strip.placement.y = "right",
        strip.text.x = element_text(size = 10),
        strip.text.y = element_text(size = 10, color = "white", face = "bold"),
        strip.background = element_blank(),
        strip.x = element_blank()) +
  scale_y_continuous(position = "right")+
  #scale_y_log10() +
    ylab("Promoter mean methylation") +
  xlab("Time (week)")
fig_met_in_Male_FGC
```

```{r}
bilan %>% 
  filter(CT_group == "X-linked MethDep", sex == "Male", type == "FGC") %>% 
  mutate(timepoints = case_when(time_week <= 7 ~ "6-7W",
                                time_week >= 17 ~ ">17W")) %>% 
  ungroup() %>% 
  group_by(timepoints) %>% 
  dplyr::summarise("methylation (mean)" = mean(mean_met, na.rm = TRUE),
                   "methylation (median)" = median(mean_met, na.rm = TRUE)) %>% 
  DT::datatable()

```

## Fig 6D: Exp + Meth

```{r}
gridExtra::grid.arrange(fig_expression_in_Male_FGC, fig_met_in_Male_FGC, ncol = 2)

pdf(file = "../figs/Fig6D_Expression_and_Met_in_Male_FGC.pdf", width = 5, height = 3)
gridExtra::grid.arrange(fig_expression_in_Male_FGC, fig_met_in_Male_FGC, ncol = 2)
dev.off()
```

# Fig 6F: expression and methylation in Male FGC: Selected examples

## Initial gene selection

```{r}
X_MethDep_genes <- c("SSX1", "VCX", "PAGE2", "MAGEC2", "MAGEB16")
notX_MethDep_genes <- c("RNF17", "CTCFL")
#not_MethDep_genes <- c("DMRT1", "SYCP1")
selected_genes <- c(X_MethDep_genes, notX_MethDep_genes)

strip <- strip_themed(background_y = elem_list_rect(fill = c(rep("indianred2", 5), rep(CT_group_colors[2], 2))))

bilan_exp_selected <-bilan_exp %>% 
  filter(sex == "Male FGCs") %>% 
  filter(gene %in% selected_genes) %>%
  mutate(gene = factor(gene, levels = selected_genes)) %>% 
  left_join(CT_genes %>% 
              dplyr::select(external_gene_name, regulated_by_methylation, X_linked) %>% 
              dplyr::rename(gene = external_gene_name)) %>% 
  filter(!is.na(regulated_by_methylation)) %>% 
  mutate(CT_group = case_when(regulated_by_methylation & X_linked ~ "X-linked MethDep",
                              regulated_by_methylation & !X_linked ~ "Not X MethDep",
                              !regulated_by_methylation ~ "Non-MethDep"))
bilan_exp_selected$gene <- factor(bilan_exp_selected$gene, levels = selected_genes)

fig_exp <- bilan_exp_selected %>% 
    ggplot(aes(x = type_sex, y = proportion_of_pos, group = gene , color = CT_group)) +
    geom_point()+
    geom_line() +
  scale_color_manual(values = CT_group_colors)+
  facet_grid2( gene ~ ., strip = strip, switch = "y") +
  theme_bw() +
  theme(aspect.ratio = .5,
        axis.title.x = element_blank(),
        axis.text.x = element_text(size = 8, face = "bold"),
        axis.title.y = element_text(size = 12),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        strip.text.y = element_text(size = 7, face = "bold"),
        strip.placement = "outside") +
  ylab("Proportion of positive cells")

bilan_met_selected <- bilan %>% 
  ungroup() %>% 
  filter(gene %in% selected_genes) %>%
  filter(type_sex == "Male FGC") %>% 
  mutate(time_week = factor(time_week))

bilan_met_selected$gene <- factor(bilan_met_selected$gene, levels = selected_genes)

fig_met <- bilan_met_selected %>% 
  ggplot(aes(x = time_week, y = mean_met, group = gene, color = CT_group)) +
  geom_point()+
    geom_line() +
  #facet_grid2( gene ~ ., strip = strip) +
  facet_grid( gene ~ .) +
  scale_color_manual(values = CT_group_colors)+
  theme_bw() +
  theme(aspect.ratio = .4,
        axis.title.x = element_blank(),
        axis.text.x = element_text(size = 8, face = "bold"),
        axis.title.y = element_text(size = 12),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        strip.text.y = element_text(size = 0, face = "bold"),
        strip.background = element_blank()) +
  ylab("Promoter mean methylation")
```

Expression

```{r}
fig_exp
```

Methylation

```{r}
fig_met
```

```{r}
#gridExtra::grid.arrange(fig_exp, fig_met, ncol = 3)

pdf(file = "../figs/Fig6F_Examples.pdf")#, width = 3, height = 5)
gridExtra::grid.arrange(fig_exp, fig_met, ncol = 2)
dev.off()
```

# Fig 7D: Methylation  in Female FGC

```{r}
met_thr <- 75

Genes_methylated_in_soma <- as_tibble(assay(mean_methylation_in_FGC), rownames = "gene") %>% 
  pivot_longer(names_to = "sample", values_to = "met", -gene) %>% 
  left_join(as_tibble(colData(mean_methylation_in_FGC)) %>% 
              dplyr::select(sample, group, type, time_week, sex, sex_time, sex_type, type) %>% 
              dplyr::rename(FGC_group = group)) %>%
  filter(type == "Soma") %>% 
  mutate(sex = case_when(sex == "F" ~ "Female",
                         sex == "M" ~ "Male")) %>% 
  group_by(gene, type, time_week, sex) %>%
  summarise(mean_met = mean(met, na.rm = TRUE)) %>% 
  ungroup() %>% 
  dplyr::select(gene, sex, mean_met) %>% 
  pivot_wider(names_from = sex, values_from = mean_met) %>% 
  filter(Female >= met_thr, Male >= met_thr) %>% 
  pull(gene)

MethDep <- CT_genes %>% filter(regulated_by_methylation)
bilan <- as_tibble(assay(mean_methylation_in_FGC), rownames = "gene") %>% 
  filter(gene %in% Genes_methylated_in_soma | gene %in% MethDep$external_gene_name) %>% 
  pivot_longer(names_to = "sample", values_to = "met", -gene) %>% 
  left_join(all_genes %>% 
              dplyr::select(external_gene_name, CT_gene_type, regulated_by_methylation, chr, X_linked) %>% 
              dplyr::rename(gene = external_gene_name)) %>% 
  filter(!is.na(regulated_by_methylation)) %>% 
  mutate(CT_group = case_when(CT_gene_type != "CT_gene" & X_linked ~ "X-linked\nNon CG",
                              CT_gene_type != "CT_gene" & chr %in% c(1:22) ~ "Autosome\nNon CG",
                              CT_gene_type == "CT_gene" & X_linked & regulated_by_methylation ~ "X-linked\nMethDep",
                              CT_gene_type == "CT_gene" & !X_linked & regulated_by_methylation ~ "Not X\nMethDep")) %>% 
  filter(!is.na(CT_group)) %>% 
  mutate(CT_group = factor(CT_group)) %>% 
  left_join(as_tibble(colData(mean_methylation_in_FGC)) %>% 
              dplyr::select(sample, group, type, time_week, sex, sex_time, sex_type, type) %>% 
              dplyr::rename(FGC_group = group)) %>%
  group_by(CT_group, gene, type, time_week, sex) %>%
  summarise(mean_met = mean(met, na.rm = TRUE),
            met_sd = sd(met, na.rm = TRUE),
            median_met = median(met, na.rm = TRUE)) %>%
  mutate(type_sex = case_when(type == "FGC" & sex == "F" ~ "Female FGC",
                              type == "FGC" & sex == "M" ~ "Male FGC",
                              type == "Soma"& sex == "F" ~ "Somatic cells",
                              type == "Soma"& sex == "M" ~ "Somatic cells")) %>% 
  mutate(sex = case_when(sex == "F" ~ "Female",
                         sex == "M" ~ "Male")) %>% 
  mutate(Group = paste0(CT_group, type, " ")) %>% 
  mutate(Group = factor(Group, levels = c("X-linked MethDep FGC",
                                          "X-linked MethDep Soma",
                                          "Not X MethDep FGC",
                                          "Not X MethDep Soma", 
                                          "Not X Non CG FGC", 
                                          "Not X Non CG Soma",
                                          "X-linked Non CGF GC",
                                          "X-linked Non CG Soma")))

bilan$time_type <- paste0(bilan$type, " ", bilan$time_week, "W")
bilan$time_type <- factor(bilan$time_type , levels = c("Soma 7W", "Soma 8W", "FGC 6W", "FGC 7W", "FGC 8W", "FGC 9W", "FGC 10W", "FGC 17W", "FGC 21W", "FGC 24W"))
bilan$CT_group <- factor(bilan$CT_group, levels = c("X-linked\nMethDep", "Not X\nMethDep", "Autosome\nNon CG", "X-linked\nNon CG" ))

bilan %>% ungroup() %>% dplyr::select(CT_group, gene) %>% unique %>% pull(CT_group) %>% table()

library(ggh4x)
strip_group_colors <- c(CT_group_colors[1] , CT_group_colors[2] , "gray40", "gray40")
names(strip_group_colors) <- levels(bilan$CT_group)
strip <- strip_themed(background_x = elem_list_rect(fill = strip_group_colors))

fig <- bilan %>% 
  filter(sex == "Female") %>% 
  ggplot(aes(x = time_type, y = mean_met, fill = CT_group)) +
  geom_boxplot(linewidth = 0.3,
               alpha = 0.5,  color = "black",
               outliers = F) +
  scale_fill_manual(values = strip_group_colors) +
  facet_grid2( ~ CT_group ,  scales = "free_x", space = "free",
             drop = TRUE, strip = strip) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, 
                                   size = 10,
                                   vjust = 0.5,  
                                   hjust = 1),
        axis.title.y = element_text(size = 12),
        axis.title.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.position = "none",
        strip.text.y = element_text(size = 12),
        strip.text.x = element_text(size = 10, color = "white", face = "bold"),
        strip.background = element_blank()) +
    ylab("Mean promoter methylation") +
  xlab("Time (week)")

fig

pdf(file = "../figs/Fig7D_met_in_Female_FGC.pdf", width = 6, height = 5)
fig
dev.off()
```

```{r}
bilan %>% 
  filter(sex == "Female", type == "FGC") %>% 
  ungroup() %>% 
  group_by(CT_group, time_week) %>% 
  dplyr::summarise("methylation (mean)" = mean(mean_met, na.rm = TRUE),
                   "methylation (median)" = median(mean_met, na.rm = TRUE)) %>% 
  DT::datatable()

bilan %>% 
  filter(sex == "Female", type == "FGC") %>% 
  ungroup() %>% 
  group_by(CT_group) %>% 
  dplyr::summarise("methylation (mean)" = mean(mean_met, na.rm = TRUE),
                   "methylation (median)" = median(mean_met, na.rm = TRUE)) %>% 
  DT::datatable()
```

# Fig 7B: Proportion of CT genes expressed in Male and Female germ cells

## Using all datasets  (FGC, oocytes and testis datasets)

- Estimate the proportion of cells in wich each gene is detected.

- Considered as "detected" if the gene is detected in more than 10% of a cell type

```{r}
FGC_sce <- CTdata::FGC_sce()
FGC_sce <- FGC_sce[, FGC_sce$low_ncounts == "False"]
FGC_sce <- FGC_sce[, FGC_sce$high_mito == "False"]

dds_FGC <- scuttle::summarizeAssayByGroup(FGC_sce, 
                            ids = colData(FGC_sce)[, "type"],
                            statistics = "prop.detected")

testis_sce <- CTdata::testis_sce()
levels(testis_sce$type)[c(7,8)] <- "Sperm"
germcells <-  c("SSC", "Spermatogonia", "Early_spermatocyte", 
                "Late_spermatocyte", "Round_spermatid", 
                "Elongated_spermatid", "Sperm")

testis_sce <- testis_sce[, testis_sce$type %in% germcells]
testis_sce$type <- factor(testis_sce$type, levels = germcells)

dds_testis <- scuttle::summarizeAssayByGroup(testis_sce, 
                            ids = colData(testis_sce)[, "type"],
                            statistics = "prop.detected")

oocytes_sce <- CTdata::oocytes_sce()
oocytes_sce <- oocytes_sce[, oocytes_sce$type %in% c("Growing oocytes", "Metaphase II")]
oocytes_sce$type <- droplevels(oocytes_sce$type)

dds_oocytes <- scuttle::summarizeAssayByGroup(oocytes_sce, 
                            ids = colData(oocytes_sce)[, "type"],
                            statistics = "prop.detected")
```

- CG Genes detected in more than 10% of any cell type is considered as expressed in this cell type

- Kept only CG genes detected in at least one cell type of any of the 3 datasets

- Removed Y-linked CG genes from the analysis

```{r}
genes_detected_in_dds_FGC <- names(dds_FGC)[rowMaxs(assay(dds_FGC[,])) > 0.1]

selected_genes <- CT_genes %>% 
  filter(chr != "Y") %>% 
  filter(external_gene_name %in% genes_detected_in_dds_FGC)

bilan_FGC <- tibble(gene = names(rowSums(assay(dds_FGC[selected_genes$external_gene_name, 
                                                   c("F_PGC", "F_GC", "F_oogonia", "F_oocyte")]) > 0.1) > 0),
                detected_in_female_FGC = as.vector(rowSums(assay(dds_FGC[selected_genes$external_gene_name, 
                                                                     c("F_PGC", "F_GC", "F_oogonia", "F_oocyte")]) > 0.1) > 0),
                detected_in_male_FGC = as.vector(rowSums(assay(dds_FGC[selected_genes$external_gene_name, 
                                                                   c("M_PGC", "M_GC", "M_pre_spermatogonia")]) > 0.1) > 0))

```

```{r}
genes_detected_in_dds_testis <- names(dds_testis)[rowMaxs(assay(dds_testis[,], na.rm = TRUE)) > 0.1]

selected_genes <- CT_genes %>% 
  filter(chr != "Y") %>% 
  filter(external_gene_name %in% genes_detected_in_dds_testis)

bilan_testis <- tibble(gene = names(rowSums(assay(dds_testis[selected_genes$external_gene_name, ]) > 0.1) > 0),
                detected_in_male_adult = as.vector(rowSums(assay(dds_testis[selected_genes$external_gene_name, ]) > 0.1) > 0))

```

```{r}
genes_detected_in_dds_oocytes <- names(dds_oocytes)[rowMaxs(assay(dds_oocytes[,], na.rm = TRUE)) > 0.1]

selected_genes <- CT_genes %>% 
  filter(chr != "Y") %>% 
  filter(external_gene_name %in% genes_detected_in_dds_oocytes)

bilan_oocytes <- tibble(gene = names(rowSums(assay(dds_oocytes[selected_genes$external_gene_name, ]) > 0.1) > 0),
                detected_in_Female_oocytes = as.vector(rowSums(assay(dds_oocytes[selected_genes$external_gene_name, ]) > 0.1) > 0))

```

```{r}
bilan <- full_join(bilan_FGC, 
          bilan_testis) %>% 
  full_join(bilan_oocytes)

bilan_by_male_female <- bilan %>% 
rowwise() %>% 
  mutate(detected_in_male = sum(detected_in_male_FGC, detected_in_male_adult, na.rm = TRUE)) %>% 
  mutate(detected_in_female = sum(detected_in_female_FGC, detected_in_Female_oocytes, na.rm = TRUE)) %>% 
  left_join(CT_genes %>% 
              dplyr::select(external_gene_name, chr, regulated_by_methylation, X_linked) %>% 
              dplyr::rename(gene = external_gene_name)) %>% 
  filter(chr!= "Y") %>% 
  mutate(group = case_when(regulated_by_methylation & X_linked ~ "X-linked\nMethDep",
                           regulated_by_methylation & !X_linked ~ "Autosomal\nMethDep",
                           !regulated_by_methylation  ~ "Non\nMethDep"))

bilan_by_group <- bilan_by_male_female %>% 
  group_by(group) %>% 
  summarise("F" = sum(detected_in_female),
            "M" = sum(detected_in_male))

flextable(bilan_by_group, 
                     cwidth = 1.5,
                     cheight = 0.25) %>%
  set_caption(as_paragraph(
    as_chunk("Among genes detected in FGC, testis or oocytes dataset", 
             props = fp_text_default(font.family = "Arial",
                                                 font.size = 10))), 
    word_stylename = "Table Caption",
    align_with_table = TRUE
  )
bilan_by_group$group <- factor(bilan_by_group$group, levels = c("X-linked\nMethDep", "Autosomal\nMethDep", "Non\nMethDep"   ))

names(CT_group_colors) <- c("X-linked\nMethDep", "Autosomal\nMethDep", "Non\nMethDep"   )
strip <- strip_themed(background_x = elem_list_rect(fill = CT_group_colors))

bilan_by_group <- bilan_by_group %>% 
  pivot_longer(names_to = "detected_where", values_to = "Number of detected CG genes", -group)

bilan_by_group$detected_where <- factor(bilan_by_group$detected_where, levels = c("M", "F"   ))

fig <- bilan_by_group %>% 
  ggplot(aes(x = detected_where, y = `Number of detected CG genes`, fill = detected_where)) +
  geom_col() +
  scale_fill_manual(values = sex_colors) +
  facet_grid2(~ group, strip = strip) +
  theme_bw() +
  theme(aspect.ratio = 1.5,
        panel.grid = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 12, face = "bold"),
        axis.text.x = element_text(size = 12, face = "bold"),
        legend.position = "none",
        strip.text.x = element_text(size = 12, color = "white", face = "bold")) +
  ylab("Number of CG\ngenes detected")

fig

pdf(file = "../figs/Fig7B_Genes_detected_in_Male_and_Female.pdf", width = 4, height = 3)
fig
dev.off()

```

```{r}
bilan_by_group$group <- gsub("\n", x = bilan_by_group$group, replacement = " ")
knitr::kable(bilan_by_group[1:2,])

chisq.test(bilan_by_group$`Number of detected CG genes`[1:2], p = c(0.5, 0.5))

knitr::kable(bilan_by_group[3:4,])
chisq.test(bilan_by_group$`Number of detected CG genes`[3:4], p = c(0.5, 0.5))

knitr::kable(bilan_by_group[5:6,])
chisq.test(bilan_by_group$`Number of detected CG genes`[5:6], p = c(0.5, 0.5))

```

# session info

```{r}
sessionInfo()
```

### 2025-03-25 5.CT_methylation_in_oocytes_and_sperm.Rmd
Added code used to generate the paper

---
title: "X-linked MethDep CG genes are more methylated in oocytes than in sperm"
subtitle: "Figures 7E + 7F"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r, echo = FALSE, Results = 'asis'}
suppressPackageStartupMessages(library("dplyr"))
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r, warning = FALSE}
library("tidyverse")
library("ComplexHeatmap")
library("CTexploreR")
library("circlize")
library("SummarizedExperiment")
library("SingleCellExperiment")
library("DESeq2")
```

#

```{r}
CT_genes_X_met <- CT_genes %>% 
  filter(CT_gene_type == "CT_gene") %>% 
  filter(X_linked, regulated_by_methylation) %>% 
  pull(external_gene_name)

gamete_colors <- c("MII oocytes" = "deeppink",
               "Sperm"= "steelblue")

MethDep_colors <- c("MethDep" = "indianred2",
"Non-MethDep"= "cyan4")
```

scWGBS data from

[Single-cell DNA methylome sequencing of human preimplantation embryos. Nat Genet 50, 12–19 (2018). Zhu, P., Guo, H., Ren, Y. et al.](https://doi.org/10.1038/s41588-017-0007-6)

# Oocytes vs Sperm

## X-linked MethDep CT genes

```{r}
non_missing_values_thr <- 5
```

Kept genes with at least `r non_missing_values_thr` methylation value per cell type

Removed Y-sperm cells from dataset for analysis of X-linked genes!

```{r}
mean_met <- CTdata::mean_methylation_in_embryo()
mean_met <- mean_met[rowRanges(mean_met)$external_gene_name %in% CT_genes_X_met,]

# => Removed Y-sperm cells from dataset for analysis of X-linked genes!
Y_sperm_cells <- as_tibble(colData(mean_met), rownames = "cell") %>% 
  filter(Genotype.of.the.embryo == "Y") %>% 
  pull(cell)

bilan_oocytes <- as_tibble(assay(mean_met), rownames = "gene") %>% 
  pivot_longer(names_to = "cell", values_to = "met", -gene) %>% 
  left_join(as_tibble(colData(mean_met), rownames = "cell")) %>% 
  filter(!low_quality) %>% 
  #filter(!cell %in% Y_sperm_cells) %>% 
  filter(Genotype.of.the.embryo != "XXX") %>% 
  filter(Stage %in% c("MII Oocyte")) %>% 
  dplyr::select(gene, cell, met) %>% 
  pivot_wider(names_from = cell, values_from = met)

bilan_sperm <- as_tibble(assay(mean_met), rownames = "gene") %>% 
  pivot_longer(names_to = "cell", values_to = "met", -gene) %>% 
  left_join(as_tibble(colData(mean_met), rownames = "cell")) %>% 
  filter(!cell %in% Y_sperm_cells) %>% 
  filter(!low_quality) %>% 
  filter(Stage %in% c("Sperm"))  %>% 
  dplyr::select(gene, cell, met) %>% 
  pivot_wider(names_from = cell, values_from = met)

mat_oocytes <- as.matrix(bilan_oocytes[,-1])
rownames(mat_oocytes) <- bilan_oocytes$gene

mat_sperm <- as.matrix(bilan_sperm[,-1])
rownames(mat_sperm) <- bilan_sperm$gene

# Remove genes with too many missing values
# Genes must have at least met values in 5 oocytes and in 5 sperm cells
genes_selected1 <- tibble(gene = rownames(mat_oocytes),
                       mean_met = rowMeans(mat_oocytes, na.rm = TRUE),
                       non_missing_values = rowSums(!is.na(mat_oocytes)) ) %>% 
  arrange(desc(non_missing_values)) %>% 
  filter(non_missing_values >= non_missing_values_thr) %>% 
  pull(gene)

genes_selected2 <- tibble(gene = rownames(mat_sperm),
                       mean_met = rowMeans(mat_sperm, na.rm = TRUE),
                       non_missing_values = rowSums(!is.na(mat_sperm)) ) %>% 
  arrange(desc(non_missing_values)) %>% 
  filter(non_missing_values >= non_missing_values_thr) %>% 
  pull(gene)

genes_selected <- BiocGenerics::intersect(genes_selected1, genes_selected2)

# Order by decreasing methylation levels in oocytes
gene_ordered <- tibble(gene = rownames(mat_oocytes[genes_selected,]),
                       mean_met = rowMeans(mat_oocytes[genes_selected,], na.rm = TRUE)) %>% 
  arrange(desc(mean_met)) %>% 
  pull(gene)

gene_reordered <- gene_ordered
# SPANXD comes first as also methylated in sperm
if ("SPANXD" %in% gene_ordered){
  gene_reordered <- c( gene_ordered[-which(gene_ordered == "SPANXD")], "SPANXD")
}

```

`r length(gene_reordered)` genes kept.

```{r}
X_linked_CT_genes_selected <- gene_reordered

h_oocytes <- Heatmap(mat_oocytes[gene_reordered, ],
               column_title = "MII oocytes",
               name = 'Met',
               col = colorRamp2(c(1:100),colorRampPalette(
                 c("moccasin","dodgerblue4"))(100)),
               na_col = "gray80",
                     left_annotation = rowAnnotation(
                       met = anno_block(
                         gp = gpar(fill = c(MethDep_colors[1]),
                                   face = "bold",
                                   col = "black", lwd = 0.2),
                         width = unit(c(0.5), "cm"),
                         labels = "X-linked MethDep", 
                         labels_gp = gpar(col = "white", fontsize = 10))),
               clustering_method_rows = "ward.D",
               cluster_rows = FALSE,
               cluster_columns = FALSE,
               show_row_names = FALSE,
               show_column_names = TRUE,
               show_heatmap_legend = TRUE,
               show_row_dend = FALSE,
               row_names_gp = gpar(fontsize = 5),
               column_names_gp = gpar(fontsize = 0),
               column_names_side = "bottom",
               row_names_side = "left",
                 width = unit(4, "cm"),
                 height = unit(6, "cm"))

gene_to_label <- c("MAGEC2", "DDX53", "MAGEB16", "MAGEB6", "TEX13C", "PASD1", "SSX1", "CT45A10", "SPANXD", "VCX", "GAGE2A")
ind_to_label <- which(rownames(mat_sperm[gene_reordered, ]) %in% gene_to_label)
ha <- rowAnnotation(lab = anno_mark(at = ind_to_label, 
                                        labels = rownames(mat_sperm[gene_reordered, ][ind_to_label,]),
                                        labels_gp = gpar(fontsize = 6)))

h_sperm <- Heatmap(mat_sperm[gene_reordered, ],
                   column_title = "Sperm",
                   name = 'Met',
                   col = colorRamp2(c(1:100),colorRampPalette(
                     c("moccasin","dodgerblue4"))(100)),
                   na_col = "gray80",
                   clustering_method_rows = "ward.D",
                   cluster_rows = FALSE,
                   cluster_columns = FALSE,
                   show_row_names = TRUE,
                   show_column_names = TRUE,
                   show_heatmap_legend = TRUE,
                   show_row_dend = FALSE,
                   row_names_gp = gpar(fontsize = 6),
                   column_names_gp = gpar(fontsize = 0),
                   column_names_side = "bottom",
                   row_names_side = "left",
                   right_annotation = ha,
                 width = unit(4, "cm"),
                 height = unit(6, "cm"))

ht_list <- h_oocytes + h_sperm

pdf(file = "../figs/Fig7E.pdf", width = 6, height = 4)
draw(ht_list, merge_legend = TRUE, ht_gap = unit(0.1, "cm"))
dev.off()

draw(ht_list, merge_legend = TRUE, ht_gap = unit(0.1, "cm"))
```

### Quantification

#### based on mean methylation values

Calculated the **mean** methylation of each gene in MII oocytes and in sperm

```{r}
bilan_mean_CT_prom <- rbind(
  tibble(gene = gene_reordered,
       mean_met = rowMeans(mat_oocytes[gene_reordered, ], na.rm = TRUE)) %>% 
  mutate(cell_type = "MII oocytes"),
  tibble(gene = gene_reordered,
       mean_met = rowMeans(mat_sperm[gene_reordered, ], na.rm = TRUE)) %>% 
  mutate(cell_type = "Sperm"))

DT::datatable(bilan_mean_CT_prom)

knitr::kable(bilan_mean_CT_prom %>% 
  group_by(cell_type) %>% 
  dplyr::summarise(Mean_prom_met = mean(mean_met), Median_prom_met = median(mean_met)))
```

t.test: Is methylation higher in oocytes than in sperm ?

```{r}
bilan_mean_CT_prom %>% 
  ggplot(aes(x = cell_type, y = mean_met, color = cell_type)) +
  geom_boxplot(alpha = 0) +
  scale_color_manual(values = gamete_colors) +
  theme_bw() +
  theme(aspect.ratio = 2,
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none") +
  ylab("Mean promoter methylation (%)") +
  xlab('')

gridExtra::grid.arrange(bilan_mean_CT_prom %>% 
                          ggplot(aes(x = mean_met)) +
                          geom_histogram(bins = 10),
                        bilan_mean_CT_prom %>% 
                          ggplot(aes(x = log1p(mean_met))) +
                          geom_histogram(bins = 10), ncol = 2)

# bilan_mean_CT_prom %>% filter(cell_type == "MII oocytes") %>% pull(gene)
# bilan_mean_CT_prom %>% filter(cell_type == "Sperm") %>% pull(gene)
```

t.test

```{r}
t.test(bilan_mean_CT_prom %>% filter(cell_type == "MII oocytes") %>% pull(mean_met) %>% log1p(),
       bilan_mean_CT_prom %>% filter(cell_type == "Sperm") %>% pull(mean_met) %>% log1p())
```

t.test but one-side

```{r}
t.test(bilan_mean_CT_prom %>% filter(cell_type == "MII oocytes") %>% pull(mean_met) %>% log1p(),
       bilan_mean_CT_prom %>% filter(cell_type == "Sperm") %>% pull(mean_met) %>% log1p(),
       alternative = "greater")
```

t.test paired

```{r}
t.test(bilan_mean_CT_prom %>% filter(cell_type == "MII oocytes") %>% pull(mean_met) %>% log1p(),
       bilan_mean_CT_prom %>% filter(cell_type == "Sperm") %>% pull(mean_met) %>% log1p(), 
       paired = TRUE)
```

t.test one-side paired

```{r}
t.test(bilan_mean_CT_prom %>% filter(cell_type == "MII oocytes") %>% pull(mean_met) %>% log1p(),
       bilan_mean_CT_prom %>% filter(cell_type == "Sperm") %>% pull(mean_met) %>% log1p(),
       alternative = "greater", 
       paired = TRUE)
```

## Other X-linked genes

All X-linked genes other than MethDep CT genes

Kept genes with at least `r non_missing_values_thr` methylation value per cell type

Removed Y-sperm cells from dataset for analysis of X-linked genes!

```{r}
mean_met <- CTdata::mean_methylation_in_embryo()
mean_met_X_chr <- mean_met[rowRanges(mean_met)$chr == "X", ]

CTP_genes_X_met <- all_genes %>%
  filter(regulated_by_methylation) %>%
  filter(CT_gene_type == "CTP_gene") %>%
  pull(external_gene_name)

X_linked_not_CT <- all_genes %>% 
  filter(chr == "X") %>% 
  filter(!external_gene_name %in% CT_genes_X_met) %>% 
  pull(external_gene_name)

mean_met_X_chr <- mean_met_X_chr[rowRanges(mean_met_X_chr)$external_gene_name %in% X_linked_not_CT,]
```

```{r}
# => Removed Y-sperm cells from dataset for analysis of X-linked genes!
Y_sperm_cells <- as_tibble(colData(mean_met_X_chr), rownames = "cell") %>% 
  filter(Genotype.of.the.embryo == "Y") %>% 
  pull(cell)

bilan_oocytes <- as_tibble(assay(mean_met_X_chr), rownames = "gene") %>% 
  pivot_longer(names_to = "cell", values_to = "met", -gene) %>% 
  left_join(as_tibble(colData(mean_met_X_chr), rownames = "cell")) %>% 
  filter(!low_quality) %>% 
  filter(Genotype.of.the.embryo != "XXX") %>% 
  filter(Stage %in% c("MII Oocyte")) %>% 
  dplyr::select(gene, cell, met) %>% 
  pivot_wider(names_from = cell, values_from = met)

bilan_sperm <- as_tibble(assay(mean_met_X_chr), rownames = "gene") %>% 
  pivot_longer(names_to = "cell", values_to = "met", -gene) %>% 
  left_join(as_tibble(colData(mean_met_X_chr), rownames = "cell")) %>% 
  filter(!cell %in% Y_sperm_cells) %>% 
  filter(!low_quality) %>% 
  filter(Stage %in% c("Sperm"))  %>% 
  dplyr::select(gene, cell, met) %>% 
  pivot_wider(names_from = cell, values_from = met)

mat_oocytes <- as.matrix(bilan_oocytes[,-1])
rownames(mat_oocytes) <- bilan_oocytes$gene
mat_sperm <- as.matrix(bilan_sperm[,-1])
rownames(mat_sperm) <- bilan_sperm$gene

# Remove genes with too many missing values
# Genes must have at least met values in 5 oocytes and in 5 sperm cells
genes_selected1 <- tibble(gene = rownames(mat_oocytes),
                       mean_met = rowMeans(mat_oocytes, na.rm = TRUE),
                       non_missing_values = rowSums(!is.na(mat_oocytes)) ) %>% 
  arrange(desc(non_missing_values)) %>% 
  filter(non_missing_values >= non_missing_values_thr) %>% 
  pull(gene)

genes_selected2 <- tibble(gene = rownames(mat_sperm),
                       mean_met = rowMeans(mat_sperm, na.rm = TRUE),
                       non_missing_values = rowSums(!is.na(mat_sperm)) ) %>% 
  arrange(desc(non_missing_values)) %>% 
  filter(non_missing_values >= non_missing_values_thr) %>% 
  pull(gene)

genes_selected <- BiocGenerics::intersect(genes_selected1, genes_selected2)

# Order by decreasing methylation levels in oocytes
gene_ordered <- tibble(gene = rownames(mat_oocytes[genes_selected,]),
                       mean_met = rowMeans(mat_oocytes[genes_selected,], na.rm = TRUE)) %>% 
  arrange(desc(mean_met)) %>% 
  pull(gene)

length(gene_ordered)

h_oocytes <- Heatmap(mat_oocytes[gene_ordered, ],
                     column_title = "MII oocytes",
                     name = 'Met',
                     col = colorRamp2(c(1:100),colorRampPalette(
                       c("moccasin","dodgerblue4"))(100)),
                     na_col = "gray80",
                     clustering_method_rows = "ward.D",
                     cluster_rows = FALSE,
                     cluster_columns = FALSE,
                     show_row_names = TRUE,
                     show_column_names = TRUE,
                     show_heatmap_legend = TRUE,
                     show_row_dend = FALSE,
                     row_names_gp = gpar(fontsize = 0),
                     column_names_gp = gpar(fontsize = 0),
                     column_names_side = "bottom",
                     row_names_side = "left",
                     width = unit(4, "cm"),
                     height = unit(8, "cm"))

h_sperm <- Heatmap(mat_sperm[gene_ordered, ],
                   column_title = "Sperm",
                   name = 'Met',
                   col = colorRamp2(c(1:100),colorRampPalette(
                     c("moccasin","dodgerblue4"))(100)),
                   na_col = "gray80",
                   clustering_method_rows = "ward.D",
                   cluster_rows = FALSE,
                   cluster_columns = FALSE,
                   show_row_names = TRUE,
                   show_column_names = TRUE,
                   show_heatmap_legend = TRUE,
                   show_row_dend = FALSE,
                   row_names_gp = gpar(fontsize = 0),
                   column_names_gp = gpar(fontsize = 0),
                   column_names_side = "bottom",
                   row_names_side = "left",
                   width = unit(4, "cm"),
                   height = unit(8, "cm"))

ht_list1 <- h_oocytes + h_sperm

draw(ht_list1, merge_legend = TRUE, ht_gap = unit(0.1, "cm"))
```

idem but cluster by methylation in sperm

```{r}
# Order by decreasing methylation levels in sperm
gene_ordered <- tibble(gene = rownames(mat_sperm[genes_selected,]),
                       mean_met = rowMeans(mat_sperm[genes_selected,], na.rm = TRUE)) %>%
  arrange(desc(mean_met)) %>%
  pull(gene)

h_oocytes <- Heatmap(mat_oocytes[gene_ordered, ],
                     column_title = "MII oocytes",
                     name = 'Met',
                     col = colorRamp2(c(1:100),colorRampPalette(
                       c("moccasin","dodgerblue4"))(100)),
                     na_col = "gray80",
                     clustering_method_rows = "ward.D",
                     cluster_rows = FALSE,
                     cluster_columns = FALSE,
                     show_row_names = TRUE,
                     show_column_names = TRUE,
                     show_heatmap_legend = TRUE,
                     show_row_dend = FALSE,
                     row_names_gp = gpar(fontsize = 0),
                     column_names_gp = gpar(fontsize = 0),
                     column_names_side = "bottom",
                     row_names_side = "left",
                     width = unit(4, "cm"),
                     height = unit(8, "cm"))

h_sperm <- Heatmap(mat_sperm[gene_ordered, ],
                   column_title = "Sperm",
                   name = 'Met',
                   col = colorRamp2(c(1:100),colorRampPalette(
                     c("moccasin","dodgerblue4"))(100)),
                   na_col = "gray80",
                   clustering_method_rows = "ward.D",
                   cluster_rows = FALSE,
                   cluster_columns = FALSE,
                   show_row_names = TRUE,
                   show_column_names = TRUE,
                   show_heatmap_legend = TRUE,
                   show_row_dend = FALSE,
                   row_names_gp = gpar(fontsize = 0),
                   column_names_gp = gpar(fontsize = 0),
                   column_names_side = "bottom",
                   row_names_side = "left",
                   width = unit(4, "cm"),
                   height = unit(8, "cm"))

ht_list2 <- h_oocytes + h_sperm

draw(ht_list2, merge_legend = TRUE, ht_gap = unit(0.1, "cm"))

```

#### t.test

t.test Is methylation higher in oocytes than in sperm ?

```{r}
gene_reordered <- gene_ordered
bilan_mean_other_prom <- rbind(
  tibble(gene = gene_reordered,
       mean_met = rowMeans(mat_oocytes[gene_reordered, ], na.rm = TRUE)) %>% 
  mutate(cell_type = "MII oocytes"),
  tibble(gene = gene_reordered,
       mean_met = rowMeans(mat_sperm[gene_reordered, ], na.rm = TRUE)) %>% 
  mutate(cell_type = "Sperm"))

DT::datatable(bilan_mean_other_prom)

knitr::kable(bilan_mean_other_prom %>% 
  group_by(cell_type) %>% 
  dplyr::summarise(Mean_prom_met = mean(mean_met), Median_prom_met = median(mean_met)))

bilan_mean_other_prom %>% 
  ggplot(aes(x = cell_type, y = mean_met, color = cell_type)) +
  geom_boxplot(alpha = 0) +
  scale_color_manual(values = gamete_colors) +
  theme_bw() +
  theme(aspect.ratio = 2,
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "none") +
  ylab("Mean promoter methylation (%)") +
  xlab('')
```

t.test

```{r}
t.test(bilan_mean_other_prom %>% filter(cell_type == "MII oocytes") %>% pull(mean_met) %>% log1p(),
       bilan_mean_other_prom %>% filter(cell_type == "Sperm") %>% pull(mean_met) %>% log1p())
```

t.test but one-side

```{r}
t.test(bilan_mean_other_prom %>% filter(cell_type == "MII oocytes") %>% pull(mean_met) %>% log1p(),
       bilan_mean_other_prom %>% filter(cell_type == "Sperm") %>% pull(mean_met) %>% log1p(),
       alternative = "greater")
```

t.test paired

```{r}
t.test(bilan_mean_other_prom %>% filter(cell_type == "MII oocytes") %>% pull(mean_met) %>% log1p(),
       bilan_mean_other_prom %>% filter(cell_type == "Sperm") %>% pull(mean_met) %>% log1p(), 
       paired = TRUE)
```

t.test one-side paired

```{r}
t.test(bilan_mean_other_prom %>% filter(cell_type == "MII oocytes") %>% pull(mean_met) %>% log1p(),
       bilan_mean_other_prom %>% filter(cell_type == "Sperm") %>% pull(mean_met) %>% log1p(),
       alternative = "greater", 
       paired = TRUE)
```

## Alltogether

```{r}
bilan <- rbind(bilan_mean_CT_prom %>% mutate(group = "X-linked CG MethDep"),
bilan_mean_other_prom %>% mutate(group = "X-linked non CG"))

# bilan %>% 
#   ggplot(aes(x = cell_type, y = mean_met, color = cell_type)) +
#   geom_boxplot(alpha = 0) +
#   facet_wrap(~ group) +
#   scale_color_manual(values = gamete_colors) +
#   theme_bw() +
#   theme(aspect.ratio = 2,
#         panel.grid.minor = element_blank(),
#         panel.grid.major = element_blank(),
#         legend.position = "none")

bilan$group <- factor(bilan$group)
levels(bilan$group) <- c("X-linked\nCG MethDep","X-linked\nnon CG")

group_colors <- c("X-linked\nCG MethDep" = "indianred2", "X-linked\nnon CG"= "gray40")
fig <- bilan %>% 
  ggplot(aes(x = cell_type, y = mean_met, color = group, fill = group)) +
  scale_color_manual(values = group_colors) +
    scale_fill_manual(values = group_colors) +
  geom_boxplot(alpha = 0.2, outliers = FALSE) +
  #facet_wrap(~ group) +
  #scale_color_manual(values = gamete_colors) +
  theme_bw() +
  theme(#aspect.ratio = 2,
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.75, 0.85),
        legend.key.spacing.y = unit(0.3, 'cm'),
        legend.background = element_blank(),
        #legend.spacing.x = unit(20.0, 'cm'),
        legend.title = element_blank(),
        legend.text = element_text(size = 8),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 12),
        axis.text = element_text(size = 12, face = "bold")) +
  ylab("Mean promoter methylation (%)")

pdf(file = paste0("../figs/Fig7F.pdf"), height = 3, width = 3)
fig
dev.off()

fig
```

## Welch test

Welch test (t.test but set var.equal = FALSE)

*Welch test is more robust than t.tests when the number of sample per group is very different*

For` X-linked MethDep`, methylation is also significantly different between `oocytes` and `sperm`:

```{r}
bilan <- bilan %>% 
  mutate(Group = paste0(group, "_", cell_type)) %>% 
  mutate(log_mean_met = log1p(mean_met))

t.test(bilan %>% 
         filter(cell_type == "MII oocytes") %>% 
         filter(group == "X-linked\nCG MethDep") %>% pull(log_mean_met),
      bilan %>% 
         filter(cell_type == "Sperm") %>% 
         filter(group == "X-linked\nCG MethDep") %>% pull(log_mean_met),
        var.equal = FALSE)
```

For` X-linked Non MethDep`, methylation is also significantly different between `oocytes` ad `sperm`:

```{r}
t.test(bilan %>% 
         filter(cell_type == "MII oocytes") %>% 
         filter(group == "X-linked\nnon CG") %>% pull(log_mean_met),
      bilan %>% 
         filter(cell_type == "Sperm") %>% 
         filter(group == "X-linked\nnon CG") %>% pull(log_mean_met),
      var.equal = FALSE)

```

Methylation is highly significantly different in oocytes between `X-linked MethDep` and `X-linked non CG`:

```{r}
t.test(bilan %>% 
         filter(cell_type == "MII oocytes") %>% 
         filter(group == "X-linked\nCG MethDep") %>% pull(log_mean_met),
      bilan %>% 
         filter(cell_type == "MII oocytes") %>% 
         filter(group == "X-linked\nnon CG") %>% pull(log_mean_met),
      var.equal = FALSE)

# wilcox.test(bilan %>% 
#          filter(cell_type == "MII oocytes") %>% 
#          filter(group == "X-linked\nCG MethDep") %>% pull(mean_met),
#       bilan %>% 
#          filter(cell_type == "MII oocytes") %>% 
#          filter(group == "X-linked\nnon CG") %>% pull(mean_met))
```

Methylation is not significantly different in sperm between `X-linked MethDep` and `X-linked non CG`:

```{r}
t.test(bilan %>% 
         filter(cell_type == "Sperm") %>% 
         filter(group == "X-linked\nCG MethDep") %>% pull(log_mean_met),
      bilan %>% 
         filter(cell_type == "Sperm") %>% 
         filter(group == "X-linked\nnon CG") %>% pull(log_mean_met),
      var.equal = FALSE)

# wilcox.test(bilan %>% 
#          filter(cell_type == "MII oocytes") %>% 
#          filter(group == "X-linked\nCG MethDep") %>% pull(mean_met),
#       bilan %>% 
#          filter(cell_type == "MII oocytes") %>% 
#          filter(group == "X-linked\nnon CG") %>% pull(mean_met))
```

# Session info

```{r}
sessionInfo()
```

### 2025-03-25 6.CT_methylation_in_early_embryos.Rmd
Added code used to generate the paper

---
title: "X-linked MethDep in early embryos, Zhu scWGBS data"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r, echo = FALSE, Results = 'asis'}
suppressPackageStartupMessages(library("dplyr"))
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
library("tidyverse")
library("ComplexHeatmap")
library("CTexploreR")
library("circlize")
library("SummarizedExperiment")
library("SingleCellExperiment")
library("DESeq2")
```

#

```{r}
legends_param <- list(
  labels_gp = gpar(col = "black", fontsize = 3),
  title_gp = gpar(col = "black", fontsize = 3, fontface = "bold"),
  simple_anno_size = unit(0.1, "cm"),
  row_names_gp = gpar(fontsize = 1),
  annotation_name_side = "left",
  border = FALSE,
  border_gp = gpar(lwd = 0.1),
  grid_width = unit(0.1, "cm"),
  grid_height = unit(0.01, "cm"),
  legend_height = unit(0.1, "cm"))

sex_colors<- c("XX" = "deeppink",
               "XY"= "steelblue",
               "X" = "gray20")

CT_genes_X_met <- all_genes %>% 
  filter(CT_gene_type == "CT_gene") %>% 
  filter(X_linked, regulated_by_methylation) %>% 
  pull(external_gene_name)
```

Data from:
[Single-cell DNA methylome sequencing of human preimplantation embryos. Nat Genet 50, 12–19 (2018). Zhu, P., Guo, H., Ren, Y. et al.](https://doi.org/10.1038/s41588-017-0007-6)

# Selection of X-linked CG MethDep genes initially methylated in oocytes

```{r}
non_missing_values_thr <- 5
met_thr_in_oocytes <- 33 #30
met_thr_in_sperm <- 33 #30
```

Selected X-linked CG MethDep genes initially methylated in oocytes

Selection procedure:

- Selected all X-linked genes with at least `r non_missing_values_thr` methylation values in each cell type (oocytes and sperm cells)

- Selected X-linked genes showing at least `r met_thr_in_oocytes`% methylation in MII oocytes (but less than `r met_thr_in_sperm`% in sperm)

- Kept Sperm cells carrying an X chromosome

```{r}
mean_met <- CTdata::mean_methylation_in_embryo()
mean_met <- mean_met[rowRanges(mean_met)$external_gene_name %in% CT_genes_X_met,]

Y_sperm_cells <- as_tibble(colData(mean_met), rownames = "cell") %>% 
  filter(Genotype.of.the.embryo == "Y") %>% 
  pull(cell)

genes_selected_in_oocytes <- as_tibble(assay(mean_met), rownames = "gene") %>% 
  pivot_longer(names_to = "cell", values_to = "met", -gene) %>% 
  left_join(as_tibble(colData(mean_met), rownames = "cell")) %>% 
  filter(!cell %in% Y_sperm_cells) %>% 
  filter(Genotype.of.the.embryo != "XXX") %>% 
  dplyr::rename(Genotype = Genotype.of.the.embryo) %>% 
  filter(!low_quality) %>% 
  filter(Stage %in% c("MII Oocyte")) %>% 
  filter(!is.na(met)) %>% 
  group_by(gene) %>% 
  dplyr::count() %>% 
  filter(n >= non_missing_values_thr) %>% 
  pull(gene)

genes_selected_in_sperm <- as_tibble(assay(mean_met), rownames = "gene") %>% 
  pivot_longer(names_to = "cell", values_to = "met", -gene) %>% 
  left_join(as_tibble(colData(mean_met), rownames = "cell")) %>% 
  filter(!cell %in% Y_sperm_cells) %>% 
  filter(Genotype.of.the.embryo != "XXX") %>% 
  dplyr::rename(Genotype = Genotype.of.the.embryo) %>% 
  filter(!low_quality) %>% 
  filter(Stage %in% c("Sperm"))%>% 
  filter(!is.na(met)) %>% 
  group_by(gene) %>% 
  dplyr::count() %>% 
  filter(n >= non_missing_values_thr) %>% 
  pull(gene)

genes_selected <- BiocGenerics::intersect(genes_selected_in_oocytes, genes_selected_in_sperm)

bilan <- as_tibble(assay(mean_met), rownames = "gene") %>% 
  filter(gene %in% genes_selected) %>% 
  pivot_longer(names_to = "cell", values_to = "met", -gene) %>% 
  left_join(as_tibble(colData(mean_met), rownames = "cell")) %>% 
  filter(!cell %in% Y_sperm_cells) %>% 
  filter(Genotype.of.the.embryo != "XXX") %>% 
  dplyr::rename(Genotype = Genotype.of.the.embryo) %>% 
  filter(!low_quality) %>% 
  filter(Stage %in% c("MII Oocyte", "Sperm")) %>% 
  group_by(gene, Stage, Genotype) %>% 
  dplyr::summarise(mean_met = mean(met, na.rm = TRUE)) %>% 
  mutate(stage_sex = paste0(Stage, '_', Genotype))

```

```{r}
genes_methylated_in_oocytes <- bilan %>% 
  filter(Stage %in% c("MII Oocyte")) %>% 
  arrange(mean_met) %>% 
  filter(mean_met >= met_thr_in_oocytes) %>% 
  pull(gene)

met_in_sperm_genes <- bilan %>% 
  filter(Stage %in% c("Sperm")) %>% 
  arrange(mean_met) %>% 
  filter(mean_met > met_thr_in_sperm) %>% 
  pull(gene)

selected_genes <- genes_methylated_in_oocytes[!genes_methylated_in_oocytes %in% met_in_sperm_genes]

bilan %>% 
  mutate(selected = ifelse(gene %in% selected_genes, "kept", "removed")) %>% 
  filter(Stage %in% c("MII Oocyte")) %>% 
  arrange(mean_met) %>% 
  left_join(all_genes %>% dplyr::rename(gene = external_gene_name) %>% dplyr::select(gene, CT_gene_type)) %>% 
  ggplot(aes(x = gene, y = mean_met, color = selected, label = gene)) +
  geom_point() +
  geom_hline(yintercept = met_thr_in_oocytes) +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom")  +
  geom_text( color = "black", size = 3)

```

```{r}
tested_genes <- as_tibble(assay(mean_met), rownames = "gene") %>% 
  filter(gene %in% selected_genes) %>% 
  pull(gene)
tested_genes
length(tested_genes)

bilan_selected <- as_tibble(assay(mean_met), rownames = "gene") %>% 
  filter(gene %in% tested_genes) %>% 
  pivot_longer(names_to = "cell", values_to = "met", -gene) %>% 
  left_join(as_tibble(colData(mean_met), rownames = "cell")) %>% 
  filter(!cell %in% Y_sperm_cells) %>% 
  filter(Genotype.of.the.embryo != "XXX") %>% 
  dplyr::rename(Genotype = Genotype.of.the.embryo) %>% 
  #filter(!low_quality) %>%  # otherwise no more XX-8-cells 
  filter(!Stage %in% c("GV Oocyte")) %>% 
  group_by(gene, Stage, Genotype) %>% 
  dplyr::summarise(mean_met = mean(met, na.rm = TRUE)) %>% 
  mutate(stage_sex = paste0(Stage, '_', Genotype))

levels(bilan_selected$Stage)[10] <- "Post-impl"

fig <- bilan_selected %>% 
  ggplot(aes(x = Stage, y = mean_met, group = Genotype)) +
  geom_boxplot(aes(x = Stage, y = mean_met, group = stage_sex, 
                   color = Genotype, fill = Genotype), 
               outliers = FALSE,alpha = 0.1) +
  scale_color_manual(values = sex_colors) +
  scale_fill_manual(values = sex_colors) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12, face = "bold"),
        axis.title.y = element_text(size = 12, face = "bold"),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(linewidth = 0.2)) +
  ylab("Mean promoter methylation (%)") +
  xlab('')

fig

pdf(file = paste0("../figs/Fig8A_CG_met_in_Male_and_Female_embryos.pdf"),
    width = 5, height = 4)
fig
dev.off()

```

# Example of genes showing a transient imprint in early embryos

```{r}
gamete_colors <- c("MII oocytes" = "deeppink",
               "Sperm"= "steelblue")

MethDep_colors <- c("MethDep" = "indianred2",
"Non-MethDep"= "cyan4")

Stage_col <- c("GV Oocyte" = "mediumvioletred", "MII Oocyte" = "brown4", 
               "Sperm" = "palegreen2", "Zygote" = "lightgoldenrod1", 
               "2-cell" = "ivory2", "4-cell" = "lightslateblue", 
               "8-cell" = "plum2", "Morula" = "deepskyblue",
               "Blastocyst" = "sienna1",  "Post-implantation" = "peachpuff")

stage_level_col <- 
  c("GV Oocyte" = "mediumvioletred", "MII Oocyte" = "brown4", 
    "Sperm" = "palegreen2", 
    "PN-early" = "yellow", "PN-mid" = "gold", "PN-late" = "gold4",
    "2-cell" = "ivory2", "4-cell" = "lightslateblue", 
    "8-cell" = "plum2", "Morula" = "deepskyblue",
    "Blastocyst (Mixed)" = "sienna1",  "Blastocyst (ICM)" = "indianred1",
    "Blastocyst (TE)" = "purple", "Heart" = "steelblue", "Villus" = "gray" )
```

```{r}
methylation_in_embryo <- CTdata::methylation_in_embryo()
load("/storage/research/dduv/cbio-lg/cluster/Packages/CTdata/inst/extdata/all_promoter_regions_hg19.rda")

legends_param <- list(
  labels_gp = gpar(col = "black", fontsize = 10),
  title_gp = gpar(col = "black", fontsize = 10, fontface = "bold"),
  simple_anno_size = unit(0.1, "cm"),
  row_names_gp = gpar(fontsize = 1),
  annotation_name_side = "left",
  border = FALSE,
  border_gp = gpar(lwd = 0.1),
  grid_width = unit(0.3, "cm"),
  grid_height = unit(0.3, "cm"),
  legend_height = unit(0.5, "cm"))
```

```{r}

promoter_methylation_by_GR <- function(GR = NULL, 
                                       nt_up = 1000,
                                       nt_down = 500,
                                       sc_or_bulk = NULL, # NULL, single or bulk
                                       genotype = NULL,
                                       stage = NULL,
                                       split_by = "stage_level",
                                       rm_low_quality = TRUE,
                                       rm_cell_with_only_missing_values = TRUE,
                                       clustRow = FALSE,
                                       fs = 6){

  x <- GR
  gene <- x$external_gene_name
  tested_gene_GR <- subsetByOverlaps(methylation_in_embryo, x)
  CpGpos <- tested_gene_GR@rowRanges@ranges@start
  TSS_hg19 <- x$TSS_liftover
  if(as.vector(x@strand) == '+'){
    CpGpositions <- CpGpos - x$TSS_liftover
  }
  if(as.vector(x@strand) == '-'){
    CpGpositions <- x$TSS_liftover  - CpGpos 
  }
  df_col <- colData(tested_gene_GR)
  if (rm_low_quality) df_col <- df_col[!is.na(df_col$low_quality) & !df_col$low_quality,]
  if (!is.null(stage)) df_col <- df_col[df_col$stage_level %in% stage, ]
  if (!is.null(genotype)) df_col <- df_col[df_col$`Genotype of the embryo` %in% genotype, ]
  df_col <- df_col[!df_col$`Genotype of the embryo` %in% "XXX", ]
  if (!is.null(sc_or_bulk)){
    if(sc_or_bulk == "single"){
      df_col <- df_col[df_col$`Single Cell or Bulk cell` == "single cell", ]
    }
    if(sc_or_bulk == "bulk"){
      df_col <- df_col[df_col$`Single Cell or Bulk cell` == "bulk cells", ]
    }
  }

  df_col <- df_col[order(df_col$stage_level, df_col$`Genotype of the embryo`),]

  y <- as_tibble(assay(tested_gene_GR)[, df_col$Sample_Name]) %>% 
    mutate(CpG = CpGpositions) %>% dplyr::select(CpG, everything()) %>% 
    filter(CpG > -nt_up & CpG < nt_down) %>% 
   #  filter(CpG > -nt_arround_TS & CpG < nt_arround_TS) %>% 
    arrange(CpG) %>% 
    pivot_longer(names_to = "Sample_Name", values_to = "met", -CpG) %>% 
    pivot_wider(names_from = CpG, values_from = met)
  mat <- as.matrix(y[, -1])
  rownames(mat) <- y$Sample_Name

  if (rm_cell_with_only_missing_values) {
    if (length(which(rowSums(is.na(mat)) == ncol(mat))) > 0) {
      mat <- mat[-which(rowSums(is.na(mat)) == ncol(mat)), ,drop = F]
      df_col <- df_col[rownames(mat),]
    }
  }

  rowSplit <- df_col$stage_level
  if (split_by == "Stage") {
    df_col <- df_col[order(df_col$Stage, df_col$`Genotype of the embryo`),] 
    rowSplit <- df_col$Stage
  }

  if (split_by == "embryo") {
    df_col$embryo <- paste0(df_col$stage_level, "_", df_col$individual)
    df_col$embryo[df_col$stage_level %in% c("GV Oocyte")] <- "GV Oocyte"
    df_col$embryo[df_col$stage_level %in% c("MII Oocyte")] <- "MII Oocyte"
    df_col$embryo[df_col$stage_level %in% c("Sperm")] <- "Sperm"
    df_col$embryo[df_col$stage_level %in% c("2-cell")] <- "2-cell"
        df_col$embryo[df_col$stage_level %in% c("4-cell")] <- "4-cell"
            df_col$embryo[df_col$stage_level %in% c("8-cell")] <- "8-cell"
    df_col$embryo[df_col$stage_level %in% c("Heart")] <- "Heart"
    df_col$embryo[df_col$stage_level %in% c("Villus")] <- "Villus"
    df_col <- df_col[order(df_col$stage_level, df_col$`Genotype of the embryo`, df_col$embryo),] 
    rowSplit <- factor(df_col$embryo, levels = unique(df_col$embryo))
  }

  Genotype <- as.matrix(df_col$`Genotype of the embryo`)
  rownames(Genotype) <- df_col$Sample_Name
  row_ha_Genotype <- rowAnnotation(Genotype = Genotype,
                              na_col = "gray80",
                              annotation_legend_param = legends_param,
                              simple_anno_size = unit(0.2, "cm"),
                              col = list(Genotype = c("XX" = "deeppink",
                                                 "XY"= "steelblue",
                                                 "X" = "pink",
                                                 "Y"= "dodgerblue",
                                                 "XXX" = "orange")),
                              annotation_name_gp = gpar(fontsize = 0))

  Stage <- as.matrix(df_col$Stage)
  rownames(Stage) <- df_col$Stage
  row_ha_Stage <- rowAnnotation(Stage = Stage,
                              na_col = "gray80",
                              annotation_legend_param = legends_param,
                              simple_anno_size = unit(0.2, "cm"),
                             # col = list(Stage = Stage_col),
                              annotation_name_gp = gpar(fontsize = 0))

  stage_level <- as.matrix(df_col$stage_level)
  rownames(stage_level) <- df_col$stage_level
  row_ha_stage_level <- rowAnnotation(stage_level = stage_level,
                              na_col = "gray80",
                              annotation_legend_param = legends_param,
                              simple_anno_size = unit(0.2, "cm"),
                              col = list(stage_level = stage_level_col),
                              annotation_name_gp = gpar(fontsize = 0))
  data_type <- as.matrix(df_col$bulk_or_single_cell)
  rownames(data_type) <- df_col$bulk_or_single_cell
  row_ha_data_type <- rowAnnotation(data_type = data_type,
                              na_col = "gray80",
                              annotation_legend_param = legends_param,
                              simple_anno_size = unit(0.2, "cm"),
                              col = list(data_type = c("bulk cells" = "black",
                                         "single cell" = "cyan")),
                              annotation_name_gp = gpar(fontsize = 0))

  h <- Heatmap(mat[rownames(df_col), ],
               column_title = paste0('Promoter methylation of ', gene, " TSS ", TSS_hg19, " (hg19)"),
               name = 'Met',
               row_split = rowSplit,
               cluster_row_slices = FALSE,
               row_gap = unit(c(1), "mm"),
               row_title_gp = gpar(fontsize = 8, fontface = "bold"),
               row_title_rot = 0,
               col = colorRamp2(c(1:100),colorRampPalette(
                 c("moccasin","dodgerblue4"))(100)),
               na_col = "gray80",
               clustering_method_rows = "ward.D",
               cluster_rows = clustRow,
               cluster_columns = FALSE,
               show_row_names = TRUE,
               show_column_names = TRUE,
               show_heatmap_legend = TRUE,
               show_row_dend = FALSE,
               right_annotation = c(row_ha_Genotype,  #row_ha_Stage,
                                    #row_ha_stage_level,
                                    #row_ha_data_type,
                                      gap = unit(1, "mm")),
               row_names_gp = gpar(fontsize = fs),
               column_names_gp = gpar(fontsize = fs),
               column_names_side = "bottom",
               row_names_side = "left")

  print(h) 
}
```

```{r}
promoter_methylation_fig <- function(GR = NULL, 
                                     title = "",
                                     nt_up = 1000,
                                     nt_down = 500,
                                     sc_or_bulk = NULL, # NULL, single or bulk
                                     genotype = NULL,
                                     stage = NULL,
                                     split_by = "stage_level",
                                     rm_low_quality = TRUE,
                                     rm_cell_with_only_missing_values = TRUE,
                                     show_legend = TRUE,
                                     cell_per_type = NULL,
                                     clustRow = FALSE,
                                     h_width = 5,
                                     h_height = 10,
                                     rowtitle_fs = 7,
                                     coltitle_fs = 7,
                                     CpG_n_fs = 4,
                                     fs = 6){
  x <- GR
  gene <- x$external_gene_name
  tested_gene_GR <- subsetByOverlaps(methylation_in_embryo, x)
  CpGpos <- tested_gene_GR@rowRanges@ranges@start
  TSS_hg19 <- x$TSS_liftover
  if(as.vector(x@strand) == '+'){
    CpGpositions <- CpGpos - x$TSS_liftover
  }
  if(as.vector(x@strand) == '-'){
    CpGpositions <- x$TSS_liftover  - CpGpos 
  }

  df_col <- colData(tested_gene_GR)
  if (rm_low_quality) df_col <- df_col[!is.na(df_col$low_quality) & !df_col$low_quality,]
  if (!is.null(stage)) df_col <- df_col[df_col$stage_level %in% stage, ]
  if (!is.null(genotype)) df_col <- df_col[df_col$`Genotype of the embryo` %in% genotype, ]
  df_col <- df_col[!df_col$`Genotype of the embryo` %in% "XXX", ]

  if (!is.null(sc_or_bulk)){
    if(sc_or_bulk == "single"){
      df_col <- df_col[df_col$`Single Cell or Bulk cell` == "single cell", ]
    }
    if(sc_or_bulk == "bulk"){
      df_col <- df_col[df_col$`Single Cell or Bulk cell` == "bulk cells", ]
    }
  }

  df_col <- df_col[!df_col$Sample_Name %in% c("Icm03-W-s64","scBS-C-MII-9"), ]

  y <- as_tibble(assay(tested_gene_GR)[, df_col$Sample_Name]) %>% 
    mutate(CpG = CpGpositions) %>% dplyr::select(CpG, everything()) %>% 
    filter(CpG > -nt_up & CpG < nt_down) %>% 
    #  filter(CpG > -nt_arround_TS & CpG < nt_arround_TS) %>% 
    arrange(CpG) %>% 
    pivot_longer(names_to = "Sample_Name", values_to = "met", -CpG) %>% 
    pivot_wider(names_from = CpG, values_from = met)
  mat <- as.matrix(y[, -1])
  rownames(mat) <- y$Sample_Name

  # calculate mean in each sample, to be able to order sequences within in group
  # in fonction of the methylation level
  df_col$median_met <- rowMedians(mat, na.rm = T)
  df_col$prop_of_missing_values <- (rowSums(is.na(mat)) / ncol(mat))

  if (rm_cell_with_only_missing_values) df_col <- df_col[df_col$prop_of_missing_values < 1, ]

  if (!is.null(cell_per_type)){
    ### Keep n max cell per group, keeping the samples with the less missing values

    kept <- as_tibble(df_col) %>% 
      filter(Stage %in% cell_per_type$Stage[1]) %>% 
      arrange(prop_of_missing_values) %>% 
      head(pull(cell_per_type[1, 2])) %>% 
      pull(Sample_Name)
    for (i in seq_along(cell_per_type$Stage)[-1]){
      tmp <- as_tibble(df_col) %>% 
        filter(Stage %in% cell_per_type$Stage[i]) %>% 
        arrange(prop_of_missing_values) %>% 
        head(pull(cell_per_type[i, 2])) %>% 
        pull(Sample_Name)
      kept <- c(kept, tmp)
    }
    df_col <- df_col[df_col$Sample_Name %in% kept, ] 
  }

  df_col <- df_col[order(df_col$stage_level, df_col$`Genotype of the embryo`, df_col$median_met),]

  rowSplit <- df_col$stage_level

  if (split_by == "Stage") {
    df_col <- df_col[order(df_col$Stage, df_col$`Genotype of the embryo`, df_col$median_met),] 
    rowSplit <- df_col$Stage
  }

  if (split_by == "embryo") {
    df_col$embryo <- paste0(df_col$stage_level, "_", df_col$individual)
    df_col$embryo[df_col$stage_level %in% c("GV Oocyte")] <- "GV Oocyte"
    df_col$embryo[df_col$stage_level %in% c("MII Oocyte")] <- "MII Oocyte"
    df_col$embryo[df_col$stage_level %in% c("Sperm")] <- "Sperm"
    df_col$embryo[df_col$stage_level %in% c("2-cell")] <- "2-cell"
    df_col$embryo[df_col$stage_level %in% c("4-cell")] <- "4-cell"
    df_col$embryo[df_col$stage_level %in% c("8-cell")] <- "8-cell"
    df_col$embryo[df_col$stage_level %in% c("Heart")] <- "Heart"
    df_col$embryo[df_col$stage_level %in% c("Villus")] <- "Villus"
    df_col <- df_col[order(df_col$stage_level, df_col$`Genotype of the embryo`, df_col$embryo, df_col$mean_met),] 
    rowSplit <- factor(df_col$embryo, levels = unique(df_col$embryo))
  }

  Genotype <- as.matrix(df_col$`Genotype of the embryo`)
  rownames(Genotype) <- df_col$Sample_Name
  row_ha_Genotype <- rowAnnotation(Genotype = Genotype,
                                   na_col = "gray80",
                                   annotation_legend_param = legends_param,
                                   simple_anno_size = unit(0.2, "cm"),
                                   col = list(Genotype = c("XX" = "deeppink",
                                                           "XY"= "steelblue",
                                                           "X" = "pink",
                                                           "Y"= "dodgerblue",
                                                           "XXX" = "orange")),
                                   annotation_name_gp = gpar(fontsize = 0))

  Stage <- as.matrix(df_col$Stage)
  rownames(Stage) <- df_col$Stage
  row_ha_Stage <- rowAnnotation(Stage = Stage,
                                na_col = "gray80",
                                annotation_legend_param = legends_param,
                                simple_anno_size = unit(0.2, "cm"),
                                col = list(Stage = Stage_col),
                                annotation_name_gp = gpar(fontsize = 0))

  stage_level <- as.matrix(df_col$stage_level)
  rownames(stage_level) <- df_col$stage_level
  row_ha_stage_level <- rowAnnotation(stage_level = stage_level,
                                       na_col = "gray80",
                                       annotation_legend_param = legends_param,
                                       simple_anno_size = unit(0.2, "cm"),
                                       col = list(stage_level = stage_level_col),
                                       annotation_name_gp = gpar(fontsize = 0))
  data_type <- as.matrix(df_col$bulk_or_single_cell)
  rownames(data_type) <- df_col$bulk_or_single_cell
  row_ha_data_type <- rowAnnotation(data_type = data_type,
                                    na_col = "gray80",
                                    annotation_legend_param = legends_param,
                                    simple_anno_size = unit(0.2, "cm"),
                                    col = list(data_type = c("bulk cells" = "black",
                                                             "single cell" = "cyan")),
                                    annotation_name_gp = gpar(fontsize = 0))

  if (all(all_of(df_col$`Genotype of the embryo`) %in% c("X", "XX"))) gender <- "Female"
  if (all(all_of(df_col$`Genotype of the embryo`) %in% c("Y", "XY"))) gender <- "Male"

  ht_opt$TITLE_PADDING = unit(c(2.5, 5), "points")
  h <- Heatmap(mat[rownames(df_col), ],
               column_title = title,
               name = "Met",
               row_split = rowSplit,
               cluster_row_slices = FALSE,
               row_gap = unit(c(3), "mm"),
               border = TRUE,
               border_gp = gpar(lwd = 0.1),
               row_title_gp = gpar( fontsize = rowtitle_fs, fontface = "bold"),
               column_title_gp = gpar(fontsize = coltitle_fs, fontface = "bold"),
               row_title_rot = 90,
               col = colorRamp2(c(1:100),colorRampPalette(
                 c("moccasin","dodgerblue4"))(100)),
               na_col = "gray80",
               clustering_method_rows = "ward.D",
               cluster_rows = clustRow,
               cluster_columns = FALSE,
               show_row_names = FALSE,
               show_column_names = TRUE,
               show_heatmap_legend = show_legend,
               show_row_dend = FALSE,
               row_names_gp = gpar(fontsize = fs),
               column_names_gp = gpar(fontsize = CpG_n_fs),
               heatmap_width = unit(h_width, "cm"),
               heatmap_height = unit(h_height, "cm"),
               column_names_side = "bottom",
               row_names_side = "left",
               heatmap_legend_param = legends_param)

  print(h) 
}

```

# MAGEC2 Figures

```{r}
my_gene <- "MAGEC2"
my_GR <- all_promoter_regions_hg19[all_promoter_regions_hg19$external_gene_name == my_gene,]

promoter_methylation_by_GR(GR = my_GR[1], 
                           nt_up = 1000,
                           nt_down = 500,
                           sc_or_bulk = NULL,
                           #genotype = "XX",
                           #stage = "MII Oocyte",
                           #split_by = "individual",
                           split_by = "Stage",
                           rm_low_quality = FALSE,
                           rm_cell_with_only_missing_values = TRUE,
                           clustRow = FALSE,
                           fs = 2)
```

Show Promoter region -/- 1000 pb around TSS

```{r}

prom_region <- c(1000, 1000)

my_GR <- all_promoter_regions_hg19[all_promoter_regions_hg19$external_gene_name == my_gene,]
gene <- my_GR$external_gene_name

cell_per_type <- tibble(Stage = c("MII Oocyte", "Sperm"),
                        n = c(10, 10))

A <- cowplot::plot_grid(grid.grabExpr(
  draw(promoter_methylation_fig(GR = my_GR[1], 
            title = "Oocytes",
            show_legend = FALSE,
            rowtitle_fs = 0,
            CpG_n_fs = 2.5,
            nt_up = prom_region[1],
            nt_down = prom_region[2],
            sc_or_bulk = NULL,
            genotype = c("X", "XX"),
            stage = c("MII Oocyte") ,
            #split_by = "individual",
            split_by = "Stage",
            rm_low_quality = FALSE,
            rm_cell_with_only_missing_values = TRUE,
            cell_per_type = cell_per_type,
            clustRow = FALSE,
            h_width = 4,
            h_height = 2,
            fs = 0)),  labels = c('A')))

B <- cowplot::plot_grid(grid.grabExpr(
  draw(promoter_methylation_fig(GR = my_GR[1], 
            title = "Sperm",
            show_legend = FALSE,
            rowtitle_fs = 0,
             CpG_n_fs = 2.5,
             nt_up = prom_region[1],
            nt_down = prom_region[2],
            sc_or_bulk = NULL,
            genotype = c("X", "XX"),
            stage = c("Sperm") ,
            #split_by = "individual",
            split_by = "Stage",
            rm_low_quality = FALSE,
            rm_cell_with_only_missing_values = TRUE,
            cell_per_type = cell_per_type,
            clustRow = FALSE,
            h_width = 4,
            h_height = 2,
            fs = 0)),  labels = c('A')))

figA <- cowplot::plot_grid(A, B, 
                   # rel_widths = c(2,1),
                   # scale = c(1, 0.8),
                   # label_size = 10, 
                   nrow = 2,
                   align = "h")

# pdf(file = paste0("../figs/Fig8C_", gene, "_Methylation_Oocytes_Sperm.pdf"))#, height = 3, width =4)
# figA
# dev.off()

# Show the same number of cells in male and female embryos
# Define the number of cells per stage manually
cell_per_type <- tibble(Stage = c("Zygote", #"2-cell", "4-cell", "8-cell", 
                                  "Morula", "Blastocyst", "Post-implantation"),
                        n = c(6, #8, 8, 8, 
                              10, 12, 4))

cell_per_type <- tibble(Stage = c("Zygote", #"2-cell", "4-cell", "8-cell", 
                                  "Morula", "Blastocyst", "Post-implantation"),
                        n = c(10, #8, 8, 8, 
                              10, 20, 4))

D <- cowplot::plot_grid(grid.grabExpr(draw(
  promoter_methylation_fig(GR = my_GR[1], 
                           title = "Female",
                           show_legend = FALSE,
                           rowtitle_fs = 8,
                           coltitle_fs = 10,
                           CpG_n_fs = 2.5,
                           nt_up = prom_region[1],
                           nt_down = prom_region[2],
                           sc_or_bulk = "single",
                           genotype = c("X", "XX"),
                           stage = c("PN-early", "PN-mid", "PN-late", "2-cell", #"4-cell", 
                                     "8-cell", "Morula", "Blastocyst (Mixed)", 
                                     "Blastocyst (ICM)", "Heart", "Villus"),    
                           #split_by = "individual",
                           split_by = "Stage",
                           rm_low_quality = FALSE, # otherwise not enough samples
                           rm_cell_with_only_missing_values = TRUE,
                           cell_per_type = cell_per_type,
                           clustRow = FALSE,
                           h_width = 4,
                           h_height =5,
                           fs = 8)),  labels = c('A')))

E <- cowplot::plot_grid(grid.grabExpr(draw(
  promoter_methylation_fig(GR = my_GR[1], 
                           title = "Male",
                           rowtitle_fs = 8,
                           coltitle_fs = 10,
                          CpG_n_fs = 2.5,
                           nt_up = prom_region[1],
                           nt_down = prom_region[2],
                           sc_or_bulk = "single",
                           genotype = c("Y", "XY"),
                           stage = c("PN-early", "PN-mid", "PN-late", "2-cell", "4-cell", 
                                     "8-cell", 
                                     "Morula", "Blastocyst (Mixed)", "Blastocyst (ICM)", "Heart", "Villus"),       
                           #split_by = "individual",
                           split_by = "Stage",
                           rm_low_quality = FALSE, # otherwise not enough samples
                           rm_cell_with_only_missing_values = TRUE,
                           cell_per_type = cell_per_type,
                           clustRow = FALSE,
                           h_width = 4,
                           h_height =5,
                           fs = 8)),  labels = c('B')))

figB<- cowplot::plot_grid(D,E, 
                   # rel_widths = c(2,1),
                   # scale = c(1, 0.8),
                   # label_size = 10, 
                   nrow = 1,
                   align = "h")

# pdf(file = paste0("../figs/Fig8C_", gene, "_Methylation_embryos.pdf"))#, height = 3, width = 3)
# figB
# dev.off()

figC <- cowplot::plot_grid(figA, figB,
                    rel_heights = c(1,1.6),
                   # scale = c(1, 0.8),
                   # label_size = 10, 
                   nrow = 2,
                   align = "h",
                   labels = my_gene)

pdf(file = paste0("../figs/Fig8C_", gene, "_Methylation_embryos.pdf"), height = 3.5, width = 4.5)
figC
dev.off()

figC
```

# DDX53 Figures

```{r}
my_gene <- "DDX53"
my_GR <- all_promoter_regions_hg19[all_promoter_regions_hg19$external_gene_name == my_gene,]

promoter_methylation_by_GR(GR = my_GR[1], 
                           nt_up = 1000,
                           nt_down = 500,
                           sc_or_bulk = NULL,
                           #genotype = "XX",
                           #stage = "MII Oocyte",
                           #split_by = "individual",
                           split_by = "Stage",
                           rm_low_quality = FALSE,
                           rm_cell_with_only_missing_values = TRUE,
                           clustRow = FALSE,
                           fs = 2)
```

Show Promoter region -/- 1000 pb around TSS

```{r}
prom_region <- c(1000, 1000)

my_GR <- all_promoter_regions_hg19[all_promoter_regions_hg19$external_gene_name == my_gene,]
gene <- my_GR$external_gene_name

cell_per_type <- tibble(Stage = c("MII Oocyte", "Sperm"),
                        n = c(10, 10))

A <- cowplot::plot_grid(grid.grabExpr(
  draw(promoter_methylation_fig(GR = my_GR[1], 
            title = "Oocytes",
            show_legend = FALSE,
            rowtitle_fs = 0,
            CpG_n_fs = 2.5,
            nt_up = prom_region[1],
            nt_down = prom_region[2],
            sc_or_bulk = NULL,
            genotype = c("X", "XX"),
            stage = c("MII Oocyte") ,
            #split_by = "individual",
            split_by = "Stage",
            rm_low_quality = FALSE,
            rm_cell_with_only_missing_values = TRUE,
            cell_per_type = cell_per_type,
            clustRow = FALSE,
            h_width = 4,
            h_height = 2,
            fs = 0)),  labels = c('A')))

B <- cowplot::plot_grid(grid.grabExpr(
  draw(promoter_methylation_fig(GR = my_GR[1], 
            title = "Sperm",
            show_legend = FALSE,
            rowtitle_fs = 0,
             CpG_n_fs = 2.5,
             nt_up = prom_region[1],
            nt_down = prom_region[2],
            sc_or_bulk = NULL,
            genotype = c("X", "XX"),
            stage = c("Sperm") ,
            #split_by = "individual",
            split_by = "Stage",
            rm_low_quality = FALSE,
            rm_cell_with_only_missing_values = TRUE,
            cell_per_type = cell_per_type,
            clustRow = FALSE,
            h_width = 4,
            h_height = 2,
            fs = 0)),  labels = c('A')))

figA <- cowplot::plot_grid(A, B, 
                   # rel_widths = c(2,1),
                   # scale = c(1, 0.8),
                   # label_size = 10, 
                   nrow = 2,
                   align = "h")

# pdf(file = paste0("../figs/Fig8C_", gene, "_Methylation_Oocytes_Sperm.pdf"))#, height = 3, width =4)
# figA
# dev.off()

# Show the same number of cells in male and female embryos
# Define the number of cells per stage manually
cell_per_type <- tibble(Stage = c("Zygote", #"2-cell", "4-cell", "8-cell", 
                                  "Morula", "Blastocyst", "Post-implantation"),
                        n = c(6, #8, 8, 8, 
                              10, 12, 4))

cell_per_type <- tibble(Stage = c("Zygote", #"2-cell", "4-cell", "8-cell", 
                                  "Morula", "Blastocyst", "Post-implantation"),
                        n = c(10, #8, 8, 8, 
                              10, 20, 4))

D <- cowplot::plot_grid(grid.grabExpr(draw(
  promoter_methylation_fig(GR = my_GR[1], 
                           title = "Female",
                           show_legend = FALSE,
                           rowtitle_fs = 8,
                           coltitle_fs = 10,
                           CpG_n_fs = 2.5,
                           nt_up = prom_region[1],
                           nt_down = prom_region[2],
                           sc_or_bulk = "single",
                           genotype = c("X", "XX"),
                           stage = c("PN-early", "PN-mid", "PN-late", "2-cell", #"4-cell", 
                                     "8-cell", "Morula", "Blastocyst (Mixed)", 
                                     "Blastocyst (ICM)", "Heart", "Villus"),    
                           #split_by = "individual",
                           split_by = "Stage",
                           rm_low_quality = FALSE, # otherwise not enough samples
                           rm_cell_with_only_missing_values = TRUE,
                           cell_per_type = cell_per_type,
                           clustRow = FALSE,
                           h_width = 4,
                           h_height =5,
                           fs = 8)),  labels = c('A')))

E <- cowplot::plot_grid(grid.grabExpr(draw(
  promoter_methylation_fig(GR = my_GR[1], 
                           title = "Male",
                           rowtitle_fs = 8,
                           coltitle_fs = 10,
                          CpG_n_fs = 2.5,
                           nt_up = prom_region[1],
                           nt_down = prom_region[2],
                           sc_or_bulk = "single",
                           genotype = c("Y", "XY"),
                           stage = c("PN-early", "PN-mid", "PN-late", "2-cell", "4-cell", 
                                     "8-cell", 
                                     "Morula", "Blastocyst (Mixed)", "Blastocyst (ICM)", "Heart", "Villus"),       
                           #split_by = "individual",
                           split_by = "Stage",
                           rm_low_quality = FALSE, # otherwise not enough samples
                           rm_cell_with_only_missing_values = TRUE,
                           cell_per_type = cell_per_type,
                           clustRow = FALSE,
                           h_width = 4,
                           h_height =5,
                           fs = 8)),  labels = c('B')))

figB<- cowplot::plot_grid(D,E, 
                   # rel_widths = c(2,1),
                   # scale = c(1, 0.8),
                   # label_size = 10, 
                   nrow = 1,
                   align = "h")

# pdf(file = paste0("../figs/Fig8C_", gene, "_Methylation_embryos.pdf"))#, height = 3, width = 3)
# figB
# dev.off()

figC <- cowplot::plot_grid(figA, figB,
                    rel_heights = c(1,1.6),
                   # scale = c(1, 0.8),
                   # label_size = 10, 
                   nrow = 2,
                   align = "h",
                   labels = my_gene)

pdf(file = paste0("../figs/Fig8C_", gene, "_Methylation_embryos.pdf"), height = 3.5, width = 4.5)
figC
dev.off()

figC
```

# MAGEB6 Figures

```{r}
my_gene <- "MAGEB6"
my_GR <- all_promoter_regions_hg19[all_promoter_regions_hg19$external_gene_name == my_gene,]

promoter_methylation_by_GR(GR = my_GR[1], 
                           nt_up = 1000,
                           nt_down = 500,
                           sc_or_bulk = NULL,
                           #genotype = "XX",
                           #stage = "MII Oocyte",
                           #split_by = "individual",
                           split_by = "Stage",
                           rm_low_quality = FALSE,
                           rm_cell_with_only_missing_values = TRUE,
                           clustRow = FALSE,
                           fs = 2)
```

Show Promoter region -/- 1000 pb around TSS

```{r}
prom_region <- c(1000, 1000)

my_GR <- all_promoter_regions_hg19[all_promoter_regions_hg19$external_gene_name == my_gene,]
gene <- my_GR$external_gene_name

cell_per_type <- tibble(Stage = c("MII Oocyte", "Sperm"),
                        n = c(10, 10))

A <- cowplot::plot_grid(grid.grabExpr(
  draw(promoter_methylation_fig(GR = my_GR[1], 
            title = "Oocytes",
            show_legend = FALSE,
            rowtitle_fs = 0,
            CpG_n_fs = 2.5,
            nt_up = prom_region[1],
            nt_down = prom_region[2],
            sc_or_bulk = NULL,
            genotype = c("X", "XX"),
            stage = c("MII Oocyte") ,
            #split_by = "individual",
            split_by = "Stage",
            rm_low_quality = FALSE,
            rm_cell_with_only_missing_values = TRUE,
            cell_per_type = cell_per_type,
            clustRow = FALSE,
            h_width = 4,
            h_height = 2,
            fs = 0)),  labels = c('A')))

B <- cowplot::plot_grid(grid.grabExpr(
  draw(promoter_methylation_fig(GR = my_GR[1], 
            title = "Sperm",
            show_legend = FALSE,
            rowtitle_fs = 0,
             CpG_n_fs = 2.5,
             nt_up = prom_region[1],
            nt_down = prom_region[2],
            sc_or_bulk = NULL,
            genotype = c("X", "XX"),
            stage = c("Sperm") ,
            #split_by = "individual",
            split_by = "Stage",
            rm_low_quality = FALSE,
            rm_cell_with_only_missing_values = TRUE,
            cell_per_type = cell_per_type,
            clustRow = FALSE,
            h_width = 4,
            h_height = 2,
            fs = 0)),  labels = c('A')))

figA <- cowplot::plot_grid(A, B, 
                   # rel_widths = c(2,1),
                   # scale = c(1, 0.8),
                   # label_size = 10, 
                   nrow = 2,
                   align = "h")

# pdf(file = paste0("../figs/Fig8C_", gene, "_Methylation_Oocytes_Sperm.pdf"))#, height = 3, width =4)
# figA
# dev.off()

# Show the same number of cells in male and female embryos
# Define the number of cells per stage manually
cell_per_type <- tibble(Stage = c("Zygote", #"2-cell", "4-cell", "8-cell", 
                                  "Morula", "Blastocyst", "Post-implantation"),
                        n = c(6, #8, 8, 8, 
                              10, 12, 4))

cell_per_type <- tibble(Stage = c("Zygote", #"2-cell", "4-cell", "8-cell", 
                                  "Morula", "Blastocyst", "Post-implantation"),
                        n = c(10, #8, 8, 8, 
                              10, 20, 4))

D <- cowplot::plot_grid(grid.grabExpr(draw(
  promoter_methylation_fig(GR = my_GR[1], 
                           title = "Female",
                           show_legend = FALSE,
                           rowtitle_fs = 8,
                           coltitle_fs = 10,
                           CpG_n_fs = 2.5,
                           nt_up = prom_region[1],
                           nt_down = prom_region[2],
                           sc_or_bulk = "single",
                           genotype = c("X", "XX"),
                           stage = c("PN-early", "PN-mid", "PN-late", "2-cell", #"4-cell", 
                                     "8-cell", "Morula", "Blastocyst (Mixed)", 
                                     "Blastocyst (ICM)", "Heart", "Villus"),    
                           #split_by = "individual",
                           split_by = "Stage",
                           rm_low_quality = FALSE, # otherwise not enough samples
                           rm_cell_with_only_missing_values = TRUE,
                           cell_per_type = cell_per_type,
                           clustRow = FALSE,
                           h_width = 4,
                           h_height =5,
                           fs = 8)),  labels = c('A')))

E <- cowplot::plot_grid(grid.grabExpr(draw(
  promoter_methylation_fig(GR = my_GR[1], 
                           title = "Male",
                           rowtitle_fs = 8,
                           coltitle_fs = 10,
                          CpG_n_fs = 2.5,
                           nt_up = prom_region[1],
                           nt_down = prom_region[2],
                           sc_or_bulk = "single",
                           genotype = c("Y", "XY"),
                           stage = c("PN-early", "PN-mid", "PN-late", "2-cell", "4-cell", 
                                     "8-cell", 
                                     "Morula", "Blastocyst (Mixed)", "Blastocyst (ICM)", "Heart", "Villus"),       
                           #split_by = "individual",
                           split_by = "Stage",
                           rm_low_quality = FALSE, # otherwise not enough samples
                           rm_cell_with_only_missing_values = TRUE,
                           cell_per_type = cell_per_type,
                           clustRow = FALSE,
                           h_width = 4,
                           h_height =5,
                           fs = 8)),  labels = c('B')))

figB<- cowplot::plot_grid(D,E, 
                   # rel_widths = c(2,1),
                   # scale = c(1, 0.8),
                   # label_size = 10, 
                   nrow = 1,
                   align = "h")

# pdf(file = paste0("../figs/Fig8C_", gene, "_Methylation_embryos.pdf"))#, height = 3, width = 3)
# figB
# dev.off()

figC <- cowplot::plot_grid(figA, figB,
                    rel_heights = c(1,1.6),
                   # scale = c(1, 0.8),
                   # label_size = 10, 
                   nrow = 2,
                   align = "h",
                   labels = my_gene)

pdf(file = paste0("../figs/Fig8C_", gene, "_Methylation_embryos.pdf"), height = 3.5, width = 4.5)
figC
dev.off()

figC
```

# Session info

```{r}
sessionInfo()
```

### 2025-03-25 7.DE_Zhu_pseusobulk.Rmd
Added code used to generate the paper

---
title: "DE analysis, F vs M embryos, Zhu scWGBS data"
subtitle: "pseudo-bulk analysis"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r, echo = FALSE, Results = 'asis'}
suppressPackageStartupMessages(library("dplyr"))
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
library("tidyverse")
library("ComplexHeatmap")
library("CTexploreR")
library("circlize")
library("SummarizedExperiment")
library("SingleCellExperiment")
library("DESeq2")
library("org.Hs.eg.db")
library("DropletUtils")
library("scater")
library("scran")
library("scuttle")
```

```{r}
load("~/cluster/Packages/CTdata/eh_data/all_genes.rda")

CT_genes_X_met <- all_genes %>% 
  filter(CT_gene_type == "CT_gene") %>% 
  filter(X_linked, regulated_by_methylation) %>% 
  pull(external_gene_name)
```

Data from:
[Single-cell DNA methylome sequencing of human preimplantation embryos. Nat Genet 50, 12–19 (2018). Zhu, P., Guo, H., Ren, Y. et al.](https://doi.org/10.1038/s41588-017-0007-6)

**Started from raw files**

- Downloaded all fastq files (including scRNAseq fastq files from ICM and TE cells)

see `/storage/research/dduv/cbio-lg/cluster/DataSets/early_embryos/Zhu_dataset/scRNAseq_reprocessed_data/` folder

- All files were reprocessed, in order to:

-- Show expression values in TPM rather than FPKM (previous data was downloaded on GEO as an count matrix of already processed data in FPKM values)

-- Use all the cells for the differential expression analysis instead of using only the few available "bulk samples"

-- Test in parallel the expression of CG X-linked MethDep when multimapping was allowed

# Pre-processing

## Quality control

```{r}
load("/storage/research/dduv/cbio-lg/cluster/DataSets/early_embryos/Zhu_dataset/scRNAseq_reprocessed_data/processed_data/count_matrix/coldata.rda")
load("/storage/research/dduv/cbio-lg/cluster/DataSets/early_embryos/Zhu_dataset/scRNAseq_reprocessed_data/processed_data/count_matrix/raw_counts.rda")

coldata <- as.data.frame(coldata) 
rownames(coldata) <- coldata$sample
coldata$gender <- ifelse(coldata$gender == "male", "M", "F")

sce <- SingleCellExperiment(assays = counts[, coldata$sample],
                      colData = coldata)

# No mitochondrial genes (not in GTF or in genome ref used for alignment?)
# grep("^MT-", x = mapIds(org.Hs.eg.db, rownames(sce), "SYMBOL", "ENSEMBL"), value = TRUE)

sce <- sce[rowSums(assay(sce)) > 0,]

sce$gender <- factor(sce$gender, levels = c("M", "F"))

rowData(sce)$gene <- mapIds(org.Hs.eg.db, rownames(sce), "SYMBOL", "ENSEMBL")

assayNames(sce) <- "counts"
rowData(sce)$is_mito <- FALSE
rowData(sce)$is_mito[grep(pattern = "^MT-", x = rowData(sce)$gene, value = FALSE)] <- TRUE
sce <- addPerCellQC(sce, subsets=list(Mito = rowData(sce)$is_mito))
# No MT genes
#table(sce$subsets_Mito_percent)

table(sce$cell_type, sce$gender)
```

### Number of detected genes

```{r}
thr_detection <- 7500

# Define outliers with low detection level
outliers <- colnames(sce)[isOutlier(sce$detected, type = "lower")]
outliers <- colnames(sce)[sce$detected < thr_detection]
##outliers

sce$outlier_detected <- FALSE
sce$outlier_detected[sce$sample %in% outliers] <- TRUE

f1 <- as_tibble(colData(sce)) %>% 
  ggplot(aes(x = 1, y = detected)) +
  geom_jitter(aes(x = 1, y = detected, color = outlier_detected)) +
  geom_violin(alpha = 0.5, outliers = F) +
  geom_hline(yintercept = thr_detection, linetype = "dashed") +
  theme(legend.position = "bottom",
        legend.title.position =  "top",
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6),
                axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  ggtitle("Number of detected genes")

f2 <- as_tibble(colData(sce)) %>% 
  ggplot(aes(x = 1, y = detected)) +
  geom_jitter(aes(x = 1, y = detected, color = bulk_or_single_cell)) +
  geom_violin(alpha = 0.5, outliers = F) +
    scale_color_manual(values = c("blue", "cyan3"))+
  theme(legend.position = "bottom",
        legend.title.position =  "top",
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6),
                axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())+
  ggtitle("Number of detected genes")

gridExtra::grid.arrange(f2, f1, ncol = 3)
```

cells with less than `r thr_detection` are considered as outliers

### Sequencing depth

```{r}
thr_sum <- 7000000
outliers <- colnames(sce)[sce$sum < thr_sum]

sce$outlier_sum <- FALSE
sce$outlier_sum[sce$sample %in% outliers] <- TRUE

f1 <- as_tibble(colData(sce)) %>% 
  ggplot(aes(x = 1, y = sum)) +
  geom_jitter(aes(x = 1, y = sum, color = outlier_sum)) +
  geom_violin(alpha = 0.5, outliers = F) +
  geom_hline(yintercept = thr_sum, linetype = "dashed") +
  theme(legend.position = "bottom",
        legend.title.position =  "top",
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6),
                axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  ggtitle("Total number of reads")

f3 <- as_tibble(colData(sce)) %>% 
  ggplot(aes(x = 1, y = sum)) +
  geom_jitter(aes(x = 1, y = sum, color = outlier_detected)) +
  geom_violin(alpha = 0.5, outliers = F) +
  geom_hline(yintercept = thr_sum, linetype = "dashed") +
  theme(legend.position = "bottom",
        legend.title.position =  "top",
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6),
                axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  ggtitle("Total number of reads")

f2 <- as_tibble(colData(sce)) %>% 
  ggplot(aes(x = 1, y = sum)) +
  geom_jitter(aes(x = 1, y = sum, color = bulk_or_single_cell)) +
  geom_violin(alpha = 0.5, outliers = F) +
  scale_color_manual(values = c("blue", "cyan3"))+
  theme(legend.position = "bottom",
        legend.title.position =  "top",
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6),
                axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())+
  ggtitle("Total number of reads")

gridExtra::grid.arrange(f2, f3, f1, ncol = 3)
```

cells with less than `r thr_sum` are considered as outliers

### Ouliers identification

```{r}
outliers <- as_tibble(colData(sce)) %>% 
  filter(outlier_sum | outlier_detected) %>% 
  pull(sample)

sce$outlier <- FALSE
sce$outlier[sce$sample %in% outliers] <- TRUE

# Define outliers with low sum value
f1 <- as_tibble(colData(sce)) %>% 
  ggplot(aes(x = 1, y = detected)) +
  geom_jitter(aes(x = 1, y = detected, color = outlier)) +
  geom_violin(alpha = 0.5, outliers = F) +
  theme(legend.position = "bottom",
        legend.title.position =  "top",
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

f2 <- as_tibble(colData(sce)) %>% 
  ggplot(aes(x = 1, y = sum)) +
  geom_jitter(aes(x = 1, y = sum, color = outlier)) +
  geom_violin(alpha = 0.5, outliers = F) +
  theme(legend.position = "bottom",
        legend.title.position =  "top",
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

gridExtra::grid.arrange(f1, f2, ncol = 2)
```

=> Removed outlier cells

```{r}
table(outlier =sce$outlier)

sce <- sce[, !sce$outlier]
```

=> `r dim(sce)[2]` cells kept

```{r}
table(sce$cell_type, sce$gender)
```

### No link between data's quality and sex factor

Checked that the data's quality is independant of sex factor:

```{r}
a <- as_tibble(colData(sce)) %>% 
  ggplot(aes(x = gender, y = detected)) +
  geom_jitter(aes(x = gender, y = detected, color = gender)) +
  geom_boxplot(alpha = 0.5, outliers = F) +
    scale_color_manual(values = c("deeppink", "steelblue"))+
  theme(legend.position = "bottom",
        legend.title.position =  "top",
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())+
  ggtitle("Number of detected genes")

b <- as_tibble(colData(sce)) %>% 
  ggplot(aes(x = gender, y = sum)) +
  geom_jitter(aes(x = gender, y = sum, color = gender)) +
  geom_boxplot(alpha = 0.5, outliers = F) +
    scale_color_manual(values = c("deeppink", "steelblue"))+
  theme(legend.position = "bottom",
        legend.title.position =  "top",
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())+
  ggtitle("Sequencing Depth")

gridExtra::grid.arrange(a,b, ncol = 2)
```

### Normalisation

```{r}
## Normalisation
sce <- logNormCounts(sce, transform = "none")
sce <- logNormCounts(sce)

# Normalise into TPMs for visualisations
featureC_results <- read_tsv("/storage/research/dduv/cbio-lg/cluster/DataSets/early_embryos/Zhu_dataset/scRNAseq_reprocessed_data/processed_data/featurecounts/SRR6010592_featurecounts.tsv", 
                             skip = 1, col_names = TRUE)

# Check the order of the genes
#unique(rownames(counts) == featureC_results$Geneid)
x1 <- counts / featureC_results$Length * 1000
total <- colSums(x1)

TPM_matrix <- as_tibble(x1, rownames = "ensembl_gene_id") %>% 
  gather(sample, counts, -ensembl_gene_id) %>% 
  left_join(enframe(total) %>% 
              dplyr::rename(sample = name, total = value)) %>% 
  mutate(TPM = counts / total * 1000000) %>% 
  dplyr::select(ensembl_gene_id, sample, TPM) %>% 
  spread(sample, TPM) 
TPM <- as.matrix(TPM_matrix[,-1])
rownames(TPM) <- TPM_matrix$ensembl_gene_id

assay(sce, "TPM") <- TPM[rownames(sce), colnames(sce)]
```

# CG genes Heatmaps

- Show X-linked MethDep CG genes detected in at least 20% of cells

- Added Ubiquitous and Y-linked genes as controls

```{r}
my_genes <- all_genes %>% 
  filter(CT_gene_type == "CT_gene") %>% 
  filter(X_linked, regulated_by_methylation) %>% 
  filter(ensembl_gene_id %in% rownames(sce))

legend_colors <- c(
  "#5E4FA2", "#3288BD", "#66C2A5", "#ABDDA4", "#E6F598",
  "#FFFFBF", "#FEE08B", "#FDAE61", "#F46D43", "#D53E4F",
  "#9E0142")

legends_param <- list(
  labels_gp = gpar(col = "black", fontsize = 7),
  title_gp = gpar(col = "black", fontsize = 7, fontface = "bold"),
  simple_anno_size = unit(0.1, "cm"),
  row_names_gp = gpar(fontsize = 7),
  annotation_name_side = "left",
  border = FALSE,
  border_gp = gpar(lwd = 0.1),
  grid_width = unit(0.2, "cm"),
  grid_height = unit(0.2, "cm"),
  legend_height = unit(0.2, "cm"))

draw_embryo_exp_Zhu_dataset <- function(genes, 
                            rm_outlier = FALSE,
                            cell_type = NULL,
                            detected_in_percent_of_cells = 0,
                            split_rows_by = "chr_met", 
                            font_size = 6,
                            gender = NULL,
                            logTransform = FALSE,
                            assayName = "TPM",
                            scale_lims = NULL,
                            clustRow = TRUE,
                            h_width = 8, 
                            h_height = 8, 
                            show_right_annot = TRUE,
                            clust_method = "centroid"){

  # Top Annotations
  df_col <- colData(sce)

  if(rm_outlier) df_col <- df_col[!df_col$outlier,]
  fig_title <- "Blastocyst cells"
  if(!is.null(cell_type)) {
    df_col <- df_col[df_col$cell_type == cell_type,]
    fig_title <- paste0(cell_type, " cells")
  }

  df_col$sex <- df_col$gender
  levels(df_col$sex) <- c("Male", "Female")
  df_col <- df_col[order(df_col$sex, df_col$cell_type ),]
  df_col$Stage <- df_col$cell_type
  df_col$bulk_or_single_cell <- df_col$bulk_or_single_cell
  df_col$group <- paste0(df_col$sex, " ", df_col$cell_type, " cells")

  column_ha_sex = HeatmapAnnotation(
    sex = df_col$sex,
    col = list(sex = c("Female" = "deeppink", "Male" = "steelblue")),
        border = FALSE, 
    simple_anno_size = unit(0.2, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

    column_ha_Stage = HeatmapAnnotation(
    CellType = df_col$Stage,
    col = list(CellType = c("ICM" = "gray", "TE" = "gray10")),
    border = FALSE, 
    simple_anno_size = unit(0.2, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

  column_ha_bulk = HeatmapAnnotation(
    bulk_or_single_cell = df_col$bulk_or_single_cell,
    col = list(bulk_or_single_cell = c("single cell" = "cyan3", "bulk cells" = "blue")),
    border = FALSE, 
    simple_anno_size = unit(0.2, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

  column_ha_outlier = HeatmapAnnotation(
    outlier = df_col$outlier,
    col = list(outlier = c("TRUE" = "cyan", "FALSE" = "gray10")),
    border = FALSE, 
    simple_anno_size = unit(0.2, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

  column_ha_sample = HeatmapAnnotation(
    sample = df_col$source_name,
    #col = list(outlier = c("TRUE" = "cyan", "FALSE" = "gray10")),
    border = FALSE, 
    simple_anno_size = unit(0.2, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

  column_ha_detected = HeatmapAnnotation(
    detected = df_col$detected,
    #col = list(outlier = c("TRUE" = "cyan", "FALSE" = "gray10")),
    border = FALSE, 
    simple_anno_size = unit(0.2, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

  ensembl_gene <- data.frame(external_gene_name = all_genes$external_gene_name)  
  rownames(ensembl_gene) <- all_genes$ensembl_gene_id

  genes <- genes[genes %in% rownames(sce)]

  mat <- assay(sce, assayName)[genes, ]
  rownames(mat) <- ensembl_gene[rownames(mat),]

  # Show only genes detected in more than x% of cells
  mat <- mat[rowSums(mat > 0) >= dim(mat)[2]*detected_in_percent_of_cells,]

  if (logTransform) mat <- log1p(mat)

   if (is.null(scale_lims)) scale_lims <- c(0, max(mat))
  heatmap_colors <- colorRamp2(
    seq(scale_lims[1], scale_lims[2], length = 11), legend_colors)

  lines <- tibble(gene = rownames(mat)) %>% 
    mutate(group = case_when(gene %in% CT_genes$external_gene_name ~ "X-linked CG MethDep",
                             !gene %in% CT_genes$external_gene_name ~ "Control")) %>% 
    mutate(group = factor(group, levels = c("X-linked CG MethDep", "Control")))

  ht1 <- Heatmap(mat[, rownames(df_col)],
                 name = assayName,
                 #column_title = fig_title,
                 clustering_method_rows = "ward.D2",
                 clustering_method_columns = "ward.D2",
                 column_split = df_col$sex,
                 row_split = lines$group,
                 row_gap = unit(c(1), "mm"),
                 col = heatmap_colors,
                 cluster_rows = clustRow,
                 cluster_columns = FALSE,
                 cluster_row_slices = FALSE,
                 cluster_column_slices = FALSE,
                 show_row_names = TRUE,
                 show_column_names = FALSE,
                 show_row_dend = FALSE,
                 column_names_gp = gpar(fontsize = 8),
                 column_names_centered = TRUE,
                 row_names_gp = gpar(fontsize = font_size),
                 row_names_side = "left",
                 border = TRUE,
                 border_gp = gpar(lwd = 0.5),
                 column_names_side = c("bottom"),
                 column_names_rot = 90,
                 column_title_gp = gpar(fontsize = 10, face = "bold"),
                 row_title_gp = gpar(fontsize = 0),
                 top_annotation = c(column_ha_Stage,
                                     column_ha_sex,
                 #                    # column_ha_sample,
                                     gap = unit(0.1, "mm")),
                 # use_raster = TRUE,
                 # raster_device = "CairoPNG",
                 # raster_quality = 10,
                 width = unit(h_width, "cm"),
                 height = unit(h_height, "cm"),
                 heatmap_legend_param = legends_param)

  ht1
}

controls <- all_genes %>% 
  filter(external_gene_name %in% c("DDX3Y", "RPS4Y1", "EIF1AY"))

fig <- draw_embryo_exp_Zhu_dataset(genes = c(my_genes$ensembl_gene_id, controls$ensembl_gene_id), 
                            clustRow = T, 
                            cell_type = NULL,
                            font_size = 5,
                            scale_lims = c(0, 100),
                            h_width = 5,
                            h_height = 4,
                            logTransform = F,
                            detected_in_percent_of_cells = 0.2)

pdf(file = "../figs/Fig8B_Heatmap_CG_metDep_in_Zhu_dataset.pdf", width = 6, height = 4)
fig

dev.off()

fig 
```

NB: Are represented on the heatmap only genes detected in more than 20% of cells

# Differential expression analysis

**pseudo-bulk samples**

Group cells by embryo

Kept embryos with more than one cell

```{r}
dds_pb <- scuttle::summarizeAssayByGroup(sce,
                            ids = colData(sce)[, "source_name"],
                            statistics = "sum")

coldata <- tibble(source_name = dds_pb$ids,
                  ncells = dds_pb$ncells) %>%
  left_join(as_tibble(colData(sce)) %>% 
              dplyr::select(source_name, cell_type, gender) %>% 
              unique()) %>%
  as.data.frame()

rownames(coldata) <- coldata$source_name

dds_pb$cell_type <- coldata[colnames(dds_pb),]$cell_type
dds_pb$gender <- coldata[colnames(dds_pb),]$gender
dds_pb$sample <- coldata[colnames(dds_pb),]$source_name
dds_pb$embryo <- paste0("embryo", gsub("Te", '', gsub("Icm", '', x = gsub("-\\w*", '', x = dds_pb$sample))))

dds_pb <- dds_pb[, dds_pb$ncells > 1]

as_tibble(colData(dds_pb)) %>% 
  DT::datatable()

dds <- DESeqDataSetFromMatrix(countData = assay(dds_pb), 
                              colData = colData(dds_pb),
                              design = ~ cell_type + gender)

dds$gender <- factor(dds$gender, levels = c("M", "F"))

dds <- estimateSizeFactors(dds)
filtering_thr <- 10
keep <- rowSums(counts(dds, normalize = TRUE) >= filtering_thr ) >= 3
dds <- DESeq(dds[keep,])

table(dds$cell_type, dds$gender)
```

## PCA

```{r}
## PCA
rld <- varianceStabilizingTransformation(dds)
rv <- rowVars(assay(rld))
selected <- order(rv, decreasing = TRUE)[seq_len(min(500, length(rv)))]
pca <- prcomp(t(assay(rld)[selected, ]))
percentVar <- pca$sdev^2/sum(pca$sdev^2)
percentVar_1 <- paste0("PC1: ", round(percentVar[1] * 100), "% variance")
percentVar_2 <- paste0("PC2: ", round(percentVar[2] * 100), "% variance")
percentVar_3 <- paste0("PC3: ", round(percentVar[3] * 100), "% variance")
percentVar_4 <- paste0("PC4: ", round(percentVar[4] * 100), "% variance")
pcaData <- as_tibble(pca$x, rownames = "samples") %>%
  dplyr::select(samples, "PC1", "PC2", "PC3", "PC4") %>%
  left_join(as_tibble(colData(dds), rownames = "samples"))

ggplot(pcaData, aes(x = PC1, y = PC2, col = gender, shape = cell_type, label = sample)) +
  geom_point(size = 4) +
  xlab(percentVar_1) +
  ylab(percentVar_2) +
  geom_text(color = "black", size = 2)

ggplot(pcaData, aes(x = PC1, y = PC2, col = cell_type, shape = gender, label = sample)) +
  geom_point(size = 4) +
  xlab(percentVar_1) +
  ylab(percentVar_2) +
  geom_text(color = "black", size = 2)

ggplot(pcaData, aes(x = PC1, y = PC2, col = embryo, shape = gender, label = sample)) +
  geom_point(size = 4) +
  xlab(percentVar_1) +
  ylab(percentVar_2) +
  geom_text(color = "black", size = 2)
```

## Volcano plot

```{r}
resultsNames(dds)
res_dds <- results(dds, 
                   name = "gender_F_vs_M", 
                   independentFiltering = FALSE)

#load("~/cluster/CBIO-templates/ensembl_2_geneName/transcripts_infos.rda")

res_tbl <- as_tibble(res_dds, rownames = "ensembl_gene_id") %>%
  left_join(all_genes %>% dplyr::select(ensembl_gene_id, external_gene_name, chr) %>% unique()) %>%
  arrange(padj)

padj_thr <- 0.05
log2FC_thr <- 1

ggplot(res_tbl %>% filter(!is.na(padj)), aes(x = pvalue)) + 
    geom_histogram(bins = 20, boundary = 0) +
  ggtitle("pvalues histogram")

res_tbl %>% 
  filter(!is.na(padj)) %>% 
  ggplot(aes(x= log2FoldChange, y = -log10(padj))) +
  geom_point(size = 0.5) +
    geom_point(data = res_tbl %>% 
               filter(chr == "X"), 
             aes(x= log2FoldChange, y = -log10(padj)), 
             color = "steelblue", size = 0.55) +
  geom_point(data = res_tbl %>% 
               filter(external_gene_name %in% CT_genes$external_gene_name), 
             aes(x= log2FoldChange, y = -log10(padj)), 
             color = "red", size = 0.55) +
  geom_point(data = res_tbl %>% 
               filter(chr == "Y"), 
             aes(x= log2FoldChange, y = -log10(padj)), 
             color = "black", size = 0.55) +
  ylab("-log10(padj)") +
  ggrepel::geom_text_repel(data = subset(res_tbl, external_gene_name %in% CT_genes$external_gene_name),
                           aes(x = log2FoldChange, y = -log10(padj), label = external_gene_name),
                           size = 2,
                           segment.size = 0.1, max.overlaps = 30) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray70") +
  geom_hline(yintercept = -log10(padj_thr), linetype = "dashed", color = "gray70") 
```

### Volcano plot for paper

```{r}
load("~/cluster/Packages/CTdata/eh_data/CT_genes.rda")

CTP_genes <- CT_genes %>% 
  filter(CT_gene_type == "CTP_gene")

CT_genes <- CT_genes %>% 
  filter(CT_gene_type == "CT_gene")

thr <- 1e-25
logFC_thr <- 10
res_tbl_modified <- res_tbl %>% 
  mutate(group = case_when(padj < thr ~ "padj_outlier",
                           abs(log2FoldChange) >= logFC_thr ~ "logFC_outlier",
                           padj > thr & abs(log2FoldChange) < logFC_thr ~ "others")) %>% 
  mutate(padj = ifelse(padj <= thr, thr, padj)) %>% 
  mutate(log2FoldChange = ifelse(log2FoldChange >= logFC_thr, logFC_thr, log2FoldChange)) %>% 
   mutate(log2FoldChange = ifelse(log2FoldChange <= -logFC_thr, -logFC_thr, log2FoldChange)) %>% 
  filter(chr %in% c(1:22, "X", "Y")) %>% 
  mutate(chr = case_when(chr %in% c("X", "Y") ~ chr,
                         !chr %in% c("X", "Y") ~ "Autosome")) %>% 
  mutate(X_CG_MethDEp = case_when(external_gene_name %in% CT_genes$external_gene_name ~ TRUE,
                             external_gene_name %in% CT_genes$external_gene_name ~ FALSE)) %>% 
  filter(!is.na(padj)) %>% 
  filter(!is.na(group)) 
  #filter(abs(log2FoldChange) > 0.05)

labeled_X_genes <- res_tbl_modified %>% 
  filter(chr == "X") %>% 
  filter(abs(log2FoldChange) > 5  & padj < 0.005) %>% 
  filter(!external_gene_name %in% CT_genes$external_gene_name) %>%
  arrange(padj)

labeled_CT_genes <- res_tbl_modified %>% 
  filter(external_gene_name %in% CT_genes_X_met & padj < 0.05 & log2FoldChange > 2) %>% 
  arrange(padj)

X_sign_genes <- res_tbl_modified %>% 
  filter(chr == "X") %>% 
  filter(padj < padj_thr) %>% 
  arrange(padj) #%>% head(30)

Y_sign_genes <- res_tbl_modified %>% 
  filter(chr == "Y") %>% 
  #filter(padj < padj_thr) %>% 
  filter(log2FoldChange < 0 ) %>% 
  arrange(padj) #%>% head(30)

set.seed = 123
fig <- res_tbl_modified %>% 
  ggplot(aes(x= log2FoldChange, y = -log10(padj))) +
  geom_point(data = res_tbl_modified,
             aes(x= log2FoldChange, y = -log10(padj), shape = group),
             size = 1, color = "gray80") + 
    geom_point(data = res_tbl_modified %>% 
               filter(external_gene_name %in% X_sign_genes$external_gene_name),
             aes(x= log2FoldChange, y = -log10(padj), shape = group),
             size = 1, color = "steelblue") +
  geom_point(data = res_tbl_modified %>% 
               filter(ensembl_gene_id %in% Y_sign_genes$ensembl_gene_id),
             aes(x= log2FoldChange, y = -log10(padj), shape = group),
             size = 1, color = "black") +
  geom_point(data = res_tbl_modified %>% 
               filter(external_gene_name %in% CT_genes_X_met),# & padj < padj_thr),
             aes(x= log2FoldChange, y = -log10(padj), shape = group),
             size = 1.2, color = "firebrick2") +
  scale_shape_manual(values = c(15, 19, 17))+
  xlab("log2FoldChange") + 
  ylab("-log10(padj)") +
  geom_vline(xintercept = 0, linetype = "dashed", 
             color = "black", linewidth = 0.2) +
  # geom_vline(xintercept = -1, linetype = "dashed",
  #            color = "black", linewidth = 0.2) +
  geom_hline(yintercept = -log10(padj_thr), 
             linetype = "dashed", color = "black", linewidth = 0.2) +
  ggrepel::geom_text_repel(data = subset(res_tbl_modified, external_gene_name %in% labeled_X_genes$external_gene_name),
                           aes(x = log2FoldChange, y = -log10(padj), label = external_gene_name),
                           size = 2, color = "steelblue",
                           segment.size = 0.1, max.overlaps = 10) +
   ggrepel::geom_text_repel(data = subset(res_tbl_modified, external_gene_name %in% labeled_CT_genes$external_gene_name),
                           aes(x = log2FoldChange, y = -log10(padj), label = external_gene_name),
                           size = 2, color = "firebrick2",
                           segment.size = 0.1, max.overlaps = 30) +
  ggtitle("Female versus male cells blastocysts cells (Zhu dataset)") +
  theme_bw() +
  theme(legend.position = "none",
        title = element_text(size = 10),
        panel.grid = element_blank()) 
fig

pdf(file = "../figs/Fig8D_Volcano_Zhu_dataset.pdf", width = 4, height = 4)
fig

dev.off()

fig 
```

- Genes with a padjusted value < `r thr` are represented as triangles

- Genes with an absolute log2FC > `r logFC_thr` are represented as squares

- Red dots = X-linked MethDep genes

- Blue dots = X-linked genes not CG genes

- Black dots = Y-linked genes

- Gray dots = other genes

# Session info

```{r}
sessionInfo()
```

### 2025-03-25 8.DE_Petropoulos_pseudobulk.Rmd
Added code used to generate the paper

---
title: "Early embryo analysis"
subtitle: "Petropoulos dataset"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r, echo = FALSE, Results = 'asis'}
suppressPackageStartupMessages(library("dplyr"))
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
library("tidyverse")
library("ComplexHeatmap")
library("CTexploreR")
library("circlize")
library("DESeq2")
library("SummarizedExperiment")
library("SingleCellExperiment")
library("DropletUtils")
library("scater")
library("scran")
library("scuttle")
```

#

```{r}
legends_param <- list(
  labels_gp = gpar(col = "black", fontsize = 3),
  title_gp = gpar(col = "black", fontsize = 3, fontface = "bold"),
  simple_anno_size = unit(0.1, "cm"),
  row_names_gp = gpar(fontsize = 1),
  annotation_name_side = "left",
  border = FALSE,
  border_gp = gpar(lwd = 0.1),
  grid_width = unit(0.1, "cm"),
  grid_height = unit(0.01, "cm"),
  legend_height = unit(0.1, "cm"))
```

```{r}
CT_genes_X_met <- all_genes %>% 
  filter(CT_gene_type == "CT_gene") %>% 
  filter(X_linked, regulated_by_methylation) %>% 
  pull(external_gene_name)
```

[Single-Cell RNA-Seq Reveals Lineage and X Chromosome Dynamics in Human Preimplantation Embryos](https://pubmed.ncbi.nlm.nih.gov/27062923/)
Petropoulos et al, Cell 2016

# Quality control

```{r}
# Start from raw counts
load(file = "/storage/research/dduv/cbio-lg/cluster/DataSets/early_embryos/Petropoulos_dataset/data/Petropoulos_sce_raw_counts.rda")
sce <- petropoulos_sce_raw_counts
rm(petropoulos_sce_raw_counts)
assayNames(sce) <- "counts"

table(sce$lineage, sce$day)
```

## Check sex attribute for each indivisual

```{r}

as_tibble(colData(sce)) %>% 
  filter(day == "E3") %>% 
  ggplot(aes(x = individual, y = gender, color = gender)) +
  geom_jitter(width = 0.1)+
  theme(axis.text.x = element_text(angle = 90))+
  facet_wrap(~ day)

# "E3.1", "E3.2", "E3.3" are ambigous
sce <- sce[, !sce$individual %in% c("E3.1", "E3.2", "E3.3")]

as_tibble(colData(sce)) %>% 
  filter(day == "E4") %>% 
  ggplot(aes(x = individual, y = gender, color = gender)) +
  geom_jitter(width = 0.1)+
  theme(axis.text.x = element_text(angle = 90))+
  facet_wrap(~ day)

as_tibble(colData(sce)) %>% 
  filter(day == "E5") %>% 
  ggplot(aes(x = individual, y = gender, color = gender)) +
  geom_jitter(width = 0.1)+
  theme(axis.text.x = element_text(angle = 90))+
  facet_wrap(~ day)
# "E5.40" is likely a Female
sce$gender[sce$individual == "E5.40"] <- "F"

as_tibble(colData(sce)) %>% 
  filter(day == "E6") %>% 
  ggplot(aes(x = individual, y = gender, color = gender)) +
  geom_jitter(width = 0.1)+
  theme(axis.text.x = element_text(angle = 90))+
  facet_wrap(~ day)

# "E6.7" is likely a Male
sce$gender[sce$individual == "E6.7"] <- "M"

as_tibble(colData(sce)) %>% 
  filter(day == "E7") %>% 
  ggplot(aes(x = individual, y = gender, color = gender)) +
  geom_jitter(width = 0.1)+
  theme(axis.text.x = element_text(angle = 90))+
  facet_wrap(~ day)
```

```{r}

# No mitochondrial genes (not in GTF or in genome ref used for alignment?)
# grep("^MT-", x = rownames(sce), value = TRUE)

sce <- sce[rowSums(assay(sce)) > 0,]

sce$gender <- factor(sce$gender, levels = c("M", "F"))

rowData(sce)$is_mito <- FALSE
rowData(sce)$is_mito[grep(pattern = "^MT-", x = rowData(sce)$gene, value = FALSE)] <- TRUE
sce <- addPerCellQC(sce, subsets=list(Mito = rowData(sce)$is_mito))
# No MT genes
#table(sce$subsets_Mito_percent)

table(sce$day, sce$gender)
```

### Number of detected genes

```{r}
thr_detection <- 7500

thr_detection <- 0

# Define outliers with low detection level
#outliers <- colnames(sce)[isOutlier(sce$detected, type = "lower")]
outliers <- colnames(sce)[sce$detected < thr_detection]
##outliers

sce$outlier_detected <- FALSE
sce$outlier_detected[sce$sample %in% outliers] <- TRUE

f1 <- as_tibble(colData(sce)) %>% 
  ggplot(aes(x = 1, y = detected)) +
  geom_jitter(aes(x = 1, y = detected, color = outlier_detected)) +
  geom_violin(alpha = 0.5, outliers = F) +
  geom_hline(yintercept = thr_detection, linetype = "dashed") +
  theme(legend.position = "bottom",
        legend.title.position =  "top",
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6),
                axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  ggtitle("Number of detected genes")

# as_tibble(colData(sce)) %>% 
#   ggplot(aes(x = 1, y = detected)) +
#   geom_jitter(aes(x = 1, y = detected, color = source_name)) +
#   geom_violin(alpha = 0.5, outliers = F)

f2 <- as_tibble(colData(sce)) %>% 
  ggplot(aes(x = 1, y = detected)) +
  geom_jitter(aes(x = 1, y = detected, color = day)) +
  geom_violin(alpha = 0.5, outliers = F) +
    #scale_color_manual(values = c("blue", "cyan3"))+
  theme(legend.position = "bottom",
        legend.title.position =  "top",
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6),
                axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())+
  ggtitle("Number of detected genes")

gridExtra::grid.arrange(f2, f1, ncol = 3)

```

### Sequencing depth

```{r}
thr_sum <- 1000000
thr_sum <- 0
#Define outliers with low detection level
#outliers <- colnames(sce)[isOutlier(sce$sum, type = "lower")]
outliers <- colnames(sce)[sce$sum < thr_sum]

sce$outlier_sum <- FALSE
sce$outlier_sum[sce$sample %in% outliers] <- TRUE

f1 <- as_tibble(colData(sce)) %>% 
  ggplot(aes(x = 1, y = sum)) +
  geom_jitter(aes(x = 1, y = sum, color = outlier_sum)) +
  geom_violin(alpha = 0.5, outliers = F) +
  geom_hline(yintercept = thr_sum, linetype = "dashed") +
  theme(legend.position = "bottom",
        legend.title.position =  "top",
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6),
                axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  ggtitle("Total number of reads")

f3 <- as_tibble(colData(sce)) %>% 
  ggplot(aes(x = 1, y = sum)) +
  geom_jitter(aes(x = 1, y = sum, color = day)) +
  geom_violin(alpha = 0.5, outliers = F) +
  #geom_hline(yintercept = thr_sum, linetype = "dashed") +
  theme(legend.position = "bottom",
        legend.title.position =  "top",
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6),
                axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  ggtitle("Total number of reads")

# as_tibble(colData(sce)) %>% 
#   ggplot(aes(x = 1, y = detected)) +
#   geom_jitter(aes(x = 1, y = detected, color = source_name)) +
#   geom_violin(alpha = 0.5, outliers = F)

gridExtra::grid.arrange(f3, f1, ncol = 2)
```

### Ouliers identification

```{r}
outliers <- as_tibble(colData(sce)) %>% 
  filter(outlier_sum | outlier_detected) %>% 
  pull(sample)

sce$outlier <- FALSE
sce$outlier[sce$sample %in% outliers] <- TRUE

# Define outliers with low sum value
f1 <- as_tibble(colData(sce)) %>% 
  ggplot(aes(x = 1, y = detected)) +
  geom_jitter(aes(x = 1, y = detected, color = outlier)) +
  geom_violin(alpha = 0.5, outliers = F) +
  theme(legend.position = "bottom",
        legend.title.position =  "top",
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

f2 <- as_tibble(colData(sce)) %>% 
  ggplot(aes(x = 1, y = sum)) +
  geom_jitter(aes(x = 1, y = sum, color = outlier)) +
  geom_violin(alpha = 0.5, outliers = F) +
  theme(legend.position = "bottom",
        legend.title.position =  "top",
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

gridExtra::grid.arrange(f1, f2, ncol = 2)
```

=> Removed outlier cells

```{r}
table(outlier =sce$outlier)

sce <- sce[, !sce$outlier]
```

=> `r dim(sce)[2]` cells kept

```{r}
table(sce$day, sce$gender)
```

### No link between data's quality and sex factor

Checked that the data's quality is independant of sex factor:

```{r}
a <- as_tibble(colData(sce)) %>% 
  ggplot(aes(x = gender, y = detected)) +
  geom_jitter(aes(x = gender, y = detected, color = gender)) +
  geom_boxplot(alpha = 0.5, outliers = F) +
    scale_color_manual(values = list("M" = "steelblue", "F" = "deeppink")) +
  theme(legend.position = "bottom",
        legend.title.position =  "top",
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())+
  ggtitle("Number of detected genes")

b <- as_tibble(colData(sce)) %>% 
  ggplot(aes(x = gender, y = sum)) +
  geom_jitter(aes(x = gender, y = sum, color = gender)) +
  geom_boxplot(alpha = 0.5, outliers = F) +
scale_color_manual(values = list("M" = "steelblue", "F" = "deeppink")) +
  theme(legend.position = "bottom",
        legend.title.position =  "top",
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())+
  ggtitle("Sequencing Depth")

gridExtra::grid.arrange(a,b, ncol = 2)
```

### Normalisation

```{r}
## Normalisation
sce <- logNormCounts(sce, transform = "none")
sce <- logNormCounts(sce) 
```

```{r}
## Feature selection
hvg <- modelGeneVar(sce)
fit_hvg <- metadata(hvg)
# plot(fit_hvg$mean, fit_hvg$var,
#      xlab = "Mean of log-expression",
#      ylab = "Variance of log-expression",
#      col = ifelse(hvg$p.value < 0.1, "red", "black"))
# curve(fit_hvg$trend(x), col = "dodgerblue", add = TRUE, lwd = 2)

topHvg <- getTopHVGs(hvg, n = 500)

rowSubset(sce) <- topHvg

## PCA
set.seed(123)
sce <- fixedPCA(sce, rank = 100, subset.row = topHvg) # rank = 50 by default
plotPCA(sce, colour_by = "day", shape = "lineage")
plotPCA(sce, colour_by = "gender")
plotPCA(sce, colour_by = "lineage")
```

idem but focus on E4

```{r}
n_cells_per_individual <- enframe(table(sce$individual), name = "individual", value = "n_cells")

sce_reduced <- sce[, sce$day == "E4"]
hvg <- modelGeneVar(sce_reduced)
fit_hvg <- metadata(hvg)
# plot(fit_hvg$mean, fit_hvg$var,
#      xlab = "Mean of log-expression",
#      ylab = "Variance of log-expression",
#      col = ifelse(hvg$p.value < 0.1, "red", "black"))
# curve(fit_hvg$trend(x), col = "dodgerblue", add = TRUE, lwd = 2)

topHvg <- getTopHVGs(hvg, n = 500)

rowSubset(sce_reduced) <- topHvg

# plotPCA(sce_reduced, colour_by = "day", shape = "lineage")
# plotPCA(sce_reduced, colour_by = "gender")

selected <- n_cells_per_individual %>% filter(n_cells> 5) %>% pull(individual)
plotPCA(sce_reduced[, sce_reduced$individual %in% selected], colour_by = "individual") 
```

idem but focus on E5

```{r}
n_cells_per_individual <- enframe(table(sce$individual), name = "individual", value = "n_cells")

sce_reduced <- sce[, sce$day == "E5"]
hvg <- modelGeneVar(sce_reduced)
fit_hvg <- metadata(hvg)
# plot(fit_hvg$mean, fit_hvg$var,
#      xlab = "Mean of log-expression",
#      ylab = "Variance of log-expression",
#      col = ifelse(hvg$p.value < 0.1, "red", "black"))
# curve(fit_hvg$trend(x), col = "dodgerblue", add = TRUE, lwd = 2)

topHvg <- getTopHVGs(hvg, n = 500)

rowSubset(sce_reduced) <- topHvg

# 
# plotPCA(sce_reduced, colour_by = "day", shape = "lineage")
# plotPCA(sce_reduced, colour_by = "gender")

selected <- n_cells_per_individual %>% filter(n_cells> 10) %>% pull(individual)
plotPCA(sce_reduced[, sce_reduced$individual %in% selected], colour_by = "individual") 
plotPCA(sce_reduced[, sce_reduced$individual %in% selected], colour_by = "gender") 
plotPCA(sce_reduced[, sce_reduced$individual %in% selected], colour_by = "sum") 
plotPCA(sce_reduced[, sce_reduced$individual %in% selected], colour_by = "detected") 
```

idem but focus on E6

```{r}
n_cells_per_individual <- enframe(table(sce$individual), name = "individual", value = "n_cells")

sce_reduced <- sce[, sce$day == "E6"]
hvg <- modelGeneVar(sce_reduced)
fit_hvg <- metadata(hvg)
# plot(fit_hvg$mean, fit_hvg$var,
#      xlab = "Mean of log-expression",
#      ylab = "Variance of log-expression",
#      col = ifelse(hvg$p.value < 0.1, "red", "black"))
# curve(fit_hvg$trend(x), col = "dodgerblue", add = TRUE, lwd = 2)

topHvg <- getTopHVGs(hvg, n = 500)

rowSubset(sce_reduced) <- topHvg

# plotPCA(sce_reduced, colour_by = "day", shape = "lineage")
# plotPCA(sce_reduced, colour_by = "gender")

selected <- n_cells_per_individual %>% filter(n_cells> 10) %>% pull(individual)
plotPCA(sce_reduced[, sce_reduced$individual %in% selected], colour_by = "individual") 
plotPCA(sce_reduced[, sce_reduced$individual %in% selected], colour_by = "gender") 
plotPCA(sce_reduced[, sce_reduced$individual %in% selected], colour_by = "lineage") 
plotPCA(sce_reduced[, sce_reduced$individual %in% selected], colour_by = "sum") 
plotPCA(sce_reduced[, sce_reduced$individual %in% selected], colour_by = "detected") 
```

idem but focus on E7

```{r}
n_cells_per_individual <- enframe(table(sce$individual), name = "individual", value = "n_cells")

sce_reduced <- sce[, sce$day == "E7"]
hvg <- modelGeneVar(sce_reduced)
fit_hvg <- metadata(hvg)
# plot(fit_hvg$mean, fit_hvg$var,
#      xlab = "Mean of log-expression",
#      ylab = "Variance of log-expression",
#      col = ifelse(hvg$p.value < 0.1, "red", "black"))
# curve(fit_hvg$trend(x), col = "dodgerblue", add = TRUE, lwd = 2)

topHvg <- getTopHVGs(hvg, n = 500)

rowSubset(sce_reduced) <- topHvg

#plotPCA(sce_reduced, colour_by = "day", shape = "lineage")
#plotPCA(sce_reduced, colour_by = "gender")

selected <- n_cells_per_individual %>% filter(n_cells> 10) %>% pull(individual)
plotPCA(sce_reduced[, sce_reduced$individual %in% selected], colour_by = "individual") 
plotPCA(sce_reduced[, sce_reduced$individual %in% selected], colour_by = "gender") 
plotPCA(sce_reduced[, sce_reduced$individual %in% selected], colour_by = "lineage") 
plotPCA(sce_reduced[, sce_reduced$individual %in% selected], colour_by = "sum") 
plotPCA(sce_reduced[, sce_reduced$individual %in% selected], colour_by = "detected") 
```

```{r}

table(sce$lineage, sce$day, sce$gender)
```

# CG genes Heatmaps

## exploratory Heatmaps

```{r}
legends_param <- list(
  labels_gp = gpar(col = "black", fontsize = 10),
  title_gp = gpar(col = "black", fontsize = 10, fontface = "bold"),
  simple_anno_size = unit(0.1, "cm"),
  row_names_gp = gpar(fontsize = 8),
  annotation_name_side = "left",
  border = FALSE,
  border_gp = gpar(lwd = 0.1),
  grid_width = unit(0.3, "cm"),
  grid_height = unit(0.3, "cm"),
  legend_height = unit(0.3, "cm"))

legend_colors <- c(
  "#5E4FA2", "#3288BD", "#66C2A5", "#ABDDA4", "#E6F598",
  "#FFFFBF", "#FEE08B", "#FDAE61", "#F46D43", "#D53E4F",
  "#9E0142")

ensembl_gene <- data.frame(gene = all_genes$external_gene_name)
rownames(ensembl_gene) <- all_genes$ensembl_gene_id

draw_embryo_exp_Petropoulos_dataset <- function(genes, 
                            detected_in_percent_of_cells = 0,
                            day = NULL,
                            font_size = 6,
                            gender = NULL,
                            scale = FALSE,
                            assayName = "normcounts",
                            scale_lims = NULL,
                            clustRow = TRUE,
                            h_width = 8, 
                            h_height = 8, 
                            show_right_annot = TRUE,
                            clust_method = "centroid"){

  # Top Annotations
  df_col <- colData(sce)
  df_col <- df_col[!is.na(df_col$gender),]
  if (!is.null(day)) df_col <- df_col[df_col$day %in% day, ]
  df_col$sex <- df_col$gender
  levels(df_col$sex) <- c("Male", "Female")
  df_col <- df_col[order(df_col$day, df_col$sex, df_col$individual, df_col$lineage),]
  df_col$group <- paste0(df_col$day, "_", df_col$gender)

  column_ha_sex = HeatmapAnnotation(
    sex = df_col$sex,
    col = list(sex = c("Female" = "deeppink", "Male" = "steelblue")),
        border = FALSE, 
    simple_anno_size = unit(0.4, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

    column_ha_individual = HeatmapAnnotation(
    individual = df_col$individual,
    #col = list(individual = c("Female" = "deeppink", "Male" = "steelblue")),
        border = FALSE, 
    simple_anno_size = unit(0.4, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

    column_ha_lineage = HeatmapAnnotation(
    lineage = df_col$lineage,
    #col = list(individual = c("Female" = "deeppink", "Male" = "steelblue")),
        border = FALSE, 
    simple_anno_size = unit(0.4, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

    column_ha_day = HeatmapAnnotation(
    day = df_col$day,
    #col = list(Stage = c("ICM" = "gray", "TE" = "gray10")),
    border = FALSE, 
    simple_anno_size = unit(0.4, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

    column_ha_sum = HeatmapAnnotation(
    sum = df_col$sum,
        border = FALSE, 
    simple_anno_size = unit(0.4, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

    column_ha_detected = HeatmapAnnotation(
    detected = df_col$detected,
        border = FALSE, 
    simple_anno_size = unit(0.4, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

  n <- which(assayNames(sce) == assayName)
  mat <- assay(sce, i = n)[genes, ]

  # Show only genes detected in more than x% of cells
  mat <- mat[rowSums(mat > 0) >= dim(mat)[2]*detected_in_percent_of_cells,]
  # gene_type <- CT_genes_types %>% 
  #   filter(gene %in% genes) %>% 
  #   mutate(gene = factor(gene, levels = genes)) %>% 
  #   arrange(gene)

  if (scale) {
    mat <- t(scale(t(mat), center = FALSE, scale = TRUE))
    assayName = "scaled_exp"
    }

  if (is.null(scale_lims)) scale_lims <- c(0, quantile(mat, .9))
  heatmap_colors <- colorRamp2(
    seq(scale_lims[1], scale_lims[2], length = 11), legend_colors)

  lines <- tibble(gene = rownames(mat)) %>% 
    mutate(group = case_when(gene %in% CT_genes$external_gene_name ~ "X-linked CG MethDep",
                             !gene %in% CT_genes$external_gene_name ~ "Control")) %>% 
    mutate(group = factor(group, levels = c("X-linked CG MethDep", "Control")))

  ht1 <- Heatmap(mat[, rownames(df_col)],
                  name = assayName,
                 clustering_method_rows = "ward.D2",
                 clustering_method_columns = "ward.D2",
                 column_split = df_col$group,
                 row_split = lines$group,
                 row_gap = unit(c(1), "mm"),
                 col = heatmap_colors,
                 cluster_rows = clustRow,
                 cluster_columns = FALSE,
                 cluster_row_slices = FALSE,
                 cluster_column_slices = FALSE,
                 show_row_names = TRUE,
                 show_column_names = FALSE,
                 show_row_dend = FALSE,
                 column_names_gp = gpar(fontsize = 8),
                 column_names_centered = TRUE,
                 row_names_gp = gpar(fontsize = font_size),
                 row_names_side = "left",
                 border = TRUE,
                 border_gp = gpar(lwd = 0.5),
                 column_names_side = c("bottom"),
                 column_names_rot = 90,
                 column_title_gp = gpar(fontsize = 8),
                 row_title_gp = gpar(fontsize = 0),
                 top_annotation = c(column_ha_day,
                                    column_ha_sex,
                                    column_ha_detected,
                                    column_ha_lineage,
                                    column_ha_sum,
                                    column_ha_individual,
                                    # column_ha_sample,
                                    gap = unit(0.1, "mm")),
                 use_raster = TRUE,
                 raster_device = "CairoPNG",
                 raster_quality = 10,
                 heatmap_legend_param = legends_param,
                 width = unit(h_width, "cm"),
                 height = unit(h_height, "cm"))

  draw(ht1, merge_legend = TRUE)
}

my_genes <- all_genes %>% 
  filter(CT_gene_type == "CT_gene") %>% 
  filter(X_linked, regulated_by_methylation) %>% 
  filter(external_gene_name %in% rownames(sce))

controls <- all_genes %>% 
  filter(external_gene_name %in% c(#"GAPDH", "ACTB", "POU5F1", "XIST", 
    "DDX3Y", "RPS4Y1", "EIF1AY"))

X_linked <- all_genes %>% 
  filter(chr == "X") %>% 
  filter(external_gene_name %in% rownames(sce)) %>% 
  filter(!external_gene_name %in% CT_genes$external_gene_name) %>% 
  pull(external_gene_name)

chr1_linked <- all_genes %>% 
  filter(chr == 1) %>% 
  filter(external_gene_name %in% rownames(sce)) %>% 
  filter(!external_gene_name %in% CT_genes$external_gene_name) %>% 
  pull(external_gene_name) 
```

```{r}

draw_embryo_exp_Petropoulos_dataset <- function(genes, 
                            detected_in_percent_of_cells = 0,
                            day = NULL,
                            font_size = 6,
                            gender = NULL,
                            scale = FALSE,
                            assayName = "normcounts",
                            scale_lims = NULL,
                            clustRow = TRUE,
                            h_width = 8, 
                            h_height = 8, 
                            show_right_annot = TRUE,
                            clust_method = "centroid"){

  # Top Annotations
  df_col <- colData(sce)
  df_col <- df_col[!is.na(df_col$gender),]
  if (!is.null(day)) df_col <- df_col[df_col$day %in% day, ]
  df_col$sex <- df_col$gender
  levels(df_col$sex) <- c("Male", "Female")
  df_col <- df_col[order(df_col$day, df_col$sex, df_col$individual, df_col$lineage),]
  df_col$group <- paste0(df_col$day, "_", df_col$gender)

  column_ha_sex = HeatmapAnnotation(
    sex = df_col$sex,
    col = list(sex = c("Female" = "deeppink", "Male" = "steelblue")),
        border = FALSE, 
    simple_anno_size = unit(0.4, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

    column_ha_individual = HeatmapAnnotation(
    individual = df_col$individual,
    #col = list(individual = c("Female" = "deeppink", "Male" = "steelblue")),
        border = FALSE, 
    simple_anno_size = unit(0.4, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

    column_ha_lineage = HeatmapAnnotation(
    lineage = df_col$lineage,
    #col = list(individual = c("Female" = "deeppink", "Male" = "steelblue")),
        border = FALSE, 
    simple_anno_size = unit(0.4, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

    column_ha_day = HeatmapAnnotation(
    day = df_col$day,
    #col = list(Stage = c("ICM" = "gray", "TE" = "gray10")),
    border = FALSE, 
    simple_anno_size = unit(0.4, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

    column_ha_sum = HeatmapAnnotation(
    sum = df_col$sum,
        border = FALSE, 
    simple_anno_size = unit(0.4, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

    column_ha_detected = HeatmapAnnotation(
    detected = df_col$detected,
        border = FALSE, 
    simple_anno_size = unit(0.4, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

  n <- which(assayNames(sce) == assayName)
  mat <- assay(sce, i = n)[genes, ]

  # Show only genes detected in more than x% of cells
  mat <- mat[rowSums(mat > 0) >= dim(mat)[2]*detected_in_percent_of_cells,]
  # gene_type <- CT_genes_types %>% 
  #   filter(gene %in% genes) %>% 
  #   mutate(gene = factor(gene, levels = genes)) %>% 
  #   arrange(gene)

  if (scale) {
    mat <- t(scale(t(mat), center = FALSE, scale = TRUE))
    assayName = "scaled_exp"
    }

  if (is.null(scale_lims)) scale_lims <- c(0, quantile(mat, .9))
  heatmap_colors <- colorRamp2(
    seq(scale_lims[1], scale_lims[2], length = 11), legend_colors)

  lines <- tibble(gene = rownames(mat)) %>% 
    mutate(group = case_when(gene %in% CT_genes$external_gene_name ~ "X-linked CG MethDep",
                             !gene %in% CT_genes$external_gene_name ~ "Control")) %>% 
    mutate(group = factor(group, levels = c("X-linked CG MethDep", "Control")))

  ht1 <- Heatmap(mat[, rownames(df_col)],
                  name = assayName,
                 clustering_method_rows = "ward.D2",
                 clustering_method_columns = "ward.D2",
                 column_split = df_col$group,
                 row_split = lines$group,
                 row_gap = unit(c(1), "mm"),
                 col = heatmap_colors,
                 cluster_rows = clustRow,
                 cluster_columns = FALSE,
                 cluster_row_slices = FALSE,
                 cluster_column_slices = FALSE,
                 show_row_names = TRUE,
                 show_column_names = FALSE,
                 show_row_dend = FALSE,
                 column_names_gp = gpar(fontsize = 8),
                 column_names_centered = TRUE,
                 row_names_gp = gpar(fontsize = font_size),
                 row_names_side = "left",
                 border = TRUE,
                 border_gp = gpar(lwd = 0.5),
                 column_names_side = c("bottom"),
                 column_names_rot = 90,
                 column_title_gp = gpar(fontsize = 8),
                 row_title_gp = gpar(fontsize = 0),
                 top_annotation = c(column_ha_day,
                                    column_ha_sex,
                                    # column_ha_sample,
                                    gap = unit(0.1, "mm")),
                 use_raster = TRUE,
                 raster_device = "CairoPNG",
                 raster_quality = 10,
                 heatmap_legend_param = legends_param,
                 width = unit(h_width, "cm"),
                 height = unit(h_height, "cm"))

  draw(ht1, merge_legend = TRUE)
}

draw_embryo_exp_Petropoulos_dataset(c("MAGEC2", "MAGEB16", "VCX", "VCX2", "VCX3B", "DDX53", #"PAGE2", #"PAGE1", 
                                    controls$external_gene_name, #"ACTB", "GAPDH", 
                                    "RPS26", "XIST"),
                                    #day = "E6",
                                    clustRow = T, 
                            scale_lims = c(0, 300),
                            detected_in_percent_of_cells = 0)

draw_embryo_exp_Petropoulos_dataset(c(my_genes$external_gene_name, 
                                    controls$external_gene_name, #"ACTB", "GAPDH", 
                                    "RPS26", "XIST"),
                                    #day = "E3",
                                    clustRow = T, 
                            scale_lims = c(0, 300),
                            detected_in_percent_of_cells = 0) 
```

### E3 embryos

```{r}
draw_embryo_exp_Petropoulos_dataset <- function(genes, 
                            detected_in_percent_of_cells = 0,
                            day = NULL,
                            font_size = 6,
                            gender = NULL,
                            scale = FALSE,
                            assayName = "normcounts",
                            scale_lims = NULL,
                            clustRow = TRUE,
                            h_width = 8, 
                            h_height = 8, 
                            show_right_annot = TRUE,
                            clust_method = "centroid"){

  # Top Annotations
  df_col <- colData(sce)
  df_col <- df_col[!is.na(df_col$gender),]
  if (!is.null(day)) df_col <- df_col[df_col$day %in% day, ]
  df_col$sex <- df_col$gender
  levels(df_col$sex) <- c("Male", "Female")
  df_col <- df_col[order(df_col$day, df_col$sex, df_col$individual, df_col$lineage),]
  df_col$group <- paste0(df_col$day, "_", df_col$gender)

  column_ha_sex = HeatmapAnnotation(
    sex = df_col$sex,
    col = list(sex = c("Female" = "deeppink", "Male" = "steelblue")),
        border = FALSE, 
    simple_anno_size = unit(0.4, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

    column_ha_individual = HeatmapAnnotation(
    individual = df_col$individual,
    #col = list(individual = c("Female" = "deeppink", "Male" = "steelblue")),
        border = FALSE, 
    simple_anno_size = unit(0.4, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

    column_ha_lineage = HeatmapAnnotation(
    lineage = df_col$lineage,
    #col = list(individual = c("Female" = "deeppink", "Male" = "steelblue")),
        border = FALSE, 
    simple_anno_size = unit(0.4, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

    column_ha_day = HeatmapAnnotation(
    day = df_col$day,
    #col = list(Stage = c("ICM" = "gray", "TE" = "gray10")),
    border = FALSE, 
    simple_anno_size = unit(0.4, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

    column_ha_sum = HeatmapAnnotation(
    sum = df_col$sum,
        border = FALSE, 
    simple_anno_size = unit(0.4, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

    column_ha_detected = HeatmapAnnotation(
    detected = df_col$detected,
        border = FALSE, 
    simple_anno_size = unit(0.4, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

  n <- which(assayNames(sce) == assayName)
  mat <- assay(sce, i = n)[genes, ]

  # Show only genes detected in more than x% of cells
  mat <- mat[rowSums(mat > 0) >= dim(mat)[2]*detected_in_percent_of_cells,]
  # gene_type <- CT_genes_types %>% 
  #   filter(gene %in% genes) %>% 
  #   mutate(gene = factor(gene, levels = genes)) %>% 
  #   arrange(gene)

  if (scale) {
    mat <- t(scale(t(mat), center = FALSE, scale = TRUE))
    assayName = "scaled_exp"
    }

  if (is.null(scale_lims)) scale_lims <- c(0, quantile(mat, .9))
  heatmap_colors <- colorRamp2(
    seq(scale_lims[1], scale_lims[2], length = 11), legend_colors)

  lines <- tibble(gene = rownames(mat)) %>% 
    mutate(group = case_when(gene %in% CT_genes$external_gene_name ~ "X-linked CG MethDep",
                             !gene %in% CT_genes$external_gene_name ~ "Control")) %>% 
    mutate(group = factor(group, levels = c("X-linked CG MethDep", "Control")))

  ht1 <- Heatmap(mat[, rownames(df_col)],
                  name = assayName,
                 clustering_method_rows = "ward.D2",
                 clustering_method_columns = "ward.D2",
                 column_split = df_col$group,
                 row_split = lines$group,
                 row_gap = unit(c(1), "mm"),
                 col = heatmap_colors,
                 cluster_rows = clustRow,
                 cluster_columns = FALSE,
                 cluster_row_slices = FALSE,
                 cluster_column_slices = FALSE,
                 show_row_names = TRUE,
                 show_column_names = FALSE,
                 show_row_dend = FALSE,
                 column_names_gp = gpar(fontsize = 8),
                 column_names_centered = TRUE,
                 row_names_gp = gpar(fontsize = font_size),
                 row_names_side = "left",
                 border = TRUE,
                 border_gp = gpar(lwd = 0.5),
                 column_names_side = c("bottom"),
                 column_names_rot = 90,
                 column_title_gp = gpar(fontsize = 8),
                 row_title_gp = gpar(fontsize = 0),
                 top_annotation = c(column_ha_day,
                                    column_ha_sex,
                                    column_ha_detected,
                                    column_ha_lineage,
                                    column_ha_sum,
                                    column_ha_individual,
                                    # column_ha_sample,
                                    gap = unit(0.1, "mm")),
                 use_raster = TRUE,
                 raster_device = "CairoPNG",
                 raster_quality = 10,
                 heatmap_legend_param = legends_param,
                 width = unit(h_width, "cm"),
                 height = unit(h_height, "cm"))

  draw(ht1, merge_legend = TRUE)
}
```

Focus on selected CG genes :

```{r}
draw_embryo_exp_Petropoulos_dataset(c("MAGEC2", "MAGEB16", "VCX", "VCX2", "VCX3B", "DDX53", #"PAGE2", #"PAGE1", 
                                    controls$external_gene_name, "ACTB", "GAPDH", "RPS26", "XIST"),
                                    day = "E3",
                                    clustRow = T, 
                            scale_lims = c(0, 300),
                            detected_in_percent_of_cells = 0) 
```

All X-linked MethDep CG genes :

```{r}
draw_embryo_exp_Petropoulos_dataset(c(my_genes$external_gene_name, controls$external_gene_name, "RPS26"),
                                    day = "E3",
                                    clustRow = T, 
                            scale_lims = c(0, 300),
                            detected_in_percent_of_cells = 0)
```

All X-linked other than CG genes:

```{r}
draw_embryo_exp_Petropoulos_dataset(c(X_linked),
                                    day = "E3",
                                    clustRow = T, 
                            #scale_lims = c(0, 500),
                            scale = T,
                            font_size = 6,
                            detected_in_percent_of_cells = 0.5) 
```

chr1-linked genes:

```{r}
draw_embryo_exp_Petropoulos_dataset(c(chr1_linked),
                                    day = "E3",
                                    clustRow = T, 
                            #scale_lims = c(0, 1),
                            scale = T,
                            font_size = 6,
                            detected_in_percent_of_cells = 0.8) 
```

Some Female cells (usually corresponding to the same embryo) express no CG genes,

but they also express some control genes at lower expression level

- bad quality embryos ? (unlikely as no link with number of detected genes or sequencing depth)

- expression embryo-specific?

- linked with ZGA that has not occured yet?

### E4 embryos

Focus on selected CG genes :

```{r}
draw_embryo_exp_Petropoulos_dataset(c("MAGEC2", "MAGEB16", "VCX", "VCX2", "VCX3B", "DDX53", #"PAGE2", #"PAGE1", 
                                    controls$external_gene_name, "ACTB", "GAPDH", "RPS26", "XIST"),
                                    day = "E4",
                                    clustRow = T, 
                            scale_lims = c(0, 300),
                            detected_in_percent_of_cells = 0) 
```

All X-linked MethDep CG genes :

```{r}
draw_embryo_exp_Petropoulos_dataset(c(my_genes$external_gene_name, controls$external_gene_name, "RPS26"),
                                    day = "E4",
                                    clustRow = T, 
                            scale_lims = c(0, 300),
                            detected_in_percent_of_cells = 0)
```

All X-linked other than CG genes:

```{r}
draw_embryo_exp_Petropoulos_dataset(c(X_linked),
                                    day = "E4",
                                    clustRow = T, 
                            #scale_lims = c(0, 500),
                            scale = T,
                            font_size = 6,
                            detected_in_percent_of_cells = 0.5) 
```
chr1-linked genes:

```{r}
draw_embryo_exp_Petropoulos_dataset(c(chr1_linked),
                                    day = "E4",
                                    clustRow = T, 
                            #scale_lims = c(0, 1),
                            scale = T,
                            font_size = 6,
                            detected_in_percent_of_cells = 0.8) 
```

### E5 embryos

Focus on selected CG genes :

```{r}
draw_embryo_exp_Petropoulos_dataset(c("MAGEC2", "MAGEB16", "VCX", "VCX2", "VCX3B", "DDX53", #"PAGE2", #"PAGE1", 
                                    controls$external_gene_name, "ACTB", "GAPDH", "RPS26", "XIST"),
                                    day = "E5",
                                    clustRow = T, 
                            scale_lims = c(0, 300),
                            detected_in_percent_of_cells = 0) 
```

All X-linked MethDep CG genes :

```{r}
draw_embryo_exp_Petropoulos_dataset(c(my_genes$external_gene_name, controls$external_gene_name, "RPS26"),
                                    day = "E5",
                                    clustRow = T, 
                            scale_lims = c(0, 300),
                            detected_in_percent_of_cells = 0)
```

All X-linked other than CG genes:

```{r}
draw_embryo_exp_Petropoulos_dataset(c(X_linked),
                                    day = "E5",
                                    clustRow = T, 
                            #scale_lims = c(0, 500),
                            scale = T,
                            font_size = 6,
                            detected_in_percent_of_cells = 0.5) 
```

chr1-linked genes:

```{r}
draw_embryo_exp_Petropoulos_dataset(c(chr1_linked),
                                    day = "E5",
                                    clustRow = T, 
                            #scale_lims = c(0, 1),
                            scale = T,
                            font_size = 6,
                            detected_in_percent_of_cells = 0.8) 
```

### E6 embryos

Focus on selected CG genes :

```{r}
draw_embryo_exp_Petropoulos_dataset(c("MAGEC2", "MAGEB16", "VCX", "VCX2", "VCX3B", "DDX53", #"PAGE2", #"PAGE1", 
                                    controls$external_gene_name, "ACTB", "GAPDH", "RPS26", "XIST"),
                                    day = "E6",
                                    clustRow = T, 
                            scale_lims = c(0, 300),
                            detected_in_percent_of_cells = 0) 
```

All X-linked MethDep CG genes :

```{r}
draw_embryo_exp_Petropoulos_dataset(c(my_genes$external_gene_name, controls$external_gene_name, "RPS26"),
                                    day = "E6",
                                    clustRow = T, 
                            scale_lims = c(0, 300),
                            detected_in_percent_of_cells = 0)
```

All X-linked other than CG genes:

```{r}
draw_embryo_exp_Petropoulos_dataset(c(X_linked),
                                    day = "E6",
                                    clustRow = T, 
                            #scale_lims = c(0, 500),
                            scale = T,
                            font_size = 6,
                            detected_in_percent_of_cells = 0.5) 
```

chr1-linked genes:

```{r}
draw_embryo_exp_Petropoulos_dataset(c(chr1_linked),
                                    day = "E6",
                                    clustRow = T, 
                            #scale_lims = c(0, 1),
                            scale = T,
                            font_size = 6,
                            detected_in_percent_of_cells = 0.8) 
```

### E7 embryos

Focus on selected CG genes :

```{r}
draw_embryo_exp_Petropoulos_dataset(c("MAGEC2", "MAGEB16", "VCX", "VCX2", "VCX3B", "DDX53", #"PAGE2", #"PAGE1", 
                                    controls$external_gene_name, "ACTB", "GAPDH", "RPS26", "XIST"),
                                    day = "E7",
                                    clustRow = T, 
                            scale_lims = c(0, 300),
                            detected_in_percent_of_cells = 0) 
```

All X-linked MethDep CG genes :

```{r}
draw_embryo_exp_Petropoulos_dataset(c(my_genes$external_gene_name, controls$external_gene_name, "RPS26"),
                                    day = "E7",
                                    clustRow = T, 
                            scale_lims = c(0, 300),
                            detected_in_percent_of_cells = 0)
```

All X-linked other than CG genes:

```{r}
draw_embryo_exp_Petropoulos_dataset(c(X_linked),
                                    day = "E7",
                                    clustRow = T, 
                            #scale_lims = c(0, 500),
                            scale = T,
                            font_size = 6,
                            detected_in_percent_of_cells = 0.5) 
```

chr1-linked genes:

```{r}
draw_embryo_exp_Petropoulos_dataset(c(chr1_linked),
                                    day = "E7",
                                    clustRow = T, 
                            #scale_lims = c(0, 1),
                            scale = T,
                            font_size = 6,
                            detected_in_percent_of_cells = 0.8) 
```

# pseudo bulk

```{r}
sce$ind_lineage <- paste0(sce$individual, '_',  sce$lineage)
dds <- scuttle::summarizeAssayByGroup(sce, 
                            ids = colData(sce)[, c("ind_lineage")],
                            statistics = "sum")

coldata <- tibble(ind_lineage = dds$ids,
                  ncells = dds$ncells) %>% 
  left_join(as_tibble(colData(sce)) %>% 
              dplyr::select(ind_lineage, day, stage, gender, lineage) %>% unique()) %>% 
  as.data.frame()

# remove E3
coldata <- coldata[coldata$day %in% c("E4", "E5", "E6", "E7"),]

#table(coldata$gender, coldata$ind_lineage)

# Remove embryo with ambigous gender
#coldata <- coldata[!coldata$ambigous, ]
rownames(coldata) <- coldata$ind_lineage

# table(ncells = coldata$ncells, day = coldata$day)
# table(ncells = coldata$ncells, day = coldata$day, coldata$gender)

number_cells_thr <- 3
n <- length(unique(dds$ids))
as_tibble(colData(dds)) %>% 
  mutate(day = substr(start = 1, stop = 2, ids),
         embryo = gsub("_.*", x = ids, replace = '' )) %>% 
  ggplot(aes( x = day, y = ncells, color = day)) +
  geom_jitter()  +
 # scale_color_manual(values = sample(scales::hue_pal()(n)))+
 theme(legend.position = "none") +
  geom_hline(yintercept = number_cells_thr, linetype = "dashed")

```

Remove Individuals with less than `r number_cells_thr` cells

Removed day3 (ZGA not fully completed yet)

design = gender + lineage + day

```{r}

# Remove Individual with less than x cells
coldata <- coldata[coldata$ncells >= number_cells_thr,]
#table(ncells = coldata$ncells, day = coldata$day, coldata$gender, coldata$lineage)

dds <- dds[, coldata$ind_lineage]
dds$day <- coldata$day
dds$stage <- coldata$stage
dds$gender <- coldata$gender
dds$lineage <- coldata$lineage
dds$ambigous <- coldata$ambigous

colData(dds)$group <- paste0(dds$day, "_", dds$lineage, "_", dds$gender)

dds <- DESeqDataSetFromMatrix(countData = assay(dds[, rownames(colData(dds))]), 
                              colData = colData(dds),
                              design = ~ gender + lineage + day)

dds$day <- factor(dds$day)
dds$gender <- factor(dds$gender, c(levels = c("M", "F")))
table(dds$day, dds$gender)
table(dds$lineage, dds$day, dds$gender)

dds <- estimateSizeFactors(dds)
filtering_thr <- 10
keep <- rowSums(counts(dds, normalize = TRUE) >= filtering_thr ) >= 5
table(keep)
dds <- DESeq(dds[keep,])

#dds <- DESeq(dds)

as_tibble(colData(dds)) %>% 
  ggplot(aes(x = ids, y = sizeFactor, fill = gender)) +
  geom_col()+
  theme(axis.text.x = element_text(angle = 90, size = 5, hjust = 1, vjust = .5)) +
  facet_wrap(~ day, scales = "free_x", ncol = 4) 
```

## PCA

```{r}
## PCA
rld <- varianceStabilizingTransformation(dds)
rv <- rowVars(assay(rld))
selected <- order(rv, decreasing = TRUE)[seq_len(min(500, length(rv)))]
pca <- prcomp(t(assay(rld)[selected, ]))
percentVar <- pca$sdev^2/sum(pca$sdev^2)
percentVar_1 <- paste0("PC1: ", round(percentVar[1] * 100), "% variance")
percentVar_2 <- paste0("PC2: ", round(percentVar[2] * 100), "% variance")
percentVar_3 <- paste0("PC3: ", round(percentVar[3] * 100), "% variance")
percentVar_4 <- paste0("PC4: ", round(percentVar[4] * 100), "% variance")
pcaData <- as_tibble(pca$x, rownames = "samples") %>%
  dplyr::select(samples, "PC1", "PC2", "PC3", "PC4") %>%
  left_join(as_tibble(colData(dds), rownames = "samples"))

ggplot(pcaData, aes(x = PC1, y = PC2, col = lineage, shape = gender)) +
  geom_point(size = 2.5) +
  xlab(percentVar_1) +
  ylab(percentVar_2) +
  facet_wrap(~ day)

ggplot(pcaData, aes(x = PC1, y = PC2, col = sizeFactor, shape = gender)) +
  geom_point(size = 2.5) +
  scale_color_continuous(type = "viridis") +
  xlab(percentVar_1) +
  ylab(percentVar_2) +
  facet_wrap(~ day)

# ggplot(pcaData, aes(x = PC3, y = PC4, col = lineage, shape = gender)) +
#   geom_point(size = 2.5) +
#   xlab(percentVar_3) +
#   ylab(percentVar_4) +
#   facet_wrap(~ day)
```
=> Separation by day and by lineage

## Volcano

```{r}
load("~/cluster/CBIO-templates/ensembl_2_geneName/transcripts_infos.rda")
transcripts_infos <- transcripts_infos %>% filter(chromosome_name %in% c(1:22, "X", "Y", "MT")) %>% 
              dplyr::select(external_gene_name, chromosome_name) %>% unique()

res_dds <- results(dds, 
                   name = "gender_F_vs_M", 
                   independentFiltering = FALSE)
res_tbl <- as_tibble(res_dds, rownames = "external_gene_name") %>% 
  left_join(transcripts_infos) %>% 
  arrange(padj)

# res_dds <- results(dds, 
#                    name = "gender_female_vs_male", 
#                    independentFiltering = TRUE)

# res_shr <- lfcShrink(dds, coef = "gender_F_vs_M")
# res_shr_tbl <- as_tibble(res_shr, rownames = "external_gene_name") %>%
#   left_join(transcripts_infos) %>%
#   arrange(padj)

#save(res_tbl, file = "../tmp_data/res_tbl_gender_F_vs_M.rda")
#save(res_shr_tbl, file = "../tmp_data/res_shr_tbl_gender_F_vs_M.rda")

padj_thr <- 0.05
log2FC_thr <- 1

ggplot(res_tbl %>% filter(!is.na(padj)), aes(x = pvalue)) + 
    geom_histogram(bins = 20, boundary = 0) +
  ggtitle("pvalues histogram")

labeled_genes <- res_tbl %>% 
  filter(chromosome_name != "Y") %>% 
  filter(padj < padj_thr) %>% 
  filter(external_gene_name %in% CT_genes$external_gene_name) %>% 
  #filter(abs(log2FoldChange) > log2FC_thr ) %>% 
  arrange(padj) #%>% head(30)

res_tbl %>% 
  filter(!is.na(padj)) %>% 
  ggplot(aes(x= log2FoldChange, y = -log10(padj))) +
  geom_point(size = 0.5) +
    geom_point(data = res_tbl %>% 
               filter(chromosome_name == "X"), 
             aes(x= log2FoldChange, y = -log10(padj)), 
             color = "steelblue", size = 0.55) +
  geom_point(data = res_tbl %>% 
               filter(external_gene_name %in% CT_genes$external_gene_name), 
             aes(x= log2FoldChange, y = -log10(padj)), 
             color = "red", size = 0.55) +
  geom_point(data = res_tbl %>% 
               filter(chromosome_name == "Y"), 
             aes(x= log2FoldChange, y = -log10(padj)), 
             color = "black", size = 0.55) +
 # xlab("log2FoldChange") + 
  ylab("-log10(padj)") +
  ggrepel::geom_text_repel(data = res_tbl %>% filter(padj < 0.05 & log2FoldChange > 3),
                           aes(x = log2FoldChange, y = -log10(padj), label = external_gene_name),
                           size = 2,
                           segment.size = 0.1, max.overlaps = 30) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray70") +
  geom_hline(yintercept = -log10(padj_thr), linetype = "dashed", color = "gray70")
```

### Volcano plot for paper

```{r}
load("~/cluster/Packages/CTdata/eh_data/CT_genes.rda")

CTP_genes <- CT_genes %>% 
  filter(CT_gene_type == "CTP_gene")

CT_genes <- CT_genes %>% 
  filter(CT_gene_type == "CT_gene")

thr <- 1e-35
logFC_thr <- 8
res_tbl_modified <- res_tbl %>% 
  mutate(group = case_when(padj < thr ~ "padj_outlier",
                           abs(log2FoldChange) >= logFC_thr ~ "logFC_outlier",
                           padj > thr & abs(log2FoldChange) < logFC_thr ~ "others")) %>% 
  mutate(padj = ifelse(padj <= thr, thr, padj)) %>% 
  mutate(log2FoldChange = ifelse(log2FoldChange >= logFC_thr, logFC_thr, log2FoldChange)) %>% 
  filter(chromosome_name %in% c(1:22, "X", "Y")) %>% 
  mutate(chr = case_when(chromosome_name %in% c("X", "Y") ~ chromosome_name,
                         !chromosome_name %in% c("X", "Y") ~ "Autosome")) %>% 
  mutate(X_CG_MethDEp = case_when(external_gene_name %in% CT_genes$external_gene_name ~ TRUE,
                             external_gene_name %in% CT_genes$external_gene_name ~ FALSE)) %>% 
  filter(!is.na(padj)) %>% 
  filter(!is.na(group)) 
  #filter(abs(log2FoldChange) > 0.05)

labeled_X_genes <- res_tbl_modified %>% 
  filter(chr == "X") %>% 
  filter(abs(log2FoldChange) > 5 & padj < 0.00005) %>% 
  filter(!external_gene_name %in% CT_genes$external_gene_name) %>% 
  #filter(external_gene_name %in% CT_genes$external_gene_name) %>% 
  #filter(abs(log2FoldChange) > log2FC_thr ) %>% 
  arrange(padj)

labeled_CT_genes <- res_tbl_modified %>% 
  #filter(chr != "Y") %>% 
  filter(external_gene_name %in% CT_genes_X_met & padj < 0.0005 & log2FoldChange > 2.5) %>% 
  #filter(external_gene_name %in% CT_genes$external_gene_name) %>% 
  #filter(abs(log2FoldChange) > log2FC_thr ) %>% 
  arrange(padj)

X_sign_genes <- res_tbl_modified %>% 
  filter(chr == "X") %>% 
  filter(padj < padj_thr) %>% 
  arrange(padj) #%>% head(30)

Y_sign_genes <- res_tbl_modified %>% 
  filter(chr == "Y") %>% 
  filter(padj < padj_thr) %>% 
  filter(log2FoldChange < 0 ) %>% 
  arrange(padj) #%>% head(30)

set.seed = 123
fig <- res_tbl_modified %>% 
  ggplot(aes(x= log2FoldChange, y = -log10(padj))) +
  geom_point(data = res_tbl_modified,
             aes(x= log2FoldChange, y = -log10(padj), shape = group),
             size = 1, color = "gray80") + 
    geom_point(data = res_tbl_modified %>% 
               filter(external_gene_name %in% X_sign_genes$external_gene_name),
             aes(x= log2FoldChange, y = -log10(padj), shape = group),
             size = 1, color = "steelblue") +
  geom_point(data = res_tbl_modified %>% 
               filter(external_gene_name %in% Y_sign_genes$external_gene_name),
             aes(x= log2FoldChange, y = -log10(padj), shape = group),
             size = 1, color = "black") +
  geom_point(data = res_tbl_modified %>% 
               filter(external_gene_name %in% CT_genes_X_met),# & padj < padj_thr),
             aes(x= log2FoldChange, y = -log10(padj), shape = group),
             size = 1.2, color = "firebrick2") +
  scale_shape_manual(values = c(15, 19, 17))+
                       # c("padj_outlier" = 17,
                       #          "others" = 15,
                       #          "logFC_outlier" = 19))+
  xlab("log2FoldChange") + 
  ylab("-log10(padj)") +
  geom_vline(xintercept = 0, linetype = "dashed", 
             color = "black", linewidth = 0.2) +
  # geom_vline(xintercept = -1, linetype = "dashed",
  #            color = "black", linewidth = 0.2) +
  geom_hline(yintercept = -log10(padj_thr), 
             linetype = "dashed", color = "black", linewidth = 0.2) +
  ggrepel::geom_text_repel(data = subset(res_tbl_modified, external_gene_name %in% labeled_X_genes$external_gene_name),
                           aes(x = log2FoldChange, y = -log10(padj), label = external_gene_name),
                           size = 3, color = "steelblue",
                           segment.size = 0.1, max.overlaps = 10) +
   ggrepel::geom_text_repel(data = subset(res_tbl_modified, external_gene_name %in% labeled_CT_genes$external_gene_name),
                           aes(x = log2FoldChange, y = -log10(padj), label = external_gene_name),
                           size = 3, color = "firebrick2",
                           segment.size = 0.1, max.overlaps = 30) +
  xlim(-logFC_thr,logFC_thr) +
  ggtitle("Petropoulos dataset") +
  theme_bw() +
  theme(legend.position = "none",
        title = element_text(size = 10),
        panel.grid = element_blank()) 
fig

pdf(file = "../figs/FigS2_Volcano_Petropoulos_dataset.pdf", width = 4, height = 4)
fig

dev.off()

```

- Genes with a padjusted value < `r thr` are represented as triangles

- Genes with an absolute log2FC > `r logFC_thr` are represented as squares

- Red dots = X-linked MethDep genes

- Blue dots = X-linked genes not CG genes

- Black dots = Y-linked genes

- Gray dots = other genes

```{r}
# pdf(file = paste0("../figs/figs_CT_in_embryo/Fig_Female_vs_Male_volcano_Zhu.pdf"),
#     width = 5, height = 4)
# fig
# dev.off()
```

```{r}
# #idem but shrinked logFC:
# 
# res_shr_tbl %>%
#   filter(!is.na(padj)) %>%
#   ggplot(aes(x= log2FoldChange, y = -log10(padj))) +
#   geom_point(size = 0.5) +
#   geom_point(size = 0.5) +
#   geom_point(data = res_shr_tbl %>%
#                filter(chromosome_name == "X"),
#              aes(x= log2FoldChange, y = -log10(padj)),
#              color = "steelblue", size = 0.55) +
#   geom_point(data = res_shr_tbl %>%
#                filter(external_gene_name %in% CT_genes$external_gene_name),
#              aes(x= log2FoldChange, y = -log10(padj)),
#              color = "red", size = 0.55) +
# 
#   geom_point(data = res_shr_tbl %>%
#                filter(chromosome_name == "Y"),
#              aes(x= log2FoldChange, y = -log10(padj)),
#              color = "black", size = 0.55) +
#   # xlab("log2FoldChange") +
#   ylab("-log10(padj)") +
#   ggrepel::geom_text_repel(data = subset(res_shr_tbl, external_gene_name %in% CT_genes$external_gene_name),
#                            aes(x = log2FoldChange, y = -log10(padj), label = external_gene_name),
#                            size = 2,
#                            segment.size = 0.1, max.overlaps = 30) +
#   geom_vline(xintercept = 0, linetype = "dashed", color = "gray70") +
#   geom_hline(yintercept = -log10(padj_thr), linetype = "dashed", color = "gray70")
```

```{r}
ordered_samples <- as_tibble(colData(dds)) %>% 
  arrange(gender, day) %>% 
  pull(ids)
plot_counts <- function(my_gene, order = ordered_samples){
  as_tibble(counts(dds[my_gene,], normalize = T), rownames = 'gene') %>%
  gather(ids, counts, -gene) %>% 
  left_join(as_tibble(colData(dds))) %>% 
  mutate(sample = factor(ids, levels = order)) %>% 
 # mutate(gene = mapIds(org.Hs.eg.db, ENSEMBL, "SYMBOL", "ENSEMBL")) %>% 
  # mutate(Gene = case_when(is.na(gene) ~ ENSEMBL,
  #                         !is.na(gene) ~ gene)) %>% 
  ggplot(aes(x = sample, y = counts, fill = gender)) +
  geom_bar(stat = 'identity')+
  facet_wrap( ~ gene, scales = "free") +
  theme(axis.text.x = element_text(size = 6, angle = 90, vjust = 0.5, hjust = 1)) +
  xlab(element_blank()) +
  ylab("Normalized counts") 
}
plot_counts("VCX2")
plot_counts(c("MAGEC2"))
# plot_counts(res_tbl %>% filter(padj < 0.05) %>% arrange(desc(log2FoldChange)) %>% pull(external_gene_name) %>% head(16))
# plot_counts("MAGEC2")
# plot_counts("XIST")
# plot_counts("POU4F1")

plot_counts_by_day <- function(my_gene, order = ordered_samples){
  as_tibble(counts(dds[my_gene,], normalize = T), rownames = 'gene') %>%
  gather(ids, counts, -gene) %>% 
  left_join(as_tibble(colData(dds))) %>% 
  mutate(sample = factor(ids, levels = order)) %>% 
 # mutate(gene = mapIds(org.Hs.eg.db, ENSEMBL, "SYMBOL", "ENSEMBL")) %>% 
  # mutate(Gene = case_when(is.na(gene) ~ ENSEMBL,
  #                         !is.na(gene) ~ gene)) %>% 
  ggplot(aes(x = sample, y = counts, fill = gender)) +
    #scale_fill_manual(values = list("M" = "steelblue", "F" = "deeppink"))+
  geom_bar(stat = 'identity')+
  facet_wrap( ~ day, scales = "free") +
  theme(axis.text.x = element_text(size = 6, angle = 90, vjust = 0.5, hjust = 1)) +
  xlab(element_blank()) +
  ylab("Normalized counts") 
}

plot_counts_by_day("MAGEC2")
```

```{r}
# pdf(file = paste0("../figs/figs_CT_in_embryo/Fig_Female_vs_Male_volcano_petropoulos_E4_E7.pdf"),
#     width = 5, height = 4)
# fig
# dev.off()
```

Expression of all X-linked_methDep **significantly DE between females and males**

```{r}
my_genes <- labeled_CT_genes %>% 
  filter(log2FoldChange > 2)
plot_counts <- function(my_gene, order = ordered_samples){
  as_tibble(counts(dds[my_gene,], normalize = T), rownames = 'gene') %>%
  gather(ids, counts, -gene) %>% 
  left_join(as_tibble(colData(dds))) %>% 
  mutate(sample = factor(ids, levels = order)) %>% 
 # mutate(gene = mapIds(org.Hs.eg.db, ENSEMBL, "SYMBOL", "ENSEMBL")) %>% 
  # mutate(Gene = case_when(is.na(gene) ~ ENSEMBL,
  #                         !is.na(gene) ~ gene)) %>% 
  ggplot(aes(x = sample, y = counts, fill = gender)) +
  geom_bar(stat = 'identity')+
  facet_wrap( ~ gene, scales = "free") +
  theme(axis.text.x = element_text(size = 0, angle = 90, vjust = 0.5, hjust = 1)) +
  xlab(element_blank()) +
  ylab("Normalized counts") 
}
plot_counts(my_genes$external_gene_name)

```

## Expression heatmap

```{r}
legends_param <- list(
  labels_gp = gpar(col = "black", fontsize = 8),
  title_gp = gpar(col = "black", fontsize = 8, fontface = "bold"),
  simple_anno_size = unit(0.1, "cm"),
  row_names_gp = gpar(fontsize = 1),
  annotation_name_side = "left",
  border = FALSE,
  border_gp = gpar(lwd = 0.1),
  grid_width = unit(0.1, "cm"),
  grid_height = unit(0.1, "cm"),
  legend_height = unit(0.1, "cm"))

legend_colors <- c(
  "#5E4FA2", "#3288BD", "#66C2A5", "#ABDDA4", "#E6F598",
  "#FFFFBF", "#FEE08B", "#FDAE61", "#F46D43", "#D53E4F",
  "#9E0142")

ensembl_gene <- data.frame(gene = all_genes$external_gene_name)
rownames(ensembl_gene) <- all_genes$ensembl_gene_id

draw_embryo_exp_Petropoulos_dataset <- function(genes, 
                                                font_size = 6,
                                                scale = FALSE,
                                                log_transform = TRUE,
                                                scale_lims = NULL,
                                                clustRow = TRUE,
                                                h_width = 8, 
                                                h_height = 8, 
                                                show_right_annot = TRUE,
                                                clust_method = "centroid"){

  # Top Annotations
  df_col <- colData(dds)
  df_col$sex <- df_col$gender
  levels(df_col$sex) <- c("Male", "Female")
  df_col <- df_col[order(df_col$sex, df_col$sex),]
  df_col$day_sex <- paste0(df_col$day, '_', df_col$gender)

  column_ha_sex = HeatmapAnnotation(
    sex = df_col$sex,
    col = list(sex = c("Female" = "deeppink", "Male" = "steelblue")),
        border = FALSE, 
    simple_anno_size = unit(0.4, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

  column_ha_day = HeatmapAnnotation(
    day = df_col$day,
    col = list(day = c("E4" = "aliceblue", "E4" = "moccasin", "E5" = "orange3", "E6" = "firebrick3", "E7" = "brown4")),
       # border = FALSE, 
    simple_anno_size = unit(0.4, "cm"), 
    annotation_legend_param = legends_param,
    annotation_name_gp = gpar(fontsize = 0))

  genes <- genes[genes %in% rownames(dds)]

  mat <- counts(dds[genes, ], normalize = TRUE)
  #rownames(mat) <- ensembl_gene[genes, ]

  units = "Counts"
  if (log_transform) {
    mat <- log1p(mat)
    units = "log1p(Counts)"
    }
  if (scale) {
    mat <- t(scale(t(mat), center = FALSE, scale = TRUE))
    units = "scaled_exp"
    }

  # gene_type <- CT_genes_types %>% 
  #   filter(gene %in% genes) %>% 
  #   mutate(gene = factor(gene, levels = genes)) %>% 
  #   arrange(gene)

  if (is.null(scale_lims)) scale_lims <- c(0, max(mat))
  heatmap_colors <- colorRamp2(
    seq(scale_lims[1], scale_lims[2], length = 11), legend_colors)

  ht1 <- Heatmap(mat[, rownames(df_col)],
                  name = units,
                 clustering_method_rows = "ward.D",
                 clustering_method_columns = "ward.D",
                 column_split = df_col$day,
                 row_gap = unit(c(1), "mm"),
                 col = heatmap_colors,
                 cluster_rows = clustRow,
                 cluster_columns = FALSE,
                 cluster_row_slices = FALSE,
                 cluster_column_slices = FALSE,
                 show_row_names = TRUE,
                 show_column_names = FALSE,
                 show_row_dend = FALSE,
                 column_names_gp = gpar(fontsize = 8),
                 column_names_centered = TRUE,
                 row_names_gp = gpar(fontsize = font_size),
                 row_names_side = "left",
                 border = TRUE,
                 border_gp = gpar(lwd = 0.5),
                 column_names_side = c("bottom"),
                 column_names_rot = 90,
                 column_title_gp = gpar(fontsize = 12),
                 row_title_gp = gpar(fontsize = 0),
                 top_annotation = c(column_ha_sex,
                                    #column_ha_day,
                                    gap = unit(0.1, "mm")),
                 use_raster = TRUE,
                 raster_device = "CairoPNG",
                 raster_quality = 10,
                 heatmap_legend_param = legends_param,
                 width = unit(h_width, "cm"),
                 height = unit(h_height, "cm"))

  draw(ht1, merge_legend = TRUE)
}
```

Expression Heatmap of all X-linked_methDep

```{r}

draw_embryo_exp_Petropoulos_dataset(CT_genes_X_met, clustRow = T, log_transform = FALSE, scale_lims = c(0,200)) 
```

```{r}
# #idem but log_transformed:
# draw_embryo_exp_Petropoulos_dataset(CT_genes_X_met, clustRow = T, log_transform = TRUE) 
```

Expression Heatmap of all X-linked_methDep **significantly DE between females and males**

and with logFC > 3

```{r}
my_genes <- labeled_CT_genes %>% 
  filter(log2FoldChange > 3)

draw_embryo_exp_Petropoulos_dataset(my_genes$external_gene_name, 
                                    clustRow = T, 
                                    h_width = 8, 
                                    h_height = 4, 
                                    log_transform = F, 
                                    scale_lims = c(0, 200))
```

```{r}
as_tibble(counts(dds[my_genes$external_gene_name,]), rownames = "gene") %>% 
  pivot_longer(names_to = "embryo", values_to = "counts", -gene) %>% 
  left_join(as_tibble(colData(dds), rownames = "embryo")) %>% 
  ggplot(aes(x = group, y = log1p(counts), color = gender)) +
  scale_color_manual(values = c("M" = "blue", "F" = "deeppink")) +
  #geom_jitter(size = 1) +
  geom_boxplot(alpha = 0, outliers = FALSE) +
  facet_wrap(~ gene, scales = "free_y") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90))

as_tibble(counts(dds["MAGEC2",]), rownames = "gene") %>% 
  pivot_longer(names_to = "embryo", values_to = "counts", -gene) %>% 
  left_join(as_tibble(colData(dds), rownames = "embryo")) %>% 
  ggplot(aes(x = group, y = log1p(counts), color = gender)) +
  scale_color_manual(values = c("M" = "blue", "F" = "deeppink")) +
  #geom_jitter(size = 1) +
  geom_boxplot(alpha = 0, outliers = FALSE) +
  facet_wrap(~ gene, scales = "free_y") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, size = 5))
```

```{r}
sessionInfo()
```

### 2025-03-26 6.CT_methylation_in_early_embryos.Rmd
Add files via upload

@@
 
+```
+## t.tests
+
+### Gametes stage
+
+```{r}
+t.test(bilan_selected %>% 
+  filter(Genotype == "XX") %>% 
+  filter(Stage == "MII Oocyte") %>% 
+  pull(mean_met),
+  bilan_selected %>% 
+  filter(Genotype == "X") %>% 
+  filter(Stage == "Sperm") %>% 
+  pull(mean_met))
+```
+
+### at zygote stage
+
+```{r}
+stage <- "Zygote"
+t.test(bilan_selected %>% 
+  filter(Genotype == "XX") %>% 
+  filter(Stage == stage) %>% 
+  pull(mean_met),
+  bilan_selected %>% 
+  filter(Genotype == "XY") %>% 
+  filter(Stage == stage) %>% 
+  pull(mean_met))
+```
+
+### at 2-cell stage
+
+```{r}
+stage <- "2-cell"
+t.test(bilan_selected %>% 
+  filter(Genotype == "XX") %>% 
+  filter(Stage == stage) %>% 
+  pull(mean_met),
+  bilan_selected %>% 
+  filter(Genotype == "XY") %>% 
+  filter(Stage == stage) %>% 
+  pull(mean_met))
+```
+
+### at 4-cell stage
+
+```{r}
+stage <- "4-cell"
+t.test(bilan_selected %>% 
+  filter(Genotype == "XX") %>% 
+  filter(Stage == stage) %>% 
+  pull(mean_met),
+  bilan_selected %>% 
+  filter(Genotype == "XY") %>% 
+  filter(Stage == stage) %>% 
+  pull(mean_met))
+```
+
+### at 8-cell stage
+
+```{r}
+stage <- "8-cell"
+t.test(bilan_selected %>% 
+  filter(Genotype == "XX") %>% 
+  filter(Stage == stage) %>% 
+  pull(mean_met),
+  bilan_selected %>% 
+  filter(Genotype == "XY") %>% 
+  filter(Stage == stage) %>% 
+  pull(mean_met))
+```
+### at Morula stage
+
+```{r}
+stage <- "Morula"
+t.test(bilan_selected %>% 
+  filter(Genotype == "XX") %>% 
+  filter(Stage == stage) %>% 
+  pull(mean_met),
+  bilan_selected %>% 
+  filter(Genotype == "XY") %>% 
+  filter(Stage == stage) %>% 
+  pull(mean_met))
+```
+
+### at Blastocyst stage
+
+```{r}
+stage <- "Blastocyst"
+t.test(bilan_selected %>% 
+  filter(Genotype == "XX") %>% 
+  filter(Stage == stage) %>% 
+  pull(mean_met),
+  bilan_selected %>% 
+  filter(Genotype == "XY") %>% 
+  filter(Stage == stage) %>% 
+  pull(mean_met))
+```
+
+### at Post-implantation stage
+
+```{r}
+stage <- "Post-impl"
+t.test(bilan_selected %>% 
+  filter(Genotype == "XX") %>% 
+  filter(Stage == stage) %>% 
+  pull(mean_met),
+  bilan_selected %>% 
+  filter(Genotype == "XY") %>% 
+  filter(Stage == stage) %>% 
+  pull(mean_met))
 ```

### 2025-03-28 2.Database_Comparison.Rmd
Updated database comparison

--- 2024-10-04 Paper_code-Database_Comparison.Rmd
@@
 ---
-title: "Code used to generate paper results : Database comparison and CT characteristics"
-author: "Julie Devis"
-date: "2023-08-28"
+title: "Database comparison"
 output: 
@@
 ```{r load_data, warning=FALSE, message=FALSE}
-load("../CTdata/eh_data/all_genes.rda")
-load("../CTdata/eh_data/CT_genes.rda")
-
-CT_and_CTP_genes <- CT_genes
-CT_genes <- (filter(CT_genes, CT_gene_type == "CT_gene"))
+CT_and_CTP_genes <- CT_genes()
+CT_genes <- (filter(CT_genes(), CT_gene_type == "CT_gene"))
 
@@
 CT lists from other databases have been checked (using GTEx and our 
-`GTEx_expression()` funtion and GeneCards) in order to remove duplicated
+`GTEx_expression()` function and GeneCards) in order to remove duplicated
 gene names or deprecated ones and allow comparison between databases.
@@
 lost[lost$external_gene_name == "C9orf131", "Gene_Name"] <- "SPATA31G1"
+lost[lost$external_gene_name == "C17orf80", "Gene_Name"] <- "MTNAP1"
 
@@
 external_names_to_keep <- gene_synonym %>%
-   filter(external_synonym %in% missing_Bruggeman) %>%
-   filter(external_gene_name %in% all_genes$external_gene_name) %>% 
+   filter(external_synonym %in% missing_Bruggeman) %>% # take those in Bruggeman that are synonym
+   filter(external_gene_name %in% all_genes$external_gene_name) %>% # but that are actually in all_genes with another name
    mutate(Gene_Name = external_gene_name)
@@
                                 all_genes$external_gene_name])
+
 ```
 
-38 genes are not found with their names in all genes
+39 genes are not found with their names in all genes
 
@@
 transcript_prob <- lost %>% 
-  filter(testis_specificity == "testis_specific" |
-           testis_specificity == "testis_preferential") %>%
-  filter(TCGA_category == "activated" | TCGA_category == "multimapping_issue") %>% 
-  filter(CCLE_category == "activated") %>% 
-  dim()
+  pull(IGV_backbone) %>% 
+  table()
 ```
